### 2013-09-30 src2/bb/JosefPartI.tex
another approach

\section{Introduction}

Suppose you have just bought yourself a little programmable toy robot
and you want to design a small DSL in Haskell to program it. The robot
can turn, move forward and has a sensor to detect whether it is facing
an obstacle or not. Inspired by the robot language in \cite{} you come
up with a set of primitives shown in figure
\ref{fig:interface}. In particular, you would like to use the
do-notation to sequence programs.

\begin{figure} 
\begin{itemize} 
  \item \verb!move      :: Program ()! 
  \item \verb!turnLeft  :: Program ()!
  \item \verb!turnRight :: Program ()!
  \item \verb!sensor    :: Program Bool!
  \item \verb!cond      :: Program Bool -> Program () -> Program () -> Program ()!
  \item \verb!while     :: Program Bool -> Program () -> Program ()!
  \item \verb!instance Monad Program!
\end{itemize} 
\label{fig:interface} 
\caption{Proposed set of basic robot operations} 
\end{figure}

\emph{Example programs}

The next task is to implement this language. Compiling EDSLs like this
to something which the physical robot can execute is well know and
there are off-the-shelf techniques to use. But there is one quirk; how
does one generate an abstract syntax tree from a type which implements
the Monad interface?

Since you want to compile the language there must be an abstract
syntax tree representation of some sort of the language. The natural
thing is just to create a new type where each language construct has
its own constructor. But should you deal with the monadic constructs?
Well, the simplest solution would be to just add them to the type as well. The resulting type

\begin{figure}
\begin{verbatim}
data BoolE = Lit Bool
         | Var String
         | (:||:) BoolE BoolE
         | (:&&:)  BoolE BoolE
         | Not BoolE

data Program a where
  Move      :: Program ()
  TurnRight :: Program ()
  TurnLeft  :: Program ()

  Sensor    :: Program BoolE
  Cond      :: BoolE -> Program () -> Program () -> Program ()
  While     :: Program BoolE -> Program () -> Program ()

  Return    :: a -> Program a 
  Bind      :: Program a -> (a -> Program b) -> Program b
\end{verbatim}
\label{fig:program}
\caption{A data type for programs}
\end{figure}

\emph{More text}

In the rest of this paper we will see that this na\"ive method
actually can be made to work and that it's a practical, compositional
technique for reifying monads.

### 2013-09-30 src2/bb/body.tex
another approach

% Look Sebastian fisher (a play on regular expression) 
%   Really goes all the way

% Student Sven is working on a Robot programming DSL supervised by Prof. Kurt. 
% Sven is beginner Haskeller
% Kurt is knowledgible (experienced)

% Introduction (Presents problem) 
%  Robot language operations. 
%  Robot exists that run programs in a simple language.
%  want to run code in robot. 
%  Haskell DSL and Do notation 
%  Compilation necessary.

% Student has Haskell School of expression 
%   But need reifyable robot programs (evaluation not enough)

% Show example program early 
%   Sven is a good student and early expresses an example program in his thought 
%   out API.
%

% How to be able to use Do notation 
%  Adds Return and Bind naively to data structure.

% Student brings --CODE SNIPPET-- to Prof. Kurt.

% Now the problem, how can this work? 
%  Prof. is familiar with Gills blog post.

% Student brings --CODE SNIPPET-- (the compiler) to Prof.

% Surprise 
% Work on generalisations  and make the compositionality explicit
%

% What about the monad laws ? 
%  Prof. asks student to verify that Monad laws hold.

\section{Introduction}
\studname{} is a computer science student working in a project involving programming 
robots in a low-level imperative language. However, \studname{} has a budding 
interest in functional programming using Haskell and has read ``The Haskell 
school of expression'' \cite{HUDAK}. He gets the idea to implement 
a language for robot control embedded in Haskell. \studname{} realises that the 
capabilities of the robot hardware are very similar to those of the 
robots that Hudak evaluates graphically in a grid world. There is however 
a very important difference; the embedded language needs to be compiled into 
some form that is understood by the robot. 
%\studname{} in a computer science student with a budding interest in functional 
%programming using Haskell. \studname{} is developing  
%an embedded domain specific language (EDSL) for robot control. Inspired 
%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
%he can run on actual robot hardware. 
%with 
%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
%he sets out.

Guided by the capabilities of the target robot, \studname{} designs an API for 
robot programming. The robot can perform operations 
such as {\em move} that steps the robot forward and {\em turn} left and right. 
The robot also has the capability to execute program loops and conditionals and 
it has a forward facing {\tt sensor} with which it can query the world.

%The robot also has a forward facing {\tt sensor} with which it can query the 
%world and also the capabilities to execute program loops and conditionals. 
\pagebreak
The API that \studname{} designs is simple:

%\begin{figure} 
\begin{small}
\begin{verbatim} 
move      :: Program () 
turnLeft  :: Program () 
turnRight :: Program ()
sensor    :: Program Bool 
cond      :: Program Bool 
             -> Program () 
             -> Program () 
             -> Program () 
while     :: Program Bool 
             -> Program () 
             -> Program () 
\end{verbatim}
\end{small} 
%\label{fig:interface} 
%\caption{Proposed set of basic robot operations}
%\end{figure}

Now, \studname{} wants to express robot programs using the Haskell
{\em do} notation. The motivation behind this is the imperative look
and feel of the operations he identified and the potential to use all
the control structures in the {\em Control.Monad} library. For this, a
{\tt Monad} instance is needed.

To be able to create a first-order representation of a computation
described using monads, \studname{} needs to solve the problem of {\em
  reifying} monads. This is the story of how \studname{} came up with
a particularly simple solution to this problem.

%\emph{*** Josef, Think about writing here}

%\begin{minipage}[]{\linewidth}
%\vspace{5mm}

%\fbox{%
%\parbox{0.8\linewidth}{What \studname{} does not know at this time is that he has stumbled over the 
%problem of {\em Monad reification}. Monad reification is the observing of the shape of the computation and this is necessary in order to compile to some low level representation.}
%}
%\end{minipage}

\subsection{Example programs}

\studname{} is quite happy with his language design and decides to
write some programs to try it out.
%experiments 
%with what some of these robot controlling programs could look like once his 
%EDSL is finished.

The first program is called {\tt sMove}, which implements a safe move 
operation. This operation can be executed by the robot even though it is facing 
an obstacle, without risk of harming robot or obstacle. 
%.It feels like a good idea to \studname{} to check the sensor before trying to 
%move the robot forward. Otherwise something in either the world or the 
%robot might suffer damage.

\begin{small} 
\begin{verbatim}
sMove :: Program () 
sMove = cond sensor turnRight move 
\end{verbatim}
\end{small}

The second program moves the robot forward until it stands directly 
in front of an obstacle such as a wall.

\begin{small} 
\begin{verbatim}
moveToWall :: Program () 
moveToWall = while ((liftM not) sensor) move
\end{verbatim}
\end{small}

Based on these examples, \studname{} is quite satisfied with the design of his
language and turns to implementing it.

\subsection{Implementation and data structures}

Enthused by the prospect of compiling these programs to the target language and 
seeing some robot action, \studname{} goes to work on the data structures.

% Since an abstract syntax representation of the robot programs is desired,
Since the language is going to be compiled, \studname{} realises that the 
booleans in his language cannot be regular Haskell booleans. The booleans 
needs to be replaced by boolean typed expressions.

\pagebreak
\begin{small} 
\begin{verbatim}
type Name = String

data BoolE = Lit Bool
           | Var Name
           | (:||:) BoolE BoolE
           | (:&&:)  BoolE BoolE
           | Not BoolE  
\end{verbatim} 
\end{small}

Following this, the  operations that should go into the {\tt Program} data type 
feel straightforward.

\begin{small} 
\begin{verbatim}
data Program a where 
  Move      :: Program () 
  TurnLeft  :: Program () 
  TurnRight :: Program () 
  Sensor    :: Program BoolE
  Cond      :: Program BoolE 
               -> Program () 
               -> Program () 
  While     :: Program BoolE 
               -> Program () 
               -> Program () 
\end{verbatim} 
\end{small}

\studname{} continues by naively adding constructors for {\tt Return} and 
{\tt Bind} to the {\tt Program} data type.

\begin{small} 
\begin{verbatim}
data Program a where 
  ...
  Return :: a -> Program a 
  Bind   :: Program a 
            -> (a -> Program b)
            -> Program b 
\end{verbatim} 
\end{small}

Then he writes down the {\tt Monad} instance.

\begin{small} 
\begin{verbatim}
instance Monad Program where 
  return = Return 
  (>>=)  = Bind 
\end{verbatim} 
\end{small}

Proud of his accomplishments, \studname{} sends his Haskell module in an 
email to Prof. \docname{}. \studname{} knows that \docname{} is teaching an 
introductory Haskell course and should be able to provide feedback.

\subsection{Problem statement} 
\emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline

\noindent Prof. \docname{} notices an email from \studname{} in his inbox and 
opens it.

\vspace{5mm}

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Dear Professor \docname{}
\newline \newline
\noindent I am CS student working in a robot control project. Usually, we 
program our robots in C but I have developed an interest in Haskell 
programming and thought it'd be natural to try implementing an EDSL. 
I know you teach an FP course and thought I would ask for your input. 
Attached to this mail is a file containing an outline of the data types I 
want to use. Does this look sensible to you? \newline \newline

\noindent Thank you \newline
\noindent \studname{}
\end{minipage} 
}

%I have progressed some towards a robot programming language. Attached 
%to this mail is a first sketch of a {\tt Program a} data type for robot 
%programs.
\vspace{5mm}

Prof. \docname{} opens the attachment and takes a look at the data type. He 
particularly notices the constructors {\tt Return} and {\tt Bind} and the 
{\tt Monad} instance. He shakes his head at \studname{}'s naivet\'e and writes an 
email back:

%\vspace{5mm}

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello \studname{}
\newline \newline
\noindent I'm afraid your implementation of the monadic primitives can never 
work. The constructor {\tt Return} can take any arbitrary value that has 
nothing to do with the language you're designing. These values may be 
strings, binary trees or higher-order functions from zygomorphisms to 
histomorphisms. The same problem goes for {\tt Bind}; it has to be able 
to handle arbitrary values. It cannot possibly work! This is a very 
difficult problem and such a naive solution is bound to fail. 
\newline \newline
%% \noindent For hints towards a solution to this I suggest that you follow 
%% this \underline{link} \cite{GillBlog}. Another idea is that you refer 
%% to the article ``Generic monadic constructs for embedded languages'' 
%% \cite{Generic}.
%% \newline \newline

\noindent Prof. \docname{} \newline
\noindent Dept. of Computer Science and Engineering \newline
\noindent Chalmers University of Technology 
\end{minipage} 
}

\vspace{5mm}

% \section{Implementation continues}
% \emph{Back to \studname{}'s implementation efforts} \newline \newline

\noindent While waiting for feedback from \docname{}, \studname{} has 
carried on trying to implement a compiler for his language.

When \docname{}'s email does arrive, \studname{} is confused. He feels he 
has already managed to compile the {\tt Program} data type into a 
representation closer to that which is executed by the robot hardware.

He writes an apprehensive email back to Prof. \docname{}.

\vspace{5mm}

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Hello Prof. \docname{} 
\newline \newline
\noindent Thanks for your feedback. But I think it does work! Attached
to this email you find a file containing my attempt at compiling the
{\tt Program} data type into a representation closer to that which is
executed by our robots. \newline \newline

\noindent I have also attached two example programs that you can compile and 
then run in the graphical simulator (see figure~\ref{fig:programs}) \newline \newline
%\vspace{5mm}

%\includegraphics[width=.49\linewidth]{./spiral1}
%\includegraphics[width=.49\linewidth]{./spiral2}

%\vspace{5mm}

%\noindent And these two images are output from running the 
%{\tt followWall} (see figure~\ref{fig:followWall}) program. \newline

%\vspace{5mm}

%\includegraphics[width=.49\linewidth]{./wall1}
%\includegraphics[width=.49\linewidth]{./wall2}

%\vspace{5mm}

\noindent \studname{} 
\end{minipage} 
}

\FloatBarrier

\begin{figure*}
\begin{tabular}{c|c}
\begin{minipage}[]{0.45\columnwidth}
\begin{small}
\begin{verbatim}
spiralIn :: Int -> Program () 
spiralIn 0 = return () 
spiralIn n = do
  replicateM_ 2 $ do
    replicateM_ n move
    turnLeft
  spiralIn (n-1) 
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\columnwidth} 
\begin{minipage}[]{0.45\columnwidth}
\includegraphics[width=\columnwidth]{./bb/spiral1}
\end{minipage}
&
\begin{minipage}[]{0.45\columnwidth}
\begin{small}
\begin{verbatim} 
followWall :: Program () 
followWall = 
  while (return true) $ 
    cond checkLeft sMove $ 
            do turnLeft
               move

checkLeft :: Program BoolE     
checkLeft = do
  TurnLeft
  s <- Sensor
  TurnRight 
  return s
\end{verbatim}
\end{small}
\end{minipage}
\hspace{0.05\columnwidth}
\begin{minipage}[]{0.45\columnwidth}
\includegraphics[width=\columnwidth]{./bb/wall2}
\end{minipage}

\end{tabular}
\caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
to \docname{} in an email}
\label{fig:programs}
\end{figure*}

\section{Compilation of the monadic robot EDSL} 
\emph{\docname{} receives \studname{}'s latest email}\newline %\newline

\noindent \docname{} looks through the {\tt Compiler} module and finds a data type describing
a first-order representation of the robot language.

\begin{small} 
\begin{verbatim}
data Prg = PMove          
         | PTurnRight
         | PTurnLeft
         | PSensor Name  
         | PCond BoolE Prg Prg
         | PWhile Name Prg Prg 
         | PSeq Prg Prg
         | PSkip       
         | PAssign Name BoolE
\end{verbatim}
\end{small}

\begin{figure} 
\begin{small}
\begin{verbatim} 
newNameSupply :: NameSupply 
split2 :: NameSupply -> 
          (NameSupply,NameSupply) 
split3 :: NameSupply -> 
          (NameSupply,NameSupply,NameSupply)
supplyValue :: NameSupply -> Int 
\end{verbatim} 
\end{small}
\caption{The interface of the splittable name supply used in the implementation 
of the compile functions} 
\label{fig:namesupply}
\end{figure}

From the {\tt Prg} data type, \docname{} presumes that the {\tt PSeq}
constructor will be the result of compiling {\tt Bind} and that
{\tt PSensor} binds a variable which can be used in {\tt PAssign} and
{\tt PWhile}. The conclusion is that the representation looks
sensible, but he is very interested in seeing the {\tt Program a -> Prg}
transformation. The compile function assumes a splittable name
supply with the interface described in figure \ref{fig:namesupply}.

\pagebreak
\begin{small} 
\begin{verbatim}
runCompile :: Program a -> Prg
runCompile prg = snd $ compile s prg 
  where 
    s = newNameSupply

compile :: NameSupply -> Program a -> (a, Prg) 
compile s Move = ((),PMove)
compile s TurnRight = ((),PTurnRight)
compile s TurnLeft  = ((),PTurnLeft)
compile s Sensor    = (Var nom,PSensor nom)
  where
    v   = supplyValue s
    nom = "v" ++ show v 
compile s (Cond b p1 p2) = 
    ((),bp `PSeq` PCond b' p1' p2') 
  where
    (s1,s2,s3) = split3 s
    (b',bp)    = compile s1 b
    (a1,p1')   = compile s2 p1
    (a2,p2')   = compile s3 p2
compile s (While wp prg) = ((),PWhile nom nwp prg') 
  where
    (s1,s2,s3) = split3 s
    (b,wp')    = compile s1 wp
    (c,prg')   = compile s2 prg 
    nom        = "v" ++ (show $ supplyValue s3)
    nwp        = (wp' `PSeq` PAssign nom b) 
compile s (Return a) = (a,PSkip)
compile s (Bind pa f) = (b, prg1 `PSeq` prg2) 
  where
    (s1,s2)  = split2 s
    (a,prg1) = compile s1 pa
    (b,prg2) = compile s2 (f a) 
\end{verbatim}
\end{small}

%\begin{small} 
%\begin{verbatim}
%compile :: Program a -> Prg
%compile prg = snd $ compile' s prg 
%  where 
%    s = unsafePerformIO $ newEnumSupply
%    
%compile' :: Supply Int -> Program a -> (a, Prg) 
%compile' s Move = ((),PMove)
%compile' s TurnRight = ((),PTurnRight)
%compile' s TurnLeft  = ((),PTurnLeft)
%compile' s Sensor    = (Var nom,PSensor nom)
%  where
%    v = supplyValue s
%    nom = "v" ++ show v 
%compile' s (Cond b p1 p2) = ((), PCond b p1' p2') 
%  where
%    (s1,s2) = split2 s
%    (a1,p1') = compile' s1 p1
%    (a2,p2') = compile' s2 p2 
%compile' s (While wp prg) = ((),PWhile nom nwp prg') 
%  where
%    (s1,s2,s3) = split3 s
%    (b,wp') = compile' s1 wp
%    (c,prg') = compile' s2 prg 
%    nom  = "v" ++ (show $ supplyValue s3)
%    nwp = (wp' `PSeq` PAssign nom b) 
%compile' s (Return a) = (a,PSkip)
%compile' s (Bind pa f) = (b, prg1 `PSeq` prg2) 
%  where
%    (s1,s2) = split2 s
%    (a,prg1) = compile' s1 pa
%    (b,prg2) = compile' s2 (f a) 
%\end{verbatim}
%\end{small}

\docname{} studies the code in disbelief, saves the module and tries
it out on some examples. The code does indeed seem to work and
\docname{}'s disbelief is replaced with enthusiasm; this appears to be
a simple and convenient way to reify monads.
%What's more it does so in a 
%compositional way. The compilation of the {\tt Return} and {\tt Bind} cases 
%are completely independent of any other construct of the language.

\docname{} sends an email to \studname{}, inviting him to a meeting.

\section{Meeting in \docname{}'s office} %\section{Related Work} 
%\emph{Meeting in \docname{}'s office} \newline

%TODO Here I want to cast light on more related work 
%and have \docname{} and \studname{} compare their approach to some of the existing ones. 
%also they should start sketching out the Idea of using {\tt CompData} instead 
%of a la carte for showing the compositionality.

%\begin{minipage}[]{\linewidth}
\begin{dialogue}

\speak{\docname{}} Hello, come in.

\speak{\studname{}} Thanks.

\speak{\docname{}} So, about this robot language! You gave me a bit of a 
surprise there. I was entirely sure that what you did was impossible.

\speak{\studname{}} Oh, But I just did the first thing that came to mind. 
There was no deep thought behind it.

\speak{\docname{}} The problem is \direct{\docname{} approaches his whiteboard}
 when you try to reify the bind constructor of your representation.
\end{dialogue}

%\mbox{ 
\begin{small}
\begin{Verbatim}[commandchars=\\\{\}] 
Bind :: Program a -> \underline{(a -> Program b)} -> Program b 
\end{Verbatim}
\end{small}
%}

\begin{dialogue}
\speak{\docname{}} You need to come up with an element of type {\tt a} to 
pass to the function.

\speak{\studname{}} I didn't realise it was problem.

\speak{\docname{}} But it is! Your solution is that you are careful
about the return types of all operations in your language. They are
either of unit type or some type which can be guaranteed to be reified.
\direct{\docname{} scribbles on his whiteboard}

\end {dialogue}

\begin{small}
\begin{Verbatim}[commandchars=\\\{\}] 
Move :: Program \underline{()}

Sensor :: Program \underline{BoolE}
\end{Verbatim}
\end{small}

\begin{dialogue}
\speak{\docname{}} This is different from how such constructs are
normally implemented. If I had implemented the language I would have
given {\tt Sensor} the type {\tt Program Bool} so that I could write
an evaluator of type {\tt Program a -> a}, or perhaps {\tt Program a -> M a}.
But then adding a monad to the language and reifying it becomes much harder.

\speak{\studname{}} Yes, now that you mention it, I have seen that style used
for defining DSLs.

\speak{\docname{}} Having constructors like {\tt Sensor} return {\tt
  BoolE} instead of {\tt Bool} is a crucial part of why your
method works. It allows the {\tt compile} function to be written in
such a way that it can generate the return values statically and pass
them as arguments to {\tt Bind}. In a sense you're doing evaluation at
the same time as compilation. There's just enough evaluation to
remove the monadic constructs. It's a very neat trick!

\speak{\studname{}} Thank you.

\speak{\docname{}} Another way to think about your trick is that you are
using a writer monad to evaluate your language and produce the first-order
syntax tree as a side effect. Then you could have used a state monad
transformer on top of that for the fresh name supply. But that's a stylistic
choice.

\speak{\studname{}} You seem to be making a big deal out of this. I
just did what I thought was the most straightforward thing to do.

\end{dialogue}

\subsection{Related work}

\begin{dialogue}

\speak{\docname{}} Others have solved this problem before but I have
never seen a solution as simple as yours and was quite surprised that
it works. For example, in this paper \direct{\docname{} shows
  \studname{} the paper \cite{Generic}}, the authors use a
continuation monad to be able to reify the monadic constructs of their
language. If they were to implement your robot language they would use
the same \texttt{Program} data type as you do. However, they would not
expose that to the user of the language. Instead they would create a
type like this:

\end{dialogue}

\begin{small}
\begin{verbatim}
data P a =
  P (forall r. ((a -> Program r) -> Program r))
\end{verbatim}
\end{small}

\begin{dialogue}
\speak{\docname{}} The type \texttt{P} is a continuation monad so the
monad instance comes naturally. Operations on \texttt{P} are defined
using \texttt{Bind} like this:
\end{dialogue}

\begin{small}
\begin{verbatim}
while :: P BoolE -> P () -> P ()
while c b =
  P (\k -> While (runP c) (runP b) `Bind` k)
\end{verbatim}
\end{small}
%\pagebreak
\begin{small}
\begin{verbatim}
cond :: BoolE -> P () -> P () -> P ()
cond b t e =
  P (\k -> Cond b (runP t) (runP e) `Bind` k)

runP :: P a -> Program a
runP (P f) = f (\a -> Return a)
\end{verbatim}
\end{small}

\begin{dialogue}
\speak{\studname{}} I see. Continuations are quite magical to me. I
could never have come up with that technique. How do they deal with
transforming higher-order programs to first-order programs?

\speak{\docname{}} That's a good question. The transformation to first
order programs is dealt with by a library called Syntactic. It is
described in a separate paper \direct{\docname{} pulls out the paper
  \cite{axelsson2012generic} from a pile of papers}. It's hard to do
an apples to apples comparison between Syntactic and your
technique. Syntactic solves a much bigger problem than what you're
doing so it is naturally more complicated.

\speak{\studname{}} Ok.

\speak{\docname{}} There are also the papers
\cite{Farmer,sculthorpe2013constrained} which solves
the problem in a manner that is more similar to yours. They also have explicit
constructors {\tt Bind} and {\tt Return} but give them a slightly different
type which guarantees that all applications of {\tt Bind} are {\em normalised},
so that all {\tt Bind}s are right-associated. Unexpectedly, this
allows them to make instances of the {\tt Monad} class but at the same time
constrain the arguments of {\tt Bind}.
Their technique is more complicated than yours, but just as with the
case of Syntactic, they solve a more general problem.

However, there is one particular thing I like about your technique.

\speak{\studname{}} What is that?

\speak{\docname{}} Your technique is compositional.

\end{dialogue}

\subsection{Compositionality}

\begin{dialogue} 
\speak{\studname{}} What does it mean that my technique is compositional?

\speak{\docname{}} The compile function is compositional because there
is one case for each constructor and each case deals with exactly one
constructor. In particular, the constructors {\tt Return} and {\tt
  Bind} are handled completely separately from all other language
constructs.

\speak{\studname{}} And that is good ?

\speak{\docname{}} Absolutely! Compositional definitions are nice
because they imply that there is no weird semantical interaction
between the constructs as they are defined independently. But it also
means that it should be possible to factor out the constructors
\texttt{Return} and \texttt{Bind} into a data type of their own. Then,
it could be combined with other data types using techniques like Data
Types \'{a} la Carte \cite{Wouter} or CompData \cite{CompData}. This
way, a language can be designed piece by piece.You can then select the
set of pieces required for a particular task or that suits a
particular brand of robots.

\speak{\studname{}} Oh! Interesting.
\end{dialogue}

\subsection{The Monad laws}

\begin{dialogue}
\speak{\docname{}} There is still one problem with your method though.

\speak{\studname{}} What's that?

\speak{\docname{}} When we make instances of the monad class we expect
certain laws to hold. \direct{\docname{} writes the laws on his whiteboard}

\end{dialogue}

\begin{small}
\begin{verbatim}
m >>= return    = m
return a >>= f  = f a
(m >>= f) >>= g = m >>= \a -> f a >>= g
\end{verbatim}
\end{small}

\begin{dialogue}
\speak{\docname{}} These laws clearly don't hold for your instance. Take the
first law for instance. The left hand side will have extra \texttt{Bind} and
\texttt{Return} constructors compared to the right hand side.

\speak{\studname{}} Hmmm. I hadn't really thought about that. But adding an
extra return at the end of a computation shouldn't make a difference
in my implementation.

\speak{\docname{}} So, you're saying that your implementation actually obeys
the first monad law?

\speak{\studname{}} Well, at least when I run my programs I will never see
any difference between the left hand side and the right hand side.

\speak{\docname{}} Aha, so what you're saying is that if we compare the
semantics of programs rather than comparing the programs themselves
then we get some useful laws. \direct{\docname{} scribbles some new laws
  on the whiteboard}

\end{dialogue}
%\end{minipage}

\begin{small}
\begin{verbatim}
eval (m >>= return)    = eval m
eval (return a >>= f)  = eval (f a)
eval ((m >>= f) >>= g) = eval (m >>= \a -> f a >>= g)
\end{verbatim}
\end{small}

%\begin{minipage}[]{\linewidth}
\begin{dialogue}
\speak{\docname{}} These laws are morally the same as the monad
laws, especially if we don't let the user of the robot language ever
compare terms in the language.

\speak{\studname{}} Yes, that captures my intuition very well.

\speak{\docname{}} Ok, good. Can you prove these equations?

\speak{\studname{}} No, I don't have any experience proving programs correct.

\speak{\docname{}} Well, it shouldn't be that difficult. Let me see
what I can come up with.

\direct{\docname{} scribbles frenetically on a piece of paper for a
  couple of minutes.}

Aha! Your technique is quite general. It can reify
monads into any kind of structure which is a monoid. Return translates
into the monoid unit and bind translates into the monoid operation.

For example, in your compilation function it is important that the semantics of
\texttt{PSkip} is the identity of the semantics of \texttt{PSeq} and that
\texttt{PSeq}
is associative.
\speak{\studname{}} Ok, that sounds good. I take that as meaning the method
is quite general?

\speak{\docname{}} Yes, requiring a monoid is a very mild restriction.

Look at the time! This was interesting, I got quite carried away. We must 
round off but please get back to me if you make any more progress on your 
robot language.

\speak{\studname{}} Thanks very much for your time. Bye!
\speak{\docname{}} Thank you.

\end{dialogue}

\section{Composing reifiable monadic languages} 
\label{sec:composing}

\emph{At \studname{}'s computer} \newline

\studname{} was really intrigued by the idea of compositionally
building embedded languages and after having read the papers
\docname{} showed him he decides to try his own approach to the problem.

Inspired by CompData he starts out by adding the code in figure
\ref{fig:compdata} to his file. The data type \texttt{:+:} is used for
composing languages, and the \texttt{:<:} typeclass provides coercions
so that the programmer doesn't have to worry about using the right
sequence of the \texttt{InjL} and \texttt{InjR} constructors to inject
terms into the composed language.

\begin{figure}
\begin{small}
\begin{verbatim}
data (e1 :+: e2) x a 
  = InjL (e1 x a) | InjR (e2 x a)
infixr :+:

class sub :<: sup where
  inj :: sub x a  -> sup x a

instance f :<: f where
  inj = id

instance (f :<: (f :+: g)) where
   inj = InjL

instance (f :<: h) => (f :<: (g :+: h)) where
  inj = InjR . inj 
\end{verbatim}
\end{small}
\caption{Data types from CompData for composing languages.}
\label{fig:compdata}
\end{figure}

\studname{} then starts to add data types for the different language
constructs in his robotic language. The different operations for the
robot are pleasingly easy to add.
%\pagebreak
\begin{small}
\begin{verbatim}
data MoveOp x a where 
  Move :: MoveOp x ()

data TurnOp x a where 
  TurnLeft  :: TurnOp x () 
  TurnRight :: TurnOp x ()

data SensorOp x a where 
  Sensor :: SensorOp x BoolE

data CondOp x a where 
  Cond :: x BoolE -> x () -> x () -> CondOp x ()

data WhileOp x a where 
  While :: x BoolE -> x () -> WhileOp x () 
\end{verbatim}
\end{small}

In order to test out these definitions, \studname{} next turns to
implementing a compiler. He realises that the compiler now needs to be
implemented as a class with one instance per compilable sub-language.

\begin{small}
\begin{verbatim}
runCompile :: Compile f => f a -> Prg
runCompile prg = snd $ compile s prg
  where s = newNameSupply

class Compile f where
  compile :: NameSupply -> f a -> (a, Prg)

instance Compile (MoveOp x) where
  compile _ Move = ((), PMove)

instance Compile (TurnOp x) where
  compile _ TurnLeft = ((), PTurnLeft)
  compile _ TurnRight = ((), PTurnRight)

instance Compile (SensorOp x) where
  compile s Sensor = (Var nom, PSensor nom)
    where v   = supplyValue s
          nom = "v" ++ show v
\end{verbatim}
\pagebreak
\begin{verbatim}
instance Compile x => Compile (CondOp x) where
  compile s (Cond b p1 p2) = 
    ((),bp `PSeq` PCond b' p1' p2') 
    where (s1,s2,s3) = split3 s
          (b',bp)    = compile s1 b
          (a1,p1')   = compile s2 p1
          (a2,p2')   = compile s3 p2 
\end{verbatim}
\end{small}
%% instance Compile x => Compile (CondOp x) where
%%   compile s (Cond b p1 p2) = ((),PCond b p1' p2') 
%%     where
%%       (s1,s2) = split2 s
%%       (a1,p1') = compile s1 p1
%%       (a2,p2') = compile s2 p2 
%\pagebreak
\begin{small}
\begin{verbatim}
instance Compile x => Compile (WhileOp x) where
  compile s (While wp p) = ((),PWhile nom nwp p')
    where (s1,s2,s3) = split3 s
          (b,wp')    = compile s1 wp
          (c,p')     = compile s2 p
          nom        = "v" ++ (show $ supplyValue s3)
          nwp        = wp' `PSeq` PAssign nom b

instance (Compile (e1 f), Compile (e2 f)) 
  => Compile ((e1 :+: e2) f) where
  compile s (InjL a) = compile s a
  compile s (InjR a) = compile s a 
\end{verbatim}
\end{small}

But when \studname{} tries to add the monadic operations he finds that
they are quite resistant to a compositional treatment. After much
struggle he comes up with the following data type definition:

\begin{small}
\begin{verbatim}
data Mops f x a where
  Oper   :: f x a -> Mops f x a
  Return :: a -> Mops f x a
  Bind   :: x a -> (a -> x b) -> Mops f x b
\end{verbatim}
\end{small}

The recursion is provided by another data type that \studname{} 
calls {\tt MonadExp}.

\begin{small}
\begin{verbatim}
data MonadExp f a = In (Mops f (MonadExp f) a)
\end{verbatim}
\end{small}

\studname{} thinks of the {\tt MonadExp} data type as a representation
of a monadic language parameterised over operations {\tt Oper} that
are constructed using the constructors of some type {\tt f}. The monad
instance for this data type is only slightly more complicated compared
to the earlier, non-compositional setting.

\begin{small}
\begin{verbatim}
instance Monad (MonadExp f) where 
  return = In . Return
  (>>=) a f = In (Bind a f) 
\end{verbatim}
\end{small}

\studname{} can now write the type of his robotic language, built
from independent pieces. It is noteworthy that the monadic expressions
have to be added on top of all the other language constructs.

\begin{small}
\begin{verbatim}
type Robot = MonadExp (MoveOp :+: TurnOp  :+:
                       CondOp :+: WhileOp :+:
                       SensorOp)
\end{verbatim}
\end{small}

Given the definition of his language \studname{} can now write an
injection function which helps writing smart constructors for all the
different robot operations.

\begin{small}
\begin{verbatim}
inject :: (sub :<: f) 
       => sub (MonadExp f) a 
       -> MonadExp f a
inject a = In $ Oper $ inj $ a

move :: (MoveOp :<: f) => MonadExp f ()
move =  inject Move

turnL :: (TurnOp :<: f) => MonadExp f ()
turnL = inject TurnLeft

turnR :: (TurnOp :<: f) => MonadExp f ()
turnR = inject TurnRight

sensor :: (SensorOp :<: f) => MonadExp f BoolE
sensor = inject Sensor 
\end{verbatim}
\end{small}

\begin{small}
\begin{verbatim}
cond :: (CondOp :<: f) 
     => MonadExp f BoolE 
     -> MonadExp f () -> MonadExp f () -> MonadExp f ()
cond b p1 p2 = inject $ Cond b p1 p2

while :: (WhileOp :<: f) 
      => MonadExp f BoolE 
      -> MonadExp f () -> MonadExp f ()
while pb p = inject $ While pb p
\end{verbatim}
\end{small}

Things are starting to fall into place. The only remaining bit is to
complete the compiler for monadic expressions.

\begin{small}
\begin{verbatim}
instance (Compile x, Compile (f x)) 
  => Compile (Mops f x) where
  compile s (Oper o) = compile s o
  compile s (Return a) = (a,PSkip)
  compile s (Bind m f) = (b,prg1 `PSeq` prg2) 
    where (s1,s2)  = split2 s
          (a,prg1) = compile s1 m
          (b,prg2) = compile s2 (f a)

instance (Compile (f (MonadExp f))) 
  => Compile (MonadExp f) where
  compile s (In a) = compile s a
\end{verbatim}
\end{small}

Now, \studname{} has all the functions he needs to compile a simple
example program.

\begin{small}
\begin{verbatim}
test1 :: Robot () 
test1 = do 
  move
  turnL
  move 
  turnL 
\end{verbatim}
\end{small}

Compiling the program above gives the expected result.

\begin{small}
\begin{verbatim}
PSeq PMove (PSeq PTurnLeft (PSeq PMove PTurnLeft))
\end{verbatim}
\end{small}

\studname{} decides to contact \docname{} again to show him what 
he has done.

\vspace{5mm}

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
Dear Prof. \docname{}
\newline \newline
\noindent I have attached a file which demonstrates that the monadic
constructs can be factored out and compiled separately, just as you
suggested.
\newline \newline

\noindent Thanks \newline
\noindent \studname{}
\end{minipage} 
}

\vspace{5mm}

\noindent\colorbox{light-gray}{
\begin{minipage}[]{0.9\linewidth}
\noindent 
\studname{},
\newline

Fascinating! You've managed to factor out the monadic operations so
that they can be compiled once and for all. Users of your library
don't have to be concerned at all with the semantics of bind and return.
Very nice!
\newline
%% \end{minipage}}

%% \noindent\colorbox{light-gray}{
%% \begin{minipage}[]{0.9\linewidth}
\noindent 
I note that the data type for monadic operations cannot be composed as
freely as the other operations, but instead has to be applied
separately as a final step. Your
solution is very similar to how variable binding is handled in the
Syntactic library \cite{syntactic}.
\newline \newline

\noindent Prof. \docname{}
\newline
\noindent Dept. of Computer Science and Engineering \newline
\noindent Chalmers University of Technology 
\end{minipage} 
}

%One of the advantages of the monad reification technique we have
%presented is that it is compositional. The monadic constructs can be
%dealt with separately from all the primitives of the embedded
%languages.

%A consequence of the compositionality is that we can use techniques
%such as CompData \cite{} to decompose the definition of DSLs into
%smaller reusable building blocks. Figure \ref{fig:compdata} shows a
%compositional implementation of the monadic constructs, separately
%from any other constructs. The type \verb!Cxt! is defined in the
%package CompData and ties the recursive knot.

%\begin{figure}
%\begin{verbatim}
%data MonadT r a where
%  Return :: a -> MonadT r a
%  Bind   :: r a -> (a -> r b) -> MonadT r b

%instance (MonadT :<: f) => Monad (Cxt h f a) where
%  return a = iReturn a
%  m >>= f  = iBind m f
%\end{verbatim}
%\label{fig:compdata}
%\caption{Compositional definition monads}
%\end{figure}

%\end{minipage}

% \docname{} calls upon \studname{} to tally up the limitations of his approach

% Monad Laws (Do the hold ?) 
% Infinite programs 
% Only certain ``kinds'' of reification!

\section{Epilogue}

%This short story has outlined the discovery process and the many turns from 
%hopefulness to despair and back to hopefulness of two people who met over 
%a language embedded in Haskell. 
%% This was the story about two people meeting over the discovery of a way to 
%% reify monads. Along the road from idea to implementation there was some ups 
%% and downs but in the end they did arrive at a compositional 
%% and reifiable way of constructing monadic EDSLs. The method, albeit not as 
%% general as briefly hoped for, is an example of a simple and elegant approach 
%% to the problem.

This paper shows a simple method of implementing monadic EDSLs. It shows a
na\"ive approach to the monad reification problem which has the additional benefit of being compositional. In section 
\ref{sec:composing}, the language is reimplemented in an extensible way and 
the compile function is shown explicitly to be compositional by treating sub 
parts of the language separately.

The example language used throughout the story is small and quite
limited. But the technique presented in this paper does scale up to
larger languages and more complicated language constructs.  It is
currently used in the implementation of Obsidian, an EDSL for general
purpose GPU programming \cite{Obsidian}. Obsidian has language
constructs which are considerably more advanced than the robot
language is this paper. One example is the sequential for-loop which
is higher-order (the type parameter {\tt t} is not relevant to the
current discussion):
\begin{small}
\begin{verbatim}
SeqFor :: EWord32 -> (EWord32 -> Program t ())
       -> Program t ()
\end{verbatim}
\end{small}
There does, however, seem to be restrictions on what kind of language
constructs the reification technique presented in this paper can deal
with. For instance, the functions from the {\tt MonadPlus} class has
resisted our attempts at adding them to an EDSL in the same way as
we've added {\tt Return} and {\tt Bind}. The exact limits of the
expressiveness of our method is currently unknown to us.

%figure~\ref{fig:spiral}, shows the output generated from running the 
 %{\tt spiralIn} program.

%\studname{} feels motivated to try writing some program that shows a little bit more
%intelligence in the robot. A small program that makes sensible decisions based
%on the sensor information is implemented.

%\begin{figure}
%\includegraphics[width=.49\linewidth]{./spiral1}
%\includegraphics[width=.49\linewidth]{./spiral2}
%\caption{Images generated by our robot simulator using the {\tt spiral} program.An important detail is that the robot simulator does not evaluate the Monadic program but rather executes a compiled first-order representation of that program.}
%\label{fig:spiral}
%\end{figure}

%\begin{figure}
%\includegraphics[width=.49\linewidth]{./wall1}
%\includegraphics[width=.49\linewidth]{./wall2}
%\caption{Images generated by the robot simulator using the {\tt followWall} program.}
%\label{fig:followwall}
%\end{figure}

%\FloatBarrier
%Set the mood. \docname{} is more of a deep ``Theory'' kind of person

%Suppose you have just bought yourself a little programmable toy robot
%and you want to design a small DSL in Haskell to program it. The robot
%can turn, move forward and has a sensor to detect whether it is facing
%an obstacle or not. Inspired by the robot language in \cite{} you come
%up with a set of primitives shown in figure
%\ref{fig:interface}. In particular, you would like to use the
%do-notation to sequence programs.

%\begin{figure} 
%\begin{itemize} 
%  \item \verb!move      :: Program ()! 
%  \item \verb!turnLeft  :: Program ()!
%  \item \verb!turnRight :: Program ()!
%  \item \verb!sensor    :: Program Bool!
%  \item \verb!cond      :: Program Bool -> Program () -> Program () -> Program ()!
%  \item \verb!while     :: Program Bool -> Program () -> Program ()!
%  \item \verb!instance Monad Program!
%\end{itemize} 
%\label{fig:interface} 
%\caption{Proposed set of basic robot operations} 
%\end{figure}

%\emph{Example programs}

%The next task is to implement this language. Compiling EDSLs like this
%to something which the physical robot can execute is well know and
%there are off-the-shelf techniques to use. But there is one quirk; how
%does one generate an abstract syntax tree from a type which implements
%the Monad interface?

%Since you want to compile the language there must be an abstract
%syntax tree representation of some sort of the language. The natural
%thing is just to create a new type where each language construct has
%its own constructor. But should you deal with the monadic constructs?
%Well, the simplest solution would be to just add them to the type as well. The resulting type

%\begin{figure}
%\begin{verbatim}
%data BoolE = Lit Bool
%         | Var String
%         | (:||:) BoolE BoolE
%         | (:&&:)  BoolE BoolE
%         | Not BoolE

%data Program a where
%  Move      :: Program ()
%  TurnRight :: Program ()
%  TurnLeft  :: Program ()

%  Sensor    :: Program BoolE
%  Cond      :: BoolE -> Program () -> Program () -> Program ()
%  While     :: Program BoolE -> Program () -> Program ()
%   
%  Return    :: a -> Program a 
%  Bind      :: Program a -> (a -> Program b) -> Program b
%\end{verbatim}
%\label{fig:program}
%\caption{A data type for programs}
%\end{figure}

%\emph{More text}

%In the rest of this paper we will see that this na\"ive method
%actually can be made to work and that it's a practical, compositional
%technique for reifying monads.

### 2013-09-30 src2/bb/paper.tex
another approach

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

%added JS
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{url}
\usepackage{xcolor}
\usepackage{dialogue}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}

\definecolor{light-gray}{gray}{0.85}
\newcommand{\docname}{Bj\"{o}rn}
\newcommand{\studname}{Benny}

\begin{document}

\exclusivelicense
\conferenceinfo{ICFP~'13}{September 25--27,2013, Boston, MA, USA} 
\copyrightyear{2013} 
\copyrightdata{978-1-4503-2326-0/13/09} 
\doi{2500365.2500611}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

%\title{Reifiable and Compositional Monadic Embedded Languages}
\title{Simple and Compositional Reification of Monadic Embedded Languages}
\subtitle{Functional Pearl}

\authorinfo{Josef Svenningsson \and Bo Joel Svensson}
           {Chalmers University of Technology}
           {\{josefs,joels\}@chalmers.se}

\maketitle

\begin{abstract}
When writing embedded domain specific languages in Haskell, it is
often convenient to be able to make an instance of the {\tt
Monad} class to take advantage of the do-notation and the extensive
monad libraries. Commonly it is desirable to compile such languages 
rather than just interpret them.
This introduces the problem of monad {\em reification}, i.e. observing 
the structure of the monadic computation.
We present a solution to the monad reification problem and
illustrate it with a small robot control language.
Monad reification is not new but the novelty
of our approach is in its directness, simplicity and compositionality.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages]
\category{D.3.4}{Programming Languages}{Processors}[Code generation]

\keywords
Compiling, embedded language, reification, monads

%\section{Plan}
%
%\begin{itemize} 
%\item The language: Robots
%\item Problem statement 
%  \begin{itemize}
%  \item Want to use DO-notation
%  \item Examples
%        Spirals

%  \item Want to be able to reify a syntax tree
%  \item why ? (Compile and transfer code to actual physical robot) 
%  \end{itemize}

%\item What is the most Naive way to approach this? 
%  \begin{itemize}
%  \item Add return and bind constructors + monad instance. 
%  \end{itemize}
%\item How can this Possibly work? 
%  \begin{itemize} 
%  \item Evaluation 
%  \item Compilations
%  \end{itemize}
%\item Compositional! 
%  \begin{itemize} 
%  \item Data types a'la carte! : NO 
%  \item Will Use the CompData approach. (which is based on a'la carte though) 
%  \end{itemize}
%\item Monad Laws 
%  \begin{itemize} 
%  \item eval (return a \verb!>>=! f) = eval (f a) 
%  \end{itemize}
%\item Related (if at all) 
%
%\end{itemize}

\input{body}
%\input{partII}
%\input{partIII}

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

\acks
We would like to thank Emil Axelsson for very valuable help 
during the implementation of the methods used in this pearl. We 
thank Mary Sheeran for suggesting the names \docname~ \& \studname. 
The ICFP reviewers provided many helpful suggestion which has improved the 
paper.

This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the Swedish Research Council.

%Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.
\begin{thebibliography}{9}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Axelsson()]{syntactic}
E.~Axelsson.
\newblock Syntactic.
\newblock http://hackage.haskell.org/package/syntactic.

\bibitem[Axelsson(2012)]{axelsson2012generic}
E.~Axelsson.
\newblock A generic abstract syntax model for embedded languages.
\newblock In \emph{Proceedings of the 17th ACM SIGPLAN international conference
  on Functional programming}, pages 323--334. ACM, 2012.

\bibitem[Bahr and Hvitved(2012)]{CompData}
P.~Bahr and T.~Hvitved.
\newblock Parametric compositional data types.
\newblock In \emph{MSFP}, pages 3--24, 2012.

\bibitem[Claessen et~al.(2012)Claessen, Sheeran, and Svensson]{Obsidian}
K.~Claessen, M.~Sheeran, and B.~J. Svensson.
\newblock Expressive {A}rray {C}onstructs in an {E}mbedded {GPU} {K}ernel
  {P}rogramming {L}anguage.
\newblock In \emph{Proceedings of the 7th workshop on Declarative aspects and
  applications of multicore programming}, DAMP '12, pages 21--30, New York, NY,
  USA, 2012. ACM.
\newblock ISBN 978-1-4503-1117-5.

\bibitem[Farmer and Gill(2012)]{Farmer}
A.~Farmer and A.~Gill.
\newblock {Haskell} {DSL}s for interactive web services.
\newblock In \emph{1st International Workshop on Cross-model Language Design
  and Implementation}, Sep 2012.
\newblock (published on workshop website,
  http://workshops.inf.ed.ac.uk/xldi2012/).

\bibitem[Hudak(1999)]{HUDAK}
P.~Hudak.
\newblock \emph{The Haskell School of Expression: Learning Functional
  Programming through Multimedia}.
\newblock Cambridge University Press, New York, NY, USA, 1999.
\newblock ISBN 0521643384.

\bibitem[Persson et~al.(2012)Persson, Axelsson, and Svenningsson]{Generic}
A.~Persson, E.~Axelsson, and J.~Svenningsson.
\newblock Generic monadic constructs for embedded languages.
\newblock In \emph{Proceedings of the 23rd international conference on
  Implementation and Application of Functional Languages}, IFL'11, pages
  85--99, Berlin, Heidelberg, 2012. Springer-Verlag.

\bibitem[Sculthorpe et~al.(2013)Sculthorpe, Bracker, Giorgidze, and
  Gill]{sculthorpe2013constrained}
N.~Sculthorpe, J.~Bracker, G.~Giorgidze, and A.~Gill.
\newblock The constrained-monad problem.
\newblock In \emph{Proceedings of the 18th ACM SIGPLAN international conference
  on Functional programming}. ACM, 2013.

\bibitem[Swierstra(2008)]{Wouter}
W.~Swierstra.
\newblock Data types \`{a} la carte.
\newblock \emph{Journal of Functional Programming}, 18\penalty0 (4):\penalty0
  423, 2008.

\end{thebibliography}

%\bibliography{paper}

%\begin{thebibliography}{}
%\softraggedright

%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...

%\end{thebibliography}

\end{document}

### 2013-09-30 src2/bb/partI.tex
another approach

\section{Imperative robot language}
In the book ``The Haskell school of expression'', \cite{HUDAK}, Paul Hudak 
describes an imperative robot language. In the book the robot language 
is based on the statemonad and robot programs are evaluated graphically 
in a grid world. What would happen if we instead wanted to actually reify 
the monadic computation and compile it into some form that could for example 
be run in actual robot hardware. It is probably unlikely that such a robot 
would support the Haskell runtime system.  
The Robot language that we use as an example is very similar to the one 
appearing in Hudak's book. Robots can {\em move} forwards, {\em turn} left 
and right, can turn a {\em pen} on and off and it can react to its 
environment via a forward facing {\em sensor}.

\begin{figure} 
\begin{itemize} 
  \item \verb!move      :: Program ()! 
  \item \verb!turnLeft  :: Program ()!
  \item \verb!turnRight :: Program ()!
  \item \verb!pen       :: Exp -> Program ()!
  \item \verb!sensor    :: Program Exp!
\end{itemize} 
\label{fig:interface} 
\caption{Proposed set of basic robot operations} 
\end{figure}

Figure \ref{fig:interface}, shows a set of basic operations that our robots 
should implement. The reason that {\tt pen} takes an expression and that 
{\tt sensor} gives an expression is that we want to be able to reify abstract 
syntax trees. Had the expressions been replaced by Haskell booleans we would 
have been forced to evaluate these programs at Haskell runtime. The {\tt Exp} 
type is a very simple boolean expression datatype.

\begin{small}
\begin{verbatim}
data Exp = Lit Bool
         | Var String
         | (:||:) Exp Exp
         | (:&&:)  Exp Exp
         | Not Exp  
\end{verbatim}
\end{small}

\subsection{Example programs}
\FloatBarrier

Here we present a few example programs that we want to be able to express in 
our language. We have chosen the examples to illustrate how useful a 
{\tt Monad} instance is for our language, it enables use of all the control 
constructs of the {\tt Control.Monad} library.

\begin{small}
\begin{verbatim} 
spiralIn :: Int -> Program () 
spiralIn 0 = return () 
spiralIn n = do
  replicateM_ 2 $ do
    replicateM_ n move
    turnLeft
  spiralIn (n-1) 
\end{verbatim}
\end{small}

Figure \ref{fig:spiral}, shows the resulting output from the program above. The
graphics in the figure is not created by evaluation but rather via compilation 
of the Monadic program into a first order representation, this is a key feature 
of our approach. 
% Jumping to quickly into the magic ?

Another example, that displays more intelligence in our robot, and also 
uses the {\tt sensor} to query the surrounding world is the {\tt followWall} 
Program. This programs the robot to always keep a wall on its left side.

\begin{small}
\begin{verbatim}
followWall :: Program () 
followWall = do
  while (return true) $ do
    b <- checkLeft
    cond b sMove (turnLeft >> move) 
  where
    sMove = do
      s <- sensor
      cond s turnRight move  
\end{verbatim}
\end{small}

\begin{figure}
\includegraphics[width=.49\linewidth]{./spiral1}
\includegraphics[width=.49\linewidth]{./spiral2}
\caption{Images generated by our robot simulator using the {\tt spiral} program.An important detail is that the robot simulator does not evaluate the Monadic program but rather executes a compiled first order representation of that program.}
\label{fig:spiral}
\end{figure}

\begin{figure}
\includegraphics[width=.49\linewidth]{./wall1}
\includegraphics[width=.49\linewidth]{./wall2}
\caption{Images generated by the robot simulator using the {\tt followWall} program.}
\label{fig:followwall}
\end{figure}

%\subsection{Using Monads and do-notation}

\FloatBarrier
\subsection{Implementation of the robot language}

\begin{small}
\begin{verbatim} 
data Program a where
  Move      :: Program ()
  TurnRight :: Program ()
  TurnLeft  :: Program ()

  Sensor    :: Program Exp

  Pen       :: Exp -> Program () 
  Cond      :: Exp 
            -> Program () 
            -> Program () 
            -> Program ()

  While     :: Program Exp 
            -> Program () 
            -> Program ()

  Return    :: a -> Program a 
  Bind      :: Program a 
            -> (a -> Program b) 
            -> Program b

\end{verbatim}
\end{small}

The \verb!Monad! instance.

\begin{small}
\begin{verbatim} 
instance Monad Program where
  return = Return
  (>>=)  = Bind
\end{verbatim}
\end{small}

The expression type represent simple boolean expressions. It captures the 
operations we want to be able to perform on the input the robot gets from its 
sensor.

\begin{small}
\begin{verbatim}
data Exp = Lit Bool
         | Var String
         | (:||:) Exp Exp
         | (:&&:)  Exp Exp
         | Not Exp  
\end{verbatim}
\end{small}

A possible extension to the capabilities of the robot would be to add integers
and arithmetic to the expression datatype. We leave this as an exercise to the 
reader.

\begin{small}
\begin{verbatim} 
printExp :: Exp -> String
printExp (Lit b) = show b
printExp (Var nom) = nom
printExp (a :||: b) = "(" ++ printExp a   ++ 
                      " || "              ++ 
                      printExp b ++ ")"
printExp (a :&&: b) = "(" ++ printExp a   ++ 
                      " && "              ++ 
                      printExp b ++ ")"
printExp (Not a)    = "(" ++ "not" ++ " " ++ 
                      printExp a ++ ")"

instance Show Exp where
  show = printExp 
\end{verbatim} 
\end{small}

Given an {\em environment}, a map from names to booleans, it is also possible 
to evalute these expressions.

\begin{small}
\begin{verbatim}
type Env = [(String, Bool)]

evalExp :: Env -> Exp -> Maybe Bool
evalExp e (Lit a) = Just a
evalExp e (Var str) = lookup str e
evalExp e (e1 :||: e2) = liftM2 (||) (evalExp e e1) 
                                     (evalExp e e2)
evalExp e (e1 :&&: e2) = liftM2 (&&) (evalExp e e1) 
                                     (evalExp e e2)
evalExp e (Not e1)     = liftM  not  (evalExp e e1)
\end{verbatim} 
\end{small}

\begin{itemize}
 \item Difference between {\tt Program Exp} and {\tt Exp}.
 \item {\tt Cond} takes {\tt Program ()}s and results in a {\tt Program ()} 
   if {\tt Cond} would have returned a {\tt Program a} we would be forced into 
   evaluation ``mode'' for conditionals. \newline 
   Another option is probably that we could have a 
   {\tt Cond :: Reifyable a => Exp -> Program a -> Program a -> Program a}. 
   Then some suitable {\tt Reifyable} class would be needed. 
\end{itemize}

%maybe show evaluate in pieces! 
% Clean up the eval code and make sure it does what it should !

%Talk about eval and explain World and Grid and Robots !

\subsection{Evaluating programs}

To be able to evaluate a robot program a world is needed. We consider the 
robot to be a part of this world. The robot moves around in a two dimensional 
grid of Cells. Each cell can contain either a wall or a ground.

\begin{small}
\begin{verbatim}
data Cell = Wall | Ground 
\end{verbatim}
\end{small}

The grid is a two dimensional array of cells.

\begin{small}
\begin{verbatim}
type Grid = Array (Int,Int) Cell
\end{verbatim}
\end{small}

There is a small set of operations defined on Grids.

\begin{small}
\begin{verbatim}
size :: Grid -> (Int,Int) 
size = (\(x,y) -> (y,x)) .  snd  . bounds

(*!) :: Grid -> (Int,Int) -> Cell
(*!) g (x,y) = (!) g (y,x)

fromList :: ((Int,Int),(Int,Int)) -> [Cell] -> Grid 
fromList = listArray
\end{verbatim}
\end{small}

More details!

\begin{small}
\begin{verbatim} 
data Direction = North | South | East | West

type Pos       = (Int,Int)

type Line      = (Pos,Pos)

data World     = World { robotDir   :: Direction, 
                         robotPos   :: Pos,
                         robotPen   :: Bool,
                         robotEnv   :: Env,
                         robotTrack :: [Line],
                         worldGrid  :: Grid}
                 deriving Show 
\end{verbatim}
\end{small}

If the result we are interested in is the state of the world after 
executing a robot program, we can define an eval function as follows

\begin{small}
\begin{verbatim} 
eval :: Program a -> World -> World 
eval p w = snd $ eval' s p w
  where
    s = unsafePerformIO $ newEnumSupply

mkName :: Supply Int -> String
mkName s = "v" ++ show (supplyValue s)
\end{verbatim}
\end{small}

The big part of the evaluation process is performed by the {\tt eval'} function 
that has the signature \newline 
\verb!Supply Int -> Program a -> World -> (a,World)!. The evaluation function is simple.  
% NOTE: it is not in the evaluation function that problems occur.

\begin{small}
\begin{verbatim}
eval' :: Supply Int -> Program a -> World -> (a,World)  
eval' s Move w = ((), w')
  where
    w' = w {robotPos = tPos (robotDir w) (robotPos w)}

eval' s TurnRight w = ((), w')
  where
    w' = w {robotDir = turnR (robotDir w)}
eval' s TurnLeft w = ((), w')
  where
    w' = w {robotDir = turnL (robotDir w)}

eval' s Sensor w = (Var nom,w') 
  where
    cell = worldGrid w *! sensorPos 
    sensorPos = tPos (robotDir w) (robotPos w)
    env' = (nom,cell == Wall) : robotEnv w
    w'   = w { robotEnv = env' } 
    nom = mkName s

eval' s (Cond b p1 p2) w =
  case evalExp (robotEnv w) b  of
    (Just True) -> eval' s p1 w
    (Just False) -> eval' s p2 w
    Nothing -> error "eval': faulty robotEnv"

eval' s (Pen b) w = ((),w') 
  where
    b' = evalExp (robotEnv w) b 
    w' = w {robotPen = fromJust b'}

eval' s (Return a) w = (a,w)
eval' s (Bind a f) w = (b,w2) 
  where
    (s1,s2) = split2 s
    (a1,w1) = eval' s1 a w
    (b,w2)  = eval' s2 (f a1) w1

tPos :: Direction -> Pos -> Pos    
tPos North (x,y) = (x,y-1)
tPos South (x,y) = (x,y+1)
tPos East  (x,y) = (x+1,y)
tPos West  (x,y) = (x-1,y)

turnR :: Direction -> Direction 
turnR North = East
turnR East = South
turnR South = West
turnR West = North

turnL = turnR . turnR . turnR   
\end{verbatim}
\end{small}

\subsection{Printing and compiling programs} 
This is where it gets interesting.

\begin{small}
\begin{verbatim}
printPrg :: Program a -> String
printPrg p = snd $ printPrg' s p
  where
    s = unsafePerformIO $ newEnumSupply

printPrg' :: Supply Int -> Program a -> (a, String)
printPrg' s Move = ((),"move\n")
printPrg' s TurnRight = ((),"turnR\n")
printPrg' s TurnLeft = ((),"turnL\n")
printPrg' s Sensor = (Var nom,nom ++ " <- sensor\n")
  where
    nom = mkName s
printPrg' s (Cond b p1 p2) = ((),str)
  where str = "if " ++ printExp b ++ "\n" ++
              "then\n" ++ snd (printPrg' s p1) ++ 
              "else\n" ++ snd (printPrg' s p2)
printPrg' s (While pb p) = ((),str)
  where
    (s1,s2) = split2 s 
    str = "while {" ++ snd (printPrg' s1 pb) ++ "}\n" ++
          "do\n" ++ 
          snd (printPrg' s1 p) ++ 
          "done\n"
printPrg' s (Pen b)  = ((),str)
  where
    str = "setPen " ++ printExp b ++ "\n"
printPrg' s (Return a) = (a, "")
printPrg' s (Bind pa f) = (b,str ++ str2) 
  where
    (s1,s2) = split2 s 
    (a,str) = printPrg' s1 pa 
    prg2 = f a
    (b,str2) = printPrg' s2 prg2

\end{verbatim}
\end{small}

%\subsection{Infinite programs}
%The spiral program can not easily be compiled.  
%
%\begin{verbatim}
%spiral n = do 
%  repeat n Move 
%  TurnLeft 
%  spiral (n+1) 
%\end{verbatim} 
%
%propose a method to compile it anyway (in stages, using lazyness).

%\begin{figure}
%\includegraphics[width=5cm]{./robot1}
%\caption{example robot graphics}
%\label{fig:ROBOT1}

%\end{figure}

%\begin{figure}[ht]
% \begin{minipage}[b]{0.45\linewidth}
%
%   \includegraphics[width=\textwidth]{./robot1}
%   %\caption{example robot graphics}
%   %\label{fig:ROBOT1}
%  \end{minipage}
%
%
%  \begin{minipage}[b]{0.45\linewidth}
%   \includegraphics[width=\textwidth]{./robot1}
%   %\caption{example robot graphics}
%   %\label{fig:ROBOT1}
%  \end{minipage}
%\end{figure}

%\begin{figure}
%\includegraphics[width=.5\linewidth]{./robot1}
%\includegraphics[width=.5\linewidth]{./robot1}
%\caption{two images  }
%\label{fig:robo}
%\end{figure}

### 2013-09-30 src2/bb/partII.tex
another approach

\section{Compositional}

One of the advantages of the monad reification technique we have
presented is that it is compositional. The monadic constructs can be
dealt with separately from all the primitives of the embedded
langauges.

A consequence of the compositionality is that we can use techniques
such as compdata \cite{} to decompose the definition of DSLs into
smaller reusable building blocks. Figure \ref{fig:compdata} shows a
compositional implementation of the monadic constructs, separately
from any other constructs. Firstly, our type \verb!MonadT! define

The type \verb!Cxt! is defined in the
package compdata and ties the recursive knot.

\begin{figure}
\begin{verbatim}
data MonadT r a where
  Return :: a -> MonadT r a
  Bind   :: r a -> (a -> r b) -> MonadT r b

instance (MonadT :<: f) => Monad (Cxt h f a) where
  return a = iReturn a
  m >>= f  = iBind m f
\end{verbatim}
\label{fig:compdata}
\caption{Compositional definition monads}
\end{figure}

### 2013-09-30 src2/bb/partIII.tex
another approach

--- 2013-09-30 src2/bb/body.tex
@@
-% Look Sebastian fisher (a play on regular expression) 
-%   Really goes all the way
-
-% Student Sven is working on a Robot programming DSL supervised by Prof. Kurt. 
-% Sven is beginner Haskeller
-% Kurt is knowledgible (experienced)
-
-% Introduction (Presents problem) 
-%  Robot language operations. 
-%  Robot exists that run programs in a simple language.
-%  want to run code in robot. 
-%  Haskell DSL and Do notation 
-%  Compilation necessary.
-
-% Student has Haskell School of expression 
-%   But need reifyable robot programs (evaluation not enough)
-
-% Show example program early 
-%   Sven is a good student and early expresses an example program in his thought 
-%   out API.
-%
-
-% How to be able to use Do notation 
-%  Adds Return and Bind naively to data structure.
-
-% Student brings --CODE SNIPPET-- to Prof. Kurt.
-
-% Now the problem, how can this work? 
-%  Prof. is familiar with Gills blog post.
-
-% Student brings --CODE SNIPPET-- (the compiler) to Prof.
-
-% Surprise 
-% Work on generalisations  and make the compositionality explicit
-%
-
-% What about the monad laws ? 
-%  Prof. asks student to verify that Monad laws hold.
-
-\section{Introduction}
-\studname{} is a computer science student working in a project involving programming 
-robots in a low-level imperative language. However, \studname{} has a budding 
-interest in functional programming using Haskell and has read ``The Haskell 
-school of expression'' \cite{HUDAK}. He gets the idea to implement 
-a language for robot control embedded in Haskell. \studname{} realises that the 
-capabilities of the robot hardware are very similar to those of the 
-robots that Hudak evaluates graphically in a grid world. There is however 
-a very important difference; the embedded language needs to be compiled into 
-some form that is understood by the robot. 
-%\studname{} in a computer science student with a budding interest in functional 
-%programming using Haskell. \studname{} is developing  
-%an embedded domain specific language (EDSL) for robot control. Inspired 
-%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
-%evaluating his robot EDSL, \studname{} wants to compile the language to a form 
-%he can run on actual robot hardware. 
-%with 
-%different motivations, \studname{} wants compile his EDSL for actual physical robots, 
-%he sets out.
-
-Guided by the capabilities of the target robot, \studname{} designs an API for 
-robot programming. The robot can perform operations 
-such as {\em move} that steps the robot forward and {\em turn} left and right. 
-The robot also has the capability to execute program loops and conditionals and 
-it has a forward facing {\tt sensor} with which it can query the world.
-
-%The robot also has a forward facing {\tt sensor} with which it can query the 
-%world and also the capabilities to execute program loops and conditionals. 
-\pagebreak
-The API that \studname{} designs is simple:
-
-%\begin{figure} 
-\begin{small}
-\begin{verbatim} 
-move      :: Program () 
-turnLeft  :: Program () 
-turnRight :: Program ()
-sensor    :: Program Bool 
-cond      :: Program Bool 
-             -> Program () 
-             -> Program () 
-             -> Program () 
-while     :: Program Bool 
-             -> Program () 
-             -> Program () 
-\end{verbatim}
-\end{small} 
-%\label{fig:interface} 
-%\caption{Proposed set of basic robot operations}
-%\end{figure}
-
-Now, \studname{} wants to express robot programs using the Haskell
-{\em do} notation. The motivation behind this is the imperative look
-and feel of the operations he identified and the potential to use all
-the control structures in the {\em Control.Monad} library. For this, a
-{\tt Monad} instance is needed.
-
-To be able to create a first-order representation of a computation
-described using monads, \studname{} needs to solve the problem of {\em
-  reifying} monads. This is the story of how \studname{} came up with
-a particularly simple solution to this problem.
-
-%\emph{*** Josef, Think about writing here}
+\section{Limitations}
 
 %\begin{minipage}[]{\linewidth}
-%\vspace{5mm}
-
-%\fbox{%
-%\parbox{0.8\linewidth}{What \studname{} does not know at this time is that he has stumbled over the 
-%problem of {\em Monad reification}. Monad reification is the observing of the shape of the computation and this is necessary in order to compile to some low level representation.}
-%}
-%\end{minipage}
-
-\subsection{Example programs}
-
-\studname{} is quite happy with his language design and decides to
-write some programs to try it out.
-%experiments 
-%with what some of these robot controlling programs could look like once his 
-%EDSL is finished.
-
-The first program is called {\tt sMove}, which implements a safe move 
-operation. This operation can be executed by the robot even though it is facing 
-an obstacle, without risk of harming robot or obstacle. 
-%.It feels like a good idea to \studname{} to check the sensor before trying to 
-%move the robot forward. Otherwise something in either the world or the 
-%robot might suffer damage.
-
-\begin{small} 
-\begin{verbatim}
-sMove :: Program () 
-sMove = cond sensor turnRight move 
-\end{verbatim}
-\end{small}
-
-The second program moves the robot forward until it stands directly 
-in front of an obstacle such as a wall.
-
-\begin{small} 
-\begin{verbatim}
-moveToWall :: Program () 
-moveToWall = while ((liftM not) sensor) move
-\end{verbatim}
-\end{small}
-
-Based on these examples, \studname{} is quite satisfied with the design of his
-language and turns to implementing it.
-
-\subsection{Implementation and data structures}
-
-Enthused by the prospect of compiling these programs to the target language and 
-seeing some robot action, \studname{} goes to work on the data structures.
-
-% Since an abstract syntax representation of the robot programs is desired,
-Since the language is going to be compiled, \studname{} realises that the 
-booleans in his language cannot be regular Haskell booleans. The booleans 
-needs to be replaced by boolean typed expressions.
-
-\pagebreak
-\begin{small} 
-\begin{verbatim}
-type Name = String
-
-data BoolE = Lit Bool
-           | Var Name
-           | (:||:) BoolE BoolE
-           | (:&&:)  BoolE BoolE
-           | Not BoolE  
-\end{verbatim} 
-\end{small}
-
-Following this, the  operations that should go into the {\tt Program} data type 
-feel straightforward.
-
-\begin{small} 
-\begin{verbatim}
-data Program a where 
-  Move      :: Program () 
-  TurnLeft  :: Program () 
-  TurnRight :: Program () 
-  Sensor    :: Program BoolE
-  Cond      :: Program BoolE 
-               -> Program () 
-               -> Program () 
-  While     :: Program BoolE 
-               -> Program () 
-               -> Program () 
-\end{verbatim} 
-\end{small}
-
-\studname{} continues by naively adding constructors for {\tt Return} and 
-{\tt Bind} to the {\tt Program} data type.
-
-\begin{small} 
-\begin{verbatim}
-data Program a where 
-  ...
-  Return :: a -> Program a 
-  Bind   :: Program a 
-            -> (a -> Program b)
-            -> Program b 
-\end{verbatim} 
-\end{small}
-
-Then he writes down the {\tt Monad} instance.
-
-\begin{small} 
-\begin{verbatim}
-instance Monad Program where 
-  return = Return 
-  (>>=)  = Bind 
-\end{verbatim} 
-\end{small}
-
-Proud of his accomplishments, \studname{} sends his Haskell module in an 
-email to Prof. \docname{}. \studname{} knows that \docname{} is teaching an 
-introductory Haskell course and should be able to provide feedback.
-
-\subsection{Problem statement} 
-\emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline
-
-\noindent Prof. \docname{} notices an email from \studname{} in his inbox and 
-opens it.
-
-\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-Dear Professor \docname{}
-\newline \newline
-\noindent I am CS student working in a robot control project. Usually, we 
-program our robots in C but I have developed an interest in Haskell 
-programming and thought it'd be natural to try implementing an EDSL. 
-I know you teach an FP course and thought I would ask for your input. 
-Attached to this mail is a file containing an outline of the data types I 
-want to use. Does this look sensible to you? \newline \newline
-
-\noindent Thank you \newline
-\noindent \studname{}
-\end{minipage} 
-}
-
-%I have progressed some towards a robot programming language. Attached 
-%to this mail is a first sketch of a {\tt Program a} data type for robot 
-%programs.
-\vspace{5mm}
-
-Prof. \docname{} opens the attachment and takes a look at the data type. He 
-particularly notices the constructors {\tt Return} and {\tt Bind} and the 
-{\tt Monad} instance. He shakes his head at \studname{}'s naivet\'e and writes an 
-email back:
-
-%\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-Hello \studname{}
-\newline \newline
-\noindent I'm afraid your implementation of the monadic primitives can never 
-work. The constructor {\tt Return} can take any arbitrary value that has 
-nothing to do with the language you're designing. These values may be 
-strings, binary trees or higher-order functions from zygomorphisms to 
-histomorphisms. The same problem goes for {\tt Bind}; it has to be able 
-to handle arbitrary values. It cannot possibly work! This is a very 
-difficult problem and such a naive solution is bound to fail. 
-\newline \newline
-%% \noindent For hints towards a solution to this I suggest that you follow 
-%% this \underline{link} \cite{GillBlog}. Another idea is that you refer 
-%% to the article ``Generic monadic constructs for embedded languages'' 
-%% \cite{Generic}.
-%% \newline \newline
-
-\noindent Prof. \docname{} \newline
-\noindent Dept. of Computer Science and Engineering \newline
-\noindent Chalmers University of Technology 
-\end{minipage} 
-}
-
-\vspace{5mm}
-
-% \section{Implementation continues}
-% \emph{Back to \studname{}'s implementation efforts} \newline \newline
-
-\noindent While waiting for feedback from \docname{}, \studname{} has 
-carried on trying to implement a compiler for his language.
-
-When \docname{}'s email does arrive, \studname{} is confused. He feels he 
-has already managed to compile the {\tt Program} data type into a 
-representation closer to that which is executed by the robot hardware.
-
-He writes an apprehensive email back to Prof. \docname{}.
-
-\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-Hello Prof. \docname{} 
-\newline \newline
-\noindent Thanks for your feedback. But I think it does work! Attached
-to this email you find a file containing my attempt at compiling the
-{\tt Program} data type into a representation closer to that which is
-executed by our robots. \newline \newline
-
-\noindent I have also attached two example programs that you can compile and 
-then run in the graphical simulator (see figure~\ref{fig:programs}) \newline \newline
-%\vspace{5mm}
-
-%\includegraphics[width=.49\linewidth]{./spiral1}
-%\includegraphics[width=.49\linewidth]{./spiral2}
-
-%\vspace{5mm}
-
-%\noindent And these two images are output from running the 
-%{\tt followWall} (see figure~\ref{fig:followWall}) program. \newline
-
-%\vspace{5mm}
-
-%\includegraphics[width=.49\linewidth]{./wall1}
-%\includegraphics[width=.49\linewidth]{./wall2}
-
-%\vspace{5mm}
-
-\noindent \studname{} 
-\end{minipage} 
-}
-
-\FloatBarrier
-
-\begin{figure*}
-\begin{tabular}{c|c}
-\begin{minipage}[]{0.45\columnwidth}
-\begin{small}
-\begin{verbatim}
-spiralIn :: Int -> Program () 
-spiralIn 0 = return () 
-spiralIn n = do
-  replicateM_ 2 $ do
-    replicateM_ n move
-    turnLeft
-  spiralIn (n-1) 
-\end{verbatim}
-\end{small}
-\end{minipage}
-\hspace{0.05\columnwidth} 
-\begin{minipage}[]{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{./bb/spiral1}
-\end{minipage}
-&
-\begin{minipage}[]{0.45\columnwidth}
-\begin{small}
-\begin{verbatim} 
-followWall :: Program () 
-followWall = 
-  while (return true) $ 
-    cond checkLeft sMove $ 
-            do turnLeft
-               move
-
-checkLeft :: Program BoolE     
-checkLeft = do
-  TurnLeft
-  s <- Sensor
-  TurnRight 
-  return s
-\end{verbatim}
-\end{small}
-\end{minipage}
-\hspace{0.05\columnwidth}
-\begin{minipage}[]{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{./bb/wall2}
-\end{minipage}
-
-\end{tabular}
-\caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
-to \docname{} in an email}
-\label{fig:programs}
-\end{figure*}
-
-\section{Compilation of the monadic robot EDSL} 
-\emph{\docname{} receives \studname{}'s latest email}\newline %\newline
-
-\noindent \docname{} looks through the {\tt Compiler} module and finds a data type describing
-a first-order representation of the robot language.
-
-\begin{small} 
-\begin{verbatim}
-data Prg = PMove          
-         | PTurnRight
-         | PTurnLeft
-         | PSensor Name  
-         | PCond BoolE Prg Prg
-         | PWhile Name Prg Prg 
-         | PSeq Prg Prg
-         | PSkip       
-         | PAssign Name BoolE
-\end{verbatim}
-\end{small}
-
-\begin{figure} 
-\begin{small}
-\begin{verbatim} 
-newNameSupply :: NameSupply 
-split2 :: NameSupply -> 
-          (NameSupply,NameSupply) 
-split3 :: NameSupply -> 
-          (NameSupply,NameSupply,NameSupply)
-supplyValue :: NameSupply -> Int 
-\end{verbatim} 
-\end{small}
-\caption{The interface of the splittable name supply used in the implementation 
-of the compile functions} 
-\label{fig:namesupply}
-\end{figure}
-
-From the {\tt Prg} data type, \docname{} presumes that the {\tt PSeq}
-constructor will be the result of compiling {\tt Bind} and that
-{\tt PSensor} binds a variable which can be used in {\tt PAssign} and
-{\tt PWhile}. The conclusion is that the representation looks
-sensible, but he is very interested in seeing the {\tt Program a -> Prg}
-transformation. The compile function assumes a splittable name
-supply with the interface described in figure \ref{fig:namesupply}.
-
-\pagebreak
-\begin{small} 
-\begin{verbatim}
-runCompile :: Program a -> Prg
-runCompile prg = snd $ compile s prg 
-  where 
-    s = newNameSupply
-
-compile :: NameSupply -> Program a -> (a, Prg) 
-compile s Move = ((),PMove)
-compile s TurnRight = ((),PTurnRight)
-compile s TurnLeft  = ((),PTurnLeft)
-compile s Sensor    = (Var nom,PSensor nom)
-  where
-    v   = supplyValue s
-    nom = "v" ++ show v 
-compile s (Cond b p1 p2) = 
-    ((),bp `PSeq` PCond b' p1' p2') 
-  where
-    (s1,s2,s3) = split3 s
-    (b',bp)    = compile s1 b
-    (a1,p1')   = compile s2 p1
-    (a2,p2')   = compile s3 p2
-compile s (While wp prg) = ((),PWhile nom nwp prg') 
-  where
-    (s1,s2,s3) = split3 s
-    (b,wp')    = compile s1 wp
-    (c,prg')   = compile s2 prg 
-    nom        = "v" ++ (show $ supplyValue s3)
-    nwp        = (wp' `PSeq` PAssign nom b) 
-compile s (Return a) = (a,PSkip)
-compile s (Bind pa f) = (b, prg1 `PSeq` prg2) 
-  where
-    (s1,s2)  = split2 s
-    (a,prg1) = compile s1 pa
-    (b,prg2) = compile s2 (f a) 
-\end{verbatim}
-\end{small}
-
-%\begin{small} 
-%\begin{verbatim}
-%compile :: Program a -> Prg
-%compile prg = snd $ compile' s prg 
-%  where 
-%    s = unsafePerformIO $ newEnumSupply
-%    
-%compile' :: Supply Int -> Program a -> (a, Prg) 
-%compile' s Move = ((),PMove)
-%compile' s TurnRight = ((),PTurnRight)
-%compile' s TurnLeft  = ((),PTurnLeft)
-%compile' s Sensor    = (Var nom,PSensor nom)
-%  where
-%    v = supplyValue s
-%    nom = "v" ++ show v 
-%compile' s (Cond b p1 p2) = ((), PCond b p1' p2') 
-%  where
-%    (s1,s2) = split2 s
-%    (a1,p1') = compile' s1 p1
-%    (a2,p2') = compile' s2 p2 
-%compile' s (While wp prg) = ((),PWhile nom nwp prg') 
-%  where
-%    (s1,s2,s3) = split3 s
-%    (b,wp') = compile' s1 wp
-%    (c,prg') = compile' s2 prg 
-%    nom  = "v" ++ (show $ supplyValue s3)
-%    nwp = (wp' `PSeq` PAssign nom b) 
-%compile' s (Return a) = (a,PSkip)
-%compile' s (Bind pa f) = (b, prg1 `PSeq` prg2) 
-%  where
-%    (s1,s2) = split2 s
-%    (a,prg1) = compile' s1 pa
-%    (b,prg2) = compile' s2 (f a) 
-%\end{verbatim}
-%\end{small}
-
-\docname{} studies the code in disbelief, saves the module and tries
-it out on some examples. The code does indeed seem to work and
-\docname{}'s disbelief is replaced with enthusiasm; this appears to be
-a simple and convenient way to reify monads.
-%What's more it does so in a 
-%compositional way. The compilation of the {\tt Return} and {\tt Bind} cases 
-%are completely independent of any other construct of the language.
-
-\docname{} sends an email to \studname{}, inviting him to a meeting.
-
-\section{Meeting in \docname{}'s office} %\section{Related Work} 
-%\emph{Meeting in \docname{}'s office} \newline
-
-%TODO Here I want to cast light on more related work 
-%and have \docname{} and \studname{} compare their approach to some of the existing ones. 
-%also they should start sketching out the Idea of using {\tt CompData} instead 
-%of a la carte for showing the compositionality.
-
-%\begin{minipage}[]{\linewidth}
-\begin{dialogue}
-
-\speak{\docname{}} Hello, come in.
-
-\speak{\studname{}} Thanks.
-
-\speak{\docname{}} So, about this robot language! You gave me a bit of a 
-surprise there. I was entirely sure that what you did was impossible.
-
-\speak{\studname{}} Oh, But I just did the first thing that came to mind. 
-There was no deep thought behind it.
-
-\speak{\docname{}} The problem is \direct{\docname{} approaches his whiteboard}
- when you try to reify the bind constructor of your representation.
-\end{dialogue}
-
-%\mbox{ 
-\begin{small}
-\begin{Verbatim}[commandchars=\\\{\}] 
-Bind :: Program a -> \underline{(a -> Program b)} -> Program b 
-\end{Verbatim}
-\end{small}
-%}
-
-\begin{dialogue}
-\speak{\docname{}} You need to come up with an element of type {\tt a} to 
-pass to the function.
-
-\speak{\studname{}} I didn't realise it was problem.
-
-\speak{\docname{}} But it is! Your solution is that you are careful
-about the return types of all operations in your language. They are
-either of unit type or some type which can be guaranteed to be reified.
-\direct{\docname{} scribbles on his whiteboard}
-
-\end {dialogue}
-
-\begin{small}
-\begin{Verbatim}[commandchars=\\\{\}] 
-Move :: Program \underline{()}
-
-Sensor :: Program \underline{BoolE}
-\end{Verbatim}
-\end{small}
-
-\begin{dialogue}
-\speak{\docname{}} This is different from how such constructs are
-normally implemented. If I had implemented the language I would have
-given {\tt Sensor} the type {\tt Program Bool} so that I could write
-an evaluator of type {\tt Program a -> a}, or perhaps {\tt Program a -> M a}.
-But then adding a monad to the language and reifying it becomes much harder.
-
-\speak{\studname{}} Yes, now that you mention it, I have seen that style used
-for defining DSLs.
-
-\speak{\docname{}} Having constructors like {\tt Sensor} return {\tt
-  BoolE} instead of {\tt Bool} is a crucial part of why your
-method works. It allows the {\tt compile} function to be written in
-such a way that it can generate the return values statically and pass
-them as arguments to {\tt Bind}. In a sense you're doing evaluation at
-the same time as compilation. There's just enough evaluation to
-remove the monadic constructs. It's a very neat trick!
-
-\speak{\studname{}} Thank you.
-
-\speak{\docname{}} Another way to think about your trick is that you are
-using a writer monad to evaluate your language and produce the first-order
-syntax tree as a side effect. Then you could have used a state monad
-transformer on top of that for the fresh name supply. But that's a stylistic
-choice.
-
-\speak{\studname{}} You seem to be making a big deal out of this. I
-just did what I thought was the most straightforward thing to do.
-
-\end{dialogue}
-
-\subsection{Related work}
-
-\begin{dialogue}
-
-\speak{\docname{}} Others have solved this problem before but I have
-never seen a solution as simple as yours and was quite surprised that
-it works. For example, in this paper \direct{\docname{} shows
-  \studname{} the paper \cite{Generic}}, the authors use a
-continuation monad to be able to reify the monadic constructs of their
-language. If they were to implement your robot language they would use
-the same \texttt{Program} data type as you do. However, they would not
-expose that to the user of the language. Instead they would create a
-type like this:
-
-\end{dialogue}
-
-\begin{small}
-\begin{verbatim}
-data P a =
-  P (forall r. ((a -> Program r) -> Program r))
-\end{verbatim}
-\end{small}
-
-\begin{dialogue}
-\speak{\docname{}} The type \texttt{P} is a continuation monad so the
-monad instance comes naturally. Operations on \texttt{P} are defined
-using \texttt{Bind} like this:
-\end{dialogue}
-
-\begin{small}
-\begin{verbatim}
-while :: P BoolE -> P () -> P ()
-while c b =
-  P (\k -> While (runP c) (runP b) `Bind` k)
-\end{verbatim}
-\end{small}
-%\pagebreak
-\begin{small}
-\begin{verbatim}
-cond :: BoolE -> P () -> P () -> P ()
-cond b t e =
-  P (\k -> Cond b (runP t) (runP e) `Bind` k)
-
-runP :: P a -> Program a
-runP (P f) = f (\a -> Return a)
-\end{verbatim}
-\end{small}
-
-\begin{dialogue}
-\speak{\studname{}} I see. Continuations are quite magical to me. I
-could never have come up with that technique. How do they deal with
-transforming higher-order programs to first-order programs?
-
-\speak{\docname{}} That's a good question. The transformation to first
-order programs is dealt with by a library called Syntactic. It is
-described in a separate paper \direct{\docname{} pulls out the paper
-  \cite{axelsson2012generic} from a pile of papers}. It's hard to do
-an apples to apples comparison between Syntactic and your
-technique. Syntactic solves a much bigger problem than what you're
-doing so it is naturally more complicated.
-
-\speak{\studname{}} Ok.
-
-\speak{\docname{}} There are also the papers
-\cite{Farmer,sculthorpe2013constrained} which solves
-the problem in a manner that is more similar to yours. They also have explicit
-constructors {\tt Bind} and {\tt Return} but give them a slightly different
-type which guarantees that all applications of {\tt Bind} are {\em normalised},
-so that all {\tt Bind}s are right-associated. Unexpectedly, this
-allows them to make instances of the {\tt Monad} class but at the same time
-constrain the arguments of {\tt Bind}.
-Their technique is more complicated than yours, but just as with the
-case of Syntactic, they solve a more general problem.
-
-However, there is one particular thing I like about your technique.
-
-\speak{\studname{}} What is that?
-
-\speak{\docname{}} Your technique is compositional.
-
-\end{dialogue}
-
-\subsection{Compositionality}
-
-\begin{dialogue} 
-\speak{\studname{}} What does it mean that my technique is compositional?
-
-\speak{\docname{}} The compile function is compositional because there
-is one case for each constructor and each case deals with exactly one
-constructor. In particular, the constructors {\tt Return} and {\tt
-  Bind} are handled completely separately from all other language
-constructs.
-
-\speak{\studname{}} And that is good ?
-
-\speak{\docname{}} Absolutely! Compositional definitions are nice
-because they imply that there is no weird semantical interaction
-between the constructs as they are defined independently. But it also
-means that it should be possible to factor out the constructors
-\texttt{Return} and \texttt{Bind} into a data type of their own. Then,
-it could be combined with other data types using techniques like Data
-Types \'{a} la Carte \cite{Wouter} or CompData \cite{CompData}. This
-way, a language can be designed piece by piece.You can then select the
-set of pieces required for a particular task or that suits a
-particular brand of robots.
-
-\speak{\studname{}} Oh! Interesting.
-\end{dialogue}
-
-\subsection{The Monad laws}
-
 \begin{dialogue}
@@
 
-\speak{\docname{}} When we make instances of the monad class we expect
+\speak{\docname{}} When we make instances of the Monad class we expect
 certain laws to hold. \direct{\docname{} writes the laws on his whiteboard}
@@
 \end{dialogue}
+%\end{minipage}
 
@@
 
+%\begin{minipage}[]{\linewidth}
 \begin{dialogue}
@@
 \begin{dialogue}
-\speak{\docname{}} These laws are morally the same as the monad
+
+\speak{\docname{}} These laws a morally the same as the monad
 laws, especially if we don't let the user of the robot language ever
@@
 \speak{\docname{}} Ok, good. Can you prove these equations?
+\end{dialogue}
+%\end{minipage}
 
-\speak{\studname{}} No, I don't have any experience proving programs correct.
-
-\speak{\docname{}} Well, it shouldn't be that difficult. Let me see
-what I can come up with.
-
-\direct{\docname{} scribbles frenetically on a piece of paper for a
-  couple of minutes.}
-
-Aha! Your technique is quite general. It can reify
-monads into any kind of structure which is a monoid. Return translates
-into the monoid unit and bind translates into the monoid operation.
-
-For example, in your compilation function it is important that the semantics of
-\texttt{PSkip} is the identity of the semantics of \texttt{PSeq} and that
-\texttt{PSeq}
-is associative.
-\speak{\studname{}} Ok, that sounds good. I take that as meaning the method
-is quite general?
-
-\speak{\docname{}} Yes, requiring a monoid is a very mild restriction.
-
-Look at the time! This was interesting, I got quite carried away. We must 
-round off but please get back to me if you make any more progress on your 
-robot language.
-
-\speak{\studname{}} Thanks very much for your time. Bye!
-\speak{\docname{}} Thank you.
-
-\end{dialogue}
-
-\section{Composing reifiable monadic languages} 
-\label{sec:composing}
-
-\emph{At \studname{}'s computer} \newline
-
-\studname{} was really intrigued by the idea of compositionally
-building embedded languages and after having read the papers
-\docname{} showed him he decides to try his own approach to the problem.
-
-Inspired by CompData he starts out by adding the code in figure
-\ref{fig:compdata} to his file. The data type \texttt{:+:} is used for
-composing languages, and the \texttt{:<:} typeclass provides coercions
-so that the programmer doesn't have to worry about using the right
-sequence of the \texttt{InjL} and \texttt{InjR} constructors to inject
-terms into the composed language.
-
-\begin{figure}
-\begin{small}
-\begin{verbatim}
-data (e1 :+: e2) x a 
-  = InjL (e1 x a) | InjR (e2 x a)
-infixr :+:
-
-class sub :<: sup where
-  inj :: sub x a  -> sup x a
-
-instance f :<: f where
-  inj = id
-
-instance (f :<: (f :+: g)) where
-   inj = InjL
-
-instance (f :<: h) => (f :<: (g :+: h)) where
-  inj = InjR . inj 
-\end{verbatim}
-\end{small}
-\caption{Data types from CompData for composing languages.}
-\label{fig:compdata}
-\end{figure}
-
-\studname{} then starts to add data types for the different language
-constructs in his robotic language. The different operations for the
-robot are pleasingly easy to add.
-%\pagebreak
-\begin{small}
-\begin{verbatim}
-data MoveOp x a where 
-  Move :: MoveOp x ()
-
-data TurnOp x a where 
-  TurnLeft  :: TurnOp x () 
-  TurnRight :: TurnOp x ()
-
-data SensorOp x a where 
-  Sensor :: SensorOp x BoolE
-
-data CondOp x a where 
-  Cond :: x BoolE -> x () -> x () -> CondOp x ()
-
-data WhileOp x a where 
-  While :: x BoolE -> x () -> WhileOp x () 
-\end{verbatim}
-\end{small}
-
-In order to test out these definitions, \studname{} next turns to
-implementing a compiler. He realises that the compiler now needs to be
-implemented as a class with one instance per compilable sub-language.
-
-\begin{small}
-\begin{verbatim}
-runCompile :: Compile f => f a -> Prg
-runCompile prg = snd $ compile s prg
-  where s = newNameSupply
-
-class Compile f where
-  compile :: NameSupply -> f a -> (a, Prg)
-
-instance Compile (MoveOp x) where
-  compile _ Move = ((), PMove)
-
-instance Compile (TurnOp x) where
-  compile _ TurnLeft = ((), PTurnLeft)
-  compile _ TurnRight = ((), PTurnRight)
-
-instance Compile (SensorOp x) where
-  compile s Sensor = (Var nom, PSensor nom)
-    where v   = supplyValue s
-          nom = "v" ++ show v
-\end{verbatim}
-\pagebreak
-\begin{verbatim}
-instance Compile x => Compile (CondOp x) where
-  compile s (Cond b p1 p2) = 
-    ((),bp `PSeq` PCond b' p1' p2') 
-    where (s1,s2,s3) = split3 s
-          (b',bp)    = compile s1 b
-          (a1,p1')   = compile s2 p1
-          (a2,p2')   = compile s3 p2 
-\end{verbatim}
-\end{small}
-%% instance Compile x => Compile (CondOp x) where
-%%   compile s (Cond b p1 p2) = ((),PCond b p1' p2') 
-%%     where
-%%       (s1,s2) = split2 s
-%%       (a1,p1') = compile s1 p1
-%%       (a2,p2') = compile s2 p2 
-%\pagebreak
-\begin{small}
-\begin{verbatim}
-instance Compile x => Compile (WhileOp x) where
-  compile s (While wp p) = ((),PWhile nom nwp p')
-    where (s1,s2,s3) = split3 s
-          (b,wp')    = compile s1 wp
-          (c,p')     = compile s2 p
-          nom        = "v" ++ (show $ supplyValue s3)
-          nwp        = wp' `PSeq` PAssign nom b
-
-instance (Compile (e1 f), Compile (e2 f)) 
-  => Compile ((e1 :+: e2) f) where
-  compile s (InjL a) = compile s a
-  compile s (InjR a) = compile s a 
-\end{verbatim}
-\end{small}
-
-But when \studname{} tries to add the monadic operations he finds that
-they are quite resistant to a compositional treatment. After much
-struggle he comes up with the following data type definition:
-
-\begin{small}
-\begin{verbatim}
-data Mops f x a where
-  Oper   :: f x a -> Mops f x a
-  Return :: a -> Mops f x a
-  Bind   :: x a -> (a -> x b) -> Mops f x b
-\end{verbatim}
-\end{small}
-
-The recursion is provided by another data type that \studname{} 
-calls {\tt MonadExp}.
-
-\begin{small}
-\begin{verbatim}
-data MonadExp f a = In (Mops f (MonadExp f) a)
-\end{verbatim}
-\end{small}
-
-\studname{} thinks of the {\tt MonadExp} data type as a representation
-of a monadic language parameterised over operations {\tt Oper} that
-are constructed using the constructors of some type {\tt f}. The monad
-instance for this data type is only slightly more complicated compared
-to the earlier, non-compositional setting.
-
-\begin{small}
-\begin{verbatim}
-instance Monad (MonadExp f) where 
-  return = In . Return
-  (>>=) a f = In (Bind a f) 
-\end{verbatim}
-\end{small}
-
-\studname{} can now write the type of his robotic language, built
-from independent pieces. It is noteworthy that the monadic expressions
-have to be added on top of all the other language constructs.
-
-\begin{small}
-\begin{verbatim}
-type Robot = MonadExp (MoveOp :+: TurnOp  :+:
-                       CondOp :+: WhileOp :+:
-                       SensorOp)
-\end{verbatim}
-\end{small}
-
-Given the definition of his language \studname{} can now write an
-injection function which helps writing smart constructors for all the
-different robot operations.
-
-\begin{small}
-\begin{verbatim}
-inject :: (sub :<: f) 
-       => sub (MonadExp f) a 
-       -> MonadExp f a
-inject a = In $ Oper $ inj $ a
-
-move :: (MoveOp :<: f) => MonadExp f ()
-move =  inject Move
-
-turnL :: (TurnOp :<: f) => MonadExp f ()
-turnL = inject TurnLeft
-
-turnR :: (TurnOp :<: f) => MonadExp f ()
-turnR = inject TurnRight
-
-sensor :: (SensorOp :<: f) => MonadExp f BoolE
-sensor = inject Sensor 
-\end{verbatim}
-\end{small}
-
-\begin{small}
-\begin{verbatim}
-cond :: (CondOp :<: f) 
-     => MonadExp f BoolE 
-     -> MonadExp f () -> MonadExp f () -> MonadExp f ()
-cond b p1 p2 = inject $ Cond b p1 p2
-
-while :: (WhileOp :<: f) 
-      => MonadExp f BoolE 
-      -> MonadExp f () -> MonadExp f ()
-while pb p = inject $ While pb p
-\end{verbatim}
-\end{small}
-
-Things are starting to fall into place. The only remaining bit is to
-complete the compiler for monadic expressions.
-
-\begin{small}
-\begin{verbatim}
-instance (Compile x, Compile (f x)) 
-  => Compile (Mops f x) where
-  compile s (Oper o) = compile s o
-  compile s (Return a) = (a,PSkip)
-  compile s (Bind m f) = (b,prg1 `PSeq` prg2) 
-    where (s1,s2)  = split2 s
-          (a,prg1) = compile s1 m
-          (b,prg2) = compile s2 (f a)
-
-instance (Compile (f (MonadExp f))) 
-  => Compile (MonadExp f) where
-  compile s (In a) = compile s a
-\end{verbatim}
-\end{small}
-
-Now, \studname{} has all the functions he needs to compile a simple
-example program.
-
-\begin{small}
-\begin{verbatim}
-test1 :: Robot () 
-test1 = do 
-  move
-  turnL
-  move 
-  turnL 
-\end{verbatim}
-\end{small}
-
-Compiling the program above gives the expected result.
-
-\begin{small}
-\begin{verbatim}
-PSeq PMove (PSeq PTurnLeft (PSeq PMove PTurnLeft))
-\end{verbatim}
-\end{small}
-
-\studname{} decides to contact \docname{} again to show him what 
-he has done.
-
-\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-Dear Prof. \docname{}
-\newline \newline
-\noindent I have attached a file which demonstrates that the monadic
-constructs can be factored out and compiled separately, just as you
-suggested.
-\newline \newline
-
-\noindent Thanks \newline
-\noindent \studname{}
-\end{minipage} 
-}
-
-\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-\studname{},
-\newline
-
-Fascinating! You've managed to factor out the monadic operations so
-that they can be compiled once and for all. Users of your library
-don't have to be concerned at all with the semantics of bind and return.
-Very nice!
-\newline
-%% \end{minipage}}
-
-%% \noindent\colorbox{light-gray}{
-%% \begin{minipage}[]{0.9\linewidth}
-\noindent 
-I note that the data type for monadic operations cannot be composed as
-freely as the other operations, but instead has to be applied
-separately as a final step. Your
-solution is very similar to how variable binding is handled in the
-Syntactic library \cite{syntactic}.
-\newline \newline
-
-\noindent Prof. \docname{}
-\newline
-\noindent Dept. of Computer Science and Engineering \newline
-\noindent Chalmers University of Technology 
-\end{minipage} 
-}
-
-%One of the advantages of the monad reification technique we have
-%presented is that it is compositional. The monadic constructs can be
-%dealt with separately from all the primitives of the embedded
-%languages.
-
-%A consequence of the compositionality is that we can use techniques
-%such as CompData \cite{} to decompose the definition of DSLs into
-%smaller reusable building blocks. Figure \ref{fig:compdata} shows a
-%compositional implementation of the monadic constructs, separately
-%from any other constructs. The type \verb!Cxt! is defined in the
-%package CompData and ties the recursive knot.
-
-%\begin{figure}
-%\begin{verbatim}
-%data MonadT r a where
-%  Return :: a -> MonadT r a
-%  Bind   :: r a -> (a -> r b) -> MonadT r b
-
-%instance (MonadT :<: f) => Monad (Cxt h f a) where
-%  return a = iReturn a
-%  m >>= f  = iBind m f
-%\end{verbatim}
-%\label{fig:compdata}
-%\caption{Compositional definition monads}
-%\end{figure}
-
-%\end{minipage}
+\emph{Prove the monad laws}\newline \newline
 
@@
 % Infinite programs 
-% Only certain ``kinds'' of reification!
+% Only certains ``kinds'' of reification!
+
+\section{Benefits}
 
@@
 
-%This short story has outlined the discovery process and the many turns from 
-%hopefulness to despair and back to hopefulness of two people who met over 
-%a language embedded in Haskell. 
-%% This was the story about two people meeting over the discovery of a way to 
-%% reify monads. Along the road from idea to implementation there was some ups 
-%% and downs but in the end they did arrive at a compositional 
-%% and reifiable way of constructing monadic EDSLs. The method, albeit not as 
-%% general as briefly hoped for, is an example of a simple and elegant approach 
-%% to the problem.
+The trick which we've shown in this paper scales up nicely to larger
+embedded languages. The language Obsidian already use it.
 
-This paper shows a simple method of implementing monadic EDSLs. It shows a
-na\"ive approach to the monad reification problem which has the additional benefit of being compositional. In section 
-\ref{sec:composing}, the language is reimplemented in an extensible way and 
-the compile function is shown explicitly to be compositional by treating sub 
-parts of the language separately.
+% we should always point of the positives as well!
 
-The example language used throughout the story is small and quite
-limited. But the technique presented in this paper does scale up to
-larger languages and more complicated language constructs.  It is
-currently used in the implementation of Obsidian, an EDSL for general
-purpose GPU programming \cite{Obsidian}. Obsidian has language
-constructs which are considerably more advanced than the robot
-language is this paper. One example is the sequential for-loop which
-is higher-order (the type parameter {\tt t} is not relevant to the
-current discussion):
-\begin{small}
-\begin{verbatim}
-SeqFor :: EWord32 -> (EWord32 -> Program t ())
-       -> Program t ()
-\end{verbatim}
-\end{small}
-There does, however, seem to be restrictions on what kind of language
-constructs the reification technique presented in this paper can deal
-with. For instance, the functions from the {\tt MonadPlus} class has
-resisted our attempts at adding them to an EDSL in the same way as
-we've added {\tt Return} and {\tt Bind}. The exact limits of the
-expressiveness of our method is currently unknown to us.
-
-%figure~\ref{fig:spiral}, shows the output generated from running the 
- %{\tt spiralIn} program.
-
-%\studname{} feels motivated to try writing some program that shows a little bit more
-%intelligence in the robot. A small program that makes sensible decisions based
-%on the sensor information is implemented.
-
-%\begin{figure}
-%\includegraphics[width=.49\linewidth]{./spiral1}
-%\includegraphics[width=.49\linewidth]{./spiral2}
-%\caption{Images generated by our robot simulator using the {\tt spiral} program.An important detail is that the robot simulator does not evaluate the Monadic program but rather executes a compiled first-order representation of that program.}
-%\label{fig:spiral}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=.49\linewidth]{./wall1}
-%\includegraphics[width=.49\linewidth]{./wall2}
-%\caption{Images generated by the robot simulator using the {\tt followWall} program.}
-%\label{fig:followwall}
-%\end{figure}
-
-%\FloatBarrier
-%Set the mood. \docname{} is more of a deep ``Theory'' kind of person
-
-%Suppose you have just bought yourself a little programmable toy robot
-%and you want to design a small DSL in Haskell to program it. The robot
-%can turn, move forward and has a sensor to detect whether it is facing
-%an obstacle or not. Inspired by the robot language in \cite{} you come
-%up with a set of primitives shown in figure
-%\ref{fig:interface}. In particular, you would like to use the
-%do-notation to sequence programs.
-
-%\begin{figure} 
-%\begin{itemize} 
-%  \item \verb!move      :: Program ()! 
-%  \item \verb!turnLeft  :: Program ()!
-%  \item \verb!turnRight :: Program ()!
-%  \item \verb!sensor    :: Program Bool!
-%  \item \verb!cond      :: Program Bool -> Program () -> Program () -> Program ()!
-%  \item \verb!while     :: Program Bool -> Program () -> Program ()!
-%  \item \verb!instance Monad Program!
-%\end{itemize} 
-%\label{fig:interface} 
-%\caption{Proposed set of basic robot operations} 
-%\end{figure}
-
-%\emph{Example programs}
-
-%The next task is to implement this language. Compiling EDSLs like this
-%to something which the physical robot can execute is well know and
-%there are off-the-shelf techniques to use. But there is one quirk; how
-%does one generate an abstract syntax tree from a type which implements
-%the Monad interface?
-
-%Since you want to compile the language there must be an abstract
-%syntax tree representation of some sort of the language. The natural
-%thing is just to create a new type where each language construct has
-%its own constructor. But should you deal with the monadic constructs?
-%Well, the simplest solution would be to just add them to the type as well. The resulting type
-
-%\begin{figure}
-%\begin{verbatim}
-%data BoolE = Lit Bool
-%         | Var String
-%         | (:||:) BoolE BoolE
-%         | (:&&:)  BoolE BoolE
-%         | Not BoolE
-
-%data Program a where
-%  Move      :: Program ()
-%  TurnRight :: Program ()
-%  TurnLeft  :: Program ()
-
-%  Sensor    :: Program BoolE
-%  Cond      :: BoolE -> Program () -> Program () -> Program ()
-%  While     :: Program BoolE -> Program () -> Program ()
-%   
-%  Return    :: a -> Program a 
-%  Bind      :: Program a -> (a -> Program b) -> Program b
-%\end{verbatim}
-%\label{fig:program}
-%\caption{A data type for programs}
-%\end{figure}
-
-%\emph{More text}
-
-%In the rest of this paper we will see that this na\"ive method
-%actually can be made to work and that it's a practical, compositional
-%technique for reifying monads.
+% we should always point of the positives as well!

### 2013-09-30 src2/bb/partINew.tex
another approach

--- 2013-09-30 src2/bb/body.tex
@@
 \studname{} is a computer science student working in a project involving programming 
-robots in a low-level imperative language. However, \studname{} has a budding 
+robots in a low level imperative language. However, \studname{} has a budding 
 interest in functional programming using Haskell and has read ``The Haskell 
-school of expression'' \cite{HUDAK}. He gets the idea to implement 
-a language for robot control embedded in Haskell. \studname{} realises that the 
-capabilities of the robot hardware are very similar to those of the 
+school of expression'' \cite{HUDAK}. Naturally he gets the idea to implement 
+a language for robot control embedded in Haskell. \studname{} realise that the 
+capabilities of their robot hardware is very similar to those of the 
 robots that Hudak evaluates graphically in a grid world. There is however 
@@
 Guided by the capabilities of the target robot, \studname{} designs an API for 
-robot programming. The robot can perform operations 
+robot programming. Amongst the capabilities of the robot are operations 
 such as {\em move} that steps the robot forward and {\em turn} left and right. 
-The robot also has the capability to execute program loops and conditionals and 
-it has a forward facing {\tt sensor} with which it can query the world.
+The robot also has a forward facing {\tt sensor} with which it can query the 
+world and also the capabilities to execute program loops and conditionals.
 
-%The robot also has a forward facing {\tt sensor} with which it can query the 
-%world and also the capabilities to execute program loops and conditionals. 
-\pagebreak
 The API that \studname{} designs is simple:
@@
 sensor :: Program Bool 
-cond      :: Program Bool 
-             -> Program () 
-             -> Program () 
-             -> Program () 
-while     :: Program Bool 
-             -> Program () 
-             -> Program () 
+cond :: Bool -> Program () -> Program () -> Program () 
+while :: Program Bool -> Program () -> Program () 
 \end{verbatim}
@@
 
-Now, \studname{} wants to express robot programs using the Haskell
-{\em do} notation. The motivation behind this is the imperative look
-and feel of the operations he identified and the potential to use all
-the control structures in the {\em Control.Monad} library. For this, a
-{\tt Monad} instance is needed.
-
-To be able to create a first-order representation of a computation
-described using monads, \studname{} needs to solve the problem of {\em
-  reifying} monads. This is the story of how \studname{} came up with
-a particularly simple solution to this problem.
-
-%\emph{*** Josef, Think about writing here}
+Now, \studname{} wants to express robot programs using the Haskell {\em do} 
+notation. The motivation behind this is the imperative look and feel 
+of the operations he identified and the potential to use all the control structures 
+in the {\em Control.Monad} library. For this, a {\tt Monad} instance is needed.
 
 %\begin{minipage}[]{\linewidth}
-%\vspace{5mm}
+\vspace{5mm}
 
-%\fbox{%
-%\parbox{0.8\linewidth}{What \studname{} does not know at this time is that he has stumbled over the 
-%problem of {\em Monad reification}. Monad reification is the observing of the shape of the computation and this is necessary in order to compile to some low level representation.}
-%}
+\fbox{%
+\parbox{0.8\linewidth}{What \studname{} does not know at this time is that he has stumbled over the 
+problem of {\em Monad reification}. Monad reification is the observing of the shape of the computation and this is necessary in order to compile to some low level representation.}
+}
 %\end{minipage}
@@
 \subsection{Example programs}
-
-\studname{} is quite happy with his language design and decides to
-write some programs to try it out.
+After listing the desired basic operations of the language, \studname{} tries to write 
+a few very simple example programs.
 %experiments 
@@
 
-The first program is called {\tt sMove}, which implements a safe move 
+The first program written is one called sMove, that implements a safe move 
 operation. This operation can be executed by the robot even though it is facing 
-an obstacle, without risk of harming robot or obstacle. 
+an obstacle without risk of harming robot or obstacle. 
 %.It feels like a good idea to \studname{} to check the sensor before trying to 
@@
 sMove :: Program () 
-sMove = cond sensor turnRight move 
+sMove = do 
+  s <- sensor 
+  cond s turnRight move 
 \end{verbatim}
@@
 
-The second program moves the robot forward until it stands directly 
+The second program he writes moves the robot forward until it stands directly 
 in front of an obstacle such as a wall.
@@
 
-% Since an abstract syntax representation of the robot programs is desired,
-Since the language is going to be compiled, \studname{} realises that the 
-booleans in his language cannot be regular Haskell booleans. The booleans 
-needs to be replaced by boolean typed expressions.
+Since an abstract syntax representation of the robot programs is desired, \studname{} 
+realises that the booleans of some of the operations needs to be replaced by 
+boolean typed expressions.
 
-\pagebreak
 \begin{small} 
 \begin{verbatim}
-type Name = String
-
 data BoolE = Lit Bool
-           | Var Name
+           | Var String
            | (:||:) BoolE BoolE
@@
 
-Following this, the  operations that should go into the {\tt Program} data type 
-feel straightforward.
+following this, the  operations that should go into the {\tt Program} data type 
+feels straightforward.
 
@@
   Sensor :: Program BoolE
-  Cond      :: Program BoolE 
-               -> Program () 
-               -> Program () 
-  While     :: Program BoolE 
-               -> Program () 
-               -> Program () 
+  Cond :: BoolE -> Program () -> Program () 
+  While :: Program BoolE -> Program () -> Program () 
 \end{verbatim} 
@@
 \studname{} continues by naively adding constructors for {\tt Return} and 
-{\tt Bind} to the {\tt Program} data type.
+{\tt Bind} to the {\tt Program a} data type.
 
@@
   Return :: a -> Program a 
-  Bind   :: Program a 
-            -> (a -> Program b)
-            -> Program b 
+  Bind :: Program a -> (a -> Program b) -> Program b 
 \end{verbatim} 
@@
 
-Proud of his accomplishments, \studname{} sends his Haskell module in an 
-email to Prof. \docname{}. \studname{} knows that \docname{} is teaching an 
-introductory Haskell course and should be able to provide feedback.
+Proud of his accomplishments, \studname{} sends his Haskell module in an email to Prof. \docname{}.
+\studname{} knows that \docname{} is teaching an introductory Haskell course and should be 
+able to provide feedback.
 
-\subsection{Problem statement} 
-\emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline
+\section{Problem statement} 
+\emph{Meanwhile in Professor \docname{}'s Office}\newline \newline
 
-\noindent Prof. \docname{} notices an email from \studname{} in his inbox and 
-opens it.
+\noindent Prof. \docname{} notices an email from \studname{} in his inbox and opens it.
 
@@
 programming and thought it'd be natural to try implementing an EDSL. 
-I know you teach an FP course and thought I would ask for your input. 
-Attached to this mail is a file containing an outline of the data types I 
+I know you teach an FP course and thought id ask for your input. 
+Atached to this mail is a file containing an outline of the data types I 
 want to use. Does this look sensible to you? \newline \newline
@@
 particularly notices the constructors {\tt Return} and {\tt Bind} and the 
-{\tt Monad} instance. He shakes his head at \studname{}'s naivet\'e and writes an 
+{\tt Monad} instance. He shakes his head at \studname{}'s naivety and writes an 
 email back:
 
-%\vspace{5mm}
+\vspace{5mm}
 
@@
 \noindent I'm afraid your implementation of the monadic primitives can never 
-work. The constructor {\tt Return} can take any arbitrary value that has 
+work. The constructor {\tt Return} can take any arbitrary value which has 
 nothing to do with the language you're designing. These values may be 
-strings, binary trees or higher-order functions from zygomorphisms to 
-histomorphisms. The same problem goes for {\tt Bind}; it has to be able 
+strings, binary trees of higher order functions from zygomorphisms to 
+histomorphisms. The same problem goes for {\tt Bind}, it has to be able 
 to handle arbitrary values. It cannot possibly work! This is a very 
 difficult problem and such a naive solution is bound to fail.
-\newline \newline
-%% \noindent For hints towards a solution to this I suggest that you follow 
-%% this \underline{link} \cite{GillBlog}. Another idea is that you refer 
-%% to the article ``Generic monadic constructs for embedded languages'' 
-%% \cite{Generic}.
-%% \newline \newline
+
+\noindent For hints towards a solution to this I suggest that you follow 
+this \underline{link} \cite{GillBlog}. Another idea is that you refer 
+to the article ``Generic monadic constructs for embedded languages'' 
+\cite{Generic}. \newline \newline
 
@@
 
-% \section{Implementation continues}
-% \emph{Back to \studname{}'s implementation efforts} \newline \newline
+\section{Implementation continues}
+\emph{Back to \studname{}'s implementation efforts} \newline \newline
 
-\noindent While waiting for feedback from \docname{}, \studname{} has 
-carried on trying to implement a compiler for his language.
+\noindent While waiting for feedback from \docname{}, \studname{} has carried on trying to implement
+a compiler for his language.
 
-When \docname{}'s email does arrive, \studname{} is confused. He feels he 
-has already managed to compile the {\tt Program} data type into a 
-representation closer to that which is executed by the robot hardware.
+When \docname{}'s email does arrive, \studname{} is confused. He feels he has already managed 
+to compile the {\tt Program a} data type into a representation closer to that 
+which is executed by the robot hardware.
 
@@
 \newline \newline
-\noindent Thanks for your feedback. But I think it does work! Attached
-to this email you find a file containing my attempt at compiling the
-{\tt Program} data type into a representation closer to that which is
-executed by our robots. \newline \newline
-
-\noindent I have also attached two example programs that you can compile and 
-then run in the graphical simulator (see figure~\ref{fig:programs}) \newline \newline
-%\vspace{5mm}
-
-%\includegraphics[width=.49\linewidth]{./spiral1}
-%\includegraphics[width=.49\linewidth]{./spiral2}
-
-%\vspace{5mm}
-
-%\noindent And these two images are output from running the 
-%{\tt followWall} (see figure~\ref{fig:followWall}) program. \newline
-
-%\vspace{5mm}
-
-%\includegraphics[width=.49\linewidth]{./wall1}
-%\includegraphics[width=.49\linewidth]{./wall2}
-
-%\vspace{5mm}
+\noindent Thanks for your feedback. But I think it does work! attached 
+to this email (Compiler.hs) you find my attempt at compiling the 
+{\tt Program a} into a representation closer to that which is executed 
+by our robots. \newline \newline
 
@@
 
-\FloatBarrier
-
-\begin{figure*}
-\begin{tabular}{c|c}
-\begin{minipage}[]{0.45\columnwidth}
-\begin{small}
-\begin{verbatim}
-spiralIn :: Int -> Program () 
-spiralIn 0 = return () 
-spiralIn n = do
-  replicateM_ 2 $ do
-    replicateM_ n move
-    turnLeft
-  spiralIn (n-1) 
-\end{verbatim}
-\end{small}
-\end{minipage}
-\hspace{0.05\columnwidth} 
-\begin{minipage}[]{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{./bb/spiral1}
-\end{minipage}
-&
-\begin{minipage}[]{0.45\columnwidth}
-\begin{small}
-\begin{verbatim} 
-followWall :: Program () 
-followWall = 
-  while (return true) $ 
-    cond checkLeft sMove $ 
-            do turnLeft
-               move
-
-checkLeft :: Program BoolE     
-checkLeft = do
-  TurnLeft
-  s <- Sensor
-  TurnRight 
-  return s
-\end{verbatim}
-\end{small}
-\end{minipage}
-\hspace{0.05\columnwidth}
-\begin{minipage}[]{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{./bb/wall2}
-\end{minipage}
-
-\end{tabular}
-\caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
-to \docname{} in an email}
-\label{fig:programs}
-\end{figure*}
+\vspace{5mm}
 
 \section{Compilation of the monadic robot EDSL} 
-\emph{\docname{} receives \studname{}'s latest email}\newline %\newline
+\emph{\docname{} receives \studname{}'s latest email}\newline \newline
 
 \noindent \docname{} looks through the {\tt Compiler} module and finds a data type describing
-a first-order representation of the robot language.
+a first order representation of the robot language.
 
@@
          | PTurnLeft
-         | PSensor Name  
+         | PSensor Var    
          | PCond BoolE Prg Prg
-         | PWhile Name Prg Prg 
+         | PWhile Var Prg Prg 
          | PSeq Prg Prg
          | PSkip       
-         | PAssign Name BoolE
+         | PAssign Var BoolE
 \end{verbatim}
@@
 
-\begin{figure} 
+From the {\tt Prg} data type, \docname{} presumes that the {\tt PSeq} constructor 
+will replace occurences of {\tt Bind} and that the variable binding of the 
+{\tt Sensor} construct is reified using {\tt PAssign}. The conclusion is
+that the representation looks sensible but he is very interested in seeing 
+the {\tt Program a -> Prg} transformation.
+
 \begin{small} 
 \begin{verbatim}
-newNameSupply :: NameSupply 
-split2 :: NameSupply -> 
-          (NameSupply,NameSupply) 
-split3 :: NameSupply -> 
-          (NameSupply,NameSupply,NameSupply)
-supplyValue :: NameSupply -> Int 
-\end{verbatim} 
-\end{small}
-\caption{The interface of the splittable name supply used in the implementation 
-of the compile functions} 
-\label{fig:namesupply}
-\end{figure}
+compile :: Program a -> Prg
+compile prg = snd $ compile' s prg 
+  where 
+    s = unsafePerformIO $ newEnumSupply
 
-From the {\tt Prg} data type, \docname{} presumes that the {\tt PSeq}
-constructor will be the result of compiling {\tt Bind} and that
-{\tt PSensor} binds a variable which can be used in {\tt PAssign} and
-{\tt PWhile}. The conclusion is that the representation looks
-sensible, but he is very interested in seeing the {\tt Program a -> Prg}
-transformation. The compile function assumes a splittable name
-supply with the interface described in figure \ref{fig:namesupply}.
-
-\pagebreak
-\begin{small} 
-\begin{verbatim}
-runCompile :: Program a -> Prg
-runCompile prg = snd $ compile s prg 
-  where 
-    s = newNameSupply
-
-compile :: NameSupply -> Program a -> (a, Prg) 
-compile s Move = ((),PMove)
-compile s TurnRight = ((),PTurnRight)
-compile s TurnLeft  = ((),PTurnLeft)
-compile s Sensor    = (Var nom,PSensor nom)
+compile' :: Supply Int -> Program a -> (a, Prg) 
+compile' s Move = ((),PMove)
+compile' s TurnRight = ((),PTurnRight)
+compile' s TurnLeft  = ((),PTurnLeft)
+compile' s Sensor    = (Var nom,PSensor nom)
   where
@@
     nom = "v" ++ show v 
-compile s (Cond b p1 p2) = 
-    ((),bp `PSeq` PCond b' p1' p2') 
+compile' s (Cond b p1 p2) = ((), PCond b p1' p2') 
+  where
+    (s1,s2) = split2 s
+    (a1,p1') = compile' s1 p1
+    (a2,p2') = compile' s2 p2 
+compile' s (While wp prg) = ((),PWhile nom nwp prg') 
   where
     (s1,s2,s3) = split3 s
-    (b',bp)    = compile s1 b
-    (a1,p1')   = compile s2 p1
-    (a2,p2')   = compile s3 p2
-compile s (While wp prg) = ((),PWhile nom nwp prg') 
-  where
-    (s1,s2,s3) = split3 s
-    (b,wp')    = compile s1 wp
-    (c,prg')   = compile s2 prg 
+    (b,wp') = compile' s1 wp
+    (c,prg') = compile' s2 prg 
     nom  = "v" ++ (show $ supplyValue s3)
     nwp = (wp' `PSeq` PAssign nom b) 
-compile s (Return a) = (a,PSkip)
-compile s (Bind pa f) = (b, prg1 `PSeq` prg2) 
+compile' s (Return a) = (a,PSkip)
+compile' s (Bind pa f) = (b, prg1 `PSeq` prg2) 
   where
     (s1,s2) = split2 s
-    (a,prg1) = compile s1 pa
-    (b,prg2) = compile s2 (f a) 
+    (a,prg1) = compile' s1 pa
+    (b,prg2) = compile' s2 (f a) 
 \end{verbatim}
@@
 
-%\begin{small} 
-%\begin{verbatim}
-%compile :: Program a -> Prg
-%compile prg = snd $ compile' s prg 
-%  where 
-%    s = unsafePerformIO $ newEnumSupply
-%    
-%compile' :: Supply Int -> Program a -> (a, Prg) 
-%compile' s Move = ((),PMove)
-%compile' s TurnRight = ((),PTurnRight)
-%compile' s TurnLeft  = ((),PTurnLeft)
-%compile' s Sensor    = (Var nom,PSensor nom)
-%  where
-%    v = supplyValue s
-%    nom = "v" ++ show v 
-%compile' s (Cond b p1 p2) = ((), PCond b p1' p2') 
-%  where
-%    (s1,s2) = split2 s
-%    (a1,p1') = compile' s1 p1
-%    (a2,p2') = compile' s2 p2 
-%compile' s (While wp prg) = ((),PWhile nom nwp prg') 
-%  where
-%    (s1,s2,s3) = split3 s
-%    (b,wp') = compile' s1 wp
-%    (c,prg') = compile' s2 prg 
-%    nom  = "v" ++ (show $ supplyValue s3)
-%    nwp = (wp' `PSeq` PAssign nom b) 
-%compile' s (Return a) = (a,PSkip)
-%compile' s (Bind pa f) = (b, prg1 `PSeq` prg2) 
-%  where
-%    (s1,s2) = split2 s
-%    (a,prg1) = compile' s1 pa
-%    (b,prg2) = compile' s2 (f a) 
-%\end{verbatim}
-%\end{small}
-
-\docname{} studies the code in disbelief, saves the module and tries
-it out on some examples. The code does indeed seem to work and
-\docname{}'s disbelief is replaced with enthusiasm; this appears to be
-a simple and convenient way to reify monads.
-%What's more it does so in a 
-%compositional way. The compilation of the {\tt Return} and {\tt Bind} cases 
-%are completely independent of any other construct of the language.
+\docname{} studies the code in disbelief but saves the module to his harddrive 
+and tries it out on some examples. The code does indeed seem to work and 
+\docname{}'s disbelief is replaced with enthusiasm, this appears to be a simple 
+and conveniant way to reify monads. what's more it does so in a 
+compositional way. The compilation of the {\tt Return} and {\tt Bind} cases 
+are completly independent of any other construct of the language.
 
@@
 
-\section{Meeting in \docname{}'s office} %\section{Related Work} 
-%\emph{Meeting in \docname{}'s office} \newline
+\section{Related Work} 
+\emph{Meeting in \docname{}'s office} \newline
 
@@
 
-\speak{\docname{}} So, about this robot language! You gave me a bit of a 
+\speak{\docname{}} So, about this robot language! you gave me a bit of a 
 surprise there. I was entirely sure that what you did was impossible.
@@
 \speak{\docname{}} The problem is \direct{\docname{} approaches his whiteboard}
- when you try to reify the bind constructor of your representation.
+ when you try to reify the bind constructor of your representation..
 \end{dialogue}
@@
 \begin{dialogue}
-\speak{\docname{}} You need to come up with an element of type {\tt a} to 
-pass to the function.
+\speak{\docname{}} you need to come up with an element of type a to pass to the 
+function.
 
@@
 
-\speak{\docname{}} But it is! Your solution is that you are careful
-about the return types of all operations in your language. They are
-either of unit type or some type which can be guaranteed to be reified.
-\direct{\docname{} scribbles on his whiteboard}
+\speak{\docname{}} But it is. I have never seen a solution as naive as yours 
+though and was quite surprised that it works.
+
+\speak{\docname{}} For example, in this paper \cite{Generic}, the authors use 
+a continuation monad to be able to reify the monadic constructs of their 
+language
+
+\end{dialogue}
+\emph{Have \docname{} outline the continuation monad approach on the whiteboard}
+
+%\emph{Maybe here kurt can show related work and explain why these related
+%approaches exist\cite{GillBlog}
+%\cite{Farmer}
+%\cite{Generic}
+%\cite{Wouter}
+%\cite{CompData}
+%\cite{Compile} 
+%\cite{MonadReader15}}
+
+\begin{dialogue}
+\speak{\docname{}} Your solution is that you are careful about the return types 
+of other operations in your language. They are either of unit type or some 
+in itself reifyable expression type. \direct{\docname{} scribbles on  his whiteboard}
 
@@
 
+\emph{lots more text here}
+
 \begin{dialogue}
-\speak{\docname{}} This is different from how such constructs are
-normally implemented. If I had implemented the language I would have
-given {\tt Sensor} the type {\tt Program Bool} so that I could write
-an evaluator of type {\tt Program a -> a}, or perhaps {\tt Program a -> M a}.
-But then adding a monad to the language and reifying it becomes much harder.
 
-\speak{\studname{}} Yes, now that you mention it, I have seen that style used
-for defining DSLs.
+\speak{\docname{}} About compositionality, I think that we can make this explicit 
+by using an approach based on \cite{Wouter} or \cite{CompData}.
 
-\speak{\docname{}} Having constructors like {\tt Sensor} return {\tt
-  BoolE} instead of {\tt Bool} is a crucial part of why your
-method works. It allows the {\tt compile} function to be written in
-such a way that it can generate the return values statically and pass
-them as arguments to {\tt Bind}. In a sense you're doing evaluation at
-the same time as compilation. There's just enough evaluation to
-remove the monadic constructs. It's a very neat trick!
+\speak{\docname{}}  Bye!  
+\end{dialogue}
+%\end{minipage}
 
-\speak{\studname{}} Thank you.
+\section{Execution of the compiled representation} 
+\emph{The discussion turns to practical}\newline \newline 
+\FloatBarrier
 
-\speak{\docname{}} Another way to think about your trick is that you are
-using a writer monad to evaluate your language and produce the first-order
-syntax tree as a side effect. Then you could have used a state monad
-transformer on top of that for the fresh name supply. But that's a stylistic
-choice.
+%\begin{minipage}[]{\linewidth}
+\begin{dialogue}
 
-\speak{\studname{}} You seem to be making a big deal out of this. I
-just did what I thought was the most straightforward thing to do.
+\speak{\docname{}} Let us see if we can implement some more interesting 
+example programs in your language. Can you do spiral ?
+
+\speak{\studname{}} No. Infinite programs are ruled out by my compilation 
+function.
+
+\speak{\docname{}} It is true that your compilation function rules out 
+infinite programs but that is necessary to rule out such programs anyway 
+if you want them to compile for our robot hardware. 
+\speak{\docname{}} Write a spiralIn program with a well defined termination 
+condition.
+
+\direct{\studname{} opens his laptop and starts hacking..} 
+\end {dialogue} 
+% \end{minipage}
+
+\begin{small}
+\begin{verbatim}
+spiralIn :: Int -> Program () 
+spiralIn 0 = return () 
+spiralIn n = do
+  replicateM_ 2 $ do
+    replicateM_ n move
+    turnLeft
+  spiralIn (n-1) 
+\end{verbatim}
+\end{small}
+
+% \begin{minipage}[]{\linewidth}
+\begin{dialogue}
+
+\speak{\docname{}} yes. That should work. Do we have any way to verify this 
+program.
+
+\speak{\studname{}} We can run it through the graphical simulator!
+
+\speak{\docname{}} Show me!
 
@@
 
-\subsection{Related work}
+\includegraphics[width=.49\linewidth]{./spiral1}
+\includegraphics[width=.49\linewidth]{./spiral2}
 
@@
 
-\speak{\docname{}} Others have solved this problem before but I have
-never seen a solution as simple as yours and was quite surprised that
-it works. For example, in this paper \direct{\docname{} shows
-  \studname{} the paper \cite{Generic}}, the authors use a
-continuation monad to be able to reify the monadic constructs of their
-language. If they were to implement your robot language they would use
-the same \texttt{Program} data type as you do. However, they would not
-expose that to the user of the language. Instead they would create a
-type like this:
+\speak{\docname{}} Nice!
+
+\speak{\docname{}} Now, let us try something more advanced. Can you program 
+the robot to always keep a wall on his left hand side?
 
@@
 \begin{verbatim} 
-data P a =
-  P (forall r. ((a -> Program r) -> Program r))
+followWall :: Program () 
+followWall = do
+  while (return true) $ do
+    b <- checkLeft
+    cond b sMove (turnLeft >> move) 
 \end{verbatim}
@@
 \begin{dialogue} 
-\speak{\docname{}} The type \texttt{P} is a continuation monad so the
-monad instance comes naturally. Operations on \texttt{P} are defined
-using \texttt{Bind} like this:
+\speak{\docname{}} Yes. that looks about right. Let us see it. 
 \end{dialogue}
 
-\begin{small}
-\begin{verbatim}
-while :: P BoolE -> P () -> P ()
-while c b =
-  P (\k -> While (runP c) (runP b) `Bind` k)
-\end{verbatim}
-\end{small}
-%\pagebreak
-\begin{small}
-\begin{verbatim}
-cond :: BoolE -> P () -> P () -> P ()
-cond b t e =
-  P (\k -> Cond b (runP t) (runP e) `Bind` k)
-
-runP :: P a -> Program a
-runP (P f) = f (\a -> Return a)
-\end{verbatim}
-\end{small}
-
-\begin{dialogue}
-\speak{\studname{}} I see. Continuations are quite magical to me. I
-could never have come up with that technique. How do they deal with
-transforming higher-order programs to first-order programs?
-
-\speak{\docname{}} That's a good question. The transformation to first
-order programs is dealt with by a library called Syntactic. It is
-described in a separate paper \direct{\docname{} pulls out the paper
-  \cite{axelsson2012generic} from a pile of papers}. It's hard to do
-an apples to apples comparison between Syntactic and your
-technique. Syntactic solves a much bigger problem than what you're
-doing so it is naturally more complicated.
-
-\speak{\studname{}} Ok.
-
-\speak{\docname{}} There are also the papers
-\cite{Farmer,sculthorpe2013constrained} which solves
-the problem in a manner that is more similar to yours. They also have explicit
-constructors {\tt Bind} and {\tt Return} but give them a slightly different
-type which guarantees that all applications of {\tt Bind} are {\em normalised},
-so that all {\tt Bind}s are right-associated. Unexpectedly, this
-allows them to make instances of the {\tt Monad} class but at the same time
-constrain the arguments of {\tt Bind}.
-Their technique is more complicated than yours, but just as with the
-case of Syntactic, they solve a more general problem.
-
-However, there is one particular thing I like about your technique.
-
-\speak{\studname{}} What is that?
-
-\speak{\docname{}} Your technique is compositional.
-
-\end{dialogue}
-
-\subsection{Compositionality}
-
-\begin{dialogue} 
-\speak{\studname{}} What does it mean that my technique is compositional?
-
-\speak{\docname{}} The compile function is compositional because there
-is one case for each constructor and each case deals with exactly one
-constructor. In particular, the constructors {\tt Return} and {\tt
-  Bind} are handled completely separately from all other language
-constructs.
-
-\speak{\studname{}} And that is good ?
-
-\speak{\docname{}} Absolutely! Compositional definitions are nice
-because they imply that there is no weird semantical interaction
-between the constructs as they are defined independently. But it also
-means that it should be possible to factor out the constructors
-\texttt{Return} and \texttt{Bind} into a data type of their own. Then,
-it could be combined with other data types using techniques like Data
-Types \'{a} la Carte \cite{Wouter} or CompData \cite{CompData}. This
-way, a language can be designed piece by piece.You can then select the
-set of pieces required for a particular task or that suits a
-particular brand of robots.
-
-\speak{\studname{}} Oh! Interesting.
-\end{dialogue}
-
-\subsection{The Monad laws}
-
-\begin{dialogue}
-\speak{\docname{}} There is still one problem with your method though.
-
-\speak{\studname{}} What's that?
-
-\speak{\docname{}} When we make instances of the monad class we expect
-certain laws to hold. \direct{\docname{} writes the laws on his whiteboard}
-
-\end{dialogue}
-
-\begin{small}
-\begin{verbatim}
-m >>= return    = m
-return a >>= f  = f a
-(m >>= f) >>= g = m >>= \a -> f a >>= g
-\end{verbatim}
-\end{small}
-
-\begin{dialogue}
-\speak{\docname{}} These laws clearly don't hold for your instance. Take the
-first law for instance. The left hand side will have extra \texttt{Bind} and
-\texttt{Return} constructors compared to the right hand side.
-
-\speak{\studname{}} Hmmm. I hadn't really thought about that. But adding an
-extra return at the end of a computation shouldn't make a difference
-in my implementation.
-
-\speak{\docname{}} So, you're saying that your implementation actually obeys
-the first monad law?
-
-\speak{\studname{}} Well, at least when I run my programs I will never see
-any difference between the left hand side and the right hand side.
-
-\speak{\docname{}} Aha, so what you're saying is that if we compare the
-semantics of programs rather than comparing the programs themselves
-then we get some useful laws. \direct{\docname{} scribbles some new laws
-  on the whiteboard}
-
-\end{dialogue}
-%\end{minipage}
-
-\begin{small}
-\begin{verbatim}
-eval (m >>= return)    = eval m
-eval (return a >>= f)  = eval (f a)
-eval ((m >>= f) >>= g) = eval (m >>= \a -> f a >>= g)
-\end{verbatim}
-\end{small}
-
-%\begin{minipage}[]{\linewidth}
-\begin{dialogue}
-\speak{\docname{}} These laws are morally the same as the monad
-laws, especially if we don't let the user of the robot language ever
-compare terms in the language.
-
-\speak{\studname{}} Yes, that captures my intuition very well.
-
-\speak{\docname{}} Ok, good. Can you prove these equations?
-
-\speak{\studname{}} No, I don't have any experience proving programs correct.
-
-\speak{\docname{}} Well, it shouldn't be that difficult. Let me see
-what I can come up with.
-
-\direct{\docname{} scribbles frenetically on a piece of paper for a
-  couple of minutes.}
-
-Aha! Your technique is quite general. It can reify
-monads into any kind of structure which is a monoid. Return translates
-into the monoid unit and bind translates into the monoid operation.
-
-For example, in your compilation function it is important that the semantics of
-\texttt{PSkip} is the identity of the semantics of \texttt{PSeq} and that
-\texttt{PSeq}
-is associative.
-\speak{\studname{}} Ok, that sounds good. I take that as meaning the method
-is quite general?
-
-\speak{\docname{}} Yes, requiring a monoid is a very mild restriction.
-
-Look at the time! This was interesting, I got quite carried away. We must 
-round off but please get back to me if you make any more progress on your 
-robot language.
-
-\speak{\studname{}} Thanks very much for your time. Bye!
-\speak{\docname{}} Thank you.
-
-\end{dialogue}
-
-\section{Composing reifiable monadic languages} 
-\label{sec:composing}
-
-\emph{At \studname{}'s computer} \newline
-
-\studname{} was really intrigued by the idea of compositionally
-building embedded languages and after having read the papers
-\docname{} showed him he decides to try his own approach to the problem.
-
-Inspired by CompData he starts out by adding the code in figure
-\ref{fig:compdata} to his file. The data type \texttt{:+:} is used for
-composing languages, and the \texttt{:<:} typeclass provides coercions
-so that the programmer doesn't have to worry about using the right
-sequence of the \texttt{InjL} and \texttt{InjR} constructors to inject
-terms into the composed language.
-
-\begin{figure}
-\begin{small}
-\begin{verbatim}
-data (e1 :+: e2) x a 
-  = InjL (e1 x a) | InjR (e2 x a)
-infixr :+:
-
-class sub :<: sup where
-  inj :: sub x a  -> sup x a
-
-instance f :<: f where
-  inj = id
-
-instance (f :<: (f :+: g)) where
-   inj = InjL
-
-instance (f :<: h) => (f :<: (g :+: h)) where
-  inj = InjR . inj 
-\end{verbatim}
-\end{small}
-\caption{Data types from CompData for composing languages.}
-\label{fig:compdata}
-\end{figure}
-
-\studname{} then starts to add data types for the different language
-constructs in his robotic language. The different operations for the
-robot are pleasingly easy to add.
-%\pagebreak
-\begin{small}
-\begin{verbatim}
-data MoveOp x a where 
-  Move :: MoveOp x ()
-
-data TurnOp x a where 
-  TurnLeft  :: TurnOp x () 
-  TurnRight :: TurnOp x ()
-
-data SensorOp x a where 
-  Sensor :: SensorOp x BoolE
-
-data CondOp x a where 
-  Cond :: x BoolE -> x () -> x () -> CondOp x ()
-
-data WhileOp x a where 
-  While :: x BoolE -> x () -> WhileOp x () 
-\end{verbatim}
-\end{small}
-
-In order to test out these definitions, \studname{} next turns to
-implementing a compiler. He realises that the compiler now needs to be
-implemented as a class with one instance per compilable sub-language.
-
-\begin{small}
-\begin{verbatim}
-runCompile :: Compile f => f a -> Prg
-runCompile prg = snd $ compile s prg
-  where s = newNameSupply
-
-class Compile f where
-  compile :: NameSupply -> f a -> (a, Prg)
-
-instance Compile (MoveOp x) where
-  compile _ Move = ((), PMove)
-
-instance Compile (TurnOp x) where
-  compile _ TurnLeft = ((), PTurnLeft)
-  compile _ TurnRight = ((), PTurnRight)
-
-instance Compile (SensorOp x) where
-  compile s Sensor = (Var nom, PSensor nom)
-    where v   = supplyValue s
-          nom = "v" ++ show v
-\end{verbatim}
-\pagebreak
-\begin{verbatim}
-instance Compile x => Compile (CondOp x) where
-  compile s (Cond b p1 p2) = 
-    ((),bp `PSeq` PCond b' p1' p2') 
-    where (s1,s2,s3) = split3 s
-          (b',bp)    = compile s1 b
-          (a1,p1')   = compile s2 p1
-          (a2,p2')   = compile s3 p2 
-\end{verbatim}
-\end{small}
-%% instance Compile x => Compile (CondOp x) where
-%%   compile s (Cond b p1 p2) = ((),PCond b p1' p2') 
-%%     where
-%%       (s1,s2) = split2 s
-%%       (a1,p1') = compile s1 p1
-%%       (a2,p2') = compile s2 p2 
-%\pagebreak
-\begin{small}
-\begin{verbatim}
-instance Compile x => Compile (WhileOp x) where
-  compile s (While wp p) = ((),PWhile nom nwp p')
-    where (s1,s2,s3) = split3 s
-          (b,wp')    = compile s1 wp
-          (c,p')     = compile s2 p
-          nom        = "v" ++ (show $ supplyValue s3)
-          nwp        = wp' `PSeq` PAssign nom b
-
-instance (Compile (e1 f), Compile (e2 f)) 
-  => Compile ((e1 :+: e2) f) where
-  compile s (InjL a) = compile s a
-  compile s (InjR a) = compile s a 
-\end{verbatim}
-\end{small}
-
-But when \studname{} tries to add the monadic operations he finds that
-they are quite resistant to a compositional treatment. After much
-struggle he comes up with the following data type definition:
-
-\begin{small}
-\begin{verbatim}
-data Mops f x a where
-  Oper   :: f x a -> Mops f x a
-  Return :: a -> Mops f x a
-  Bind   :: x a -> (a -> x b) -> Mops f x b
-\end{verbatim}
-\end{small}
-
-The recursion is provided by another data type that \studname{} 
-calls {\tt MonadExp}.
-
-\begin{small}
-\begin{verbatim}
-data MonadExp f a = In (Mops f (MonadExp f) a)
-\end{verbatim}
-\end{small}
-
-\studname{} thinks of the {\tt MonadExp} data type as a representation
-of a monadic language parameterised over operations {\tt Oper} that
-are constructed using the constructors of some type {\tt f}. The monad
-instance for this data type is only slightly more complicated compared
-to the earlier, non-compositional setting.
-
-\begin{small}
-\begin{verbatim}
-instance Monad (MonadExp f) where 
-  return = In . Return
-  (>>=) a f = In (Bind a f) 
-\end{verbatim}
-\end{small}
-
-\studname{} can now write the type of his robotic language, built
-from independent pieces. It is noteworthy that the monadic expressions
-have to be added on top of all the other language constructs.
-
-\begin{small}
-\begin{verbatim}
-type Robot = MonadExp (MoveOp :+: TurnOp  :+:
-                       CondOp :+: WhileOp :+:
-                       SensorOp)
-\end{verbatim}
-\end{small}
-
-Given the definition of his language \studname{} can now write an
-injection function which helps writing smart constructors for all the
-different robot operations.
-
-\begin{small}
-\begin{verbatim}
-inject :: (sub :<: f) 
-       => sub (MonadExp f) a 
-       -> MonadExp f a
-inject a = In $ Oper $ inj $ a
-
-move :: (MoveOp :<: f) => MonadExp f ()
-move =  inject Move
-
-turnL :: (TurnOp :<: f) => MonadExp f ()
-turnL = inject TurnLeft
-
-turnR :: (TurnOp :<: f) => MonadExp f ()
-turnR = inject TurnRight
-
-sensor :: (SensorOp :<: f) => MonadExp f BoolE
-sensor = inject Sensor 
-\end{verbatim}
-\end{small}
-
-\begin{small}
-\begin{verbatim}
-cond :: (CondOp :<: f) 
-     => MonadExp f BoolE 
-     -> MonadExp f () -> MonadExp f () -> MonadExp f ()
-cond b p1 p2 = inject $ Cond b p1 p2
-
-while :: (WhileOp :<: f) 
-      => MonadExp f BoolE 
-      -> MonadExp f () -> MonadExp f ()
-while pb p = inject $ While pb p
-\end{verbatim}
-\end{small}
-
-Things are starting to fall into place. The only remaining bit is to
-complete the compiler for monadic expressions.
-
-\begin{small}
-\begin{verbatim}
-instance (Compile x, Compile (f x)) 
-  => Compile (Mops f x) where
-  compile s (Oper o) = compile s o
-  compile s (Return a) = (a,PSkip)
-  compile s (Bind m f) = (b,prg1 `PSeq` prg2) 
-    where (s1,s2)  = split2 s
-          (a,prg1) = compile s1 m
-          (b,prg2) = compile s2 (f a)
-
-instance (Compile (f (MonadExp f))) 
-  => Compile (MonadExp f) where
-  compile s (In a) = compile s a
-\end{verbatim}
-\end{small}
-
-Now, \studname{} has all the functions he needs to compile a simple
-example program.
-
-\begin{small}
-\begin{verbatim}
-test1 :: Robot () 
-test1 = do 
-  move
-  turnL
-  move 
-  turnL 
-\end{verbatim}
-\end{small}
-
-Compiling the program above gives the expected result.
-
-\begin{small}
-\begin{verbatim}
-PSeq PMove (PSeq PTurnLeft (PSeq PMove PTurnLeft))
-\end{verbatim}
-\end{small}
-
-\studname{} decides to contact \docname{} again to show him what 
-he has done.
-
-\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-Dear Prof. \docname{}
-\newline \newline
-\noindent I have attached a file which demonstrates that the monadic
-constructs can be factored out and compiled separately, just as you
-suggested.
-\newline \newline
-
-\noindent Thanks \newline
-\noindent \studname{}
-\end{minipage} 
-}
-
-\vspace{5mm}
-
-\noindent\colorbox{light-gray}{
-\begin{minipage}[]{0.9\linewidth}
-\noindent 
-\studname{},
-\newline
-
-Fascinating! You've managed to factor out the monadic operations so
-that they can be compiled once and for all. Users of your library
-don't have to be concerned at all with the semantics of bind and return.
-Very nice!
-\newline
-%% \end{minipage}}
-
-%% \noindent\colorbox{light-gray}{
-%% \begin{minipage}[]{0.9\linewidth}
-\noindent 
-I note that the data type for monadic operations cannot be composed as
-freely as the other operations, but instead has to be applied
-separately as a final step. Your
-solution is very similar to how variable binding is handled in the
-Syntactic library \cite{syntactic}.
-\newline \newline
-
-\noindent Prof. \docname{}
-\newline
-\noindent Dept. of Computer Science and Engineering \newline
-\noindent Chalmers University of Technology 
-\end{minipage} 
-}
-
-%One of the advantages of the monad reification technique we have
-%presented is that it is compositional. The monadic constructs can be
-%dealt with separately from all the primitives of the embedded
-%languages.
-
-%A consequence of the compositionality is that we can use techniques
-%such as CompData \cite{} to decompose the definition of DSLs into
-%smaller reusable building blocks. Figure \ref{fig:compdata} shows a
-%compositional implementation of the monadic constructs, separately
-%from any other constructs. The type \verb!Cxt! is defined in the
-%package CompData and ties the recursive knot.
-
-%\begin{figure}
-%\begin{verbatim}
-%data MonadT r a where
-%  Return :: a -> MonadT r a
-%  Bind   :: r a -> (a -> r b) -> MonadT r b
-
-%instance (MonadT :<: f) => Monad (Cxt h f a) where
-%  return a = iReturn a
-%  m >>= f  = iBind m f
-%\end{verbatim}
-%\label{fig:compdata}
-%\caption{Compositional definition monads}
-%\end{figure}
+\includegraphics[width=.49\linewidth]{./wall1}
+\includegraphics[width=.49\linewidth]{./wall2}
 
 %\end{minipage}
-
-% \docname{} calls upon \studname{} to tally up the limitations of his approach
-
-% Monad Laws (Do the hold ?) 
-% Infinite programs 
-% Only certain ``kinds'' of reification!
-
-\section{Epilogue}
-
-%This short story has outlined the discovery process and the many turns from 
-%hopefulness to despair and back to hopefulness of two people who met over 
-%a language embedded in Haskell. 
-%% This was the story about two people meeting over the discovery of a way to 
-%% reify monads. Along the road from idea to implementation there was some ups 
-%% and downs but in the end they did arrive at a compositional 
-%% and reifiable way of constructing monadic EDSLs. The method, albeit not as 
-%% general as briefly hoped for, is an example of a simple and elegant approach 
-%% to the problem.
-
-This paper shows a simple method of implementing monadic EDSLs. It shows a
-na\"ive approach to the monad reification problem which has the additional benefit of being compositional. In section 
-\ref{sec:composing}, the language is reimplemented in an extensible way and 
-the compile function is shown explicitly to be compositional by treating sub 
-parts of the language separately.
-
-The example language used throughout the story is small and quite
-limited. But the technique presented in this paper does scale up to
-larger languages and more complicated language constructs.  It is
-currently used in the implementation of Obsidian, an EDSL for general
-purpose GPU programming \cite{Obsidian}. Obsidian has language
-constructs which are considerably more advanced than the robot
-language is this paper. One example is the sequential for-loop which
-is higher-order (the type parameter {\tt t} is not relevant to the
-current discussion):
-\begin{small}
-\begin{verbatim}
-SeqFor :: EWord32 -> (EWord32 -> Program t ())
-       -> Program t ()
-\end{verbatim}
-\end{small}
-There does, however, seem to be restrictions on what kind of language
-constructs the reification technique presented in this paper can deal
-with. For instance, the functions from the {\tt MonadPlus} class has
-resisted our attempts at adding them to an EDSL in the same way as
-we've added {\tt Return} and {\tt Bind}. The exact limits of the
-expressiveness of our method is currently unknown to us.
 
@@
 %\includegraphics[width=.49\linewidth]{./spiral2}
-%\caption{Images generated by our robot simulator using the {\tt spiral} program.An important detail is that the robot simulator does not evaluate the Monadic program but rather executes a compiled first-order representation of that program.}
+%\caption{Images generated by our robot simulator using the {\tt spiral} program.An important detail is that the robot simulator does not evaluate the Monadic program but rather executes a compiled first order representation of that program.}
 %\label{fig:spiral}

### 2013-09-30 src2/expressive/abstract.tex
another approach

%\begin{abstract}
%This is the text of the abstract.

\section*{Abstract}
Graphics Processing Units (GPUs) are powerful computing devices 
that with the advent of CUDA/OpenCL are becomming useful for general 
purpose computations. 
Obsidian is an embedded domain specific language that generates CUDA kernels
from functional descriptions.
A symbolic array construction allows us to guarantee that intermediate
arrays are fused away. However, the current array construction has
some drawbacks; in particular, arrays cannot be combined efficiently.
We add a new type of {\em push arrays} to the existing Obsidian system in 
order to solve this problem. The two array types complement each other,
and enable the definition of combinators that both
take apart and combine arrays, and that result in efficient generated code.
This extension to Obsidian is demonstrated on a sequence of
sorting kernels, with good results.
The case study also illustrates the use of combinators for expressing
the structure of parallel algorithms.
The work presented is preliminary, and 
the combinators presented must be generalised.
However, the raw speed of the generated kernels bodes well.
%\end{abstract}

### 2013-09-30 src2/expressive/applications.tex
another approach

\section{Application}
\label{sec:MARY}

\subsection{Sorting on a GPU}
In this section, we introduce combinators that express patterns
of computation on whole arrays, and show their application to the
development of fast sorting kernels.
By kernel, we mean a computation that is performed by multiple
threads, each performing the same computation, in a single block.
The computation is performed entirely on the GPU, operating on
a short array, which has been placed into shared memory.
On the GPU on which we perform measurements, the maximum number
of threads in a single block is 512. Thus, we will build and benchmark
a sequence of
kernels that sort 512 inputs.
Our first kernels are implemented using one thread per array element.
Next, we show how Push arrays allow us to move to having each
thread operate on two array elements, giving a substantial performance
improvement.

Sorting kernels are typically used as building blocks in larger
programs to sort much larger sequences of inputs. In section~\ref{sec:benchmarks},
we show how to build a sorter for large arrays from small building blocks, including
small kernels for sorting and merging that are generated from Obsidian.
Our small kernels are constructed in the form of sorting and merging {\em networks},
building on Batcher's bitonic merger~\cite{Batcher} and
on the periodic balanced merger~\cite{PeriodicBalanced}.
We chose also to implement the large sorter used  in benchmarking the
small kernels as a sorting network. However, large
sorters that are not themselves sorting networks (with typical examples being radix sort and quicksort) often call small sorting networks when they need to sort small arrays during their execution.
Thus, small, fast sorting kernels have a variety of uses.

\subsection{Describing Batcher's bitonic merger}

The bitonic merger is typically presented as a recursive
construction and we have earlier explored ways to describe
and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LavaSorter}.
Here, we consider iterative descriptions using similar combinators.

Figure~\ref{fig:bitonicMerger} illustrates the merger for 16 inputs.
Data flows from left to right.
The vertical lines indicate components that operate on two array elements, placing the minimum onto the lower (abstract) wire, and the maximum onto the other output
of the component.
The leftmost {\em stage} operates (for $n=16$) on elements that are 
$8$ apart. the next stage deals with elements that are $4$ apart, and so on.

\begin{figure}
\centering
\includegraphics[scale=0.25]{./expressive/bitonic}
\caption{A diagram of a 16 input bitonic merging network, using
a style that is standard in the literature. Note
that in each stage containing 8 min/max or comparator components, all
8 operate on independent parts of the input and so can proceed in parallel.
}
\label{fig:bitonicMerger}
\end{figure}

We introduce a combinator {\tt ilv1}, for {\em interleave}, that captures this pattern.
{\tt ilv1 i f g} applies {\tt f} to elements {\small $2^i$} apart,
producing the output at the lower of the two input indices; it applies {\tt g}
to the same pairs of elements, producing the output on the upper index.
Defining {\tt stage i} to be {\tt ilv1 i min max},
the four stages in the diagram are simply
{\tt stage} applied to{\small  $3$, $2$ $1$} and {\small $0$}.
The definition of {\tt ilv1} makes use of the fact that
flipping the bit $i$ of an index (using
the function {\tt flipBit}) gives the index of the element
that will be combined with it using the functions{\tt f} and {\tt g}.
The decision about whether to apply {\tt f} or {\tt g} is made
by looking at the value of bit $i$.
As we shall see, the use of the Obsidian {\tt ifThenElse}
produces conditionals in the resulting CUDA.
\begin{codesize}
\begin{verbatim}
lowBit :: Int -> UWordE -> Exp Bool
lowBit i ix = (ix .&. bit i) ==* 0

flipBit :: Bits a => Int -> a -> a
flipBit = flip complementBit

ilv1 :: Choice a => 
        Int -> (b -> b-> a) -> (b -> b -> a) -> 
        Array b -> Array a
ilv1 i f g arr = Array ixf (len arr)
  where
    ixf ix = let l = arr ! ix
                 r = arr ! newix
                 newix = flipBit i ix
             in (ifThenElse (lowBit i ix) (f l r) (g l r))
\end{verbatim}
\end{codesize}
\noindent
Expressing {\tt ilv1} using bit-flipping may seem strange, but it has
the advantage that it actually applies the desired pattern of
computation repeatedly over larger input arrays.
Now, for {\small $2^n$} inputs, a Haskell list containing the $n$ calls
of this interleave combinator are built:
\begin{codesize}
\begin{verbatim}
bmerge :: Int -> [Array IntE -> Array IntE]
bmerge n = [istage (n-i) | i <- [1..n]]
  where istage i = ilv1 i min max
\end{verbatim}
\end{codesize}
\noindent
Finally, the {\tt compose} function makes each element of the list
into a kernel (using {\tt map pure}) and places a {\tt sync} between
each kernel (using {\tt composeS}).
\begin{codesize}
\begin{verbatim}
compose :: (Scalar a) => 
           [Array (Exp a) -> Array (Exp a)] 
           -> Array (Exp a) -> Kernel (Array (Exp a))
compose = composeS . map pure

runm k = putStrLn$ CUDA.genKernel "bitonicMerge" 
         (compose (bmerge k)) (namedArray "inp" (2^k))
\end{verbatim}
\end{codesize}
\noindent
Note that {\tt bmerge k} works on inputs of length {\small }$2^{k+j}$, for $j > 0$, applying the merger to sub-sequences of length {\small $2^k$}.
The CUDA code for {\tt bmerge 4} on $16$ inputs
(with some newlines inserted) is
\begin{codesize}
\begin{verbatim}
*Main> runm 4
__global__ void bitonicMerge(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] 
     = ((tid&8)==0) 
      ? min(input0[((bid*16)+tid)],input0[((bid*16)+(tid^8))]) 
      : max(input0[((bid*16)+tid)],input0[((bid*16)+(tid^8))]);
  __syncthreads();
  (( int *)(sbase + 64))[tid] 
     = ((tid&4)==0) 
      ? min((( int *)sbase)[tid],(( int *)sbase)[(tid^4)]) 
      : max((( int *)sbase)[tid],(( int *)sbase)[(tid^4)]);
  __syncthreads();
  (( int *)sbase)[tid] 
     = ((tid&2)==0) 
      ? min((( int *)(sbase+64))[tid],(( int *)(sbase+64))[(tid^2)]) 
      : max((( int *)(sbase+64))[tid],(( int *)(sbase+64))[(tid^2)]);
  __syncthreads();
  (( int *)(sbase + 64))[tid] 
     = ((tid&1)==0) 
      ? min((( int *)sbase)[tid],(( int *)sbase)[(tid^1)]) 
      : max((( int *)sbase)[tid],(( int *)sbase)[(tid^1)]);
  __syncthreads();
  result0[((bid*16)+tid)] = (( int *)(sbase+64))[tid];
\end{verbatim}
\end{codesize}

%% ** TODO Note that bmerge k not only works on inputs of length $2^{k+j}$,
%% for $j > 0$, applying the merger to sub-sequences of length $2^k$.

\subsection{Modifying the bitonic merger}

The bitonic merger for which we have just generated a kernel is known
to sort so-called bitonic sequences, which include sequences whose first
half is sorted in one direction and whose second half is sorted in the other
direction. This fact can be used to build the well-known bitonic sorting
network. However, a GPU implementation typically needs to check, for each comparator,
whether or not it should sort upwards or downwards, see for instance
the simple CUDA implementation shown in Appendix A.
We choose here to modify the merger so that it sorts two concatenated
sequences that are sorted in the {\em same} direction.
We do this by using a well-known trick, reversing half of the input to the merger. It turns out that
we can also reverse the same half of the output of the first stage of
the network, without affecting overall behaviour.
The resulting network, {\tt tmerge}, shown in Figure~\ref{fig:mixedMerger},
encourages us to develop a new combinator to describe the characteristic
V-shaped pattern that results in the first stage.
The combinator is modelled on {\tt ilv1}. The only difference
is that the ``partner'' of an index is found not by flipping bit $i$, but
by flipping bits $0$ to $i$, using function {\tt flipLSBsTo}. The implementation
of {\tt vee1} is got from that for {\tt ilv1} by replacing the call of
{\tt flipBit} by one of {\tt flipLSBsTo} (and we could also have chosen
to make a more generic function that is parameterised on this {\em partner} function).
\begin{codesize}
\begin{verbatim}
flipLSBsTo :: Int -> UWordE -> UWordE
flipLSBsTo i = (`xor` (oneBits (i+1)))

vee1 :: Choice a => 
        Int -> (b -> b-> a) -> (b -> b -> a) -> 
        Array b -> Array a
vee1 i f g arr = Array ixf (len arr)
  where
    ixf ix = let l = arr ! ix
                 r = arr ! newix
                 newix = flipLSBsTo i ix
             in (ifThenElse (lowBit i ix) (f l r) (g l r))
\end{verbatim}
\end{codesize}

\begin{figure}
\centering
\includegraphics[scale=0.25]{./expressive/mixed}
\caption{16 input merging network, {\em tmerge}. The first stage
is made with a new combinator that we call vee, while
the remaining stages are as in the bitonic merger.
This network sorts an input that consists of two half-sized
sorted sequences, giving the opportunity to build a tree-shaped
sorting network.}
\label{fig:mixedMerger}
\end{figure}

\begin{codesize}
\begin{verbatim}
tmerge :: Int -> [Array IntE -> Array IntE]
tmerge n = vstage (n-1): [istage (n-i) | i <- [2..n]]
  where
    vstage i = vee1 i min max 
    istage i = ilv1 i min max
\end{verbatim}
\end{codesize}

Now that we have a merger that sorts sub-sequences containing two concatenated sorted sequences, it
is easy to make a tree of them.
The list of kernels to be composed now becomes
\begin{codesize}
\begin{verbatim}
tsort1 :: Int -> [Array IntE -> Array IntE]
tsort1 n = concat [tmerge i | i <- [1..n]]
\end{verbatim}
\end{codesize}
\noindent
The resulting sorting network is shown in Figure~\ref{fig:mixedsorter}.

\begin{figure}
\centering
\includegraphics[scale=0.33]{./expressive/mmixedsorter}
\caption{A 16-input sorter made from a tree of {\em tmerge} mergers. 8 2-input mergers
feed 4 4-input mergers, followed  by 2 8-input and one 16-input merger.
A sorter on 8 inputs is shaded, as is a merger on 8 inputs, above it. 
}
\label{fig:mixedsorter}
\end{figure}
\noindent
The following call writes the resulting CUDA to file {\tt tsort1.cu}
and this is the first {\em generated} CUDA kernel whose performance is measured in section~\ref{sec:benchmarks}.

\begin{codesize}
\begin{verbatim}
runs1 k
  = writeFile "tsort1.cu" $ CUDA.genKernel "tsort1" 
      (compose (tsort1 k)) (namedArray "inp" (2^k))
\end{verbatim}
\end{codesize}

The generated code
% for $8$ inputs is shown in Appendix A. It 
uses
one thread per array element (just as the {\tt bitonicMerge}
kernel shown above did). The next step is to move to using Push as
well as Pull arays, so as to be able to generate more efficient code
from essentially the same sorter construction.

\subsection{New combinator implementations using Push arrays}

%% ** TODO make code for ilv2 more comprehensible :)

It is in combining the results of the {\tt f} and {\tt g} functions
that we run into difficulty using just Pull arrays.
Earlier, we saw how to use Push arrays to implement {\tt concP},
which concatenates two arrays.
Here, we use exactly the same approach to make a new version of
the {\em interleave} combinator. The results of applying
the {\tt f}s and {\tt gs} are combined into a Push array, in the right
order.
\pagebreak
%% ** TODO More explanation needed.
\begin{codesize}
\begin{verbatim}
ixMap :: (UWordE -> UWordE) -> ArrayP a -> ArrayP a 
ixMap f (ArrayP p n) = ArrayP (ixMap' f p) n

ixMap' :: (UWordE -> UWordE) 
         -> P (UWordE, a)
         -> P (UWordE, a) 
ixMap' f p = \g -> p (\(i,a) -> g (f i,a))

insertZero :: Int -> UWordE -> UWordE
insertZero 0 a = a `shiftL` 1
insertZero i a 
  = a + (a .&. fromIntegral (complement (oneBits i :: Word32)))

ilv2 :: Choice b => 
        Int -> (a -> a -> b) -> (a -> a -> b) -> 
        Array a -> ArrayP b
ilv2 i f g (Array ixf n) 
   = ArrayP (\k -> app a5 k *>* app a6 k) n
  where
    n2 = n `div` 2
    a1 = Array (ixf . left) (n-n2)
    a2 = Array (ixf . right) n2
    a3 = zipWith f a1 a2
    a4 = zipWith g a1 a2
    a5 = ixMap left (push a3)
    a6 = ixMap right (push a4)
    left = insertZero i
    right = flipBit i  . left
    app (ArrayP f _) a = f a
\end{verbatim}
\end{codesize}
\noindent
This new combinator can now replace {\tt ilv1} in the bitonic
merger, giving a kernel that runs considerably faster. We will use
that kernel to build a large sorter later.

The implemenation of {\tt vee2} is almost identical to
that of {\tt ilv2}, with {\tt flipBit i}
replaced by {\tt flipLSBsTo} as before (so that, again, one would in
fact make a more generic function for building such combinators).
Now, we just need to replace the {\tt ilv1} and {\tt vee1} combinators
in the tree sorter with {\tt ilv2} and {\tt vee2} , to get a verrsion that uses half as many threads:
\begin{codesize}
\begin{verbatim}
tmerge2 :: Int -> [Array IntE -> ArrayP IntE]
tmerge2 n = vstage (n-1) : [ istage (n-i) | i <- [2..n]]
  where
    vstage i = vee2 i min max
    istage i = ilv2 i min max

tsort2 :: Int -> [Array IntE -> ArrayP IntE]
tsort2 n = concat [tmerge2 i | i <- [1..n]]
\end{verbatim}
\end{codesize}
As we shall see in section~\ref{sec:benchmarks}, the resulting code
is significantly faster. This is because it uses one thread per two array elements,
and the code no longer contains any conditionals.

In order to go faster still, we resort to building a different sorting network, of
exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\cite{PeriodicBalanced}.
This involves the introduction of one new combinator that can be seen as a mixture
of the {\tt ilv} and {\tt vee} combinators already introduced.

\subsection{A sorter built from the balanced periodic merger}
First, we note that the balanced periodic merger contains multiple
uses of the now familiar vee-shaped pattern, see Figure~\ref{fig:periodicMerger}.
\begin{figure}
\centering
\includegraphics[scale=0.25]{./expressive/balanced}
\caption{16 input periodic balanced merging network}
\label{fig:periodicMerger}
\end{figure}

\begin{codesize}
\begin{verbatim}
bpmerge2 :: Int -> [Array IntE -> ArrayP IntE]
bpmerge2 n = [vstage (n-i) | i <- [1..n]]
  where vstage i = vee2 i min max
\end{verbatim}
\end{codesize}

Now Dowd et al proved that the balanced periodic merger sorts two {\em interleaved} sorted sequences. So, taking an iterative view of the resulting sorter,
we want to build a tree of mergers as before, but the smaller mergers should
be interleaved, rather than operating on adjacent sub-sequences.
There should be one merger on the right hand end of the network; left of
that, there should be two mergers that operate on the odd and even elements,
and each of them should in turn be fed by two interleaved mergers, and so.
The sorter is illustrated, for $16$ inputs in Figure~\ref{fig:vsorter}.
Just to the left of the final balanced merger, one of the two interleaved
mergers is shown using dotted lines. It operates on a completely different set of inputs from the other 8-input merger in the same part of the tree.

\begin{figure*}
\centering
\includegraphics[scale=0.5]{./expressive/vsorter}
\caption{A sorter based on the idea that the periodic
balanced merger network sorts two interleaved sorted sequences.
It consists of two half-sized sorters, one working on the
odd elements of the input and one on the even, followed by
the balanced merger. The diagram indicates using dotted lines the
balanced merger that is the final (rightmost) part of
one of the half-size sorters.}
\label{fig:vsorter}
\end{figure*}

The most straightforward way to give an iterative description of
this sorter is
to introduce a new combinator that is a combination of
{\tt ilv} and {\tt vee}.
The only thing that we need to change is the {\em partner} function.
This time we will flip not the least significant bits from position $0$ to
position $i$ but from position $i$ to $i+j$.

\begin{codesize}
\begin{verbatim}
-- flip bits from position i to position i+j inclusive
flipBitsFrom :: Bits a => Int -> Int -> a -> a
flipBitsFrom i j a = a `xor` (fromIntegral mask)
  where
    mask = (oneBits (j + 1):: Word32) `shiftL` i

ilvVee1 :: Choice a => 
           Int -> Int -> 
           (b -> b-> a) -> (b -> b -> a) -> 
           Array b -> Array a
ilvVee1 i j f g arr = Array ixf (len arr)
  where
    ixf ix = let l = arr ! ix
                 r = arr ! newix
                 newix = flipBitsFrom i j ix
             in (ifThenElse (lowBit (i+j) ix) (f l r) (g l r))
\end{verbatim}
\end{codesize}

\begin{codesize}
\begin{verbatim}
ilvVee2 :: Choice b => Int -> Int -> 
           (a -> a -> b) -> (a -> a -> b) -> 
           Array a -> ArrayP b
ilvVee2 i j f g (Array ixf n) 
  = ArrayP (\k -> app a5 k *>* app a6 k) n
  where
    n2 = n `div` 2
    a1 = Array (ixf . left) (n-n2)
    a2 = Array (ixf . right) n2
    a3 = zipWith f a1 a2
    a4 = zipWith g a1 a2
    a5 = ixMap left (push a3)
    a6 = ixMap right (push a4)
    left = insertZero (i+j)
    right = flipBitsFrom i j . left
    app (ArrayP f _) a = f a
\end{verbatim}
\end{codesize}

For both variants of the combinator, we simply add to the {\tt ilv} definitions
a new {\tt Int} parameter, {\tt j}, and replace {\tt flipBit i} by {\tt flipBitsFrom i j}. We also insert the zero bit (when calculating the left index)
at position {\tt i + j} rather than just at position {\tt i}.
{\tt ilvVee} is a generalisation of both {\tt ilv} and {\tt vee}.
{\tt ilvVee i 0} has the same behaviour as {\tt ilv i}, and {\tt ilvVee 0 (j-1)}
is the same as {\tt vee j}. The {\tt i} parameter controls the degree
of interleaving, and the {\tt j} parameter controls the size of
the vee-shaped blocks.

For $16$ inputs, the parameters
to {\tt ilvVee} that describe the periodic merger on the right of the construction
are $i=0$ (for no interleaving) paired with $3$, $2$, $1$ and $0$
for the decreasing size of the vee-shaped blocks.
Next, to the left, the mergers are interleaved ($i=1$) and there are three
stages with
vee-shaped blocks of decreasing size ($j = 2, 1, 0$), see Figure~\ref{fig:vsorter}.
The following code gives an iterative description of the construction
for {\small $2^n$} inputs:

\begin{codesize}
\begin{verbatim}
vsort :: Int -> Array IntE -> Kernel (Array IntE)
vsort n = composeS . map pure $ [istage (n-i) (i-j) 
                                | i <- [1..n], j <- [1..i]]
  where istage i j = ilvVee2 i j min max
\end{verbatim}
\end{codesize}

\noindent
The resulting generated code uses one thread per
$2$ indices.

\subsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
\begin{figure}
\begin{codesize}
\begin{verbatim}
  unsigned int arrayLength = 1 << LOG_L_SIZE;
  unsigned int diff = LOG_L_SIZE - LOG_S_SIZE;
  unsigned int blocks = arrayLength / S_SIZE;
  unsigned int threads = S_SIZE / 2;

  sortSmall<<<blocks, threads,4096>>>(din,din);

  for(int i = 0 ; i < diff ; i += 1){ 
    vSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<i)*S_SIZE);

    for(int j = i-1; j >= 0; j -= 1)
      iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);

    bmergeSmall<<<blocks,threads,4096>>>(din,din);}
\end{verbatim}
\end{codesize}
\caption{CUDA code for our large sorter. {\tt sortSmall} and {\tt bmergeSmall}
are replaced by Obsidian-generated kernels in the experiments. {\tt vSwap} and {\tt iSwap} are handwritten CUDA kernels
that perform one column of compare and swap operations in the vee and ilv shapes respectively, and are
parameterised on the stride. (Our generated kernels have fixed input and output size.)}
\label{fig:CUDAsort}
\end{figure}
We have measured the performance of generated 512-input sorting and merging kernels by plugging them into a larger sorter written in CUDA.
The sorter has exactly the structure shown in Figure~\ref{fig:mixedsorter}.
Figure~\ref{fig:newmixedsorter} shows the location of smaller sorters and mergers, and of vSwap and iSwap kernels for $16$ inputs.
Larger sorters simply have more columns of mergers, each preceded by a vSwap and a number of iSwaps.
The overall structure of the resulting CUDA code is shown in Figure~\ref{fig:CUDAsort}.
Because {\tt iSwap} is used repeatedly, we also wrote kernels
corresponding to compositions of two and three of them (avoiding
memory accesses between the columns).
That is, we replaced the loop containing {\tt iSwap} above by
\pagebreak
\begin{codesize}
\begin{verbatim}
for(int j = i-1; j >= 0; j -= 3){ 
  if (j==0) 
    iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
    else 
    {if (j==1) 
       iSwap2<<<blocks/4,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
       else 
       iSwap3<<<blocks/8,threads*2,0>>>(din,din,(1<<j)*S_SIZE);}}
\end{verbatim}
\end{codesize}
\noindent
The Obsidian implementation and all code for examples in the paper are available at \url{http://www.cse.chalmers.se/~joels/expressive.html}.

\begin{figure}
\centering
\includegraphics[scale=0.33]{./expressive/newmixedSorter}
\caption{
This diagram shows the tree-shaped sorting network that we saw earlier, but with 4 input sorting and merging kernels indicated. This is the structure of the network that we have used to implement large sorters, in which the small kernels having {\small $2^9 = 512$} inputs in all cases. For {\small $2^{24}$} inputs overall, the resulting sorter
has one column of small sorters and $24-9=15$ of small mergers.}
%% both ofs in the above sentence should remain :)
\label{fig:newmixedsorter}
\end{figure}

Table~\ref{fig:table} shows the performance figures for the large sorter
for $5$ different $512-$input small sorter kernels.
{\tt tsort1} is defined above, and is a variant of bitonic sort
that does not require {\tt if} statements to determine the direction
of sorting of pairs, as all comparison operations place the minimum at the
same index as the lower input.
{\tt tsort2} is the same construction, but built with the combinators {\tt ilv2}
and {\tt vee2} that result in the use of one thread per two indices.
{\tt vsort}, the fastest kernel, uses a generalisation of those
two combinators, and again uses one thread per two indices.
{\tt vsort1} is the same construction as {\tt vsort}, but with {\tt ilvVee2} replaced
by {\tt ilvVee1}, and so using one thread per index.
As a reference small kernel implementation, we include a simple hand-coded bitonic sort
that uses one thread per index (see code in Appendix).
None of the kernels has been subject to optimisations related to warp size.

\input{./expressive/table1.tex}

We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:table2}.
The {\tt tsort1} and {\tt vsort1} kernels, which are built using only pull arrays, 
(and which have one index per thread in the generated code) are
noticeably slower than those that have two indices per thread.
The generation of the latter kernels is made possible by the introduction of 
push arrays.

\input{./expressive/table2.tex}

Although none of our generated kernels is optimised (for example with respect to warp size), 
their performance is, nevertheless, very good. We are working on automated warp size-related 
optimisations. It would also be interesting to explore the fusion of adjacent columns of 
comparators in the small kernels; omitting a {\tt sync} would cause this fusion to happen
but we also need to modify the threading behaviour (doubling the number of indices per thread).

%The final version of the paper will contain benchmarking against a highly optimised sorter kernel obtained from our colleagues in the computer graphics group at Chalmers. For now, we emphasise that the {\tt vsort} kernel is so fast that we expect it to compare favourably. Given the simplicity of its description in Obsidian, we feel that this is a good way to write fast kernels.

### 2013-09-30 src2/expressive/discussion.tex
another approach

\section {Discussion}

%% ** TODO. Mention CSE. warp-related opt.

Push arrays form a new approach to array representation in DSELs.
We do not know of similar approaches in the literature, despite
the fact that the notions of demand and data flow
may feel familiar to the reader who considers Pull and Push arrays.
The addition of Push arrays to Obsidian seems highly beneficial. With 
this new feature, the user gains finer grained control over the code generated
and the resulting CUDA kernels perform considerably better than before.
This 
was illustrated in the series of sorters explored in section~\ref{sec:MARY}.
The performance of {\tt vsort} is sufficiently good that it
can be used as a first phase in a larger sorter (written in CUDA) that
can sort 16M elements in 96 ms, while an i7-920 CPU takes
around 2740 ms. Further speed improvements look possible, both in the coordination code
and in the kernels. An obvious next step would be to investigate the generation of the {\tt iSwap} and
{\tt vSwap} kernels from Obsidian. (This is not currently possible because of assumptions that we made about the interfaces to kernels and about how {\em thread ids} are used. We will look into ways to relax
our assumptions.)

The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
description language~\cite{LavaSorter} was that a relatively small set of combinators went a long way. So, although we introduced three
combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
We expect to provide the user with a well-documented set of combinators,
so that users can get access to this style of programming without having
to develop their own combinators, and without having to think too much
about bit-hacking.
The bit-manipulation approach chosen to define our combinators automatically
created functions that apply to sub-sequences of the input that
are of an appropriate length.

In this paper, we made combinators for the special case of two input, two
output operations (built from two two-input funtions that we typically called
{\tt f} and {\tt g}). 
This approach should be generalised to deal with blocks that have $2^k$ inputs
and outputs.
Also, we made a compound combinator from {\tt ilv} and {\tt vee},
but generalising to more than two input components would allow
for composing combinators, and indeed for recursive descriptions
that could be unrolled. Then, ignoring {\tt syncs}, a recursive description of
{\tt vsort} could be something like 
\begin{codesize}
\begin{verbatim}
vsortR 0 = id
vsortR n = bpmergeR n . ilv2 1 (vsortR (n-1))
\end{verbatim}
\end{codesize}
\noindent
It would then be necessary to optimise the code generated from
multiple applications of {\tt ilv2 1}, for example, whereas here
we have forced the user to figure out both the unrolling and the combinations.
Moving to more general combinators would also give the opportunity to
provide predefined combinators that capture more of the commonly used threading patterns (for instance $k$ indices per thread rather than the $1$ and $2$ shown here).

The integration of Push arrays into Obsidian raises some new questions.
Previously, there was a direct correspondence between the length 
of an array and the number of threads used to compute it, which
allowed the user to write an initial program without worrying about
threads at all, and then to tweak the Obsidian program if he was not satisfied
with the threading behaviour of the resulting kernel. 
Now, as can be seen in the {\tt catArrayPs} example and in the sorters, this correspondence
can be broken using Push arrays. The {\tt catArrayPs} example and two
of the sorters use half as many threads as the 
number of elements. For users who are very concerned about the speed of
the generated kernels, getting this control through using Push arrays in
a particular pattern is clearly a good thing. But adding a second, different
way to control thread use in the generated code certainly complicates matters, and further case studies are needed to confirm that the complication pays off.

The addition of Push arrays also adds the possibility to include potentially 
unsafe operations in Obsidian, for example by writing multiple array elements to the same index, or by discarding elements.
This new expressiveness will have to be carefully controlled. On the positive side, it offers the possibility to encode functions like {\tt filter} from Haskell
that
are simply not expressible using only Pull arrays. Being able to implement {\tt filter} would make programming kernels in obsidian feel much more like programming in Haskell -- a welcome loosening of the strait-jacket.
Once that is done, it will be time to develop a very simple coordination language to allow programming of entire GPU applications that make use of
the kind of small kernel building blocks developed here.

### 2013-09-30 src2/expressive/intro.tex
another approach

\section{Introduction}

%\input{table}

%----------------------------------------------------------------------------

Graphics Processing Units (GPUs) are parallel computers with hundreds 
to thousands of processing elements. The CUDA and OpenCL languages  
make available the power of the GPU to programmers interested in general 
purpose computations. In CUDA and OpenCL, the programmer writes {\em kernels}, 
Single Program Multiple Data (SPMD) programs that are executed by groups 
of threads on the available processing elements of the GPU.

CUDA and OpenCL are general purpose programming languages, mirroring the 
increased capabilities of a modern GPU to target that domain. However, these 
languages lack compositionality. 
Also, being based in C/C++ means that the core idea in a program may
not be easily visible.

\subsection{Embedded DSLs for GPGPU programming} 
We are aiming for a GPU programming language that is more
concise than mainstream
languages such as CUDA and OpenCL. 
Obsidian is a
domain specific embedded language (DSEL) implemented in Haskell.
When an Obsidian program is run, a representation of the program is created 
as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
The program representation generated when running an Obsidian program 
is compiled into CUDA code. We are also working on an OpenCL backend.

Our approach is different from that of other Haskell DSELs targetting
GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
We do not try to abstract away from 
all the peculiarities of GPU programming, but rather provide a higher 
level language in which to experiment with them.
For instance,
Accelerate provides a standard set of basic operations such as 
{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
Obsidian, on the other hand, allows the user to experiment with the
{\em generation} of small kernels for fixed size array inputs
from higher level descriptions.
It is intended to allow the user to play with the kinds of tradeoffs that are important
when writing such high performance building blocks; in this paper, the main consideration
is the number of array elements of the input and output that are manipulated
by a single thread in the generated CUDA code.
An important aspect of Obsidian is the symbolic array representation used, along
with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
allows the programmer to guide code generation and control parallelism and thread
use~\cite{JSLIC}.

%Today there are number of EDSLs written in Haskell targeting 
%GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
%another by having a slightly different approach. We do not try to abstract away from 
%all the peculiarities of GPU programming but rather giving a higher 
%level language in which to experiment with them.
%trying to appeal to a slightly different target audience. 
%We try to cater not for the advanced functional programmer who wants to 
%run his programs on GPUs, but rather to offer a higher level 
%language to the CUDA/OpenCL hacker. 
%In this language we want the user to be 
%able to quickly experiment with the kind of tradeoffs that are important 
%when writing a high performance GPU kernel.

%Unlike the other approaches that are complementary to our Haskell GPU EDSL, 
%Obsidian does not provide the usual set of basic operations such as 
%{\tt Map}, {\tt Reduce} and {\tt ZipWith} as built in skeletons. We 
%want a language where the programmer can experiment with implementing 
%his or her own versions of these basic operations. So far, we make
%this possible in Obsidian through a {\tt Sync} operation and the Array 
%representation we use. The {\tt Sync} operation, using target domain 
%nomenclature, is the programmers tool to guide code generation and 
%obtain parallelism. The presentation of Obsidian here will be somewhat 
%cursory, for a more complete view of the various versions and approaches 
%we have tried, see~\cite{JSLIC}.
%More details on this later.

In Obsidian, a kernel that sums an array can be expressed as: 
\begin{codesize}
\begin{verbatim}
sum :: Array IntE -> Kernel (Array IntE) 
sum arr | len arr == 1 = return arr
        | otherwise    = (pure (fmap (uncurry (+)) . pair) 
                          ->- sync 
                          ->- sum) arr                       
\end{verbatim}
\end{codesize}
The result of running this kernel on an eight element input array, 
{\tt runKernel sum (namedArray ``input'' 8)}, is an intermediate representation 
of the computation (shown in slightly pretty-printed form): 
\begin{codesize}
\begin{verbatim}
arr0 = malloc(16)
par i 4 {
arr0[i] = ( +  input[( *  i 2 )] input[( +  ( *  i 2 ) 1 )] );
}Sync
arr1 = malloc(8)
par i 2 {
arr1[i] = ( +  arr0[( *  i 2 )] arr0[( +  ( *  i 2 ) 1 )] );
}Sync
arr2 = malloc(4)
par i 1 {
arr2[i] = ( +  arr1[( *  i 2 )] arr1[( +  ( *  i 2 ) 1 )] );
}Sync
\end{verbatim}
\end{codesize}
The named intermediate arrays in this representation are then laid 
out in GPU shared memory and CUDA code can be generated (here for arrays of length eight)\footnote{An alignment qualifier for shared memory 
has been omitted to save space in the listings showing generated code}: 
%\pagebreak
\begin{codesize}
\begin{verbatim}
__global__ void sum(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] = 
    (input0[((bid*8)+(tid*2))]+
     input0[((bid*8)+((tid*2)+1))]);
  __syncthreads();
  if (tid<2){
    (( int *)(sbase + 16))[tid] = 
      ((( int *)sbase)[(tid*2)]+
       (( int *)sbase)[((tid*2)+1)]);
  }
  __syncthreads();
  if (tid<1){
    (( int *)sbase)[tid] = 
      ((( int *)(sbase+16))[(tid*2)]+
       (( int *)(sbase+16))[((tid*2)+1)]);
  }
  __syncthreads();
  if (tid<1){
    result0[(bid+tid)] = (( int *)sbase)[tid];
  }
}
\end{verbatim}
\end{codesize}
%The CUDA code above is generated by issuing the command:
%\begin{codesize}
%\begin{verbatim}
%genKernel ``sum'' sum (namedArray undefined 8 :: Array IntE)
%\end{verbatim}
%\end{codesize}
%\noindent
%and the kernel code is generated for arrays of length eight.

%----------------------------------------------------------------------------
\subsection{Arrays in Obsidian}

An array is represented by an indexing function and a length:
\begin{codesize} 
\begin{verbatim}
data Array a = Array (UWordE -> a) Word32 
\end{verbatim}
\end{codesize}
This array representation has served us well. It has these properties:
\begin{itemize} 
\item Fusion of operations is automatic.
\item It naturally describes a data-parallel computation suitable for CUDA/OpenCL generation.
\item Many basic operations can be implemented: {\tt map}, {\tt zipWith} etc.
\end{itemize}
Using this array representation in a DSEL is not new; the first occurence that we know of is in
Pan~\cite{PAN}. Similar array representations have also later been 
used in Feldspar~\cite{FELDSPAR}, and more recently also in the Repa library~\cite{REPA}.
Functions for indexing and getting the length of arrays are as follows:
\begin{codesize} 
\begin{verbatim}
(!) :: Array a -> UWordE -> a 
(Array ixf _) ! ix = ixf ix

len :: Array a -> Word32
len (Array _ n) = n 
\end{verbatim}
\end{codesize}
%Implementing a {\em Functor} instance for the {\tt Array} datatype is possible:
A {\em Functor} instance for the {\tt Array} datatype is
\begin{codesize} 
\begin{verbatim}
instance Functor Array where 
  fmap f arr = Array (\ix -> f (arr ! ix)) (len arr) 
\end{verbatim}
\end{codesize}
Now, composed applications of {\tt fmap} will be automatically fused. This is 
illustrated in the example program below and the CUDA generated
from it.  
\begin{codesize} 
\begin{verbatim}
mapFusion :: Array IntE -> Kernel (Array IntE) 
mapFusion = pure (fmap (+1) . fmap (*2)) 
\end{verbatim}
\end{codesize}

\begin{codesize} 
\begin{verbatim}
__global__ void mapFusion(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = ((input0[((bid*32)+tid)]*2)+1);
}
\end{verbatim}
\end{codesize}
Both of these code listings need explanation. In the Haskell 
code, {\tt mapFusion} has type {\tt Array IntE -> Kernel (Array IntE)};
{\tt Kernel} is a state monad that accumulates CUDA code as well as provides 
new names for intermediate arrays. Neither of these features of the monad 
is activated by this example though. The {\tt pure} function is defined 
using the monad's {\tt return} as {\tt pure f a = return (f a)}. In this case, 
it lifts a function of type {\tt Array IntE -> Array IntE} into 
a kernel.

The generated CUDA code computes the result array using a number of threads 
equal to the length of that array. In this case, the kernel was generated
to deal with arrays of length $32$. The important detail to notice in the 
CUDA code is that there is no intermediate array created between the 
{\tt (*2)} and the {\tt (+1)} operations.

The {\tt mapFusion} example could just as well have been implemented using 
the kernel sequential composition combinator, {\tt ->-}. 
\begin{codesize} 
\begin{verbatim}
mapFusion :: Array IntE -> Kernel (Array IntE) 
mapFusion = pure (fmap (*2)) ->- pure (fmap (+1)) 
\end{verbatim}
\end{codesize}
Exactly the same CUDA code is then generated.

In some cases, it is necessary to force computation of 
intermediate arrays. This can be used to share partial computations between 
threads and to expose parallelism. In Obsidian, the tool for this
is called {\tt sync}, a built-in kernel. Using {\tt sync} as follows prevents fusion of the
two operations:
\begin{codesize} 
\begin{verbatim}
mapUnFused :: Array IntE -> Kernel (Array IntE) 
mapUnFused = pure (fmap (*2)) ->- sync ->- pure (fmap (+1))
\end{verbatim}
\end{codesize}
The generated CUDA code now stores an intermediate result in local shared
memory before moving on. 
\begin{codesize} 
\begin{verbatim}
__global__ void mapUnFused(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] = (input0[((bid*32)+tid)]*2);
  __syncthreads();
  result0[((bid*32)+tid)] = ((( int *)sbase)[tid]+1);
}
\end{verbatim}
\end{codesize}
Intermediate arrays are laid out in the {\tt sbase} array in shared memory. 
Since we may store arrays of many different types in the same locations 
of the shared memory at different times during the execution, the type casts
used in the code above are necessary.

%----------------------------------------------------------------------------
\subsection{Sync and parallelism}

The {\tt sync} operation also enables the writing of parallel reduction 
kernels. A reduction operation is an operation that takes an array as input
and
produces a singleton array as output.

First, we define {\tt zipWith} and {\tt halve} on Obsidian arrays.
\begin{codesize} 
\begin{verbatim}
zipWith :: (a -> b -> c) -> Array a -> Array b -> Array c
zipWith op a1 a2 = Array (\ix -> (a1 ! ix) `op` (a2 ! ix)) 
                   (min (len a1) (len a2))

splitAt :: Word32 -> Array a -> (Array a, Array a) 
splitAt n arr = 
  (Array (\ix -> arr ! ix) n , 
   Array (\ix -> arr ! (ix + fromIntegral n)) (len arr - n))

halve arr = splitAt ((len arr) `div` 2) arr
\end{verbatim}
\end{codesize}
A reduction kernel that takes an array whose length is a power of two 
and gives an array of length one can be defined recursively. 
Defining kernels recursively results in completely unrolled CUDA kernels,
and kernel input size must be known at compile time.
The approach to reduction taken here is to split the input array 
into two halves and then apply {\tt zipWith} of the combining function to the 
two halves, repeating the process until the length is one. 
\begin{codesize} 
\begin{verbatim}
reduceS :: (a -> a -> a) -> Array a -> Kernel (Array a) 
reduceS op arr | len arr == 1 = return arr
               | otherwise    = 
                 (pure ((uncurry (zipWith op)) . halve)
                  ->- reduceS op) arr
\end{verbatim}
\end{codesize}
%This Obsidian Kernel takes an array of length $2n$ as input and the output 
Since the output of this kernel is of length one, and the number 
of elements in the output array specifies the number of threads used to compute it, 
this function, {\tt reduceS}, defines a sequential reduction. 
The generated code for arrays of length eight is 
\pagebreak
%The code is generated with the 
%command {\tt genKernel ``reduceSAdd'' (reduce (+)) (namedArray undefined 8 :: Array IntE)}, 
%meaning that a CUDA reduce-with-addition kernel for arrays of length eight is generated. 
\begin{codesize} 
\begin{verbatim}
__global__ void reduceSAdd(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[(bid+tid)] = 
    (((input0[((bid*8)+tid)]+
       input0[((bid*8)+(tid+4))])+
      (input0[((bid*8)+(tid+2))]+
       input0[((bid*8)+((tid+2)+4))]))+
     ((input0[((bid*8)+(tid+1))]+
       input0[((bid*8)+((tid+1)+4))])+
      (input0[((bid*8)+((tid+1)+2))]+
       input0[((bid*8)+(((tid+1)+2)+4))])));
}
\end{verbatim}
\end{codesize}
Sequential reduction is not very interesting for GPU execution, but 
the fix is simple. A well placed use of {\tt sync} indicates that 
we want to compute, after each {\tt zipWith} phase, the intermediate 
arrays using as many threads as that intermediate array is long. 
The effect is shown in the code below. 
\begin{codesize} 
\begin{verbatim}
reduce :: Syncable Array a 
          => (a -> a -> a) -> Array a -> Kernel (Array a)
reduce op arr | len arr == 1 = return arr
              | otherwise    = 
                (pure ((uncurry (zipWith op)) . halve)
                 ->- sync 
                 ->- reduce op) arr
\end{verbatim}
\end{codesize}
The CUDA code for reduction with addition on eight
elements is
\begin{codesize} 
\begin{verbatim}
__global__ void reduceAdd(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] = 
    (input0[((bid*8)+tid)]+
     input0[((bid*8)+(tid+4))]);
  __syncthreads();
  if (tid<2){
    (( int *)(sbase + 16))[tid] = 
      ((( int *)sbase)[tid]+
       (( int *)sbase)[(tid+2)]);   
  }
  __syncthreads();
  if (tid<1){
    (( int *)sbase)[tid] = 
      ((( int *)(sbase+16))[tid]+
       (( int *)(sbase+16))[(tid+1)]);    
  }
  __syncthreads();
  if (tid<1){
    result0[(bid+tid)] = (( int *)sbase)[tid];  
  }
}
\end{verbatim}
\end{codesize}           
In this generated CUDA, three phases can be identified. 
The first uses four threads to compute a four element intermediate array;
the second uses two threads, and so on. At the very end, a single thread
copies the result from local shared memory to global memory.

\subsection{Drawbacks of Obsidian Arrays}
\label{sec:Drawbacks}

%*** THE STUFF BELOW IS IN CONFLICT WITH WHAT I ADDED ABOVE!
%When using Obsidian we have available to us a toolbox of higher order
%functions that are familiar to functional programmers. One example of such a 
%function is {\tt zipWith}.

%\begin{codesize} 
%\begin{verbatim}
%example1 :: (Array IntE, Array IntE) 
%            -> Kernel (Array IntE)
%example1 (arr1,arr2) = return$ zipWith (+) arr1 arr2
%\end{verbatim}
%\end{codesize}

%The body of this function looks very familiar to Haskell programmers. The 
%type of the function however is a little bit different than what we are 
%used to from everyday Haskell usage. In Obsidian we have a concept called 
%a Kernel. Computations of types {\tt a -> Kernel b} can be turned into 
%runnable GPU code by providing an input. 
%*** VERY VAGUE AND ODD 
%*** TODO: Explain kernel ?

%The following code listing shows the CUDA code we can generate from 
%the Obsidian program given above:

% *** TODO: Needs a code size between tiny and small.

%\begin{codesize}
%\begin{verbatim}
%__global__ void example1(int *input0,int *input1,int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;

%  result0[tid] = (input0[((bid*32)+tid)]+
%                  input1[((bid*32)+tid)]);

%}
%\end{verbatim}
%\end{codesize}

%The CUDA code above writes into a result array the pairwise sums from 
%its two input arrays, as expected. 
%*** There are many details here. tid, bid, 32. explain now or push to later?

% \subsection{Obsidian arrays}

%In Obsidian an array is represented by an indexing function and a length. 
%\begin{codesize}
%\begin{verbatim}
%data Array a = Array (IxExp -> a) Word32
%\end{verbatim}
%\end{codesize}

%And {\tt IxExp} is defined as {\tt UWordE}, that is an expression 
%representing a unsigned 32bit value. The length of an array is represented
%with a unsigned 32bit value as well but here it has been beneficial 
%to use a built in Haskell type in order to be able to use this 
%information at program generation time rather than program run time.

%This array representation specifies how to compute each index of an Arrays. 
%If we are interested in the value at location 5 in an array all we need 
%to do is apply the indexing function to the literal expression 5 and 
%we get a description of how to compute that value.

%This array representation have been working fairly well in many situations. 
%In {\tt example1} where we used {\tt zipWith} we got just about the resulting
%CUDA code that a CUDA programmer would write himself. For some other examples 
%the situation is much worse.

The previous subsection described positive aspects of the array 
representation that we have used so far. There are, however, 
circumstances in which this Array representation is too restricted.

Take the problem of concatenating two arrays. Using the array representation 
described above, the only way to concatenate two arrays is to introduce 
a conditional into the indexing function. If {\tt f} and {\tt g} are the 
indexing functions of two arrays that are to be concatenated, and {\tt n1}
is the length of the first array, the indexing 
function of the result must be
\begin{codesize}
\begin{verbatim}
new ix = if (ix < n1) 
         then f ix 
         else g (ix - n1)
\end{verbatim}
\end{codesize}
The following program
concatenates two arrays: 
\begin{codesize}
\begin{verbatim}
catArrays :: (Array IntE, Array IntE) -> Kernel (Array IntE)
catArrays = pure conc
\end{verbatim}
\end{codesize}
%(arr1,arr2) = return$ conc (arr1,arr2)
When it is used to generate a CUDA kernel that concatenates
two arrays of length $16$, the following code is the result: 
\begin{codesize}
\begin{verbatim}
__global__ void catArrays(int *input0,int *input1,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = 
    (tid<16) ? input0[((bid*16)+tid)] : 
               input1[((bid*16)+(tid-16))];
}
\end{verbatim}
\end{codesize}
%__global__ void catArrays(int *input0,
%                          int *input1,
%                          int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;
%  
%  result0[((bid*64)+tid)] = 
%    (tid<32) ? input0[((bid*32)+tid)] : 
%               input1[((bid*32)+(tid-32))];
%  
%}
Now, conditionals like these are {\em bad} in code 
to execute on a GPU, with its wide-SIMD data-parallel model. Separating
the operation into two assignments and using half as many threads gives
much higher performance.
\begin{codesize}
\begin{verbatim}
__global__ void catArraysByHand(int *input0,
                                int *input1,
                                int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = input0[((bid*16)+tid)];
  result0[((bid*32)+tid+16)] = input1[((bid*16)+tid)];  
}
\end{verbatim}
\end{codesize}
There are cases where code with conditionals is not that bad. 
An expert on NVIDIA GPUs in particular may say that code with 
the condition {\tt (tid < 32)} is fine, since 32 is the SIMD width 
of those GPUs. However, any number that is not a multiple of 32 would 
lead to poor performance, so in general this is a problem. 
 Worse still, {\em zipping} two arrays together and then 
{\em unpairing} (to get an array of elements)
leads to code that takes two different paths depending 
on odd or even {\em thread id}. When a GPU executes such code, it 
shuts down half of the threads and computes the two paths in sequence. 
\begin{codesize}
\begin{verbatim}
zippUnpair :: (Array IntE, Array IntE) -> Kernel (Array IntE) 
zippUnpair = pure (unpair . zipp)
\end{verbatim}
\end{codesize}
The {\tt zipp} and {\tt unpair} operations are defined as follows: 
\begin{codesize}
\begin{verbatim}
zipp :: (Array a, Array b) -> Array (a, b)             
zipp (arr1,arr2) = 
     Array (\ix -> (arr1 ! ix, arr2 ! ix)) 
           (min (len arr1) (len arr2))

unpair :: Choice a => Array (a,a) -> Array a
unpair arr = 
    let n = len arr
    in  Array (\ix -> ifThenElse ((mod ix 2) ==* 0) 
                      (fst (arr ! (ix `shiftR` 1)))
                      (snd (arr ! (ix `shiftR` 1)))) (2*n)
\end{verbatim}
\end{codesize}
Code generated from the {\tt zippUnpair} program exhibits really 
poor performance; at any time half of the threads are shut down. 
\begin{codesize}
\begin{verbatim}
__global__ void zippUnpair(int *input0,
                           int *input1,
                           int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*64)+tid)] = 
    ((tid%2)==0) ? input0[((bid*32)+(tid>>1))] : 
                   input1[((bid*32)+(tid>>1))];
}
\end{verbatim}
\end{codesize}
If we wrote this CUDA program by hand, we would, again, split it up into 
two phases so that all threads can progress in parallel.

%Even though this is a pretty artificial example, it illustrates the 
%problem well. The situation is just as bad when using the function
%{\tt unpair}. This function takes an array of pairs and returns an 
%array of elements such that position 0 and 1 in the result array holds 
%the values of the pair at index 0 in the input array. Using this operation 
%results in a conditional statement that takes different execution paths 
%on odd and even numbered threads.

The arrays described so far, with an indexing function and a length, have 
been nicknamed {\em Pull arrays} for how they describe how to 
compute an element by {\em pulling} data from a number 
of places. Using just Pull arrays, we have been unable to solve the 
problems described so far in this section. The solution is to add a complementary array type to Obsidian.

%For a more in depth view of the parts of Obsidian that has been showed 
%in this section see~\cite{JSLIC}.

### 2013-09-30 src2/expressive/paperFinal.tex
another approach

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{url}

\newcommand{\codesize}{\fontsize{7}{8}\selectfont}

\begin{document}
\conferenceinfo{DAMP'12,} {January 28, 2012, Philadelphia, PA, USA.}
\CopyrightYear{2012}
\copyrightdata{978-1-4503-1117-5/12/01}

%\conferenceinfo{WXYZ '05}{date, City.} 
%\copyrightyear{2005} 
%\copyrightdata{[to be supplied]}

\titlebanner{}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

%\title{Towards an Implementation of Push Arrays in Obsidian}
\title{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
%\subtitle{Subtitle Text, if any}

%\authorinfo{Name1}
%           {Affiliation1}
%           {Email1}
\authorinfo{Koen Claessen \and Mary Sheeran \and Bo Joel Svensson}
           {Chalmers University of Technology, 
             Department of Computer Science and Engineering,
             Gothenburg, Sweden}
           {koen@chalmers.se/ms@chalmers.se/joels@chalmers.se}

\maketitle

% ABSTRACT !!! -----------------------------------------------------------------
\begin{abstract}
%This is the text of the abstract.

Graphics Processing Units (GPUs) are powerful computing devices 
that with the advent of CUDA/OpenCL are becomming useful for general 
purpose computations. 
Obsidian is an embedded domain specific language that generates CUDA kernels
from functional descriptions.
A symbolic array construction allows us to guarantee that intermediate
arrays are fused away. However, the current array construction has
some drawbacks; in particular, arrays cannot be combined efficiently.
We add a new type of {\em push arrays} to the existing Obsidian system in 
order to solve this problem. The two array types complement each other,
and enable the definition of combinators that both
take apart and combine arrays, and that result in efficient generated code.
This extension to Obsidian is demonstrated on a sequence of
sorting kernels, with good results.
The case study also illustrates the use of combinators for expressing
the structure of parallel algorithms.
The work presented is preliminary, and 
the combinators presented must be generalised.
However, the raw speed of the generated kernels bodes well.
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
\category{D.3.4}{Programming Languages}{Processors}[Code generation]

\terms
Languages, Algorithms, Performance

\keywords
Arrays, Data parallelism, Embedded Domain Specific Language, General Purpose GPU programming, Haskell

%\input{intro}
\section{Introduction}

%\input{table}

%----------------------------------------------------------------------------

Graphics Processing Units (GPUs) are parallel computers with hundreds 
to thousands of processing elements. The CUDA and OpenCL languages  
make available the power of the GPU to programmers interested in general 
purpose computations. In CUDA and OpenCL, the programmer writes {\em kernels}, 
Single Program Multiple Data (SPMD) programs that are executed by groups 
of threads on the available processing elements of the GPU.

CUDA and OpenCL are general purpose programming languages, mirroring the 
increased capabilities of a modern GPU to target that domain. However, these 
languages lack compositionality. 
Also, being based in C/C++ means that the core idea in a program may
not be easily visible.

\subsection{Embedded DSLs for GPGPU programming} 
We are aiming for a GPU programming language that is more
concise than mainstream
languages such as CUDA and OpenCL. 
Obsidian is a
domain specific embedded language (DSEL) implemented in Haskell.
When an Obsidian program is run, a representation of the program is created 
as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
The program representation generated when running an Obsidian program 
is compiled into CUDA code. We are also working on an OpenCL backend.

Our approach is different from that of other Haskell DSELs targetting
GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
We do not try to abstract away from 
all the peculiarities of GPU programming, but rather provide a higher 
level language in which to experiment with them.
For instance,
Accelerate provides a standard set of basic operations such as 
{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
Obsidian, on the other hand, allows the user to experiment with the
{\em generation} of small kernels for fixed size array inputs
from higher level descriptions.
It is intended to allow the user to play with the kinds of tradeoffs that are important
when writing such high performance building blocks; in this paper, the main consideration
is the number of array elements of the input and output that are manipulated
by a single thread in the generated CUDA code.
An important aspect of Obsidian is the symbolic array representation used, along
with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
allows the programmer to guide code generation and control parallelism and thread
use~\cite{JSLIC}.

%Today there are number of EDSLs written in Haskell targeting 
%GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
%another by having a slightly different approach. We do not try to abstract away from 
%all the peculiarities of GPU programming but rather giving a higher 
%level language in which to experiment with them.
%trying to appeal to a slightly different target audience. 
%We try to cater not for the advanced functional programmer who wants to 
%run his programs on GPUs, but rather to offer a higher level 
%language to the CUDA/OpenCL hacker. 
%In this language we want the user to be 
%able to quickly experiment with the kind of tradeoffs that are important 
%when writing a high performance GPU kernel.

%Unlike the other approaches that are complementary to our Haskell GPU EDSL, 
%Obsidian does not provide the usual set of basic operations such as 
%{\tt Map}, {\tt Reduce} and {\tt ZipWith} as built in skeletons. We 
%want a language where the programmer can experiment with implementing 
%his or her own versions of these basic operations. So far, we make
%this possible in Obsidian through a {\tt Sync} operation and the Array 
%representation we use. The {\tt Sync} operation, using target domain 
%nomenclature, is the programmers tool to guide code generation and 
%obtain parallelism. The presentation of Obsidian here will be somewhat 
%cursory, for a more complete view of the various versions and approaches 
%we have tried, see~\cite{JSLIC}.
%More details on this later.

In Obsidian, a kernel that sums an array can be expressed as: 
\begin{codesize}
\begin{verbatim}
sum :: Array IntE -> Kernel (Array IntE) 
sum arr | len arr == 1 = return arr
        | otherwise    = (pure (fmap (uncurry (+)) . pair) 
                          ->- sync 
                          ->- sum) arr                       
\end{verbatim}
\end{codesize}
The result of running this kernel on an eight element input array, 
{\tt runKernel sum (namedArray ``input'' 8)}, is an intermediate representation 
of the computation (shown in slightly pretty-printed form): 
\begin{codesize}
\begin{verbatim}
arr0 = malloc(16)
par i 4 {
arr0[i] = ( +  input[( *  i 2 )] input[( +  ( *  i 2 ) 1 )] );
}Sync
arr1 = malloc(8)
par i 2 {
arr1[i] = ( +  arr0[( *  i 2 )] arr0[( +  ( *  i 2 ) 1 )] );
}Sync
arr2 = malloc(4)
par i 1 {
arr2[i] = ( +  arr1[( *  i 2 )] arr1[( +  ( *  i 2 ) 1 )] );
}Sync
\end{verbatim}
\end{codesize}
The named intermediate arrays in this representation are then laid 
out in GPU shared memory and CUDA code can be generated (here for arrays of length eight)\footnote{An alignment qualifier for shared memory 
has been omitted to save space in the listings showing generated code}: 
%\pagebreak
\begin{codesize}
\begin{verbatim}
__global__ void sum(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] = 
    (input0[((bid*8)+(tid*2))]+
     input0[((bid*8)+((tid*2)+1))]);
  __syncthreads();
  if (tid<2){
    (( int *)(sbase + 16))[tid] = 
      ((( int *)sbase)[(tid*2)]+
       (( int *)sbase)[((tid*2)+1)]);
  }
  __syncthreads();
  if (tid<1){
    (( int *)sbase)[tid] = 
      ((( int *)(sbase+16))[(tid*2)]+
       (( int *)(sbase+16))[((tid*2)+1)]);
  }
  __syncthreads();
  if (tid<1){
    result0[(bid+tid)] = (( int *)sbase)[tid];
  }
}
\end{verbatim}
\end{codesize}
%The CUDA code above is generated by issuing the command:
%\begin{codesize}
%\begin{verbatim}
%genKernel ``sum'' sum (namedArray undefined 8 :: Array IntE)
%\end{verbatim}
%\end{codesize}
%\noindent
%and the kernel code is generated for arrays of length eight.

%----------------------------------------------------------------------------
\subsection{Arrays in Obsidian}

An array is represented by an indexing function and a length:
\begin{codesize} 
\begin{verbatim}
data Array a = Array (UWordE -> a) Word32 
\end{verbatim}
\end{codesize}
This array representation has served us well. It has these properties:
\begin{itemize} 
\item Fusion of operations is automatic.
\item It naturally describes a data-parallel computation suitable for CUDA/OpenCL generation.
\item Many basic operations can be implemented: {\tt map}, {\tt zipWith} etc.
\end{itemize}
Using this array representation in a DSEL is not new; the first occurence that we know of is in
Pan~\cite{PAN}. Similar array representations have also later been 
used in Feldspar~\cite{FELDSPAR}, and more recently also in the Repa library~\cite{REPA}.
Functions for indexing and getting the length of arrays are as follows:
\begin{codesize} 
\begin{verbatim}
(!) :: Array a -> UWordE -> a 
(Array ixf _) ! ix = ixf ix

len :: Array a -> Word32
len (Array _ n) = n 
\end{verbatim}
\end{codesize}
%Implementing a {\em Functor} instance for the {\tt Array} datatype is possible:
A {\em Functor} instance for the {\tt Array} datatype is
\begin{codesize} 
\begin{verbatim}
instance Functor Array where 
  fmap f arr = Array (\ix -> f (arr ! ix)) (len arr) 
\end{verbatim}
\end{codesize}
Now, composed applications of {\tt fmap} will be automatically fused. This is 
illustrated in the example program below and the CUDA generated
from it.  
\begin{codesize} 
\begin{verbatim}
mapFusion :: Array IntE -> Kernel (Array IntE) 
mapFusion = pure (fmap (+1) . fmap (*2)) 
\end{verbatim}
\end{codesize}

\begin{codesize} 
\begin{verbatim}
__global__ void mapFusion(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = ((input0[((bid*32)+tid)]*2)+1);
}
\end{verbatim}
\end{codesize}
Both of these code listings need explanation. In the Haskell 
code, {\tt mapFusion} has type {\tt Array IntE -> Kernel (Array IntE)};
{\tt Kernel} is a state monad that accumulates CUDA code as well as provides 
new names for intermediate arrays. Neither of these features of the monad 
is activated by this example though. The {\tt pure} function is defined 
using the monad's {\tt return} as {\tt pure f a = return (f a)}. In this case, 
it lifts a function of type {\tt Array IntE -> Array IntE} into 
a kernel.

The generated CUDA code computes the result array using a number of threads 
equal to the length of that array. In this case, the kernel was generated
to deal with arrays of length $32$. The important detail to notice in the 
CUDA code is that there is no intermediate array created between the 
{\tt (*2)} and the {\tt (+1)} operations.

The {\tt mapFusion} example could just as well have been implemented using 
the kernel sequential composition combinator, {\tt ->-}. 
\begin{codesize} 
\begin{verbatim}
mapFusion :: Array IntE -> Kernel (Array IntE) 
mapFusion = pure (fmap (*2)) ->- pure (fmap (+1)) 
\end{verbatim}
\end{codesize}
Exactly the same CUDA code is then generated.

In some cases, it is necessary to force computation of 
intermediate arrays. This can be used to share partial computations between 
threads and to expose parallelism. In Obsidian, the tool for this
is called {\tt sync}, a built-in kernel. Using {\tt sync} as follows prevents fusion of the
two operations:
\begin{codesize} 
\begin{verbatim}
mapUnFused :: Array IntE -> Kernel (Array IntE) 
mapUnFused = pure (fmap (*2)) ->- sync ->- pure (fmap (+1))
\end{verbatim}
\end{codesize}
The generated CUDA code now stores an intermediate result in local shared
memory before moving on. 
\begin{codesize} 
\begin{verbatim}
__global__ void mapUnFused(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] = (input0[((bid*32)+tid)]*2);
  __syncthreads();
  result0[((bid*32)+tid)] = ((( int *)sbase)[tid]+1);
}
\end{verbatim}
\end{codesize}
Intermediate arrays are laid out in the {\tt sbase} array in shared memory. 
Since we may store arrays of many different types in the same locations 
of the shared memory at different times during the execution, the type casts
used in the code above are necessary.

%----------------------------------------------------------------------------
\subsection{Sync and parallelism}

The {\tt sync} operation also enables the writing of parallel reduction 
kernels. A reduction operation is an operation that takes an array as input
and
produces a singleton array as output.

First, we define {\tt zipWith} and {\tt halve} on Obsidian arrays.
\begin{codesize} 
\begin{verbatim}
zipWith :: (a -> b -> c) -> Array a -> Array b -> Array c
zipWith op a1 a2 = Array (\ix -> (a1 ! ix) `op` (a2 ! ix)) 
                   (min (len a1) (len a2))

splitAt :: Word32 -> Array a -> (Array a, Array a) 
splitAt n arr = 
  (Array (\ix -> arr ! ix) n , 
   Array (\ix -> arr ! (ix + fromIntegral n)) (len arr - n))

halve arr = splitAt ((len arr) `div` 2) arr
\end{verbatim}
\end{codesize}
A reduction kernel that takes an array whose length is a power of two 
and gives an array of length one can be defined recursively. 
Defining kernels recursively results in completely unrolled CUDA kernels,
and kernel input size must be known at compile time.
The approach to reduction taken here is to split the input array 
into two halves and then apply {\tt zipWith} of the combining function to the 
two halves, repeating the process until the length is one. 
\begin{codesize} 
\begin{verbatim}
reduceS :: (a -> a -> a) -> Array a -> Kernel (Array a) 
reduceS op arr | len arr == 1 = return arr
               | otherwise    = 
                 (pure ((uncurry (zipWith op)) . halve)
                  ->- reduceS op) arr
\end{verbatim}
\end{codesize}
%This Obsidian Kernel takes an array of length $2n$ as input and the output 
Since the output of this kernel is of length one, and the number 
of elements in the output array specifies the number of threads used to compute it, 
this function, {\tt reduceS}, defines a sequential reduction. 
The generated code for arrays of length eight is 
\pagebreak
%The code is generated with the 
%command {\tt genKernel ``reduceSAdd'' (reduce (+)) (namedArray undefined 8 :: Array IntE)}, 
%meaning that a CUDA reduce-with-addition kernel for arrays of length eight is generated. 
\begin{codesize} 
\begin{verbatim}
__global__ void reduceSAdd(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[(bid+tid)] = 
    (((input0[((bid*8)+tid)]+
       input0[((bid*8)+(tid+4))])+
      (input0[((bid*8)+(tid+2))]+
       input0[((bid*8)+((tid+2)+4))]))+
     ((input0[((bid*8)+(tid+1))]+
       input0[((bid*8)+((tid+1)+4))])+
      (input0[((bid*8)+((tid+1)+2))]+
       input0[((bid*8)+(((tid+1)+2)+4))])));
}
\end{verbatim}
\end{codesize}
Sequential reduction is not very interesting for GPU execution, but 
the fix is simple. A well placed use of {\tt sync} indicates that 
we want to compute, after each {\tt zipWith} phase, the intermediate 
arrays using as many threads as that intermediate array is long. 
The effect is shown in the code below. 
\begin{codesize} 
\begin{verbatim}
reduce :: Syncable Array a 
          => (a -> a -> a) -> Array a -> Kernel (Array a)
reduce op arr | len arr == 1 = return arr
              | otherwise    = 
                (pure ((uncurry (zipWith op)) . halve)
                 ->- sync 
                 ->- reduce op) arr
\end{verbatim}
\end{codesize}
The CUDA code for reduction with addition on eight
elements is
\begin{codesize} 
\begin{verbatim}
__global__ void reduceAdd(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] = 
    (input0[((bid*8)+tid)]+
     input0[((bid*8)+(tid+4))]);
  __syncthreads();
  if (tid<2){
    (( int *)(sbase + 16))[tid] = 
      ((( int *)sbase)[tid]+
       (( int *)sbase)[(tid+2)]);   
  }
  __syncthreads();
  if (tid<1){
    (( int *)sbase)[tid] = 
      ((( int *)(sbase+16))[tid]+
       (( int *)(sbase+16))[(tid+1)]);    
  }
  __syncthreads();
  if (tid<1){
    result0[(bid+tid)] = (( int *)sbase)[tid];  
  }
}
\end{verbatim}
\end{codesize}           
In this generated CUDA, three phases can be identified. 
The first uses four threads to compute a four element intermediate array;
the second uses two threads, and so on. At the very end, a single thread
copies the result from local shared memory to global memory.

\subsection{Drawbacks of Obsidian Arrays}
\label{sec:Drawbacks}

%*** THE STUFF BELOW IS IN CONFLICT WITH WHAT I ADDED ABOVE!
%When using Obsidian we have available to us a toolbox of higher order
%functions that are familiar to functional programmers. One example of such a 
%function is {\tt zipWith}.

%\begin{codesize} 
%\begin{verbatim}
%example1 :: (Array IntE, Array IntE) 
%            -> Kernel (Array IntE)
%example1 (arr1,arr2) = return$ zipWith (+) arr1 arr2
%\end{verbatim}
%\end{codesize}

%The body of this function looks very familiar to Haskell programmers. The 
%type of the function however is a little bit different than what we are 
%used to from everyday Haskell usage. In Obsidian we have a concept called 
%a Kernel. Computations of types {\tt a -> Kernel b} can be turned into 
%runnable GPU code by providing an input. 
%*** VERY VAGUE AND ODD 
%*** TODO: Explain kernel ?

%The following code listing shows the CUDA code we can generate from 
%the Obsidian program given above:

% *** TODO: Needs a code size between tiny and small.

%\begin{codesize}
%\begin{verbatim}
%__global__ void example1(int *input0,int *input1,int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;

%  result0[tid] = (input0[((bid*32)+tid)]+
%                  input1[((bid*32)+tid)]);

%}
%\end{verbatim}
%\end{codesize}

%The CUDA code above writes into a result array the pairwise sums from 
%its two input arrays, as expected. 
%*** There are many details here. tid, bid, 32. explain now or push to later?

% \subsection{Obsidian arrays}

%In Obsidian an array is represented by an indexing function and a length. 
%\begin{codesize}
%\begin{verbatim}
%data Array a = Array (IxExp -> a) Word32
%\end{verbatim}
%\end{codesize}

%And {\tt IxExp} is defined as {\tt UWordE}, that is an expression 
%representing a unsigned 32bit value. The length of an array is represented
%with a unsigned 32bit value as well but here it has been beneficial 
%to use a built in Haskell type in order to be able to use this 
%information at program generation time rather than program run time.

%This array representation specifies how to compute each index of an Arrays. 
%If we are interested in the value at location 5 in an array all we need 
%to do is apply the indexing function to the literal expression 5 and 
%we get a description of how to compute that value.

%This array representation have been working fairly well in many situations. 
%In {\tt example1} where we used {\tt zipWith} we got just about the resulting
%CUDA code that a CUDA programmer would write himself. For some other examples 
%the situation is much worse.

The previous subsection described positive aspects of the array 
representation that we have used so far. There are, however, 
circumstances in which this Array representation is too restricted.

Take the problem of concatenating two arrays. Using the array representation 
described above, the only way to concatenate two arrays is to introduce 
a conditional into the indexing function. If {\tt f} and {\tt g} are the 
indexing functions of two arrays that are to be concatenated, and {\tt n1}
is the length of the first array, the indexing 
function of the result must be
\begin{codesize}
\begin{verbatim}
new ix = if (ix < n1) 
         then f ix 
         else g (ix - n1)
\end{verbatim}
\end{codesize}
The following program
concatenates two arrays: 
\begin{codesize}
\begin{verbatim}
catArrays :: (Array IntE, Array IntE) -> Kernel (Array IntE)
catArrays = pure conc
\end{verbatim}
\end{codesize}
%(arr1,arr2) = return$ conc (arr1,arr2)
When it is used to generate a CUDA kernel that concatenates
two arrays of length $16$, the following code is the result: 
\begin{codesize}
\begin{verbatim}
__global__ void catArrays(int *input0,int *input1,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = 
    (tid<16) ? input0[((bid*16)+tid)] : 
               input1[((bid*16)+(tid-16))];
}
\end{verbatim}
\end{codesize}
%__global__ void catArrays(int *input0,
%                          int *input1,
%                          int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;
%  
%  result0[((bid*64)+tid)] = 
%    (tid<32) ? input0[((bid*32)+tid)] : 
%               input1[((bid*32)+(tid-32))];
%  
%}
Now, conditionals like these are {\em bad} in code 
to execute on a GPU, with its wide-SIMD data-parallel model. Separating
the operation into two assignments and using half as many threads gives
much higher performance.
\begin{codesize}
\begin{verbatim}
__global__ void catArraysByHand(int *input0,
                                int *input1,
                                int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = input0[((bid*16)+tid)];
  result0[((bid*32)+tid+16)] = input1[((bid*16)+tid)];  
}
\end{verbatim}
\end{codesize}
There are cases where code with conditionals is not that bad. 
An expert on NVIDIA GPUs in particular may say that code with 
the condition {\tt (tid < 32)} is fine, since 32 is the SIMD width 
of those GPUs. However, any number that is not a multiple of 32 would 
lead to poor performance, so in general this is a problem. 
 Worse still, {\em zipping} two arrays together and then 
{\em unpairing} (to get an array of elements)
leads to code that takes two different paths depending 
on odd or even {\em thread id}. When a GPU executes such code, it 
shuts down half of the threads and computes the two paths in sequence. 
\begin{codesize}
\begin{verbatim}
zippUnpair :: (Array IntE, Array IntE) -> Kernel (Array IntE) 
zippUnpair = pure (unpair . zipp)
\end{verbatim}
\end{codesize}
The {\tt zipp} and {\tt unpair} operations are defined as follows: 
\begin{codesize}
\begin{verbatim}
zipp :: (Array a, Array b) -> Array (a, b)             
zipp (arr1,arr2) = 
     Array (\ix -> (arr1 ! ix, arr2 ! ix)) 
           (min (len arr1) (len arr2))

unpair :: Choice a => Array (a,a) -> Array a
unpair arr = 
    let n = len arr
    in  Array (\ix -> ifThenElse ((mod ix 2) ==* 0) 
                      (fst (arr ! (ix `shiftR` 1)))
                      (snd (arr ! (ix `shiftR` 1)))) (2*n)
\end{verbatim}
\end{codesize}
Code generated from the {\tt zippUnpair} program exhibits really 
poor performance; at any time half of the threads are shut down. 
\begin{codesize}
\begin{verbatim}
__global__ void zippUnpair(int *input0,
                           int *input1,
                           int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*64)+tid)] = 
    ((tid%2)==0) ? input0[((bid*32)+(tid>>1))] : 
                   input1[((bid*32)+(tid>>1))];
}
\end{verbatim}
\end{codesize}
If we wrote this CUDA program by hand, we would, again, split it up into 
two phases so that all threads can progress in parallel.

%Even though this is a pretty artificial example, it illustrates the 
%problem well. The situation is just as bad when using the function
%{\tt unpair}. This function takes an array of pairs and returns an 
%array of elements such that position 0 and 1 in the result array holds 
%the values of the pair at index 0 in the input array. Using this operation 
%results in a conditional statement that takes different execution paths 
%on odd and even numbered threads.

The arrays described so far, with an indexing function and a length, have 
been nicknamed {\em Pull arrays} for how they describe how to 
compute an element by {\em pulling} data from a number 
of places. Using just Pull arrays, we have been unable to solve the 
problems described so far in this section. The solution is to add a complementary array type to Obsidian.

%For a more in depth view of the parts of Obsidian that has been showed 
%in this section see~\cite{JSLIC}.

%\input{push} 
\section{Push Arrays}

In order to improve low level control for the programmer, {\em Push
arrays} are added to Obsidian. The old Pull arrays are still available,
along with the new array type.

Some operations, typically involving taking arrays apart,
are easily described using Pull arrays, giving
efficient code. In those cases, using a Push array would add complexity in the 
implementation for no performance benefit.
Other operations cannot be implemented efficiently
with Pull arrays, but Push arrays then provide the solution.
This duality is apparent when looking at operations on Pull arrays such as
{\tt halve} and {\tt conc} (for concatenate). 
The {\tt halve} function is efficient since it introduces no diverging 
conditionals. The {\tt conc} function, on the other hand, introduces conditionals.
Concatening two arrays using the {\tt concP} combinator, implemented on Push arrays, 
allows us to generate the desired code:

\begin{codesize}
\begin{verbatim}
catArrayPs :: (Array IntE, Array IntE) -> Kernel (ArrayP Int)
catArrayPs = pure concP
\end{verbatim}
\end{codesize}

\begin{codesize}
\begin{verbatim}
__global__ void catArrayPs(int *input0,int *input1,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*32)+tid)] = input0[((bid*16)+tid)];
  result0[((bid*32)+(16+tid))] = input1[((bid*16)+tid)];

}
\end{verbatim}
\end{codesize}

\noindent
Compared to the CUDA code for {\tt catArrays}, this kernel 
uses only $16$ threads instead of $32$. At each step of the computation, all 
the threads are fully busy doing exactly the same thing, which is the preferred 
mode of execution on the target platform.

%The performance effects of these unwanted conditionals, compared to code generated 
%using the new Push arrays, will be illustrated in section~\ref{sec:MARY}.

%The {\tt halve} operation has appeared previously and the {\tt pair} 
%operation is its inverse. Implementing the {\tt pair} operation using 
%Pull arrays is very easy and also efficient:

%\begin{codesize}
%\begin{verbatim}
%pair :: Array a -> Array (a,a)
%pair (Array ixf n) = Array (\ix -> (ixf (ix*2),ixf (ix*2+1))) n'
%  where 
%    n' = n `div` 2 
%\end{verbatim}
%\end{codesize}

%It is efficient because it introduces no diverging conditionals. There is
%just a tiny bit of extra arithmetic in the indexing function. On the 
%other hand {\tt unpair}, as we saw, introduced an unnecessary and 
%inefficient conditional. It turns out that {\tt unpair} can be 
%efficiently implemented on a Push array.

\subsection {What are Push Arrays?}

The idea behind Push arrays is to have a way to describe where 
elements are supposed to end up. In some sense, a Push array produces 
a collection of Index/Value pairs. %This makes a Push array a lot more 
%powerful than a Pull array. 
This makes Push arrays complementary to Pull arrays. For example, it 
is possible for a Push array to output several elements at the same index 
(which we probably need to control carefully). Push arrays should permit 
us to provide more expressive operations on arrays to the user, including 
an operation similar to Haskell's {\tt filter} on lists. Here, we consider 
a different advantage of adding push arrays: finer control over patterns of 
thread use in generated code.

%% Will it really give nondeterminism or will the last one written win, or some such?

A Push array consists of three parts: a function in continuation passing style, 
a {\tt Program} datatype and an array datatype.

\begin{codesize}
\begin{verbatim}
type P a = (a -> Program) -> Program 
\end{verbatim}
\end{codesize}

\noindent
For another example of using continuations and a more complete
description of their meaning and application, see~\cite{POORKOEN}.

The {\tt Program} datatype has now been adopted as Obsidian's internal 
representation of CUDA programs.

\begin{codesize}
\begin{verbatim}
data Program
  = Skip
  | forall a. Scalar a => Assign Name UWordE (Exp a) 
  | Par (UWordE -> Program) Word32   
  | Allocate Name Word32 Type 
  | Synchronize 
  | ProgramSeq Program 
               Program 
\end{verbatim}
\end{codesize}

\noindent
Even Obsidian programs 
that never explicitly uses a Push array will also be represented by this 
datatype.

Note that the {\tt Par} constructor, 
the {\em parallel for loop}, could potentially introduce nesting, which would lead 
to {\em nested data-parallelism}. We do not compile nested data parallelism 
into CUDA, and right now this is guaranteed by taking care not to introduce 
any nesting in the library functions provided. Some of the 
simpler cases of nestedness should be possible to take care of quite easily. 
For example, one extra level of nesting could be done by sequential execution 
in each thread of the GPU; using sequential computations per thread has 
been shown to be beneficial~\cite{OLAMARCUS}. But for the general case 
of arbitrary nesting, some method of flattening is needed. 
We also assume that both {\tt Allocate} and {\tt Synchronize} occur only at the top level in objects of type {\tt Program}.

Now, a Push array is a function in continuation passing style
coupled with a length.

\begin{codesize}
\begin{verbatim}
data ArrayP a = ArrayP (P (UWordE, a)) Word32
\end{verbatim}
\end{codesize}

There is a function that takes an array and turns it into a Push array, 
called {\tt push}. This function is defined for both Pull and Push arrays:

\begin{codesize}
\begin{verbatim} 
class Pushable a where 
  push :: a e -> ArrayP e

instance Pushable ArrayP where 
  push = id

instance Pushable Array where   
  push (Array ixf n) =
     ArrayP (\func -> Par (\i -> func (i,(ixf i))) n) n 
\end{verbatim}
\end{codesize}

Going in the other direction, from a Push array to a Pull array, is a 
costly operation; it involves writing all the elements to GPU memory 
followed by creating a Pull array that represents reading them. 
The task of writing intermediate values to memory 
has traditionally been up to the {\tt sync} operation in Obsidian.
Therefore, in this version, {\tt sync} is overloaded to operate on both Pull and 
Push arrays.  This means that the 
{\tt sync} operation can be used both on arrays of type {\tt Array} 
and of type {\tt ArrayP}. The result type, however, is always {\tt Array}.

When a Push array is synced, it is applied to a continuation that writes the
elements into a named array in memory. The name to use is obtained through the 
{\tt Kernel} monad.

\begin{codesize}
\begin{verbatim}
targetArray :: Scalar a => Name -> (UWordE,Exp a) -> Program
targetArray n (i,a) = Assign n i a 
\end{verbatim}
\end{codesize}

After applying the Push array to {\tt targetArray <name>}, the {\tt sync}
operation proceeds by storing away a representation of the program that 
computes the array called {\tt <name>}; it returns a Pull array 
that reads elements from that same array.

Now we have seen enough of the implementation of Push arrays to be able 
to look at some operations. Earlier, we saw that the
array concatenation function
{\tt conc} on Pull arrays leads to inefficient code. The Push 
version of this operation, called {\tt concP} can be implemented 
as follows:

\begin{codesize}
\begin{verbatim}
concP :: (Pushable arr1,
          Pushable arr2) => (arr1 a, arr2 a) -> ArrayP a     
concP (arr1,arr2) = 
  ArrayP (\func -> f func
                   *>* 
                   g (\(i,a) -> func (fromIntegral n1 + i,a)))
         (n1+n2)
  where 
     ArrayP f n1 = push arr1
     ArrayP g n2 = push arr2
\end{verbatim}
\end{codesize}

\noindent
The function {\tt concP} takes two arrays, that can be Push or Pull arrays, 
and concatenates them into a single Push array. It does so by creating a 
sequential program, using the {\tt *>*} operator for {\tt Program} sequential
composition. An example use of this combinator has already been displayed in 
the {\tt catArrayPs} example.

%If the {\tt catArrays} kernel is reimplemented using {\tt concP}, CUDA code 
%without conditionals is generated. The new version of {\tt catArrays}, 
%called {\tt catArrayPs} is:

%\begin{codesize}
%\begin{verbatim}
%catArrayPs :: (Array IntE, Array IntE) -> Kernel (ArrayP Int)
%catArrayPs = pure concP
%\end{verbatim}
%\end{codesize}

%And the CUDA code it generates is:

%\begin{codesize}
%\begin{verbatim}
%__global__ void catArrayPs(int *input0,int *input1,int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;
%  
%  result0[((bid*32)+tid)] = input0[((bid*16)+tid)];
%  result0[((bid*32)+(16+tid))] = input1[((bid*16)+tid)];
%  
%}
%\end{verbatim}
%\end{codesize}
%__global__ void catArrayPs(int *input0,
%                           int *input1,
%                           int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;
%  
%  result0[((bid*64)+tid)] = input0[((bid*32)+tid)];
%  result0[((bid*64)+(32+tid))] = input1[((bid*32)+tid)];
%}

The {\tt zippUnpair} example shows a drawback similar to that of {\tt catArrays} using Pull arrays. 
In this case, the problem is that the {\tt unpair} function introduces 
a conditional that takes different paths depending on odd or even {\em thread id}. 
A Push array implementation of the {\tt unpair} operation called {\tt unpairP}
can be given as follows:

\begin{codesize}
\begin{verbatim}
unpairP :: Pushable arr => arr (a,a) -> ArrayP a 
unpairP arr =  ArrayP (\k -> f (everyOther k)) (2 * n)
  where 
    ArrayP f n = push arr

everyOther :: ((UWordE, a) -> Program ()) 
              -> (UWordE, (a,a)) -> Program ()
everyOther f  = \(ix,(a,b)) -> f (ix * 2,a) *>* f (ix * 2 + 1,b)  
\end{verbatim}
\end{codesize}

\noindent
Just like {\tt concP}, this function takes either a Push or Pull array 
as input, and produces a Push array as result.

Rewriting the example from earlier using {\tt unpairP} gives:

\begin{codesize}
\begin{verbatim}
zippUnpairP :: (Array IntE, Array IntE) -> Kernel (ArrayP IntE) 
zippUnpairP = pure (unpairP . zipp)
\end{verbatim}
\end{codesize}

\noindent
In this case, the generated code looks as follows:

\begin{codesize}
\begin{verbatim}
__global__ void zippUnpairP(int *input0,int *input1,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;

  result0[((bid*64)+(tid*2))] = input0[((bid*32)+tid)];
  result0[((bid*64)+((tid*2)+1))] = input1[((bid*32)+tid)];
}
\end{verbatim}
\end{codesize}

\noindent
Again, we get CUDA code that uses half as many threads as the inefficient 
version, but all threads are occupied at all times. This uses
the resources more efficiently.

%\subsection{Push arrays and Obsidian}

Being able to generate the kind of code that we have just seen is something we have desired for a long time. We believe that 
Push arrays are an important tool for obtaining high performance kernels. The results 
in section~\ref{sec:MARY} bear this out.

%Section~\ref{sec:BENEFITS} showed some examples where Push arrays 
%was helpful. Lets look at how the operations on Push arrays used there 
%are implemented. First there was {\tt concP}:

%The function {\tt concP} can be used to concatenate any arrays that we can 
%perform the {\tt push} function on. For example both input arrays could be 
%normal Pull arrays. Concatenating the two arrays uses sequential composition 
%of programs, {\tt *>*}, which is implemented using the {\tt ProgramSeq} 
%constructor.

%The next function used in the examples was {\tt unpairP}

%\subsection {Benefits of Push Arrays}
%\label{sec:BENEFITS}

%in section~\ref{sec:Drawbacks} some drawbacks of Pull arrays where 
%highlighted. If we use Push arrays we can improve the generated code 
%in those examples. Push arrays here have type {\tt ArrayP} and operations 
%that operate on Push arrays or give Push arrays as result end their names
%with a capital p.

%In the case of array concatenation using Push arrays we get:

%\begin{codesize}
%\begin{verbatim}
%catArrayPs :: (Array (Exp Int), Array (Exp Int)) 
%           -> Kernel (ArrayP (Exp Int))
%catArrayPs = pure concP
%\end{verbatim}
%\end{codesize}
%(arr1,arr2) = 
%   return$ concP (arr1, arr2)

%The function {\tt push} is used to turn a Pull array into a Push array. 
%Now, precisely the code we desired can be generated:

%\pagebreak

%\begin{codesize}
%\begin{verbatim}
%__global__ void catArrayPs(int *input0,
%                           int *input1,
%                           int *result0){
%  unsigned int tid = threadIdx.x;
%  unsigned int bid = blockIdx.x;

%  result0[((bid*64)+tid)] = input0[((bid*32)+tid)];
%  result0[((bid*64)+(32+tid))] = input1[((bid*32)+tid)];

%}
%\end{verbatim}
%\end{codesize}

%The next example uses {\tt unpair}:

%\subsection { Thoughts about Push arrays in Obsidian }

%\input{applications}
\section{Application}
\label{sec:MARY}

\subsection{Sorting on a GPU}
In this section, we introduce combinators that express patterns
of computation on whole arrays, and show their application to the
development of fast sorting kernels.
By kernel, we mean a computation that is performed by multiple
threads, each performing the same computation, in a single block.
The computation is performed entirely on the GPU, operating on
a short array, which has been placed into shared memory.
On the GPU on which we perform measurements, the maximum number
of threads in a single block is 512. Thus, we will build and benchmark
a sequence of
kernels that sort 512 inputs.
Our first kernels are implemented using one thread per array element.
Next, we show how Push arrays allow us to move to having each
thread operate on two array elements, giving a substantial performance
improvement.

Sorting kernels are typically used as building blocks in larger
programs to sort much larger sequences of inputs. In section~\ref{sec:benchmarks},
we show how to build a sorter for large arrays from small building blocks, including
small kernels for sorting and merging that are generated from Obsidian.
Our small kernels are constructed in the form of sorting and merging {\em networks},
building on Batcher's bitonic merger~\cite{Batcher} and
on the periodic balanced merger~\cite{PeriodicBalanced}.
We chose also to implement the large sorter used  in benchmarking the
small kernels as a sorting network. However, large
sorters that are not themselves sorting networks (with typical examples being radix sort and quicksort) often call small sorting networks when they need to sort small arrays during their execution.
Thus, small, fast sorting kernels have a variety of uses.

\subsection{Describing Batcher's bitonic merger}

The bitonic merger is typically presented as a recursive
construction and we have earlier explored ways to describe
and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LavaSorter}.
Here, we consider iterative descriptions using similar combinators.

Figure~\ref{fig:bitonicMerger} illustrates the merger for 16 inputs.
Data flows from left to right.
The vertical lines indicate components that operate on two array elements, placing the minimum onto the lower (abstract) wire, and the maximum onto the other output
of the component.
The leftmost {\em stage} operates (for $n=16$) on elements that are 
$8$ apart. the next stage deals with elements that are $4$ apart, and so on.

\begin{figure}
\centering
\includegraphics[scale=0.25]{bitonic}
\caption{A diagram of a 16 input bitonic merging network, using
a style that is standard in the literature. Note
that in each stage containing 8 min/max or comparator components, all
8 operate on independent parts of the input and so can proceed in parallel.
}
\label{fig:bitonicMerger}
\end{figure}

We introduce a combinator {\tt ilv1}, for {\em interleave}, that captures this pattern.
{\tt ilv1 i f g} applies {\tt f} to elements {\small $2^i$} apart,
producing the output at the lower of the two input indices; it applies {\tt g}
to the same pairs of elements, producing the output on the upper index.
Defining {\tt stage i} to be {\tt ilv1 i min max},
the four stages in the diagram are simply
{\tt stage} applied to{\small  $3$, $2$ $1$} and {\small $0$}.
The definition of {\tt ilv1} makes use of the fact that
flipping the bit $i$ of an index (using
the function {\tt flipBit}) gives the index of the element
that will be combined with it using the functions{\tt f} and {\tt g}.
The decision about whether to apply {\tt f} or {\tt g} is made
by looking at the value of bit $i$.
As we shall see, the use of the Obsidian {\tt ifThenElse}
produces conditionals in the resulting CUDA.
\begin{codesize}
\begin{verbatim}
lowBit :: Int -> UWordE -> Exp Bool
lowBit i ix = (ix .&. bit i) ==* 0

flipBit :: Bits a => Int -> a -> a
flipBit = flip complementBit

ilv1 :: Choice a => 
        Int -> (b -> b-> a) -> (b -> b -> a) -> 
        Array b -> Array a
ilv1 i f g arr = Array ixf (len arr)
  where
    ixf ix = let l = arr ! ix
                 r = arr ! newix
                 newix = flipBit i ix
             in (ifThenElse (lowBit i ix) (f l r) (g l r))
\end{verbatim}
\end{codesize}
\noindent
Expressing {\tt ilv1} using bit-flipping may seem strange, but it has
the advantage that it actually applies the desired pattern of
computation repeatedly over larger input arrays.
Now, for {\small $2^n$} inputs, a Haskell list containing the $n$ calls
of this interleave combinator are built:
\begin{codesize}
\begin{verbatim}
bmerge :: Int -> [Array IntE -> Array IntE]
bmerge n = [istage (n-i) | i <- [1..n]]
  where istage i = ilv1 i min max
\end{verbatim}
\end{codesize}
\noindent
Finally, the {\tt compose} function makes each element of the list
into a kernel (using {\tt map pure}) and places a {\tt sync} between
each kernel (using {\tt composeS}).
\begin{codesize}
\begin{verbatim}
compose :: (Scalar a) => 
           [Array (Exp a) -> Array (Exp a)] 
           -> Array (Exp a) -> Kernel (Array (Exp a))
compose = composeS . map pure

runm k = putStrLn$ CUDA.genKernel "bitonicMerge" 
         (compose (bmerge k)) (namedArray "inp" (2^k))
\end{verbatim}
\end{codesize}
\noindent
Note that {\tt bmerge k} works on inputs of length {\small }$2^{k+j}$, for $j > 0$, applying the merger to sub-sequences of length {\small $2^k$}.
The CUDA code for {\tt bmerge 4} on $16$ inputs
(with some newlines inserted) is
\begin{codesize}
\begin{verbatim}
*Main> runm 4
__global__ void bitonicMerge(int *input0,int *result0){
  unsigned int tid = threadIdx.x;
  unsigned int bid = blockIdx.x;
  extern __shared__ unsigned char sbase[];
  (( int *)sbase)[tid] 
     = ((tid&8)==0) 
      ? min(input0[((bid*16)+tid)],input0[((bid*16)+(tid^8))]) 
      : max(input0[((bid*16)+tid)],input0[((bid*16)+(tid^8))]);
  __syncthreads();
  (( int *)(sbase + 64))[tid] 
     = ((tid&4)==0) 
      ? min((( int *)sbase)[tid],(( int *)sbase)[(tid^4)]) 
      : max((( int *)sbase)[tid],(( int *)sbase)[(tid^4)]);
  __syncthreads();
  (( int *)sbase)[tid] 
     = ((tid&2)==0) 
      ? min((( int *)(sbase+64))[tid],(( int *)(sbase+64))[(tid^2)]) 
      : max((( int *)(sbase+64))[tid],(( int *)(sbase+64))[(tid^2)]);
  __syncthreads();
  (( int *)(sbase + 64))[tid] 
     = ((tid&1)==0) 
      ? min((( int *)sbase)[tid],(( int *)sbase)[(tid^1)]) 
      : max((( int *)sbase)[tid],(( int *)sbase)[(tid^1)]);
  __syncthreads();
  result0[((bid*16)+tid)] = (( int *)(sbase+64))[tid];
\end{verbatim}
\end{codesize}

%% ** TODO Note that bmerge k not only works on inputs of length $2^{k+j}$,
%% for $j > 0$, applying the merger to sub-sequences of length $2^k$.

\subsection{Modifying the bitonic merger}

The bitonic merger for which we have just generated a kernel is known
to sort so-called bitonic sequences, which include sequences whose first
half is sorted in one direction and whose second half is sorted in the other
direction. This fact can be used to build the well-known bitonic sorting
network. However, a GPU implementation typically needs to check, for each comparator,
whether or not it should sort upwards or downwards, see for instance
the simple CUDA implementation shown in Appendix A.
We choose here to modify the merger so that it sorts two concatenated
sequences that are sorted in the {\em same} direction.
We do this by using a well-known trick, reversing half of the input to the merger. It turns out that
we can also reverse the same half of the output of the first stage of
the network, without affecting overall behaviour.
The resulting network, {\tt tmerge}, shown in Figure~\ref{fig:mixedMerger},
encourages us to develop a new combinator to describe the characteristic
V-shaped pattern that results in the first stage.
The combinator is modelled on {\tt ilv1}. The only difference
is that the ``partner'' of an index is found not by flipping bit $i$, but
by flipping bits $0$ to $i$, using function {\tt flipLSBsTo}. The implementation
of {\tt vee1} is got from that for {\tt ilv1} by replacing the call of
{\tt flipBit} by one of {\tt flipLSBsTo} (and we could also have chosen
to make a more generic function that is parameterised on this {\em partner} function).
\begin{codesize}
\begin{verbatim}
flipLSBsTo :: Int -> UWordE -> UWordE
flipLSBsTo i = (`xor` (oneBits (i+1)))

vee1 :: Choice a => 
        Int -> (b -> b-> a) -> (b -> b -> a) -> 
        Array b -> Array a
vee1 i f g arr = Array ixf (len arr)
  where
    ixf ix = let l = arr ! ix
                 r = arr ! newix
                 newix = flipLSBsTo i ix
             in (ifThenElse (lowBit i ix) (f l r) (g l r))
\end{verbatim}
\end{codesize}

\begin{figure}
\centering
\includegraphics[scale=0.25]{mixed}
\caption{16 input merging network, {\em tmerge}. The first stage
is made with a new combinator that we call vee, while
the remaining stages are as in the bitonic merger.
This network sorts an input that consists of two half-sized
sorted sequences, giving the opportunity to build a tree-shaped
sorting network.}
\label{fig:mixedMerger}
\end{figure}

\begin{codesize}
\begin{verbatim}
tmerge :: Int -> [Array IntE -> Array IntE]
tmerge n = vstage (n-1): [istage (n-i) | i <- [2..n]]
  where
    vstage i = vee1 i min max 
    istage i = ilv1 i min max
\end{verbatim}
\end{codesize}

Now that we have a merger that sorts sub-sequences containing two concatenated sorted sequences, it
is easy to make a tree of them.
The list of kernels to be composed now becomes
\begin{codesize}
\begin{verbatim}
tsort1 :: Int -> [Array IntE -> Array IntE]
tsort1 n = concat [tmerge i | i <- [1..n]]
\end{verbatim}
\end{codesize}
\noindent
The resulting sorting network is shown in Figure~\ref{fig:mixedsorter}.

\begin{figure}
\centering
\includegraphics[scale=0.33]{mmixedsorter}
\caption{A 16-input sorter made from a tree of {\em tmerge} mergers. 8 2-input mergers
feed 4 4-input mergers, followed  by 2 8-input and one 16-input merger.
A sorter on 8 inputs is shaded, as is a merger on 8 inputs, above it. 
}
\label{fig:mixedsorter}
\end{figure}
\noindent
The following call writes the resulting CUDA to file {\tt tsort1.cu}
and this is the first {\em generated} CUDA kernel whose performance is measured in section~\ref{sec:benchmarks}.

\begin{codesize}
\begin{verbatim}
runs1 k
  = writeFile "tsort1.cu" $ CUDA.genKernel "tsort1" 
      (compose (tsort1 k)) (namedArray "inp" (2^k))
\end{verbatim}
\end{codesize}

The generated code
% for $8$ inputs is shown in Appendix A. It 
uses
one thread per array element (just as the {\tt bitonicMerge}
kernel shown above did). The next step is to move to using Push as
well as Pull arays, so as to be able to generate more efficient code
from essentially the same sorter construction.

\subsection{New combinator implementations using Push arrays}

%% ** TODO make code for ilv2 more comprehensible :)

It is in combining the results of the {\tt f} and {\tt g} functions
that we run into difficulty using just Pull arrays.
Earlier, we saw how to use Push arrays to implement {\tt concP},
which concatenates two arrays.
Here, we use exactly the same approach to make a new version of
the {\em interleave} combinator. The results of applying
the {\tt f}s and {\tt gs} are combined into a Push array, in the right
order.
\pagebreak
%% ** TODO More explanation needed.
\begin{codesize}
\begin{verbatim}
ixMap :: (UWordE -> UWordE) -> ArrayP a -> ArrayP a 
ixMap f (ArrayP p n) = ArrayP (ixMap' f p) n

ixMap' :: (UWordE -> UWordE) 
         -> P (UWordE, a)
         -> P (UWordE, a) 
ixMap' f p = \g -> p (\(i,a) -> g (f i,a))

insertZero :: Int -> UWordE -> UWordE
insertZero 0 a = a `shiftL` 1
insertZero i a 
  = a + (a .&. fromIntegral (complement (oneBits i :: Word32)))

ilv2 :: Choice b => 
        Int -> (a -> a -> b) -> (a -> a -> b) -> 
        Array a -> ArrayP b
ilv2 i f g (Array ixf n) 
   = ArrayP (\k -> app a5 k *>* app a6 k) n
  where
    n2 = n `div` 2
    a1 = Array (ixf . left) (n-n2)
    a2 = Array (ixf . right) n2
    a3 = zipWith f a1 a2
    a4 = zipWith g a1 a2
    a5 = ixMap left (push a3)
    a6 = ixMap right (push a4)
    left = insertZero i
    right = flipBit i  . left
    app (ArrayP f _) a = f a
\end{verbatim}
\end{codesize}
\noindent
This new combinator can now replace {\tt ilv1} in the bitonic
merger, giving a kernel that runs considerably faster. We will use
that kernel to build a large sorter later.

The implemenation of {\tt vee2} is almost identical to
that of {\tt ilv2}, with {\tt flipBit i}
replaced by {\tt flipLSBsTo} as before (so that, again, one would in
fact make a more generic function for building such combinators).
Now, we just need to replace the {\tt ilv1} and {\tt vee1} combinators
in the tree sorter with {\tt ilv2} and {\tt vee2} , to get a verrsion that uses half as many threads:
\begin{codesize}
\begin{verbatim}
tmerge2 :: Int -> [Array IntE -> ArrayP IntE]
tmerge2 n = vstage (n-1) : [ istage (n-i) | i <- [2..n]]
  where
    vstage i = vee2 i min max
    istage i = ilv2 i min max

tsort2 :: Int -> [Array IntE -> ArrayP IntE]
tsort2 n = concat [tmerge2 i | i <- [1..n]]
\end{verbatim}
\end{codesize}
As we shall see in section~\ref{sec:benchmarks}, the resulting code
is significantly faster. This is because it uses one thread per two array elements,
and the code no longer contains any conditionals.

In order to go faster still, we resort to building a different sorting network, of
exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\cite{PeriodicBalanced}.
This involves the introduction of one new combinator that can be seen as a mixture
of the {\tt ilv} and {\tt vee} combinators already introduced.

\subsection{A sorter built from the balanced periodic merger}
First, we note that the balanced periodic merger contains multiple
uses of the now familiar vee-shaped pattern, see Figure~\ref{fig:periodicMerger}.
\begin{figure}
\centering
\includegraphics[scale=0.25]{balanced}
\caption{16 input periodic balanced merging network}
\label{fig:periodicMerger}
\end{figure}

\begin{codesize}
\begin{verbatim}
bpmerge2 :: Int -> [Array IntE -> ArrayP IntE]
bpmerge2 n = [vstage (n-i) | i <- [1..n]]
  where vstage i = vee2 i min max
\end{verbatim}
\end{codesize}

Now Dowd et al proved that the balanced periodic merger sorts two {\em interleaved} sorted sequences. So, taking an iterative view of the resulting sorter,
we want to build a tree of mergers as before, but the smaller mergers should
be interleaved, rather than operating on adjacent sub-sequences.
There should be one merger on the right hand end of the network; left of
that, there should be two mergers that operate on the odd and even elements,
and each of them should in turn be fed by two interleaved mergers, and so.
The sorter is illustrated, for $16$ inputs in Figure~\ref{fig:vsorter}.
Just to the left of the final balanced merger, one of the two interleaved
mergers is shown using dotted lines. It operates on a completely different set of inputs from the other 8-input merger in the same part of the tree.

\begin{figure*}
\centering
\includegraphics[scale=0.5]{vsorter}
\caption{A sorter based on the idea that the periodic
balanced merger network sorts two interleaved sorted sequences.
It consists of two half-sized sorters, one working on the
odd elements of the input and one on the even, followed by
the balanced merger. The diagram indicates using dotted lines the
balanced merger that is the final (rightmost) part of
one of the half-size sorters.}
\label{fig:vsorter}
\end{figure*}

The most straightforward way to give an iterative description of
this sorter is
to introduce a new combinator that is a combination of
{\tt ilv} and {\tt vee}.
The only thing that we need to change is the {\em partner} function.
This time we will flip not the least significant bits from position $0$ to
position $i$ but from position $i$ to $i+j$.

\begin{codesize}
\begin{verbatim}
-- flip bits from position i to position i+j inclusive
flipBitsFrom :: Bits a => Int -> Int -> a -> a
flipBitsFrom i j a = a `xor` (fromIntegral mask)
  where
    mask = (oneBits (j + 1):: Word32) `shiftL` i

ilvVee1 :: Choice a => 
           Int -> Int -> 
           (b -> b-> a) -> (b -> b -> a) -> 
           Array b -> Array a
ilvVee1 i j f g arr = Array ixf (len arr)
  where
    ixf ix = let l = arr ! ix
                 r = arr ! newix
                 newix = flipBitsFrom i j ix
             in (ifThenElse (lowBit (i+j) ix) (f l r) (g l r))
\end{verbatim}
\end{codesize}

\begin{codesize}
\begin{verbatim}
ilvVee2 :: Choice b => Int -> Int -> 
           (a -> a -> b) -> (a -> a -> b) -> 
           Array a -> ArrayP b
ilvVee2 i j f g (Array ixf n) 
  = ArrayP (\k -> app a5 k *>* app a6 k) n
  where
    n2 = n `div` 2
    a1 = Array (ixf . left) (n-n2)
    a2 = Array (ixf . right) n2
    a3 = zipWith f a1 a2
    a4 = zipWith g a1 a2
    a5 = ixMap left (push a3)
    a6 = ixMap right (push a4)
    left = insertZero (i+j)
    right = flipBitsFrom i j . left
    app (ArrayP f _) a = f a
\end{verbatim}
\end{codesize}

For both variants of the combinator, we simply add to the {\tt ilv} definitions
a new {\tt Int} parameter, {\tt j}, and replace {\tt flipBit i} by {\tt flipBitsFrom i j}. We also insert the zero bit (when calculating the left index)
at position {\tt i + j} rather than just at position {\tt i}.
{\tt ilvVee} is a generalisation of both {\tt ilv} and {\tt vee}.
{\tt ilvVee i 0} has the same behaviour as {\tt ilv i}, and {\tt ilvVee 0 (j-1)}
is the same as {\tt vee j}. The {\tt i} parameter controls the degree
of interleaving, and the {\tt j} parameter controls the size of
the vee-shaped blocks.

For $16$ inputs, the parameters
to {\tt ilvVee} that describe the periodic merger on the right of the construction
are $i=0$ (for no interleaving) paired with $3$, $2$, $1$ and $0$
for the decreasing size of the vee-shaped blocks.
Next, to the left, the mergers are interleaved ($i=1$) and there are three
stages with
vee-shaped blocks of decreasing size ($j = 2, 1, 0$), see Figure~\ref{fig:vsorter}.
The following code gives an iterative description of the construction
for {\small $2^n$} inputs:

\begin{codesize}
\begin{verbatim}
vsort :: Int -> Array IntE -> Kernel (Array IntE)
vsort n = composeS . map pure $ [istage (n-i) (i-j) 
                                | i <- [1..n], j <- [1..i]]
  where istage i j = ilvVee2 i j min max
\end{verbatim}
\end{codesize}

\noindent
The resulting generated code uses one thread per
$2$ indices.

\subsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
\begin{figure}
\begin{codesize}
\begin{verbatim}
  unsigned int arrayLength = 1 << LOG_L_SIZE;
  unsigned int diff = LOG_L_SIZE - LOG_S_SIZE;
  unsigned int blocks = arrayLength / S_SIZE;
  unsigned int threads = S_SIZE / 2;

  sortSmall<<<blocks, threads,4096>>>(din,din);

  for(int i = 0 ; i < diff ; i += 1){ 
    vSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<i)*S_SIZE);

    for(int j = i-1; j >= 0; j -= 1)
      iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);

    bmergeSmall<<<blocks,threads,4096>>>(din,din);}
\end{verbatim}
\end{codesize}
\caption{CUDA code for our large sorter. {\tt sortSmall} and {\tt bmergeSmall}
are replaced by Obsidian-generated kernels in the experiments. {\tt vSwap} and {\tt iSwap} are handwritten CUDA kernels
that perform one column of compare and swap operations in the vee and ilv shapes respectively, and are
parameterised on the stride. (Our generated kernels have fixed input and output size.)}
\label{fig:CUDAsort}
\end{figure}
We have measured the performance of generated 512-input sorting and merging kernels by plugging them into a larger sorter written in CUDA.
The sorter has exactly the structure shown in Figure~\ref{fig:mixedsorter}.
Figure~\ref{fig:newmixedsorter} shows the location of smaller sorters and mergers, and of vSwap and iSwap kernels for $16$ inputs.
Larger sorters simply have more columns of mergers, each preceded by a vSwap and a number of iSwaps.
The overall structure of the resulting CUDA code is shown in Figure~\ref{fig:CUDAsort}.
Because {\tt iSwap} is used repeatedly, we also wrote kernels
corresponding to compositions of two and three of them (avoiding
memory accesses between the columns).
That is, we replaced the loop containing {\tt iSwap} above by
\pagebreak
\begin{codesize}
\begin{verbatim}
for(int j = i-1; j >= 0; j -= 3){ 
  if (j==0) 
    iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
    else 
    {if (j==1) 
       iSwap2<<<blocks/4,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
       else 
       iSwap3<<<blocks/8,threads*2,0>>>(din,din,(1<<j)*S_SIZE);}}
\end{verbatim}
\end{codesize}
\noindent
The Obsidian implementation and all code for examples in the paper are available at \url{http://www.cse.chalmers.se/~joels/expressive.html}.

\begin{figure}
\centering
\includegraphics[scale=0.33]{newmixedSorter}
\caption{
This diagram shows the tree-shaped sorting network that we saw earlier, but with 4 input sorting and merging kernels indicated. This is the structure of the network that we have used to implement large sorters, in which the small kernels having {\small $2^9 = 512$} inputs in all cases. For {\small $2^{24}$} inputs overall, the resulting sorter
has one column of small sorters and $24-9=15$ of small mergers.}
%% both ofs in the above sentence should remain :)
\label{fig:newmixedsorter}
\end{figure}

Table~\ref{fig:table} shows the performance figures for the large sorter
for $5$ different $512-$input small sorter kernels.
{\tt tsort1} is defined above, and is a variant of bitonic sort
that does not require {\tt if} statements to determine the direction
of sorting of pairs, as all comparison operations place the minimum at the
same index as the lower input.
{\tt tsort2} is the same construction, but built with the combinators {\tt ilv2}
and {\tt vee2} that result in the use of one thread per two indices.
{\tt vsort}, the fastest kernel, uses a generalisation of those
two combinators, and again uses one thread per two indices.
{\tt vsort1} is the same construction as {\tt vsort}, but with {\tt ilvVee2} replaced
by {\tt ilvVee1}, and so using one thread per index.
As a reference small kernel implementation, we include a simple hand-coded bitonic sort
that uses one thread per index (see code in Appendix).
None of the kernels has been subject to optimisations related to warp size.

\input{table1.tex}

We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:table2}.
The {\tt tsort1} and {\tt vsort1} kernels, which are built using only pull arrays, 
(and which have one index per thread in the generated code) are
noticeably slower than those that have two indices per thread.
The generation of the latter kernels is made possible by the introduction of 
push arrays.

\input{table2.tex}

Although none of our generated kernels is optimised (for example with respect to warp size), 
their performance is, nevertheless, very good. We are working on automated warp size-related 
optimisations. It would also be interesting to explore the fusion of adjacent columns of 
comparators in the small kernels; omitting a {\tt sync} would cause this fusion to happen
but we also need to modify the threading behaviour (doubling the number of indices per thread).

%The final version of the paper will contain benchmarking against a highly optimised sorter kernel obtained from our colleagues in the computer graphics group at Chalmers. For now, we emphasise that the {\tt vsort} kernel is so fast that we expect it to compare favourably. Given the simplicity of its description in Obsidian, we feel that this is a good way to write fast kernels.

%\input{discussion}

\section {Discussion}

%% ** TODO. Mention CSE. warp-related opt.

Push arrays form a new approach to array representation in DSELs.
We do not know of similar approaches in the literature, despite
the fact that the notions of demand and data flow
may feel familiar to the reader who considers Pull and Push arrays.
The addition of Push arrays to Obsidian seems highly beneficial. With 
this new feature, the user gains finer grained control over the code generated
and the resulting CUDA kernels perform considerably better than before.
This 
was illustrated in the series of sorters explored in section~\ref{sec:MARY}.
The performance of {\tt vsort} is sufficiently good that it
can be used as a first phase in a larger sorter (written in CUDA) that
can sort 16M elements in 96 ms, while an i7-920 CPU takes
around 2740 ms. Further speed improvements look possible, both in the coordination code
and in the kernels. An obvious next step would be to investigate the generation of the {\tt iSwap} and
{\tt vSwap} kernels from Obsidian. (This is not currently possible because of assumptions that we made about the interfaces to kernels and about how {\em thread ids} are used. We will look into ways to relax
our assumptions.)

The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
description language~\cite{LavaSorter} was that a relatively small set of combinators went a long way. So, although we introduced three
combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
We expect to provide the user with a well-documented set of combinators,
so that users can get access to this style of programming without having
to develop their own combinators, and without having to think too much
about bit-hacking.
The bit-manipulation approach chosen to define our combinators automatically
created functions that apply to sub-sequences of the input that
are of an appropriate length.

In this paper, we made combinators for the special case of two input, two
output operations (built from two two-input funtions that we typically called
{\tt f} and {\tt g}). 
This approach should be generalised to deal with blocks that have $2^k$ inputs
and outputs.
Also, we made a compound combinator from {\tt ilv} and {\tt vee},
but generalising to more than two input components would allow
for composing combinators, and indeed for recursive descriptions
that could be unrolled. Then, ignoring {\tt syncs}, a recursive description of
{\tt vsort} could be something like 
\begin{codesize}
\begin{verbatim}
vsortR 0 = id
vsortR n = bpmergeR n . ilv2 1 (vsortR (n-1))
\end{verbatim}
\end{codesize}
\noindent
It would then be necessary to optimise the code generated from
multiple applications of {\tt ilv2 1}, for example, whereas here
we have forced the user to figure out both the unrolling and the combinations.
Moving to more general combinators would also give the opportunity to
provide predefined combinators that capture more of the commonly used threading patterns (for instance $k$ indices per thread rather than the $1$ and $2$ shown here).

The integration of Push arrays into Obsidian raises some new questions.
Previously, there was a direct correspondence between the length 
of an array and the number of threads used to compute it, which
allowed the user to write an initial program without worrying about
threads at all, and then to tweak the Obsidian program if he was not satisfied
with the threading behaviour of the resulting kernel. 
Now, as can be seen in the {\tt catArrayPs} example and in the sorters, this correspondence
can be broken using Push arrays. The {\tt catArrayPs} example and two
of the sorters use half as many threads as the 
number of elements. For users who are very concerned about the speed of
the generated kernels, getting this control through using Push arrays in
a particular pattern is clearly a good thing. But adding a second, different
way to control thread use in the generated code certainly complicates matters, and further case studies are needed to confirm that the complication pays off.

The addition of Push arrays also adds the possibility to include potentially 
unsafe operations in Obsidian, for example by writing multiple array elements to the same index, or by discarding elements.
This new expressiveness will have to be carefully controlled. On the positive side, it offers the possibility to encode functions like {\tt filter} from Haskell
that
are simply not expressible using only Pull arrays. Being able to implement {\tt filter} would make programming kernels in obsidian feel much more like programming in Haskell -- a welcome loosening of the strait-jacket.
Once that is done, it will be time to develop a very simple coordination language to allow programming of entire GPU applications that make use of
the kind of small kernel building blocks developed here.

\appendix
\section{Appendix}
\begin{codesize}
\begin{verbatim}
__device__ inline void swap(int & a, int & b)
{
   int tmp = a;
   a = b;
   b = tmp;
}

__global__ static void bitonicSort(int * values, int *results)
{
   extern __shared__ int shared[];

   const unsigned int tid = threadIdx.x;
   const unsigned int bid = blockIdx.x;

   // Copy input to shared mem.
   shared[tid] = values[(bid*NUM) + tid];

   __syncthreads();

   // Parallel bitonic sort.
   for (unsigned int k = 2; k <= NUM; k *= 2)
   { 
     // bitonic merge
     for (unsigned int j = k / 2; j>0; j /= 2)
       {
           unsigned int ixj = tid ^ j;

           if (ixj > tid)
           {
               if ((tid & k) == 0)
               {
                   if (shared[tid] > shared[ixj])
                   {
                       swap(shared[tid], shared[ixj]);
                   }
               }
               else
               {
                   if (shared[tid] < shared[ixj])
                   {
                       swap(shared[tid], shared[ixj]);
                   }
               }
           }
           __syncthreads();
       }
   }

   // Write result.
   results[(bid*NUM) + tid] = shared[tid];
}
\end{verbatim}
\end{codesize}

\acks
This research has been funded by the Swedish Foundation for Strategic Research (which funds
the RAW FP Project) and
by the Swedish Research Council.

% We recommend abbrvnat bibliography style.

%\bibliographystyle{abbrvnat}
%\bibliography{paper}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Axelsson et~al.(2011)Axelsson, Claessen, Sheeran, Svenningsson,
  Engdal, and Persson]{FELDSPAR}
E.~Axelsson, K.~Claessen, M.~Sheeran, J.~Svenningsson, D.~Engdal, and
  A.~Persson.
\newblock {T}he {D}esign and {I}mplementation of {F}eldspar an {E}mbedded
  {L}anguage for {D}igital {S}ignal {P}rocessing.
\newblock In \emph{Proceedings of the 22nd international conference on
  Implementation and application of functional languages}, IFL'10, pages
  121--136, Berlin, Heidelberg, 2011. Springer-Verlag.

\bibitem[Batcher(1968)]{Batcher}
K.~E. Batcher.
\newblock Sorting {N}etworks and {T}heir {A}pplications.
\newblock In \emph{Proc. AFIPS Spring Joint Computer Conference 32}, pages
  307--314, 1968.

\bibitem[Billeter et~al.(2009)Billeter, Olsson, and Assarsson]{OLAMARCUS}
M.~Billeter, O.~Olsson, and U.~Assarsson.
\newblock Efficient stream compaction on wide {SIMD} many-core architectures.
\newblock In \emph{Proc. Conf. on High Performance Graphics}, HPG '09, pages
  159--166. ACM, 2009.

\bibitem[Chakravarty et~al.(2011)Chakravarty, Keller, Lee, McDonell, and
  Grover]{ACCELERATE}
M.~M. Chakravarty, G.~Keller, S.~Lee, T.~L. McDonell, and V.~Grover.
\newblock Accelerating {H}askell array codes with multicore {GPU}s.
\newblock In \emph{Proc. sixth workshop on {D}eclarative {A}spects of
  {M}ulticore {P}rogramming}, DAMP '11, pages 3--14. ACM, 2011.

\bibitem[Claessen(1999)]{POORKOEN}
K.~Claessen.
\newblock A poor man's concurrency monad.
\newblock \emph{J. Funct. Program.}, 9\penalty0 (3):\penalty0 313--323, 1999.

\bibitem[Claessen et~al.(2001)Claessen, Sheeran, and Singh]{LavaSorter}
K.~Claessen, M.~Sheeran, and S.~Singh.
\newblock The {D}esign and {V}erification of a {S}orter {C}ore.
\newblock In \emph{Proc. Int. Conf. on Correct Hardware Design and Verification
  Methods (CHARME)}, volume 2144 of \emph{Springer LNCS}, pages 355--369, 2001.

\bibitem[Dowd et~al.(1989)Dowd, Perl, and Larry~Rudolph]{PeriodicBalanced}
M.~Dowd, Y.~Perl, and M.~S. Larry~Rudolph.
\newblock The {P}eriodic {B}alanced {S}orting {N}etwork.
\newblock \emph{Journal of the ACM}, 36:\penalty0 738--757, 1989.

\bibitem[Elliott(2003)]{PAN}
C.~Elliott.
\newblock Functional {I}mages.
\newblock In \emph{The Fun of Programming}, ``Cornerstones of Computing''
  series. Palgrave, Mar. 2003.

\bibitem[Elliott et~al.(2003)Elliott, Finne, and de~Moor]{COMPILEEDSL}
C.~Elliott, S.~Finne, and O.~de~Moor.
\newblock Compiling embedded languages.
\newblock \emph{Journal of Functional Programming}, 13\penalty0 (2), 2003.

\bibitem[{J}oel {S}vensson(2011)]{JSLIC}
{J}oel {S}vensson.
\newblock {O}bsidian: {GPU} {K}ernel {P}rogramming in {H}askell.
\newblock Technical Report 77L, {C}omputer {S}cience and {E}nginering,
  {C}halmers {U}niversity of {T}echnology, {G}othenburg, 2011.
\newblock Thesis for the degree of Licentiate of Philosophy.

\bibitem[Keller et~al.(2010)Keller, Chakravarty, Leshchinskiy, Peyton~Jones,
  and Lippmeier]{REPA}
G.~Keller, M.~M. Chakravarty, R.~Leshchinskiy, S.~Peyton~Jones, and
  B.~Lippmeier.
\newblock Regular, shape-polymorphic, parallel arrays in {H}askell.
\newblock In \emph{Proceedings of the 15th ACM SIGPLAN international conference
  on Functional programming}, ICFP '10, pages 261--272, New York, NY, USA,
  2010. ACM.

\bibitem[Larsen(2011)]{BARRACUDA}
B.~Larsen.
\newblock Simple optimizations for an applicative array language for graphics
  processors.
\newblock In \emph{Proc. sixth workshop on {D}eclarative {A}spects of
  {M}ulticore {P}rogramming}, DAMP '11, pages 25--34. ACM, 2011.

\bibitem[Mainland and Morrisett(2010)]{NIKOLA}
G.~Mainland and G.~Morrisett.
\newblock Nikola: embedding compiled {GPU} functions in {H}askell.
\newblock In \emph{Proc. third ACM Symposium on Haskell}. ACM, 2010.

\bibitem[Sheeran(1989)]{sortsRuby}
M.~Sheeran.
\newblock Describing {B}utterfly {N}etworks in {R}uby.
\newblock In \emph{Functional Programming}, pages 182--205. Springer Workshops
  in Computing, 1989.

%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

### 2013-09-30 src2/expressive/push.tex
another approach

--- 2013-09-30 src2/expressive/paperFinal.tex
@@
-%-----------------------------------------------------------------------------
-%
-%               Template for sigplanconf LaTeX Class
-%
-% Name:         sigplanconf-template.tex
-%
-% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
-%               file for SIGPLAN conference proceedings.
-%
-% Author:       Paul C. Anagnostopoulos
-%               Windfall Software
-%               978 371-2316
-%               paul@windfall.com
-%
-% Created:      15 February 2005
-%
-%-----------------------------------------------------------------------------
-
-\documentclass[]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{url}
-
-\newcommand{\codesize}{\fontsize{7}{8}\selectfont}
-
-\begin{document}
-\conferenceinfo{DAMP'12,} {January 28, 2012, Philadelphia, PA, USA.}
-\CopyrightYear{2012}
-\copyrightdata{978-1-4503-1117-5/12/01}
-
-%\conferenceinfo{WXYZ '05}{date, City.} 
-%\copyrightyear{2005} 
-%\copyrightdata{[to be supplied]}
-
-\titlebanner{}        % These are ignored unless
-\preprintfooter{}   % 'preprint' option specified.
-
-%\title{Towards an Implementation of Push Arrays in Obsidian}
-\title{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
-%\subtitle{Subtitle Text, if any}
-
-%\authorinfo{Name1}
-%           {Affiliation1}
-%           {Email1}
-\authorinfo{Koen Claessen \and Mary Sheeran \and Bo Joel Svensson}
-           {Chalmers University of Technology, 
-             Department of Computer Science and Engineering,
-             Gothenburg, Sweden}
-           {koen@chalmers.se/ms@chalmers.se/joels@chalmers.se}
-
-\maketitle
-
-% ABSTRACT !!! -----------------------------------------------------------------
-\begin{abstract}
-%This is the text of the abstract.
-
-Graphics Processing Units (GPUs) are powerful computing devices 
-that with the advent of CUDA/OpenCL are becomming useful for general 
-purpose computations. 
-Obsidian is an embedded domain specific language that generates CUDA kernels
-from functional descriptions.
-A symbolic array construction allows us to guarantee that intermediate
-arrays are fused away. However, the current array construction has
-some drawbacks; in particular, arrays cannot be combined efficiently.
-We add a new type of {\em push arrays} to the existing Obsidian system in 
-order to solve this problem. The two array types complement each other,
-and enable the definition of combinators that both
-take apart and combine arrays, and that result in efficient generated code.
-This extension to Obsidian is demonstrated on a sequence of
-sorting kernels, with good results.
-The case study also illustrates the use of combinators for expressing
-the structure of parallel algorithms.
-The work presented is preliminary, and 
-the combinators presented must be generalised.
-However, the raw speed of the generated kernels bodes well.
-\end{abstract}
-
-%\category{CR-number}{subcategory}{third-level}
-\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
-\category{D.3.4}{Programming Languages}{Processors}[Code generation]
-
-\terms
-Languages, Algorithms, Performance
-
-\keywords
-Arrays, Data parallelism, Embedded Domain Specific Language, General Purpose GPU programming, Haskell
-
-%\input{intro}
-\section{Introduction}
-
-%\input{table}
-
-%----------------------------------------------------------------------------
-
-Graphics Processing Units (GPUs) are parallel computers with hundreds 
-to thousands of processing elements. The CUDA and OpenCL languages  
-make available the power of the GPU to programmers interested in general 
-purpose computations. In CUDA and OpenCL, the programmer writes {\em kernels}, 
-Single Program Multiple Data (SPMD) programs that are executed by groups 
-of threads on the available processing elements of the GPU.
-
-CUDA and OpenCL are general purpose programming languages, mirroring the 
-increased capabilities of a modern GPU to target that domain. However, these 
-languages lack compositionality. 
-Also, being based in C/C++ means that the core idea in a program may
-not be easily visible.
-
-\subsection{Embedded DSLs for GPGPU programming} 
-We are aiming for a GPU programming language that is more
-concise than mainstream
-languages such as CUDA and OpenCL. 
-Obsidian is a
-domain specific embedded language (DSEL) implemented in Haskell.
-When an Obsidian program is run, a representation of the program is created 
-as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
-The program representation generated when running an Obsidian program 
-is compiled into CUDA code. We are also working on an OpenCL backend.
-
-Our approach is different from that of other Haskell DSELs targetting
-GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
-We do not try to abstract away from 
-all the peculiarities of GPU programming, but rather provide a higher 
-level language in which to experiment with them.
-For instance,
-Accelerate provides a standard set of basic operations such as 
-{\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
-with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
-Obsidian, on the other hand, allows the user to experiment with the
-{\em generation} of small kernels for fixed size array inputs
-from higher level descriptions.
-It is intended to allow the user to play with the kinds of tradeoffs that are important
-when writing such high performance building blocks; in this paper, the main consideration
-is the number of array elements of the input and output that are manipulated
-by a single thread in the generated CUDA code.
-An important aspect of Obsidian is the symbolic array representation used, along
-with its associated {\tt sync} operation. As we shall see, the {\tt sync} operation
-allows the programmer to guide code generation and control parallelism and thread
-use~\cite{JSLIC}.
-
-%Today there are number of EDSLs written in Haskell targeting 
-%GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
-%another by having a slightly different approach. We do not try to abstract away from 
-%all the peculiarities of GPU programming but rather giving a higher 
-%level language in which to experiment with them.
-%trying to appeal to a slightly different target audience. 
-%We try to cater not for the advanced functional programmer who wants to 
-%run his programs on GPUs, but rather to offer a higher level 
-%language to the CUDA/OpenCL hacker. 
-%In this language we want the user to be 
-%able to quickly experiment with the kind of tradeoffs that are important 
-%when writing a high performance GPU kernel.
-
-%Unlike the other approaches that are complementary to our Haskell GPU EDSL, 
-%Obsidian does not provide the usual set of basic operations such as 
-%{\tt Map}, {\tt Reduce} and {\tt ZipWith} as built in skeletons. We 
-%want a language where the programmer can experiment with implementing 
-%his or her own versions of these basic operations. So far, we make
-%this possible in Obsidian through a {\tt Sync} operation and the Array 
-%representation we use. The {\tt Sync} operation, using target domain 
-%nomenclature, is the programmers tool to guide code generation and 
-%obtain parallelism. The presentation of Obsidian here will be somewhat 
-%cursory, for a more complete view of the various versions and approaches 
-%we have tried, see~\cite{JSLIC}.
-%More details on this later.
-
-In Obsidian, a kernel that sums an array can be expressed as: 
-\begin{codesize}
-\begin{verbatim}
-sum :: Array IntE -> Kernel (Array IntE) 
-sum arr | len arr == 1 = return arr
-        | otherwise    = (pure (fmap (uncurry (+)) . pair) 
-                          ->- sync 
-                          ->- sum) arr                       
-\end{verbatim}
-\end{codesize}
-The result of running this kernel on an eight element input array, 
-{\tt runKernel sum (namedArray ``input'' 8)}, is an intermediate representation 
-of the computation (shown in slightly pretty-printed form): 
-\begin{codesize}
-\begin{verbatim}
-arr0 = malloc(16)
-par i 4 {
-arr0[i] = ( +  input[( *  i 2 )] input[( +  ( *  i 2 ) 1 )] );
-}Sync
-arr1 = malloc(8)
-par i 2 {
-arr1[i] = ( +  arr0[( *  i 2 )] arr0[( +  ( *  i 2 ) 1 )] );
-}Sync
-arr2 = malloc(4)
-par i 1 {
-arr2[i] = ( +  arr1[( *  i 2 )] arr1[( +  ( *  i 2 ) 1 )] );
-}Sync
-\end{verbatim}
-\end{codesize}
-The named intermediate arrays in this representation are then laid 
-out in GPU shared memory and CUDA code can be generated (here for arrays of length eight)\footnote{An alignment qualifier for shared memory 
-has been omitted to save space in the listings showing generated code}: 
-%\pagebreak
-\begin{codesize}
-\begin{verbatim}
-__global__ void sum(int *input0,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-  extern __shared__ unsigned char sbase[];
-  (( int *)sbase)[tid] = 
-    (input0[((bid*8)+(tid*2))]+
-     input0[((bid*8)+((tid*2)+1))]);
-  __syncthreads();
-  if (tid<2){
-    (( int *)(sbase + 16))[tid] = 
-      ((( int *)sbase)[(tid*2)]+
-       (( int *)sbase)[((tid*2)+1)]);
-  }
-  __syncthreads();
-  if (tid<1){
-    (( int *)sbase)[tid] = 
-      ((( int *)(sbase+16))[(tid*2)]+
-       (( int *)(sbase+16))[((tid*2)+1)]);
-  }
-  __syncthreads();
-  if (tid<1){
-    result0[(bid+tid)] = (( int *)sbase)[tid];
-  }
-}
-\end{verbatim}
-\end{codesize}
-%The CUDA code above is generated by issuing the command:
-%\begin{codesize}
-%\begin{verbatim}
-%genKernel ``sum'' sum (namedArray undefined 8 :: Array IntE)
-%\end{verbatim}
-%\end{codesize}
-%\noindent
-%and the kernel code is generated for arrays of length eight.
-
-%----------------------------------------------------------------------------
-\subsection{Arrays in Obsidian}
-
-An array is represented by an indexing function and a length:
-\begin{codesize} 
-\begin{verbatim}
-data Array a = Array (UWordE -> a) Word32 
-\end{verbatim}
-\end{codesize}
-This array representation has served us well. It has these properties:
-\begin{itemize} 
-\item Fusion of operations is automatic.
-\item It naturally describes a data-parallel computation suitable for CUDA/OpenCL generation.
-\item Many basic operations can be implemented: {\tt map}, {\tt zipWith} etc.
-\end{itemize}
-Using this array representation in a DSEL is not new; the first occurence that we know of is in
-Pan~\cite{PAN}. Similar array representations have also later been 
-used in Feldspar~\cite{FELDSPAR}, and more recently also in the Repa library~\cite{REPA}.
-Functions for indexing and getting the length of arrays are as follows:
-\begin{codesize} 
-\begin{verbatim}
-(!) :: Array a -> UWordE -> a 
-(Array ixf _) ! ix = ixf ix
-
-len :: Array a -> Word32
-len (Array _ n) = n 
-\end{verbatim}
-\end{codesize}
-%Implementing a {\em Functor} instance for the {\tt Array} datatype is possible:
-A {\em Functor} instance for the {\tt Array} datatype is
-\begin{codesize} 
-\begin{verbatim}
-instance Functor Array where 
-  fmap f arr = Array (\ix -> f (arr ! ix)) (len arr) 
-\end{verbatim}
-\end{codesize}
-Now, composed applications of {\tt fmap} will be automatically fused. This is 
-illustrated in the example program below and the CUDA generated
-from it.  
-\begin{codesize} 
-\begin{verbatim}
-mapFusion :: Array IntE -> Kernel (Array IntE) 
-mapFusion = pure (fmap (+1) . fmap (*2)) 
-\end{verbatim}
-\end{codesize}
-
-\begin{codesize} 
-\begin{verbatim}
-__global__ void mapFusion(int *input0,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-
-  result0[((bid*32)+tid)] = ((input0[((bid*32)+tid)]*2)+1);
-}
-\end{verbatim}
-\end{codesize}
-Both of these code listings need explanation. In the Haskell 
-code, {\tt mapFusion} has type {\tt Array IntE -> Kernel (Array IntE)};
-{\tt Kernel} is a state monad that accumulates CUDA code as well as provides 
-new names for intermediate arrays. Neither of these features of the monad 
-is activated by this example though. The {\tt pure} function is defined 
-using the monad's {\tt return} as {\tt pure f a = return (f a)}. In this case, 
-it lifts a function of type {\tt Array IntE -> Array IntE} into 
-a kernel.
-
-The generated CUDA code computes the result array using a number of threads 
-equal to the length of that array. In this case, the kernel was generated
-to deal with arrays of length $32$. The important detail to notice in the 
-CUDA code is that there is no intermediate array created between the 
-{\tt (*2)} and the {\tt (+1)} operations.
-
-The {\tt mapFusion} example could just as well have been implemented using 
-the kernel sequential composition combinator, {\tt ->-}. 
-\begin{codesize} 
-\begin{verbatim}
-mapFusion :: Array IntE -> Kernel (Array IntE) 
-mapFusion = pure (fmap (*2)) ->- pure (fmap (+1)) 
-\end{verbatim}
-\end{codesize}
-Exactly the same CUDA code is then generated.
-
-In some cases, it is necessary to force computation of 
-intermediate arrays. This can be used to share partial computations between 
-threads and to expose parallelism. In Obsidian, the tool for this
-is called {\tt sync}, a built-in kernel. Using {\tt sync} as follows prevents fusion of the
-two operations:
-\begin{codesize} 
-\begin{verbatim}
-mapUnFused :: Array IntE -> Kernel (Array IntE) 
-mapUnFused = pure (fmap (*2)) ->- sync ->- pure (fmap (+1))
-\end{verbatim}
-\end{codesize}
-The generated CUDA code now stores an intermediate result in local shared
-memory before moving on. 
-\begin{codesize} 
-\begin{verbatim}
-__global__ void mapUnFused(int *input0,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-  extern __shared__ unsigned char sbase[];
-  (( int *)sbase)[tid] = (input0[((bid*32)+tid)]*2);
-  __syncthreads();
-  result0[((bid*32)+tid)] = ((( int *)sbase)[tid]+1);
-}
-\end{verbatim}
-\end{codesize}
-Intermediate arrays are laid out in the {\tt sbase} array in shared memory. 
-Since we may store arrays of many different types in the same locations 
-of the shared memory at different times during the execution, the type casts
-used in the code above are necessary.
-
-%----------------------------------------------------------------------------
-\subsection{Sync and parallelism}
-
-The {\tt sync} operation also enables the writing of parallel reduction 
-kernels. A reduction operation is an operation that takes an array as input
-and
-produces a singleton array as output.
-
-First, we define {\tt zipWith} and {\tt halve} on Obsidian arrays.
-\begin{codesize} 
-\begin{verbatim}
-zipWith :: (a -> b -> c) -> Array a -> Array b -> Array c
-zipWith op a1 a2 = Array (\ix -> (a1 ! ix) `op` (a2 ! ix)) 
-                   (min (len a1) (len a2))
-
-splitAt :: Word32 -> Array a -> (Array a, Array a) 
-splitAt n arr = 
-  (Array (\ix -> arr ! ix) n , 
-   Array (\ix -> arr ! (ix + fromIntegral n)) (len arr - n))
-
-halve arr = splitAt ((len arr) `div` 2) arr
-\end{verbatim}
-\end{codesize}
-A reduction kernel that takes an array whose length is a power of two 
-and gives an array of length one can be defined recursively. 
-Defining kernels recursively results in completely unrolled CUDA kernels,
-and kernel input size must be known at compile time.
-The approach to reduction taken here is to split the input array 
-into two halves and then apply {\tt zipWith} of the combining function to the 
-two halves, repeating the process until the length is one. 
-\begin{codesize} 
-\begin{verbatim}
-reduceS :: (a -> a -> a) -> Array a -> Kernel (Array a) 
-reduceS op arr | len arr == 1 = return arr
-               | otherwise    = 
-                 (pure ((uncurry (zipWith op)) . halve)
-                  ->- reduceS op) arr
-\end{verbatim}
-\end{codesize}
-%This Obsidian Kernel takes an array of length $2n$ as input and the output 
-Since the output of this kernel is of length one, and the number 
-of elements in the output array specifies the number of threads used to compute it, 
-this function, {\tt reduceS}, defines a sequential reduction. 
-The generated code for arrays of length eight is 
-\pagebreak
-%The code is generated with the 
-%command {\tt genKernel ``reduceSAdd'' (reduce (+)) (namedArray undefined 8 :: Array IntE)}, 
-%meaning that a CUDA reduce-with-addition kernel for arrays of length eight is generated. 
-\begin{codesize} 
-\begin{verbatim}
-__global__ void reduceSAdd(int *input0,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-
-  result0[(bid+tid)] = 
-    (((input0[((bid*8)+tid)]+
-       input0[((bid*8)+(tid+4))])+
-      (input0[((bid*8)+(tid+2))]+
-       input0[((bid*8)+((tid+2)+4))]))+
-     ((input0[((bid*8)+(tid+1))]+
-       input0[((bid*8)+((tid+1)+4))])+
-      (input0[((bid*8)+((tid+1)+2))]+
-       input0[((bid*8)+(((tid+1)+2)+4))])));
-}
-\end{verbatim}
-\end{codesize}
-Sequential reduction is not very interesting for GPU execution, but 
-the fix is simple. A well placed use of {\tt sync} indicates that 
-we want to compute, after each {\tt zipWith} phase, the intermediate 
-arrays using as many threads as that intermediate array is long. 
-The effect is shown in the code below. 
-\begin{codesize} 
-\begin{verbatim}
-reduce :: Syncable Array a 
-          => (a -> a -> a) -> Array a -> Kernel (Array a)
-reduce op arr | len arr == 1 = return arr
-              | otherwise    = 
-                (pure ((uncurry (zipWith op)) . halve)
-                 ->- sync 
-                 ->- reduce op) arr
-\end{verbatim}
-\end{codesize}
-The CUDA code for reduction with addition on eight
-elements is
-\begin{codesize} 
-\begin{verbatim}
-__global__ void reduceAdd(int *input0,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-  extern __shared__ unsigned char sbase[];
-  (( int *)sbase)[tid] = 
-    (input0[((bid*8)+tid)]+
-     input0[((bid*8)+(tid+4))]);
-  __syncthreads();
-  if (tid<2){
-    (( int *)(sbase + 16))[tid] = 
-      ((( int *)sbase)[tid]+
-       (( int *)sbase)[(tid+2)]);   
-  }
-  __syncthreads();
-  if (tid<1){
-    (( int *)sbase)[tid] = 
-      ((( int *)(sbase+16))[tid]+
-       (( int *)(sbase+16))[(tid+1)]);    
-  }
-  __syncthreads();
-  if (tid<1){
-    result0[(bid+tid)] = (( int *)sbase)[tid];  
-  }
-}
-\end{verbatim}
-\end{codesize}           
-In this generated CUDA, three phases can be identified. 
-The first uses four threads to compute a four element intermediate array;
-the second uses two threads, and so on. At the very end, a single thread
-copies the result from local shared memory to global memory.
-
-\subsection{Drawbacks of Obsidian Arrays}
-\label{sec:Drawbacks}
-
-%*** THE STUFF BELOW IS IN CONFLICT WITH WHAT I ADDED ABOVE!
-%When using Obsidian we have available to us a toolbox of higher order
-%functions that are familiar to functional programmers. One example of such a 
-%function is {\tt zipWith}.
-
-%\begin{codesize} 
-%\begin{verbatim}
-%example1 :: (Array IntE, Array IntE) 
-%            -> Kernel (Array IntE)
-%example1 (arr1,arr2) = return$ zipWith (+) arr1 arr2
-%\end{verbatim}
-%\end{codesize}
-
-%The body of this function looks very familiar to Haskell programmers. The 
-%type of the function however is a little bit different than what we are 
-%used to from everyday Haskell usage. In Obsidian we have a concept called 
-%a Kernel. Computations of types {\tt a -> Kernel b} can be turned into 
-%runnable GPU code by providing an input. 
-%*** VERY VAGUE AND ODD 
-%*** TODO: Explain kernel ?
-
-%The following code listing shows the CUDA code we can generate from 
-%the Obsidian program given above:
-
-% *** TODO: Needs a code size between tiny and small.
-
-%\begin{codesize}
-%\begin{verbatim}
-%__global__ void example1(int *input0,int *input1,int *result0){
-%  unsigned int tid = threadIdx.x;
-%  unsigned int bid = blockIdx.x;
-
-%  result0[tid] = (input0[((bid*32)+tid)]+
-%                  input1[((bid*32)+tid)]);
-
-%}
-%\end{verbatim}
-%\end{codesize}
-
-%The CUDA code above writes into a result array the pairwise sums from 
-%its two input arrays, as expected. 
-%*** There are many details here. tid, bid, 32. explain now or push to later?
-
-% \subsection{Obsidian arrays}
-
-%In Obsidian an array is represented by an indexing function and a length. 
-%\begin{codesize}
-%\begin{verbatim}
-%data Array a = Array (IxExp -> a) Word32
-%\end{verbatim}
-%\end{codesize}
-
-%And {\tt IxExp} is defined as {\tt UWordE}, that is an expression 
-%representing a unsigned 32bit value. The length of an array is represented
-%with a unsigned 32bit value as well but here it has been beneficial 
-%to use a built in Haskell type in order to be able to use this 
-%information at program generation time rather than program run time.
-
-%This array representation specifies how to compute each index of an Arrays. 
-%If we are interested in the value at location 5 in an array all we need 
-%to do is apply the indexing function to the literal expression 5 and 
-%we get a description of how to compute that value.
-
-%This array representation have been working fairly well in many situations. 
-%In {\tt example1} where we used {\tt zipWith} we got just about the resulting
-%CUDA code that a CUDA programmer would write himself. For some other examples 
-%the situation is much worse.
-
-The previous subsection described positive aspects of the array 
-representation that we have used so far. There are, however, 
-circumstances in which this Array representation is too restricted.
-
-Take the problem of concatenating two arrays. Using the array representation 
-described above, the only way to concatenate two arrays is to introduce 
-a conditional into the indexing function. If {\tt f} and {\tt g} are the 
-indexing functions of two arrays that are to be concatenated, and {\tt n1}
-is the length of the first array, the indexing 
-function of the result must be
-\begin{codesize}
-\begin{verbatim}
-new ix = if (ix < n1) 
-         then f ix 
-         else g (ix - n1)
-\end{verbatim}
-\end{codesize}
-The following program
-concatenates two arrays: 
-\begin{codesize}
-\begin{verbatim}
-catArrays :: (Array IntE, Array IntE) -> Kernel (Array IntE)
-catArrays = pure conc
-\end{verbatim}
-\end{codesize}
-%(arr1,arr2) = return$ conc (arr1,arr2)
-When it is used to generate a CUDA kernel that concatenates
-two arrays of length $16$, the following code is the result: 
-\begin{codesize}
-\begin{verbatim}
-__global__ void catArrays(int *input0,int *input1,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-
-  result0[((bid*32)+tid)] = 
-    (tid<16) ? input0[((bid*16)+tid)] : 
-               input1[((bid*16)+(tid-16))];
-}
-\end{verbatim}
-\end{codesize}
-%__global__ void catArrays(int *input0,
-%                          int *input1,
-%                          int *result0){
-%  unsigned int tid = threadIdx.x;
-%  unsigned int bid = blockIdx.x;
-%  
-%  result0[((bid*64)+tid)] = 
-%    (tid<32) ? input0[((bid*32)+tid)] : 
-%               input1[((bid*32)+(tid-32))];
-%  
-%}
-Now, conditionals like these are {\em bad} in code 
-to execute on a GPU, with its wide-SIMD data-parallel model. Separating
-the operation into two assignments and using half as many threads gives
-much higher performance.
-\begin{codesize}
-\begin{verbatim}
-__global__ void catArraysByHand(int *input0,
-                                int *input1,
-                                int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-
-  result0[((bid*32)+tid)] = input0[((bid*16)+tid)];
-  result0[((bid*32)+tid+16)] = input1[((bid*16)+tid)];  
-}
-\end{verbatim}
-\end{codesize}
-There are cases where code with conditionals is not that bad. 
-An expert on NVIDIA GPUs in particular may say that code with 
-the condition {\tt (tid < 32)} is fine, since 32 is the SIMD width 
-of those GPUs. However, any number that is not a multiple of 32 would 
-lead to poor performance, so in general this is a problem. 
- Worse still, {\em zipping} two arrays together and then 
-{\em unpairing} (to get an array of elements)
-leads to code that takes two different paths depending 
-on odd or even {\em thread id}. When a GPU executes such code, it 
-shuts down half of the threads and computes the two paths in sequence. 
-\begin{codesize}
-\begin{verbatim}
-zippUnpair :: (Array IntE, Array IntE) -> Kernel (Array IntE) 
-zippUnpair = pure (unpair . zipp)
-\end{verbatim}
-\end{codesize}
-The {\tt zipp} and {\tt unpair} operations are defined as follows: 
-\begin{codesize}
-\begin{verbatim}
-zipp :: (Array a, Array b) -> Array (a, b)             
-zipp (arr1,arr2) = 
-     Array (\ix -> (arr1 ! ix, arr2 ! ix)) 
-           (min (len arr1) (len arr2))
-
-unpair :: Choice a => Array (a,a) -> Array a
-unpair arr = 
-    let n = len arr
-    in  Array (\ix -> ifThenElse ((mod ix 2) ==* 0) 
-                      (fst (arr ! (ix `shiftR` 1)))
-                      (snd (arr ! (ix `shiftR` 1)))) (2*n)
-\end{verbatim}
-\end{codesize}
-Code generated from the {\tt zippUnpair} program exhibits really 
-poor performance; at any time half of the threads are shut down. 
-\begin{codesize}
-\begin{verbatim}
-__global__ void zippUnpair(int *input0,
-                           int *input1,
-                           int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-
-  result0[((bid*64)+tid)] = 
-    ((tid%2)==0) ? input0[((bid*32)+(tid>>1))] : 
-                   input1[((bid*32)+(tid>>1))];
-}
-\end{verbatim}
-\end{codesize}
-If we wrote this CUDA program by hand, we would, again, split it up into 
-two phases so that all threads can progress in parallel.
-
-%Even though this is a pretty artificial example, it illustrates the 
-%problem well. The situation is just as bad when using the function
-%{\tt unpair}. This function takes an array of pairs and returns an 
-%array of elements such that position 0 and 1 in the result array holds 
-%the values of the pair at index 0 in the input array. Using this operation 
-%results in a conditional statement that takes different execution paths 
-%on odd and even numbered threads.
-
-The arrays described so far, with an indexing function and a length, have 
-been nicknamed {\em Pull arrays} for how they describe how to 
-compute an element by {\em pulling} data from a number 
-of places. Using just Pull arrays, we have been unable to solve the 
-problems described so far in this section. The solution is to add a complementary array type to Obsidian.
-
-%For a more in depth view of the parts of Obsidian that has been showed 
-%in this section see~\cite{JSLIC}.
-
-%\input{push} 
 \section{Push Arrays}
@@
 %\subsection { Thoughts about Push arrays in Obsidian }
-
-%\input{applications}
-\section{Application}
-\label{sec:MARY}
-
-\subsection{Sorting on a GPU}
-In this section, we introduce combinators that express patterns
-of computation on whole arrays, and show their application to the
-development of fast sorting kernels.
-By kernel, we mean a computation that is performed by multiple
-threads, each performing the same computation, in a single block.
-The computation is performed entirely on the GPU, operating on
-a short array, which has been placed into shared memory.
-On the GPU on which we perform measurements, the maximum number
-of threads in a single block is 512. Thus, we will build and benchmark
-a sequence of
-kernels that sort 512 inputs.
-Our first kernels are implemented using one thread per array element.
-Next, we show how Push arrays allow us to move to having each
-thread operate on two array elements, giving a substantial performance
-improvement.
-
-Sorting kernels are typically used as building blocks in larger
-programs to sort much larger sequences of inputs. In section~\ref{sec:benchmarks},
-we show how to build a sorter for large arrays from small building blocks, including
-small kernels for sorting and merging that are generated from Obsidian.
-Our small kernels are constructed in the form of sorting and merging {\em networks},
-building on Batcher's bitonic merger~\cite{Batcher} and
-on the periodic balanced merger~\cite{PeriodicBalanced}.
-We chose also to implement the large sorter used  in benchmarking the
-small kernels as a sorting network. However, large
-sorters that are not themselves sorting networks (with typical examples being radix sort and quicksort) often call small sorting networks when they need to sort small arrays during their execution.
-Thus, small, fast sorting kernels have a variety of uses.
-
-\subsection{Describing Batcher's bitonic merger}
-
-The bitonic merger is typically presented as a recursive
-construction and we have earlier explored ways to describe
-and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LavaSorter}.
-Here, we consider iterative descriptions using similar combinators.
-
-Figure~\ref{fig:bitonicMerger} illustrates the merger for 16 inputs.
-Data flows from left to right.
-The vertical lines indicate components that operate on two array elements, placing the minimum onto the lower (abstract) wire, and the maximum onto the other output
-of the component.
-The leftmost {\em stage} operates (for $n=16$) on elements that are 
-$8$ apart. the next stage deals with elements that are $4$ apart, and so on.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.25]{bitonic}
-\caption{A diagram of a 16 input bitonic merging network, using
-a style that is standard in the literature. Note
-that in each stage containing 8 min/max or comparator components, all
-8 operate on independent parts of the input and so can proceed in parallel.
-}
-\label{fig:bitonicMerger}
-\end{figure}
-
-We introduce a combinator {\tt ilv1}, for {\em interleave}, that captures this pattern.
-{\tt ilv1 i f g} applies {\tt f} to elements {\small $2^i$} apart,
-producing the output at the lower of the two input indices; it applies {\tt g}
-to the same pairs of elements, producing the output on the upper index.
-Defining {\tt stage i} to be {\tt ilv1 i min max},
-the four stages in the diagram are simply
-{\tt stage} applied to{\small  $3$, $2$ $1$} and {\small $0$}.
-The definition of {\tt ilv1} makes use of the fact that
-flipping the bit $i$ of an index (using
-the function {\tt flipBit}) gives the index of the element
-that will be combined with it using the functions{\tt f} and {\tt g}.
-The decision about whether to apply {\tt f} or {\tt g} is made
-by looking at the value of bit $i$.
-As we shall see, the use of the Obsidian {\tt ifThenElse}
-produces conditionals in the resulting CUDA.
-\begin{codesize}
-\begin{verbatim}
-lowBit :: Int -> UWordE -> Exp Bool
-lowBit i ix = (ix .&. bit i) ==* 0
-
-flipBit :: Bits a => Int -> a -> a
-flipBit = flip complementBit
-
-ilv1 :: Choice a => 
-        Int -> (b -> b-> a) -> (b -> b -> a) -> 
-        Array b -> Array a
-ilv1 i f g arr = Array ixf (len arr)
-  where
-    ixf ix = let l = arr ! ix
-                 r = arr ! newix
-                 newix = flipBit i ix
-             in (ifThenElse (lowBit i ix) (f l r) (g l r))
-\end{verbatim}
-\end{codesize}
-\noindent
-Expressing {\tt ilv1} using bit-flipping may seem strange, but it has
-the advantage that it actually applies the desired pattern of
-computation repeatedly over larger input arrays.
-Now, for {\small $2^n$} inputs, a Haskell list containing the $n$ calls
-of this interleave combinator are built:
-\begin{codesize}
-\begin{verbatim}
-bmerge :: Int -> [Array IntE -> Array IntE]
-bmerge n = [istage (n-i) | i <- [1..n]]
-  where istage i = ilv1 i min max
-\end{verbatim}
-\end{codesize}
-\noindent
-Finally, the {\tt compose} function makes each element of the list
-into a kernel (using {\tt map pure}) and places a {\tt sync} between
-each kernel (using {\tt composeS}).
-\begin{codesize}
-\begin{verbatim}
-compose :: (Scalar a) => 
-           [Array (Exp a) -> Array (Exp a)] 
-           -> Array (Exp a) -> Kernel (Array (Exp a))
-compose = composeS . map pure
-
-runm k = putStrLn$ CUDA.genKernel "bitonicMerge" 
-         (compose (bmerge k)) (namedArray "inp" (2^k))
-\end{verbatim}
-\end{codesize}
-\noindent
-Note that {\tt bmerge k} works on inputs of length {\small }$2^{k+j}$, for $j > 0$, applying the merger to sub-sequences of length {\small $2^k$}.
-The CUDA code for {\tt bmerge 4} on $16$ inputs
-(with some newlines inserted) is
-\begin{codesize}
-\begin{verbatim}
-*Main> runm 4
-__global__ void bitonicMerge(int *input0,int *result0){
-  unsigned int tid = threadIdx.x;
-  unsigned int bid = blockIdx.x;
-  extern __shared__ unsigned char sbase[];
-  (( int *)sbase)[tid] 
-     = ((tid&8)==0) 
-      ? min(input0[((bid*16)+tid)],input0[((bid*16)+(tid^8))]) 
-      : max(input0[((bid*16)+tid)],input0[((bid*16)+(tid^8))]);
-  __syncthreads();
-  (( int *)(sbase + 64))[tid] 
-     = ((tid&4)==0) 
-      ? min((( int *)sbase)[tid],(( int *)sbase)[(tid^4)]) 
-      : max((( int *)sbase)[tid],(( int *)sbase)[(tid^4)]);
-  __syncthreads();
-  (( int *)sbase)[tid] 
-     = ((tid&2)==0) 
-      ? min((( int *)(sbase+64))[tid],(( int *)(sbase+64))[(tid^2)]) 
-      : max((( int *)(sbase+64))[tid],(( int *)(sbase+64))[(tid^2)]);
-  __syncthreads();
-  (( int *)(sbase + 64))[tid] 
-     = ((tid&1)==0) 
-      ? min((( int *)sbase)[tid],(( int *)sbase)[(tid^1)]) 
-      : max((( int *)sbase)[tid],(( int *)sbase)[(tid^1)]);
-  __syncthreads();
-  result0[((bid*16)+tid)] = (( int *)(sbase+64))[tid];
-\end{verbatim}
-\end{codesize}
-
-%% ** TODO Note that bmerge k not only works on inputs of length $2^{k+j}$,
-%% for $j > 0$, applying the merger to sub-sequences of length $2^k$.
-
-\subsection{Modifying the bitonic merger}
-
-The bitonic merger for which we have just generated a kernel is known
-to sort so-called bitonic sequences, which include sequences whose first
-half is sorted in one direction and whose second half is sorted in the other
-direction. This fact can be used to build the well-known bitonic sorting
-network. However, a GPU implementation typically needs to check, for each comparator,
-whether or not it should sort upwards or downwards, see for instance
-the simple CUDA implementation shown in Appendix A.
-We choose here to modify the merger so that it sorts two concatenated
-sequences that are sorted in the {\em same} direction.
-We do this by using a well-known trick, reversing half of the input to the merger. It turns out that
-we can also reverse the same half of the output of the first stage of
-the network, without affecting overall behaviour.
-The resulting network, {\tt tmerge}, shown in Figure~\ref{fig:mixedMerger},
-encourages us to develop a new combinator to describe the characteristic
-V-shaped pattern that results in the first stage.
-The combinator is modelled on {\tt ilv1}. The only difference
-is that the ``partner'' of an index is found not by flipping bit $i$, but
-by flipping bits $0$ to $i$, using function {\tt flipLSBsTo}. The implementation
-of {\tt vee1} is got from that for {\tt ilv1} by replacing the call of
-{\tt flipBit} by one of {\tt flipLSBsTo} (and we could also have chosen
-to make a more generic function that is parameterised on this {\em partner} function).
-\begin{codesize}
-\begin{verbatim}
-flipLSBsTo :: Int -> UWordE -> UWordE
-flipLSBsTo i = (`xor` (oneBits (i+1)))
-
-vee1 :: Choice a => 
-        Int -> (b -> b-> a) -> (b -> b -> a) -> 
-        Array b -> Array a
-vee1 i f g arr = Array ixf (len arr)
-  where
-    ixf ix = let l = arr ! ix
-                 r = arr ! newix
-                 newix = flipLSBsTo i ix
-             in (ifThenElse (lowBit i ix) (f l r) (g l r))
-\end{verbatim}
-\end{codesize}
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.25]{mixed}
-\caption{16 input merging network, {\em tmerge}. The first stage
-is made with a new combinator that we call vee, while
-the remaining stages are as in the bitonic merger.
-This network sorts an input that consists of two half-sized
-sorted sequences, giving the opportunity to build a tree-shaped
-sorting network.}
-\label{fig:mixedMerger}
-\end{figure}
-
-\begin{codesize}
-\begin{verbatim}
-tmerge :: Int -> [Array IntE -> Array IntE]
-tmerge n = vstage (n-1): [istage (n-i) | i <- [2..n]]
-  where
-    vstage i = vee1 i min max 
-    istage i = ilv1 i min max
-\end{verbatim}
-\end{codesize}
-
-Now that we have a merger that sorts sub-sequences containing two concatenated sorted sequences, it
-is easy to make a tree of them.
-The list of kernels to be composed now becomes
-\begin{codesize}
-\begin{verbatim}
-tsort1 :: Int -> [Array IntE -> Array IntE]
-tsort1 n = concat [tmerge i | i <- [1..n]]
-\end{verbatim}
-\end{codesize}
-\noindent
-The resulting sorting network is shown in Figure~\ref{fig:mixedsorter}.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.33]{mmixedsorter}
-\caption{A 16-input sorter made from a tree of {\em tmerge} mergers. 8 2-input mergers
-feed 4 4-input mergers, followed  by 2 8-input and one 16-input merger.
-A sorter on 8 inputs is shaded, as is a merger on 8 inputs, above it. 
-}
-\label{fig:mixedsorter}
-\end{figure}
-\noindent
-The following call writes the resulting CUDA to file {\tt tsort1.cu}
-and this is the first {\em generated} CUDA kernel whose performance is measured in section~\ref{sec:benchmarks}.
-
-\begin{codesize}
-\begin{verbatim}
-runs1 k
-  = writeFile "tsort1.cu" $ CUDA.genKernel "tsort1" 
-      (compose (tsort1 k)) (namedArray "inp" (2^k))
-\end{verbatim}
-\end{codesize}
-
-The generated code
-% for $8$ inputs is shown in Appendix A. It 
-uses
-one thread per array element (just as the {\tt bitonicMerge}
-kernel shown above did). The next step is to move to using Push as
-well as Pull arays, so as to be able to generate more efficient code
-from essentially the same sorter construction.
-
-\subsection{New combinator implementations using Push arrays}
-
-%% ** TODO make code for ilv2 more comprehensible :)
-
-It is in combining the results of the {\tt f} and {\tt g} functions
-that we run into difficulty using just Pull arrays.
-Earlier, we saw how to use Push arrays to implement {\tt concP},
-which concatenates two arrays.
-Here, we use exactly the same approach to make a new version of
-the {\em interleave} combinator. The results of applying
-the {\tt f}s and {\tt gs} are combined into a Push array, in the right
-order.
-\pagebreak
-%% ** TODO More explanation needed.
-\begin{codesize}
-\begin{verbatim}
-ixMap :: (UWordE -> UWordE) -> ArrayP a -> ArrayP a 
-ixMap f (ArrayP p n) = ArrayP (ixMap' f p) n
-
-ixMap' :: (UWordE -> UWordE) 
-         -> P (UWordE, a)
-         -> P (UWordE, a) 
-ixMap' f p = \g -> p (\(i,a) -> g (f i,a))
-
-insertZero :: Int -> UWordE -> UWordE
-insertZero 0 a = a `shiftL` 1
-insertZero i a 
-  = a + (a .&. fromIntegral (complement (oneBits i :: Word32)))
-
-ilv2 :: Choice b => 
-        Int -> (a -> a -> b) -> (a -> a -> b) -> 
-        Array a -> ArrayP b
-ilv2 i f g (Array ixf n) 
-   = ArrayP (\k -> app a5 k *>* app a6 k) n
-  where
-    n2 = n `div` 2
-    a1 = Array (ixf . left) (n-n2)
-    a2 = Array (ixf . right) n2
-    a3 = zipWith f a1 a2
-    a4 = zipWith g a1 a2
-    a5 = ixMap left (push a3)
-    a6 = ixMap right (push a4)
-    left = insertZero i
-    right = flipBit i  . left
-    app (ArrayP f _) a = f a
-\end{verbatim}
-\end{codesize}
-\noindent
-This new combinator can now replace {\tt ilv1} in the bitonic
-merger, giving a kernel that runs considerably faster. We will use
-that kernel to build a large sorter later.
-
-The implemenation of {\tt vee2} is almost identical to
-that of {\tt ilv2}, with {\tt flipBit i}
-replaced by {\tt flipLSBsTo} as before (so that, again, one would in
-fact make a more generic function for building such combinators).
-Now, we just need to replace the {\tt ilv1} and {\tt vee1} combinators
-in the tree sorter with {\tt ilv2} and {\tt vee2} , to get a verrsion that uses half as many threads:
-\begin{codesize}
-\begin{verbatim}
-tmerge2 :: Int -> [Array IntE -> ArrayP IntE]
-tmerge2 n = vstage (n-1) : [ istage (n-i) | i <- [2..n]]
-  where
-    vstage i = vee2 i min max
-    istage i = ilv2 i min max
-
-tsort2 :: Int -> [Array IntE -> ArrayP IntE]
-tsort2 n = concat [tmerge2 i | i <- [1..n]]
-\end{verbatim}
-\end{codesize}
-As we shall see in section~\ref{sec:benchmarks}, the resulting code
-is significantly faster. This is because it uses one thread per two array elements,
-and the code no longer contains any conditionals.
-
-In order to go faster still, we resort to building a different sorting network, of
-exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\cite{PeriodicBalanced}.
-This involves the introduction of one new combinator that can be seen as a mixture
-of the {\tt ilv} and {\tt vee} combinators already introduced.
-
-\subsection{A sorter built from the balanced periodic merger}
-First, we note that the balanced periodic merger contains multiple
-uses of the now familiar vee-shaped pattern, see Figure~\ref{fig:periodicMerger}.
-\begin{figure}
-\centering
-\includegraphics[scale=0.25]{balanced}
-\caption{16 input periodic balanced merging network}
-\label{fig:periodicMerger}
-\end{figure}
-
-\begin{codesize}
-\begin{verbatim}
-bpmerge2 :: Int -> [Array IntE -> ArrayP IntE]
-bpmerge2 n = [vstage (n-i) | i <- [1..n]]
-  where vstage i = vee2 i min max
-\end{verbatim}
-\end{codesize}
-
-Now Dowd et al proved that the balanced periodic merger sorts two {\em interleaved} sorted sequences. So, taking an iterative view of the resulting sorter,
-we want to build a tree of mergers as before, but the smaller mergers should
-be interleaved, rather than operating on adjacent sub-sequences.
-There should be one merger on the right hand end of the network; left of
-that, there should be two mergers that operate on the odd and even elements,
-and each of them should in turn be fed by two interleaved mergers, and so.
-The sorter is illustrated, for $16$ inputs in Figure~\ref{fig:vsorter}.
-Just to the left of the final balanced merger, one of the two interleaved
-mergers is shown using dotted lines. It operates on a completely different set of inputs from the other 8-input merger in the same part of the tree.
-
-\begin{figure*}
-\centering
-\includegraphics[scale=0.5]{vsorter}
-\caption{A sorter based on the idea that the periodic
-balanced merger network sorts two interleaved sorted sequences.
-It consists of two half-sized sorters, one working on the
-odd elements of the input and one on the even, followed by
-the balanced merger. The diagram indicates using dotted lines the
-balanced merger that is the final (rightmost) part of
-one of the half-size sorters.}
-\label{fig:vsorter}
-\end{figure*}
-
-The most straightforward way to give an iterative description of
-this sorter is
-to introduce a new combinator that is a combination of
-{\tt ilv} and {\tt vee}.
-The only thing that we need to change is the {\em partner} function.
-This time we will flip not the least significant bits from position $0$ to
-position $i$ but from position $i$ to $i+j$.
-
-\begin{codesize}
-\begin{verbatim}
--- flip bits from position i to position i+j inclusive
-flipBitsFrom :: Bits a => Int -> Int -> a -> a
-flipBitsFrom i j a = a `xor` (fromIntegral mask)
-  where
-    mask = (oneBits (j + 1):: Word32) `shiftL` i
-
-ilvVee1 :: Choice a => 
-           Int -> Int -> 
-           (b -> b-> a) -> (b -> b -> a) -> 
-           Array b -> Array a
-ilvVee1 i j f g arr = Array ixf (len arr)
-  where
-    ixf ix = let l = arr ! ix
-                 r = arr ! newix
-                 newix = flipBitsFrom i j ix
-             in (ifThenElse (lowBit (i+j) ix) (f l r) (g l r))
-\end{verbatim}
-\end{codesize}
-
-\begin{codesize}
-\begin{verbatim}
-ilvVee2 :: Choice b => Int -> Int -> 
-           (a -> a -> b) -> (a -> a -> b) -> 
-           Array a -> ArrayP b
-ilvVee2 i j f g (Array ixf n) 
-  = ArrayP (\k -> app a5 k *>* app a6 k) n
-  where
-    n2 = n `div` 2
-    a1 = Array (ixf . left) (n-n2)
-    a2 = Array (ixf . right) n2
-    a3 = zipWith f a1 a2
-    a4 = zipWith g a1 a2
-    a5 = ixMap left (push a3)
-    a6 = ixMap right (push a4)
-    left = insertZero (i+j)
-    right = flipBitsFrom i j . left
-    app (ArrayP f _) a = f a
-\end{verbatim}
-\end{codesize}
-
-For both variants of the combinator, we simply add to the {\tt ilv} definitions
-a new {\tt Int} parameter, {\tt j}, and replace {\tt flipBit i} by {\tt flipBitsFrom i j}. We also insert the zero bit (when calculating the left index)
-at position {\tt i + j} rather than just at position {\tt i}.
-{\tt ilvVee} is a generalisation of both {\tt ilv} and {\tt vee}.
-{\tt ilvVee i 0} has the same behaviour as {\tt ilv i}, and {\tt ilvVee 0 (j-1)}
-is the same as {\tt vee j}. The {\tt i} parameter controls the degree
-of interleaving, and the {\tt j} parameter controls the size of
-the vee-shaped blocks.
-
-For $16$ inputs, the parameters
-to {\tt ilvVee} that describe the periodic merger on the right of the construction
-are $i=0$ (for no interleaving) paired with $3$, $2$, $1$ and $0$
-for the decreasing size of the vee-shaped blocks.
-Next, to the left, the mergers are interleaved ($i=1$) and there are three
-stages with
-vee-shaped blocks of decreasing size ($j = 2, 1, 0$), see Figure~\ref{fig:vsorter}.
-The following code gives an iterative description of the construction
-for {\small $2^n$} inputs:
-
-\begin{codesize}
-\begin{verbatim}
-vsort :: Int -> Array IntE -> Kernel (Array IntE)
-vsort n = composeS . map pure $ [istage (n-i) (i-j) 
-                                | i <- [1..n], j <- [1..i]]
-  where istage i j = ilvVee2 i j min max
-\end{verbatim}
-\end{codesize}
-
-\noindent
-The resulting generated code uses one thread per
-$2$ indices.
-
-\subsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
-\begin{figure}
-\begin{codesize}
-\begin{verbatim}
-  unsigned int arrayLength = 1 << LOG_L_SIZE;
-  unsigned int diff = LOG_L_SIZE - LOG_S_SIZE;
-  unsigned int blocks = arrayLength / S_SIZE;
-  unsigned int threads = S_SIZE / 2;
-
-  sortSmall<<<blocks, threads,4096>>>(din,din);
-
-  for(int i = 0 ; i < diff ; i += 1){ 
-    vSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<i)*S_SIZE);
-
-    for(int j = i-1; j >= 0; j -= 1)
-      iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
-
-    bmergeSmall<<<blocks,threads,4096>>>(din,din);}
-\end{verbatim}
-\end{codesize}
-\caption{CUDA code for our large sorter. {\tt sortSmall} and {\tt bmergeSmall}
-are replaced by Obsidian-generated kernels in the experiments. {\tt vSwap} and {\tt iSwap} are handwritten CUDA kernels
-that perform one column of compare and swap operations in the vee and ilv shapes respectively, and are
-parameterised on the stride. (Our generated kernels have fixed input and output size.)}
-\label{fig:CUDAsort}
-\end{figure}
-We have measured the performance of generated 512-input sorting and merging kernels by plugging them into a larger sorter written in CUDA.
-The sorter has exactly the structure shown in Figure~\ref{fig:mixedsorter}.
-Figure~\ref{fig:newmixedsorter} shows the location of smaller sorters and mergers, and of vSwap and iSwap kernels for $16$ inputs.
-Larger sorters simply have more columns of mergers, each preceded by a vSwap and a number of iSwaps.
-The overall structure of the resulting CUDA code is shown in Figure~\ref{fig:CUDAsort}.
-Because {\tt iSwap} is used repeatedly, we also wrote kernels
-corresponding to compositions of two and three of them (avoiding
-memory accesses between the columns).
-That is, we replaced the loop containing {\tt iSwap} above by
-\pagebreak
-\begin{codesize}
-\begin{verbatim}
-for(int j = i-1; j >= 0; j -= 3){ 
-  if (j==0) 
-    iSwap<<<blocks/2,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
-    else 
-    {if (j==1) 
-       iSwap2<<<blocks/4,threads*2,0>>>(din,din,(1<<j)*S_SIZE);
-       else 
-       iSwap3<<<blocks/8,threads*2,0>>>(din,din,(1<<j)*S_SIZE);}}
-\end{verbatim}
-\end{codesize}
-\noindent
-The Obsidian implementation and all code for examples in the paper are available at \url{http://www.cse.chalmers.se/~joels/expressive.html}.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.33]{newmixedSorter}
-\caption{
-This diagram shows the tree-shaped sorting network that we saw earlier, but with 4 input sorting and merging kernels indicated.
-This is the structure of the network that we have used to implement large sorters, in which the small kernels having {\small $2^9 = 512$} inputs in all cases.
-For {\small $2^{24}$} inputs overall, the resulting sorter
-has one column of small sorters and $24-9=15$ of small mergers.}
-%% both ofs in the above sentence should remain :)
-\label{fig:newmixedsorter}
-\end{figure}
-
-Table~\ref{fig:table} shows the performance figures for the large sorter
-for $5$ different $512-$input small sorter kernels.
-{\tt tsort1} is defined above, and is a variant of bitonic sort
-that does not require {\tt if} statements to determine the direction
-of sorting of pairs, as all comparison operations place the minimum at the
-same index as the lower input.
-{\tt tsort2} is the same construction, but built with the combinators {\tt ilv2}
-and {\tt vee2} that result in the use of one thread per two indices.
-{\tt vsort}, the fastest kernel, uses a generalisation of those
-two combinators, and again uses one thread per two indices.
-{\tt vsort1} is the same construction as {\tt vsort}, but with {\tt ilvVee2} replaced
-by {\tt ilvVee1}, and so using one thread per index.
-As a reference small kernel implementation, we include a simple hand-coded bitonic sort
-that uses one thread per index (see code in Appendix).
-None of the kernels has been subject to optimisations related to warp size.
-
-\input{table1.tex}
-
-We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:table2}.
-The {\tt tsort1} and {\tt vsort1} kernels, which are built using only pull arrays, 
-(and which have one index per thread in the generated code) are
-noticeably slower than those that have two indices per thread.
-The generation of the latter kernels is made possible by the introduction of 
-push arrays.
-
-\input{table2.tex}
-
-Although none of our generated kernels is optimised (for example with respect to warp size), 
-their performance is, nevertheless, very good. We are working on automated warp size-related 
-optimisations. It would also be interesting to explore the fusion of adjacent columns of 
-comparators in the small kernels; omitting a {\tt sync} would cause this fusion to happen
-but we also need to modify the threading behaviour (doubling the number of indices per thread).
-
-%The final version of the paper will contain benchmarking against a highly optimised sorter kernel obtained from our colleagues in the computer graphics group at Chalmers.
-For now, we emphasise that the {\tt vsort} kernel is so fast that we expect it to compare favourably.
-Given the simplicity of its description in Obsidian, we feel that this is a good way to write fast kernels.
-
-%\input{discussion}
-
-\section {Discussion}
-
-%% ** TODO. Mention CSE. warp-related opt.
-
-Push arrays form a new approach to array representation in DSELs.
-We do not know of similar approaches in the literature, despite
-the fact that the notions of demand and data flow
-may feel familiar to the reader who considers Pull and Push arrays.
-The addition of Push arrays to Obsidian seems highly beneficial. With 
-this new feature, the user gains finer grained control over the code generated
-and the resulting CUDA kernels perform considerably better than before.
-This 
-was illustrated in the series of sorters explored in section~\ref{sec:MARY}.
-The performance of {\tt vsort} is sufficiently good that it
-can be used as a first phase in a larger sorter (written in CUDA) that
-can sort 16M elements in 96 ms, while an i7-920 CPU takes
-around 2740 ms. Further speed improvements look possible, both in the coordination code
-and in the kernels. An obvious next step would be to investigate the generation of the {\tt iSwap} and
-{\tt vSwap} kernels from Obsidian.
-(This is not currently possible because of assumptions that we made about the interfaces to kernels and about how {\em thread ids} are used.
-We will look into ways to relax
-our assumptions.)
-
-The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
-description language~\cite{LavaSorter} was that a relatively small set of combinators went a long way. So, although we introduced three
-combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which
-includes the other two, we do not believe that every new kernel development exercise would demand a completely new set of combinators.
-We expect to provide the user with a well-documented set of combinators,
-so that users can get access to this style of programming without having
-to develop their own combinators, and without having to think too much
-about bit-hacking.
-The bit-manipulation approach chosen to define our combinators automatically
-created functions that apply to sub-sequences of the input that
-are of an appropriate length.
-
-In this paper, we made combinators for the special case of two input, two
-output operations (built from two two-input funtions that we typically called
-{\tt f} and {\tt g}). 
-This approach should be generalised to deal with blocks that have $2^k$ inputs
-and outputs.
-Also, we made a compound combinator from {\tt ilv} and {\tt vee},
-but generalising to more than two input components would allow
-for composing combinators, and indeed for recursive descriptions
-that could be unrolled. Then, ignoring {\tt syncs}, a recursive description of
-{\tt vsort} could be something like 
-\begin{codesize}
-\begin{verbatim}
-vsortR 0 = id
-vsortR n = bpmergeR n . ilv2 1 (vsortR (n-1))
-\end{verbatim}
-\end{codesize}
-\noindent
-It would then be necessary to optimise the code generated from
-multiple applications of {\tt ilv2 1}, for example, whereas here
-we have forced the user to figure out both the unrolling and the combinations.
-Moving to more general combinators would also give the opportunity to
-provide predefined combinators that capture more of the commonly used threading patterns (for instance $k$ indices per thread rather than the $1$ and $2$ shown here).
-
-The integration of Push arrays into Obsidian raises some new questions.
-Previously, there was a direct correspondence between the length 
-of an array and the number of threads used to compute it, which
-allowed the user to write an initial program without worrying about
-threads at all, and then to tweak the Obsidian program if he was not satisfied
-with the threading behaviour of the resulting kernel. 
-Now, as can be seen in the {\tt catArrayPs} example and in the sorters, this correspondence
-can be broken using Push arrays. The {\tt catArrayPs} example and two
-of the sorters use half as many threads as the 
-number of elements. For users who are very concerned about the speed of
-the generated kernels, getting this control through using Push arrays in
-a particular pattern is clearly a good thing. But adding a second, different
-way to control thread use in the generated code certainly complicates matters, and further case studies are needed to confirm that the complication pays off.
-
-The addition of Push arrays also adds the possibility to include potentially 
-unsafe operations in Obsidian, for example by writing multiple array elements to the same index, or by discarding elements.
-This new expressiveness will have to be carefully controlled. On the positive side, it offers the possibility to encode functions like {\tt filter} from Haskell
-that
-are simply not expressible using only Pull arrays.
-Being able to implement {\tt filter} would make programming kernels in obsidian feel much more like programming in Haskell -- a welcome loosening of the strait-jacket.
-Once that is done, it will be time to develop a very simple coordination language to allow programming of entire GPU applications that make use of
-the kind of small kernel building blocks developed here.
-
-\appendix
-\section{Appendix}
-\begin{codesize}
-\begin{verbatim}
-__device__ inline void swap(int & a, int & b)
-{
-   int tmp = a;
-   a = b;
-   b = tmp;
-}
-
-__global__ static void bitonicSort(int * values, int *results)
-{
-   extern __shared__ int shared[];
-
-   const unsigned int tid = threadIdx.x;
-   const unsigned int bid = blockIdx.x;
-
-   // Copy input to shared mem.
-   shared[tid] = values[(bid*NUM) + tid];
-
-   __syncthreads();
-
-   // Parallel bitonic sort.
-   for (unsigned int k = 2; k <= NUM; k *= 2)
-   { 
-     // bitonic merge
-     for (unsigned int j = k / 2; j>0; j /= 2)
-       {
-           unsigned int ixj = tid ^ j;
-
-           if (ixj > tid)
-           {
-               if ((tid & k) == 0)
-               {
-                   if (shared[tid] > shared[ixj])
-                   {
-                       swap(shared[tid], shared[ixj]);
-                   }
-               }
-               else
-               {
-                   if (shared[tid] < shared[ixj])
-                   {
-                       swap(shared[tid], shared[ixj]);
-                   }
-               }
-           }
-           __syncthreads();
-       }
-   }
-
-   // Write result.
-   results[(bid*NUM) + tid] = shared[tid];
-}
-\end{verbatim}
-\end{codesize}
-
-\acks
-This research has been funded by the Swedish Foundation for Strategic Research (which funds
-the RAW FP Project) and
-by the Swedish Research Council.
-
-% We recommend abbrvnat bibliography style.
-
-%\bibliographystyle{abbrvnat}
-%\bibliography{paper}
-
-% The bibliography should be embedded for final submission.
-
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Axelsson et~al.(2011)Axelsson, Claessen, Sheeran, Svenningsson,
-  Engdal, and Persson]{FELDSPAR}
-E.~Axelsson, K.~Claessen, M.~Sheeran, J.~Svenningsson, D.~Engdal, and
-  A.~Persson.
-\newblock {T}he {D}esign and {I}mplementation of {F}eldspar an {E}mbedded
-  {L}anguage for {D}igital {S}ignal {P}rocessing.
-\newblock In \emph{Proceedings of the 22nd international conference on
-  Implementation and application of functional languages}, IFL'10, pages
-  121--136, Berlin, Heidelberg, 2011. Springer-Verlag.
-
-\bibitem[Batcher(1968)]{Batcher}
-K.~E. Batcher.
-\newblock Sorting {N}etworks and {T}heir {A}pplications.
-\newblock In \emph{Proc. AFIPS Spring Joint Computer Conference 32}, pages
-  307--314, 1968.
-
-\bibitem[Billeter et~al.(2009)Billeter, Olsson, and Assarsson]{OLAMARCUS}
-M.~Billeter, O.~Olsson, and U.~Assarsson.
-\newblock Efficient stream compaction on wide {SIMD} many-core architectures.
-\newblock In \emph{Proc. Conf. on High Performance Graphics}, HPG '09, pages
-  159--166. ACM, 2009.
-
-\bibitem[Chakravarty et~al.(2011)Chakravarty, Keller, Lee, McDonell, and
-  Grover]{ACCELERATE}
-M.~M. Chakravarty, G.~Keller, S.~Lee, T.~L. McDonell, and V.~Grover.
-\newblock Accelerating {H}askell array codes with multicore {GPU}s.
-\newblock In \emph{Proc. sixth workshop on {D}eclarative {A}spects of
-  {M}ulticore {P}rogramming}, DAMP '11, pages 3--14. ACM, 2011.
-
-\bibitem[Claessen(1999)]{POORKOEN}
-K.~Claessen.
-\newblock A poor man's concurrency monad.
-\newblock \emph{J. Funct. Program.}, 9\penalty0 (3):\penalty0 313--323, 1999.
-
-\bibitem[Claessen et~al.(2001)Claessen, Sheeran, and Singh]{LavaSorter}
-K.~Claessen, M.~Sheeran, and S.~Singh.
-\newblock The {D}esign and {V}erification of a {S}orter {C}ore.
-\newblock In \emph{Proc. Int. Conf. on Correct Hardware Design and Verification
-  Methods (CHARME)}, volume 2144 of \emph{Springer LNCS}, pages 355--369, 2001.
-
-\bibitem[Dowd et~al.(1989)Dowd, Perl, and Larry~Rudolph]{PeriodicBalanced}
-M.~Dowd, Y.~Perl, and M.~S. Larry~Rudolph.
-\newblock The {P}eriodic {B}alanced {S}orting {N}etwork.
-\newblock \emph{Journal of the ACM}, 36:\penalty0 738--757, 1989.
-
-\bibitem[Elliott(2003)]{PAN}
-C.~Elliott.
-\newblock Functional {I}mages.
-\newblock In \emph{The Fun of Programming}, ``Cornerstones of Computing''
-  series. Palgrave, Mar. 2003.
-
-\bibitem[Elliott et~al.(2003)Elliott, Finne, and de~Moor]{COMPILEEDSL}
-C.~Elliott, S.~Finne, and O.~de~Moor.
-\newblock Compiling embedded languages.
-\newblock \emph{Journal of Functional Programming}, 13\penalty0 (2), 2003.
-
-\bibitem[{J}oel {S}vensson(2011)]{JSLIC}
-{J}oel {S}vensson.
-\newblock {O}bsidian: {GPU} {K}ernel {P}rogramming in {H}askell.
-\newblock Technical Report 77L, {C}omputer {S}cience and {E}nginering,
-  {C}halmers {U}niversity of {T}echnology, {G}othenburg, 2011.
-\newblock Thesis for the degree of Licentiate of Philosophy.
-
-\bibitem[Keller et~al.(2010)Keller, Chakravarty, Leshchinskiy, Peyton~Jones,
-  and Lippmeier]{REPA}
-G.~Keller, M.~M. Chakravarty, R.~Leshchinskiy, S.~Peyton~Jones, and
-  B.~Lippmeier.
-\newblock Regular, shape-polymorphic, parallel arrays in {H}askell.
-\newblock In \emph{Proceedings of the 15th ACM SIGPLAN international conference
-  on Functional programming}, ICFP '10, pages 261--272, New York, NY, USA,
-  2010. ACM.
-
-\bibitem[Larsen(2011)]{BARRACUDA}
-B.~Larsen.
-\newblock Simple optimizations for an applicative array language for graphics
-  processors.
-\newblock In \emph{Proc. sixth workshop on {D}eclarative {A}spects of
-  {M}ulticore {P}rogramming}, DAMP '11, pages 25--34. ACM, 2011.
-
-\bibitem[Mainland and Morrisett(2010)]{NIKOLA}
-G.~Mainland and G.~Morrisett.
-\newblock Nikola: embedding compiled {GPU} functions in {H}askell.
-\newblock In \emph{Proc. third ACM Symposium on Haskell}. ACM, 2010.
-
-\bibitem[Sheeran(1989)]{sortsRuby}
-M.~Sheeran.
-\newblock Describing {B}utterfly {N}etworks in {R}uby.
-\newblock In \emph{Functional Programming}, pages 182--205. Springer Workshops
-  in Computing, 1989.
-
-%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-%P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
-
-\end{document}

### 2013-09-30 src2/expressive/table.tex
another approach

% table

\begin{figure}
\begin{center}
  \begin{tabular}{  l | r | r}
    %\hline
    Kernel               & Threads   & ms / grid   \\ \hline
    bitonic\_sort(CUDA)  & 512       & 30.28       \\      
    tsort1               & 512       & 15.82       \\
    tsort2               & 256       & 11.64       \\ 
    vsort1               & 512       & 14.96       \\ 
    vsort                & 256       &  9.33       \\ 
    %\hline
  \end{tabular}
\end{center}
\label{fig:table}
\caption{Table is showing the time it takes for a grid of 32678 blocks of 
threads to execute.Each grid was launched 1000 times, the total execution time 
was then divided by 1000 to get the listed figures. The GPU used is a NVIDIA GTX480} 
\end{figure}

### 2013-09-30 src2/expressive/table1.tex
another approach

% table

\begin{table}
\begin{center}
  \begin{tabular}{| l | r | r | r | r | r | r | }
    \hline
        k                             &  $20$ &  $21$ & $22$& $23$ & $24$ & $24$(CPU)  \\ \hline
    bitonic(CUDA)(512)                &  5 & 12 & 25 & 54 & 117 & 2741   \\      
    tsort1(512)                       &  4 & 10 & 22 & 47 & 102 & 2742    \\
    tsort2(256)                       &  4 & 9 & 23 & 45 & 98 & 2741     \\ 
    vsort1(512)                       &  4 & 10 & 21 & 47 & 102 & 2747    \\ 
    vsort(256)                        &  4 & 9 & 20 & 44 & 96 & 2783    \\ 
    \hline
  \end{tabular}
\end{center}
\label{fig:table}
\caption{Sorting time (ms) for $2^k$ inputs, $20 \leq k \leq 24$.  The GPU used is an NVIDIA GTX480.
Each line shows the result using a particular small $512$ input sort kernel with the
indicated number of threads as {\tt sortSmall} in the CUDA code above. The small merge kernel used
is our generated {\tt bmerge2}, with two indices per thread and $512$ inputs. Memory transfer time (GPU-CPU)
is not shown. In the $2^{24}$ input case using {\tt vsort}, the total sorting time, including
memory transfers was 141 ms.
The rightmost column shows the time taken for the C quicksort function {\tt qsort}, compiled with
{\tt gcc -O3}, to sort the same inputs as those sorted on the GPU on an i7-920 2.8GHz CPU.
} 
\end{table}

%% bitonic 30203
%%

### 2013-09-30 src2/expressive/table2.tex
another approach

% table

\begin{table}
\begin{center}
  \begin{tabular}{| l | r | r |   }

    \hline
                              & time & percent \\ \hline
    bitonic(CUDA)             & 30203 & 32.07\\
    tsort1                    & 15823 & 19.72  \\
    vsort1                       & 14955 & 18.76 \\ 
    tsort2                      & 11562 & 15.37    \\ 
    vsort                       & 9228 &  12.67 \\ 
    bitonic (NVIDIA SDK)        & 23815 & 6.55 \\
    \hline
  \end{tabular}
\end{center}
\label{tab:table2}
\caption{time: GPU time (in $\mu$s) spent in calls to small sorter kernels 
in the initial phase of the sorter for $2^{24}$ inputs. percent: percentage 
of total GPU time spent in the small sort kernel.
The final line shows the numbers for the bitonic code (for large sorts) that 
is distributed with the the NVIDIA SDK. Its structure also starts with a phase 
of small sorting kernels. It takes 274 ms to sort $2^{24}$ key-value pairs on 
an NVIDIA GTX480 GPU. It could be sped up by using our {\tt vsort} kernel, 
and also by (hand) fusing single ``column'' kernels as we did with {\tt iSwap}, 
although the need to calculate directions of comparators in the bitonic network 
would complicate this. Our CUDA code is simpler because all comparators
point in the same direction in the sorter construction.
} 
\end{table}

%% bitonic 30203
%% tsort1 15823
%% tsort2 11562
%% vsort1 14955
%% vsort 9228

### 2013-09-30 src2/thesis.tex
another approach

\documentclass[a4paper]{book}

%\usepackage[latin9]{inputenc}
\usepackage[utf8x]{inputenc}
\usepackage[pdftex]{graphicx}
\usepackage{pdfpages}
\usepackage[british]{babel}
\usepackage{placeins}
\usepackage{dialogue}
\usepackage{fancyvrb}

\usepackage{t1enc}         
\usepackage{times}
\usepackage{url}
\usepackage{enumerate}

\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{lingmacros}
\usepackage{subfig}

\pagestyle{plain}
%\pagenumbering{arabic}

%stuff
\usepackage{latexsym}
\usepackage{tikz,pgflibraryshapes}
\usepackage{booktabs}
%%

\newcommand{\bequ}{\begin{quote}}
\newcommand{\enqu}{\end{quote}}
\newcommand{\eop}[1]{\mbox{\sl #1}}
\newcommand{\sube}[2]{#1_{#2}}
\newcommand{\id}[1]{{\it #1}}
\newcommand{\kwd}[1]    {\mbox{\textbf{#1}}}
\newcommand{\str}[1]{{\id "#1"}}
\newcommand{\lemon}{\emph{lemon}}
\newcommand{\textt}[1]{\texttt{#1}}
\newcommand{\txttt}[1]{\texttt{#1}}

\renewcommand*{\thesection}{\arabic{section}}

\begin{document}

%\input{titlepage}

\clearpage

\section*{Abstract}
Abstract abstract

\vspace{5mm}

\noindent

\textbf{Keywords:} Keyword1, Keyword2

\clearpage

\section*{Acknowledgements}
ACK ACK ACK

\vspace{5mm}

FUNDING FUNDING

\tableofcontents

\newpage

\chapter{Introduction}

\clearpage{}
\section*{Additional Publications}

Papers not included in thesis

\bibliographystyle{splncs}
\bibliography{intro}

\chapter{Papers}

\cleardoublepage

\section*{\centering Expressive Array Constructs in an Embedded GPU Kernel Programming Language}

\begin{center} 
Koen Claessen, Mary Sheeran, Bo Joel Svensson
\end{center}

\input{./expressive/paperThesis}

\cleardoublepage

\section*{\centering Simple and Compositional Reification of Monadic Embedded Languages}

\begin{center} 
Josef Svenningsson, Bo Joel Svensson
\end{center}

\input{./bb/paperThesis}

\cleardoublepage

\section{Paper2}

%\input{}

\cleardoublepage

\end{document}

### 2013-09-30 src2/csort/Algorithm.tex
apa

\section{Counting Sort}
\label{sec:algorithm}

Counting sort, like radix sort, relies on array indexing rather than
comparison to order elements. The key idea of counting sort is to
count the number of occurrences of each element, and then perform a
prefix sum over all the counts. The prefix sum generates an array
which, for each element in the original array, will point to its
position in the final sorted array. In order for this to work,
counting sort needs a lower and upper bound on the values of the
elements in the input array. Counting sort is typically explained in
terms of sorting integers and we will do the same here.

The array \verb!{5,2,5,7,1}! will be used as a running example. In
addition, the range \verb!(1,10)! will be the approximation of the
range of the keys in the input. A diagrammatic illustration of our
running example is presented in figure \ref{fig:example}.

\begin{figure}
\begin{small} 
\begin{verbatim}
countingSort :: (Int,Int) 
                -> Array Int Int 
                -> Array Int Int
countingSort range input =
  reconstruct $
  scanlArray (+) 0 $
  histogram range input

histogram :: (Int,Int) 
             -> Array Int Int 
             -> Array Int Int
histogram range = 
  accumArray (\i _ -> i+1) 0 range . elems . fmap dup

reconstruct :: Array Int Int -> Array Int Int
reconstruct arr = 
  array (0,arr!l-1)
    [ (a,i)
    | (i,e) <- init (assocs arr)
    , a <- [ e .. arr!(i+1) - 1] ]
  where (_,l) = bounds arr

dup a = (a,a)
\end{verbatim}
\end{small}
\caption{A Haskell specification of Counting Sort. The function
  \texttt{scanlArray} is not standard but works much like
  \texttt{scanl}
  for lists.}
\label{fig:haskell}
\end{figure}

%% \begin{figure}
%% \begin{center}
%% \includegraphics[height=4cm]{./CountingSortDiagram}
%% \end{center}
%% \caption{An example run of counting sort in diagrammatic form}
%% \label{fig:example}
%% \end{figure}
\begin{figure}
\input{./csort/csort.tikz}
\caption{An example run of counting sort in diagrammatic form}
\label{fig:example}
\end{figure}

Counting sort consists of three steps: histogram creation, prefix sum
and reconstruction.

The histogram creation step computes a new array which records how
many times an integer occurs in the input array, i.e. the new array is
a histogram of the input array. The size and indices of the histogram
array are determined by the range which approximates the lower and
upper bounds of the keys.

Computing the histogram of \verb!{5,2,5,7,1}! with the range \verb!(1,10)!, 
results in \verb!{1,1,0,0,2,0,1,0,0,0}!.
%When running the example array \verb!{5,2,5,7,1}! through \\
%\verb!histogram! together with the range \verb!(1,10)! the resulting
%histogram is \verb!{1,1,0,0,2,0,1,0,0,0}!. 
Indexing in this array
starts at \verb!1! because that's the lower bound given in the
range. Each element in the histogram indicates how many times the
corresponding value occurs in the input array. For example, \verb!5!
occurs twice and \verb!8! occurs zero times.

The second step of the algorithm computes the prefix sum (or inclusive
scan with plus) of the histogram array, and prepends a zero.  For input array
\verb!{5,2,5,7,1}!  and range \verb!(1,10)!, the histogram array is
\verb!{1,1,0,0,2,0,1,0,0,0}! and its prefix sum is {\tt
  \{0,1,2,2,2,4,\allowbreak{}4,5,5,5,5\}}.

%\verb!{0,1,2,2,2,4,4,5,5,5,5}!. This array starts with 0 and is one
%longer than the histogram array.

The array computed by the prefix sum is called the position array. It
indicates where in the final sorted array each value is placed. The
indices of the position array correspond to values in the final array
and the elements correspond to positions in the final array. The
reconstruct step takes care of distributing the values into the final
sorted array. For each index in the position array, the difference
between the element at that index and the index plus one is
computed. This difference indicates how many times the value occurs in
the final array.

The position array in our running example is {\tt\{0,1,2,2,2,4,\allowbreak{}4,5,5,5,5\}}.
Indexing starts at \verb!1! just as with the histogram. As an example,
element \verb!1! occurs once in the final array because the
difference between the element at position \verb!1! and \verb!2! in
the position array is precisely one. Also, \verb!1! is placed
on index \verb!0!. Furthermore, element \verb!5! occurs twice in
the final array (the difference between \verb!2! and \verb!4! in the
position array), on indices \verb!2! and \verb!3!. Continuing this
process with all elements will result in the following final array:
\verb!{1,2,5,5,7}!.

A functional description of counting sort can be found in figure
\ref{fig:haskell}. The main function is \verb!countingSort! which
takes two arguments: a range with a safe approximation of the lower
and upper bounds of the values in the input array, and the input array
itself. The histogram step is implemented by the function
\verb!histogram!, the prefix sum with \verb!scanlArray! and the
reconstruct step with the function \verb!reconstruct!.

\subsection{Occurrence sort: Removing duplicates}

There is an interesting variation of counting sort which removes
duplicate elements. It seems to be folklore; it is mentioned on
Wikipedia \cite{wikipedia}. However, we haven't found any description
of it in the literature. It is particularly interesting because it
allows for an efficient data-parallel implementation.
%% something which
%% will be explored in section \ref{sec:parallel}. Here we focus on the 
%% functional specification of the algorithm.
%% The rest of this
%% section will focus on the functional specification of the algorithm.

\begin{figure}
\begin{small}
\begin{verbatim}
countingSort :: (Int,Int) 
                -> Array Int Int 
                -> Array Int Int
countingSort range input =
  reconstruct $
  scanlArray (+) 0 $
  occurs range input

occurs :: (Int,Int) 
          -> Array Int Int 
          -> Array Int Int
occurs range = 
  accumArray (\_ _ -> 1) 0 range . elems . fmap dup
\end{verbatim}
\end{small}
\caption{A Haskell specification of occurrence sort}
\label{fig:duphaskell}
\end{figure}

Figure \ref{fig:duphaskell} shows the changes we make 
to counting sort in order to obtain occurrence sort.
%% counting sort compute a sorted array and at the same time remove
%% duplicate elements. 
Only the \verb!histogram! function needs changing
so that instead of computing an actual histogram which counts the
number of occurrences of an element, it only records if an element
occurs or not with a \verb!1! or a \verb!0!. Since the function
no longer computes a histogram it has been renamed to
\verb!occurs!.

The prefix sum and the reconstruction step can be kept intact. They
will still compute the correct result.

Compared to the full counting sort, occurrence sort has some
interesting properties which make it more suited for parallel execution.
In particular, the histogram construction phase and the reconstruction
phase allow for less synchronisation.

When constructing the histogram in the counting sort algorithm, an
index has to be incremented each time the corresponding value is found
in the input array. When parallelising this operation, it is important
that the increments are done atomically, which incurs a cost. The occurs
phase presented above doesn't need to employ any special mechanism to
ensure atomicity, since it writes a single word with the value $1$ to
memory.

Given the array \verb!{5,2,5,7,1}! as input and \verb!(1,10)!  as
range, the occurs step will now compute
\verb!{1,1,0,0,1,0,1,0,0,0}!. The prefix sum will in turn produce the
following position array: \verb!{0,1,2,2,2,3,3,4,4,4,4}!. The
reconstruct step computes \verb!{1,2,5,7}!. In particular,
there is only one occurrence of \verb!5! in the final array.

### 2013-09-30 src2/csort/GPUs.tex
apa

\section{GPUs and CUDA} 
%\FloatBarrier
% This section provides an overview of GPU programming and architecture.

The GPUs we target in this paper and with Obsidian are NVIDIA GPUs which
support CUDA. Here we provide some background information on GPUs 
that we hope will aid in the understanding of example programs throughout this paper. 
%% This section briefly outlines features of GPUs that have 
%% influenced the current version of Obsidian and that we make use of when 
%% implementing the algorithms in following sections.

GPUs supporting CUDA are based on a scalable design. The GPU consists of a 
number of so-called multiprocessors (MPs) each consisting of a number of 
processing elements (PEs); sometimes these are referred to as cores or as 
stream processors. An MP also contains a local shared memory through which 
threads running on the PEs can communicate. The key that 
makes the architecture scalable is that a GPU consists of one or more 
of these MPs. This also influences the programming model significantly; since
threads can only communicate in a synchronised manner using the local memory 
(or in a very limited way using atomic operations) the programmer must 
partition the computation in such a way that its communication patterns 
follow this architectural constraint. The GPU also has access to a larger 
memory called the {\em global} or {\em device} memory. The sizes of these 
memories and the number of MPs and PEs per MP vary slightly between 
generations of GPUs. For example, the GPU used in the benchmarks 
(section \ref{sec:Benchmarks}) has a total of 1344 single-precision floating
point CUDA cores and has 2 GB global memory. The very latest GPU architectures 
from NVIDIA slightly digress from what is outlined here but mostly still 
adhere to the same programming model. For exact details refer to \cite{KEPLER}.

The programming model that CUDA exposes is called single program multiple 
threads (SPMT) and is a slight variation of the SPMD concept. This programming 
model reflects the GPU architecture. In CUDA, the programmer writes a single 
program that is executed by many threads. Because of the scalable architecture,
with its one or more MP, these threads are grouped into {\em blocks} 
(now up to 1024 threads per block). A block of threads is executed on an MP; 
thus only threads within that block can communicate using the shared memory. 
The same CUDA program can be executed on any GPU along the scale, from the 
smallest with only one MP to the largest. The only difference is in the number 
of blocks of threads that can be executed in parallel. This constraint implies 
that all blocks must be free of any sequential dependencies. The collection of 
all blocks is called a {\em Grid}.

A CUDA program has two parts; there is a coordination program that runs on the 
CPU and there are kernels that execute on the GPU MPs. The program describing 
a kernel is parameterised over a blockId and a threadId so that 
decisions can be made from that information during execution. A typical 
CUDA program starts out by loading data into local memory using some 
function of blockId and threadId. It then computes on the local memory 
and uses a syncthreads primitive when exchanging values between threads. 
When the local computation is done the final results are written back into 
global memory again using some function of blockId and threadId. 
%A detail 
%here is that within a block, a smaller number of threads (32 of them) 
%called a {\em Warp} can communicate safely without using the synchronisation 
%primitive. This comes from how the GPU executes the warps; this is done 
%in SIMD fashion, all threads progress at the same pace (essentially that 
%they share a single instruction pointer).

For more CUDA and GPU specifics see references \cite{KEPLER,CUDA}.

\begin{figure}
\begin{small}
\begin{verbatim}
__global__ void addv(float *i0, 
                     float *i1, 
                     float *r){ 
  unsigned int ix = 
     blockIdx.x * blockDim.x + threadIdx.x;

  extern __shared__ float sm[]; 
  sm[threadIdx.x] = i0[ix] + i1[ix]; 
  r[ix] = sm[threadIdx.x];
} 
\end{verbatim}
\end{small}
\caption{The code illustrates elementwise addition of vectors. Shared memory 
(the sm array) is used to allow the {\tt addv} program to be run with the 
result {\tt r} pointing to the same memory area as either of the inputs.} 
%But mostly shared memory is used to illustrate the syntax involved.}
\label{fig:cudaKernel1}
\end{figure}

\begin{figure}
\begin{small}
\begin{verbatim}
#define BLOCK_SIZE 32
#define BLOCKS 4 
#define N (BLOCKS * BLOCK_SIZE)

int main(int argc, char **argv){
  float *v1, *v2, *r;
  float *dv1, *dv2, *dr;

  v1 = (float*)malloc(N*sizeof(float));
  v2 = (float*)malloc(N*sizeof(float)); 
  r = (float*)malloc(N*sizeof(float));

  //Generate or read input data.
  ...

  //Allocate arrays in Global memory
  cudaMalloc((void**)&dv1, sizeof(float) * N ); 
  cudaMalloc((void**)&dv2, sizeof(float) * N ); 
  cudaMalloc((void**)&dr, sizeof(float) * N );

  //Copy data into Global memory
  cudaMemcpy(dv1, v1, sizeof(float) * N, 
                      cudaMemcpyHostToDevice);
  cudaMemcpy(dv2, v2, sizeof(float) * N, 
                      cudaMemcpyHostToDevice);

  //Launch the vector add kernel.
  addv<<<BLOCKS, BLOCK_SIZE, BLOCK_SIZE * sizeof(float)>>>
      (dv1,dv2,dr);

  //Launch further kernels on the data.
  ...

  cudaMemcpy(r, dr, sizeof(float) * N , 
                    cudaMemcpyDeviceToHost);

  cudaFree(dv1);
  cudaFree(dv2);
  cudaFree(dr);

  //Show or further process results on the CPU. 
  ... 
}

\end{verbatim}
\end{small} 
\caption{This code starts a CUDA kernel on  
the GPU. The syntax $<<<${\tt nb,nt,sm}$>>>$ is used to set up a kernel
launch configuration. The {\tt nb}, {\tt nt} and {\tt sm} quantities refer to 
the number of blocks, the number of threads/block and the amount of shared 
memory.}
\label{fig:cudaCoord}
\end{figure}

%\FloatBarrier

### 2013-09-30 src2/csort/Obsidian.tex
apa

\section{Obsidian} 
\label{sec:Obsidian} 
%\FloatBarrier

Obsidian is a programming language for expressing general purpose 
GPU kernels, compute kernels, in a high level and functional style.  
Obsidian is embedded in Haskell.
Specifically, Obsidian is a library of functions for generating 
and combining abstract syntax trees.
%That Obsidian is embedded in Haskell means that it is a library of 
%functions for generating and combining abstract syntax trees. 
By using features of Haskell such as overloading and higher-order-functions, 
the library/language border is blurred and we call the result an embedded 
language. From the abstract syntax trees generated while running an 
Obsidian program, NVIDIA CUDA code is generated.

\subsection{A brief history of Obsidian} 
\label{sec:OBSHist}

In the Obsidian project we experiment with combinators and language 
features for GPU kernel implementation. We seek to strike a balance 
between low-level control and useful high-level abstractions. Over time, this 
has led to a number of different versions of Obsidian. In reference \cite{JSLIC}, 
two versions of Obsidian differing by having a {\em monadic} or a limited {\em arrow} 
like interface are described. We have settled on using the monadic style.

One of the array representations that Obsidian uses, the Pull array, was present already 
in the earlier work but went under a different name. It was not until discovering 
a complementary array representation, that we call push arrays, that the traditional 
array was renamed to pull.

The Obsidian implementation of pull arrays was influenced by Pan's representation 
of images as functions from coordinates to colour values \cite{ELLIJFP}. A pull array 
is a function from index to value and an associated length.

Push arrays were added to Obsidian first in reference \cite{PUSH}. A push array
specifies where elements are to end up in a result array, rather than where 
they come from (as in a pull array). In essence, a push array is a
computation which writes an array to memory, but is parameterised on
the {\em write function} which writes a single element into memory.
The push array function can then 
use the write function zero, one, or several times.

Push arrays were added to Obsidian with the purpose of solving some performance 
issues that we had been struggling with for a long time. These problems were 
mainly concerned with concatenating or combining arrays in an efficient way.
Push arrays have also been adopted by others and are used in the implementation 
of filters in reference~\cite{FPCDSL} and in the new Nikola\footnote{github.com/mainland/nikola/blob/master/src/Data/Array/Nikola/Repr/Push.hs}.

Both push and pull arrays are virtual, in the sense that they do not directly 
correspond to any data in a region of memory. This virtual nature of the array 
representations gives us fusion of operations for free. The programmer has 
to explicitly force an array (using a {\tt force} function) to actually compute 
the values and store them in memory and to prevent operations from fusing.

In the version of Obsidian used in this paper\footnote{github.com/svenssonjoel/Obsidian/branches/February2013}, new features were added.  We 
introduce two new variants of push and pull arrays called {\tt GlobPush} and 
{\tt GlobPull}. This introduces a distinction between local arrays, short 
enough to be computed by a block of the GPU, and global arrays that represent 
computations spread over blocks of the GPU. This is an important step when 
it comes to Obsidian's capabilities; we can now express in what order a kernel
fetches elements from global memory. 
Earlier versions of Obsidian were limited to straight-line block indexing. A kernel 
generated by older Obsidian accessed blocks of elements in a fixed way (thread block 
i accessed elements using {\tt (i * blocksize) + f tid}, where {\tt f} can permute 
indices within a block but there is no similar way to permute the actual blocks).

In order to implement histograms we also need to add  atomic operations to the language. The addition of atomic operations here is 
done in a rather ad hoc way and pushes us to program in a very low-level and 
imperative style. We strive to incorporate much of CUDA's low-level functionality 
into Obsidian and adding atomic operations is an attempt in that 
direction.

The version of Obsidian used in this paper also introduces a new monadic representation 
of GPU programs. We call the monad {\tt Program}. The {\tt Program} type has a 
parameter that represents at what level in the GPU hierarchy it executes. There 
are sequential {\em Thread} programs and parallel {\em Block} and {\em Grid} programs. Type synonyms 
are available, {\tt TProgram}, {\tt BProgram} and {\tt GProgram}.

Many of the new features we mention above have been further refined and 
are present in versions of Obsidian following the one we use in this paper. For 
example, in the master branch of Obsidian\footnote{github.com/svenssonjoel/Obsidian} we build on the ideas in this paper. 
There, global and local arrays are represented by the same data type. That 
work also goes further by implementing hierarchy generic functions that are 
applicable at different levels of the Thread, Block and Grid hierarchy. 
We are also experimenting with the addition of mutable arrays in order to 
better integrate the atomic operations\footnote{github.com/svenssonjoel/Obsidian/tree/mutable}.

%% Obsidian raises the level of abstraction for GPU programming but wants 
%% to keep low level control. To facilitate this, Obsidian supplies different 
%% kinds of arrays. There are {\tt Push} and {\tt Pull} arrays for local 
%% computations and in the current version of 
%% Obsidian\footnote{github.com/SvenssonJoel/Obsidian} we add global arrays 
%% of types {\tt GlobPush} and {\tt GlobPull} to the language.

%% Both push and pull arrays are in some sense virtual. They do not necessarily 
%% represent data that exists in memory. Instead, the push and pull arrays 
%% represent ways to compute an array. Concretely, a pull array is function from 
%% index to element. A push array is a function that when given a write-function 
%% (a function that takes an index and an element to a program 
%% that represents the writing of that element into an array) gives 
%% a program that represents writing a number of elements. In 
%% Obsidian, there are three kinds of programs, represented by a {\tt Program} data 
%% type. TPrograms, BPrograms and GPrograms represent what a thread, block of 
%% threads and a grid of blocks can do. Thread programs are the result type in 
%% the so-called write-functions. BPrograms operate on local arrays and 
%% grid programs operate on global arrays. 
%% Since push and pull arrays are virtual operations upon them are fused 
%% automatically. The standard example is map-fusion {\tt (map a . map b = map (a . b))}. 
%% The programmer decides when, and if, intermediate values are computed and 
%% stored in memory by using a {\tt force} function.
%% %In order to ensure that an intermediate value is computed there is a {\tt force}
%% %function. 
%% The {\tt force} function takes a push or pull array and gives 
%% a block program resulting in a pull array. The block program writes the data 
%% contained in the array into shared memory and the return array symbolises 
%% reading from that memory.

%% %Hence, when working with local push and 
%% %pull arrays the associated programs are thread and block programs and when 
%% %working with global arrays the grid programs also come into play.

%% The push and pull arrays have different strengths. Pull arrays 
%% make it easy to implement functions such as {\tt map} and {\tt zipWith}. 
%% % and 
%% %also give map-fusion for free. A pull array is also easy to parallelise;  
%% %thread $i$ is in charge of computing element $i$.  
%% %The push arrays provide ways to compute arrays in a way so that thread $i$ 
%% %computes other elements than $i$. Actually, any thread be in charge of the 
%% %computation of any element of the result. 
%% Push arrays allow thread $i$ to compute array elements at indices other 
%% than $i$; this enables efficient array concatenation as well 
%% as the data scattering operations used later. For more information about 
%% push arrays, see reference \cite{PUSH}.

%Both push and pull arrays are easy to parallelise.

%In the current version of Obsidian\footnote{github.com/SvenssonJoel/Obsidian} 
%we are exploring ways to increase the scope of what kinds of compute kernels 
%we can describe and obtain efficient code for. One aspect of this exploration 
%is the addition of a new kind of array representation that we call 
%{\em push} arrays \cite{PUSH}. Previous versions of Obsidian 
%(described in \cite{JSLIC}) represented arrays as functions 
%from indices to values,we call these {\em pull} arrays. Another addition 
%is global arrays. Earlier versions of Obsidian has been focused on 
%expressing the kind of computations that takes place locally, in a block.

%The pull representation of arrays make it easy to describe 
%functions that take arrays apart and gives fusion of operations (for 
%example map-fusion) for free. However, Combining arrays, such as the 
%array concatenate operation, is not efficient using this array 
%representation. We have had problems in generating good CUDA code 
%for such functions. For example concatenation of pull arrays 
%leads to conditionals in the generated code of a kind that can have 
%negative performance impact on the GPUs we are targeting. 
%The implementation of the algorithm described in this paper 
%uses both the pull and the push arrays offered by Obsidian.

%The GPUs targeted by Obsidian has an hierarchical structure 
%in more senses than just its memory system. Threads are only 
%allowed to communicate using local shared memory in small groups 
%(thread blocks). 
%To match the hierarchal architecture of GPUs, Obsidian has local arrays 
%of type {\tt Pull a} and {\tt Push a}. These local arrays are used in the 
%implementation of programs that can be run by a block, in Obsidian these 
%programs are represented by a data type called {\tt BProgram}. There 
%are also global arrays types {\tt GlobPull a} and {\tt GlobPush a} that 
%are used in the implementation of programs executed by more than one MP, 
%these programs are represented by a {\tt GProgram} data type.   
%and global arrays of 
%types {\tt GlobPull a} and {\tt GlobPush a}.
%Global arrays are used when simple permutations of a large input array is 
%needed. Here it is important to that no communication needing 
%synchronisations occurs. Global arrays are also used as the inputs 
%and outputs of a top level Obsidian program (a program you want to generate runnable CUDA code from). In the cases where threads need to cooperate 
%on solving a task the local arrays should be used. 
\subsection{Obsidian programming example} 
\label{sec:sklansky}
\FloatBarrier

In the different variants of counting sort that are developed in this
paper, the prefix sum operation is the same. It is therefore natural
to use as an introductory example of how Obsidian programs are
written.

%We want to be able to sort large arrays and to do this we need an 
%implementation of prefix sum that is larger (on more elements) than what 
%fits in a block. 
We need to operate on arrays that are too long to fit in a single block.
%This large prefix sum is implemented as shown in
The resulting prefix sum is implemented as shown in reference 
\cite{LargeScan}, see figure \ref{fig:prefixsum}. There are three steps 
involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
the prefix sum of the auxiliary array is  computed. The prefix sums used 
in phase A and B of the 
figure can both be generated from the same Obsidian program. Finally the 
local prefix sums and the result of the auxiliary prefix sum are combined, 
forming a large prefix sum over the entire array.

\begin{figure}
\includegraphics[width=\linewidth]{./csort/prefixsum}
\caption{Illustration of how to combine many local prefix sum computations 
 into a large one. The algorithm has three steps: A compute local prefix sums, 
 B recursively compute prefix sums on the maximums of the local prefix sum 
 calculations, C distribute summed maximums over the local results.}
\label{fig:prefixsum}
\end{figure}

\begin{figure} 
\begin{center}
\includegraphics[width=\linewidth]{./csort/prefixzoom}
\caption{Close-up on the prefix sum and distribute operations used in Figure \ref{fig:prefixsum}.} 
%Zooms in on the shapes used in figure \ref{fig:prefixsum}. The left 
%shape symbolises a prefix sum and the right shows the distribute operation}
\label{fig:prefixzoom}
\end{center}
\end{figure}

The figure also shows what kernels we need to produce. First, something that 
performs a number of local prefix sums in parallel over the full length 
of an array is needed. The prefix sum algorithm that we implement here is 
based on divide and conquer and originates from Sklansky \cite{Sklansky}. 
% A kernel that performs the distribution (part C in the figure) is also needed.

The code below can be thought of as a program generator. Given an integer, $n$,
it gives a prefix computation for $2^n$ elements. Here we go directly to 
generating the prefix sums operations (with the operator + hard coded) rather 
than a higher order function that can be used to more generically.

\begin{small}
\begin{Verbatim}[samepage=true]
sklansky :: Int 
            -> Pull EWord32 
            -> BProgram (Pull EWord32)
sklansky 0 arr = return arr
sklansky n arr = do 
    let arr1 = binSplit (n-1) fan arr
    arr2 <- force arr1
    sklansky (n-1) arr2
\end{Verbatim}
\end{small}

%\begin{small}
%\begin{verbatim} 
%sklansky :: Scalar a => Int -> (Exp a -> Exp a -> Exp a)
%            -> Pull (Exp a) -> BProgram (Pull (Exp a))
%sklansky 0 op arr = return arr
%sklansky n op arr = do 
%    let arr1 = binSplit (n-1) (fan op) arr
%    arr2 <- force arr1
%    sklansky (n-1) op arr2
%\end{verbatim}
%\end{small}

The {\tt sklansky} code above is short but still contains much that needs 
explanation. First, the type specifies that the function takes a normal 
Haskell Int and an input array ({\tt Pull EWord32}). The result is a {\tt BProgram} that 
produces a result array. 
%The code above could be thought of as a program 
%generator for a supplied number,$n$, and operation (for example (+)); it gives 
%you a program that performs all prefix sums operation on an array. 
The function {\tt binSplit} is in the Obsidian library and is used 
to implement divide and conquer algorithms. This function splits an array 
recursively a given number of times and then applies some computation on all 
parts. In this case the {\tt fan} operation implemented below is used.

%To be able to implement prefix sum on the GPU we need to split the work 
%up into different phases. While computing a prefix sum in parallel , values 
%needs to be communicated between threads. This indicates that a kernel that 
%utilises local shared memory is a good idea. In CUDA the threads that can 
%communicate via shared memory all belong to the same block (group of threads). 
%These blocks can contain up to 1024 threads but often much fewer to 
%conserve resources and allow more blocks to run in parallel. In this example 
%a 256 threads (and elements) kernel is created. In the 
%benchmarks (section \ref{sec:Benchmarks}) prefix sum kernels of many 
%different sizes are used but all of these can be generated from the same 
%Obsidian description. In order to be able to implement a large prefix sum 
%the local kernel is also designed to output its maximal element to 
%a separate array. This separate array can be used to compute values to adjust 
%the local sums by. This follows the schema developed in \cite{LargeScan}.

%The local prefix sum is executed in parallel over the large array. 
%The large array is thus some multiple of the number of elements that the 
%local kernel can process. These sub results (both local scans and recursively 
%scaned block maximi) are then combined using another kernel called 
%{\tt distribute}.

%\begin{verbatim} 
%sklansky :: Scalar a => Int -> (Exp a -> Exp a -> Exp a)
%            -> Pull (Exp a) -> BProgram (Pull (Exp a))
%sklansky 0 op arr = return arr
%sklansky n op arr = do 
%    let arr1 = twoK (n-1) (fan op) arr
%    arr2 <- force arr1
%    sklansky (n-1) op arr2
%\end{verbatim}

\begin{small}
\begin{Verbatim}[samepage=true]
fan arr =  a1 `conc`  fmap (+ (last a1)) a2 
    where 
      (a1,a2) = halve arr
\end{Verbatim} 
\end{small}
%\begin{small}
%\begin{verbatim} 
%fan op arr =  a1 `conc`  fmap (op c) a2 
%    where 
%      (a1,a2) = halve arr
%      c = a1 ! (fromIntegral (len a1 - 1))
%\end{verbatim} 
%\end{small}

The {\tt fan} operation splits an array down the middle and then adds 
the element at the highest index of the first half to all elements of 
the second. Then the two parts are concatenated back together. 
%The {\tt sklansky} function uses the {\tt force} function that ensures 
%that the array is computed and the elements stored in memory. This can be 
%used to introduce parallelism but also, by leaving it out, to ensure fusion 
%of computations.

In order to obtain both the local prefix sum results and the maximum 
element of the block, a small wrapper function is created.

\begin{small}
\begin{Verbatim}[samepage=true]
sklanskyMax :: Int 
               -> Pull EWord32 
               -> BProgram (Pull EWord32, 
                            Pull EWord32)
sklanskyMax n arr = do
  res <- sklansky n arr
  return (res,singleton (last res))
\end{Verbatim}
\end{small} 
%\begin{small}
%\begin{verbatim}
%sklanskyMax :: Scalar a => Int -> (Exp a -> Exp a -> Exp a)
%     -> Pull (Exp a) -> BProgram (Pull (Exp a), Pull (Exp a))
%sklanskyMax n op arr = do
%  res <- sklansky n op arr
%  return (res,singleton (last res))
%\end{verbatim}
%\end{small}

All this wrapper does is compute the prefix sum followed by returning the  
result and a singleton array containing the maximum, {\tt last}, element.

%This function is currently the 
%target of much experimentation with regards to how exactly it should 
%be implemented and what exact type it should have.
This local prefix sum is run on sub-parts of the global array. The function
{\tt mapDist} splits up the array, runs a local computation on each part using
the MPs of the GPU; when it finishes, the result of that computation is an array
that remains distributed across the local memories of the GPU MPs.

\begin{small}
\begin{Verbatim}[samepage=true]
sklanskyG :: Int 
             -> GlobPull EWord32
             -> GProgram (GlobPush EWord32, 
                          GlobPush EWord32)
sklanskyG logbsize input =
  toGProgram (mapDist (sklanskyMax logbsize) 
                      (2^logbsize) 
                      input)
\end{Verbatim}  
\end{small}
%\begin{small}
%\begin{verbatim} 
%sklanskyG :: (Num (Exp a), Scalar a)
%             => Int
%             -> GlobPull (Exp a)
%             -> GProgram (GlobPush  (Exp a), GlobPush (Exp a))
%sklanskyG logbsize input = toGProgram $
%    mapDist (sklanskyMax logbsize (+)) (2^logbsize) input
%\end{verbatim}  
%\end{small}

In the {\tt sklanskyG} function the local prefix combinator implemented 
so far is turned into a global prefix sum computation. 
The {\tt toGProgram} function takes a function from blockId to 
a {\tt BProgram} and gives back a {\tt GProgram}, that is a program running 
on the full GPU (across many MPs).

The CUDA code generated from {\tt sklanskyG} is shown in figure
\ref{fig:genSklansky}. Writing that type of CUDA code by hand is both
tedious and error prone. The Obsidian code is both shorter, more
compositional and easier to read.

One kernel remains to be implemented, {\tt distribute}.

\begin{small}
\begin{Verbatim}[samepage=true]
distribute :: EWord32 
              -> GlobPull EWord32
              -> GlobPull EWord32 
              -> GlobPull EWord32
distribute bs maxs inputs = zipWithG (+) maxs' inputs 
  where 
    maxs' = repeat bs maxs
    repeat n = ixMap (\ix -> ix `div` n)
\end{Verbatim} 
\end{small}
%\begin{small}
%\begin{verbatim} 
%distribute :: Num a => Exp Word32 -> GlobPull a 
%              -> GlobPull a -> GlobPull a
%distribute bs maxs inputs = zipWithG (+) maxs' inputs 
%  where 
%    maxs' = repeat bs maxs
%    repeat n gp = ixMap (\ix -> ix `div` n) gp 
%\end{verbatim} 
%\end{small}
%repeat n (GlobPull ixf) = GlobPull $ \ix -> ixf (ix `div` n)

The {\tt distribute} kernel takes three inputs, a number, $n$,  and two global pull 
arrays. The elements of the first array are repeated $n$ times and then 
the result is element wise added to the second input array. The code generated 
from this kernel is shown in figure \ref{fig:genDistrib}. The 
{\tt distribute} kernel is used in the implementation of a large prefix sum 
(over more elements then what can fit in shared memory) from local prefix sum 
computations. This is indicated in figure~\ref{fig:prefixsum}.

\begin{figure} 
\newcommand{\myfontsize}{\fontsize{7}{9}\selectfont}
{\myfontsize
\begin{verbatim} 
__global__ void sklansky(uint32_t *input0,
                         uint32_t *output0,
                         uint32_t *output1){
  extern __shared__ unsigned char sbase[];
  ((uint32_t *)sbase)[tid] = 
    (((tid&0x1)<0x1) 
      ? input0[((bid*32)+((tid&0xFFFFFFFE)|(tid&0x1)))] 
      : (input0[((bid*32)+((tid&0xFFFFFFFE)|0x0))]+
         input0[((bid*32)+((tid&0xFFFFFFFE)|(tid&0x1)))]));
  __syncthreads();
  ((uint32_t *)(sbase + 128))[tid] = 
    (((tid&0x3)<0x2) 
      ? ((uint32_t *)sbase)[((tid&0xFFFFFFFC)|(tid&0x3))] 
      : (((uint32_t *)sbase)[((tid&0xFFFFFFFC)|0x1)]+
         ((uint32_t *)sbase)[((tid&0xFFFFFFFC)|(tid&0x3))]));
  __syncthreads();
  ((uint32_t *)sbase)[tid] = 
    (((tid&0x7)<0x4) 
      ? ((uint32_t *)(sbase+128))[((tid&0xFFFFFFF8)|(tid&0x7))] 
      : (((uint32_t *)(sbase+128))[((tid&0xFFFFFFF8)|0x3)]+
        ((uint32_t *)(sbase+128))[((tid&0xFFFFFFF8)|(tid&0x7))]));
  __syncthreads();
  ((uint32_t *)(sbase + 128))[tid] = 
    (((tid&0xF)<0x8) 
      ? ((uint32_t *)sbase)[((tid&0xFFFFFFF0)|(tid&0xF))] 
      : (((uint32_t *)sbase)[((tid&0xFFFFFFF0)|0x7)]+
         ((uint32_t *)sbase)[((tid&0xFFFFFFF0)|(tid&0xF))]));
  __syncthreads();
  ((uint32_t *)sbase)[tid] = 
    ((tid<0x10) 
      ? ((uint32_t *)(sbase+128))[tid] 
      : (((uint32_t *)(sbase+128))[0xF]+
         ((uint32_t *)(sbase+128))[tid]));
  __syncthreads();
  ((uint32_t *)(sbase + 128))[tid] = ((uint32_t *)sbase)[tid];
  if (tid<1){
    ((uint32_t *)(sbase + 256))[tid] = ((uint32_t *)sbase)[31];
  }
  __syncthreads();
  output0[((bid*32)+tid)] = ((uint32_t *)(sbase+128))[tid];
  if (tid<1){
    output1[(bid+tid)] = ((uint32_t *)(sbase+256))[tid];
  }
}
\end{verbatim} 
}
\caption{CUDA code generated from the Obsidian {\tt sklanskyG} prefix sum program. 
 This example shows the 32 element version of the prefix sum computation}
\label{fig:genSklansky} 
\end{figure}

\begin{figure} 
\begin{small}
\begin{verbatim} 
__global__ void distribute(uint32_t s0, 
                           uint32_t *input1,
                           uint32_t *input2,
                           uint32_t *output0){
  output0[((bid*bDim)+tid)] = 
      (input1[(((bid*bDim)+tid)/s0)]+
       input2[((bid*bDim)+tid)]);

}
\end{verbatim} 
\end{small}
\caption{CUDA code generated from the Obsidian {\tt distribute} function.}
\label{fig:genDistrib} 
\end{figure}

\subsection{A note about generated code}

All generated code shown in this paper has been edited slightly by hand. This 
is partly to increase readability but also to conserve space. The changes 
that have been made are entirely cosmetic. Line breaks have been inserted in too
long lines. All occurrences of {\tt blockIdx.x}, {\tt blockDim.x} and 
{\tt threadIdx.x} have been replaced with {\tt bid}, {\tt bDim} and {\tt tid}.
Some decimal constants have been changed to hexadecimal.
No other changes have been made.

%\FloatBarrier

### 2013-09-30 src2/csort/benchmarks.tex
apa

\section{Performance evaluation} 
\label{sec:Benchmarks}

\FloatBarrier

This section presents the experimental results. Our implementations of
counting sort are compared to the implementation of sorting from the
Thrust library \cite{THRUST}. Thrust is a C++ library, developed by
NVIDIA, which provides abstractions similar to the standard template
library, but targets the GPU. The reason we have chosen to compare with
Thrust is that it has similar goals to Obsidian: to provide high-level 
constructs while generating efficient GPU code.

All the data being sorted in the measurements are 32 bit unsigned
integer valued keys.
The convention used in the charts is that the y-axis represents time in ms. 
The different experiments are called {\em sorted} for sorting of an 
already sorted array, and {\em unique} for sorting of randomised arrays where each 
element occurs exactly once. There are also a number of experiments called 
T$x$ for a number $x$;  the arrays used in these experiments contain
elements from an interval of numbers (0 to $2^x-1$).

Figures \ref{fig:chart1} and \ref{fig:chart3} show the results of
comparing our full counting sort to Thrust sort and occurrence sort
 to Thrust's {\tt sort} and {\tt unique} functions.
%% Figure \ref{fig:chart1}  show the results of
%% comparing our full counting sort to Thrust sort and our duplicate
%% removing sort to Thrust sort followed by unique.
Our implementation of
counting sort consistently outperforms Thrust, except for small ranges
of values. The reason counting sort performance degrades is that the
number of threads used in the reconstruct step is proportional to the
range, and therefore at small ranges the execution becomes more
sequential.

We compare occurrence sort to sort followed by unique from
the Thrust library because that seems to be the current best practice
\cite{REMOVEDUPS}. Our implementation is always at least twice as fast
but in many cases it outperforms Thrust by more than a factor of four.
Our method is clearly beneficial for removing duplicates.

%We are comparing 
%our implementations of counting sort to the sorting algorithm present 
%in the Thrust library. 
%The counting sort that removes duplicates is compared to Thrust's sort 
%followed by unique (unique removes all but one element in groups 
%of consecutively equal elements). These results can be found in 
%figures \ref{fig:chart1} and \ref{fig:chart3}.

%We also compare our two variants of counting sort directly to each
We also compare occurrence sort and counting sort to each 
other.  The result of this is shown in figure \ref{fig:chart2}. We
can see that the occurrence sort is typically a factor of
two faster than the regular sorting algorithm. The speedup comes from
the fact that the {\tt occurs} kernel doesn't need to perform any kind
of synchronisation, whereas the {\tt histogram} kernel needs to use
atomic operations.

The charts included in this paper show the performance when sorting eight
respectively 32 million ($2^{23}$ resp. $2^{25}$) elements. However, we 
ran all benchmarks with a number of elements, $N$, equal to $2^{n}$ for $n$ 
between 20 and 25. We found that the interesting comparison is not in what 
happens as $N$ grows larger but rather what happens as we vary the ranges 
of the elements contained in the arrays. 
%% when the ranges of the 
%% elements contained in the array is increased.

All timing was performed on a system with an NVIDIA GTX670 GPU.

\begin{figure*}
\includegraphics[width=\linewidth]{./csort/chart1}
\caption{On the left, counting sort is compared to the Thrust library. 
On the right, occurrence sort is compared to the {\tt sort} followed by {\tt unique} 
function in Thrust. Both charts show running time for 8 Million elements.}
%The left chart shows running time of full counting sort compared to 
%the sort function in the Thrust library. The right chart shows running time 
%of counting sort that removes duplicates compared to the sort followed by 
%unique function of the Thrust library. 8 Million elements where used in these 
%experiments.}
\label{fig:chart1}
\end{figure*}

\begin{figure*}
\includegraphics[width=\linewidth]{./csort/chart3}
\caption{This chart shows same comparison as figure \ref{fig:chart1}, but for 
32 Million elements.}
%The left chart shows running time of full counting sort compared to 
%the sort function in the Thrust library. The right chart shows running time 
%of counting sort that removes duplicates compared to the sort followed by 
%unique function of the Thrust library. 32 Million elements where used in 
%these experiments.}
\label{fig:chart3}
\end{figure*}

\begin{figure*}
\includegraphics[width=\linewidth]{./csort/chart2}
\caption{These two charts compares occurrence sort to  counting sort. 
The left chart is for 8 Million elements and the right chart is 
for 32 Million elements.}
\label{fig:chart2}
\end{figure*}

\subsection{CUDA framework for the benchmarks}

To obtain the timing measurements above, a 
framework\footnote{Source code needed to repeat our experiments is available at \url{www.cse.chalmers.se/~joels/csort.html}} 
written in CUDA was used. 
This framework looks very similar to the CUDA code in figure \ref{fig:cudaCoord}. 
The main difference is that a number of kernels are run in sequence, as 
shown in figure \ref{fig:csortframe}. The timing is performed on the GPU 
computation only, as illustrated by figure \ref{fig:timing}.

\begin{figure} 
\begin{small}
\begin{verbatim} 
float total_time;

cudaEventRecord(start,0);
for (int i = 0; i < 1000; ++i) { 
    // perform counting sort. 
}
cudaEventRecord(stop,0);
cudaEventSynchronize(stop);

cudaEventElapsedTime(&total_time,start,stop);
\end{verbatim}
\end{small}
\caption{The timing methodology used in performance experiments. A CUDA event 
is recorded before and after the execution on the GPU. Their respective recorded times
are then compared.}
\label{fig:timing}
\end{figure}

\begin{figure} 
\begin{small}
\begin{verbatim} 
histogram<<<NB,BS,0>>>(dinput,dhistoutput);

scan256<<<NB,BS,2052>>>
    (dhistoutput,dscanoutput+1,dmaxs+1);

//dmaxs needs to be scaned (dmaxs2 is ignored) 
scan256<<<SMALL,BS,2052>>>
    (dmaxs+1,dmaxs+1,dmaxs1+1);

scan16<<<1,SMALL,132>>>
    (dmaxs1+1,dmaxs1+1,dmaxs2);

// distribute can be in-place regarding dresult 
// since it reads and writes in the exact 
// same location (per thread)
distribute<<<SMALL,BS,0>>>
    (BS,dmaxs1,dmaxs+1,dmaxs+1);
distribute<<<NB,BS,0>>>
    (BS,dmaxs,dscanoutput+1,dscanoutput+1);

reconstruct<<<NB,BS,0>>>
    (dscanoutput,dresult);   
\end{verbatim}
\end{small}
\caption{The kernel launch sequence used in the timing of counting sort for 
1 Million elements. Any other input data size would vary only some of the prefix 
sum (scan in the code) sizes.}
\label{fig:csortframe}
\end{figure}

% \FloatBarrier

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/chartObs2M}
%\caption{Another chart example!}
%\label{fig:chart2}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/chartObs4M}
%\caption{Another chart example!}
%\label{fig:chart2}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/chartObs8M}
%\caption{Another chart example!}
%\label{fig:chart2}
%\end{figure}

%\begin{figure}[ht]
%\begin{minipage}[b]{0.5\linewidth}
%\centering
%\includegraphics[width=\textwidth]{./img/chartObs4M}
%\caption{This is to the left}
%\label{fig:figure1}
%\end{minipage}
%\hspace{0.5cm}
%\begin{minipage}[b]{0.5\linewidth}
%\centering
%\includegraphics[width=\textwidth]{./img/chartObs8M}
%\caption{This is to the right}
%\label{fig:figure2}
%\end{minipage}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/chart3}
%\caption{Shows execution of the counting sort algorithm on 4 million 
%         random elements.}
%\label{fig:chart3}
%\end{figure}

%Figures \ref{fig:chart1},\ref{fig:chart2} and \ref{fig:chart3} show how our algorithm 
%compares to a thrust implementation for 4,8 and 16 million elements in 
%various ranges.

% >>> What can be said more about the numbers ???

### 2013-09-30 src2/csort/discussion.tex
apa

\section {Discussion}

It is instructive to compare the code of the two variants of counting
sort. The standard algorithm requires mutations and its parallel
implementation is rather complicated. The occurrence sort variation
avoids key synchronisations, making the algorithm
data-parallel and this gives it a performance advantage. However, in
applying the trick to remove synchronisation we have changed the
semantics of the function. In this case, it is still useful, albeit less
generally applicable. The idea of removing synchronisation is a
general and important idea for speeding up algorithms, but it is not
clear how general the method used in this paper is.  We currently do
not know of any other algorithms which could benefit from a similar
trick.

If the desired effect is to sort and remove duplicates 
(for an example of this see \cite{REMOVEDUPS}) then 
there is a clear benefit to using occurrence sort
over the {\tt sort} combined with {\tt unique} method.

In this paper, we are entirely focused on the implementation and
performance of kernels. Reference \cite{CSORT} goes into the details
of performing data transfers in parallel with ongoing computations. We
have not addressed this issue at all in our implementations. This is
also the reason why we haven't performed any benchmarks against the algorithm in
\cite{CSORT}, because it would compare very different things. As a
piece of future work, it would be interesting to add the capability of
streaming data from the main memory to the GPU to our algorithms.

In counting sort, an auxiliary array holding the count of how many times 
each element occurs in the input array, the histogram array, is created. 
The size of this array is 
based on the range of elements occurring in the input. This means that 
more memory is needed for larger ranges and may imply that counting sort 
is suitable only up to some point. A related algorithm, known as radix 
sort \cite{Knuth}, addresses this issue by using multiple counting sort 
like passes, one for each digit position and thus limits this auxiliary 
memory usage to an array of size related to the radix (number of unique 
digits) used in the representation of the values being sorted. During 
our exploration of the counting sort algorithm on a GPU, we were not much 
concerned by memory usage; all data and auxiliary 
arrays fit easily in the device memory. The charts (\ref{fig:chart1}, \ref{fig:chart3} and \ref{fig:chart2}) indicate that our implementation offers the most 
benefit in the ranges labelled as $T10$ to $T17$; these are in fact relatively 
small ranges (1024 to 131072 unique values). Our current hypothesis 
is that the decrease in benefit of counting sort from range $T18$ and onward
is due to more and more complicated memory access patterns in the 
reconstruction phase. However, more in-depth profiling is needed to 
understand where the decrease in benefit comes from.
%%  However, we postulate that 
%% the decrease in benefit following $T18$ and onward are due to more complicated
%% memory access patterns during reconstruction

Obsidian is work in progress but this paper shows a step forward in its
capabilities. New kinds of kernels that operate on global arrays can be 
implemented. We also added atomic operations to the language. However, 
when dealing with atomic operations, we express the Obsidian program 
at a very low level. 
%% The exact implementation of global arrays is not yet fully 
%% decided upon. For example, the global arrays shown in this paper does 
%% not have a length associated with them. The size of a global array is rather
%% decided as a multiple of the size the kernel operates on in the case of 
%% local computations. In the case of computations that do not use any local
%% memory the length of global arrays is more free. These sizes are set when 
%% launching the kernels. 
%% Sometimes we made use of very low-level capabilities of Obsidian; when dealing 
%%with atomic operations. 
One task for future work is to find some way to raise 
the level of abstraction in that area. Another future direction in exploring 
the counting sort algorithm is to consider further optimisations.
%% Another future direction in exploring 
%% the counting sort algorithm specifically is to see what other kind of  
%% optimisations we can perform. 
One possibility is to work with sequentiality 
(more work per thread). Continuing exploration in that direction could lead 
to further improvements of Obsidian when it comes to low-level control.

In the current version of Obsidian, we need to write some CUDA 
code by hand in order to implement the algorithm we describe. 
In the future we want to be able to do all coding without leaving 
Haskell.

%% There is already an example of using Obsidian as the code 
%% generating back end of such an EDSL. This EDSL is called PEGGY and 
%% is described in \cite{cole2012beauty}.

%% This case study turned out an interesting exercise in very low-level 
%% programming in Obsidian. Having the ability to drop down to a CUDA-like 
%% language allows us to make use of atomic operations...

### 2013-09-30 src2/csort/intro.tex
apa

\section{Introduction}

Sorting is an ever important, ever fascinating field of computer
science. The introduction of GPUs has introduced new challenges in
designing fast sorting algorithms.

This paper focuses on counting sort together with a variation which
removes duplicate elements. We call this variation
{\em occurrence sort}. Counting sort is a non-comparing sort
suitable for parallel implementation. The occurrence sort variation 
presented here is interesting because it seems to be a particularly
good fit for executing on the GPU. It has a very natural functional,
data-parallel implementation. We believe we are the first to study
this variation in the literature.

These algorithms are explored in Obsidian \cite{JSLIC,PUSH}, a domain
specific language targeting GPU programming. The goal of Obsidian is
to strike a balance between high-level constructs and low-level control. 
We want to provide the programmer with a set of tools for low-level 
experimentation with the details that influence performance when 
implementing GPU kernels. The counting sort
case study shows that Obsidian generates competitive kernels with
relatively little programmer effort. That Obsidian is an embedded language 
allows us to rapidly experiment with the addition of features and with varying 
programming idioms. The version used in this case study
adds global arrays and atomic instructions to Obsidian, see section 
\ref{sec:OBSHist}.

The contributions of this paper are:
\begin{itemize}
\item We provide measurements (section \ref{sec:Benchmarks}) showing
  that counting sort is a competitive algorithm for sorting keys on the GPU,
  outperforming the sorting implementation in the library
  Thrust\cite{THRUST} in many cases.
\item Occurrence sort is shown to be particularly
  suitable for implementing on the GPU (section \ref{sec:occur}) and
  performs well (section \ref{sec:Benchmarks}).
\item The Obsidian implementation of the two sorting algorithms is
  detailed along with the generated CUDA (sections
  \ref{sec:Obsidian} and \ref{sec:parallel}).
\end{itemize}

\subsection{Related work}

Sorting has applications in the computer graphics field \cite{sintorn}. 
Example instances of sorting and duplicate element removal in computer graphics 
can be identified in references \cite{Olsson} and \cite{Karras}. Another example 
of where sorting and the removal of duplicates have applications is 
in databases \cite{REMOVEDUPS}.

%% Sorting on the GPU is interesting both for computer graphics
%% applications \cite{SINTORN} and for using the GPU to offload the
%% processor, for example in database applications \cite{REMOVEDUPS}.

The paper \cite{CSORT} implements counting sort in CUDA, and also optimisations 
that overlap computations with transferring data to and from the GPU.
Our work differs by being implemented in a high-level embedded
language and also by implementation of the occurrence sort variant of counting sort. We have not been concerned with
trying to overlap computations and data transfer, but have instead
focused solely on computations within the GPU.

Obsidian is a high-level embedded language for GPU
programming. Other such approaches include Accelerate
\cite{Acc} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
higher level approach compared to Obsidian and abstract more from GPU
details. Another example of providing a high-level interface to
programming the GPU is the C++ library Thrust \cite{THRUST}.
%\cite{FELDSPAR}

%% With the Obsidian language we raise the level of abstraction for 
%% GPU kernel implementation. In order to generate CUDA code that performs well, 
%% Obsidian tries to strike a balance between high level constructs and low level 
%% control.

%% This paper describes a case study. We implement a sorting algorithm known 
%% as counting sort. As an extra bonus we explore a variant of counting sort 
%% that can be given a nice purely functional description and lends itself well 
%% to parallelisation. The experimental results in section \ref{sec:Benchmarks} 
%% compares the performance of the counting sort algorithm we implement here to 
%% the NVIDIA Thrust \cite{THRUST} library. The Obsidian generated kernels perform
%% very well at relatively light work effort.

%% Counting sort places new demands on Obsidian and points out areas of 
%% improvement. Some left to do, some already implemented and used here. 
%% First, in order to implement counting sort we added atomic operations to 
%% Obsidian. However, currently you can only use them by very low level features 
%% of Obsidian. This is shown in section \ref{sec:parallel}. This paper also 
%% shows for the first time the new global arrays and what they can be used for.

%Vague intro ideas.
%\begin{itemize}
%  \item Quickly, what is our idea. 
%  \item Why is this sorting useful.
%        (Or is i just a fun experiment.) 
%  \item GPUs are pretty cool 
%  \item This is an Obsidian case-study. 
%  \item Sneak in ref to Feldspar.
%  
%\end{itemize}

%\cite{CSORT} 
%\cite{FELDSPAR}

### 2013-09-30 src2/csort/paper.tex
apa

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}

\usepackage{graphicx}
\usepackage{placeins} 
\usepackage{url}
\usepackage[T1]{fontenc}
\usepackage{tikz}
\usepackage{fancyvrb}

\begin{document}

\exclusivelicense
\conferenceinfo{FHPC~'13}{September 23 2013, Boston, MA, USA} 
\copyrightyear{2013} 
\copyrightdata{978-1-4503-2381-9/13/09} 
\doi{2502323.2502325}

\titlebanner{DRAFT}        % These are ignored unless
\preprintfooter{Counting Sort}   % 'preprint' option specified.

%\title{Efficient Counting Sort Implementations using an Embedded GPU Programming Language}
%\title{Implementing Counting Sort for GPUs using an Embedded Language}
\title{Counting and Occurrence Sort for GPUs using an Embedded Language}
%\subtitle{Subtitle Text, if any}

\authorinfo{Josef Svenningsson \and Bo Joel Svensson \and Mary Sheeran}
           {Dept, of Computer Science and Engineering \\ 
            Chalmers University of Technology}
           {\{josefs, joels, ms\}@chalmers.se}

\maketitle

\begin{abstract}
This paper investigates two sorting algorithms: counting sort and a
variation, occurrence sort,  which also removes duplicate elements, 
and examines their suitability for running on the GPU. The duplicate 
removing variation
turns out to have a natural functional, data-parallel implementation
which makes it particularly interesting for GPUs.

The algorithms are implemented in Obsidian, a high-level domain
specific language for GPU programming.

Measurements show that our implementations in many cases outperform
the sorting algorithm provided by the library Thrust. Furthermore, occurrence
sort is another factor of two faster than
ordinary counting sort. We conclude that counting sort is an important
contender when considering sorting algorithms for the GPU, and that 
occurrence sort is highly preferable when applicable. We also show 
that Obsidian can produce very competitive code.
\end{abstract}

%% \category{CR-number}{subcategory}{third-level}
%% <<<<<<< HEAD

%% %% \terms
%% %% term1, term2

%% %% \keywords
%% %% keyword1, keyword2
%% \category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
%% \category{D.3.4}{Programming Languages}{Processors}[Code generation]

%% %\terms
%% %term1, term2
%% \terms
%% Languages, Performance
%% =======
\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
\category{D.3.4}{Programming Languages}{Processors}[Code generation]
%% \terms
%% term1, term2
%% >>>>>>> 3b812338c4f7649982b17fb45df51f13d4285484

%\keywords
%keyword1, keyword2
\keywords
%% <<<<<<< HEAD
%% Data parallelism, array programming, embedded language
%% =======
Sorting, embedded language

%% >>>>>>> 3b812338c4f7649982b17fb45df51f13d4285484

\input{intro.tex}
\input{Algorithm.tex}
\input{GPUs.tex}
\input{Obsidian.tex} 
\input{ObsCudaAlgo.tex}
\input{benchmarks.tex}
\input{discussion.tex}

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

\acks
This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the Swedish Research Council.

\bibliographystyle{abbrvnat}
%% \bibliography{paper}

\begin{thebibliography}{18}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[wik()]{wikipedia}
Wikipedia article: Counting sort.
\newblock \url{http://en.wikipedia.org/wiki/Counting_sort}.

\bibitem[Chakravarty et~al.(2011)Chakravarty, Keller, Lee, McDonell, and
  Grover]{Acc}
M.~M. Chakravarty, G.~Keller, S.~Lee, T.~L. McDonell, and V.~Grover.
\newblock {Accelerating Haskell array codes with multicore GPUs}.
\newblock In \emph{Proc. of the sixth workshop on Declarative aspects of
  multicore programming}, DAMP '11. ACM, 2011.

\bibitem[Claessen et~al.(2012)Claessen, Sheeran, and Svensson]{PUSH}
K.~Claessen, M.~Sheeran, and B.~J. Svensson.
\newblock {E}xpressive {A}rray {C}onstructs in an {E}mbedded {GPU} {K}ernel
  {P}rogramming {L}anguage.
\newblock In \emph{Proceedings of the 7th workshop on Declarative aspects and
  applications of multicore programming}, DAMP '12, 2012.

\bibitem[Elliott et~al.(2003)Elliott, Finne, and de~Moor]{ELLIJFP}
C.~Elliott, S.~Finne, and O.~de~Moor.
\newblock Compiling embedded languages.
\newblock \emph{Journal of Functional Programming}, 13\penalty0 (2), 2003.
\newblock URL \url{http://conal.net/papers/jfp-saig/}.

\bibitem[Harris et~al.(2007)Harris, Sengupta, and Owens]{LargeScan}
M.~Harris, S.~Sengupta, and J.~D. Owens.
\newblock Parallel {P}refix {S}um ({S}can) with {CUDA}.
\newblock In H.~Nguyen, editor, \emph{GPU Gems 3}. Addison Wesley, 2007.

\bibitem[{J}oel {S}vensson(2011)]{JSLIC}
{J}oel {S}vensson.
\newblock {O}bsidian: {GPU} {K}ernel {P}rogramming in {H}askell.
\newblock Technical Report 77L, {C}omputer {S}cience and {E}nginering,
  {C}halmers {U}niversity of {T}echnology, {G}othenburg, 2011.
\newblock Thesis for the degree of Licentiate of Philosophy.

\bibitem[Karras(2012)]{Karras}
T.~Karras.
\newblock {Maximizing Parallelism in the Construction of BVHs, Octrees, and k-d
  Trees}.
\newblock In \emph{Proc. of the Fourth ACM SIGGRAPH / Eurographics conference
  on High-Performance Graphics}, EGGH-HPG'12. Eurographics Association, 2012.

\bibitem[Knuth(1998)]{Knuth}
D.~E. Knuth.
\newblock \emph{The art of computer programming, volume 3: (2nd ed.) sorting
  and searching}.
\newblock Addison Wesley Longman Publishing Co., Inc., Redwood City, CA, USA,
  1998.
\newblock ISBN 0-201-89685-0.

\bibitem[Kolonias et~al.(2011)Kolonias, Voyiatzis, Goulas, and Housos]{CSORT}
V.~Kolonias, A.~G. Voyiatzis, G.~Goulas, and E.~Housos.
\newblock {Design and implementation of an efficient integer count sort in CUDA
  GPUs}.
\newblock \emph{{Concurrency And Computing: Practice And Experiance}},
  23\penalty0 (18), Dec. 2011.

\bibitem[Krueger et~al.(2011)Krueger, Grund, Jaeckel, Zeier, and
  Plattner]{REMOVEDUPS}
J.~Krueger, M.~Grund, I.~Jaeckel, A.~Zeier, and H.~Plattner.
\newblock {Applicability of GPU Computing for Efficient Merge in In-Memory
  Databases}.
\newblock In \emph{The Second International Workshop on Accelerating Data
  Management Systems using Modern Processor and Storage Architectures (ADMS
  11)}, 2011.
\newblock \url{http://www.adms-conf.org/p19-KRUEGER.pdf}.

\bibitem[Kulkarni and Newton(2013)]{FPCDSL}
A.~Kulkarni and R.~R. Newton.
\newblock {E}mbrace, {D}efend, {E}xtend: {A} {M}ethodology for {E}mbedding
  {P}reexisting {DSL}s, 2013.
\newblock Functional Programming Concepts in Domain-Specific Languages
  (FPCDSL'13).

\bibitem[Mainland and Morrisett(2010)]{NIKOLA}
G.~Mainland and G.~Morrisett.
\newblock {Nikola: embedding compiled {GPU} functions in Haskell}.
\newblock \emph{SIGPLAN Not.}, 45\penalty0 (11), 2010.

\bibitem[{NVIDIA}({\natexlab{a}})]{CUDA}
{NVIDIA}.
\newblock {CUDA C Programming Guide}, {\natexlab{a}}.
\newblock URL
  \url{http://docs.nvidia.com/cuda/cuda-c-programming-guide/index.html}.

\bibitem[{NVIDIA}({\natexlab{b}})]{KEPLER}
{NVIDIA}.
\newblock {NVIDIAs Next Generation CUDA Compute Architecture: Kepler GK110},
  {\natexlab{b}}.
\newblock URL
  \url{http://www.nvidia.com/content/PDF/kepler/NVIDIA-Kepler-GK110-Architecture-Whitepaper.pdf}.

\bibitem[{NVIDIA}({\natexlab{c}})]{THRUST}
{NVIDIA}.
\newblock {NVIDIA Thrust Library}, {\natexlab{c}}.
\newblock URL \url{https://developer.nvidia.com/thrust}.

\bibitem[Olsson et~al.(2012)Olsson, Billeter, and Assarsson]{Olsson}
O.~Olsson, M.~Billeter, and U.~Assarsson.
\newblock Clustered deferred and forward shading.
\newblock In \emph{Proc. of the Fourth ACM SIGGRAPH / Eurographics conference
  on High-Performance Graphics}, EGGH-HPG'12. Eurographics Association, 2012.

\bibitem[Sintorn and Assarsson(2008)]{sintorn}
E.~Sintorn and U.~Assarsson.
\newblock Fast parallel {GPU}-sorting using a hybrid algorithm.
\newblock \emph{Journal of Parallel and Distributed Computing}, 68\penalty0
  (10), 2008.

\bibitem[Sklansky(1960)]{Sklansky}
J.~Sklansky.
\newblock {Conditional Sum Addition Logic}.
\newblock \emph{Trans. IRE}, EC-9\penalty0 (2), June 1960.

\end{thebibliography}

%% % The bibliography should be embedded for final submission.

%% \begin{thebibliography}{}
%% \softraggedright

%% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%% P. Q. Smith, and X. Y. Jones. ...reference text...

%% \end{thebibliography}

\end{document}

### 2013-09-30 src2/embarbb/abstract.tex
apa

Nowadays, performance in processors is increased by adding more cores or
wider vector units, or by combining accelerators like GPUs and traditional cores on a chip. 
Programming for these diverse architectures is a challenge. We would 
like to exploit all the resources at hand without putting too much burden 
on the programmer. Ideally, the programmer should be presented with a machine 
model abstracted from the specific number of cores, SIMD width or the existence of 
a GPU or not. Intel's Array Building Blocks (ArBB) is a system that takes on these 
challenges. ArBB is a language for data parallel and nested data parallel 
programming, embedded in C++. By offering a retargetable dynamic compilation 
framework, it provides vectorisation and threading to programmers without 
the need to write highly architecture specific code. We aim to bring the 
same benefits to the Haskell programmer by implementing a Haskell frontend
(embedding) of the ArBB system. We call this embedding EmbArBB. 
We use standard Haskell embedded language procedures to provide an interface 
to the ArBB functionality in Haskell. EmbArBB is work in progress and 
does not currently support all of the ArBB functionality.
Some small programming examples illustrate how the Haskell 
embedding is used to write programs. ArBB code is short and to the point 
in both C++ and Haskell.
Matrix multiplication has been benchmarked in sequential C++, ArBB in C++, EmbArBB and the Repa library.
The C++ and the Haskell embeddings have almost identical performance, showing 
that the Haskell embedding does not impose any large extra overheads. 
Two image processing algorithms have also been benchmarked against Repa. In these benchmarks
at least, EmbArBB performance
is much better than that of the Repa library, indicating that building on ArBB may be a cheap and easy approach to exploiting data parallelism in Haskell.

### 2013-09-30 src2/embarbb/conclusion.tex
apa

\section{Conclusion} 
We have shown that Intel's Array Building Blocks (ArBB) provides
an interface that is well suited to functional programming. 
The programs are quite close to mathematical specifications, in the style of NESL~\cite{NESL}. We have only just completed the embedding of the part of ArBB that deals with nested vectors, and we need to tackle many more case studies.
In our case studies using dense vectors, ArBB seems to do a good job of efficiently using parallel
hardware resources -- both cores and vector units.
By
embedding ArBB, we can, with little implementation effort, provide
quite an attractive data parallel programming language in Haskell.
Our benchmarks are preliminary and small, but they show very good performance.
Once we have completed the ArBB embedding, we will have an interesting platform on which to
experiment with and develop new programming idioms that exploit the fact
that we have a data parallel programming language embedded in an
expressive, strongly typed host language.
We feel that work in this area (as distinct from implementation methods)
is overdue. We would be happy to receive suggestions
for interesting case studies or collaborations.

### 2013-09-30 src2/embarbb/future.tex
apa

\section{Future Work} 
\label{sec:FutureWork}

%The Haskell - EmbArBB interface needs to be improved. 
%Currently there is a 
%transfer of data into ArBB before a function is executed and the results 
%are directly transferred back into Haskell.
%This is not ideal when several ArBB functions are composed, as this
%will lead to unnecessary copying of data into and out of ArBB.
%One possible solution would be to have the {\tt ArBB} monad keep track of what arrays have been bound 
%to ArBB variables and stored into ArBB memory.
%Another problem with the interface is the way {\tt map} works. Currently 
%{\tt map} takes an identifier pointing out a captured function. These
%captured functions are maintained by the {\tt ArBB} monad, which means that 
%either a function that maps some other function over data needs to be defined 
%in the {\tt ArBB} monad, or it needs to be passed the function as an argument 
%at capture time. Neither of those alternatives is desirable and new solutions
%will have to be found.

%One thought is to 
%also let the {\tt ArBB} monad keep track of what arrays have been bound 
%to ArBB variables and stored into ArBB memory. Currently there is a 
%transfer of data into ArBB before a function is executed and the results 
%are directly transferred back into Haskell. This is limiting in case 
%the programmer wants to execute several ArBB functions in sequence, each 
%on the result of the previous function. In this scenario data will be 
%copied back and forwards needlessly between function executions in the current 
%implementation.

The C++ embedding of ArBB allows for dense containers of structs in some cases. 
The operations on vectors supplied by the ArBB virtual machine are exclusively 
over vectors of scalar types. So the C++ embedding must be performing a AOS to SOA 
(Array of Struct to Struct of Array) transformation. The Haskell embedding 
does not implement any similar transformation. This is an important addition 
that would for example make implementing functions on complex numbers easier.

We stress that this paper presents first steps in the implementation of 
EmbArBB. Our benchmarks, while promising, are very limited. We must devote 
effort to developing a suite of interesting, larger data parallel programs 
for use in benchmarking EmbArBB. It is particularly important to explore the 
nested vectors, and the effects of the limitation to one level of nesting 
in ArBB. Those parts of ArBB that support nested vectors seem to be less well 
developed than those supporting dense vectors, as evidenced by the sample 
applications distributed with ArBB, none of which uses nesting. We expect 
ArBB to become more complete, and perhaps we will be able to contribute 
interesting examples both in the C++ and Haskell embeddings.

Having exercised EmbArBB more thoroughly, we will assess the results of 
the benchmarking and experiments with programming idioms, and decide on future
research directions. We expect to focus on ways to provide users with an interface that is more 
functional in style than the current C++ oriented one. 
%ArBB has nested vectors that are used to implement nested data parallelism, that 
%is data parallelism with irregularly shaped data, in the style of
%NESL~\cite{NESL}. The Haskell embedding has so 
%far been focused entirely on the dense functionality of ArBB. 
%The next step will be to find ways to embed the nested operations too.
%Looking on the 
%nested operations and seeing how these can be embedded as well is next in line.

%GPUs are efficient on flat data parallel workloads, and approaches
%to GPU programming often offer only flat data parallelism.
%Although there is some promising recent work on nested data parallelism
%in a functional setting
%on GPUs~\cite{NestedGPU}, this area is generally less well explored.
%We are enticed by the idea of
%using EmbArBB as a 
%platform for exploration of GPU programming and nested data-parallelism.
%Implementing a GPU backend to EmbArBB would require applying 
%GPU specific optimisations to the EmbArBB AST (expression datatype). 
%The question of whether to perform the necessary optimisations
%at the EmbArBB level, or later, inside ArBB, is an interesting
%one that we would like to explore.

%However 
%if implementing a GPU backend on the EmbArBB level is the right thing to do 
%or if rather it should be implemented as part of the ArBB system? There 
%could be a point to explore in the setting of EmbArBB the space of different 
%optimisations and transformations needed to run the functions on a GPU before 
%doing it in the more complicated (perhaps) setting of full ArBB.

### 2013-09-30 src2/embarbb/implementation.tex
apa

\section{Implementation} 
EmbArBB is a deeply embedded language. There is a constructor for 
every operation that ArBB can perform. A deep embedding is useful when 
there is need to apply optimisations or transformations to the AST before 
executing the operations. However, EmbArBB currently relies entirely on 
ArBB to perform optimisations to the program. The only optimisation 
performed on the EmbArBB side is sharing detection. Detecting sharing 
reduces the number of calls into the ArBB C library. Should we add a 
GPU backend to EmbArBB, further GPU specific optimisations of the AST 
will likely become necessary.

\subsection{Vectors} 
EmbArBB has support for one, two and three dimensional dense vectors,
represented by a datatype called {\tt DVector}.

\begin{verbatim}
import qualified Data.Vector.Storable as V

data DVector d a 
  = Vector {vectorData  :: V.Vector a, 
            vectorShape :: d} 
\end{verbatim}

The payload data in a {\tt DVector} is stored in a vector from the 
{\tt Data.Vector} library. There is also a {\tt d} parameter that specifies 
the shape of the DVector.

%The {\tt DVector} type has two parameters. The first parameter {\tt d} 
%is a type level representation of the dimensionality of a vector. 
%The dimensionality is encoded using a datatype:

The {\tt d} parameter is used for a type level representation of the 
dimensionality of a vector, as in the Repa library~\cite{REPA}. The 
dimensionality is encoded using the following types together with 
the {\tt Int} type.

\begin{verbatim} 
data a :. b = a :. b  
infixl :.

data Z = Z 
\end{verbatim}

%There are also easy to use synonyms for the allowed cases:
The dimensions zero to three are represented as follows:

%% MS : Shouldn't there be a ref. to Repa (or to whatever paper first
%% used these kinds of types for arrays) here??  I put a suggestion
%% but it should be changed to the correct ref.

\begin{verbatim} 
type Dim0 = Z             
type Dim1 = Dim0 :. Int
type Dim2 = Dim1 :. Int 
type Dim3 = Dim2 :. Int
\end{verbatim}

For example in the reduction functions provided by EmbArBB the 
output vector is of a dimensionality one less than the input vector 
used. Below is the type of {\tt addReduce} to illustrate this.

\begin{verbatim} 
addReduce :: Num a 
           => Exp USize 
           -> Exp (DVector (t:.Int) a) 
           -> Exp (DVector t a)
\end{verbatim}

A somewhat unfortunate side effect of this is that currently 
EmbArBB has two kinds of scalars,
{\tt Exp (DVector Dim0 a)} and {\tt Exp a}. The result of reducing 
a one-dimensional vector is a zero-dimensional vector. An operation 
called {\tt index0} converts from zero-dimensional vectors 
to scalars.

Irregular container, in ArBB called Nested vectors, are represented in 
EmbArBB by the type {\tt NVector}. Currently there are no versions of the 
{\tt copyIn}, {\tt copyOut} or {\tt new} functions for nested vectors. 
The programmer must transfer dense vectors into the ArBB heap and then 
apply nesting to them as part of the computation to perform thereupon. This 
means that having a concrete representation for a NVector is currently 
not useful. In the hope of being able to implement some of the transfer 
functions, even without direct support from the ArBB VM, we represent 
NVectors as a vector of dense data together with a vector containing 
segment lengths.

\begin{verbatim} 
data NVector a = 
       NVector { nVectorData     :: V.Vector a
               , nVectorNesting  :: V.Vector USize}
\end{verbatim}

As part of the interface between Haskell and EmbArBB there are also mutable 
vectors, of a type called {\tt BEDVector}. These are vectors that reside in 
the ArBB heap and are represented only by an integer identifier with which 
the actual data can be accessed from ArBB. These are used to store the inputs 
and outputs of calls to {\tt execute} on a captured function.

\begin{verbatim}
data BEDVector d a = 
       BEDVector { beDVectorID :: Integer
                  , beDVectorShape :: d}  
\end{verbatim}

New {\tt BEVector}s are created using the function {\tt new}.
The function {\tt copyIn} copies a {\tt DVector} from Haskell to the ArBB heap 
and there is a function {\tt copyOut} to retrieve data from ArBB.

\subsection{The language} 
The EmbArBB language is implemented as a set of library functions, operating on an expression datatype:

\begin{verbatim} 
data Expr = Lit Literal
          | Var Variable

          | Index0 Expr 
          | ResIndex Expr Int

          | Call (R GenRecord) [Expr]  
          | Map  (R GenRecord) [Expr]

          | While ([Expr] -> Expr)  
                  ([Expr] -> [Expr])  
                  [Expr]

          | If Expr Expr Expr 
          | Op Op [Expr]   
\end{verbatim} 
Most of the ArBB functionality is taken care of by the {\tt Op} constructor. 
The datatype {\tt Op} used to represent operations has over 120 
constructors, so only a selection is shown in figure~\ref{fig:OPS}. 
Having all these 120+ operations taken care of by one constructor in 
the expression datatype simplifies the implementation of the backend, 
since all of these operations are handled in a very similar way. The few 
remaining special ArBB capabilities such as loops, and function mapping 
and calling are handled by their own cases, discussed below. There are 
also some implementation specific details that result in their own 
constructors in the {\tt Expr} type, namely {\tt Index0} for turning a 
zero dimensional DVector into a scalar, and {\tt resIndex} that helps with 
implementing operations that have more than one result. An example of an 
such an operation is {\tt SortRank}, which returns both the sorted result 
of an input vector and a vector of indices that specifies a permutation 
that would have sorted the input vector.

The {\tt Expr} type also has constructors for the {\em call} and {\em map} 
functionality. To call a function means to apply it to input data. The 
{\tt map} operation specifies elementwise application of a function over 
vectors, like NESL's {\em apply-to-each}~\cite{NESL}, so it corresponds 
to Haskell's {\tt map} and {\tt zipWith}. Both {\tt Call} and {\tt Map} 
take a {\tt(R GenRecord)}. This is inspired by the delayed expressions 
that enable the implementation of {\tt vapply} in Nikola~\cite{NIKOLA}. 
The {\tt R} is the reification monad used to create DAGs (directed acyclic 
graphs) from embedded language functions. These DAGs are part of a 
{\tt GenRecord} that contains all information that the ArBB code generator 
needs to generate the ArBB function.

The {\tt While} loop is represented using higher order abstract syntax, 
that is using functions to represent the condition and body. The state 
is represented by a list of expressions; this needs to be generalised 
somewhat in order to support loops with general tuples in the state. 
Something more structured than a list is needed for this.

The expression data type used in EmbArBB is untyped (not using a GADT). 
A typed interface to the language is supplied using the same phantom types 
method as used in {\em pan}~\cite{ELLIJFP} and many other Haskell embedded 
languages since then. The choice to use an untyped {\tt Expr} datatype was 
based on the wish to keep the backend (ArBB code generation) as simple as 
possible.

%Currently there 
%are very few {\tt expr} to {\tt expr} transformations taking place before the code 
%generation so the benefits of having the typed representation would be small.
% ** Have commented this out for now as it may cause confusion, esp if
% the implementation changes before the workshop

\begin{verbatim}
-- Phantom types
type Exp a = E Expr 
\end{verbatim}

As an example, the operation {\tt addReduce}, which reduces a vector across a specified dimension, is implemented as follows in the EmbArBB library:

\begin{verbatim}
addReduce :: Num a 
           => Exp USize 
           -> Exp (DVector (t:.Int) a) 
           -> Exp (DVector t a) 
addReduce (E lev) (E vec) = 
  E $ Op AddReduce [vec,lev]
\end{verbatim}

%The {\tt addReduce} function takes a {\tt DVector} of any dimensionality greater than 0,
%thats what the (():.t) in its type specifies, and a integer ({\tt USize}) that specifies 
%the level to reduce across. The result is a {\tt DVector} of dimensionality one less 
%than the input vector.

\begin{figure} 
\begin{minipage}{\linewidth}
\begin{Verbatim}[frame=single]
data Op =  
   -- elementwise and scalar
     Add | Sub | Mul | Div | Max | Min 
   | Sin | Cos | Exp 
   ...

   -- operations on vectors 
   | Gather | Scatter | Shuffle | Unshuffle
   | RepeatRow | RepeatCol | RepeatPage 
   | Rotate | Reverse | Length | Sort 
   | AddReduce | AddScan | AddMerge
   ... 
\end{Verbatim}
\end{minipage}
\caption{  ArBB scalar, elementwise and vector operations, which are handled by the {\tt Op} constructor in the {\tt Expr} datatype.
          This is just a selection from the more than 120 different operations ArBB provides.}
\label{fig:OPS}
\end{figure}

\FloatBarrier

\subsection{Interfacing with Haskell and Code generation}

The interface between ArBB and Haskell consists of the {\tt DVector}, {\tt NVector}, {\tt BEDVector} and {\tt BEScalar} types, 
the {\tt capture}, and {\tt execute} functions, and the ArBB monad with its {\tt withArBB} 
``run''-function. This section describes what happens when the programmer captures 
an embedded language function, and when {\tt execute} is called on a captured
function.

Capture and execution of functions takes place in the {\tt ArBB} monad, which
manages state of type {\tt ArBBState}:

\begin{verbatim}
data ArBBState = 
  ArBBState 
    { arbbFunMap :: Map.Map Integer 
                            ArBBFun        
    , arbbVarMap :: Map.Map Integer 
                            VM.Variable 
    , arbbUnique :: Integer }

type ArBBFun = (VM.ConvFunction, [Type], [Type])
\end{verbatim}

This state contains a map from function names to ArBB functions and their input and 
output types. The {\tt VM.ConvFunction} is how ArBB functions are represented by 
the ArBB-VM bindings. There is a map from vector and scalar IDs to their 
corresponding ArBB variables. The last item in {\tt ArBBState} is an Integer 
that is used to generate new function names and variable IDs as the programmer 
captures more functions or creates new arrays on the ArBB heap. 
Now, the {\tt ArBB} monad is defined as:

\begin{verbatim}
type ArBB a = StateT ArBBState VM.EmitArbb a  
\end{verbatim}

{\tt VM.EmitArbb} is also a concept from the virtual machine bindings. It manages 
low-level functions (the {\tt VM.ConvFunction} functions) and implements an interface to 
the low level ArBB-VM API. 
{\tt EmitArBB} is the ArBB code generating monad from the arbb-vm bindings.
For more or less everything that can be done with the arbb-vm bindings, there is a function
of the form 
\begin{verbatim}
f :: arg1 -> ... argn -> EmitArBB out
\end{verbatim}
For example, for generating an operation node (such as +) in the ArBB IR, there is a
function of type
\begin{verbatim}
op_ :: Opcode 
     -> [Variable] 
     -> [Variable] 
     -> EmitArbb ()
\end{verbatim}
\noindent
while the function for generating a while loop in the ArBB IR has type
\begin{verbatim}
while_ :: (EmitArbb Variable) 
        -> EmitArbb a 
        -> EmitArbb a
\end{verbatim}
The details of the translation are omitted for brevity, as the approach is
standard.

When a function {\tt f} of type 
\begin{verbatim}
Exp tin1 -> ... -> Exp tinN -> Exp tout
\end{verbatim}
is captured, it is 
first applied to expressions that represent variable names. For each of the 
inputs, {\tt (tin1,..,tinN}), a variable is created. The result is an expression 
(or expressions) representing the function {\tt f}. On this expression, sharing 
detection is performed, and a directed acyclic graph (DAG) is created. 
The method of sharing detection used is based on the StableNames method~\cite{Gill}.

Then, the code generation is implemented using a very 
direct approach; no extra optimisations or transformations are applied. 
This is a reasonable choice, since the whole point of ArBB is that the 
built-in JIT compiler knows and performs architecture specific optimisations. 
Sharing detection on the Haskell side makes sense because code generation for 
each node in the DAG results in at least two calls into the ArBB-VM API, which 
means going through the FFI and incurring the associated cost. Detecting the
sharing already in the host language should give a smaller workload for the 
ArBB JIT compiler, thus reducing the time spent on JITing.
It remains to be seen how important JIT cost will be in practice, however,
as we expect it to be amortised over a large number of executions of 
the JITed code. We will need to conduct experiments with a suite of larger 
examples in order to decide if sharing detection on the Haskell side is 
worthwhile.

%Now, in a system such as this the time it 
%takes to JIT is of lesser importance, the JIT cost will likely be amortised over 
%a large number of executions of the JITed code.

Applying {\tt capture} to a function gives an object of type

\begin{verbatim}
type FunctionID = Integer

data Function i o = Function FunctionID
\end{verbatim}

The {\tt i} and {\tt o} parameters to {\tt Function} represent the input and output 
types of the captured function. As an example, capturing 
\begin{verbatim}
f :: Exp (DVector Dim1 Word32) 
   -> Exp (DVector Dim1 Word32)
\end{verbatim}
results in an object of type
\begin{verbatim}
Function (BEDVector Dim1 Word32) 
         (BEDVector Dim1 Word32)
\end{verbatim}
This is just phantom types placed over a function name that is just a {\tt String},
but it does offer a typed interface for the {\tt capture} and {\tt execute} functions.

The {\tt execute} function that launches a captured function takes a {\tt Function i o} 
object, inputs of type {\tt i} and outputs of type {\tt o}.
The function name is looked up in the {\tt ArBB} environment (the monad). The inputs
and outputs are also looked up in the environment and then the function 
is executed.

\begin{verbatim} 
execute :: (VariableList a, VariableList b) 
         => Function a b -> a -> b -> ArBB ()       
execute (Function fid) a b = 
  do 
    (ArBBState mf mv _) <- S.get 
    case Map.lookup fid  mf of 
      Nothing -> error "execute: Invalid function" 
      (Just (f,tins,touts)) -> 
        do 
          ins <- vlist a 
          outs <- vlist b

          liftVM$ VM.execute_ f outs ins 
\end{verbatim}

%The {\tt arbbUp}, {\tt arbbDown} and {\tt arbbAlloc} functions are provided by the 
%{\tt ArBBin} and {\tt ArBBOut} classes. Output storage needs to be allocated before the
%function is executed. The {\tt arbbAlloc b} uses the sizes and shapes of the output 
%to allocate proper storage for the results in ArBB memory.

The {\tt vlist} function goes through the heterogeneous list of inputs or outputs 
and looks up each of the elements in the {\tt arbbVarMap}; the result is a Haskell 
list of {\tt VM.Variable}. The function {\tt VM.execute\_} is part of the virtual 
machine API bindings, and corresponds directly to a C function in that library.

%EmbArBB is built on top of the Haskell ArBB-VM bindings~

%\begin{figure}
%\begin{center}
%\includegraphics[width=5cm]{./img/stack}
%\end{center}
%\caption{EmbArBB is implemented on top of Haskell bindings to the ArBB virtual machine C API}
%\label{fig:stack}
%\end{figure}

\FloatBarrier
%\subsection{ArBB Code generation}
%Describe what happens on ``Capture'' and ``Execute''

%\subsection{summary}

### 2013-09-30 src2/embarbb/introduction.tex
apa

\section{Introduction}

% --------------------------------------------------------------------------
%*** something about multi-core CPUs with vector units (intro).
Modern hardware architectures provide much parallelism in the form of
multi- and many-core processors and heterogeneous combinations of the two.
We would like to exploit this parallelism without over-burdening the programmer, and without sacrificing too much performance.
Ideally, the programmer should be encouraged to write the code in a way that is not too closely linked to the precise combination of cores, SIMD vector processing power and GPU
that is currently targeted.
It should also be possible to easily retarget the code to new architectures.
As processor architectures develop and are combined at an ever increasing pace, these requirements place considerable demands on programming languages
and libraries for parallel programming.

Intel's Array Building Blocks (ArBB) is one approach to the problem of how to increase productivity for programmers who need to exploit hardware parallelism, but for whom very low level parallel programming (as typically found in High Performance Computing) is too difficult or time consuming~\cite{ARBB2011}. ArBB is a retargettable dynamic compilation framework that is provided as an embedded language in C++. It aims to give the programmer access to data and thread parallelism, while protecting him from common pitfalls such as race conditions and deadlock.

Many of the ideas in ArBB feel familiar to functional programmers.
For example, when referring to collection data types, reference~\cite{ARBB2011} states that ``Expressions over collections always act as if a completely new collection were created and assignment between collections always behaves as if the entire input collection were copied.''
This rather functional view of data is accompanied by extensive optimisation that (as the paper states) removes almost all copying.
A key to the approach is the use of standard patterns of computation such as scan and reduce -- familiar as higher order functions in a functional setting.
The system tries to fuse away intermediate data-structures
when these patterns are composed.
The patterns are deterministic, so that problems with race conditions are avoided by design. Programs are stated to be short and close to the mathematical specification, and that, again, is a familiar mantra. The system allows C++ users to construct code objects called closures.
The first time the {\em call} operation is applied to
a particular function, ArBB captures the sequence of operations
over ArBB types and creates an intermediate representation. 
This representation is then compiled to
vector machine language, which is
cached for later use.  The
machine code is also invoked (in parallel on multiple
cores, if appropriate). The next time the function
is called, ArBB reuses the cached machine
code rather than regenerating it.
Although dynamically typed closures are available, ArBB is generally
statically typed.

%% Note: I felt that this para was too technical to go right at the beginning
%% Maybe want to add back the mentions of Sandy Bridge and MMX etc.
%% to a later discussion.

%Multi-, many-core and heterogeneous processors with both on the same chip
%(such as for example Sandy Bridge CPUs with a built in GPU) are 
%now common. Programming for a range of slightly different set-ups 
%of number of cores, SIMD-width, existence of GPU or not can be impractical 
%if a target architecture specific APIs and methods needs to be used. No doubt
%in the cases where every last drop of performance needs to be squeezed out 
%of a system, architecture specific methods will be used. Maybe in the 
%form of CUDA for a GPU or MMX,SSE,AVX intrinsics in the CPU case. When Good 
%alround performance across a line of different architectures is desired 
%a retargettable just-in-time (JIT) compiled approach should be a good idea. 
%Intel Array Building Blocks (ArBB) is such a system.

%ArBB is programming language that aspires to
%make it easier to program modern multi-core CPUs. ArBB gives the 
%programmer a set of operations that when used results in automatic 
%vectorisation and threading. 
%Amongst these operations are scans (prefix sums)  
%and reductions over vectors, elementwise operations and more. 
%The compilation 
%process also tries to fuse away intermediate datastructures~\cite{ARBB2011}.

ArBB is an embedded language, implemented
as a library in the host language C++. The embedding makes 
heavy use of the C++ template machinery. There is also a lower level 
interface to ArBB, the ArBB Virtual Machine C API~\cite{arbbvm}. The low level C API is not
intended for use directly by applications programmers, but rather as a base 
for implementations of alternative ArBB frontends (embeddings)~\cite{ARBB2011}.
The ArBB VM implements the compilation and parallelisation.
Reference~\cite{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
steps towards using ArBB as a backend to the embedded language Accelerate~\cite{accelerate}.

The design of ArBB seems particularly well suited to a functional setting.
In this paper, we present work in progress on embedding ArBB in Haskell. We call the embedding EmbArBB.
It is our hope that using 
Haskell as a host language will be just as user friendly and efficient as 
the C++ version. EmbArBB does not yet include all of the functionality of ArBB,
and this is discussed further in section~\ref{sec:FutureWork}.
However,  our first benchmarks show that the performance of EmbArBB is very similar to that of ArBB in C++. This bodes well. 
%We will consider using EmbArBB in the next instance of a course at Chalmers on parallel functional programming.

EmbArBB is available at GitHub as 
\newline
\url{github.com/svenssonjoel/EmbArBB}.

### 2013-09-30 src2/embarbb/listfun.tex
apa

% -- Get the number of columns
% getNCols :: Exp (DVector (t:.Int:.Int) a) 
%           -> Exp USize

\begin{figure}
\begin{small}
\begin{Verbatim}
-- Scatter values into a vector 
-- resolves collisions by adding elements
addMerge :: Exp (DVector (t:.Int) USize) -- indices
         -> Exp USize                    -- res length
         -> Exp (DVector (t:.Int) a)     -- src
         -> Exp (DVector (t:.Int) a)

-- Reduce a vector using addition
addReduce :: Num a 
          => Exp USize   -- rows, cols or pages
          -> Exp (DVector (t:.Int) a) 
          -> Exp (DVector t a)

-- Segmented version of addReduce
addReduceSeg :: Num a 
             => Exp (NVector a)  -- nested input
             -> Exp (DVector Dim1 a)

-- Create a nested vector from a dense
applyNesting :: Exp USize  -- lengths or offsets
             -> Exp (DVector Dim1 USize) -- nesting
             -> Exp (DVector Dim1 a)     
             -> Exp (NVector a)

-- Create a vector with a constant value
constVector :: Exp USize  -- length
            -> Exp a      -- value
            -> Exp (DVector Dim1 a)

-- Extract a column from a 2D vector
extractCol :: Exp USize              -- col. index
           -> Exp (DVector Dim2 a)  
           -> Exp (Vector a)

-- Fill a portion of a vector with a constant 
fill :: Exp a                -- fill value 
     -> Exp USize            -- start 
     -> Exp USize            -- end 
     -> Exp (DVector Dim1 a) -- dst
     -> Exp (DVector Dim1 a)

-- Gather elements from a vectors 
gather1D :: Exp (DVector Dim1 USize)  -- indices
         -> Exp a                     -- default
         -> Exp (DVector Dim1 a)      -- values
         -> Exp (DVector Dim1 a)

-- Get the number of rows 
getNRows :: Exp (DVector (t:.Int:.Int) a) 
         -> Exp USize

-- Turn a zero dimensional vector into a scalar 
index0 :: Exp (DVector Z a) -> Exp a

-- Create a 2D vector by repeating a 1D vector
repeatRow :: Exp USize            -- #repetitions
          -> Exp (DVector Dim1 a) -- row
          -> Exp (DVector Dim2 a)

-- Replace one column in a 2D vector 
replaceCol :: Exp USize            -- col
           -> Exp (DVector Dim1 a) -- new values
           -> Exp (DVector Dim2 a) 
           -> Exp (DVector Dim2 a)

-- Convert every element of a vector to USize
vecToUSize :: Exp (Vector a) 
           -> Exp (Vector USize)

\end{Verbatim}
\end{small}
\caption{A list of EmbArBB functions that are used in the examples with 
         short descriptions} 
\label{fig:listoffun}
\end{figure}

\begin{figure}
\fbox{
\parbox{8cm}{
\begin{itemize}
\item {\tt ISize} is an integer type used to specify for example offsets.
\item {\tt USize} is an unsigned integer type used for lengths or indices.
\item {\tt Boolean} replaces the {\tt Bool} type for EmbArBB programs. The 
reason for this is that ArBB internally represents booleans as 8bit words.
while the {\tt Storable} instance for {\tt Bool} does not. 
\item {\tt DVector} is the type of regular shaped vectors.
\item {\tt NVector} is the type of irregularly shaped vectors.
\end{itemize}
}
}
\caption{A list of EmbArBB types with short descriptions} 
\label{fig:listoftyp}
\end{figure}

### 2013-09-30 src2/embarbb/paper.tex
apa

\documentclass{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

%\usepackage[none]{hyphenat}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{url}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{fancyvrb}

\begin{document}
\conferenceinfo{FHPC'12,} {September 15, 2012, Copenhagen, Denmark.}
\CopyrightYear{2012}
\copyrightdata{978-1-4503-1577-7/12/09}

\titlebanner{}        % These are ignored unless
\preprintfooter{}   % 'preprint' option specified.

% \title{}
\title{Parallel Programming in Haskell Almost for Free} 
\subtitle{an embedding of Intel's Array Building Blocks}

\authorinfo{Bo Joel Svensson}
           {Chalmers University of Technology}
           {joels@chalmers.se}
\authorinfo{Mary Sheeran}
           {Chalmers University of Technology}
           {ms@chalmers.se}

\maketitle

\begin{abstract}
\input{abstract.tex}
\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
\category{D.3.4}{Programming Languages}{Processors}[Code generation]

\terms
Languages, Performance

\keywords
Data parallelism, array programming, dynamic compilation, embedded language

%\input{outline.tex}

\input{introduction.tex}
\input{related.tex} 
\input{programming.tex}
\input{implementation.tex}
\input{results.tex}

%\input{discussion}
\input{future.tex}
\input{conclusion.tex}

%\cite{joelryan}
%\cite{arbbpb}
%\cite{arbbcpp}
%\cite{arbbvm}
%\cite{ARBB2011}
%\cite{REPA}
%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

\acks

%** Note. Joel should thank Intel. We should ack funding from VR and SSF.
Svensson was first introduced to ArBB while on a three month internship at Intel
during 2011. Thanks go to Ryan R. Newton for inspiring supervision 
during that internship.
%while 
%working on Haskell bindings to ArBB at Intel as well as trying to understand 
%the intricate details of the inside of Data.Array.Accelerate.

This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the
Swedish Research Council.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright
\bibitem[Axelsson et~al.(2010)Axelsson, Claessen, D{\'e}vai, Horv{\'a}th,
  Keijzer, Lyckeg{\aa}rd, Persson, Sheeran, Svenningsson, and
  Vajda]{FELDSPAR2010}
E.~Axelsson, K.~Claessen, G.~D{\'e}vai, Z.~Horv{\'a}th, K.~Keijzer,
  B.~Lyckeg{\aa}rd, A.~Persson, M.~Sheeran, J.~Svenningsson, and A.~Vajda.
\newblock Feldspar: {A} domain specific language for digital signal processing
  algorithms.
\newblock In \emph{8th ACM/IEEE International Conference on Formal Methods and
  Models for Codesign (MEMOCODE 2010)}, pages 169--178. IEEE Computer Society,
  2010.

\bibitem[Blelloch(1996)]{NESL}
G.~Blelloch.
\newblock Programming {P}arallel {A}lgorithms.
\newblock \emph{Communications of the ACM}, 39\penalty0 (3), 1996.

\bibitem[Chakravarty et~al.(2011)Chakravarty, Keller, Lee, McDonell, and
  Grover]{accelerate}
M.~M. Chakravarty, G.~Keller, S.~Lee, T.~L. McDonell, and V.~Grover.
\newblock {A}ccelerating {H}askell {A}rray {C}odes with {M}ulticore {GPU}s.
\newblock In \emph{Proceedings of the sixth workshop on Declarative aspects of
  multicore programming}, DAMP '11, pages 3--14, New York, NY, USA, 2011. ACM.
\newblock ISBN 978-1-4503-0486-3.
%\newblock \doi{10.1145/1926354.1926358}.
\newblock URL \url{http://doi.acm.org/10.1145/1926354.1926358}.

\bibitem[Elliott et~al.(2003)Elliott, Finne, and de~Moor]{ELLIJFP}
C.~Elliott, S.~Finne, and O.~de~Moor.
\newblock Compiling embedded languages.
\newblock \emph{Journal of Functional Programming}, 13\penalty0 (2), 2003.
\newblock URL \url{http://conal.net/papers/jfp-saig/}.

\bibitem[Gill(2009)]{Gill}
A.~Gill.
\newblock Type-{S}afe {O}bservable {S}haring in {H}askell.
\newblock In \emph{Proceedings of the 2009 ACM SIGPLAN Haskell Symposium},
  09/2009 2009.
\newblock URL
  \url{http://www.ittc.ku.edu/csdl/fpg/sites/default/files/Gill-09-TypeSafeRei%
fication.pdf}.

\bibitem[Hughes and Sheeran(2012)]{TFPIEPFP}
J.~Hughes and M.~Sheeran.
\newblock Teaching parallel functional programming at {C}halmers.
\newblock In \emph{presentation at Trends in Functional Programming in
  Education, Workshop associated with Conf. on Trends in Functional
  Programming, St. Andrews}, 2012.

\bibitem[Intel()]{arbbvm}
Intel.
\newblock {I}ntel(r) {A}rray {B}uilding {B}locks {V}irtual {M}achine
  {S}pecification.
\newblock \url{http://software.intel.com/sites/whatif/arbb/arbb_vm.pdf}.

\bibitem[Keller et~al.(2010)Keller, Chakravarty, Leshchinskiy, Peyton~Jones,
  and Lippmeier]{REPA}
G.~Keller, M.~M. Chakravarty, R.~Leshchinskiy, S.~Peyton~Jones, and
  B.~Lippmeier.
\newblock Regular, shape-polymorphic, parallel arrays in haskell.
\newblock In \emph{Proceedings of the 15th ACM SIGPLAN international conference
  on Functional programming}, ICFP '10, pages 261--272, New York, NY, USA,
  2010. ACM.
\newblock ISBN 978-1-60558-794-3.
%\newblock \doi{10.1145/1863543.1863582}.
\newblock URL \url{http://doi.acm.org/10.1145/1863543.1863582}.

\bibitem[Lippmeier et~al.(2012)Lippmeier, Chakravarty, Keller, Leshchinskiy,
  and Jones]{DPH}
B.~Lippmeier, M.~M.~T. Chakravarty, G.~Keller, R.~Leshchinskiy, and S.~P.
  Jones.
\newblock {W}ork {E}fficient {H}igher-{O}rder {V}ectorisation.
\newblock \url{http://www.cse.unsw.edu.au/~chak/papers/replicate-tr.pdf}, 2012.
\newblock ICFP'12.

\bibitem[Mainland and {M}orrisett(2010)]{NIKOLA}
G.~Mainland and G.~{M}orrisett.
\newblock Nikola: {E}mbedding {C}ompiled {GPU} {F}unctions in {H}askell.
\newblock In \emph{Proceedings of the third ACM Haskell symposium}, pages
  67--78. ACM, 2010.
\newblock ISBN 978-1-4503-0252-4.
\newblock URL \url{http://doi.acm.org/10.1145/1863523.1863533}.

\bibitem[Newburn et~al.(2011)Newburn, So, Liu, McCool, Ghuloum, Toit, Wang, Du,
  Chen, Wu, Guo, Liu, and Zhang]{ARBB2011}
C.~J. Newburn, B.~So, Z.~Liu, M.~McCool, A.~Ghuloum, S.~D. Toit, Z.~G. Wang,
  Z.~H. Du, Y.~Chen, G.~Wu, P.~Guo, Z.~Liu, and D.~Zhang.
\newblock Intel's array building blocks: A retargetable, dynamic compiler and
  embedded language.
\newblock In \emph{Proceedings of the 2011 9th Annual IEEE/ACM International
  Symposium on Code Generation and Optimization}, CGO '11, pages 224--235,
  Washington, DC, USA, 2011. IEEE Computer Society.
\newblock ISBN 978-1-61284-356-8.

\bibitem[Sheeran(2004)]{LavaMultipliers}
M.~Sheeran.
\newblock Generating fast multipliers using clever circuits.
\newblock In \emph{Int. Conf. on Formal Methods in Computer Aided Design
  (FMCAD)}, volume 3312 of \emph{LNCS}, pages 6--20, 2004.

\bibitem[Sheeran(2011)]{SheeranJFP}
M.~Sheeran.
\newblock Functional and dynamic programming in the design of parallel prefix
  networks.
\newblock \emph{J. Funct. Program.}, 21\penalty0 (1):\penalty0 59--114, 2011.

\bibitem[Svensson and Newton(2011)]{joelryan}
B.~J. Svensson and R.~Newton.
\newblock {P}rogramming {F}uture {P}arallel {A}rchitectures with {H}askell and
  {A}r{BB}.
\newblock \url{http://faspp.ac.upc.edu/faspp11/pdf/faspp11-final12.pdf}, 2011.
\newblock Presented at the workshop: {F}uture {A}rchitectural {S}upport for
  {P}arallel {P}rogramming (FASPP), in conjuction with ISCA '11.

\end{thebibliography}
%\bibliography{paper}

\end{document}

### 2013-09-30 src2/embarbb/programming.tex
apa

\section{Programming in ArBB}
\label{sec:ProgrammingArBB}

What the authors of ArBB call {\em latent parallelism} is expressed both
by using operations like scan and reduce (or fold) over vectors and by
mapping functions over collection data-types.
ArBB provides both dense and nested vectors. 
%The nested vectors permit 
%nested data parallelism (and are one of the attractions of ArBB).
% but we have not yet included them in our embedding. Thus, we consider only dense vectors here.
Dense vectors can be one, two or three dimensional; they correspond
to ordinary arrays.

%The vectors 
%in ArBB can be one, two or three dimensional dense vectors (in other words arrays)
%or nested vectors that allow the programmer to utilise nested data parallelism. 
%A nested vector is represented by an array of elements and an array holding a 
%nesting descriptor. 
% BJS: I have seen this elsewhere but cannot find it! (wanted to \cite something)
%The ArBB system offers the programmer {\em dense} and {\em nested} vectors 
%and operations on these. A dense vector (or simply put, an array) can 
%be one, two or three dimensional. A nested vector is represented by an 
%array of elements and an array holding a nesting descriptor. Nested vectors 
%can be used to express nested data-parallelism.

Matrix-vector multiplication can be expressed in ArBB and C++ like this: 
\begin{Verbatim}[samepage=true] 
void arbb_matrix_vector(const dense<f32, 2>& a, 
                        const dense<f32>& x,
                        dense<f32>& b) {
  b = add_reduce(a * repeat_row(x, a.num_rows()));
}
\end{Verbatim} 
\noindent
This function takes a dense matrix (a two dimensional array), {\tt a}, and a dense vector, {\tt x},
and returns a dense vector {\tt b}.
It
uses {\tt add\_reduce} and {\tt repeat\_row}, which are built in ArBB 
operations. Note that we are not provided with a general reduction operator
that takes a function as parameter,
but rather with specific instances.
%The resulting code is automatically vectorised and threaded.

The matrix-vector multiplication code above is very concise, but some further
steps are necessary before it can be run on real data. The code 
below shows how to set up a 4x4 scaling matrix and a vector to 
multiply by that matrix using ArBB. This entails allocating 
memory on the ArBB heap and copying data into it, reminiscent of the copying of
data from
host to 
device that is common in general purpose GPU (GPGPU) programming .

%[samepage=true] 
\begin{Verbatim} 
int main(void)
{
  float matrix[4][4] = {{2.0,0.0,0.0,0.0},
                        {0.0,2.0,0.0,0.0},
                        {0.0,0.0,2.0,0.0},
                        {0.0,0.0,0.0,2.0}};
  float vector[4] = {1.0,2.0,3.0,4.0};

  // set up the matrix in ArBB 
  dense<f32, 2> a(4, 4);
  range<f32> write_a = a.write_only_range();
  float* a_ = &write_a[0];
  memcpy(a_,matrix,16*sizeof(float));

  // set up the vector in ArBB 
  dense<f32> x(4);
  range<f32> write_x = x.write_only_range();
  float* x_ = &write_x[0];
  memcpy(x_, vector,4*sizeof(float));

  // Room for the result
  dense<f32> arbb_b(4);
  const_range<f32> read_arbb_b;

  call(arbb_matrix_vector)(a, x, arbb_b);
  read_arbb_b = arbb_b.read_only_range();

  for (int i = 0; i < 4; ++i) {
    printf("%f ", read_arbb_b[i]);
  }

  return 0;
}
\end{Verbatim}

After setting up and copying the data into ArBB, the function {\tt arbb\_matrix\_vector} 
is {\em called} using the {\tt call} command. At this point, a number of
things happen. If the function is called for the first time, it is 
{\em captured}. This means that the function is run as a C++ function, which produces 
an intermediate representation (IR) of the computation (which is standard in such embeddings).
For example, if the function being captured contains a normal C++ {\em for loop}, it will be unrolled, much in the same way as recursion in a
Haskell embedded DSL results in unrolled code. After the function is captured, 
the IR is further optimised and finally executed. Compilation and capture 
only occur the first time a function is called. The compilation procedure
is outlined in~\cite{ARBB2011}.

Nested vectors in ArBB allow the creation of an array of dense vectors of varying length -- giving the ability to express less regular parallelism. There is a limit, though, in that only one level of such nesting is allowed. Section~\ref{sec:smvm} shows a small example that uses this nestedness in EmbArBB to implement sparse matrix vector multiplication. The program in ArBB itself is very similar. We note, though, that one of the samples distributed with ArBB is also sparse matrix vector multiplication, but implemented using dense vectors and a function that can take sections of a vector. We will need to experiment further with nestedness and when it should be used.

%In this document we describe early stages of work with a Haskell Embedding 
%of ArBB. We call this embedded language EmbArBB. It is our hope that using 
%Haskell as a host language will be just as user friendly and efficient as 
%the C++ version. Since this is work-in-progress there are places where 
%EmbArBB is lacking functionality that the C++ embedding has. This is the 
%case with nested vectors for example, more details in the Future Work section.

%\FloatBarrier
\section{Programming in EmbArBB} 
\label{sec:Programming}

%** Note. The intro should mention that the current version of EmbArBB
%is available on github and say how to fetch it, I suppose.

In EmbArBB, ArBB functions are expressed as Haskell functions on expressions
(abstract syntax trees). This is standard Haskell embedded language procedure. 
A function that adds a constant to all elements of a vector is

\begin{verbatim}
addconst :: Num a 
          => Exp a 
          -> Exp (DVector Dim1 a) 
          -> Exp (DVector Dim1 a) 
addconst s v = v + ss 
    where 
      ss = constVector (length v) s 
\end{verbatim}

In this function {\tt (+)} is used for elementwise addition of two vectors. 
The function {\tt constVector} is part of the EmbArBB library and creates a vector 
of a given length containing the same value at each index.

EmbArBB supports one, two and three dimensional dense vectors. These vectors are 
represented by a datatype called {\tt DVector} (for Dense Vector). {\tt DVector} takes two 
arguments, the first specifying its dimensionality and the second its payload 
type. Hence, an array of 32 bit words has type {\tt DVector Dim1 Word32}.
The one dimensional {\tt DVector} also has the alias {\tt Vector}. The type 
of the {\tt addconst} function above could also have been written:

\begin{verbatim}
addconst :: Num a 
          => Exp a 
          -> Exp (Vector a) 
          -> Exp (Vector a) 
\end{verbatim}

The nested vectors of ArBB correspond to {\tt NVectors} in EmbArBB. 
%These vectors are called nested vectors in ArBB but are limited to one level of nesting.

The {\tt addconst} function needs to be {\em captured} before it can be executed. 
The term {\em capture} is borrowed from ArBB nomenclature and corresponds to compiling 
the embedded language function into an ArBB function. In EmbArBB, a function is 
captured using the function {\tt capture}; in the C++ embedding, a function is 
captured when it is called for the first time. The same could have been done in 
the Haskell embedding of course, but we chose to make
capture explicit for simplicity.
The {\tt capture} function takes a Haskell function as input and produces an 
opaque typed function identifier as output. The identifier points out the function 
in an environment that is managed by a monad called {\tt ArBB}. To make this 
concrete, Figure~\ref{fig:code1} shows the complete procedure from capturing 
to execution.

%\begin{verbatim} 
%main = 
%  withArBB $ 
%  do 
%     f <- capture addconst
%     ...    
%\end{verbatim}

The {\tt withArBB} function (line two) is the ``run''-function of the {\tt ArBB} monad. 
It turns something of type {\tt (ArBB a)} into {\tt (IO a)}. This is how ArBB 
computations are interfaced with Haskell. A {\tt withArBB} session also sets the scope 
for any captured functions or vectors created in the ArBB monad.

The result of {\tt capture addconst} on line four in the code has type 
%Function (Float :- Vector Float) 
%         (MDVector Dim1 Float)
\begin{verbatim}
Function (BEScalar Float :- BEDVector Dim1 Float) 
         (BEDVector Dim1 Float)
\end{verbatim}
representing a function that takes a float and a vector of floats as 
input and gives a vector of floats as result. The ``BE'' prefixes 
in those names ({\tt BEScalar},{\tt BEDvector}) refers to this being vectors and scalar that 
reside in the ArBB heap (Backend vectors). BEVectors and Scalars are mutable and the 
vector used to store the result needs to be allocated by the programmer. This interface 
seems rather imperative but ArBB requires output data to be
placed into an output vector of the correct size. 
Because of this, the options were either to try to infer result size (which can be dependent on the value of input data) or to have the programmer
supply storage for the result. We chose the latter, and thus also avoid
additional runtime overhead due to size inference.

%This means that the options where to either somehow 
%infer the result size automatically (if that is even feasible, output size 
%could potentially be data dependent) or to let the programmer supply storage 
%for the result. Inferring size would surely also impose extra runtime overhead.

Continuing with the example: the function, once captured, can be executed 
on data. On line five, a {\tt DVector} is copied into ArBB using the {\tt copyIn} function. 
A vector to hold the result is created using the {\tt new} function on line eight. The 
{\tt new} function takes a shape description ({\tt (Z:.10)}, meaning a one dimensional 
vector of length 10) and an element to fill the vector with initially.
A scalar {\tt c} is created using {\tt mkScalar} on line nine. Next, on line ten, the 
function is executed by issuing 
\begin{verbatim}
execute f (c :- x) r1
\end{verbatim}
The {\tt (c :- x)} is a heterogeneous list of inputs. 
The {\tt :-} operator is similar to normal cons ({\tt :}) on Haskell lists.  
The output is stored in {\tt r1}. Had there been more outputs, the output
list would have also had a heterogeneous list type of the form
{\tt (r1 :- ... :- rn)}. If the result is a 
scalar, a storage location can be created using a function called 
{\tt mkScalar} that takes its initial value as input.

All that remains is to copy the output vector out of ArBB and show the result; this 
is done on lines eleven and twelve in the figure.

%main = 
% withArBB $ 
%  do 
%     f <- capture addconst
%     let x = fromList [0..10 :: Float]      
%     r1 <- liftIO$ new1D 10   
%     execute f (1 :- x)  r1   
%     r <- liftIO$ freeze r1             
%     liftIO$ putStrLn$ show r   
\begin{figure}
\begin{Verbatim}[numbers=left,frame=single] 
main = 
  withArBB $ 
  do 
     f <- capture addconst             
     x <- copyIn $ mkDVector 
                     (V.fromList [1..10 :: Float]) 
                     (Z:.10)
     r1 <- new (Z:.10) 0 
     c <- mkScalar 1      
     execute f (c :- x)  r1              
     r <- copyOut r1              
     liftIO$ putStrLn$ show r
\end{Verbatim}
\caption{The code shows how to capture a function, upload data to ArBB and 
         how to execute the captured function.}
\label{fig:code1}
\end{figure}  
%$
%** Note, I think it would be best to put the code above into a figure and then just refer to it in the discussion, rather than have three versions of it of
%increasing size.

In this example, we have seen all of the important parts of the
Haskell to EmbArBB interface: how embedded language functions are compiled and
executed, and how to transfer data from Haskell into the ArBB world.
The following subsections show EmbArBB versions of some common 
data parallel computations.

\input{./embarbb/listfun}

In the examples one can assume that following imports have been made: 
\begin{verbatim}
import Data.Vector as V
import Prelude as P
\end{verbatim}
In this way, wherever there is potential for mixup between EmbArBB functions 
and Prelude or Data.Vector functions, these are prefixed by either ``{\tt V.}'' or 
``{\tt P.}''.

% \FloatBarrier 
\subsection{Saxpy} 
{\em Saxpy} is a vector operation that gets its name from the 
description of what it does ``{\em S}ingle-precision {\em A}lpha {\em X} {\em P}lus {\em Y}''. 
There is also a double-precision version called {\em daxpy}. Here we can use a single source 
to obtain both versions:
%. The code listing below {\em axpy} illustrates this:

\begin{verbatim} 
saxpy :: Num a 
      => Exp a 
      -> Exp (DVector Dim1 a) 
      -> Exp (DVector Dim1 a) 
      -> Exp (DVector Dim1 a) 
saxpy s x y = (ss*x) + y 
    where 
      ss = constVector (length x) s
\end{verbatim}

Now, {\tt capture} can be used to instantiate the function at either float or double 
types.

\begin{verbatim}
main = 
  withArBB $ 
  do 
     f <- capture saxpy

     let v1 = V.fromList [1,3..10::Float] 
         v2 = V.fromList [2,4..10::Float]

     x <- copyIn $ mkDVector v1 (Z:.5) 
     y <- copyIn $ mkDVector v2 (Z:.5)

     r1 <- new (Z:.5) 0

     c <- mkScalar 1

     execute f (c :- x :- y)  r1

     r <- copyOut r1

     liftIO$ putStrLn$ show r
\end{verbatim}

%If instead the double precision version was desired change just the 
%type at which the function is captured. In the example above this 
%amounts to changing {\tt ::Float} to {\tt ::Double} in the definitions 
%of {\tt v1} and {\tt v2}.

Changing {\tt Float} to {\tt Double} in the definitions 
of {\tt v1} and {\tt v2} gives the double precision version of the function.
%Here, the two different choices for the type of the data lead to
%two different generated ArBB functions. 
A function needs to be captured at every type at which it is to be used. This 
is because the resulting ArBB function is not polymorphic over something corresponding 
to {\tt Num} types (while the embedded language function is).

\subsection{Matrix-vector multiplication} 
In section~\ref{sec:ProgrammingArBB}, matrix-vector multiplication was shown using the C++ ArBB interface. 
Here, EmbArBB is used to implement the same function. The function definition itself 
is very similar to the C++ one:

\begin{verbatim} 
matVec :: Exp (DVector Dim2 Float) 
        -> Exp (DVector Dim1 Float) 
        -> Exp (DVector Dim1 Float) 
matVec m v = addReduce rows 
           $ m * (repeatRow (getNRows m) v)

\end{verbatim}

The {\tt addReduce} function takes a parameter that specifies whether to reduce 
rows or columns.

% ** Note. What is a level specifier?

The following Haskell {\tt main} function completes the comparison to the C++ version shown in the Introduction:

\begin{verbatim}
main =  
  withArBB $ 
  do 
     f <- capture matVec  
     let m1 = V.fromList [2,0,0,0,
                          0,2,0,0,
                          0,0,2,0,
                          0,0,0,2]
         v1 = V.fromList [1,2,3,4]

     m <- copyIn $ mkDVector m1 (Z:.4:.4) 
     v <- copyIn $ mkDVector v1 (Z:.4)

     r1 <- new (Z:.4) 0

     execute f (m :- v)  r1

     r <- copyOut r1

     liftIO$ putStrLn$ show r
\end{verbatim}
%$ -- fixes syntax highlighting
The complete Haskell version of this program is slightly shorter than the corresponding 
C++ version. However, the C++ version could probably be made shorter as well; the Haskell and C++ versions must be considered similar in implementation complexity.

%This is 
%a very short example and cannot be expected to show off the strengths of either language.

\subsection{Matrix-matrix multiplication}

%Matrix multiplication is studied in the benchmarks in section~\ref{sec:Benchmarks}.
The following EmbArBB implementation of matrix-matrix multiplication is used as a benchmark in section~\ref{sec:Benchmarks}:

\begin{verbatim} 
matmul :: Exp (DVector Dim2 Float) 
        -> Exp (DVector Dim2 Float) 
        -> Exp (DVector Dim2 Float) 
matmul a b = fst $ while cond body (a,0)
  where 
    m = getNRows a
    n = getNCols b 
    cond (c,i) = i <* n
    body (c,i) = 
      let mult = a * repeatRow m (extractCol i b)
          col  = addReduce rows mult 
      in (replaceCol i col c, i+1) 
\end{verbatim}

This example uses a {\tt while} loop. In the C++ embedding of ArBB, the programmer 
has the option of using a C++ {\tt while} loop or a special {\tt while\_} loop. The 
normal C++ while loop unrolls at capture time; the {\tt while\_} is kept by ArBB 
and corresponds to an ArBB loop. The two loops differ in the kinds of
values on which they can depend.
The C++ loop can only depend on C++ values, so
the number of iterations in a normal {\tt while} loop must be known at capture 
time. All of these concepts have Haskell counterparts. The {\tt while} loop used in the 
example corresponds to the {\tt while\_} loop and will remain in the generated ArBB 
code; it can depend on ArBB values at runtime.
Haskell recursion is used to get the kind of unrolling 
that one gets by using a C++ {\tt while} loop.

The EmbArBB {\tt while} loop takes three parameters, a condition, a body and an initial state. 
In every iteration of the loop, the body is applied to the state and the condition checked.

%Using {\tt a}, that is the input matrix, as initial state may seem strange but it relies 
%on that the implementation of the while loop creates a copy of the initial state that 
%is used in the iterations. This makes the {\tt while} loop slightly unclear in what it 
%does and is something that must be improved as future work. An improvement could be 
%if the programmer specified {\tt (newMatrix m n,0)} as initial state.

\subsection{Histogram}
\FloatBarrier

The histogram of a grayscale image contains the frequency with 
which each shade occurs in the image. In this example, the image used represents 
the shade of each pixel with a single byte, so that
256 different shades are possible. An array of length 256 is created to hold at index 
{\tt i} the number of occurrences of the shade {\tt i} in the image. In essence, this 
is an array of buckets.

\begin{verbatim}
histogram :: Exp (DVector Dim2 Word8)  
           -> Exp (Vector Word32)
histogram input = addMerge (vecToUSize flat) 256 cv
    where
      flat = flatten input
      cv = constVector (r * c) 1 
      r  = getNRows input 
      c  = getNCols input

\end{verbatim}

The {\tt addMerge} operation takes a vector of inputs, a vector of 
indices and finally a size (specifying the result size). Elements from the 
input vector ({\tt cv} in the example) are placed into the result
at the index specified at 
the same location in the indices vector. Elements placed at the same
index are added together. Here, the input vector contains all ones, 
while the image is {\em cast} into a vector of indices.
The result 
is that index {\tt i} of the output vector contains the number of times
shade {\tt i} appears in the image.
The histogram
computation becomes very concise through the use of the {\tt addMerge} operation.

The code below creates the image shown in figure~\ref{fig:histogram}. It 
takes as input a vector on frequencies and outputs a two dimensional vector, 
a grayscale image.

\begin{verbatim}
histImage :: Exp (Vector Word32) 
           -> Exp (DVector Dim2 Word8) 
histImage input = fst $ while cond body (cvn,0)    
    where 
      cond (img,i) = i <* n 
      body (img,i) = (replaceCol i col' img,i+1) 
          where 
            val = input ! i 
            col = extractCol i img
            col' = fill black 0 n col
            n = 255 - scale 255 m val

      n = length input 
      cv = constVector (n*n) white  
      cvn = setRegularNesting2D n n cv
      m   = index0 (maxReduce rows input)
      black = 0 
      white = 255
\end{verbatim}

\begin{figure} 
\includegraphics[width=.5\linewidth]{./embarbb/img/window}
\includegraphics[width=.5\linewidth]{./embarbb/img/histogram}
\caption{ The righ-hand image visualises the frequency with which 
          different shades of gray occur in the left-hand one.}
\label{fig:histogram}
\end{figure}

\FloatBarrier
\subsection{Sobel edge detection filter}
\label{sec:Sobel}
\FloatBarrier

Sobel edge detection is an example of a stencil computation
over an array. At each element of the array, a computation is 
performed that depends on that element and on elements close by. Which of the nearby 
elements to use, and how much they influence the result is typically described using a matrix 
(in the two dimensional case) called the stencil. Below are two stencils used in the 
sobel edge detection filter:

\begin{minipage}{0.5\linewidth}
\[
G_x = \begin{bmatrix*}[r]
        -1 & 0 & 1 \\ 
        -2 & 0 & 2 \\
        -1 & 0 & 1 
      \end{bmatrix*}
\]
\end{minipage}
\begin{minipage}{0.5\linewidth}\centering
\[
G_y = \begin{bmatrix*}[r]
        -1 & -2 & -1 \\ 
         0 &  0 &  0 \\
         1 &  2 &  1 
      \end{bmatrix*}
\]
\end{minipage}

In ArBB, the {\em map} operation gives the programmer a way to implement stencil computation. 
It takes a function and a vector on which to apply the function at each element. 
Inside the function being mapped, the programmer may use a set of {\em getNeighbor} functions 
to access nearby elements.

The elements used by the stencils are kept (in the form of
coordinates relative to the centre point of the matrix) in two Haskell lists. This is an 
example of how host language (Haskell) features are used as a kind of 
scripting aid.

\begin{verbatim} 
s1, s2 :: [(Exp ISize,Exp ISize)]
s1 = [(1,1),(0,1),(-1,1),(1,-1),(0,-1),(-1,-1)]
s2 = [(1,1),(1,0),(1,-1),(-1,1),(-1,0),(-1,-1)]
\end{verbatim}

The actual weights are placed in a separate list.

\begin{verbatim} 
coeffs :: [Exp Float] 
coeffs = [-1,-2,-1,1,2,1] 
\end{verbatim}

The following functions implement the stencil computation part of the program
using the elements pointed out by {\tt s1} and {\tt s2} together with  the weights in {\tt coeffs}. 
Each of the six neighbours is multiplied by the appropriate weight. 
Haskell's {\tt foldl} function is used to sum up the values, to
give the result for the location in question. Using a Haskell fold means that the 
summation is unrolled.

\begin{verbatim}
gx :: Exp Word8 -> Exp Float  
gx x = P.foldl (+) 0 
     $ P.zipWith (*) [toFloat (getNeighbor2D x a b) 
                      / 255 
                     | (a,b) <- s1] coeffs

gy :: Exp Word8 -> Exp Float 
gy x = P.foldl (+) 0 
     $ P.zipWith (*) [toFloat (getNeighbor2D x a b) 
                      / 255
                     | (a,b) <- s2] coeffs 
\end{verbatim}

The helper functions {\tt convertToWord8} and {\tt clamp} take 
care of converting floats between 0 and 1 to bytes between 0 and 255, and 
of clamping floating point values into the 0 to 1 range.

\begin{verbatim}
convertToWord8 :: Exp Float -> Exp Word8 
convertToWord8 x = toWord8 $ (clamp x)  * 255

clamp :: Exp Float -> Exp Float
clamp x = max 0 (min x 1)  
\end{verbatim}

The complete kernel, the program that is to be executed at every index of the image,
is given below. The earlier parts are brought together into a function that 
transforms one {\tt Word8} into another.

\begin{verbatim} 
kernel :: Exp Word8 -> Exp Word8 
kernel x = convertToWord8 $ body x
  where 
    body x = sqrt (x' * x' + y' * y') 
     where 
       x' = gx x 
       y' = gy x
\end{verbatim}
\noindent
In order to execute the kernel, an ArBB {\tt map} is used. 
%ArBB map allows the function being 
%mapped to access elements neighbouring the current index by the {\tt getNeighbor} 
%functions.

\begin{verbatim}
sobel :: Exp (DVector Dim2 Word8) 
         -> Exp (DVector Dim2 Word8) 
sobel image = map kernel image  
\end{verbatim}

%In the current version of EmbArBB, the {\tt sobel} function needs to take as input 
%the function that it passes to map. This gives the {\tt sobel} function the somewhat
%complicated type that it has above. The reason the function needs to be passed as input 
%is that the {\tt kern} used in {\tt map kern image} is an opaque {\tt Function} object, 
%that is a result of calling {\tt capture}. Capturing is an operation in the {\tt ArBB} 
%monad and it is in that monad that the actual map of opaque {\tt Function} objects to 
%already generated ArBB functions is kept. 
%We see this as a wart on the current implementation.
%In future work, we would like to simplify matters for the user.

The implementation of {\tt sobel} is pleasingly simple and concise in this setting. 
Despite the fact that this is a small example, the value of being able to use Haskell
lists and operations is already becoming obvious.
%Making 
%use of Haskell lists and operations as powerful scripting tools.

\begin{figure}
\includegraphics[width=.5\linewidth]{./embarbb/img/window}
\includegraphics[width=.5\linewidth]{./embarbb/img/sobout}
\caption{ The image on the right shows the result of applying the sobel edge 
          detection filter to that on the left. }
\label{fig:sobel}
\end{figure}

%\subsection{Functions, Capturing and Executing}

%\subsection{Moving data to and from ArBB}

%\subsection{Programming Examples}
\FloatBarrier

\subsection{Image filtering} 
\label{sec:Blur}
The blur operation, which is a kind of spatial linear filter, shows another way to implement a stencil operation. The {\tt mapStencil} function used is a composite 
function in the EmbArBB library; it is not provided as a primitive from 
the ArBB VM.

\begin{verbatim} 
blur :: Exp (DVector Dim2 Word8) 
        -> Exp (DVector Dim2 Word8) 
blur image = vec2DToWord8 (res `div`  all16) 
  where 
    all16 = constVector2D (getNRows image) 
                          (getNCols image)
                          16
    res = mapStencil (Stencil [1,2,1
                              ,2,4,2   
                              ,1,2,1] (Z:.3:.3)) 
                     image'

    image' = vec2DToWord32 image        
\end{verbatim}

In Repa, similar {\tt mapStencil} functionality is present but using Template 
Haskell to give a safer way to specify the stencil. In our version there is 
no protection agains mischievous stencil specification. This guarantees only
that if the programmer says the stencil is two dimensional then it can only 
be applied to two dimensional vectors. Applying similar Template Haskell based
extension here as well is probably quite easy now that Repa has shown how.

\subsection{Sparse matrix multiplication}\label{sec:smvm}
A sparse matrix can be represented by three vectors, one (cidx) containing
column indices, one (offsets) indicating indices in the vector of values
at which the first non-zero element of each row appears, and one containing the
values themselves. This is known as the Compressed Sparse Row (CSR) format.
Given such a matrix and a dense vector, the relevant elements of the dense vector
can be gathered into a new vector using the column indices. The resuting vector
is multiplied pointwise by the vector of non-zero values from the sparse matrix. All that remains, then, is to divide the result into rows and sum each one. The offsets vector
points to where the division should happen.
The code in EmbArBB is as follows:

\begin{verbatim} 
smvm :: Exp (Vector USize)
     -> Exp (Vector USize)
     -> Exp (Vector Float)
     -> Exp (Vector Float)
smvm mval cidx os vec = addReduceSeg nps 
    where 
      ps = mval * gather1D cidx 0 vec
      nps = applyNesting offsets os ps 
\end{verbatim}

The multiplication acts pointwise on the elements of {\tt vals}
and the vector (of the same length) containing relevant elements of the dense vector (produced using {\tt gather1D}).
The function {\tt applyNesting} produces
a nested vector of rows. The function {\tt addReduceSeg} sums each
of the rows (or segments) in that array, producing the non-zero elements
of the result vector. Note that there is irregular data parallelism
here because the rows may have different lengths. We would hope
to get parallelism both in the individual summations and between summations.
This is a well known algorithm that can be traced back to Blelloch~\cite{NESL}.

%% NOTE ** I took out the 0 in the parameters to gather1D. How about
%% we just assume it is forward for now (and put in the zero in the call
%% to Gather)

%\begin{verbatim}
%import Data.Vector.Storable as V

%data DVector d a = Vector (V.Vector a) Dim

%data Dim = Zero 
%         | One Int
%         | Two Int Int 
%         | Three Int Int Int 
%\end{verbatim}

### 2013-09-30 src2/embarbb/related.tex
apa

\section{Related Work} 
\subsection{Data.Array.Accelerate} 
Accelerate~\cite{accelerate} is an embedded language for general purpose, 
flat data parallel computations on GPUs.

The Accelerate programmer expresses algorithms using collective operations 
over vectors. These collective operations are similar to the 
parallel patterns of ArBB, but are generally higher order. That is, where 
ArBB and EmbArBB have {\tt addReduce}, {\tt mulReduce} and so on, Accelerate
has a single higher-order fold function. This means that the Accelerate library
gets away with a much smaller set of operations, while maintaining a higher 
level of expressivity in the language.

Accelerate arrays have their dimensionality (shape) encoded in the type 
and this was the inspiration for the similar approach taken in EmbArBB. In 
Accelerate, a one dimensional array of integers has type {\tt Array DIM1 Int}.
However, Accelerate does not limit the dimensionality to one, two or three,
as ArBB does, but rather supports arbitrary dimensionality.

\subsection{Data Parallel Haskell} 
Data Parallel Haskell (DPH) is an extension to GHC for nested data parallelism. 
This is one thing that DPH and ArBB have in common.

In DPH, the programmer can use parallel arrays of type {\tt [:e:]} that look
similar to normal Haskell lists, but give access to data parallel execution.
The similarity to Haskell list processing doesn't end there. The operations 
on these parallel arrays are also reminiscent of Haskell's normal list processing functions.
For example, {\tt mapP}, {\tt unzipP} and {\tt sumP} are parallel versions of 
these well known functions.

DPH relies on a technique called flattening to transform its nested data parallelism 
into flat data parallelism. 
A recent article about DPH is reference~\cite{DPH}.

\subsection{Feldspar} 
Feldspar is a language for Digital Signal Processing (DSP) programming
developed at Chalmers, ELTE University and Ericsson 
(the telecom company)~\cite{FELDSPAR2010}. The functional {\tt while} loop 
of EmbArBB is inspired by the same concept in Feldspar.

Feldspar is based on a deeply embedded core language and implements 
a vector library as a shallow embedding on top of that. 
This means that there are no vector specific 
constructs in the core abstract syntax.

%The most recent release of Feldspar uses the {\em Syntactic} library
%for manipulating abstract syntax~\cite{SyntacticICFP12,FELDSYNTACTIC}. 
%We will consider switching to using the Syntactic library in the continued
%implementation of EmbArbb.
%Initial discussions with our colleague Emil Axelsson, the implementor of Syntactic, indicate that
%we should encounter no major difficulties.

\subsection{Nikola} 
Nikola is an embedded language for GPU programming\cite{NIKOLA}, also
in Haskell.  
During the early phases of implementation of EmbArBB, the source code of 
Nikola was often studied for inspiration.

The embedding used in Nikola makes use of an untyped expression 
data structure wrapped up in phantom types, in the style of Pan~\cite{ELLIJFP}.
The EmbArBB embedding works in the same way. 
One of the listed strengths of Nikola is the ability to generate GPU functions 
from Haskell functions, which enables function reuse and the ability to 
amortise the cost of code generation over several launches in an easy 
way. The ability the generate target language function from Haskell functions 
is a present in EmbArBB as well.

Nikola, like Accelerate, provides a set of higher-order functions with
general {\tt map}, {\tt reduce} and {\tt zipWith} functionality.

\subsection{Repa} 
Repa is a library for regular, shape polymorphic parallel arrays~\cite{REPA}. Repa 
uses a concept of {\tt delayed} arrays to obtain fusion of operations, such as the 
typical map fusion:
\begin{verbatim}
map f . map g  = map (f . g)
\end{verbatim} 
A delayed array 
has a representation that is quite direct to parallelise; it is represented 
as a function from an index-space to an element and the extents of that same 
index-space. Concretely:

\begin{verbatim} 
data DArray sh e = DArray sh (sh -> e)  
\end{verbatim}

Parallelising the computation of such an array is done by splitting up the 
index-space over the available parallel resources of the system. Repa does 
not use SIMD (vector) instructions; this is something that ArBB does and 
that EmbArBB gets for free.

Repa is compared to EmbArBB in the benchmarks in section~\ref{sec:Benchmarks}.

%\subsection{Obsidian}
%Really ? \cite{Obsidian}

\section{Motivation}

We have two main motivations for implementing EmbArBB.

Firstly, we are interested in exploring programming idioms for
data parallel programming in a functional setting. Embedding ArBB gives
a quick route to a platform for such exploration, without too much implementation effort. This is what the ``almost for free'' in the title is intended to refer to.
We wish to explore ways to make use of the fact that we have an array programming library embedded in a sophisticated, strongly typed host language. In our earlier work on the embedded hardware description language Lava, we investigated various approaches to exploiting the host language during netlist synthesis~\cite{LavaMultipliers}, and we have also experimented with the use of search and dynamic programming in generating parallel prefix (or scan) networks~\cite{SheeranJFP}. We intend to apply similar methods to the development of data parallel programs once the EmbArBB implementation is more complete and stable.

%It would be interesting to explore ways of using the host language in more sophisticated ways, rather as we have done for hardware generation in Lava~\cite{LavaMultipliers}.

A second motivation is the desire to teach NESL-style data parallel programming
in a Masters course on parallel functional programming that has recently been
introduced at Chalmers~\cite{TFPIEPFP}. The first instance of the course (in Spring 2012)
covered Blelloch's NESL language, with its
associated cost model~\cite{NESL}, but did not provide any satisfactory way for the students
to experience real, nested data parallel programming. This was due not only to a lack of suitable hardware, but also to deficiencies in the available tools.
We used the Repa library to get flat data parallelism, but then suffered from
the lack of a built-in scan primitive (as many of Blelloch's NESL examples use scan). Perhaps as a result
of needing scan, we did not get good performance.
We have not done enough examples or experiments yet, but it seems to us that EmbArBB
could be used to give students an experience of real parallel programming in a NESL-like
functional language. It remains to be seen whether the limited degree of nesting allowed in ArBB can be offset, at least to some extent, by clever use of the host programming language during generation of the desired abstract syntax tree.

### 2013-09-30 src2/embarbb/results.tex
apa

\section{Benchmarks}
\label{sec:Benchmarks}
%\FloatBarrier

In this section, matrix multiply, Sobel edge detection and an image blur
algorithm are used as benchmarks in comparing sequential C++ code, ArBB, EmbArBB, 
and a Haskell library called Repa. 
Repa provides regular shape polymorphic arrays; it permits
parallel execution of the resulting code, making use of multiple
cores, but not of SIMD parallelism~\cite{REPA}.
The Repa versions of the benchmarks used in the comparison 
come from the {\tt repa-examples-3.2.1.1} package on Hackage.

The processor used for all measurements is a four core Intel Core-I7 930 at 
2.80Ghz.

\subsection{matrix-matrix multiplication}
The matrix multiplication benchmarks consist of square matrices 
of sizes 256x256, 384x384, 512x512, 640x640 and 768x768. 
Figure~\ref{fig:mmchart1} shows the runtimes for the four implementations 
being compared, using the best settings in numbers of cores or threads found 
by previous experiments. The sequential C++ and ArBB versions come 
directly from the ArBB distribution.

\begin{figure}
\includegraphics[width=\linewidth]{./embarbb/img/mmchart1}
\caption{Shows in a log-scale the execution time of matrix-matrix multiplication
         comparing ArBB, EmbArBB, Repa and sequential C++. The ArBB and 
         EmbArBB lines are indistinguishable.}
\label{fig:mmchart1}
\end{figure}

\subsection{Sobel edge detection}
In the Repa Sobel code, only the applications of the stencils are timed. This 
part is defined as follows 
\begin{verbatim}
gradientX :: Monad m => Image -> m Image
gradientX img
        = computeP
        $ forStencil2 (BoundConst 0) img
              [stencil2| -1  0  1
                         -2  0  2
                         -1  0  1 |]

gradientY :: Monad m => Image -> m Image
gradientY img
        = computeP
        $ forStencil2 (BoundConst 0) img
             [stencil2|  1  2  1
                         0  0  0
                        -1 -2 -1 |] 
\end{verbatim}

A corresponding EmbArBB code to use in this comparison was implemented.

%\begin{verbatim} 
%gx :: Exp Float -> Exp Float  
%gx x = foldl (+) 0 
%     $ P.zipWith (*) [getNeighbor2D x a b
%                     | (a,b) <- s1] coeffs

%gy :: Exp Float -> Exp Float 
%gy x = foldl (+) 0 
%     $ P.zipWith (*) [getNeighbor2D x a b
%                     | (a,b) <- s2] coeffs 
%\end{verbatim}

\begin{verbatim} 
gx :: Exp (DVector Dim2 Float) 
    -> Exp (DVector Dim2 Float)  
gx = mapStencil 
         (Stencil [-1,0,1
                  ,-2,0,2
                  ,-1,0,1] (Z:.3:.3))

gy :: Exp (DVector Dim2 Float) 
    -> Exp (DVector Dim2 Float) 
gy = mapStencil 
         (Stencil [ 1, 2, 1
                  , 0, 0, 0
                  ,-1,-2,-1] (Z:.3:.3))
\end{verbatim}

%% ** This para needs to be fixed.

We compare these two examples, in which only the applications of the stencils
are timed; we also time the complete Sobel implementation in EmbArBB (shown
in section~\ref{sec:Sobel}).
The Sobel benchmark is run om images of sizes 256x256, 512x512, 1024x1024, 2048x2048 and 4096x4096.
Again, Figure~\ref{fig:sobelchart1} shows that EmbArBB performs well.  As we had expected, the 
Haskell embedding, once a function is captured, seems to impose little or no overhead 
compared to the C++ implementation.

\begin{figure}
\includegraphics[width=\linewidth]{./embarbb/img/sobelchart1}
\caption{Shows in a log-scale the execution time of a key part of the sobel edge 
         detection program. The chart compares EmbArBB to REPA and also displays 
         for reference the execution times of the full sobel program as implemented 
         in section~\ref{sec:Sobel}, called {\em EmbArBB*} in the chart.}
\label{fig:sobelchart1}
\end{figure}

\subsection{Blur}

The {\tt blur} benchmark is performed using an algorithm similar to the example 
code in section~\ref{sec:Blur} but with the following changes. The image used
is in RGB color. This means that the stencil needs to be applied three times, 
once to each color plane. Also the image is converted into a form 
where each color intensity is represented by a {\tt Double}.

Here as well as in the sobel case the Repa code from \newline{\tt repa-examples-3.2.1.1}
times only the actual computational kernel (the application of the stencils) 
including conversion to Doubles. The corresponding EmbArBB part was broken 
out and timed separately as well. The chart~\ref{fig:blurchart} shows comparison
of runtime for the key part of the computation but also adds the full execution 
time of the EmbArBB version (called {\em EmbArBB*} in the chart). 
The full EmbArBB implementation of the blur filter including conversion into Doubles, 
image decomposition into R,G and B planes,  application of the stencil and
reconstructing a planar RGB image in the end.

\begin{figure}
\includegraphics[width=\linewidth]{./embarbb/img/blurchart}
\caption{Shows the execution time of a 7x7 blur filter applied to images 
         of various sizes.}
\label{fig:blurchart}
\end{figure}

%\begin{figure}

\begin{table}
    \begin{small}
    \begin{tabular}{|l|r|r|r|r|}
        \hline
        ~             & 256x256 & 512x512 & 1024x1024 & 2048x2048 \\ \hline
        Repa 3x3      & 12 & 27 &  72 & 190 \\ 
        EmbArBB 3x3   &  1 &  2 &  20 &  77 \\
        EmbArBB* 3x3  &  2 &  9 &  33 & 153 \\  \hline 
        Repa 5x5      & 13 & 32 &  87 & 254 \\
        EmbArBB 5x5   &  1 &  5 &  28 & 105 \\ 
        EmbArBB* 5x5  &  3 & 12 &  46 & 204 \\  \hline 
        Repa 7x7      & 20 & 48 & 119 & 368 \\ 
        EmbArBB 7x7   &  2 & 10 &  46 & 176 \\ 
        EmbArBB* 7x7  &  4 & 17 &  63 & 270 \\
        \hline
    \end{tabular}
    \end{small}
\caption{The table shows execution times (rounded to ms) for various image and stencil size 
         combinations in both Repa and EmbArBB.}
\label{fig:blurtable}

\end{table}

\subsection{About the numbers} 
This section presents three benchmarks, two of which compare to Repa only and 
one that compares to Repa, ArBB in C++ and sequential C++ code. In all of these comparisons,
JIT compilation time is excluded.

The comparison to ArBB in C++ shows that the Haskell embedding does not impose any 
extra overhead (at least not in this benchmark); this matched our expectations. More comparisons to the C++ version of ArBB are needed to confirm these first impressions about overhead in the Haskell embedding. If future directions of EmbArBB development develop techniques (such as size inference) that impose a runtime overhead, then the 
comparison in execution to the C++ version becomes more important.

% ;) 
The comparisons to Repa all show that the performance of EmbArBB compares favourably. This can be attributed to the way in which ArBB's developers at Intel have incorporated vectorisation and threading.

%\end{figure}

%In the Sobel case, EmbArBB and Repa are compared but the complete version 
%of sobel as implemented in section~\ref{sec:Sobel} is also shown for reference. The EmbArBB 
%version of the code is shown earlier in section~\ref{sec:Programming}.

%Figures~\ref{fig:arbbchart} and~\ref{fig:embchart} show running times on ArBB and 
%EmbArBB when utilising one,two or four cores. The nearly identical figures 
%illustrate the parallel speedup obtained by ArBB on this particular problem.

%Figure~\ref{fig:repachart} shows similar parallel speedups for Repa when changing 
%the numbers of threads that it is allowed to use.

%% ** Maybe need some structuring here.

%\begin{verbatim}
%mmultP  :: Monad m
%        => Array U DIM2 Double 
%        -> Array U DIM2 Double 
%        -> m (Array U DIM2 Double)
%
%mmultP arr brr 
% = do   trr      <- transpose2P brr
%        let (Z :. h1  :. _)  = extent arr
%        let (Z :. _   :. w2) = extent brr
%        computeP 
%         $ fromFunction (Z :. h1 :. w2)
%         $ \ix   -> R.sumAllS 
%                  $ R.zipWith (*)
%                        (unsafeSlice arr (Any :. (row ix) :. All))
%                        (unsafeSlice trr (Any :. (col ix) :. All))
%
%transpose2P
%        :: Monad m
%        => Array U DIM2 Double 
%        -> m (Array U DIM2 Double)
%
%transpose2P arr
% = computeUnboxedP
% $ unsafeBackpermute new_extent swap arr
% where  swap (Z :. i :. j)      = Z :. j :. i
%        new_extent              = swap (extent arr)
%
%
%-- | Take the row number of a rank-2 index.
%row :: DIM2 -> Int
%row (Z :. r :. _) = r
%
%
%-- | Take the column number of a rank-2 index.
%col :: DIM2 -> Int
%col (Z :. _ :. c) = c
%
%\end{verbatim}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/chart1}
%\caption{Running time and parallel speedup of matrix 
%         multiplication on matrices of different size using ArBB}
%\label{fig:arbbchart}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/arbb1}
%\caption{Execution time for matrix-matrix multiplication using ArBB (C++).
%         Chart shows execution time for various sizes of matrices using 
%         either one two or four cores.}
%\label{fig:arbbchart}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/embarbb1}
%\caption{Execution time for matrix-matrix multiplication using EmbArBB.
%         Chart shows execution time for various sizes of matrices using 
%         either one, two or four cores.}
%\label{fig:embchart}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/repa1}
%\caption{Execution time for matrix-matrix multiplication using Repa.
%         Chart shows execution time for various sizes of matrices using 
%         either one, two, four or eight threads.}
%\label{fig:repachart}
%\end{figure}

%\begin{figure}
%\includegraphics[width=\linewidth]{./img/compare2}
%\caption{CAPTION HERE}
%\label{fig:compare2chart}
%\end{figure}

\FloatBarrier

### 2013-09-30 src2/ifl/GPUs.tex
apa

\section{Graphics Processing Units, accessible high performance parallel computing}
\label{sec:gpu}

In the development of microprocessors, the addition of new cores is now the
way forward, rather than the improvement of single thread performance.
Graphics processing units (GPUs) have moved from being specialised graphics
engines to being suitable to tackle applications with high computational
demands. For a recent survey of the hardware, programming methods and tools,
and successful applications, the reader is referred to~\cite{GPUComputing}.
Figure~\ref{fig:image}, taken from that paper, and due to NVIDIA, shows the
architecture of a modern GPU from NVIDIA. It contains 16 multiprocessors, 
grouped in pairs that share a texture fetch unit (TF in the figure). The 
texture fetch unit is of little importance when using the GPU for general 
purpose computations. Each multiprocessor has 8 stream processors (marked 
SP in the figure). These stream processors has access to 16kB of shared memory.

See reference~\cite{GPUComputing} for information about
the very similar AMD GPU architecture. We have used the NVIDIA architecture,
but developments are similar at AMD. Intel's Larrabee processor points to a
future in which each individual core is considerably more powerful than in
today's GPUs~\cite{IntelLarrabee}.

%along with shared data and instruction caches, and 16kB of shared memory.

The question of how to program powerful data-parallel processors is likely
to continue to be an interesting one. Unlike for current multicore machines,
the question here is how to keep a large number of small processors
productively occupied. NVIDIA's solution has been to develop the
architecture and the programming model in parallel. The result is called
CUDA -- an extension of C designed to allow developers to exploit the power
of GPUs. Reference~\cite{GPUCudaLuebke} gives a very brief but illuminating
introduction to CUDA for potential new users. The idea is that the user
writes small blocks of straightforward C code, which should then run in
thousands or millions of threads. We borrow the example from the above
introduction. To add two $N \times N$ matrices on a CPU, using C, one would
write something like

\begin{code}
// add 2 matrices on the CPU:
void addMatrix(float *a, float *b, float *c, int N)
{
  int i, j, index;
  for (i = 0; i < N; i++) {
    for (j = 0; j < N; j++) {
      index = i + j * N;
      c[index]=a[index] + b[index];
    }
  }
}
\end{code}
\FloatBarrier

\begin{figure}%
\begin{center}
\includegraphics[height = 6cm]{./ifl/image_top.pdf}
%\includegraphics[width = 14cm]{image.pdf}
\caption{The NVIDIA 8800GTX GPU architecture, with 8 pairs of multiprocessors. Diagram courtesy of NVIDIA.}\label{fig:image}
\end{center}
\end{figure}

\noindent In CUDA, one writes a similar C function, called a {\em kernel},
to compute one element of the matrix. Then, the kernel is invoked as many
times as the matrix has elements, resulting in many threads, which can be
run in parallel. A predefined structure called {\tt threadIdx} is used to
label each of these many threads, and can be referred to in the kernel.
\begin{code}
// add 2 matrices on the GPU (simplified)
__global__ void addMatrix(float *a,float *b, float *c, int N)
{
  int i= threadIdx.x;
  int j= threadIdx.y;
  int index= i + j * N;
  c[index]= a[index] + b[index];
}

void main()
{
  // run addMatrix in 1 block of NxN threads:
  dim3 blocksize(N, N),
  addMatrix<<<1, blocksize>>>(a, b, c, N);
}
\end{code}
Here, a two dimensional {\em thread block} of size $N \times N$ is created.

CUDA uses {\em barrier synchronisation} and {\em shared memory} for
introducing communication between threads. Contents of shared memory (16kB
per multiprocessor in the architecture shown in Figure~\ref{fig:image}) is
visible to all threads in a thread block. It is very much faster to access
this shared memory than to access the global device memory. We shall see
later that Obsidian provides users with both shared and global arrays,
giving the user control over where data is to be stored.

Since many threads are now writing and reading from the same shared memory,
it is necessary to have a mechanism that enables the necessary
synchronisation between threads. CUDA provides a barrier synchronisation
mechanism called {\tt \_\_syncthreads()}. Only when all threads in a block
have reached this barrier can any of them proceed. This allows the
programmer to ensure safe access to the shared memory for the many threads
in a thread block.

Now, a {\em grid} is a collection of thread blocks. Each thread block runs
on a single multiprocessor, and the CUDA system can schedule these individual
blocks in order to maximise the use of GPU resources. A complete program
then consists not only of the kernel definitions, but also of code, to be
run on the CPU, to launch a kernel on the GPU, examine the results and
possibly launch new kernels. In this paper, we will not go into details
about how kernels are coordinated, but will concentrate on how to write
individual kernels, as this is the part of Obsidian that is most developed.
In Obsidian, we write code that looks like the Lava descriptions in
section~\ref{sec:combinators}, and we generate CUDA code like that shown
above. This is a considerably more complex process than the generation of
netlists in Lava.

### 2013-09-30 src2/ifl/combinators.tex
apa

\section{Connection patterns for hardware design and parallel programming}\label{sec:combinators}

Connection patterns that capture common ways to connect sub-circuits into larger structures have been central to our research on functional and relational languages for hardware design.
Inspired by Backus' FP language, Sheeran's early work on $\mu$FP made use of
{\em combining forms} with geometric interpretations~\cite{SheeranLFP84}. This approach
to capturing circuit regularity was also influenced by contact
with designers of regular array circuits in industry -- see reference~\cite{SheeranJUCS} for an overview of this and much other work on functional programming and hardware design.

Later work on (our) Ruby considered the use of binary relations, rather than functions in specifying hardware~\cite{LyngbyRuby}. Lava builds upon these ideas, but also gains much in expressiveness and flexibility by being embedded in Haskell~\cite{lavaICFP,ClaessenThesis}.
The user writes what look like circuit descriptions, but are in fact circuit {\em generators}. Commonly used {\em connection patterns} are captured by higher order functions.

For example, an important pattern is parallel prefix or scan. Given inputs $[x_0, x_1 \ldots x_{n-1}]$, the prefix problem is to compute each $x_0 \circ x_1 \circ \ldots \circ x_j$ for $0 \leq j < n$, for $\circ$ an associative, but not necessarily commutative, operator. For example, the prefix sum of {\tt [1..10]} is {\tt [1,3,6,10,15,21,28,36,45,55]}. There is an obvious sequential solution, but in circuit design one is often aiming for a circuit that exploits parallelism, and so is faster (but also larger). In a construction attributed to Sklansky, one can perform the prefix
calculation by first, recursively, performing the prefix calculation on 
each half of the input, and then combining (via the operator) the last output of the first of these
recursive calls with each of the outputs of the second.
For instance, to calculate the prefix sum of {\tt [1..10]}, one can compute the prefix
sums of {\tt [1..5]} and {\tt [6..10]}, giving {\tt [1,3,6,10,15]}
and {\tt [6,13,21,30,40]}, respectively. 
The final step is to add the last element of the output of the first recursive call ({\tt 15}) to each element of
the output of the second.

To express the construction in Lava, we make use of two connection patterns.
{\tt two :: ([a] -> [b]) -> [a] -> [b]} applies its component to
the top and bottom halves of the input list, concatenating the two sub-lists
that result from these applications.
Thus, {\tt two (sklansky plus)} applied to {\tt [1..10]} gives {\tt [1,3,6,10,15,6,13,21,30,40]}.
Left-to-right serial composition has type
{\tt (a -> b) -> (b -> c) -> a -> c} and is written as infix {\tt ->-}.
The description of the construction mixes the use of connection patterns, giving a form of reuse, with the naming of ``wires''.

\begin{code}
sklansky :: ((t, t) -> t) -> [t] -> [t]
sklansky op [a] = [a]
sklansky op as = (two (sklansky op) ->- sfan) as
  where
    sfan as = a1s ++ a2s'
      where
        (a1s,a2s) = splitAt ((length as + 1) `div` 2) as
        a2s'      = [op(last a1s,a) | a <- a2s]

*Main> simulate (sklansky plus) [1..10]
[1,3,6,10,15,21,28,36,45,55]
\end{code}

Lava supports simulation, formal verification and netlist generation from definitions like this. Circuit descriptions are {\em run} (in fact symbolically evaluated) in order to produce an intermediate representation, which is in turn written out in various formats (for fixed size instances). So this is an example of {\em staged programming}~\cite{Taha03}.

The Sklansky construction is one way to implement parallel prefix, and there are many others, see for instance Hinze's excellent survey~\cite{Hinze}.
Those who develop prefix algorithms suitable for hardware implementation use a standard notation to represent
the resulting networks. Data flows from top to bottom and
the least significant input is at top left. Black dots represent
operators.
For example, Figure~\ref{fig:skl32} shows the
recursive Sklansky construction for 32 inputs. 
%
\begin{figure}%
\begin{center}
\includegraphics[bb=0 0 123 748, angle=270, scale=0.4]{./ifl/skl32.pdf}
\caption{The Sklansky construction for $32$ inputs. 
It recursively computes the parallel prefix for each half of the inputs (corresponding to the use of {\tt two} in the definition) and then combines the last output of the lower (left) half with each of the outputs of the upper (right) half. The dotted box outlines the recursive call on the lower half of the inputs.}\label{fig:skl32}
\end{center}
\end{figure}

In this work, we plan to investigate the use of connection patterns, and more generally an emphasis on {\em structure}, in parallel programming.
We have chosen to target GPUs partly because of available expertise among our colleagues at Chalmers, and partly because reading papers about General Purpose GPU (GPGPU) programming gave us a sense of d{\'e}j\`a vu. Programs are illustrated graphically, and bear a remarkable resemblance to circuit modules that we have
generated in the past using Lava. We see an opportunity here, as there is an extensive literature, going back to the 1960s, about implementing algorithms on silicon that may provide clues
about implementing algorithms on GPUs. This literature does not seem to have yet been scrutinised by the Data Parallel Programming or GPGPU communities.
This is possibly because GPUs are moving closer to simply being data parallel machines, and so work on library functions has taken inspiration from earlier work on Data Parallel Programming, such as Blelloch's NESL~\cite{NESL}. But some of the restrictions from the early data parallel machines no longer hold today; for instance broadcasting a value to many processors was expensive in the past, but is much easier to do on modern GPUs. So a construction like Sklansky, which requires such broadcasting, should now be reconsidered, and indeed we have found it to give good results in our initial experiments (writing directly in CUDA). In general, it makes sense to spread the net beyond the standard data parallel programming literature when looking for inspiration in parallel algorithm design. We plan to explore the use of ``old'' circuit design ideas in programming library functions for GPUs.

Below, we briefly review modern GPUs and a standard programming model.

### 2013-09-30 src2/ifl/conclusion.tex
apa

\section{Conclusion} \label{sec:conc}

Obsidian provides a good interface for experimenting with algorithms on GPUs. 
The earlier version described in~\cite{JMT} showed that it is possible to 
generate efficient CUDA code from the kind of high level descriptions we are 
interested in. For the kernel level, the work in progress described in this paper enhances the  expressive power of Obsidian, extending the range of algorithms that can be described, as well as the degree of control exercised by the user. 
Future work will concentrate on improving the performance of the resulting 
applications, as well as on support for the kernel coordination level.

### 2013-09-30 src2/ifl/futurework.tex
apa

% ------------------------------------------------------------------------------
\section{Future work} \label{sec:fut}

The version of Obsidian described here is at a very experimental stage. 
The quality of the C code generated needs to improve to get performance 
on par with the previous version. The previous version however, was very
limited in what you could express. This older version is described in
\cite{JMT}.
There is a clear opportunity to perform classic compiler optimisations on 
the IC formed by running an Obsidian program. Currently this is not done at all.

Ways to describe the coordination 
of kernels in code that is still short and sweet are also needed. Some experiments using 
methods similar to Lava's netlist generation have been performed, but the
resulting performance is not yet satisfactory. In CUDA, Kernel coordination 
is in part described in the actual kernel code. Kernels decide which parts 
of the given data to use. As future work we will approach the kernel 
coordination problem at a lower more CUDA-like level. We will, of necessity,
need to develop programming idioms or combinators that express how data is 
placed in the memory hierarchy. The isolation of this as a central question 
is one of the more unexpected and interesting results of the project.
Right now work is focused on developing combinators that are more clever 
in their treatment of {\tt sync}s. This leads to new data structures that 
allow the merging of {\tt sync}s. This new approach seems to make efficient
implementations of combinators such as {\tt two} possible.

%We also need to experiment with ways of ``moving syncs around'' in our 
%generated code, in order to avoid unnecessary serialisation caused by having 
%too many syncs of small groups of threads.

### 2013-09-30 src2/ifl/intro.tex
apa

\section{Introduction}

There is a pressing need for new and better ways to program parallel
machines. Graphics Processing Units (GPUs) are one kind of such parallel
machines that provide a lot of computing power. Modern GPUs have become
extremely interesting for {\em general purpose} programming, i.e. computing
functions that have little or nothing to do with graphics programming.

We aim to develop high-level methods and tools, based on functional
programming, for low-level general purpose programming of GPUs. The methods
are high-level because we are making use of common powerful programming
abstractions in functional programming, such as higher-order functions and
polymorphism. At the same time, we still want to provide control to the
programmer on important low-level details such as how much parallelism is
introduced, and memory layout. (These aims are in contrast to other
approaches to GPU programming, where the programmer expresses the intent in
a high-level language, and then lets a smart compiler try to do its best.)

Based on our earlier work on structural hardware design, we plan to
investigate whether or not a {\em structure-oriented programming style} can
be used in programming modern GPUs. We are developing Obsidian, a domain
specific language embedded in Haskell. The aim is to make extensive use of
higher order functions capturing {\em connection patterns}, and from these
compact descriptions to generate code to run on the GPUs. Our
hardware-oriented view of programming also leads us to investigate the use
of algorithmic ideas from the hardware design community in parallel
programming.

In the rest of the paper, we first briefly review our earlier work on
hardware description languages (Sect.\ \ref{sec:combinators}), and introduce
the GPU architecture we are working with (Sect.\ \ref{sec:gpu}). After
that, we describe the current status of the language Obsidian, and show
examples (Sect.\ \ref{sec:obsidian}) and experimental results (Sect.\
\ref{sec:results}). Obsidian is currently very much work in progress; we
discuss motivations and current shortcomings (Sect.\ \ref{sec:disc}) and
future directions (Sect.\ \ref{sec:fut}).

### 2013-09-30 src2/ifl/obsidian.tex
apa

% ------------------------------------------------------------------------------
\section{Obsidian: a domain specific embedded language for GPU programming}
\label{sec:obsidian}

To introduce Obsidian, we consider the implementation of a parallel prefix 
kernel. The implementation bears a close resemblance to the Lava implementation 
from section~\ref{sec:combinators}:

% ------------------------------------------------------------------------------
%sklansky :: (Choice a, Syncable (Arr s a)) => 
%            Int -> (a -> a -> a) -> Arr s a -> W (Arr s a) 
%sklansky 0 op = return 
%sklansky n op = two (sklansky (n-1) op) ->- sfan ->- sync
%    where sfan arr = do 
%            (a1,a2) <- halve arr
%            let m = len a1
%                c = a1 ! (m-1)
%            a2' <- fun (op c) a2
%            conc (a1,a2')

\begin{code}
sklansky :: (Choice a, Syncable (Arr s a)) => 
             (a -> a -> a) -> Arr s a -> W (Arr s a) 
sklansky op arr 
   | len arr == 1 = return arr
   | otherwise = (two (sklansky op) ->- sfan ->- sync) arr
    where sfan arr = do 
            (a1,a2) <- halve arr
            let m = len a1
                c = a1 ! (fromInt (m-1))
            a2' <- fun (op c) a2
            conc (a1,a2')

\end{code}
% ------------------------------------------------------------------------------
The most notable differences between the two implementations are that Obsidian 
functions are monadic and that a datatype {\tt Arr} is used where Haskell lists
are used in Lava. The pattern matching on the list used in Lava is here 
replaced by guards. The function {\tt len :: Arr s a -> Int} gives the 
length of the array. These differences lead to a slightly different programming
style.

%Another effect
%of using the {\tt Arr} type rather than lists is the need to pass a size
%parameter to the {\tt sklansky} function.  However, this limitation comes from 
%a design decision that may be subject to change shortly.

The Obsidian version of the {\tt sklansky} function implements the sought
recursive parallel prefix algorithm, but it 
contains no information about where in the memory hierarchy the intermediate
results are to be held. The following program uses {\tt sklansky} from 
above but turns it into a concrete kernel that computes all the partial sums
of an array of integers:

% ------------------------------------------------------------------------------
\begin{code}
scan_add_kernel :: GArr IntE -> W (GArr IntE) 
scan_add_kernel = cache ->- sklansky (+) ->- wb ->- sync
\end{code}
% ------------------------------------------------------------------------------
The function {\tt cache} specifies that if the array is stored it should be
stored in the on-chip shared memory. Actually storing an array is done using
the {\tt sync} function, which functionally is the identity function, but has
the extra effect of synchronising all processes after writing their data in
shared memory, such that they can exchange intermediate results. Using {\tt
sync} here allows for computations or transformations to be performed on the
data as it is being read in from global device memory. In the {\tt
scan\_add\_kernel} above this means that the first {\tt sklansky} stage will
be computed with global data as input, putting its result into shared memory.
A kernel computing the same thing but using the memory differently can be
implemented like this:
% ------------------------------------------------------------------------------
\begin{code}
scan_add_kernel2 :: GArr IntE -> W (GArr IntE) 
scan_add_kernel2 = cache ->- sync ->- sklansky (+) ->- wb ->- 
                     sync
\end{code}
% ------------------------------------------------------------------------------
Here the array is first stored into shared memory. The {\tt sklansky} stages
are then computed entirely in shared memory. The write-back function, {\tt wb}, 
works in a very similar way but specifies that the array should be moved back 
into the global memory.

Now, {\tt scan\_add\_kernel} can be launched on the GPU from within a GHCI 
session using a function called {\tt execute}:

% ------------------------------------------------------------------------------
\begin{code}
execute :: ExecMode -> (GArr (Exp a) -> W (GArr (Exp b)) -> 
              [Exp a] -> IO [Exp b]
\end{code}
% ------------------------------------------------------------------------------
Here {\tt ExecMode} can be either {\tt GPU} for launching the kernel on the 
GPU or {\tt EMU} for running it in emulation mode on the system's CPU. Below is 
the result\footnote{The output has been shortened to fit on a line.} of 
launching the scan kernel on example input:

% ------------------------------------------------------------------------------

%  \DefineVerbatimEnvironment
%     {code}{Verbatim}
%     {fontsize=\small,xleftmargin=0.6cm,samepage=true}

%\begin{small}

\begin{code}
*Main> execute GPU scan_add_kernel [1..256]
[0,1,3,6,10,15, ... ,32131,32385,32640] 
\end{code}

%\end{small}
% [E (LitInt 1),E (LitInt 3),E (LitInt 6), ... ,E (LitInt 32896)] 
% ------------------------------------------------------------------------------

Beyond the combinators described so far, we have experimented with combinators 
and permutations  needed for certain iterative sorting networks. Amongst these 
are {\tt evens} that applies a function to each pair of elements of an array. 
Together with {\tt rep} that repeats a computation a given number of times and 
a permutation called {\tt riffle} a shuffle exchange network can be defined:

% ------------------------------------------------------------------------------

\begin{code}
shex n f = rep n (riffle ->- evens f ->- sync)
\end{code}

% ------------------------------------------------------------------------------
The shuffle exchange network can be used to implement a merger useful in sorters.

% ------------------------------------------------------------------------------
\subsection{Implementation}

As seen in the examples, an Obsidian program is built from functions between 
arrays. These arrays are of type {\tt Arr s a}. There are also type synonyms
{\tt GArr} and {\tt SArr} implemented as follows:

% ------------------------------------------------------------------------------
\begin{code}
data Arr s a = Arr (IxExp ->  a, Int)
type GArr a = Arr Global a
type SArr a = Arr Shared a 
\end{code}
% ------------------------------------------------------------------------------
In Obsidian an array is represented by a function from indices to values and an 
integer giving the length of the array.

In most cases the {\tt a} in {\tt Arr s a} will be of an expression type: 
\begin{code} 
data DExp = LitInt Int
          | LitBool Bool
          | LitFloat Float 
          | BinOp Op2 DExp DExp 
          | UnOp  Op1 DExp 
          | If DExp DExp DExp 
          | Variable Name 
          | Index DExp DExp  
            deriving(Eq,Show)
\end{code}

The above expressions are dynamic in the sense that they can be used to 
represent values of {\tt Int}, {\tt Float} and {\tt Bool} type. 
This follows the approach from Compiling Embedded Languages 
\cite{Elliott03:CompileDSEL-JFP}. However, 
to obtain a typed environment in which to operate, phantom types are used.

\begin{code} 
type Exp a = E DExp

type IntE   = Exp Int
type FloatE = Exp Float
type BoolE  = Exp Bool
type IxExp  = IntE 
\end{code}

As an example consider the program: 
% ------------------------------------------------------------------------------
\begin{code} 
add_one :: GArr IntE -> W (GArr IntE)
add_one = fun (+1) ->- sync
\end{code} 
% ------------------------------------------------------------------------------
This program adds 1 to each element of an array of integers. The function {\tt fun}
has type {\tt (a -> b) -> Arr s a -> Arr s b}. {\tt fun} performs for arrays what 
{\tt map} does for lists. 
The {\tt sync} function used in the example has the effect that
the array being synced upon is written to memory. At this point the type of 
the array determines where in memory it is stored. An array of type {\tt SArr} 
will end up in the {\em shared memory} (which is currently 16KB per multi-processor). 
In this version of Obsidian it is up to the programmer to make sure that the 
array fits in the memory. An array of type  {\tt GArr} ends up in the global 
device memory, roughly a gigabyte on current graphics cards.  Using {\tt sync} 
can have performance implications since it facilitates sharing of computed 
values. 
%However, functionally it is the identity function.

To generate CUDA code from the Obsidian program {\tt add\_one}, it is applied to
a symbolic array of a given concrete length (in this example 256 elements):

% ------------------------------------------------------------------------------
\begin{code}
input :: GArr IntE 
input = mkArray (\ix -> (index (variable "input") ix)) 256
\end{code}
% ------------------------------------------------------------------------------

Applying {\tt fun (+1)} to this input array results in an array with the 
following indexing function:
\begin {code}
(\ix -> (E (BinOp Add (Index (Variable "input") ix) (LitInt 1)))) 
\end{code}

At the code generation phase this function is evaluated using a variable 
representing a thread Id. The result is an expression looking as follows:
\begin {code}
E (BinOp Add (Index (Variable "input") (Variable "tid")) (LitInt 1))
\end{code}

This expression is a direct description of what is to be computed. 
%Figure~\ref{fig:expressions} illustrates this pictorially.
% ------------------------------------------------------------------------------
%\begin{figure}%
%\begin{center}
%\includegraphics[scale=.25]{expressions.png}
%\caption{Expressions and arrays.}
%\label{fig:expressions}
%\end{center}
%\end{figure} 
% ------------------------------------------------------------------------------

% ------------------------------------------------------------------------------
Importantly, the basic library functions can be implemented using the 
{\tt Arr s a} type and thus be applicable both to shared and global arrays.
The library function {\tt rev} that reverses an array is shown as an example
of this:

% ------------------------------------------------------------------------------
\begin{code}
rev :: Arr s a -> W (Arr s a) 
rev arr = let n = len arr 
          in  return $ mkArray (\ix -> arr ! ((n - 1) - ix)) n
\end{code}
%$% 
% ------------------------------------------------------------------------------
The function {\tt rev} uses {\tt mkArray} to create an array whose indexing 
function reverses the order of the elements of the given array {\tt arr}. 
The Obsidian program {\tt rev $-$>$-$ sync} 
corresponds\footnote{in the real IC {\tt arrx} and {\tt arry} are replaced by identifiers generated in the{\tt W} monad}
to the following lines of C code:

\begin{code}
arrx[ThreadIdx.x] = arry[n - 1 - ThreadIdx.x];  
__syncthreads();
\end{code}

When an Obsidian function, such as the {\tt scan\_add\_kernel} from the 
previous section, is run, two data structures are accumulated into a monad 
called {\tt W}. The first is intermediate code {\tt IC} and the second a symbol 
table. The {\tt W} monad is a writer monad extended with some extra 
functionality for generating identifier names and to maintain the symbol table. 
You can think of the {\tt W} monad as:

\begin{code}
type W a = WriterT (IxExp -> IC) (State (SymbolTable,Int)) a
\end{code}

The IC used here is just a list of statements, (less important statements have 
been removed to save space ({\tt ...})). The {\tt IC} contains a 
subset of CUDA. In this version of Obsidian not much more than the 
{\tt Synchronize} and assignment statements of CUDA are used. 
%Looping constructs are not 
%part of the {\tt Statement} type. We are experimenting with 
%generating unrolled loops only:

% ------------------------------------------------------------------------------
%\begin{samepage}
\begin{code} 
data Statement = Synchronize 
               | DExp ::= DExp 
               -- used later in code generation
               | IfThenElse BoolE [Statement] [Statement]
                 deriving (Show,Eq)

type IC = [Statement]
\end{code} 
%\end{samepage}
% ------------------------------------------------------------------------------

The symbol table is a mapping from names to types and sizes: 
\begin{code} 
type SymbolTable = Map Name (Type,Int) 
\end{code}

Information needs to be stored into the SymbolTable whenever new intermediate
arrays are created. We have chosen to put this power in the hands of the programmer
using the {\tt sync} function. The {\tt sync} function is overloaded for 
a number of different array types:

%So, how does all the needed information end up in the IC and SymbolTable? 
%This is the responsibility of the {\tt sync} function. The {\tt sync} 
%function is overloaded for a number of different array types:

\begin{code}
class Syncable a where 
    sync :: a -> W a
    commit :: a -> W a 
\end{code}

The types of the {\tt sync} and the related {\tt commit} function
are shown in the class declaration above. To illustrate what {\tt sync} does, 
one instance of its implementation is shown:

\begin{code}
instance TypeOf (Exp a) => Syncable (GArr (Exp a)) where
    sync arr = do
      arr' <- commit arr
      write $ \ix -> [Synchronize]
      return arr'
    commit arr = do 
      let n = len arr
      var  <- newGlobalArray (typeOf (arr ! (E (LitInt 0))))  n 
      write $ \ix -> [(unE (index var ix)) ::= unE (arr ! ix)] 
      return $ mkArray (\ix -> index var ix) n
\end{code}
%$

The {\tt sync} function commits its argument array and 
thereafter writes \newline {\tt [Synchronize]} into the W monad. To see what 
this means, one should also look at the {\tt commit} function, in which
a new array is created of the same size and type 
as the given array. In the next step, an assignment statement is written into 
{\tt W} monad (added to the intermediate code). It
assigns the values computed in the given array to the newly created array.
From the intermediate code and the symbol table accumulated into the {\tt W} 
monad, C code is generated following a procedure outlined in 
figure~\ref{fig:code_gen}.
% ------------------------------------------------------------------------------
%\FloatBarrier
\begin{figure}%
\begin{center}
\includegraphics[scale=.3]{./ifl/stages2.png}
\caption{Steps involved in generating CUDA C code from Obsidian description.
The boxes represent functions and the arrows represent data structures.}
\label{fig:code_gen}
\end{center}
\end{figure}
%\FloatBarrier
% ------------------------------------------------------------------------------

The first step depicted in the figure is the running the Obsidian program. 
This builds two data structures {\tt IC} and {\tt SymbolTable}. 
The IC goes through a simple liveness analysis where for each statement
information about what data elements, in this case arrays,  are alive at 
that point is added. An array is alive if it is used in any of the following 
statements or if it is considered a result of the program. The result of this 
pass is a new {\em IC} where each statement also has a set of names pointing 
out arrays that are alive at that point.

\begin{code} 
type ICLive = [(Statement, Set Name)]
\end{code}

Now, the symbol table together with the {\tt ICLive} object is used to lay out 
the arrays in memory. Arrays that had type {\tt SArr} are assigned storage in 
the shared memory and arrays of type {\tt GArr} in Global memory. The result 
of this stage is a {\em Memory Map}. This is a mapping from names to positions 
in memory. The picture also shows that another output from this stage is 
intermediate code annotated with thread information, call it {\tt ICT}. This 
is now done in a separate pass over the IC but it could be fused with the 
memory mapping stage, saving a pass over the IC. 
The ICT is just a list of statements and the number of threads assigned to 
compute them:

\begin{code}
type ICT = [(Statement,Int)]
\end{code}
% \noindent
This enables the final pass over the IC to move thread information
into the actual IC as conditionals. The resulting IC is used to output
CUDA code.

To illustrate this, the code generated from a simple Obsidian program is 
shown. The example is very artificial and uses sync excessively in order
to create more intermediate arrays :

\begin{code}
rev_add :: GArr IntE -> W (GArr IntE)
rev_add = rev ->- sync ->- fun (+1) ->- sync
\end{code}

Figure~\ref{fig:code1} shows the CUDA C code generated from the {\tt rev\_add} 
program. Here it is visible how intermediate arrays are assigned memory in 
global memory. The global memory is pointed to by {\tt gbase}:

\begin{figure}
\begin{code} 
__global__ static void rev_add(int *source0,char *gbase){
extern __shared__ char sbase[] __attribute__ ((aligned(4)));
const int tid = threadIdx.x;
const int n0 __attribute__ ((unused)) = 256;
((int *)(gbase+0))[tid] = source0[((256 - 1) - tid)];
__syncthreads();
 ((int *)(gbase+1024))[tid] = (((int *)(gbase+0))[tid] + 1);
__syncthreads();
\end{code}
\caption{Generated CUDA code.} 
\label{fig:code1}
\end{figure}

The code in figure~\ref{fig:code1} does however not show how the {\tt ICT} 
is used in assigning work to threads. To show this  a small part of the 
generated CUDA code from the {\tt scan\_add\_kernel} is given in 
figure~\ref{fig:code2}. Notice the conditional {\tt if (tid < 128)}. 
This line effectively shuts down half of the threads. It can also be seen 
here how shared memory is used, pointed to by {\tt sbase}. Moreover, from the 
line with {\tt (63 < 64) ? ...} it becomes clear that there is room for some
obvious optimisations.

\begin{figure}
\begin{code} 
 __syncthreads();
 if (tid < 128){
   ((int *)(sbase+1520))[tid] = ((tid < 64) ?
   ((tid < 64) ?
   ((int *)(sbase+496))[tid] : 
   ((int *)(sbase+752))[(tid - 64)]) : 
   (((63 < 64) ?
   ((int *)(sbase+496))[63] : 
   ((int *)(sbase+752))[(63 - 64)]) + ((tid < 64) ?
   ((int *)(sbase+496))[tid] : 
   ((int *)(sbase+752))[(tid - 64)])));
 }
\end{code} 
\caption{A small part of the CUDA code generated from the recursive implementation of the Sklansky parallel prefix algorithm.}
\label{fig:code2}
\end{figure}
\noindent

% ------------------------------------------------------------------------------
\FloatBarrier

### 2013-09-30 src2/ifl/paper.tex
apa

\documentclass[10pt]{llncs}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}

\usepackage{alltt}
\usepackage{placeins}

\usepackage{fancyvrb}

   \DefineVerbatimEnvironment
     {code}{Verbatim}
     {fontsize=\small,xleftmargin=0.6cm,samepage=true}

\begin{document}

\title{Obsidian:\\A Domain Specific Embedded Language\\for Parallel Programming of Graphics Processors}

\author{Joel Svensson \and Mary Sheeran \and Koen Claessen}
\institute{Chalmers University of Technology, G\"oteborg, Sweden}

\date{}
\maketitle
\begin{abstract}

We present a domain specific language, embedded in Haskell, for general
purpose parallel programming on GPUs. Our intention is to explore the use of
{\em connection patterns} in parallel programming. We briefly present our
earlier work on hardware generation, and outline the current state of GPU
architectures and programming models. Finally, we present the current status
of the {\em Obsidian} project, which aims to make GPU programming easier,
without relinquishing detailed control of GPU resources. Both a programming
example and some details of the implementation are presented. This is a
report on work in progress.

\end{abstract}
\input{intro}
\input{combinators}   %% uncomment when file available
\input{GPUs}           %% I decided to put a separate section on GPUs, rather
                       %% than trying to cover it in the intro.

\input{obsidian}
\input{results}
\input{related}
\input{futurework}
\input{conclusion}
\bibliographystyle{plain}
\bibliography{ifl08}

\end{document}

### 2013-09-30 src2/ifl/paperThesis.tex
apa

--- 2013-09-30 src2/ifl/paper.tex
@@
-\documentclass[10pt]{llncs}
+%\documentclass[10pt]{llncs}
 
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{subfigure}
+%\usepackage{amsmath}
+%\usepackage{graphicx}
+%\usepackage{subfigure}
 
-\usepackage{alltt}
-\usepackage{placeins}
+%\usepackage{alltt}
+%\usepackage{placeins}
 
-\usepackage{fancyvrb}
+%\usepackage{fancyvrb}
 
@@
 
-\begin{document}
+%\begin{document}
 
-\title{Obsidian:\\A Domain Specific Embedded Language\\for Parallel Programming of Graphics Processors}
+%\title{Obsidian:\\A Domain Specific Embedded Language\\for Parallel Programming of Graphics Processors}
 
-\author{Joel Svensson \and Mary Sheeran \and Koen Claessen}
-\institute{Chalmers University of Technology, G\"oteborg, Sweden}
+%\author{Joel Svensson \and Mary Sheeran \and Koen Claessen}
+%\institute{Chalmers University of Technology, G\"oteborg, Sweden}
 
-\date{}
-\maketitle
-\begin{abstract}
+%\date{}
+%\maketitle
+
+\section*{Abstract}
 
@@
 
-\end{abstract}
-\input{intro}
-\input{combinators}   %% uncomment when file available
-\input{GPUs}           %% I decided to put a separate section on GPUs, rather
+%\end{abstract}
+\input{./ifl/intro}
+\input{./ifl/combinators}   %% uncomment when file available
+\input{./ifl/GPUs}           %% I decided to put a separate section on GPUs, rather
                        %% than trying to cover it in the intro.
 
-\input{obsidian}
-\input{results}
-\input{related}
-\input{futurework}
-\input{conclusion}
-\bibliographystyle{plain}
-\bibliography{ifl08}
+\input{./ifl/obsidian}
+\input{./ifl/results}
+\input{./ifl/related}
+\input{./ifl/futurework}
+\input{./ifl/conclusion}
+%\bibliographystyle{plain}
+%\bibliography{ifl08}
 
-\end{document}
+%\end{document}

### 2013-09-30 src2/ifl/related.tex
apa

%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% RELATED WORK
%% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Discussion} \label{sec:disc}
\subsection{Our influences}
As mentioned above, our earlier work on Lava has provided the inspiration for
the combinator-oriented or hardware-like style of programming that we are exploring in Obsidian.
On the other hand, the {\em implementation} of Obsidian has been much influenced by
{Pan}, an embedded language for image synthesis developed by 
Conal Elliot~\cite{PAN}. Because of the computational complexity of image generation, C 
code is generated. This C code can then be compiled by an optimising compiler. 
Many ideas from the paper 
``Compiling Embedded Languages'', describing the implementation of Pan have been 
used in the implementation of Obsidian~\cite{Elliott03:CompileDSEL-JFP}.

\subsection{Related work on GPU and GPGPU programming languages}
We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}. Lucas distinguishes carefully between languages (such as CG and HLSL) that aim to raise the level of abstraction at which graphics-oriented GPU programs are written, and those that attempt to abstract the entire GPU, and so must also provide a means to express the placing of programs on the GPU, feeding such programs with data, reading the results back to the CPU, and so on, as well as deciding to what extent the programmer should be involved
in stipulating those tasks.
In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
PyGPU is a language for image processing embedded in Python~\cite{LUNDPAPER}. 
PyGPU uses the introspective abilities of Python and thus bypasses
the need to implement new loop structures and conditionals for the embedded 
language. In Python it is possible to access the bytecode of a function and 
from that extract information about loops and conditionals. 
Programs written in PyGPU can be compiled and run on a GPU. 
Vertigo is another embedded language by Conal Elliot~\cite{VERTIGO}. It
is a language for 3D graphics that targets the DirectX 8.1 shader model, and
can be used to describe geometry, shaders and to generate textures. 
%Each sublanguage is given formal semantics \cite{VERTIGO}. From programs 
%written in Vertigo, assembly language programs are generated for execution on a 
%GPU.

The more general purpose languages aim to abstract away from the graphics heritage of GPUs, and target a larger group of programmers.
The thesis by Lucas presents CGiS, an imperative data-parallel programming
language that targets both GPUs and SIMD capable CPUs -- with the aim being
a combination of a high degree of abstraction and a close resemblance to traditional programming languages~\cite{Lucas08}.
BrookGPU (which is usually just called Brook) is a classic example of a language~\cite{Brook} designed to raise the level of abstraction at which GPGPU programming is done. It is an extension of C with embedded kernels, aimed at arithmetic-intense data parallel computations. C is used to declare streams\footnote{Brook is referred to as a ``stream processing'' language, but this means something different from what the reader might expect: a stream in this context is
a possibly multi-dimensional array of elements, each of which can be processed separately, in parallel.}, CG/HLSL (the lower level GPU languages) to declare kernels, while
function calls to a runtime library direct the execution of the program.
Brook had significant impact in that it raised the level of abstraction at which GPGPU programming can be done. The language Sh also aimed to raise the level of abstraction at which GPUs were programmed~\cite{Sh}. Sh was an embedded language in C++, so our work is close in spirit to it. Sh has since evolved into the RapidMind development platform~\cite{RapidMind}, which now supports multicores and Cell processors as well as GPUs. 
The RapidMind programming model has arrays as first class types. It
has been influenced by functional languages like NESL and SETL, and its program objects are pure functions. Thus it supports both functional and imperative styles of programming.
A recent PhD thesis by Jansen asserts that there are some problems with RapidMind's use of macros to embed the GPU programming language in C++, including the inability to pass
kernels (or shader programs) as classes~\cite{JansenThesis}; the thesis proposes GPU++ and claims improvement over previous approaches, particularly through the exploitation of automatic partitioning of the programs onto the available GPU hardware, and through compiler optimisations that improve runtime performance.

Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\cite{Accelerator}.
Data Parallel Haskell~\cite{DPH} incorporates Nested Data Parallelism in the style of NESL~\cite{NESL} into Haskell. GPUGen, like Obsidian, aims
to support GPGPU programming from Haskell~\cite{GPUGen}. It works by translating Haskell's intermediate language, Core, into CUDA, for collective data operations such as scan, fold and map. The intention is to plug GPUGen into the Nested Data Parallel framework of the Glasgow Haskell Compiler. Our impression is that we wish to expose considerably more detail about the GPU to the programmer, but we do not yet have sufficient information about GPUGen to be able to do a more complete comparison. Finally, we mention the Spiral project, which develops methods and tools for automatically generating high performance libraries for a variety of platforms, in domains such a signal processing, multiplication and sorting~\cite{Spiral}.
The tuning of an algorithm for a given platform is expressed as an optimisation problem, and the domain specific mathematical structure of the algorithm is used
to create a feedback-driven optimiser. The results are indeed impressive, and we feel that the approach based on an algebra of what we would call combinators
will interest functional programmers. We hope to experiment with similar
search and learning based methods, having applied similar ideas in the simpler setting of arithmetic data-path generation in Lava.

\subsection{Lessons learned so far in the project}
Our first lesson has been the gradual realisation that a key aspect of a usable
GPU programming language that exposes details of the GPU architecture to the user is the means to express where and when data is placed in and
read from the memory hierarchy.
We are accustomed, from our earlier experience in hardware design, to describing
and generating networks of communicating components -- something like data-flow graphs. We are, however, unused to needing to express choices about the use
of the various levels in a memory hierarchy. We believe that we need to develop programming idioms and language support for this. It seems likely, too, that such idioms will not be quite as specific to GPU programming as other aspects of our embedded language development. How to deal with control of access to
a memory hierarchy in a parallel system seems to be a central problem that must be tackled if we are to develop better parallel programming methods in general.
A typical example of a generic approach to this problem is
the language Sequoia, which aims to provide programmers with a means
to express how the memory hierarchy is to be used, where a relatively abstract description of
the platform, viewed as a tree of processing nodes and memories, is a parameter~\cite{Sequoia}. Thus, programmers should write very generic code, which can be compiled for many different platforms. This kind of platform independence is not our aim here, and we would like to experiment with programming idioms for control of memory access for the particular case of a CPU plus some form of highly parallel co-processor that accelerates some computations.

A second lesson concerns ways to think about synchronisation on the GPU. We naively assumed that {\tt sync} would have nice compositional behaviour, but we have found that in reality one can really only sync at the top level.
The reason why the CUDA code generated from the {\tt sklanky} example works poorly on the GPU is that it uses {\tt syncs} in a way that leads to unwanted
serialisation of computations. Looking at our generated code, we see that it may be possible to make major improvements by being cleverer about the placement
of {\tt sync}s. For instance, the semantics of {\tt two} guarantees that the two components act on distinct data, and this can be exploited in the placing of {\tt sync}s in the generated code.

Finally, we have found that we need to think harder about the two levels of
abstraction: writing the kernels themselves and kernel coordination. This paper concerned the kernel level. We do not yet have a satisfactory solution to the question of how best to express kernel coordination. This question is closely related to that about how to express memory use.

### 2013-09-30 src2/ifl/results.tex
apa

\section{Results} \label{sec:results}

Apart from the parallel prefix algorithm shown in this paper we have used
Obsidian to implement sorters. For the sorters, the generated C code 
performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\cite{LavaSorter},
has a running time of 95$\mu$secs for 256 elements (the largest size that we can cope with in a single kernel). This running time can be compared to the 
 28$\mu$secs running time of the bitonic sort example supplied with the CUDA 
SDK. However, for 256 inputs, bitonic sort has a depth (counted in number of comparators between input and output), of 36 compared to 
Vsort's 64. So  we feel confident that we can make a sorter that improves considerably on our Vsort by implementing a recursive algorithm for which the corresponding network depth is less.
(We implemented the periodic sorter in an earlier version of the system, in which recursion was not available.) The point here is not to be as fast as hand-crafted library code, but to come close enough to allow the user to quickly construct short readable programs that give decent performance. The results for sorting are promising in this respect.
Sadly, the results for the Sklansky example are rather poor, and we will return to this point in the following section.

The programs reported here were run on an NVIDIA 8800GTS and timed using 
the CUDA profiler.

### 2013-09-30 src2/papp/paperThesis.tex
apa

%\documentclass[3p,times]{elsarticle}

%\usepackage{ecrc} % [headings]{espcrc1}

%\volume{00}

%\firstpage{1}

%\journalname{Procedia Computer Science}

%\jid{procs}

%\jnltitlelogo{Procedia Computer Science}
%\setlength{\parindent}{0pt}
%\setlength{\parskip}{1ex plus 0.5ex minus 0.2ex}

%\usepackage{color}
%\usepackage{graphicx}
%\usepackage{subfigure}
%\usepackage{placeins}
%\usepackage[font=sf]{caption}
%\usepackage[figuresright]{rotating}

% set the starting page if not 1
% \setcounter{page}{17}

% add words to TeX's hyphenation exception list
%\hyphenation{author another created codependent financial paper re-commend-ed Post-Script}

%-------------------------------------------------------------------------------

%\title{Obsidian: GPU Computing Using Haskell (DRAFT)} 
%\author{Joel Svensson, Koen Claessen and Mary Sheeran} 
%\runauth{Joel Svensson, Koen Claessen and Mary Sheeran}

%% \date{\today}

% \runtitle{Obsidian: GPU Computing Using Haskell}
%\runauth{Joel Svensson, Koen Claessen and Mary Sheeran}

%\begin{document}
%\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

%\dochead{}
%% Use \dochead if there is an article header, e.g. \dochead{Short communication}

%\title{GPGPU Kernel Implementation and Refinement using Obsidian}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

%\author{Joel Svensson}
%\author{Koen Claessen}
%\author{Mary Sheeran}
%\affiliation{Chalmers University of Technology}

%\address{CSE Dept., Chalmers University of Technology, Gothenburg, Sweden}

\section*{Abstract}
Obsidian is a domain specific language for data-parallel programming 
on graphics processors (GPUs). It is embedded in the functional 
programming language Haskell. The user writes code using constructs 
familiar from Haskell (like {\tt map} and {\tt reduce}), recursion 
and some specially designed combinators for combining GPU programs. 
NVIDIA CUDA code is generated from these high level descriptions, and 
passed to the {\tt nvcc} compiler~\cite{CUDAGuide2.0}. Currently, we 
consider only the generation of single kernels, and not their coordination.

This paper is focussed on how the user should work with Obsidian, 
starting with an obviously correct (or well-tested) description of the 
required function, and refining it by the introduction of constructs to 
give finer control of the computation on the GPU. For some combinators, 
this approach results in CUDA code with satisfactory performance, promising 
increased productivity, as the high level descriptions are short and 
uncluttered. But for other combinators, the performance of generated code 
is not yet satisfactory. Ways to tackle this problem and plans to integrate 
Obsidian with another higher-level embedded language for GPU programming in 
Haskell are briefly discussed.

%Obsidian is a domain specific language for Data-parallel 
%programming on graphics processors (GPUs) embedded in Haskell. 
%From high level descriptions, code for execution on a GPU is generated.
%Obsidian generates NVIDIA CUDA code that is passed to the {\tt nvcc} 
%compiler \cite{CUDAGuide2.0}.

%This paper is mainly focused on the usage of Obsidian. The implementation
%of Obsidian itself and a discussion of how the CUDA code is generated can 
%be found in \cite{JSTECH}.

%\vspace{1pc}
%\end{abstract}

%\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
%Data-parallel \sep Embedded language \sep GPUs \sep Haskell  
%% PACS codes here, in the form: \PACS code \sep code

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

%\end{keyword}

%\end{frontmatter}

%-------------------------------------------------------------------------------
\section{Introduction}

Multicore and manycore processors are becoming increasingly common. Modern 
graphics processing units (GPUs) are examples of manycore processors. Today 
GPUs come with hundreds of processing elements capable of managing thousands 
of threads \cite{Brief8800}. The Obsidian project is about exploring ways 
to program these new machines.

Obsidian is a domain specific language for general purpose programming on 
GPUs (GPGPU), capable of generating code for modern NVIDIA GPUs. 
In~\cite{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
interface. In the current version of Obsidian the monad has been replaced 
by another data structure, closely related to Arrows~\cite{Arrows}, representing GPU programs. Although the details are 
left out because of space limitations, the use of this GPU program representation 
can be seen in section~\ref{sec:GPUPrograms}. For a description of the implementation 
see~\cite{JSTECH} and for more general information on embedded language 
implementation see~\cite{Elliott03:CompileDSEL-JFP}.

GPU design is driven by the performance demands of graphics applications. 
The kind of processing that is common in graphics falls in the data-parallel 
category \cite{GEMS2}.
% One example of a computation from graphics is the 
%transformation of vertices between different coordinate systems. That is, 
%each vertex is multiplied by the same transformation matrix, fitting  the 
%data-parallel paradigm perfectly.

%-------------------------------------------------------------------------------
\subsection{NVIDIA GPUs and CUDA}

Starting with the 8000 series of GPUs (the G80 architecture), NVIDIA's 
GPUs came with a unified architecture, meaning that all the 
processing elements on the GPU are of the same kind. This was different from 
the previous generation's GPUs, which typically had two kinds of 
processing elements, fragment and vertex processors.  
%There was one kind of processor designed to process fragment/pixel programs and another to process vertex data. 
Now these two kinds 
of processors are replaced by a single kind with capabilities surpassing both 
of the old ones. This new unified architecture together with development tools 
for GPGPU programming go under the 
name CUDA (Compute Unified Device Architecture) 
\cite{CUDAGuide2.0}. CUDA offers the GPGPU programmer a C %,wwwcuda,Brief8800
compiler and libraries for CUDA enabled GPUs (NVIDIA 8000 series and above).

Figure~\ref{fig:gpu}, shows a conceptual picture of a CUDA enabled GPU. The GPU 
has a number of {\em Multiprocessors} (MP in the picture). These MPs contain 
a number of small processing elements called {\em Streaming Processors} (SP). 
These SPs operate in an SIMD fashion. All SPs in a given MP execute 
the same instruction each clock cycle.

Each MP is capable of maintaining a large number of threads in flight at the 
same time. A group of threads running on an MP is referred to as a {\em block}. 
A block can contain more threads then there are processing elements; today a 
block can hold up to 1024 threads. There is a scheduler mapping these threads 
over the SPs available. Threads are scheduled in groups called {\em warps} 
consisting of 32 threads that are executed in an SIMD fashion on the MPs.
Conditionals that take different paths within a warp have a negative effect 
on performance; the two diverging paths will in fact be executed sequentially 
\cite{BestPrac}.
%There are however, more threads in a
%warp then there are SPs, so the SPs are switching threads to work on 
%between every instruction. This means that conditionals that
%take different paths within a warp have a negative effect on performance;
%the two diverging paths will in fact be executed sequentially \cite{BestPrac}.

The MPs also have local memory that is shared between all the 
SPs of that MP, and thus referred to as {\em Shared Memory} in the figure. 
It can be used to exchange information between threads running 
on the SPs; it can also be used as a software managed cache. 
Threads within a warp can safely communicate using shared memory without 
the need for any synchronisation. Threads from different warps, however, need 
to use a synchronisation primitive to ensure a coherent view of the shared 
memory. In CUDA C this primitive is called {\tt \_\_syncthreads()}. The
{\tt \_\_syncthreads()} primitive provides a barrier that all threads of a 
block must reach before any is allowed to proceed.

The {\em Global memory} is located off chip and is accessible by all MPs. 
In current GPUs, accesses to this memory are uncached.

\begin{figure}
\begin{center}
\includegraphics[width=.65\linewidth]{./papp/pictures/gpu.jpg}
\caption{Conceptual image of a CUDA enabled GPU.}
\label{fig:gpu}
\end{center}
\end{figure}

\FloatBarrier

% ------------------------------------------------------------------------------
\subsection{Aims of Obsidian}

The initial goal of Obsidian is to simplify the development of GPU kernels, 
the building blocks of larger GPU programs. In the future, ways to combine 
kernels into larger algorithms will also be explored.

When implementing an algorithm in CUDA, what to compute and how to compute it 
become codependent. Choices such as how many elements the kernel operates on 
and how many threads it uses affect each other greatly. Once a kernel 
is designed with a particular number of elements per thread, this becomes hard to tweak. In CUDA the programmer writes a single program 
parameterised over a thread identity. This program is then executed by 
several threads. Taking this point of view often means that what elements to use 
needs to be computed from the threadIDs. Coming up with these indexing 
computations is not always trivial.

In Obsidian the program is described as 
a computation between arrays, and combinators are used instead of direct indexing 
into structures. 
Obsidian provides an environment where it is easy to experiment with different
partitionings and choices when implementing an algorithm. It is 
possible to write a simple running first prototype version of a kernel 
without thinking about architectural details. The prototype implementation 
can then be refined into a more efficient implementation. 
The aim of Obsidian is to raise the level of abstraction for the GPGPU 
programmer and to relieve the programmer of details such as laying things out 
in memory. Performance affecting decisions should be easy to make 
and change without major rewrites of the code.

%-------------------------------------------------------------------------------
\section{Programming in Obsidian}

Obsidian is a language for GPGPU programming embedded in Haskell. Many of 
the language features resemble those of Lava, a hardware description and 
verification language \cite{lavaICFP}. The justification to use language 
constructs similar to that of a hardware description language came from 
the observation that GPGPU algorithms often were explained using circuit-like 
pictures, see for example \cite{GEMS3}.

%Obsidian can be used to describe hardware like algorithms. These are algorithms
%where the number of inputs and outputs are fixed, not dependent on 
%the values of inputs. 
%There are a large number of algorithms falling into this category. For example, 
%there are numerous sorting algorithms (sorting networks) and parallel prefix 
%algorithms implementable in this way. Also, there are examples in the 
%literature where GPGPU programmers implement their kernels to operate on very 
%specific array sizes \cite{GEMS3}. In a later stage, when the kernels are 
%composed into algorithms on large arrays, support for variable length 
%arrays is introduced.

Obsidian can be explained as two sub-languages. First there is a language of 
arrays and operations on arrays, and second, a language that enables mapping of the 
array language programs onto the GPU.

%-------------------------------------------------------------------------------
\subsection{Array Language}
\label{sec:ArrayLanguage}
\FloatBarrier
Arrays in Obsidian do not, like in C, name an area of memory. Instead, an 
array is represented by an abstract data type, called {\tt Arr}, supporting 
the following operations: {\tt (!)} for indexing, {\tt len} 
returns the length of an array and {\tt mkArr} that given a function from 
indexes to elements and a length gives an array. For example using these 
operations reversing an array can be accomplished as follows:

\begin{small}
\begin{verbatim}
rev :: Arr a -> Arr a 
rev arr  = mkArr ixf n
    where 
        ixf ix = arr ! (fromIntegral (n-1) - ix)
        n = len arr 
\end{verbatim}
\end{small}
\noindent
Array reversal is polymorphic in the element type which gives the {\tt rev} 
program the type {\tt Arr a -> Arr a}. 
Internally an array is represented by the computation that gives its elements. 
The array type consists of two parts, a function from indices to elements and an 
integer representing its length:
%-------------------------------------------------------------------------------
\begin{small}
\begin {verbatim} 
data Arr a = Arr (IndexE -> a) Int 
\end{verbatim}
\end{small}
%-------------------------------------------------------------------------------
%An array is a function from an index expression, {\tt IndexE}, to elements.
\noindent 
The length of the array is static, known at compile time, and is represented
by an {\tt Int}. The elements of an array can be {\tt Int}, {\tt Float} or 
{\tt Bool} valued expressions, represented by the types {\tt IntE}, 
{\tt FloatE} and {\tt BoolE}. Arrays can also contain arrays and tuples 
as elements.

Obsidian provides a number of functions on this array type. For example, 
a function can be mapped over an array using {\tt fmap}. 
%see figure~\ref{fig:zipunzip}:

%-------------------------------------------------------------------------------
\begin{small}
\begin{verbatim}
fmap :: (a -> b) -> Arr a -> Arr b
\end{verbatim}
\end{small}
%-------------------------------------------------------------------------------
\noindent
The array type is also an instance of {\tt Foldable}, so there is a function 
{\tt foldr} (often called {\tt reduce}) defined on arrays:

\begin{small}
\begin{verbatim}
foldr :: (a -> b -> b) -> b -> Arr a -> b
\end{verbatim}
\end{small}
\noindent
Two other basic functions on arrays that are available are {\tt pair} and 
{\tt unpair}: 
\begin{small}
\begin{verbatim}
pair :: Arr a -> Arr (a,a)
unpair  :: Choice a => Arr (a,a) -> Arr a
\end{verbatim}
\end{small}
\noindent
The function {\tt pair} takes an array and returns an array of pairs where 
the first element of the input array is paired up with the second, the 
third with the forth and so on. The {\tt unpair} function does the 
opposite. % see figure~\ref{fig:pairunpair}.

The {\tt Choice} class contains those types that have an {\tt ifThenElse} 
function defined on them:  
\begin{small}
\begin{verbatim}
ifThenElse :: Choice a => BoolE -> a -> a -> a
\end{verbatim}
\end{small}
\noindent
Another example of a function given in the array library is {\tt zipp} of type{\tt (Arr a, Arr b) -> Arr (a, b)}.
%\begin{small}
%\begin{verbatim}
%zipp :: (Arr a, Arr b) -> Arr (a, b)
%\end{verbatim}
%\end{small}
This function performs on arrays what the normal Haskell {\tt zip} 
does on lists. However, the input to {\tt zipp} is a pair of arrays.

%Array programs like those described here make up the building blocks used
%to form larger GPU programs. How to create a GPU program from these is 
%shown in the following section.

%-------------------------------------------------------------------------------

%\FloatBarrier
\subsection{GPU Programs}
\label{sec:GPUPrograms}

The second layer of Obsidian offers a data type that represents a GPU program
with input {\tt a} and output {\tt b}:

\begin{small}
\begin{verbatim} 
data a :-> b = ... 
\end{verbatim}
\end{small} 
\noindent
%The details of {\tt a :-> b} are shown in section~\ref{sec:Implementation}. 
\noindent
Informally we can think of {\tt a\nolinebreak~:->\nolinebreak~b} as representing programs that operate 
as illustrated in figure~\ref{fig:program}. This figure 
shows a program that performs some computation using a number of threads 
followed by a barrier synchronisation, and so on. The contents of the 
boxes marked with {\em Pure} can be thought of as containing an array 
program.% such as those in section~\ref{sec:ArrayLanguage}.
The type constructor {\tt :->} is a variant of an arrow, which in turn is a generalisation of a monad~\cite{JSTECH}.
%
\begin{figure}
\begin{center}
\includegraphics[width=.60\linewidth]{./papp/pictures/prg_intuit.jpg}
\caption{A GPU program of type  {\tt a :-> b} can be thought 
of as pure computations interspersed by syncs.}
\label{fig:program}
\end{center}
\end{figure}
\noindent
One way to create a GPU program is by using the function {\tt pure} of type {\tt (a -> b) -> a :-> b}.
%\begin{small}
%\begin{verbatim}
%pure :: (a -> b) -> a :-> b
%\end{verbatim}
%\end{small}
For example the array language program {\tt fmap (+1)}, that increments 
every element of an array, can be lifted to a GPU program: 
%
\begin{small}
\begin{verbatim}
incr :: Arr IntE :-> Arr IntE 
incr = pure $ fmap (+1)
\end{verbatim}
\end{small} 
 \noindent
GPU programs can be executed on the GPU from
{\em GHCi} (the Haskell interpreter) using the function {\tt execute}:

\begin{small}
\begin{verbatim}
execute :: (Flatten a, Flatten b) => (Arr a :-> Arr b) -> [a] -> IO [b]
\end{verbatim}
\end{small}
\noindent
%The class {\tt Flatten} will be explained in detail in 
%section~\ref{sec:Implementation}, but 
Instances of {\tt Flatten} are all the types that can be stored in the 
GPU memory. Examples of types that are in {\tt Flatten} are {\tt IntE}, 
{\tt FloatE}, {\tt BoolE}. Arrays and pairs of things that are in Flatten 
are also instances of {\tt Flatten}.
Here, {\tt execute} is used in order to run an instance of the {\tt incr} 
program on the GPU:

\begin{small}
\begin{verbatim}
*Obsidian> execute incr [0..9]
[1,2,3,4,5,6,7,8,9,10]
\end{verbatim}
\end{small}
\noindent
The elements of the Haskell list given to {\tt execute} are used to create 
an input array to the kernel. Following this, the kernel is executed on the
GPU and the result is read back and presented as a Haskell list.

In the example above, the {\tt execute} function generates the following kernel 
corresponding to the given GPU program:
\begin{small}
\begin{verbatim}
__global__ void generated(word* input,word* result){
  unsigned int tid = (unsigned int)threadIdx.x;
  extern __shared__ unsigned int s_data[];
  word __attribute__((unused)) *sm1 = &s_data[0];
  word __attribute__((unused)) *sm2 = &s_data[0];
  ix_int(result,tid) = (ix_int(input,tid) + 1);
}
\end{verbatim}
\end{small}
\noindent
The generated kernel has two arguments, an input array of words and an 
output array of words. Words represent 32-bit quantities that can be either
floating point or integer valued. This particular kernel does not use 
any shared memory; the incremented values are stored directly into the 
result array that resides in global memory.

Given two GPU programs, {\tt f} and {\tt g}, of suitable types, a 
composite GPU program can be created by passing the output of {\tt f} to the 
input of {\tt g}. In Obsidian this is done using the composition operator (or combinator)
{\tt (->-)}:

\begin{small}
\begin{verbatim}
(->-) :: (a :-> b) -> (b :-> c) -> (a :-> c)
\end{verbatim}
\end{small}
\noindent
The following illustrates the use of {\tt (->-)} by implementing a 
program that increments every element of an array but also reverses 
the array:

\begin{small}
\begin{verbatim}
increv :: Arr IntE :-> Arr IntE 
increv = pure (fmap (+1)) ->- pure rev

*Obsidian> execute increv [0..9]
[10,9,8,7,6,5,4,3,2,1]
\end{verbatim}
\end{small}
\noindent
The code generated from the {\tt increv} program is very similar to that 
of {\tt inc} but the indexing is reversed:

\begin{small}
\begin{verbatim}
__global__ void generated(word* input,word* result){
  unsigned int tid = (unsigned int)threadIdx.x;
  extern __shared__ unsigned int s_data[];
  word __attribute__((unused)) *sm1 = &s_data[0];
  word __attribute__((unused)) *sm2 = &s_data[0];
  ix_int(result,tid) = (ix_int(input,(9 - tid)) + 1);
}
\end{verbatim}
\end{small}
\noindent
The code generated for the {\tt incr} and {\tt increv} examples uses 
$10$ threads to compute the resulting array. By default, the result will 
be computed using a number of threads equal to the number of elements in 
the return array.

The {\tt increv} program can also be specified with an explicit storing 
of intermediate values between the {\tt rev} and the {\tt fmap (+1)}. 
This is accomplished using a primitive GPU program called {\tt sync}:

\begin{small}
\begin{verbatim}
sync  :: Flatten a => Arr a :-> Arr a

increv :: Arr IntE :-> Arr IntE 
increv = pure (fmap (+1)) ->- sync ->- pure rev
\end{verbatim}
\end{small} 
\noindent
This version of {\tt increv} computes the same result as the previous one. 
However, it does so by computing {\tt fmap (+1)} on the array, storing the
intermediate result in shared memory followed by computing the reverse.
The CUDA C code for this version of {\tt increv} looks like this. Notice 
how the shared memory is used and the call to {\tt \_\_syncthreads()}: 
%
\begin{small}
\begin{verbatim}
__global__ void generated(word* input,word* result){
  unsigned int tid = (unsigned int)threadIdx.x;
  extern __shared__ unsigned int s_data[];
  word __attribute__((unused)) *sm1 = &s_data[0];
  word __attribute__((unused)) *sm2 = &s_data[8];
  ix_int(sm1,tid) = (ix_int(input,tid) + 1);
  __syncthreads();
  ix_int(result,tid) = ix_int(sm1,(7 - tid));
}
\end{verbatim}
\end{small}

\section{Outline of Code Generation}
\label{sec:CodeGen}
\FloatBarrier

This section briefly reviews the code generation process using 
a hyphothetical program as example. The program under consideration 
is this:

\begin{small}
\begin{verbatim}
prg = pure (fmap f) ->- sync -> rev ->- sync ->- pure (fmap g) 
\end{verbatim}
\end{small}

This program applies some function {\tt f} to all elements of 
an array. The array is then reversed and lastly a function {\tt g} 
is applied to all elements. Between every operation there is a {\tt sync}.
The program {\tt prg} is an element of the datatype {\tt (a :-> b)}. 
To see what happens during code generation, it is necessary to know 
the implementation of this type.
\begin{small}
\begin{verbatim} 
data a :-> b 
   = Pure (a -> b) 
   | Sync (a -> Arr FData) (Arr FData :-> b) 
\end{verbatim}
\end{small} 
\noindent
The type {\tt (a :-> b)} is essentially a list of computations. The 
function {\tt pure} is directly implemented using the constructor 
{\tt Pure}. The function {\tt sync} is implemented in terms of the 
constructor {\tt Sync}:

\begin{small}
\begin{verbatim} 
sync :: Flatten a => (Arr a :-> Arr a)
sync = Sync (fmap toFData) (pure (fmap fromFData))
\end{verbatim}
\end{small} 
\noindent
The example program {\tt prg} would be represented as follows: 
%
\begin{small}
\begin{verbatim} 
Sync (fmap (toFData . f)) (Sync ((fmap toFData) . rev . (fmap fromFData)) (Pure (fmap g)))   
\end{verbatim}
\end{small} 
\noindent
From this representation, which is essentially the list {\tt [fmap f, rev, fmap g]},
the CUDA code is generated as outlined in figure~\ref{fig:codegen}. 
An input array is created and applied to the first stage of the computation, 
{\tt (fmap f)}, resulting in: 
\begin{small}
\begin{verbatim}
Arr (\ix -> f(index (variable ``input'') ix)) n. 
\end{verbatim}
\end{small}

The information given by this array is used to construct the C assignment 
statement {\tt sm1[tid] = f(input[tid])} as seen in the figure.

Following this a new Obsidian level array that indexed into {\tt sm1} is 
created and used as input to the second stage of the computation.

%Following this a new Obsidian level array that looks as follows is created: 
%\begin{small}
%\begin{verbatim}
%Arr (\ix -> index (variable ``sm1'') ix) n. 
%\end{verbatim}
%\end{small}

%This expression is used to generate a 
%C assignment. The array is stored into one of two shared memory arrays 
%This array  is used as input to the next phase. The computation is then double 
%buffered between the two shared memory arrays. 
%giving an expression.

\begin{figure}
\begin{center}
\includegraphics[width=.65\linewidth]{./papp/pictures/codegenexample.jpg}
\caption{Outline of the code generation procedure.}
\label{fig:codegen}
\end{center}
\end{figure}

%-------------------------------------------------------------------------------

\section{Case Studies}
\label{sec:CaseStudies}

\FloatBarrier
This section describes a few slightly larger Obsidian programs. The main purpose of this section is to show 
how we want to write GPGPU programs using Obsidian. 
%How the performance 
%of the generated code can be improved is discussed in section~\ref{sec:FutureWork}  
%%% Move this later.  XXX

\subsection{Sorting}

Sorting is a popular function to place on the GPU, see~\cite{SORT1,SORT3}.
{\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
merger merges two sorted arrays. In Obsidian this merger and sorter 
can be implemented using the combinators {\tt two}, {\tt ilv}, {\tt evens}
and {\tt odds}. For a detailed description of these combinators see
~\cite{lavaSorter}.

\begin{small}
\begin{verbatim}
mergeOE :: (Choice a, Flatten a) => Int -> ((a,a) -> (a,a)) -> (Arr a :-> Arr a)
mergeOE 1 f = pure (evens f)
mergeOE n f = ilv (mergeOE (n-1) f) ->- sync ->- pure (odds f)
\end{verbatim}
\end{small}
\noindent
This is actually a pattern into which any two-input two-output function
can be plugged. However, we choose the following {\em comparator} function (a compare and swap operation). The resulting merger can be executed on the GPU.
\begin{small}
\begin{verbatim}
cmp :: (Ordered a, Choice (a, a)) => (a, a) -> (a, a)
cmp (a,b) = ifThenElse (a <* b)  (a,b) (b,a)

*Obsidian> execute GPU (mergeOE 3 cmp) ([1,3,5,7,2,4,6,8] :: [IntE]) 
[1,2,3,4,5,6,7,8]
\end{verbatim}
\end{small}
\noindent
Given this merger, the sorter is implemented by recursively sorting 
two halves of the input array followed by an application of the merger:

\begin{small}
\begin{verbatim}
sortOE :: Int -> (Arr IntE :-> Arr IntE)
sortOE  0 =  pure id
sortOE  n =  two (sortOE (n-1)) ->- sync ->- mergeOE n cmp

*Obsidian> execute GPU (sortOE 3) [6,0,1,3,4,2,5,7]
[0,1,2,3,4,5,6,7]
\end{verbatim}
\end{small}

The performance of the code generated from this sorter specification is poor. It
runs about three times slower than the example Bitonic
sorter distributed with the CUDA SDK. 
The reason is that nested applications of the {\tt two} and {\tt ilv} 
combinators are hard to optimise. We expect that single (parameterised) combinators corresponding
to particular nestings of these combinators will bring significant performance improvement.
%
%The reason for the poor performance in this case is the code 
%condition of the code generated from the deep nestling of applications of 
%{\tt two} and {\tt ilv} that the code generator currently does a very bad job 
%at optimising. There may be alternative ways to implement these combinators 
%or entirely different ones should replace them.

%-------------------------------------------------------------------------------
\subsection{Parallel Prefix}
\FloatBarrier

This subsection shows the implementation of a parallel prefix (also called scan) 
kernel, known as {\tt sklansky} after J. Sklansky \cite{Sklansky:1960:CSA}.
This kernel will then be optimised step-by-step using Obsidian. This 
optimisation effort uses an experimental feature of Obsidian called {\tt syncHow}. 
This is different from the normal {\tt sync} in the way it assigns work to threads. 
For a more thorough explanation of {\tt syncHow} see ~\cite{JSTECH}.
For an explanation of the parallel-prefix operation see~\cite{BlellochTR90}.

The sklansky parallel prefix algorithm 
is implemented by splitting the inputs in two halves and recursively 
applying sklansky to both halves. The two sub-results are then 
joined by applying the operation between the highest index of the first
sub-result and all the elements in the second sub-result, this is done 
using a function called {\tt fan}:

\begin{small}
\begin{verbatim}
fan op arr = conc (a1, (fmap (op c) a2)) 
    where (a1,a2) = halve arr
          c       = a1 ! (fromIntegral (len a1 - 1))
\end{verbatim}
\end{small}
\noindent
The {\tt sklansky} function is now implemented using 
{\tt two} and {\tt fan}:

\begin{small}
\begin{verbatim}
sklansky :: (Flatten a, Choice a) => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
sklansky 0 op = pure id
sklansky n op = two (sklansky (n-1) op) ->- pure (fan op) ->- sync
\end{verbatim}
\end{small}

Executing ({\tt sklansky 3 (+)}) on the GPU given input {\tt [0..7]} computes
the prefix sum as expected, returning {\tt [0,1,3,6,10,15,21,28]} 
%Executing a small instance of the prefix network above on the GPU has 
%the following result:

%\begin{small}
%\begin{verbatim}
%*Obsidian> execute (sklansky 3 (+)) ([0..7] :: [IntE])
%[0,1,3,6,10,15,21,28]
%\end{verbatim}
%\end{small}

If {\tt sklansky} is used to generate code for an array of 512 elements,
it uses 512 threads to calculate the prefix sums. However, the number of 
applications of the {\tt op} operator that is needed in any stage of the 
algorithm is only 256. This indicates that 
using 256 threads to compute the result would give more efficient
use of the GPU's resources.

Since the code generated by Obsidian is not by default {\em in-place} with regard 
to shared memory, each thread in the 256 threaded program needs to both perform 
the operation between two elements and copy one value unchanged. This is desired
because then each thread performs the exact same operations,
which means that there is no risk for divergence within a warp.

This perfect division of labour is not obtainable with the current 
implementation of the {\tt How} argument to {\tt sync}. However, one 
division of the work that has experimentally been shown to perform well, see 
table in next section,  
is to let each thread $tid$ perform the work of $tid$ and $tid + 256$. This 
program is shown below:

\begin{small}
\begin{verbatim}
sklansky1 :: (Flatten a, Choice a) => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
sklansky1 0 op = pure id
sklansky1 n op = two (sklansky1 (n-1) op) ->- pure (fan op) ->- syncHow (pairNth 256) 
\end{verbatim}
\end{small}

Another optimisation to apply to the sklansky algorithm is the removal of 
unnecessary {\tt \_\_syncthreads()} calls from the generated code, this is 
done by using {\tt syncWarp}. It is up to the programmer to ensure that it 
is safe to use {\tt syncWarp}. For a sklansky network of size 32, it should 
be safe to leave out the {\tt \_\_syncthreads} since all of the communication 
stays within a warp. Using this information leads to the following code where 
the sklansky networks of size 32 or smaller use {\tt syncWarpHow}.

\begin{small}
\begin{verbatim} 
sklansky2 :: (Flatten a, Choice a) => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
sklansky2 0 op = pure id
sklansky2 n op = two (sklansky2 (n-1) op) ->- pure (fan op) 
                   ->-  if n <= 5 
                         then syncWarpHow (pairNth 256)
                         else syncHow (pairNth 256)
\end{verbatim}
\end{small}

%% XXX Why is there no explicit 32 anywhere in the vicinity of SyncWarpHow?

A last tweak to force the algorithm to compute {\em in-place}. This 
is accomplished using the primitives {\tt syncIP} and {\tt syncIPHow}, which
are variants of {\tt sync} that enforce {\em in-place} computation.

%-------------------------------------------------------------------------------
\subsubsection{Parallel Prefix Sums on large arrays}
\FloatBarrier

The Sklansky kernels given above can be used in an algorithm that computes 
the parallel prefix of a large array. This is done in the same way as in
~\cite{harris07parallel}.
The table below shows the results of using the different parallel prefix 
kernels from above in an algorithm for computing the prefix sums of 
$2^{20}$ elements.

\begin{small}
\begin{center}
  \begin{tabular}{  l | r | r |r | r  }
    %\hline
    Kernel          & In-place & Sync in Warp & Threads & ms   \\ \hline
    NVIDIA SDK      & Yes      & N/A          & 256     & 0.64 \\
    Hand Optimised  & Yes      & No           & 256     & 0.74 \\ 
    sklansky        & No       & yes          & 512     & 1.06 \\ 
    sklansky1       & No       & yes          & 256     & 0.89 \\ 
    sklansky2       & No       & No           & 256     & 0.86 \\ 
    sklansky3       & Yes      & No           & 256     & 0.79 \\
    %\hline
  \end{tabular}
\end{center}
\end{small}

The table above shows the running times of six different Sklansky kernels. 
All runtime measurements where performed on an NVIDIA 9800GX2 using one GPU. 
The code labeled {\em Hand Optimised} was written directly in CUDA. 
This kernel was the result of two afternoons of optimisation effort by 
two people. The three following kernels {\em sklansky} to {\em sklansky2} are 
generated from the given Obsidian programs. The version called {\em sklansky3} 
is identical to {\tt sklansky2} except is in-place. Even the very first version, 
{\em sklansky} performs reasonably well, but by using the experimental features 
the performance can be pushed quite close to the hand optimised version.  
The fastest kernel, called {\em NVIDIA SDK}, is the one supplied 
with the CUDA SDK. This kernel is based on the Brent-Kung network and its 
implementation is shown in~\cite{harris07parallel}. The {\tt NVIDIA SDK} 
kernel is highly optimized with regards to its memory access pattern and 
so the code is considerably more complicated than that of any of the Obsidian parallel-prefix kernels.

%-------------------------------------------------------------------------------
\FloatBarrier
\section{Future work}
\label{sec:FutureWork}

Obsidian is work in progress and as such it changes a lot. Future work will both
explore improvements to code generation for kernels and ways to coordinate kernels.

In order to get to the quite efficient version of the {\tt sklansky} kernel 
in section~\ref{sec:CaseStudies}, the experimental {\tt How} argument to {\tt sync} was needed. 
This needs to be 
explored further. {\tt How} as it is today, any function of type
{\tt IndexE -> [IndexE]}, offers too much freedom and with that the risk of
introducing errors. We are searching for an elegant model for expressing how and what 
to compute, and for ways to integrate this with
the language. This is our biggest research challenge.

Section~\ref{sec:CaseStudies}, on case studies, showed that it is possible to
generate quite efficient code from Obsidian programs. The current version 
generates efficient code for very specific uses of the {\tt two} combinator, 
for example the kind used in the {\tt sklansky} parallel prefix network. More 
work needs to be done in order to find a good way to produce efficient code 
for a larger set of combinators. 
This may very well involve designing new combinators that do not need
to be nested as the current ones do.

The CUDA code generated by Obsidian uses very few registers. In fact,
the generated code is under-utilizing the resources of the GPU. The generated 
code also contains many common subexpressions. This hints that applying a 
step of common subexpression elimintaion would be beneficial, increasing 
the register use and reducing the repeated computation of values. Performing
common subexpression elimination would also lower the instruction count,
which is quite high to begin with, because loops are unrolled. 
The register usage and instruction count values on which these conclusions 
are based where obtained using the CUDA profiler supplied as part of the 
CUDA toolkit~\cite{wwwcuda}.

Obsidian can currently only be used to generate kernels, the small 
building blocks used to form larger GPU algorithms. As future work, 
methods of describing kernel coordination in a high level fashion will 
be investigated.

At the University of New South Wales, Manuel Chakravarty 
et. al. are developing another language for GPGPU programming
called Accelerate. Accelerate is also embedded in Haskell, but intends
to be at a higher 
level of abstraction than Obsidian. Accelerate supplies the programmer 
with a collection of basic building blocks for data parallel programming. 
Amongst these building blocks are {\tt map}, {\tt zipWith} and {\tt scan}.
With the Accelerate team at UNSW, we are looking into 
ways of combining the two complementary approaches. In this setting Obsidian 
could be used to generate the underlying kernels (the building blocks) that the Accelerate 
programmer uses to build a GPGPU application.

A limitation of Obsidian is the inability to express algorithms where 
the length of the output array is dependent on the input data. 
An example of such a data dependent algorithm is {\tt filter}. The {\tt filter} 
function takes a sequence of elements and a predicate and produces a sequence 
of those elements for whom the predicate holds. Future work 
will also consist of searching for a suitable minimal language construct
to add that enables expressing such algorithms.

%-------------------------------------------------------------------------------
\section{Related work}

GPUs are becoming more and more interesting to use in non-graphical applications.
A modern GPU is a manycore machine with, today, hundreds of processing elements. 
The question of how to program these machines arises. NVIDIA's answer is CUDA
\cite{CUDAGuide2.0}. CUDA supplies a slightly extended version of C in which 
the programmer can specify GPU kernels and the controlling CPU program in the 
same language. In CUDA the programmer writes a single program parameterised 
over a thread identity. Compare this to Obsidian where the program describes 
a computation over an array.

There are a number of other C/C++ based languages that target 
GPGPU programmers, for example Brook\cite{Brook} and 
RapidMind\cite{RapidMind}. Brook, CUDA and RapidMind are major improvements from 
what was previously available for the programmer interested in general purpose 
computations on the GPU. 
%Before these, the GPGPU programmer only had the 
%graphics API to work with and needed to translate his or her programs into 
%graphics vocabulary in order to use the GPU.

Higher level approaches are also being investigated. PyGPU embedds a GPGPU 
programming language in Python\cite{PyGPU}. PyGPU makes use of Python's 
introspective abilities to generate efficient code.

Like Obsidian, Accelerate is embedded in Haskell \cite{GPUGEN}. Accelerate 
however, is higher level language than Obsidian. Where the purpose of 
Obsidian is to implement basic algorithmic building blocks such as 
reductions and prefix sums, Accelerate provide these building blocks as 
primitives.

Vertigo is another GPU programming language embedded in Haskell\cite{Vertigo}.
Vertigo can be used to describe procedureal surfaces and textures and generates
efficient DirectX9 shader code.

There are also many examples of domain specific languages (DSLs)
that target not only GPUs but also other platforms, such as multicore machines.
Of particular interest to us is Microsoft's Accelerator project~\cite{Accelerator}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
%However, Vertigo is targeting graphical applications.

%There are also many examples of languages that do not specifically target 
%GPUs but experiment with new methods and ways of parallel programming. 
%In this category we have Sequoia, where the memory hierarchy is in focus. 
%Sequoia programs can for example be compiled to the Cell BE architecture. 
%Data-Parallel Haskell extends Haskell with parallel arrays and operations 
%on parallel arrays\cite{DPH}. Data-Parallel Haskell implements the nested 
%data-parallel paradigm and is in that sense following in the path of 
%NESL\cite{NESL}.

%\begin{itemize}
%\item Chapel \cite{Chapel}
%\item DPH \cite{DPH}
%\item Fortress \cite{fortress}
%\item GPUGen \cite{GPUGEN}
%\item Lava \cite{lavaTutorial}
%\item RapidMind \cite{RapidMind}
%\item Sequoia \cite{SEQUOIA}
%\item X10 \cite{X10}
%\end{itemize}

\section{Discussion and conclusion}

Obsidian is work in progress and there are many loose ends to tie up and 
paths left to explore. In section~\ref{sec:CaseStudies}, the 
strengths of Obsidian show; it is possible to express quite complex algorithms
using short and elegant programs. The case studies also show that it is possible
to generate quite efficient code from these high level descriptions. However, 
this needs more work in order to more reliably produce efficient code and
to broaden the available set of combinators for combining GPU programs.

%Another benefit of a higher level language such as Obsidian compared to 
%CUDA is the ability to reuse code. There is an example of this too in the 
%case studies section where a single reduction program can be used to generate
%code to compute the maximum, minimum and sum very easily.

%Compositionally, or the ability to use already written programs as building
%blocks in other programs is another area where Obsidian is an improvement 
%over working in languages such as C for GPU programming. If in CUDA you have
%one Kernel that computes the

%In Obsidian it is easy to describe an initial prototype solution to a problem,
%such as for example the {\tt mySum} kernels in section~\ref{sec:SyncPar}. 
%It requires no deep knowledge of the GPU architecture but works out of the 
%box. The prototype implementation can be tweaked into a more efficient 
%implementation by performing small changes to the Obsidian code that often 
%lead to quite large differences in the generated C code. An example of this 
%is {\tt mySum1} against {\tt mySum2} where adding a single {\tt sync} leads 
%to a radically different C program.

Obsidian offers the GPGPU programmer a higher level language, while trying 
not to sacrifice too much performance. When Programming in CUDA C, the indexing
arithmetic often gets quite complex. This is a common trait of data-parallel 
programming in C-like languages. One goal of Obsidian is to be able to express 
these algorithms without the complex index manipulations; instead the data 
access pattern is captured in the use of functions such as {\tt pair} and 
{\tt two} or in the recursive structure of the Obsidian program.

%Obsidian is also an improvement over CUDA in the area of code reuse. The 
%{\tt reduce} function from section~\ref{sec:CaseStudies} is an example of this. 
%Obsidian is also compositional in another way than CUDA. Reusing kernels as
%building blocks in other kernels is in CUDA not realistic, while in Obsidian 
%it is the preferred way to write programs. Take as an example of this the 
%mergers and sorters in section~\ref{sec:CaseStudies}. In CUDA you would design 
%one kernel from scratch implementing the merger and sorter simultaneously.

This paper presented Obsidian an embedded language for GPGPU programming that
offers a higher level of abstraction compared to languages such as CUDA. 
Obsidian allows the programmer to think more of the algorithm and less of 
architectural details of the GPU. The contributions of Obsidian to the GPGPU 
field is a higher level programming environment that eases experimentation.

\section*{Acknowledgments}
\noindent
This research is funded by the Swedish Research Council.

%-------------------------------------------------------------------------------

%-------------------------------------------------------------------------------

%\bibliographystyle{elsarticle-num}
%\bibliography{obs_okt}

%\begin{figure}
%\centering
%\begin{minipage}{.5\linewidth}
%  \centering
%  \includegraphics[width=.9\linewidth]{./Pictures/sklansky-g1}
%  \caption{Stage 1}
%  \label{fig:fig_G1}
%\end{minipage}%
%\begin{minipage}{.5\linewidth}
%  \centering
%  \includegraphics[width=.9\linewidth]{./Pictures/sklansky-g2}
%  \caption{\small{Stage 2}}
%  \label{fig:fig_G2}
%\end{minipage}
%\end{figure}

%\end{document}

### 2013-09-30 src2/thesis.tex
apa

@@
 \usepackage{amssymb}
+\usepackage{mathtools}
 %\usepackage{lingmacros}
@@
 
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
 \cleardoublepage
 
-\section*{\centering Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+\chapter{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
+%\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
+%Parallel Programming of Graphics Processors}
+
+\begin{center} 
+Bo Joel Svensson, Mary Sheeran, Koen Claessen
+\end{center}
+
+\input{./ifl/paperThesis}
+
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
+\cleardoublepage
+
+\chapter{Obsidian: GPU Computing Using Haskell}
+%\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}
+
+\begin{center} 
+Bo Joel Svensson, Koen Claessen, Mary Sheeran
+\end{center}
+
+\input{./papp/paperThesis}
+
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
+\cleardoublepage
+
+\chapter{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+%\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
 
@@
 
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
 \cleardoublepage
 
-\section*{\centering Simple and Compositional Reification of Monadic Embedded Languages}
+\chapter{Counting and Occurrence Sort for GPUs using an Embedded Language}
+%\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}
+
+\begin{center} 
+Josef Svenningsson, Bo Joel Svensson, Mary Sheeran 
+\end{center}
+
+\input{./csort/paperThesis}
+
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
+\cleardoublepage
+
+\chapter{Simple and Compositional Reification of Monadic Embedded Languages}
+%\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
 
@@
 
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
 \cleardoublepage
 
-\section{Paper2}
+\chapter{Parallel Programming in Haskell Almost for Free}
+%\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
 
-%\input{}
+\begin{center} 
+Bo Joel Svensson, Mary Sheeran
+\end{center}
+
+\input{./embarbb/paperThesis}
 

### 2013-09-30 src2/expressive/abstract.tex
apa

@@
 
-\section*{Abstract}
+\subsection*{Abstract}
 Graphics Processing Units (GPUs) are powerful computing devices 

### 2013-09-30 src2/expressive/applications.tex
apa

@@
-\section{Application}
+\subsection{Application}
 \label{sec:MARY}
 
-\subsection{Sorting on a GPU}
+\subsubsection{Sorting on a GPU}
 In this section, we introduce combinators that express patterns
@@
 
-\subsection{Describing Batcher's bitonic merger}
+\subsubsection{Describing Batcher's bitonic merger}
 
@@
 
-\subsection{Modifying the bitonic merger}
+\subsubsection{Modifying the bitonic merger}
 
@@
 
-\subsection{New combinator implementations using Push arrays}
+\subsubsection{New combinator implementations using Push arrays}
 
@@
 
-\subsection{A sorter built from the balanced periodic merger}
+\subsubsection{A sorter built from the balanced periodic merger}
 First, we note that the balanced periodic merger contains multiple
@@
 
-\subsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
+\subsubsection{Measuring performance of the generated kernels}\label{sec:benchmarks}
 \begin{figure}

### 2013-09-30 src2/expressive/intro.tex
apa

@@
-\section{Introduction}
+\subsection{Introduction}
 
@@
 
-\subsection{Embedded DSLs for GPGPU programming} 
+\subsubsection{Embedded DSLs for GPGPU programming} 
 We are aiming for a GPU programming language that is more
@@
 %----------------------------------------------------------------------------
-\subsection{Arrays in Obsidian}
+\subsubsection{Arrays in Obsidian}
 
@@
 %----------------------------------------------------------------------------
-\subsection{Sync and parallelism}
+\subsubsection{Sync and parallelism}
 
@@
 
-\subsection{Drawbacks of Obsidian Arrays}
+\subsubsection{Drawbacks of Obsidian Arrays}
 \label{sec:Drawbacks}

### 2013-09-30 src2/ifl/related.tex
apa

@@
 
-\section{Discussion} \label{sec:disc}
-\subsection{Our influences}
+\subsection{Discussion} \label{sec:disc}
+\subsubsection{Our influences}
 As mentioned above, our earlier work on Lava has provided the inspiration for
@@
 
-\subsection{Related work on GPU and GPGPU programming languages}
+\subsubsection{Related work on GPU and GPGPU programming languages}
 We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}.
@@
 
-\subsection{Lessons learned so far in the project}
+\subsubsection{Lessons learned so far in the project}
 Our first lesson has been the gradual realisation that a key aspect of a usable

### 2013-09-30 src2/papp/paperThesis.tex
apa

@@
 
-\section*{Abstract}
+\subsection*{Abstract}
 Obsidian is a domain specific language for data-parallel programming 
@@
 %-------------------------------------------------------------------------------
-\section{Introduction}
+\subsection{Introduction}
 
@@
 left out because of space limitations, the use of this GPU program representation 
-can be seen in section~\ref{sec:GPUPrograms}. For a description of the implementation 
+can be seen in subsection~\ref{sec:GPUPrograms}. For a description of the implementation 
 see~\cite{JSTECH} and for more general information on embedded language 
@@
 %-------------------------------------------------------------------------------
-\subsection{NVIDIA GPUs and CUDA}
+\subsubsection{NVIDIA GPUs and CUDA}
 
@@
 % ------------------------------------------------------------------------------
-\subsection{Aims of Obsidian}
+\subsubsection{Aims of Obsidian}
 
@@
 %-------------------------------------------------------------------------------
-\section{Programming in Obsidian}
+\subsection{Programming in Obsidian}
 
@@
 %-------------------------------------------------------------------------------
-\subsection{Array Language}
+\subsubsection{Array Language}
 \label{sec:ArrayLanguage}
@@
 %to form larger GPU programs. How to create a GPU program from these is 
-%shown in the following section.
+%shown in the following subsection.
 
@@
 %\FloatBarrier
-\subsection{GPU Programs}
+\subsubsection{GPU Programs}
 \label{sec:GPUPrograms}
@@
 \noindent
-%The details of {\tt a :-> b} are shown in section~\ref{sec:Implementation}. 
+%The details of {\tt a :-> b} are shown in subsection~\ref{sec:Implementation}. 
 \noindent
@@
 boxes marked with {\em Pure} can be thought of as containing an array 
-program.% such as those in section~\ref{sec:ArrayLanguage}.
+program.% such as those in subsection~\ref{sec:ArrayLanguage}.
 The type constructor {\tt :->} is a variant of an arrow, which in turn is a generalisation of a monad~\cite{JSTECH}.
@@
 %The class {\tt Flatten} will be explained in detail in 
-%section~\ref{sec:Implementation}, but 
+%subsection~\ref{sec:Implementation}, but 
 Instances of {\tt Flatten} are all the types that can be stored in the 
@@
 
-\section{Outline of Code Generation}
+\subsection{Outline of Code Generation}
 \label{sec:CodeGen}
@@
 
-This section briefly reviews the code generation process using 
+This subsection briefly reviews the code generation process using 
 a hyphothetical program as example. The program under consideration 
@@
 
-\section{Case Studies}
+\subsection{Case Studies}
 \label{sec:CaseStudies}
@@
 \FloatBarrier
-This section describes a few slightly larger Obsidian programs. The main purpose of this section is to show 
+This subsection describes a few slightly larger Obsidian programs. The main purpose of this subsection is to show 
 how we want to write GPGPU programs using Obsidian. 
 %How the performance 
-%of the generated code can be improved is discussed in section~\ref{sec:FutureWork}  
+%of the generated code can be improved is discussed in subsection~\ref{sec:FutureWork}  
 %%% Move this later.  XXX
 
-\subsection{Sorting}
+\subsubsection{Sorting}
 
@@
 %-------------------------------------------------------------------------------
-\subsection{Parallel Prefix}
+\subsubsection{Parallel Prefix}
 \FloatBarrier
@@
 division of the work that has experimentally been shown to perform well, see 
-table in next section,  
+table in next subsection,  
 is to let each thread $tid$ perform the work of $tid$ and $tid + 256$. This 
@@
 \FloatBarrier
-\section{Future work}
+\subsection{Future work}
 \label{sec:FutureWork}
@@
 In order to get to the quite efficient version of the {\tt sklansky} kernel 
-in section~\ref{sec:CaseStudies}, the experimental {\tt How} argument to {\tt sync} was needed. 
+in subsection~\ref{sec:CaseStudies}, the experimental {\tt How} argument to {\tt sync} was needed. 
 This needs to be 
@@
 
-Section~\ref{sec:CaseStudies}, on case studies, showed that it is possible to
+Subsection~\ref{sec:CaseStudies}, on case studies, showed that it is possible to
 generate quite efficient code from Obsidian programs. The current version 
@@
 %-------------------------------------------------------------------------------
-\section{Related work}
+\subsection{Related work}
 
@@
 
-\section{Discussion and conclusion}
+\subsection{Discussion and conclusion}
 
 Obsidian is work in progress and there are many loose ends to tie up and 
-paths left to explore. In section~\ref{sec:CaseStudies}, the 
+paths left to explore. In subsection~\ref{sec:CaseStudies}, the 
 strengths of Obsidian show; it is possible to express quite complex algorithms
@@
 %CUDA is the ability to reuse code. There is an example of this too in the 
-%case studies section where a single reduction program can be used to generate
+%case studies subsection where a single reduction program can be used to generate
 %code to compute the maximum, minimum and sum very easily.
@@
 %In Obsidian it is easy to describe an initial prototype solution to a problem,
-%such as for example the {\tt mySum} kernels in section~\ref{sec:SyncPar}. 
+%such as for example the {\tt mySum} kernels in subsection~\ref{sec:SyncPar}. 
 %It requires no deep knowledge of the GPU architecture but works out of the 
@@
 %Obsidian is also an improvement over CUDA in the area of code reuse. The 
-%{\tt reduce} function from section~\ref{sec:CaseStudies} is an example of this. 
+%{\tt reduce} function from subsection~\ref{sec:CaseStudies} is an example of this. 
 %Obsidian is also compositional in another way than CUDA. Reusing kernels as
@@
 %it is the preferred way to write programs. Take as an example of this the 
-%mergers and sorters in section~\ref{sec:CaseStudies}. In CUDA you would design 
+%mergers and sorters in subsection~\ref{sec:CaseStudies}. In CUDA you would design 
 %one kernel from scratch implementing the merger and sorter simultaneously.
@@
 
-\section*{Acknowledgments}
+\subsection*{Acknowledgments}
 \noindent

### 2013-09-30 src2/thesis.tex
apa

@@
 
-\chapter{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
+\section{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
 %\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
@@
 
-\chapter{Obsidian: GPU Computing Using Haskell}
+\section{Obsidian: GPU Computing Using Haskell}
 %\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}
@@
 
-\chapter{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+\section{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
 %\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
@@
 
-\chapter{Counting and Occurrence Sort for GPUs using an Embedded Language}
+\section{Counting and Occurrence Sort for GPUs using an Embedded Language}
 %\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}
@@
 
-\chapter{Simple and Compositional Reification of Monadic Embedded Languages}
+\section{Simple and Compositional Reification of Monadic Embedded Languages}
 %\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
@@
 
-\chapter{Parallel Programming in Haskell Almost for Free}
+\section{Parallel Programming in Haskell Almost for Free}
 %\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}

### 2013-09-30 src2/embarbb/implementation.tex
apa

@@
-\section{Implementation} 
+\subsection{Implementation} 
 EmbArBB is a deeply embedded language. There is a constructor for 
@@
 
-\subsection{Vectors} 
+\subsubsection{Vectors} 
 EmbArBB has support for one, two and three dimensional dense vectors,
@@
 
-\subsection{The language} 
+\subsubsection{The language} 
 The EmbArBB language is implemented as a set of library functions, operating on an expression datatype:
@@
 
-\subsection{Interfacing with Haskell and Code generation}
+\subsubsection{Interfacing with Haskell and Code generation}
 

### 2013-09-30 src2/embarbb/programming.tex
apa

@@
-\section{Programming in ArBB}
+\subsection{Programming in ArBB}
 \label{sec:ProgrammingArBB}
@@
 %\FloatBarrier
-\section{Programming in EmbArBB} 
+\subsection{Programming in EmbArBB} 
 \label{sec:Programming}
@@
 % \FloatBarrier 
-\subsection{Saxpy} 
+\subsubsection{Saxpy} 
 {\em Saxpy} is a vector operation that gets its name from the 
@@
 
-\subsection{Matrix-vector multiplication} 
+\subsubsection{Matrix-vector multiplication} 
 In section~\ref{sec:ProgrammingArBB}, matrix-vector multiplication was shown using the C++ ArBB interface. 
@@
 
-\subsection{Matrix-matrix multiplication}
+\subsubsection{Matrix-matrix multiplication}
 
@@
 
-\subsection{Histogram}
+\subsubsection{Histogram}
 \FloatBarrier
@@
 \FloatBarrier
-\subsection{Sobel edge detection filter}
+\subsubsection{Sobel edge detection filter}
 \label{sec:Sobel}
@@
 
-\subsection{Image filtering} 
+\subsubsection{Image filtering} 
 \label{sec:Blur}
@@
 
-\subsection{Sparse matrix multiplication}\label{sec:smvm}
+\subsubsection{Sparse matrix multiplication}\label{sec:smvm}
 A sparse matrix can be represented by three vectors, one (cidx) containing

### 2013-09-30 src2/embarbb/related.tex
apa

@@
-\section{Related Work} 
-\subsection{Data.Array.Accelerate} 
+\subsection{Related Work} 
+\subsubsection{Data.Array.Accelerate} 
 Accelerate~\cite{accelerate} is an embedded language for general purpose, 
@@
 
-\subsection{Data Parallel Haskell} 
+\subsubsection{Data Parallel Haskell} 
 Data Parallel Haskell (DPH) is an extension to GHC for nested data parallelism. 
@@
 
-\subsection{Feldspar} 
+\subsubsection{Feldspar} 
 Feldspar is a language for Digital Signal Processing (DSP) programming
@@
 
-\subsection{Nikola} 
+\subsubsection{Nikola} 
 Nikola is an embedded language for GPU programming\cite{NIKOLA}, also
@@
 
-\subsection{Repa} 
+\subsubsection{Repa} 
 Repa is a library for regular, shape polymorphic parallel arrays~\cite{REPA}. Repa 
@@
 
-\section{Motivation}
+\subsection{Motivation}
 

### 2013-09-30 src2/embarbb/results.tex
apa

@@
-\section{Benchmarks}
+\subsection{Benchmarks}
 \label{sec:Benchmarks}
@@
 
-\subsection{matrix-matrix multiplication}
+\subsubsection{matrix-matrix multiplication}
 The matrix multiplication benchmarks consist of square matrices 
@@
 
-\subsection{Sobel edge detection}
+\subsubsection{Sobel edge detection}
 In the Repa Sobel code, only the applications of the stencils are timed. This 
@@
 
-\subsection{Blur}
+\subsubsection{Blur}
 
@@
 
-\subsection{About the numbers} 
+\subsubsection{About the numbers} 
 This section presents three benchmarks, two of which compare to Repa only and 

### 2013-09-30 src2/arbb/haskell_arbb.tex
apa

%% bare_conf.tex
%% V1.3
%% 2007/01/11
%% by Michael Shell
%% See:
%% http://www.michaelshell.org/
%% for current contact information.
%%
%% This is a skeleton file demonstrating the use of IEEEtran.cls
%% (requires IEEEtran.cls version 1.7 or later) with an IEEE conference paper.
%%
%% Support sites:
%% http://www.michaelshell.org/tex/ieeetran/
%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
%% and
%% http://www.ieee.org/

\documentclass[conference]{IEEEtran}

\usepackage{graphicx, color, code}
\usepackage[cmex10]{amsmath}
\usepackage{url}
% \usepackage{bookman}  %% bold teletype fonts, but don't know how to prevent it from changing the whole document

% Haven't figured out how to get this to work properly:
%\usepackage[\texttt{cmbtt}]{bold-extra}
% \usepackage{bold-extra}

\usepackage{courier}
\usepackage{fancyvrb}

% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}

%====================================================================================================
%% FIRST, MISC USEFUL DEFINITIONS
%====================================================================================================

\definecolor{darkgreen}{rgb}{0, 0.5, 0}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{mygrey}{rgb}{0.7,0.7,0.7}
\newcommand{\textblue}[1]{\textcolor{blue}{#1}}

\newcommand{\cde}[1]{{\footnotesize \tt #1}}

% \newcommand{\kw}[1]{{\bf{#1}}}
\newcommand{\kw}[1]{{\textbf{#1}}}

\newcommand{\comm}[1]{{\em \textcolor{darkgreen}{#1}}}  %% Code comments

\newcommand{\note}[1]{\begin{itemize}\item \textcolor{blue}{#1} \end{itemize}}

\newenvironment{mycode}
{% This is the begin code
  \footnotesize
  \noindent
  \begin{code}\noindent\vspace{-1mm}}
%  \begin{code}\noindent}
{% This is the end code
  \end{code} 
}

\newenvironment{inlinecode}
{
  \begin{center}
  \begin{minipage}{4in}
  \begin{mycode}
} 
{
  \end{mycode}
  \end{minipage}
  \end{center}
  \vspace{-1.5ex}
}

%% ============================================================
%% USE TO DISABLE COMMENTS:
 \def\noeditingmarks{}
%% ============================================================
\ifx\noeditingmarks\undefined
   \newcommand{\textred}[1]{\textcolor{red}{#1}}
   \newcommand{\pgwrapper}[2]{\textred{#1: #2}}
%   \newcommand{\pgwrapper}[2]{}
   \newcommand{\rnote}[1]{\begin{itemize}\item{\textcolor{blue}{#1}}\end{itemize}}
   \newcommand{\new}[1]{\textcolor{blue}{#1}}
   \newcommand{\const}[1]{\textred{#1}}

%% RRN: Dim the background for Ryan's eyes:
%%   What I need here is a way to do it based on user or host name...
% \pagecolor{mygrey}

\else
   \newcommand{\textred}[1]{#1}
   \newcommand{\pgwrapper}[2]{}
   \newcommand{\rnote}[1]{}
   \newcommand{\new}[1]{#1}
   \newcommand{\const}[1]{#1}
\fi

\newcommand{\js}[1]{\pgwrapper{BJS}{#1}}
\newcommand{\rn}[1]{\pgwrapper{RRN}{#1}}

%% Can we call our ``things'' 
%% ArBB/Haskell for the bindings part ? 
%% Accelerate/ArBB for the Accelerate backend ?
%% This would work with the convention we adopted for 
%%      ArBB/C++  
\newcommand{\systemname}[0]{{Harbb}}
%\newcommand{\systemname}[0]{\textred{HArBB}}

%% This could in theory be separate from the project name..
\newcommand{\accarbb}[0]{\systemname{}}

%% NOTE: THIS SHOULD BE THE SAME AS SYSTEMNAME:
\newcommand{\arbbvmH}[0]{ArBB/Haskell}

%====================================================================================================
% END DEFINITIONS
%====================================================================================================

\begin{document}
% \title{Painless Portable Vector Parallelism\\ with Haskell and Intel ArBB}
\title{Programming Future Parallel Architectures\\ with Haskell and Intel ArBB}
% \title{Embedded Domain-Specific Languages Pave the Way for New Parallel Architectures}

\author{\IEEEauthorblockN{Bo Joel Svensson}
\IEEEauthorblockA{Dept. of Computer Science and Engineering\\
Chalmers University of Technology\\
Gothenburg, Sweden\\
Email: joels@chalmers.se}
\and
\IEEEauthorblockN{Ryan Newton}
\IEEEauthorblockA{Intel Corporation\\
Hudson, MA\\
Email: rrnewton@gmail.com
}}
% Email: ryan.r.newton@intel.com

\maketitle

\begin{abstract}
%
New parallel architectures, such 
\new{as Cell, Intel MIC, GPUs, and tiled architectures, enable} high
performance but are often hard to program.
What is needed is a bridge between high-level programming models 
where programmers are most productive and modern parallel architectures. 
We propose that that bridge is Embedded Domain Specific Languages (EDSLs).

\new{One attractive target for EDSLs is}
Intel ArBB, a virtual machine for parallel, vectorized computations.
% that serves as a compilation target for EDSLs.
We propose to wed ArBB with the functional programming language
Haskell, {using an EDSL} that generates code for the ArBB VM.  This Haskell integration provides 
% a determinism guarantee 
added safety guarantees compared to other ArBB interfaces.
%
Further, our prototype, \systemname{}, is one of the first EDSL implementations with
optimized backends for multiple parallel architectures (CPU, \new{NVIDIA GPU}, and
others), allowing portability of source code over devices
and their accelerators.

\end{abstract}
\IEEEpeerreviewmaketitle

% ====================================================================================================
\section{Introduction}
% ====================================================================================================

Are radical new parallel architectures market-feasible if they require
significant changes for programmers?  The jury is out.  In recent
years we have seen difficult-to-program chips suffer 
% \cite{cell} 
(e.g. Cell)
and
GPU vendors strive to enable more traditional programming features
\cite{fermi} (e.g. C++).  There is an increasing
tension between ease of programming and efficiency.

The tension  shows  across diverse chip markets.  For example,
small embedded devices are most power efficient if their processors and operating systems
omit programming features such as virtual memory and threads \cite{tinyos}.  
At the other end of the power spectrum, GPU's
graphics performance may suffer due to inclusion of hardware to ease
GPGPU programming.  In short, there is an opportunity cost to
including extra hardware for programmability.

In this paper, we argue 
that a specific technique holds the greatest promise of solving the programmability dilemma.
%
Domain-specific languages, {\em embedded} within general purpose
languages (EDSLs) 
 enable familiar programming models 
{\em and} flexible
mapping onto new hardware.  The key to having this cake and eating it
too is {\em metaprogramming}.  Familiar programming features are
present, but are eliminated at an intermediate (metaprogram evaluation)
phase and therefore do not reach the parallel hardware itself.

% We present our ongoing work on a
In pursuit of this vision, we offer a
% 
 new EDSL implementation, called \accarbb{}, that combines
existing systems, Accelerate \cite{Accelerate} and ArBB \cite{ArBB}, to produce a unified
high-level programming environment equally suited to multicore,
vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \cite{larrabee}).
%
\if{0}
ArBB is based on the RapidMind \cite{RapidMind} system which was an embedded 
DSL targeting multicore CPUs as well as GPUs. ArBB does not \new{yet} include a GPU 
backend like RapidMind, targeting multicore CPUs and their vector 
units. 
\fi{}
\systemname{} is a single EDSL implementation 
with independently optimized backends \new{by different teams; 
namely, the ArBB backend for CPU/MIC, and a CUDA backend for NVIDIA GPU.}
This makes
\systemname{} an appealing platform for fair CPU/GPU comparisons, 
as well as a compelling programming model for single-source portable performance across a range of
parallel architectures, \new{present and future}.
%%To our knowledge, \systemname{} is the first EDSL implementation with
%%independently optimized backends for CPUs and GPUs.  This makes
%%\systemname{} an appealing platform for fair CPU/GPU comparisons, 
%%as well as a compelling programming
%%model for single-source portable performance across a range of
%%parallel architectures.

% ====================================================================================================
\section{Embedded Domain-Specific Languages}
% ====================================================================================================

Domain-specific languages---from Makefiles and \LaTeX{} to
Matlab---are almost too ubiquitous to notice.  Most relevant to our
purposes, domain-specific languages (DSLs) that target a narrow domain and expose communication
patterns to the compiler 
 have achieved performance-portability across a wide range of parallel
architectures.  
StreamIt\cite{streamit} is a good example.
% (for example, StreamIt\cite{streamit})
% The high-water mark in this area is perhaps  StreamIt\cite{streamit}.
%

% It has been noted \cite{not-sure} that 
% Yet the few DSLs that gain popularity lose simplicity\cite{not-sure}.  
% Yet popular DSLs rarely remain simple \cite{not-sure}.  
DSLs may start out simple and focused, but if they gain popularity
they quickly grow in complexity to rival full-blown languages.
Feature creep
can make DSL implementations complex and expensive to maintain.
Further, non-standard DSL syntax and features present a learning curve for users.  In the
last ten years an attractive solution to this dilemma has emerged: {\em
  embed} each DSL into a general-purpose host language that can provide
common functionality with familiar syntax and semantics.
% in place of the DSL.

When embedding, host language programs {generate} DSL programs;
% Embedding means that the host language is used to generate DSL programs; 
the {\em deeper} the embedding, the more integrated the
DSL into the syntax and type-system of the host language.
%
% Typically, %% pruning weasel words
A key host-language feature for embedding is that
language constructs can be overloaded to operate over {\em abstract
  syntax trees} (ASTs)\footnote{This ad-hoc polymorphism is accomplished,
for example, through operator-overloading in C++ or type classes in
Haskell.}.  For example, the following simple function operates on scalars:

\vspace{1mm}
\begin{code}
  float f(float x) \{ return (2*x - 1); \}
\end{code}
\noindent
But simply by changing the types, \cde{f} might be lifted to operate on {\em
  expressions} (which, when evaluated, will yield \cde{floats}):

\vspace{1mm}
\begin{code}
  \kw{exp<float>} f(\kw{exp<float>} x) \{
    return (2*x - 1);
  \}
\end{code}

A common arrangement is for the host language program to execute at
runtime but to generate ASTs that are executed by a just-in-time
(DSL) compiler.  This use of metaprogramming (program generation)
differs from the more common usage of preprocessors and 
% Lisp/Scheme macros \cite{scheme-macros}
macros, which typically add extra
phases of computation {\em before}  compile time---increasing
the number of compile-time rather than runtime phases.

% EDSLs typically add additional runtime phases.  That is, the host language program 
% executes at runtime but uses a JIT to process generated ASTs.
%% which are
%% usually used to perform computation {\em earlier} than the usual
%% compile time, embedded DSLs (EDSLs) typically defer computation

One reason that EDSLs are good for productivity is that the programmer
gains the software engineering benefits of the host language
(object-orientation, higher-order-functions, modules, etc), while not
paying the cost at runtime for additional layers of abstraction or
indirection.  Indeed, the embedded languages for 
performance-oriented EDSLs are often simple, first-order
languages without pointers \cite{wavescript, Accelerate}.

% , paul-lius-thingy

% regiment,
As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
decade \cite{vertigo, wavescript} but are gaining steam
recently 
%\cite{sejits, stanford-ppl} 
\cite{copperhead, stanford-ppl} 
and are beginning to appear in
commercial products \cite{ArBB}.
%
Further, EDSL techniques have spread beyond their origin in 
 the programming languages community.
% 
For example, both Stanford's Parallel Programming Laboratory (PPL) and
the Berkeley Parlab are creating EDSLs as their flagship parallel
programming solutions for domains such as machine learning and
rendering \cite{stanford-ppl}.
  Moreover, EDSLs need not be hosted by esoteric research languages---Intel's
 ArBB embeds an array language in C++ and Berkeley's
Copperhead \cite{copperhead} generates CUDA from simple Python code.

%\rnote{Is there a good motivating example program?  What are some of
%  the Metaocaml motivating examples?}

{For the remainder of this paper we will focus our discussion
  on the Intel ArBB VM, a virtual machine for just-in-time
  generation of vector codes, \new{which implements a restricted} domain-specific
  language of array computations.  In this paper we introduce
  {\em High-level ArBB}, 
  (\systemname{}), an EDSL that internally uses the ArBB
  VM.  \new{While Intel's ArBB package already includes an EDSL targeting the VM (for C++)},
\systemname{} \new{offers additional advantages, including}
  more succinct programs and additionally safety
  guarantees---namely, complete deterministic-by-construction parallel
  programs (across both host and VM languages).}

% ====================================================================================================
% \section{Accelerate and ArBB}
%\section{\accarbb{}:  ArBB and Accelerate}
\section{\accarbb{} $=$ ArBB $+$ Accelerate}
\label{sec:accelerate-arbb}
% ====================================================================================================

Our first \systemname{} prototype adapts an existing EDSL called {\em
  Accelerate} (Data.Array.Accelerate).  Accelerate targets high-level
data-parallel programming in Haskell.
Previous work on Accelerate has focused on developing a CUDA-backend
for GPU programming.  In this paper we describe our effort to retarget
Accelerate to ArBB.

With respect to determinism guarantees, the existing Intel ArBB product
represents an integration of the safe (ArBB VM) with the unsafe (C++).
In the Haskell context, because purely functional computations are
guaranteed deterministic (even when executed in parallel), and because
ArBB computations invoked by Haskell functions are themselves free of
side-effects, \systemname{} achieves a
guarantee of determinism for complete programs that combine both
Haskell computation and ArBB VM computation.
% which offers portability.
% (The 'H' in \systemname{} therefore stands for Haskell as well as ``High-level''.)

%% {\em Accelerate} (Data.Array.Accelerate) is an existing EDSL for 
%% high-level data-parallel programming in Haskell~\cite{Accelerate}. 
The Accelerate programming model consists of collective 
operations that can be performed over arrays, together with a simple 
language of scalar expressions\new{---in the current release, the Haskell type system enforces that parallelism not be nested.}
\if{0}
The Haskell type 
system is used to enforce that no nested data-parallelism can be expressed 
\new{(in the current release, this may be relaxed in the future).}
\fi{}
Accelerate's collective operations include {\em Map} (akin to parallel
for loops) {\em ZipWith} (a generalization of elementwise vector addition) and 
{\em Fold} (sum generalized)---familiar operations for programmers versed in the functional paradigm.
All of these collective operations are easily parallelizable. 
% In the case of {\em Map} and {\em ZipWith} the potential for parallelism is abundant.

%% dotProd :: Vector Float -> 
%%            Vector Float -> 
%%            Acc (Scalar Float) 
\begin{mycode}
dotProd (xs :: Vector Float) 
        (ys :: Vector Float) = 
  \kw{let} xs\_ = use xs 
      ys\_ = use ys 
  \kw{in}  fold (+) 0 (zipWith (*) xs\_ ys\_) 
\end{mycode}

The Accelerate code listing above specifies a function that takes two 1-dimensional 
arrays as inputs (of type \cde{Vector Float}, e.g. a vector of floats). The result of the function is a single
scalar. 
% (\cde{Scalar Float})
The \cde{use} function is applied to an array to convert it 
for use in the collective operations provided by Accelerate;
\cde{use} 
% is actually a function with type \cde{Vector a -> Acc (Vector a)} and 
may result in
copying the array to, for example, the GPU in the case of Accelerate's 
CUDA backend.

After applying \cde{use} to bring in input data, the programmer then
constructs a data-parallel program from collective operations.
Above, \cde{fold} is a function that takes three arguments, 
here those arguments are \cde{(+)},  $0$ and \cde{zipWith (*) xs\_ ys\_}. 
\cde{zipWith}, in turn, is a function taking three arguments, \cde{(*)}, 
\cde{xs\_} and \cde{ys\_}. The \cde{zipWith} operation here applies pairwise 
multiplication to the two arrays and the \cde{fold} sums up all the elements 
into a single scalar.

%\rn{Following para slightly redundant but ok...}
%Accelerate makes GPU programming accessible to the functional programmer 
%by offering a familiar set of operations. This set of operations also have 
%the beneficial property that they have inherent parallelism and are 
%easily implementable on the GPU.

Accelerate's CUDA backend implements collective operations using
a hand-tuned ``skeleton'' for each operation (and possibly for
different hardware versions). The kernel---for example, the function
to mapped over the dataset---is 
instantiated into the body of the skeleton code. The resulting 
program is compiled using the NVIDIA CUDA compiler and dynamically linked 
into the running Haskell program.

%\subsection{Intel Array Building Blocks (ArBB)}
\section{Intel Array Building Blocks (ArBB)}
\label{sec:arbb}

%ArBB is embedded in the C++ programming language but it also exposes a 
%very low level API that is purely C. 
%{\em ArBB} (Array Building Blocks) is also an embedded DSL. 
The operations exposed by ArBB are similar to those of Accelerate. 
In ArBB, parallel computations are expressed using a set of built-in 
primitives. All vectorization and threading is managed internally by 
ArBB. The programmer uses collective operations with a clear semantics 
such as \cde {add\_reduce} that computes the sum of the elements in a given 
array. ArBB also has language constructions for control flow, conditionals
and loops. These operations have their usual sequential semantics and are not 
parallelized by the system, rather, only specific collective operations are executed
in parallel.

Today's existing ArBB product is embedded in C++ and provides special
types for scalars and arrays (e.g. \cde{dense<f32>} rather than \cde{vector<float>}).
Using ArBB/C++ to express the dot product computation can be done 
as follows:

%\vspace{2mm}
\begin{code}
\kw{void} dot\_product(\kw{const} dense<f32>& a, 
                 \kw{const} dense<f32>& b,
                 f32& c) 
\{ 
  c = add\_reduce(a * b);
\}
\end{code}

% The above program takes two arrays of type \cde{dense<f32>} as input. 
% \cde{dense<f32>} specifies a 1-dimensional array of 32-bit floating
% point values. The result is a single 32-bit float.

Note that arithmetic operators such as (*) are overloaded 
for to operate on arrays as well as scalars (e.g. \cde{a * b} above).
Thus, the above C++ function multiplies two arrays before summing them with \cde {add\_reduce}:

The code listing below is indicative to the amount of glue code needed to invoke an 
ArBB computation\footnote{In OpenCL and CUDA the glue code situation is even worse}. It shows how the dot product code is launched using {\tt call} and 
how data is bound, {\tt bind}, for use in ArBB.

\vspace{2mm}
\begin{code}
\kw{int} main() 
\{ 
  double a[SIZE];
  double b[SIZE];

  for ( \kw{int} i = 0; i < SIZE; ++i) \{ 
    a[i] = ...; b[i] = ...;
  \}
  dense<f32> va, vb;
  f32 vc;
  bind(va, a, SIZE); 
  bind(vb, b, SIZE); 
  call(dot\_product)(va, vb, &vc); 
  ...
\}
\end{code}

%\rnote{SPECIFICALLY ADDRESS THE FUNCTIONALITY GAP?  E.G general
%  reduce.  Any other examples?}

The model provided by Accelerate is slightly richer than that of ArBB.
Even the two very simple \cde {dot\_product} examples above manage to illustrate 
this. In Accelerate there is a more general reduction primitive called \cde {fold} 
where in ArBB there are specific reductions, \cde {add\_reduce}, \cde {mul\_reduce} 
and so on. Not visible in these small examples is another difference, Accelerate 
operations are generalized to arbitrary dimensions while ArBB operations are 
limited to 1, 2, and 3 dimensions (and 0, i.e. scalar). These differences aside,
the ArBB and Accelerate programming models are very similar.

% ====================================================================================================
\section{Implementation of \systemname{}} 
% ====================================================================================================

%\rnote{Sadly the Haskell details don't buy us much here... what are
%  the {\em conceptual} problems solved in mapping the Accelerate
%  abstraction to the ArBB one?  The data models are similar, but
%  there are always mismatches -- dimensionality support, missing
%  reduction.}

%\rnote{What is the technique that you are using to address these sorts
%  of problems?  Is there anything to say about it?}

\begin{figure}
  \begin{center}
   \includegraphics[width=2.0in]{figure_architecture}    
  \end{center}
\vspace{-5mm}
  \caption{Architecture of the \systemname{} system.  \rn{(ROUGH.)}}
  \label{f:architecture}
\end{figure}

Using Accelerate and ArBB together, we propose a layered architecture
for \systemname{},
% Our goal is to make the benefits of ArBB available to the Haskell programmer. 
% \js{(Or is our goal to give an even better ``front-end'' to ArBB compared to C++)}
% We propose that this is best accomplished using a layered approach, as
pictured in Figure~\ref{f:architecture}. 
The host language execution (via Haskell in this case) executes the
programmer's source code, generating a vector program in the
restricted language of the Accelerate front-end.
Then either Accelerate backend---ArBB or CUDA---may be used.
The 
bottom layer of the ArBB backend consists of a direct mapping of the ArBB virtual machine
API (VMAPI) into Haskell including one-to-one bindings for each C 
function. 
To [partially] automate the creation of these bindings we used the C2HS system \cite{C2HS}. The 
\arbbvmH{} bindings are very low-level.
%and make little use of the high-level programming methods that
% Haskell programmers want. 
The idea 
is that the \arbbvmH{} bindings should be used to implement
backends for higher-level data-parallel EDSLs.

%% To illustrate our position we 
%% have started working on a backend to the successful DSL Accelerate 
%% (Data.Array.Accelerate) using our \arbbvmH{} bindings.

To address the functionality mismatch between Accelerate and ArBB,
when possible we reencode the users Accelerate program using existing
ArBB mechanisms.  Our prototype does not support 100\% of Accelerate's
compute model (for example, only supporting  up to three-dimensional
arrays), but the remaining functionality can be mapped onto ArBB in
time using known methods---for example, our compiler could map higher
dimensional arrays onto a specific lower-dimensional data-layout.

%% The programming models supported by ArBB and Accelerate are very similar. 
%% Both obtain their parallelism by exposing collective operations 
%% such as {\em Map}, {\em Reduce}, {\em ZipWith} and {\em Scan} (all prefixes operations). However, 
%% the functionality match is not 100\%. For example, Accelerate supports {\em Map} operations 
%% over N-dimensional Arrays while ArBB supports them only up to 3-dimensions. 
%% Similar inconsistencies apply to the other operations as well. \textred{The current
%% ArBB backend simply disallows Accelerate programs using too high dimensionality
%% of the arrays.} \rn{Sure, we want to say that we're not done closing
%%   this gap (for
%%   accuracy) but more importantly we want to identify whether
%%   techniques exist to close the gap.  Presumably, yes?  We can
%%   ourselves collapse the dimensionality of arrays before giving them
%%   to ArBB?  Are there any operations that would defeat that technique?}
%

One example of a functionality discrepancy bridged by our
implementation is reductions.  As mentioned in \ref{sec:arbb},
Accelerate allows the programmer to reduce using an arbitrary
associative function, but ArBB has only built-in reductions with fixed
operations (add, multiply, xor, etc). We plan to provide general reductions in
the ArBB backend by a two-fold strategy:

\begin{enumerate}
\item Attempt to map an Accelerate reduction directly onto an ArBB
  primitive such as \cde{add\_reduce}.
\item Apply a general reduction technique based on $log(N)$ map
  operations over successively halving array sizes.  Essentially, cut
  the array in half, combine the halves, repeat. In ~\cite{reduction}
  this approach is explained in the context of CUDA. 
\end{enumerate}

%% \rn{Is the technique worthy of
%%   description?  Can we just cite the CUDA documentation or something
%%   since it's common?}

In our current experiments, approach (2) is significantly slower.
Therefore, the ideal would be that ArBB exposed general reduction
directly in its programming model.  We expect this functionality to be
added in future releases.  In the meantime we plan to explore a
technique that would allow us to maximize the number of situations in
which (1) above applies.  Namely, a reduce operation can often be {\em
  factored} into a map followed by a reduce.  For example, a reduction
that multiplies each input number by a coefficient and sums the
results can be split into a map phase for the multiplication followed
by the built-in \cde {add\_reduce} operator.

%% Currently, our reduction implementations in ArBB 
%% are many times 
%% slower (\textasciitilde 20x) than their built-in counterparts. 
%% % To ameliorate this present performance problem we attempt to map
%% % programmers 
%% Our present workaround for this performance problem is to let 
%% \accarbb{} inspect the reduction function and switch to efficient 
%% built in  ArBB reduction where possible.  
%% \rn{Future work may be to decompose reductions into map+reduce so as
%%   to allow the reduce to become a builtin.}

Another choice faced by our implementation is the granularity at which the
ArBB JIT is invoked.  Specifically, should each collective operation result in its
own call to the ArBB JIT (in ArBB terminology, {\em
  immediate-mode}, akin to the pre-OpenGL 3.0 immediate mode), or
should multiple collective operations be placed together
inside an ArBB function and passed to the JIT?
We will call the latter approach {\em retained-mode}.

%% In immediate-mode the 
%% operations called from the ArBB API are executed
%% synchronously---before each API call returns. In retained-mode, ArBB
%% collects calls and, only when results are requested, the 
%% entire computation is optimized 
%% and executed.

Retained-mode generally offers performance benefits; a bigger 
chunk is given to the JIT compiler, enabling cross-optimization
between collective operations. 
Our prototype Accelerate backend uses 
ArBB in  
a combination of immediate- and retained-mode. The main collective 
operations are compiled using the retained-mode. For example, in the case 
of a \cde{map f}  operation first the function to be mapped is created and 
compiled using retained-mode then a small {\em mapper} function is 
also created and compiled using retained-mode. 
%% The mapper function is 
%% of course very small and therefore does not give much to the JIT compiler 
%% to work on optimization-wise. 
Between the collective operations the backend
needs to perform data management and copying, which 
are performed in immediate-mode. It is our belief that 
\systemname{} would benefit from using retained-mode exclusively 
but we leave that as future work.

% ====================================================================================================
\section{Preliminary Results}
% ====================================================================================================

%\rnote{One tiny benchmark if we can manage it.}

%\rnote{Ideally: Something like blackscholes running Arbb/C++,
%  \systemname{}, CPU+GPU, possibly compared to other blackscholes
%  implementations -- say Cilk on CPU and hand-coded CUDA (I can
%  provide the former I think).  Perhaps serial, non-vectorized CPU for
%  comparison.}

%\rnote{Point 1 -- High level thing is within X\% of low level thing.
%  The usual.  But we also want to say that \systemname{} is no worse
%  than ArBB/C++ -- the metalanguage doesn't matter, be it Haskell,
%  Python, or C++.}

%\rnote{Point 2 -- It's important to do a good job on the CPU side.
%  Even if you don't beat GPU there's a big difference between
%  vectorizing/not-vectorizing (and using multicore of coarse).  Can't
%  be ignored.}

%\rnote{Bonus points -- if we could show how bad ``naive'' ports are that
%  would be nice.  How well does your C CPU code run on GPU unmodified
%  (or minimally ported)?}

%\rnote{CPU/GPU systems that do a good job of both.  Perhaps quickly
%  dismiss those that make no effort on the CPU side.}

Black-Scholes option pricing is a finance-related benchmark that has been used in similar
DSLs targeting GPUs \cite{Accelerate, nikola}. Since we are re-using the 
Accelerate front-end, we can directly use the Black-Scholes benchmark that 
is shipped with that system.  Figure \ref{fig:blackscholes} shows the
complete code listing for an Accelerate Black-Scholes function which
can be executed on GPUs or any processor targeted by ArBB.

% blackscholes :: Vector (Float, Float, Float) 
%
%                -> Acc (Vector (Float, Float))
%\fbox{

%       r     = \textit{constant} riskfree
%       v     = \textit{constant} volatility

\begin{figure}
\begin{footnotesize}
\begin{Verbatim}[frame=single, commandchars=\\\{\}]
blackscholes (xs :: Vector (Float,Float,Float)) = 
  map kernel (\textit{use} xs)

kernel x =
  \kw{let} (price, strike, years) = \textit{unlift} x
      r     = 0.02 \textsf{\textit{-- riskfree constant}}
      v     = 0.30 \textsf{\textit{-- volatility constant}}
      sqrtT = sqrt years
      d1    = (log (price / strike) + 
                   (r + 0.5 * v * v) * years) / 
              (v * sqrtT)
      d2    = d1 - v * sqrtT
      cnd d = d >* 0 ? (1.0 - cndfn d, cndfn d)
      cndD1 = cnd d1
      cndD2 = cnd d2
      expRT = exp (-r * years)
  \kw{in} \textit{lift} ( price * cndD1 - 
            strike * expRT * cndD2
          , strike * expRT * (1.0 - cndD2) - 
            price * (1.0 - cndD1))

cndfn d =
  \kw{let} poly  = horner coeff
      coeff = [0, 0.31, -0.35, 1.78, -1.82, 1.33]
      rsqrt = 0.39894228040143267793994
      k     = 1.0 / (1.0 + 0.2316419 * abs d)
  \kw{in} rsqrt * exp (-0.5*d*d) * poly k

horner coeff x = 
  \kw{let} madd a b = b*x + a
  \kw{in}  foldr1 madd coeff
\end{Verbatim}
\end{footnotesize}
\caption{Complete code listing for a Black-Scholes function expressed
  in Haskell syntax using the Accelerate and \systemname{} libraries.
  Invocations of the functions \cde{use}, \cde{lift} and \cde{unlift}
%  and \cde{constant} 
represent additional boilerplate added for
  conversion in and out of Accelerate types.  Specifically, \cde{lift} and
  \cde{unlift} convert tuples and handle the fact that Accelerate
  arrays of tuples are really implemented as tuples of arrays.
  Otherwise, the program is identical to a plain Haskell
  implementation.}
\label{fig:blackscholes}
\end{figure}

%% \end{code}
%%  %605993438
%% % horner :: Num a => [a] -> a -> a
%% \begin{code}
%% \footnotesize

The kernel of this algorithm performs arithmetic on triples of
floating point numbers, creating pairs of floats as results. The problem 
is embarrassingly parallel, consisting of independent computations
for every element of an array (a map).

\begin{figure}
  \begin{center}
   \includegraphics[width=3.5in]{jit}    
  \end{center}
\vspace{-5mm}
  \caption{Running time experiments including JIT-time.} 
  \label{f:jit}
\end{figure}

\begin{figure}
  \begin{center}
   \includegraphics[width=3.5in]{nojit}    
  \end{center}
\vspace{-5mm}
  \caption{Running time experiments JIT-time excluded.}
  \label{f:nojit}
\end{figure}

Figures~\ref{f:jit} and \ref{f:nojit} show preliminary results obtained on 
the Black-Scholes benchmark. The figures where obtained on a system with a 4-core 
Intel Core I7 975 machine with HyperThreading. The GPU used was a NVIDIA GTX480.

Figure~\ref{f:jit} shows running times obtained when JIT-time is included. 
% What this figure mainly tells us is that the JITing overhead 
JIT compilation time is much larger 
in the CUDA backend than the ArBB one.  Part of this difference 
can be attributed to the fact that 
the CUDA backend calls an external compiler (\cde{nvcc}), which takes its input in a file and runs in
a separate process.  ArBB, on the other hand, has a library interface to its  JIT-compiler. 
Figure~\ref{f:nojit} shows results obtained when pre-compiling the
CUDA functions eliminating JIT overhead.
In Accelerate, this happens automatically when the same kernel is
invoked repeatedly, because the Accelerate CUDA backed 
uses a caching scheme to avoid unnecessary JIT invocations.  (The
caching functionality has not yet been duplicated in the 
ArBB backend.)

These early results demonstrates the principle that even when a kernel
executes with higher throughput on a GPU, in a particular program it
is difficult to decide whether a computation is worth moving to a GPU,
incurring extra data-movement and possibly extra JIT compilation
\cite{wheres-the-data-paper}.  Specifically, we see that the
Accelerate Black-Scholes program from Figure~\ref{fig:blackscholes}
performs better on the CPU if it executes once (even on a large window
of data) whereas the GPU would yield better performance in a sustained
series of executions.
%
Because both CPU and GPU execution may be desirable---and the
selection may be dynamic---it is beneficial to have a single source code
that is portable across both.  
% Accelerate combined with \systemname{} makes this possible.

% ====================================================================================================
\section{Related Work}
% ====================================================================================================

OpenCL~\cite{opencl08} is a programming model very similar to CUDA but with the aspiration 
to offer both acceleration of computations on GPUs or to multicore CPUs. 
OpenCL JIT compiles the kernels for the particular hardware available and is
in that sense similar to ArBB.  
OpenCL programs are relatively low-level and require a large amount of
boilerplate to create and invoke.  In this sense they occupy a very
different niche than Accelerate.

Microsoft Accelerator~\cite{accelerator} is an embedded language with similar 
aspirations as ArBB, that is, to target a diverse range of architectures using the 
same source code. Accelerate can be used from the C\# language or the 
functional F\# language and targets GPUs or of CPUs and their
vector units.
% However, being part of the .Net framework implies limited accessibility 
%for platforms other than Microsoft Windows.

\rn{What's the comparison with our stuff?}

%Most work on partitioning workloads between 
Many CPU and GPU comparisons, and some CPU/GPU workload partitioners \cite{merge}, rely on
redundant hand-written versions of all kernels
% \cite{cpu-gpu-partitioning} 
(though some systems like Qilin \cite{qilin} allow a single source code).
It is difficult in this kind of scenario
to make fair comparisons, controlling for the
amount of effort put into the respective implementations.
For example, comparing unoptimized serial CPU implementations vs. GPU
ones is not informative \cite{debunking-dubey}.
In \systemname{} controlling for effort
need happen only once---both CUDA and ArBB are
independently optimized by their respective teams of engineers---not for each benchmark.

% ====================================================================================================
% \section{Future work and Conclusions}
%\section{Future Work}
% ====================================================================================================

%\rn{need to be careful about being too repetitive...}
%The \accarbb{} backend is currently work-in-progress. The Accelerate
%functionality is not yet completely covered by the ArBB backend. Currently 
%the ArBB back-ed is operational for Accelerate programs using {\em Map}, 
%{\em ZipWith} and some simple instances of the Accelerate {\em Fold} operation.
%All of these also have the limitation that they only work for up to and 
%including 3-dimensional arrays using the ArBB backend while the Accelerate
%model allows the N-dimensional case. Allowing the N-dimensional case 
%in the ArBB backend for the supported operations range from a simple exercise 
%to more intricate problems needing some thought. For example, in the case 
%{\em Map} operations the needed transformation is simple (Accelerate Maps 
%over N-dimensional arrays is a per-element operation). One could flatten 
%the N-dimensional array to a 1-dimensional array (remembering its old shape) 
%perform the map and then transform the array back into N-dimensions. In 
%the case of {\em Fold} (reductions) the situation becomes more tricky. 
%transforming the arrays to arrays of lesser dimensionality then requires 
%a so-called segmented {\em Fold} operation.

%The current \accarbb{} backend uses a hybrid of immediate-mode 
%and retained-mode programming styles. Moving more towards 
%retained-mode only may have performance benefits since it would give the 
%ArBB compiler larger programs to optimize. Exploring this aspect is left 
%as future work.

% ====================================================================================================
\section{Discussion and Conclusions}
% ====================================================================================================

We have demonstrated that an EDSL such as Accelerate is sufficiently
platform-independent to break free of its original hardware target
(CUDA/GPU) and create efficient programs on other architectures.  This
gives us hope that \systemname{}/Accelerate programs will be
forward-portable to future parallel architectures and instruction
sets.

The EDSL approach changes the playing field for the designer
of compiler backends.  Rather than contending with full blown
languages and their complexities (e.g. pointers, aliasing,
inheritance, virtual functions, etc), compiler backends can focus on
simple value-oriented compute languages.

%% Achieving the ultimate goal of performance portability requires
%% progress on two research problems.  The first is EDSLs themselves and
%% their integration into host languages.  The second problem

But the EDSL method solves only part of the performance-portability problem.
Simple as EDSL target languages may be, there remains a
substantial challenge in mapping them efficiently to the diversity
of parallel architectures available now and in the near future.
%especially without requiring a large amount of additional per-platform
%user annotations.
\textred{For example, the idiosyncrasies of memory bank access on
  NVIDIA GPUs must be taken into account to generate efficient
  implementations of the high-level collective operations that we have
  discussed.}

This is a compiler backend research challenge.
The {\em skeletons} method mentioned in
Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
are the optimizations studied in the Copperhead \cite{copperhead} and
Obsidian \cite{obsidian} projects.
%
On the other hand, systems that rely on advanced optimizations
typically suffer to some extent from performance-predictability
problems.  Thus achieving portable, predictable performance on a wide range
of architectures---even for the simplest target languages---will
be the subject of much future work.

{
\small
% \bibliographystyle{plainnat}
\bibliographystyle{IEEEtran}
\bibliography{IEEEabrv,haskell_arbb}
}

\end{document}

### 2013-09-30 src2/arbb/paperThesis.tex
apa

--- 2013-09-30 src2/arbb/haskell_arbb.tex
@@
 
-\documentclass[conference]{IEEEtran}
+%\documentclass[conference]{IEEEtran}
 
-\usepackage{graphicx, color, code}
-\usepackage[cmex10]{amsmath}
-\usepackage{url}
+%\usepackage{graphicx, color, code}
+%\usepackage[cmex10]{amsmath}
+%\usepackage{url}
 % \usepackage{bookman}  %% bold teletype fonts, but don't know how to prevent it from changing the whole document
@@
 
-\usepackage{courier}
-\usepackage{fancyvrb}
+%\usepackage{courier}
+%\usepackage{fancyvrb}
 
 % correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
+%\hyphenation{op-tical net-works semi-conduc-tor}
 
@@
 {% This is the begin code
-  \footnotesize
-  \noindent
-  \begin{code}\noindent\vspace{-1mm}}
+  \begin{verbatim}}%
+  %\footnotesize
+  %\noindent
+  %\begin{code}\noindent\vspace{-1mm}}
 %  \begin{code}\noindent}
 {% This is the end code
-  \end{code} 
+  \end{verbatim}
+  %\end{code} 
 }
@@
 
-\begin{document}
+%\begin{document}
 % \title{Painless Portable Vector Parallelism\\ with Haskell and Intel ArBB}
-\title{Programming Future Parallel Architectures\\ with Haskell and Intel ArBB}
+%\title{Programming Future Parallel Architectures\\ with Haskell and Intel ArBB}
 % \title{Embedded Domain-Specific Languages Pave the Way for New Parallel Architectures}
 
-\author{\IEEEauthorblockN{Bo Joel Svensson}
-\IEEEauthorblockA{Dept. of Computer Science and Engineering\\
-Chalmers University of Technology\\
-Gothenburg, Sweden\\
-Email: joels@chalmers.se}
-\and
-\IEEEauthorblockN{Ryan Newton}
-\IEEEauthorblockA{Intel Corporation\\
-Hudson, MA\\
-Email: rrnewton@gmail.com
-}}
+%\author{\IEEEauthorblockN{Bo Joel Svensson}
+%\IEEEauthorblockA{Dept. of Computer Science and Engineering\\
+%Chalmers University of Technology\\
+%Gothenburg, Sweden\\
+%Email: joels@chalmers.se}
+%\and
+%\IEEEauthorblockN{Ryan Newton}
+%\IEEEauthorblockA{Intel Corporation\\
+%Hudson, MA\\
+%Email: rrnewton@gmail.com
+%}}
 % Email: ryan.r.newton@intel.com
 
-\maketitle
+%\maketitle
 
-\begin{abstract}
+%\begin{abstract}
 %
+\subsection*{Abstract}
 New parallel architectures, such 
@@
 
-\end{abstract}
-\IEEEpeerreviewmaketitle
+%\end{abstract}
+%\IEEEpeerreviewmaketitle
 
 % ====================================================================================================
-\section{Introduction}
+\subsection{Introduction}
 % ====================================================================================================
@@
 % ====================================================================================================
-\section{Embedded Domain-Specific Languages}
+\subsection{Embedded Domain-Specific Languages}
 % ====================================================================================================
@@
 %\section{\accarbb{}:  ArBB and Accelerate}
-\section{\accarbb{} $=$ ArBB $+$ Accelerate}
+\subsection{\accarbb{} $=$ ArBB $+$ Accelerate}
 \label{sec:accelerate-arbb}
@@
 %%            Acc (Scalar Float) 
-\begin{mycode}
+\begin{verbatim}
 dotProd (xs :: Vector Float) 
@@
   \kw{in}  fold (+) 0 (zipWith (*) xs\_ ys\_) 
-\end{mycode}
+\end{verbatim}
 
@@
 %\subsection{Intel Array Building Blocks (ArBB)}
-\section{Intel Array Building Blocks (ArBB)}
+\subsection{Intel Array Building Blocks (ArBB)}
 \label{sec:arbb}
@@
 % ====================================================================================================
-\section{Implementation of \systemname{}} 
+\subsection{Implementation of \systemname{}} 
 % ====================================================================================================
@@
   \begin{center}
-   \includegraphics[width=2.0in]{figure_architecture}    
+   \includegraphics[width=2.0in]{./arbb/figure_architecture}    
   \end{center}
@@
 % ====================================================================================================
-\section{Preliminary Results}
+\subsection{Preliminary Results}
 % ====================================================================================================
@@
   \begin{center}
-   \includegraphics[width=3.5in]{jit}    
+   \includegraphics[width=3.5in]{./arbb/jit}    
   \end{center}
@@
   \begin{center}
-   \includegraphics[width=3.5in]{nojit}    
+   \includegraphics[width=3.5in]{./arbb/nojit}    
   \end{center}
@@
 % ====================================================================================================
-\section{Related Work}
+\subsection{Related Work}
 % ====================================================================================================
@@
 % ====================================================================================================
-\section{Discussion and Conclusions}
+\subsection{Discussion and Conclusions}
 % ====================================================================================================
@@
 
-{
-\small
+%{
+%\small
 % \bibliographystyle{plainnat}
-\bibliographystyle{IEEEtran}
-\bibliography{IEEEabrv,haskell_arbb}
-}
+%\bibliographystyle{IEEEtran}
+%\bibliography{IEEEabrv,haskell_arbb}
+%}
 
-\end{document}
+%\end{document}

### 2013-10-01 src2/arbb/haskell_arbb.tex
apa

@@
 As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
-decade \cite{vertigo, wavescript} but are gaining steam
+decade \cite{VERTIGO, wavescript} but are gaining steam
 recently 

### 2013-10-01 src2/arbb/paperThesis.tex
apa

@@
 
-\newenvironment{mycode}
-{% This is the begin code
-  \begin{verbatim}}%
-  %\footnotesize
-  %\noindent
-  %\begin{code}\noindent\vspace{-1mm}}
-%  \begin{code}\noindent}
-{% This is the end code
-  \end{verbatim}
-  %\end{code} 
-}
+%% \newenvironment{mycode}
+%% {% This is the begin code
+%%   \begin{Verbatim}[commandchars=\\\{\}]}%
+%%   %\footnotesize
+%%   %\noindent
+%%   %\begin{code}\noindent\vspace{-1mm}}
+%% %  \begin{code}\noindent}
+%% {% This is the end code
+%%   \end{Verbatim}
+%%   %\end{code} 
+%% }
 
-\newenvironment{inlinecode}
-{
-  \begin{center}
-  \begin{minipage}{4in}
-  \begin{mycode}
-} 
-{
-  \end{mycode}
-  \end{minipage}
-  \end{center}
-  \vspace{-1.5ex}
-}
+%% \newenvironment{inlinecode}
+%% {
+%%   \begin{center}
+%%   \begin{minipage}{4in}
+%%   \begin{mycode}
+%% } 
+%% {
+%%   \end{mycode}
+%%   \end{minipage}
+%%   \end{center}
+%%   \vspace{-1.5ex}
+%% }
 
@@
  new EDSL implementation, called \accarbb{}, that combines
-existing systems, Accelerate \cite{Accelerate} and ArBB \cite{ArBB}, to produce a unified
+existing systems, Accelerate \cite{ACCELERATEDAMP11} and ArBB \cite{ArBB}, to produce a unified
 high-level programming environment equally suited to multicore,
@@
 \vspace{1mm}
-\begin{code}
+\begin{Verbatim}[commandchars=\\\{\}]
   float f(float x) \{ return (2*x - 1); \}
-\end{code}
+\end{Verbatim}
 \noindent
@@
 \vspace{1mm}
-\begin{code}
+\begin{Verbatim}[commandchars=\\\{\}]
   \kw{exp<float>} f(\kw{exp<float>} x) \{
@@
   \}
-\end{code}
+\end{Verbatim}
 
@@
 performance-oriented EDSLs are often simple, first-order
-languages without pointers \cite{wavescript, Accelerate}.
+languages without pointers \cite{wavescript, ACCELERATEDAMP11}.
 
@@
 %%            Acc (Scalar Float) 
-\begin{verbatim}
+\begin{Verbatim}[commandchars=\\\{\}]
 dotProd (xs :: Vector Float) 
@@
   \kw{in}  fold (+) 0 (zipWith (*) xs\_ ys\_) 
-\end{verbatim}
+\end{Verbatim}
 
@@
 %\vspace{2mm}
-\begin{code}
+\begin{Verbatim}[commandchars=\\\{\}]
 \kw{void} dot\_product(\kw{const} dense<f32>& a, 
@@
                  f32& c) 
-\{ 
+{ 
   c = add\_reduce(a * b);
-\}
-\end{code}
+}
+\end{Verbatim}
 
@@
 \vspace{2mm}
-\begin{code}
+\begin{Verbatim}[commandchars=\\\{\}]
 \kw{int} main() 
-\{ 
+{ 
   double a[SIZE];
@@
 
-  for ( \kw{int} i = 0; i < SIZE; ++i) \{ 
+  for ( \kw{int} i = 0; i < SIZE; ++i) { 
     a[i] = ...; b[i] = ...;
-  \}
+  }
   dense<f32> va, vb;
@@
   ...
-\}
-\end{code}
+}
+\end{Verbatim}
 
@@
 Black-Scholes option pricing is a finance-related benchmark that has been used in similar
-DSLs targeting GPUs \cite{Accelerate, nikola}. Since we are re-using the 
+DSLs targeting GPUs \cite{ACCELERATEDAMP11, NIKOLA}. Since we are re-using the 
 Accelerate front-end, we can directly use the Black-Scholes benchmark that 
@@
 
-Microsoft Accelerator~\cite{accelerator} is an embedded language with similar 
+Microsoft Accelerator~\cite{ACCELERATOR} is an embedded language with similar 
 aspirations as ArBB, that is, to target a diverse range of architectures using the 
@@
 are the optimizations studied in the Copperhead \cite{copperhead} and
-Obsidian \cite{obsidian} projects.
+Obsidian \cite{OBSIDIAN-TECHREPORT} projects.
 %

### 2013-10-01 src2/bb/body.tex
apa

@@
 
-\section{Introduction}
+\subsection{Introduction}
 \studname{} is a computer science student working in a project involving programming 
@@
 
-\subsection{Example programs}
+\subsubsection{Example programs}
 
@@
 
-\subsection{Implementation and data structures}
+\subsubsection{Implementation and data structures}
 
@@
 
-\subsection{Problem statement} 
+\subsubsection{Problem statement} 
 \emph{Meanwhile in Professor \docname{}'s Office}\newline %\newline
@@
 \begin{figure*}
-\begin{tabular}{c|c}
-\begin{minipage}[]{0.45\columnwidth}
+%\begin{tabular}{c|c}
+\begin{minipage}[]{0.50\linewidth}
 \begin{small}
@@
 \end{minipage}
-\hspace{0.05\columnwidth} 
-\begin{minipage}[]{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{./bb/spiral1}
+\hspace{0.05\linewidth} 
+\begin{minipage}[]{0.45\linewidth}
+\includegraphics[width=\linewidth]{./bb/spiral1}
 \end{minipage}
-&
-\begin{minipage}[]{0.45\columnwidth}
+%&
+\begin{minipage}[]{0.50\linewidth}
 \begin{small}
@@
 \end{minipage}
-\hspace{0.05\columnwidth}
-\begin{minipage}[]{0.45\columnwidth}
-\includegraphics[width=\columnwidth]{./bb/wall2}
+\hspace{0.05\linewidth}
+\begin{minipage}[]{0.45\linewidth}
+\includegraphics[width=\linewidth]{./bb/wall2}
 \end{minipage}
 
-\end{tabular}
+%\end{tabular}
 \caption{The {\tt spiralIn} and {\tt followWall} programs sent from \studname{}
@@
 
-\section{Compilation of the monadic robot EDSL} 
+\subsection{Compilation of the monadic robot EDSL} 
 \emph{\docname{} receives \studname{}'s latest email}\newline %\newline
@@
 
-\pagebreak
+%\pagebreak
 \begin{small} 
@@
 
-\section{Meeting in \docname{}'s office} %\section{Related Work} 
+\subsection{Meeting in \docname{}'s office} %\section{Related Work} 
 %\emph{Meeting in \docname{}'s office} \newline
@@
 
-\subsection{Related work}
+\subsubsection{Related work}
 
@@
 
-\subsection{Compositionality}
+\subsubsection{Compositionality}
 
@@
 
-\subsection{The Monad laws}
+\subsubsection{The Monad laws}
 
@@
 
-\section{Composing reifiable monadic languages} 
+\subsection{Composing reifiable monadic languages} 
 \label{sec:composing}
@@
 
-\section{Epilogue}
+\subsection{Epilogue}
 
@@
 currently used in the implementation of Obsidian, an EDSL for general
-purpose GPU programming \cite{Obsidian}. Obsidian has language
+purpose GPU programming \cite{Obsidian-Expressive}. Obsidian has language
 constructs which are considerably more advanced than the robot

### 2013-10-01 src2/csort/GPUs.tex
apa

@@
-\section{GPUs and CUDA} 
+\subsection{GPUs and CUDA} 
 %\FloatBarrier
@@
 generations of GPUs. For example, the GPU used in the benchmarks 
-(section \ref{sec:Benchmarks}) has a total of 1344 single-precision floating
+(section \ref{sec:CSORTBenchmarks}) has a total of 1344 single-precision floating
 point CUDA cores and has 2 GB global memory. The very latest GPU architectures 

### 2013-10-01 src2/csort/intro.tex
apa

@@
 \begin{itemize}
-\item We provide measurements (section \ref{sec:Benchmarks}) showing
+\item We provide measurements (section \ref{sec:CSORTBenchmarks}) showing
   that counting sort is a competitive algorithm for sorting keys on the GPU,
@@
   suitable for implementing on the GPU (section \ref{sec:occur}) and
-  performs well (section \ref{sec:Benchmarks}).
+  performs well (section \ref{sec:CSORTBenchmarks}).
 \item The Obsidian implementation of the two sorting algorithms is

### 2013-10-01 src2/embarbb/introduction.tex
apa

@@
 Reference~\cite{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
-steps towards using ArBB as a backend to the embedded language Accelerate~\cite{accelerate}.
+steps towards using ArBB as a backend to the embedded language Accelerate~\cite{ACCELERATEDAMP11}.
 

### 2013-10-01 src2/embarbb/programming.tex
apa

@@
 in an environment that is managed by a monad called {\tt ArBB}. To make this 
-concrete, Figure~\ref{fig:code1} shows the complete procedure from capturing 
+concrete, Figure~\ref{fig:EmbArBBcode1} shows the complete procedure from capturing 
 to execution.
@@
          how to execute the captured function.}
-\label{fig:code1}
+\label{fig:EmbArBBcode1}
 \end{figure}  
@@
 %Matrix multiplication is studied in the benchmarks in section~\ref{sec:Benchmarks}.
-The following EmbArBB implementation of matrix-matrix multiplication is used as a benchmark in section~\ref{sec:Benchmarks}:
+The following EmbArBB implementation of matrix-matrix multiplication is used as a benchmark in section~\ref{sec:EmbArBBBenchmarks}:
 

### 2013-10-01 src2/embarbb/related.tex
apa

@@
 \subsubsection{Data.Array.Accelerate} 
-Accelerate~\cite{accelerate} is an embedded language for general purpose, 
+Accelerate~\cite{ACCELERATEDAMP11} is an embedded language for general purpose, 
 flat data parallel computations on GPUs.
@@
 
-Repa is compared to EmbArBB in the benchmarks in section~\ref{sec:Benchmarks}.
+Repa is compared to EmbArBB in the benchmarks in section~\ref{sec:EmbArBBBenchmarks}.
 

### 2013-10-01 src2/expressive/applications.tex
apa

@@
 construction and we have earlier explored ways to describe
-and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LavaSorter}.
+and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LAVASORTER}.
 Here, we consider iterative descriptions using similar combinators.
@@
 
-Table~\ref{fig:table} shows the performance figures for the large sorter
+Table~\ref{tab:expressive-table1} shows the performance figures for the large sorter
 for $5$ different $512-$input small sorter kernels.
@@
 
-\input{./expressive/table1.tex}
+%\input{./expressive/table1.tex}
+\begin{table}
+\begin{center}
+  \begin{tabular}{| l | r | r | r | r | r | r | }
+    \hline
+        k                             &  $20$ &  $21$ & $22$& $23$ & $24$ & $24$(CPU)  \\ \hline
+    bitonic(CUDA)(512)                &  5 & 12 & 25 & 54 & 117 & 2741   \\      
+    tsort1(512)                       &  4 & 10 & 22 & 47 & 102 & 2742    \\
+    tsort2(256)                       &  4 & 9 & 23 & 45 & 98 & 2741     \\ 
+    vsort1(512)                       &  4 & 10 & 21 & 47 & 102 & 2747    \\ 
+    vsort(256)                        &  4 & 9 & 20 & 44 & 96 & 2783    \\ 
+    \hline
+  \end{tabular}
+\end{center}
+\caption{Sorting time (ms) for $2^k$ inputs, $20 \leq k \leq 24$.  The GPU used is an NVIDIA GTX480.
+Each line shows the result using a particular small $512$ input sort kernel with the
+indicated number of threads as {\tt sortSmall} in the CUDA code above. The small merge kernel used
+is our generated {\tt bmerge2}, with two indices per thread and $512$ inputs. Memory transfer time (GPU-CPU)
+is not shown. In the $2^{24}$ input case using {\tt vsort}, the total sorting time, including
+memory transfers was 141 ms.
+The rightmost column shows the time taken for the C quicksort function {\tt qsort}, compiled with
+{\tt gcc -O3}, to sort the same inputs as those sorted on the GPU on an i7-920 2.8GHz CPU.
+} 
+\label{tab:expressive-table1}
+\end{table}
 
-We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:table2}.
+We also recorded the GPU time spent in the calls to the small sorters alone (using the NVIDIA CUDA Visual profiler), see Table~\ref{tab:expressive-table2}.
 The {\tt tsort1} and {\tt vsort1} kernels, which are built using only pull arrays, 
@@
 
-\input{./expressive/table2.tex}
+%\input{./expressive/table2.tex}
+\begin{table}
+\begin{center}
+  \begin{tabular}{| l | r | r |   }
+
+    \hline
+                              & time & percent \\ \hline
+    bitonic(CUDA)             & 30203 & 32.07\\
+    tsort1                    & 15823 & 19.72  \\
+    vsort1                       & 14955 & 18.76 \\ 
+    tsort2                      & 11562 & 15.37    \\ 
+    vsort                       & 9228 &  12.67 \\ 
+    bitonic (NVIDIA SDK)        & 23815 & 6.55 \\
+    \hline
+  \end{tabular}
+\end{center}
+\caption{time: GPU time (in $\mu$s) spent in calls to small sorter kernels 
+in the initial phase of the sorter for $2^{24}$ inputs. percent: percentage 
+of total GPU time spent in the small sort kernel.
+The final line shows the numbers for the bitonic code (for large sorts) that 
+is distributed with the the NVIDIA SDK. Its structure also starts with a phase 
+of small sorting kernels. It takes 274 ms to sort $2^{24}$ key-value pairs on 
+an NVIDIA GTX480 GPU. It could be sped up by using our {\tt vsort} kernel, 
+and also by (hand) fusing single ``column'' kernels as we did with {\tt iSwap}, 
+although the need to calculate directions of comparators in the bitonic network 
+would complicate this. Our CUDA code is simpler because all comparators
+point in the same direction in the sorter construction.
+} 
+\label{tab:expressive-table2}
+\end{table}
 

### 2013-10-01 src2/expressive/discussion.tex
apa

@@
-\section {Discussion}
+\subsection {Discussion}
 
@@
 The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
-description language~\cite{LavaSorter} was that a relatively small set of combinators went a long way. So, although we introduced three
+description language~\cite{LAVASORTER} was that a relatively small set of combinators went a long way. So, although we introduced three
 combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which

### 2013-10-01 src2/expressive/intro.tex
apa

@@
 Our approach is different from that of other Haskell DSELs targetting
-GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
+GPUs~\cite{ACCELERATEDAMP11,NIKOLA,BARRACUDA}. 
 We do not try to abstract away from 
@@
 {\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
-with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
+with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATEDAMP11}. 
 Obsidian, on the other hand, allows the user to experiment with the

### 2013-10-01 src2/ifl/related.tex
apa

@@
 
-Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\cite{Accelerator}.
+Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\cite{ACCELERATOR}.
 Data Parallel Haskell~\cite{DPH} incorporates Nested Data Parallelism in the style of NESL~\cite{NESL} into Haskell. GPUGen, like Obsidian, aims
-to support GPGPU programming from Haskell~\cite{GPUGen}.
+to support GPGPU programming from Haskell~\cite{GPUGenTalk}.
 It works by translating Haskell's intermediate language, Core, into CUDA, for collective data operations such as scan, fold and map.

### 2013-10-01 src2/ifl/results.tex
apa

@@
-\section{Results} \label{sec:results}
+\subsection{Results} \label{sec:results}
 
@@
 Obsidian to implement sorters. For the sorters, the generated C code 
-performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\cite{LavaSorter},
+performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\cite{LAVASORTER},
 has a running time of 95$\mu$secs for 256 elements (the largest size that we can cope with in a single kernel). This running time can be compared to the 

### 2013-10-01 src2/papp/paperThesis.tex
apa

@@
 and {\tt odds}. For a detailed description of these combinators see
-~\cite{lavaSorter}.
+~\cite{LAVASORTER}.
 
@@
 This subsection shows the implementation of a parallel prefix (also called scan) 
-kernel, known as {\tt sklansky} after J. Sklansky \cite{Sklansky:1960:CSA}.
+kernel, known as {\tt sklansky} after J. Sklansky \cite{Sklansky}.
 This kernel will then be optimised step-by-step using Obsidian. This 
@@
 \subsection{Future work}
-\label{sec:FutureWork}
+% \label{sec:PAPPFutureWork}
 
@@
 
-Vertigo is another GPU programming language embedded in Haskell\cite{Vertigo}.
+Vertigo is another GPU programming language embedded in Haskell\cite{VERTIGO}.
 Vertigo can be used to describe procedureal surfaces and textures and generates
@@
 that target not only GPUs but also other platforms, such as multicore machines.
-Of particular interest to us is Microsoft's Accelerator project~\cite{Accelerator}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
+Of particular interest to us is Microsoft's Accelerator project~\cite{ACCELERATOR}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
 %However, Vertigo is targeting graphical applications.

### 2013-10-01 src2/arbb/paperThesis.tex
apa

@@
 are the optimizations studied in the Copperhead \cite{copperhead} and
-Obsidian \cite{OBSIDIAN-TECHREPORT} projects.
+Obsidian \cite{JSLIC} projects.
 %

### 2013-10-01 src2/csort/Obsidian.tex
apa

@@
-\section{Obsidian} 
+\subsection{Obsidian} 
 \label{sec:Obsidian} 
@@
 
-\subsection{A brief history of Obsidian} 
+\subsubsection{A brief history of Obsidian} 
 \label{sec:OBSHist}
@@
 The Obsidian implementation of pull arrays was influenced by Pan's representation 
-of images as functions from coordinates to colour values \cite{ELLIJFP}. A pull array 
+of images as functions from coordinates to colour values \cite{PAN}. A pull array 
 is a function from index to value and an associated length.
 
-Push arrays were added to Obsidian first in reference \cite{PUSH}. A push array
+Push arrays were added to Obsidian first in reference \cite{Obsidian-Expressive}. A push array
 specifies where elements are to end up in a result array, rather than where 
@@
 %on solving a task the local arrays should be used. 
-\subsection{Obsidian programming example} 
+\subsubsection{Obsidian programming example} 
 \label{sec:sklansky}
@@
 The resulting prefix sum is implemented as shown in reference 
-\cite{LargeScan}, see figure \ref{fig:prefixsum}. There are three steps 
+\cite{ScanCUDA}, see figure \ref{fig:prefixsum}. There are three steps 
 involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
@@
 
-\subsection{A note about generated code}
+\subsubsection{A note about generated code}
 

### 2013-10-01 src2/csort/intro.tex
apa

@@
 
-These algorithms are explored in Obsidian \cite{JSLIC,PUSH}, a domain
+These algorithms are explored in Obsidian \cite{JSLIC,Obsidian-Expressive}, a domain
 specific language targeting GPU programming. The goal of Obsidian is
@@
 
-Sorting has applications in the computer graphics field \cite{sintorn}. 
+Sorting has applications in the computer graphics field \cite{SINTORNSORT}. 
 Example instances of sorting and duplicate element removal in computer graphics 
@@
 programming. Other such approaches include Accelerate
-\cite{Acc} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
+\cite{ACCELERATEDAMP11} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
 higher level approach compared to Obsidian and abstract more from GPU

### 2013-10-01 src2/embarbb/implementation.tex
apa

@@
 A typed interface to the language is supplied using the same phantom types 
-method as used in {\em pan}~\cite{ELLIJFP} and many other Haskell embedded 
+method as used in {\em pan}~\cite{COMPILEEDSL} and many other Haskell embedded 
 languages since then. The choice to use an untyped {\tt Expr} datatype was 

### 2013-10-01 src2/ifl/related.tex
apa

@@
 ``Compiling Embedded Languages'', describing the implementation of Pan have been 
-used in the implementation of Obsidian~\cite{Elliott03:CompileDSEL-JFP}.
+used in the implementation of Obsidian~\cite{COMPILEEDSL}.
 
@@
 In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
-PyGPU is a language for image processing embedded in Python~\cite{LUNDPAPER}. 
+PyGPU is a language for image processing embedded in Python~\cite{PyGPU}. 
 PyGPU uses the introspective abilities of Python and thus bypasses

### 2013-10-01 src2/papp/paperThesis.tex
apa

@@
 see~\cite{JSTECH} and for more general information on embedded language 
-implementation see~\cite{Elliott03:CompileDSEL-JFP}.
+implementation see~\cite{COMPILEEDSL}.
 
@@
 
-Sorting is a popular function to place on the GPU, see~\cite{SORT1,SORT3}.
+Sorting is a popular function to place on the GPU, see~\cite{SINTORNSORT,SORT3}.
 {\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
@@
 the parallel prefix of a large array. This is done in the same way as in
-~\cite{harris07parallel}.
+~\cite{ScanCUDA}.
 The table below shows the results of using the different parallel prefix 
@@
 with the CUDA SDK. This kernel is based on the Brent-Kung network and its 
-implementation is shown in~\cite{harris07parallel}. The {\tt NVIDIA SDK} 
+implementation is shown in~\cite{ScanCUDA}. The {\tt NVIDIA SDK} 
 kernel is highly optimized with regards to its memory access pattern and 

### 2013-10-02 src2/thesis.tex
apa
added missing file
thesis work

@@
 
+%\usepackage{code}
 \usepackage{hyperref}
@@
 
-\newcommand{\bequ}{\begin{quote}}
-\newcommand{\enqu}{\end{quote}}
-\newcommand{\eop}[1]{\mbox{\sl #1}}
-\newcommand{\sube}[2]{#1_{#2}}
-\newcommand{\id}[1]{{\it #1}}
-\newcommand{\kwd}[1]    {\mbox{\textbf{#1}}}
-\newcommand{\str}[1]{{\id "#1"}}
-\newcommand{\lemon}{\emph{lemon}}
-\newcommand{\textt}[1]{\texttt{#1}}
-\newcommand{\txttt}[1]{\texttt{#1}}
+%% \newcommand{\bequ}{\begin{quote}}
+%% \newcommand{\enqu}{\end{quote}}
+%% \newcommand{\eop}[1]{\mbox{\sl #1}}
+%% \newcommand{\sube}[2]{#1_{#2}}
+%% \newcommand{\id}[1]{{\it #1}}
+%% \newcommand{\kwd}[1]    {\mbox{\textbf{#1}}}
+%% \newcommand{\str}[1]{{\id "#1"}}
+%% \newcommand{\lemon}{\emph{lemon}}
+%% \newcommand{\textt}[1]{\texttt{#1}}
+%% \newcommand{\txttt}[1]{\texttt{#1}}
 
-\renewcommand*{\thesection}{\arabic{section}}
+%\renewcommand*{\thesection}{\arabic{section}}
+
+\newcommand{\thesistitle}{Not decided upon}
 
@@
 
-%\input{titlepage}
+\begin{titlepage}
+\begin{centering}
+{\sc Thesis for the Degree of Doctor of Philosophy}
+\vspace{30ex}
+
+{\LARGE\bf\thesistitle}
+
+\vspace{7ex}
+
+\large Bo Joel Svensson
+\vfill
+
+\vspace{1cm}
+\normalsize
+{\sc Department of Computer Science and Engineering}\\
+{\sc Chalmers University of Technology and G\"oteborg University}\\
+G\"oteborg, Sweden 2013
+
+\end{centering}
+\end{titlepage}
 
@@
 \section*{Acknowledgements}
-ACK ACK ACK
 
 \vspace{5mm}
-
-FUNDING FUNDING
+This research has been funded by the Swedish Foundation for
+Strategic Research (which funds the Resource Aware Functional 
+Programming (RAW FP) Project) and by the Swedish Research Council.
 
@@
 
+\section{Plan} 
+\begin{itemize} 
+  \item The Scope
+  \item The Problem 
+  \item The Methods
+  \item The Results 
+\end{itemize}
+
+\begin{itemize} 
+  \item The Context
+  \item Lift head from details 
+  \item Why are there interesting questions around parallelism and GPUs
+  \item The Story
+  \item Reflect on leassons learnt along the way 
+  \item Tell a story around the papers 
+  \item Future Work
+\end{itemize}
+
 \clearpage{}
-\section*{Additional Publications}
+%\section*{Additional Publications}
 
-Papers not included in thesis
+%Papers not included in thesis
 
-\bibliographystyle{splncs}
-\bibliography{intro}
+%\bibliographystyle{plain}
+%\bibliography{intro}
 
-\chapter{Papers}
+\chapter{GPU programming papers}
 
@@
 % ---------------------------------------------------------------------------
+
+\chapter{Retargetable parallel programming papers}
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
+\cleardoublepage
+
+\section{Programming Future Parallel Architectures with Haskell and Intel ArBB}
+%\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
+
+\begin{center} 
+Bo Joel Svensson, Ryan R. Newton
+\end{center}
+
+\input{./arbb/paperThesis}
+
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
 \cleardoublepage
@@
 
+\bibliographystyle{plain}
+\bibliography{thesis}
+\addcontentsline{toc}{chapter}{Bibliography}
+
 \end{document}

### 2013-10-02 src2/thesis.tex
apa

@@
 \usepackage{subfig}
+\usepackage{fancyhdr}
 
@@
 \newcommand{\thesistitle}{Not decided upon}
+\newcommand{\dept}{Department of Computer Science and Engineering}
+\newcommand{\uni}{Chalmers University of Technology and G\"oteborg University}
+\newcommand{\group}{Functional Programming Research Group}
 
@@
 \vfill
+\includegraphics[width=120mm]{./img/ChalmGUtextsvEng}\\[5mm]
+\includegraphics[height=4cm]{./img/ChalmGUmarke}\\
 
@@
 \normalsize
-{\sc Department of Computer Science and Engineering}\\
-{\sc Chalmers University of Technology and G\"oteborg University}\\
+{\sc \dept}\\
+{\sc \uni}\\
 G\"oteborg, Sweden 2013
@@
 
+% ---------------------------------------------------------------------------
+% Tryckortssida
+% ---------------------------------------------------------------------------
+
+\quad \vfill
+
+{\noindent\large\bf\thesistitle} \\
+\noindent Bo Joel Svensson \\
+\noindent ISBN \\
+
+\vspace{1cm}
+
+\noindent\copyright {\sc {Bo Joel Svensson}}, 2013 \\*[5mm]
+\vspace{1cm}
+
+\noindent Technical Report ..D\\
+\noindent ISSN                \\
+\noindent \dept \\
+\noindent \group\\
+
+\vspace{1cm}
+
+\noindent \uni \\
+\noindent SE--412~96~~G\"oteborg, Sweden\\
+\noindent Phone: +46 (0)31--772~1000 \\
+
+\vspace{1cm}
+
+\noindent Printed in Sweden\\
+\noindent Chalmers Reproservice\\
+\noindent G\"oteborg, Sweden 2013
+
+\thispagestyle{empty}
+
 \clearpage
+\pagenumbering{roman}
 
@@
 
-\newpage
+\cleardoublepage
+
+\pagenumbering{arabic}
+\pagestyle{fancy}
+\fancyfoot{}
+\fancyhead[LO]{}
+\fancyhead[RO]{\leftmark}
+\renewcommand{\headrulewidth}{0.0pt}
+\fancyhead[LE,RO]{\thepage}
 

### 2013-10-02 src2/arbb/paperThesis.tex
apa

@@
 \kw{int} main() 
-{ 
+\{ 
   double a[SIZE];
@@
 
-  for ( \kw{int} i = 0; i < SIZE; ++i) { 
+  for ( \kw{int} i = 0; i < SIZE; ++i) \{ 
     a[i] = ...; b[i] = ...;
-  }
+  \}
   dense<f32> va, vb;
@@
   ...
-}
+\}
 \end{Verbatim}

### 2013-10-02 src2/hl/paper-restruct.tex
apa

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{placeins}
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{shapes,snakes,positioning}
\usepackage{fancyvrb}
\usepackage[T1]{fontenc}
%\newcounter{arraycard}

\begin{document}

\conferenceinfo{WXYZ '05}{date, City.} 
\copyrightyear{2005} 
\copyrightdata{[to be supplied]}

\titlebanner{DRAFT}        % These are ignored unless
\preprintfooter{Obsidian: GPU kernel generation}   % 'preprint' option specified.

%\title{Embedded Language for High Performance GPU Computing}
%\title{Towards An Embedded Language\\ for High Performance GPU Computing}  
%\title{An Embedded Language for Resource Aware \\ GPU Kernel Programming}  
\title{A High-Level Embedded Language for Low-Level GPU Kernel Programming}

%\subtitle{Subtitle Text, if any}

\authorinfo{Bo Joel Svensson \and Mary Sheeran} 
           {Chalmers University of Technology}
           {\{joels,ms\} at chalmers dot se}

\maketitle
% ---------------------------------------------------------------------------
%
% --------------------------------------------------------------------------- 
\begin{abstract}

Graphics Processing Units (GPUs) offer potential for very high performance;
they are also rapidly evolving. Obsidian is an embedded language (in Haskell) 
for implementing high performance kernels to be run on GPUs. We would like to 
have our cake and eat it too; we want to raise the level of abstraction at 
which the programmer works from that of CUDA code (which we generate) and 
still to give the programmer fine control over low level details.

We present the user's view of Obsidian by showing case studies and a 
detailed optimisation effort applied to reduction kernels. Our chosen array 
representations (pull and push arrays) provide fine control over the use of 
memory, but are also sufficiently abstract to give guaranteed fusion of composed 
array operations. The case studies also show that Obsidian helps to make 
exploration of the design space for kernels easy. Code reuse 
and composition is easier in Obsidian than in CUDA.

Obsidian uses types to model the hierarchical nature of the GPU, to help 
the programmer to write suitable (and easy to compile) GPU programs. Type 
directed compilation provides the user with language constructs that can be 
used at different levels of the GPU hierarchy; the types constrain the allowed 
behaviour of programs at different levels, while remaining reasonably unobtrusive.
The way in which the hierarchical structure of the GPU architecture
has influenced the design of Obsidian is a major theme of this paper.
Thus, the paper may function as an introduction to GPUs and GPU programming 
for functional programmers.

%We illustrate this by showing details of some kernel optimisations; we make use of two different array representations, which we call pull and push arrays. Obsidian also uses types to model the hierarchical nature of the GPU, to help the programmer to write suitable (and easy to compile) GPU programs. Benchmarks show that the low-level, detailed control offered to the programmer does have an associated pay-off in running time.

%Obsidian is an embedded language for 
%implementing high performance GPU kernels while raising the abstraction 
%level, yet maintaining low-level control. We illustrate this by showing 
%case studies and a detailed optimisation effort applied to reduction kernels.

%Obsidian uses types to model the hierarchical nature of a GPU, both 
%to help the programmer write suitable GPU programs and to allow the 
%implementation of hierarchy agnostic functions.

% We also present a simple and direct method of monad reification.

Benchmarks show that the low-level, detailed control offered to the 
programmer does have an associated pay-off in running time.

%This paper is tutorial in nature and gives much detail about the choices 
%we made in relation to GPU architectures.

\end{abstract}

%\category{CR-number}{subcategory}{third-level}
\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
\category{D.3.4}{Programming Languages}{Processors}[Code generation]

%\terms
%term1, term2
\terms
Languages, Performance

%\keywords
%keyword1, keyword2
\keywords
Data parallelism, array programming, embedded language

% ---------------------------------------------------------------------------
% INTRODUCTION 
% --------------------------------------------------------------------------- 
\section{Introduction}

Graphics Processing Units (GPUs) offer potential for high 
performance implementations of data parallel computations. However, 
GPUs are a bit quirky to program; often, the programmer needs to consider 
very low-level concepts in order to make the best use of the GPU. 
GPU programming models mirror the hierarchy of the GPU, with threads, groups of
threads with special properties and then groups of such groups.
There is a matching memory hierarchy, ranging 
from slow global memory that is accessible to all
threads to fast local memory that allows threads within a single
group of threads to communicate with each other.
For NVIDIA GPUs, both the architecture and the programming model are called
CUDA (for Compute Unified Device Architecture). GPU \emph{kernels} are compiled
to run
on the GPU, while serial {\em host code} controls
how they are launched and supplied with data. Here, the word 
kernel denotes an SPMD (Single Program Multiple Data) program that is 
run on the GPU. Section \ref{sec:GPU} covers GPU architecture.

Obsidian is an embedded language for GPU programming that aims to raise the 
level of abstraction for the GPU programmer, while maintaining control over 
the low-level details that are needed for performance. 
So far,
we have chosen not to try to apply automatic transformations or compiler 
optimisation techniques to a GPU architecture agnostic program and hope 
to get performance. In Obsidian, we want to put the tools to make the 
correct decisions into the hands of the programmer. From Obsidian programs, 
CUDA kernels are generated, while host code must still be written in CUDA.
For key building blocks like scan and reduce, we see a trend towards
the use of fewer and much more complicated kernels~\cite{merrill}, so
that the host code is often relatively simple. We wish to support the construction of these sophisticated kernels.

% ---------------------------------------------------------------------------
% Motivation and Contributions
% --------------------------------------------------------------------------- 
\subsection{Motivation and contributions}

Obsidian aims to provide a higher-level language with a more concise programming style than that of CUDA. We want to achieve 
this while keeping programmer control over the details that influence 
performance. The programmer should be able to experiment with global memory 
access patterns, make tradeoffs between sequential computation per thread 
and parallel computations across threads (using shared memory) and avoid 
unnecessary synchronisation. 
%to synchronise unnecessarily. 
%We illustrate these contributions in the 
%case studies and benchmarks section. (section \ref{sec:CASESTUDIES}).

We envision the user of Obsidian as someone who is familiar with and 
interested in the details of GPU hardware. For this category of users,
the value of Obsidian can be in the rapid prototyping of programming ideas.
Obsidian is also useful for trying to understand the actual cost model
of a particular GPU, since it is so easy to generate programs that are
systematically varied.

Another possible use of Obsidian could be as the code generating backend of 
a domain specific language that could 
benefit from GPU acceleration. Reference ~\cite{BEAUTY} is an example of this.   
%One such experiment can be found in 
%

Our work on Obsidian makes the following contributions: 
\begin{itemize} 
  \item A high-level programming interface that encourages design space exploration through experimentation with 
    details of GPU kernels (section \ref{sec:CASESTUDIES}).
  \item Compositional abstractions for implementation of efficient GPU 
    Kernels (section \ref{sec:CASESTUDIES}).
  \item We use types to model the hierarchical nature of GPUs in our embedded 
    language and to rule out programs that we cannot easily compile to a GPU (section \ref{sec:Program}).
  \item We use type directed compilation of language constructs and 
    implement some GPU hierarchy generic functions (section \ref{sec:Compile}).
  \item A simple and compositional method for reification of monads in the 
    setting of a compiled embedded language (section \ref{sec:Compile}).

\end{itemize}

% ---------------------------------------------------------------------------
% The GPU 
% --------------------------------------------------------------------------- 
\subsection{The GPU and CUDA}
\label{sec:GPU}

This section provides some background information related to GPU architecture. 
Focus is placed on those aspects of GPUs that have influenced our work with 
Obsidian the most.

The GPUs we target with Obsidian are NVIDIA GPUs that support CUDA~\cite{CUDA}.
CUDA is the NVIDIA C-dialect for data-parallel programming on their GPUs. 
We try to stay within the subset of functionality that is common 
between CUDA and OpenCL~\cite{OpenCL}, as we may in future want to generate 
OpenCL programs.

The NVIDIA CUDA capable GPUs are built on a scalable architecture. A GPU 
consists of a number of {\em multiprocessors}; each multiprocessor has a 
number of processing elements (cores) and a local memory that is shared 
between threads running on the cores. Figure \ref{fig:GPU} illustrates these 
concepts. A GPU can come with as few as one of these multiprocessors.
The two GPUs used in our measurements are the GTX 680, which has eight multiprocessors, with a total of 1536 processing cores, and the GT 650M, a laptop GPU, which has 384 such cores.
On these cores, groups of 32 threads called {\em warps} are scheduled. There are a number 
of warp scheduling units per multiprocessor. Within a warp, threads execute in 
lockstep (SIMD); diverging branches, that is those that take different 
paths on different threads within a warp, are serialised, 
leading to performance penalties.

The scalable architecture design also influences the programming model. CUDA 
programs must be able to run on all GPUs from the smallest to the largest.
Hence a CUDA program must work for any number of  multiprocessors.
The CUDA programming model exposes abstractions that fit the underlying 
architecture; there are {\em threads} (executing on the cores), {\em blocks} 
of threads (groups of threads run by a multiprocessor) and finally 
the collection of all blocks, which is called the {\em grid}.

The threads within a block can use the shared memory of the multiprocessor 
to communicate with each other. A synchronisation primitive, {\tt \_\_syncthreads()}, gives all the threads within a block a coherent view
of the shared memory. There is no similar 
synchronisation primitive between threads of different blocks.

The prototypical CUDA kernel starts out by loading data from global memory.
The indices into global memory for an individual thread are expressed
in terms of the unique identifier for that block and thread.
Some access patterns allow memory reads to be \emph{coalesced}, while others do not, giving very poor performance. The patterns that lead to good performance vary somewhat
between different GPU generations, but regular, consecutive accesses by consecutive threads within a warp are best.

A CUDA program is expressed at two levels. Kernels are data-parallel programs that run on the GPU. They are launched by the controlling 
program, which runs on the CPU of the host machine.

\begin{figure} 
\def\s{.50}
\begin{tikzpicture}[scale=(\s)]

% \draw [help lines] (0,0) grid (16,4);
\newcommand{\multiproc} [1] { 
  \draw [black, fill=yellow!30] (#1,-1) -- (3+#1,-1) -- 
                                (3+#1,5) -- (#1,5) -- cycle;
  \foreach \i in {0,...,3} {
    \node [draw, fill=brown!50] (c) at (1+#1,\i+1) {C};
    \node [draw, fill=brown!50] (c) at (2+#1,\i+1) {C};
  } 
  \draw [fill=gray!40] (0.5+#1,0.5) -- (2.5+#1,0.5) -- 
                       (2.5+#1,-0.5) -- (0.5+#1,-0.5) -- cycle; 
  \node () at (1.5+#1,0) {SM};
}

\multiproc{0}
\multiproc{4}
\multiproc{8}
\multiproc{12}

% draw shared memory 
\draw [black, fill=green!50] (0,-1.5) -- (15,-1.5) -- 
                             (15,-3) -- (0,-3) -- cycle;
\node (gm) at (7.5,-2.25) {Global Memory};

\end{tikzpicture}   
\caption{ A GPU. Each multiprocessor has its own shared memory (SM). } 
\label{fig:GPU}
\end{figure}

%\emph{*** Make sure that enough is said about the realities of GPUs: warps, divergence, bank conflicts etc.}

% ---------------------------------------------------------------------------
% PROGRAMMING IN OBSIDIAN 
% --------------------------------------------------------------------------- 
\section{A taste of Obsidian programming}

Obsidian, like CUDA, differentiates between {\em Thread}, {\em Block} and 
{\em Grid} computations. The programmer specifies a sequential computation 
to be carried out by each thread. Many instances of a sequential computation,  
run in parallel over threads, to form a block. A number of such blocks run in 
parallel, to form a grid.

Obsidian is a monadic embedded language. Local computations have the form 
{\tt a0 -> .. -> an -> BProgram b}, where {\tt BProgram}
is a representation of programs that can be performed by a block of threads.

There are two different array representations, {\em Pull} and 
{\em Push} arrays. A pull array naturally represents gather operations,
while a push array captures scattering operations. Reference \cite{Obsidian} 
describes the introduction of push arrays into Obsidian.

As an example, the code below implements a program that reverses the elements 
of an short local array.

%% \begin{small}
%% \begin{verbatim}
%% reverseL :: SPull a -> BProgram (SPull a)
%% reverseL = return . reverse  
%% \end{verbatim} 
%% \end{small}

\begin{small}
\begin{verbatim}
reverseL :: SPull a
                -> BProgram (SPush Block a)
reverseL = liftM push . return . reverse 
\end{verbatim} 
\end{small}

This program does not make use of any specific details of the BProgram monad; 
instead it just composes {\tt return} and {\tt push} with the Obsidian library 
function {\tt reverse}. The {\tt push} function converts a pull array to a push 
array.

We can change the {\tt reverseL} function slightly to have the result stored 
in local (shared) memory. Although this storing in shared memory serves little purpose in this case, it illustrates the programmer's ability to use shared memory.
The {\tt force} function takes 
either a push or a pull array , computes it and places the result in shared memory. 
Using {\tt force} requires a {\tt MemoryOps} constraint on the element type.
%{\tt a} parameter.

%% \begin{small}
%% \begin{verbatim}
%% reverseL' :: MemoryOps a 
%%                  => SPull a 
%%                  -> BProgram (SPull a)
%% reverseL' = force . reverse
%% \end{verbatim} 
%% \end{small}

\begin{small}
\begin{verbatim}
reverseL' :: MemoryOps a
                 => SPull a
                 -> BProgram (SPush Block a)
reverseL' = liftM push . force . reverse
\end{verbatim} 
\end{small}

The result of {\tt force} is always a program yielding a pull array, and hence the 
use of {\tt liftM push}.
% Alternatively the first version could be implemented 
%as {\tt reverseL = liftM push . return . reverse} .

%Figures \ref{fig:rev1} and \ref{fig:rev2} show the
%the generated code.

%\emph{*** Merge these two figures}

\begin{figure} 
\begin{small} 
\begin{verbatim} 
__global__ void reverse(int32_t* input0,
                        uint32_t n0,
                        int32_t* output0){

    output0[((blockIdx.x*256)+threadIdx.x)] = 
      input0[((blockIdx.x*256)+(255-threadIdx.x))];

}

__global__ void reverse'(int32_t* input0,
                        uint32_t n0,
                        int32_t* output0){

    extern __shared__ uint8_t sbase[];
    ((int32_t*)sbase)[threadIdx.x] = 
      input0[((blockIdx.x*256)+(255-threadIdx.x))];
    __syncthreads();
    output0[((blockIdx.x*256)+threadIdx.x)] = 
      ((int32_t*)sbase)[threadIdx.x];

}

\end{verbatim}
\end{small}
\caption{Code generated from the {\tt reverses} program with the local reverse instantiated with 
{\tt reverseL} and {\tt reverseL'}. 
 {\tt reverse'} stores the array in shared memory before copying it to the global result array.} 
\label{fig:rev1} 
\end{figure}

%% \begin{figure} 
%% \begin{small} 
%% \begin{verbatim} 
%% \end{verbatim}
%% \end{small}
%% \caption{Code generated from the {\tt reverseL'} program. This program 
%% uses shared memory} 
%% \label{fig:rev2} 
%% \end{figure}

%The figures for the generated code also contains some information that is 
%absent in the Obsidian code so far. For example there are mentions of {\tt blockIdx.x} 
%and the constant 256.

A CUDA program describes what one thread does and the mapping of that 
program over many threads is done implicitly.
In Obsidian, the programmer specifies the entire kernel
program;  the whole computation is specified, including how that computation 
is spread out over blocks, to form a {\em Grid} program.

The following Obsidian code shows how to split a large array up into 
chunks of 256 elements ({\tt splitUp 256 arr}) and how the local reverse 
kernel is mapped, in parallel, over blocks using {\tt pConcatMap}.

\begin{small} 
\begin{verbatim} 
reverses :: DPull a -> DPush Grid a
reverses arr = pConcatMap reverseL $ splitUp 256 arr 
\end{verbatim} 
\end{small}

\noindent
The {\tt pConcatMap} combinator has the following type: 
%takes a local computation that results in a push array.
\begin{small}
\begin{Verbatim}[samepage=true] 
pConcatMap :: ASize l 
              => (a1 -> Program t (SPush t a)) 
              -> Pull l a1 
              -> Push (Step t) l a
\end{Verbatim}
\end{small}

{\tt pConcatMap} takes a computation resulting in a push array at a specific 
level of the GPU hierarchy, {\tt t}, and a pull array. The computation 
is applied to each element of the pull array and the results are concatenated 
into a push array at the level above {\tt t}. Here, this is used to apply a block 
level computation, giving a grid computation.

The CUDA code generated by the above \verb!reverses! function is shown in 
Figure~\ref{fig:rev1}. Note how it uses the indices of the block 
and thread (\verb!BlockIdx.x! and \verb!ThreadId.x!)
to ensure that each thread accesses the correct part of
the global input array.

%{\tt pConcatMap} takes a computation from pull array to push array at level {\tt t} of 
%the GPU hierarchy and gives a computation at level {\tt Step t}. In this case this is 
%used to map the local reverse kernel in parallel over the blocks of the GPU.

%\emph{*** Put in the type of pConcatMap and explain why it has that type}

%% \begin{small} 
%% \begin{verbatim} 
%% reverses :: DPull b -> DPush Grid b
%% reverses arr = pMap (pushM . reverseL) 
%%                     (splitUp 256 arr) 
%% \end{verbatim} 
%% \end{small}

%\emph{*** Need to say something about the generated CUDA, refer to \ref(fig:rev1}. We have assumed that it is easy to read!}

To reverse a large array (one that cannot fit in one block), the elements
in each block-sized chunk are reversed locally, and the order of those chunks is reversed.

\begin{small} 
\begin{verbatim} 
largeReverse :: DPull a
                -> Push Grid EWord32 a
largeReverse arr =
  pConcat $ reverse $ pMap reverseL $ splitUp 256 arr 
\end{verbatim} 
\end{small}

%\emph{*** Maybe something about what we mean by a global kernel}
The kernels that can be generated using Obsidian take arrays 
in global memory as inputs and output an array to global memory. 
Kernels with this behaviour have a type of the form: 
%% The examples above also show the kind of kernel that we focus on 
%% in Obsidian. An Obsidian program from which we can generate 
%% CUDA code has a type of the form:

\begin{small} 
\begin{Verbatim}[samepage=true]
myKernel :: a0 -> ... -> aN -> DPush Grid b 
\end{Verbatim} 
\end{small}

\noindent
where {\tt a0} to {\tt aN} can be pull arrays or scalars. This choice 
limits the kinds of kernels can be expressed, but
makes compilation of those kernels a lot easier.
For example, a kernel that outputs data to two different global
arrays cannot be expressed.

%\emph{*** Perhaps we should  give a concrete example of a kernel that we cannot
%express}

% ---------------------------------------------------------------------------
% Thread Block Grid Hierarchy figure 
% --------------------------------------------------------------------------- 
\begin{figure} 
\def\s{.50}

\tikzstyle{mybox} = [draw=black, fill=blue!20, very thick,
    rectangle, rounded corners, inner sep=10pt, inner ysep=20pt]

\tikzstyle{wbox} = [draw=black, very thick,
    rectangle, rounded corners, inner sep=10pt, inner ysep=20pt]

\tikzstyle{smallbox} = [draw=black, fill=green!20, very thick,
    rectangle, rounded corners, inner sep=10pt, inner ysep=7pt]

\tikzstyle{fancytitle} =[fill=red, text=white]

\begin{tikzpicture}[scale=(\s)]

%\draw [help lines] (0,0) grid (16,10);

\node [mybox] (Grid){%
    \begin{minipage}{0.15\linewidth}
        Grid
    \end{minipage}
};
\node [mybox,below=0cm of Grid.south] (Block){%
    \begin{minipage}{0.15\linewidth}
        Block
    \end{minipage}
};
\node [mybox,below=0cm of Block.south] (Thread){%
    \begin{minipage}{0.15\linewidth}
        Thread
    \end{minipage}
};

\node [smallbox,right=0cm of Grid.north east, anchor=north west] (gd) {%
    \begin{minipage}{0.25\linewidth}
        {\small Global memory}
    \end{minipage}
};

\node [smallbox,right=0cm of Block.north east, anchor=north west] (bd) {%
    \begin{minipage}{0.25\linewidth}
        {\small Shared memory}
    \end{minipage}
};

\node [smallbox,right=0cm of Block.south east, anchor=south west] (sync) {%
    \begin{minipage}{0.25\linewidth}
        {\small Syncthreads}
    \end{minipage}
};

\node [smallbox,right=0cm of Thread.north east, anchor=north west] (td) {%
    \begin{minipage}{0.25\linewidth}
        {\small Reg to reg}
    \end{minipage}
};

\node [smallbox, right=0cm of Thread.south east, anchor=south west] (seq) {%
    \begin{minipage}{0.25\linewidth}
        {\small Sequential}
    \end{minipage}
};

\node [below=2mm of Thread] (l1) {\small Level};
\node [right=11mm of l1.north east, anchor=north west] (l2) {\small Key concepts};
\node [right=16mm of l2.north east, anchor=north west] (l3) {\small Obsidian};

\node [wbox,inner ysep=15pt,right=0cm of gd.north east, anchor=north west] (obs1) {% 
  \begin{minipage}{0.35\linewidth}
        {\small pMap, generate, DPush, DPull}
    \end{minipage}
};

\node [wbox,inner ysep=15pt,below=0cm of obs1.south] (obs2) {% 
  \begin{minipage}{0.35\linewidth}
        {\small pMap, generate, force, SPush, SPull}
    \end{minipage}
};

\node [wbox,inner ysep=9.5pt, below=0cm of obs2.south] (obs3) {% 
  \begin{minipage}{0.35\linewidth}
        {\small seqFor, seqUntil \\ seqReduce\\ seqScan, SPush, SPull}
    \end{minipage}
};

\end{tikzpicture}   
\caption{The CUDA hierarchy, showing important Obsidian features at each level. } 
\label{fig:Hierarchy}
\end{figure}

% ---------------------------------------------------------------------------
% OBSIDIAN Internals 
% --------------------------------------------------------------------------- 
\section{Obsidian internals}

%% Obsidian, like cuda, differentiates between {\em Thread}, {\em Block} and 
%% {\em Grid} computations. The programmer can specify a sequential computation 
%% to be carried out by each thread. This sequential computation can be run in 
%% parallel over many threads called a block. Then a collection of such blocks, 
%% a grid, can be specified.

%% Using Obsidian, the programmer can specify computations that span many blocks.
%% However the blocks cannot communicate in a synchronised manner. This again 
%% matches CUDA, where you need to divide programs that needs to synchronise 
%% between blocks into several kernels.

Many array languages supply a fixed set
of parallel primitives like {\em Reduce}, {\em Scan} and {\em Permute}, but
do not give the user control over how they are implemented.
Obsidian programmers are, instead, encouraged to experiment with the
implementation of such primitives, to maximise performance, for example
by trading off sequential and parallel work or by varying memory access patterns.
Obsidian's two array representations (pull and push arrays) are central to providing 
this fine control.

% ---------------------------------------------------------------------------
% Elements
% --------------------------------------------------------------------------- 
\subsection{Scalars}

An {\em expression} data type defines the language
available at the scalar level of Obsidian. This is implemented as a GADT (Generalised Algebraic Data Type) to provide a type safe 
programmer's interface:

\begin{small} 
\begin{Verbatim}[samepage=false]
data Exp a where
  Literal :: Scalar a 
             => a 
             -> Exp a

  BlockIdx :: DimSpec 
              -> Exp Word32 
  ThreadIdx :: DimSpec
               -> Exp Word32

  Index   :: Scalar a => 
             (Name,[Exp Word32]) 
             -> Exp a

  If      :: Scalar a 
             => Exp Bool
             -> Exp a 
             -> Exp a 
             -> Exp a

  BinOp   :: (Scalar a,
              Scalar b, 
              Scalar c) 
             => Op ((a,b) -> c) 
             -> Exp a 
             -> Exp b 
             -> Exp c

  UnOp    :: (Scalar a, 
              Scalar b)
             => Op (a -> b)            
             -> Exp a 
             -> Exp b 
\end{Verbatim}
\end{small}
Most constructors are standard, implementing arithmetic and conditionals 
via the {\tt BinOp}, {\tt UnOp} and {\tt If} constructors. There are also two 
GPU specific variables, {\tt ThreadIdx} and {\tt BlockIdx} referring to the thread 
identity. 
%We have chosen not to have tuples in the expression data type but rather using 
%Haskell tuples to represent these.

We provide type synonyms for commonly used element types: 
\begin{small} 
\begin{Verbatim}[samepage=true]
...
type EInt32  = Exp Int32
type EInt64  = Exp Int64
...
type EWord32  = Exp Word32 
type EWord64  = Exp Word64 
type EFloat  = Exp Float  
type EDouble = Exp Double 
type EBool   = Exp Bool 
\end{Verbatim}
\end{small}

% ---------------------------------------------------------------------------
% PUSH AND PULL ARRAYS 
% --------------------------------------------------------------------------- 
\subsection{Pull arrays}

%\emph{*** I am not sure I know what gather operations really are. Is there
%a classic example that we could show?}

Pull arrays are perfectly suited for so-called gather operations (read 
many values, compute one value), as well as directly suitable for parallelisation 
on a GPU. A pull array uses the well known representation of arrays as 
functions from index to element.

\begin{small} 
\begin{verbatim} 
data Pull s a = Pull s (EWord32 -> a) 
\end{verbatim} 
\end{small}

Very similar representations are used 
in many other embedded languages. The 
embedded language Pan ~\cite{ELLIJFP} used a similar representation for 
images and it has been used since in Obsidian, Feldspar ~\cite{FELDSPAR} 
and Repa ~\cite{REPA}. In Obsidian, there is a {\tt force} function that,
when applied to a pull array, writes the elements of that array 
into shared memory using as many threads as the pull array is long. Hence, 
the {\tt force} function can be used to express parallelism in Obsidian.

%Pull arrays have a length associated with them. The length of a pull array 
%represents its inherent capability of parallelism.

The length of an array, the {\tt s} parameter,
can be either static (a Haskell known-at-compile-time value) or dynamic 
(a runtime value). Static lengths are used for local (or block) 
computations, with those lengths determining the number of threads needed and the shared memory consumption. Dynamic 
lengths are used for global computations (across many blocks). After specifying 
a local computation, it can often be applied over a variable number of blocks. 
The dynamic lengths capture this common use case.

There are type synonyms for the dynamic and static versions of  pull arrays:

\begin{small}
\begin{verbatim}
type SPull = Pull Word32 
type DPull = Pull EWord32 
\end{verbatim}
\end{small}

% ---------------------------------------------------------------------------
% Thread, Block and Grid Computations 
% --------------------------------------------------------------------------- 
\subsection{Thread, Block and Grid Computations} 
\label{sec:Program}

The programs that can be specified in Obsidian
are represented in a {\tt Program} data type.
This data type also represents the lowest level programming interface 
exposed by Obsidian. In reference ~\cite{CSORT} we make use of these low-level 
capabilities to implement histograms using atomic operations (not shown here).

The {\tt Program} data type has a parameter, {\tt t}, that can be 
either {\tt Thread}, {\tt Block} or {\tt Grid}, corresponding to the GPU hierarchy, as shown in Figure~\ref{fig:Hierarchy}.

\begin{small}
\begin{verbatim} 
data Step a -- A step in the hierarchy
data Zero

type Thread = Zero 
type Block  = Step Thread 
type Grid   = Step Block  
\end{verbatim} 
\end{small}

Below are some selected parts of the  {\tt Program} data type. There 
is a sequential, {\tt SeqFor}, loop that runs a sequential loop body a given 
number of times. In this case, both the loop body and the loop itself are 
{\tt Thread} Programs. More interesting is the {\tt ForAll} case that introduces 
 different kinds of parallelism depending on the {\tt t} parameter.
{\tt ForAll} executes a program body at the level below several times in parallel. 
If the body of the {\tt ForAll} loop is a {\tt Thread} computation the {\tt ForAll} loop 
itself is a {\tt Block} computations. Likewise, if the body is a {\tt Block} computation, 
the loop itself is a {\tt Grid} computation. 
%% {\tt ForAllBlocks} executes several blocks 
%% in parallel over a {\tt Grid}. The {\tt Thread}, {\tt Grid} and {\tt Block} 
%% parameters rule out any nesting of loops that we cannot easily compile 
%% to GPU code.

\begin{small} 
\begin{Verbatim}[samepage=true]
data Program t a where 
  Identifier :: Program t Identifier 
  Assign :: Scalar a
            => Name 
            -> [EWord32]
            -> (Exp a)
            -> Program Thread ()
  Allocate :: Name 
              -> Word32 
              -> Type 
              -> Program Block ()

  ... 
  SeqFor :: EWord32 
            -> (EWord32 -> Program Thread ())
            -> Program Thread ()
  ForAll :: EWord32 
            -> (EWord32 -> Program t ())
            -> Program (Step t) ()

  ... 
  Sync :: Program Block ()

\end{Verbatim}
\end{small}

%The listing above also includes the constructors {\tt Identifier}, {\tt Assign},
% {\tt Allocate} and {\tt Sync}. 
The {\tt Assign} and {\tt Allocate} constructors represent assignment into 
shared memory or a variable, and allocation of shared memory. The {\tt Identifier} 
constructor is used to obtain fresh Identifiers and is not meant to be made available
to the programmer. The {\tt Sync} constructor corresponds to the 
CUDA {\tt \_\_syncthreads()} function. 
%These constructors are shown here to help the presentation 
% of push arrays in the coming section.

To allow the programmer to use the monadic {\tt do} notation when constructing Obsidian 
programs, the {\tt Program} data type is made an instance of {\tt Monad}. The details 
of how this is done are outlined in section \ref{sec:Compile}.

%% This is helpful when implementing function that are program kind polymorphic. 
%% There will be examples of these kinds of functions.

%% More details about the implementation of the {\tt Program} GADT and its 
%% associated monad instance can be found in section \ref{sec:Implementation}.

% ---------------------------------------------------------------------------
% Push Arrays
% --------------------------------------------------------------------------- 
\subsection{Push arrays} 
\label{sec:PushArrays}

A push array captures the idea of scattering, focussing on where 
elements end up rather than where they come from. In Obsidian, a push array 
is defined as follows:

\begin{small}
\begin{verbatim}
data Push p s a =
  Push s ((a -> EWord32 -> TProgram ()) -> Program p ())
\end{verbatim}
\end{small}

A push array has two parts, a length and a higher-order function. 
The idea is that this function is provided with a method to write an 
element into memory (a {\em write-function}), and returns a program 
that uses that write-function a number of times. Note that the 
write-function produces a {\tt TProgram} (short for {\tt Program Thread}), This 
restricts our push arrays from representing nested data parallelism. 
This concept is best illustrated by showing how to turn a pull array into a push array.

\begin{small} 
\begin{verbatim} 
convToPush :: SPull a -> SPush Block a
convToPush arr =
  Push n $ \wf ->
   ForAll (fromIntegral n) $ \tid -> wf (arr ! tid) tid
  where
    n = len arr       
\end{verbatim}
\end{small}

\noindent
This is actually one case of the definition of the \verb!push! function, which we will see shortly.
In the push array case, the length represents the maximum index the push array 
might write to.
%The code listing below shows how a pull arrays are implemented 
%in Obsidian. The implementation of push arrays will be shown later, in section
%\ref{sec:PushArrays}.

% ---------------------------------------------------------------------------
% Interplay 
% --------------------------------------------------------------------------- 
\subsection{Push and pull array interplay} 
\label{sec:interplay}

Both push and pull arrays represent methods of computing an array. They 
do not directly correspond to actual data stored in memory, but rather 
are virtual. This enables fusion of operations on both push and pull arrays. 
One such example is map fusion: \verb!map f . map g = map (f . g)!. It is 
only by using {\tt force} that the programmer can explicitly demand that the 
operations should not be fused but that the actual intermediate result should 
be stored in memory. In Obsidian, {\tt force} is also the only way 
to convert a push array into a pull array. 
The type of {\tt force} is: 
\begin{small} 
\begin{Verbatim}[samepage=true] 
force :: (Write p, MemoryOps a) 
         => p Word32 a 
         -> Program Block (Pull Word32 a)
\end{Verbatim}
\end{small}

\noindent
where {\tt p} is an array (push or pull) with elements that can be stored into memory. 
Note that the return type of {\tt force} is a program yielding a pull array. This pull 
array represents reading from a named area of memory.

Converting a pull array to a push array is cheap and is done using a function called {\tt push} that behaves differently (sequentially or in parallel) at different levels
of the GPU hierarchy. This is implemented as a type class:

\begin{small}
\begin{Verbatim}[samepage=true] 
class Pushable t where
  push :: ASize s => Pull s e -> Push t s e 
\end{Verbatim} 
\end{small}

The {\tt ASize} class has instances for both the static and dynamic lengths 
and is used to obtain conversion to a representation used internally; it 
provides a function called {\tt sizeConv} which can convert both static and 
dynamic lengths to a {\tt Word32} expression.

%\emph{*** We have not explained what the expression data type is
%        and what EWord32 is etc.}

There are two instances of the Pushable class, one for thread programs and 
one for block programs:

\begin{small}
\begin{Verbatim}[samepage=true] 
instance Pushable Thread where
  push (Pull n ixf) =
    Push n $ 
      \wf -> SeqFor (sizeConv n) $ \i -> wf (ixf i) i

instance Pushable Block where
  push (Pull n ixf) =
    Push n $ 
      \wf -> ForAll (sizeConv n) $ \i -> wf (ixf i) i
\end{Verbatim} 
\end{small}

Now, the {\tt push} function captures just one possible way to convert 
a pull array into push array -- with one write
per thread. In the block case, it is possible to vary 
the amount of work per thread using the {\tt pushN} function provided
by the library. 
Indeed, the conversion of pull arrays into push arrays can be done in many ways. 
For example, when pushing more than one element per thread one could make 
choices between placing them consecutively or far apart.
We intend to provide a variety of such special purpose push functions,
but would also expect Obsidian users to tailor push functions to
their own particular needs.

% ---------------------------------------------------------------------------
% Compilation to CUDA 
% --------------------------------------------------------------------------- 
\subsection{Compilation to CUDA}
\label{sec:Compile}

Obsidian is a monadic language. The {\tt Program} 
data type is made a monad by adding the following constructors:

\begin{small} 
\begin{Verbatim}[samepage=true] 
data Program t a where

   ...

  Return :: a -> Program t a
  Bind   :: Program t a 
            -> (a -> Program t b) 
            -> Program t b
\end{Verbatim} 
\end{small}

and adding a very simple {\tt Monad} instance:

\begin{small} 
\begin{verbatim} 
instance Monad (Program t) where
  return = Return
  (>>=) = Bind
\end{verbatim} 
\end{small}

The choices above influence the compilation procedure. Implementing a compiler 
for the {\tt Program} data type is not entirely straightforward. Our approach 
is described in further detail in reference ~\cite{BB}; the presentation 
here focuses instead on the type directed compilation of the {\tt ForAll} 
primitives.

The first step in the compilation process is to turn the monadic {\tt Program}
into a first-order representation. This first-order representation also makes 
explicit what level of parallelism the {\tt ForAll} constructor refers to. 
It does this by having separate constructors for parallel loops over 
all threads of a block {\tt SForAll} and parallel loops over blocks 
{\tt SForAllBlocks}. The first-order representation is a list of statements. 
The extra value of type {\tt a} that is carried around with each statement 
is used to keep track of meta information in later compilation passes.

\begin{small}
\begin{Verbatim}[samepage=true]

type IMList a = [(Statement a,a)]

type IM = IMList ()

data Statement t =
  forall a. (Show a, Scalar a) 
     => SAssign Name [EWord32] (Exp a)

  ... 
  | SSeqFor String (EWord32) (IMList t)

  ... 
  | SForAll (EWord32) (IMList t) 
  | SForAllBlocks (EWord32) (IMList t)

  ... 
  | SAllocate Name Word32 Type

  ... 
  | SSynchronize

\end{Verbatim} 
\end{small}

The type directed part of the compilation is implemented using a type class 
called {\tt Compile}. The {\tt Compile} class houses just one function, called
{\tt compile}. The {\tt compile} function takes a splittable value supply and 
a {\tt Program t a} and outputs intermediate code, of type {\tt IM}, and a 
value of type {\tt a}. The significance of the value supply and of the value 
of type {\tt a} will be made clear after considering the instances of 
{\tt Compile}.

\begin{small} 
\begin{Verbatim}[samepage=true] 
class Compile t where
  compile :: Supply Int -> Program t a -> (a,IM)
\end{Verbatim} 
\end{small}

For each of the levels of the GPU hierarchy, {\tt Grid}, {\tt Block} and {\tt Thread},
there is an instance of the {\tt Compile} class. The simplest case is {\tt Thread},
which does not at all support parallel loops; it just delegates the compilation 
process to a function that takes care of all non type directed cases (called {\tt cs}).

%% \emph{*** I changed Zero to Thread to match the following two cases.
%% Is that correct?}

\begin{small} 
\begin{Verbatim}[samepage=true] 
instance Compile Thread where 
  compile s p = cs s p 
\end{Verbatim} 
\end{small}

The cases for blocks takes care of compilation of the {\tt ForAll} 
constructor into a use of the {\tt SForAll} statement. Likewise, the 
grid case compiles into an {\tt SForAllBLocks} statement.

\begin{small} 
\begin{Verbatim}[samepage=true] 
instance Compile Block where
  compile s (ForAll n f) = (a,out (SForAll n im))
    where
      p = f (ThreadIdx X)
      (a,im) = compile s p
  compile s p = cs s p

instance Compile Grid where
  compile s (ForAll n f) = (a, out (SForAllBlocks n im))
    where 
      p = f (BlockIdx X)
      (a,im) = compile s p
  compile s p = cs s p

\end{Verbatim} 
\end{small}

The {\tt out} function used in the compilation function has type \\ {\tt Statement () -> IM}
and generates a one element {\tt IM} list with unit attached as meta-data.

Currently, only the {\tt ForAll} constructor is given the type directed treatment. 
All other language features are tied to a specific level of the GPU or are
completely independent of any such concerns. Taking care of these operations 
as well as the monadic {\tt Return} and {\tt Bind} cases is in the hands of the 
{\tt cs} function. This is also where the value supply and the {\tt a} parameter 
play a role.

\begin{small}
\begin{verbatim} 
cs :: Compile t => Supply Int -> Program t a -> (a,IM) 
cs i Identifier = (supplyValue i, [])
... 
\end{verbatim}
\end{small}

The {\tt Identifier} constructor does not correspond to any code executed on the GPU. 
It just helps in reifying the {\tt Program} monad. Now it can 
be seen that the value of type {\tt a} that is being returned is the monadic value of 
the {\tt Program}. This follows the same monad reification technique as described 
in reference~\cite{BB}.

The constructors mentioned above are directly compiled into the corresponding 
statements.

\begin{small}
\begin{verbatim} 
cs i (Assign name ix e) = ((),out (SAssign name ix e))
cs i (Allocate id n t) = ((),out (SAllocate id n t))
cs i (Sync)            = ((),out (SSynchronize))
\end{verbatim}
\end{small}

Turning to the monadic {\tt Bind} and {\tt Return} constructors, 
it becomes clear why we need that extra value of type {\tt a} in the return 
type of the compile functions.
We need a value to pass to the function in 
the {\tt Bind} constructor.

\begin{small}
\begin{verbatim} 
cs i (Bind p f) = (b,im1 ++ im2) 
  where
    (s1,s2) = split2 i
    (a,im1) = compile s1 p
    (b,im2) = compile s2 (f a)

cs i (Return a) = (a,[])
\end{verbatim}
\end{small}

After compiling the monadic {\tt Program} into a list of statements, 
some characteristics are collected.

The first analysis done on the {\tt IM} 
discovers the number of threads needed per block to run the program. This 
is done by a pass over the {\tt IM} and looking at the size  
of each parallel {\tt SForAll} loop. The largest {\tt SForAll} loop present 
decides how many threads are needed. The implementation of any other 
{\tt SForAll} loop will be wrapped in a conditional that turns 
off any unused threads.

Note that there is no similar analysis to discover the maximum extents 
of {\tt SForAllBlocks} loops. The reason is that the restriction
to kernels of the form:

\begin{small} 
\begin{Verbatim}[samepage=true]
myKernel :: a0 -> ... -> aN -> DPush Grid b 
\end{Verbatim} 
\end{small}

\noindent means that there can be no sequence of {\tt SForAllBlocks} loops in the program. In other 
words, an Obsidian Program for which we can generate code will contain at most 
one {\tt SForAllBlocks} loop. Some care has also to be taken with
the language constructs that result in grid push arrays, so that none of these 
creates illegal programs.

% \emph{*** would wrong word in above sentence.}

In the current {\tt IM}, all arrays in shared memory will have unique 
names. These names are obtained via the value supply during the reification 
of the {\tt Program} monad. These names will be used in a memory mapping phase.
A {\em liveness} analysis of the arrays collects information needed to lay 
out the arrays in the shared memory of the GPU. Currently no spilling of 
arrays that do not fit into shared memory is performed, so it is very possible 
to design an Obsidian program that will not work because of shared memory 
restrictions. After the arrays are assigned addresses in the shared memory,
a renaming phase replaces array names with these locations. This is done 
while also going from the {\tt IM} data type to a third internal representation 
of the program. This third representation is a C-like AST extended with some 
GPU specific concepts (just as CUDA is an extended C). We strive to make this representation sufficiently general to be turned into 
either CUDA code or OpenCL code, but currently only output CUDA.

% ---------------------------------------------------------------------------
% More example programs 
% --------------------------------------------------------------------------- 
\section{Case studies} 
\label{sec:CASESTUDIES}

%\emph{*** SOME TEXT THAT LEADS INTO THE CASE STUDIES}

The following case studies show first a typical application, the Mandelbrot fractal, in which the required work is naturally highly parallel.
The remaining case studies consider key building blocks, reduce and scan, that have data-flow graphs involving much more communication. Then, the programmer must find ways to decompose the algorithm in order to fit within
the constraints imposed by the GPU architecture. These two case studies illustrate the use of Obsidian for exploring the design space when implementing a particular algorithm.

\subsection{Fractals}

The Mandelbrot fractal is generated by iterating a function:

\begin{equation*} 
 z_{n+1} = z_n^2 + c
\end{equation*}

\noindent where $z$ and $c$ are complex numbers. The method to generate the fractal 
presented here is based on a sequential C program from reference 
~\cite{Fractals}. However, the problem is embarrassingly parallel.

In order to get the Mandelbrot image, one lets $z_0$ be zero and maps the 
$x$ and $y$ coordinates of the image being generated to the real and imaginary 
component of the $c$ variable.

\begin{small}
\begin{verbatim} 
xmax =  1.2 :: EFloat
xmin = -2.0 :: EFloat
ymax =  1.2 :: EFloat
ymin = -1.2 :: EFloat
\end{verbatim}
\end{small}

To obtain the well known and classical image of the set, we let 
the real part of $c$ range over $-2.0$ to $1.2$ as the $x$ coordinate range from
$0$ to $512$ and similarly for the y coordinate and the imaginary component.

\begin{small}
\begin{verbatim} 
-- For generating a 512x512 image
deltaP = (xmax - xmin) / 512.0
deltaQ = (ymax - ymin) / 512.0
\end{verbatim}
\end{small}

%\emph{*** something wrong with this sentence
%          Also too many ``we''s around here
%          Is my fix correct?}
%Now, for each pixel of the image we want to generate the number of times the function 
%presented above should be iterated. 
%Now, the image is generated by iterating the function presented above. The 
%number of iterations it takes for the function to reach beyond a certain bound 
%is used to decide the color to output.
The image is generated by iterating the function presented above.
We have chosen to map the 
height of the image onto blocks of executing threads. Each row of the
image is computed by one block of threads. This means that for a 512$\times$512 
pixel image, 512 blocks each of 512 threads are needed.

The function to be iterated is defined below and called {\tt f}. This 
function will be iterated until a condition holds (defined in the function 
{\tt cond}). We count the number of iterations and if they reach 512 we 
break out of the iteration.

\begin{small} 
\begin{verbatim} 
f bid tid (x,y,iter) = 
  (xsq - ysq + (xmin + (w32ToF tid) * deltaP),
   2*x*y + (ymax - (w32ToF bid) * deltaQ),
   iter+1) 
  where
    xsq = x*x
    ysq = y*y

cond (x,y,iter) = ((xsq + ysq) <* 4) &&* iter <* 512  
  where
    xsq = x*x
    ysq = y*y

\end{verbatim} 
\end{small}

The number of iterations that are executed is used to decide
which colour to assign to the corresponding pixel. In the function below,
{\tt seqUntil} iterates {\tt f} until the condition {\tt cond} 
holds. Then the number of iterations is extracted and used to compute 
a colour value (out of 16 possible values).

%\emph{*** Shouldn't mod and times 16 be done with shifts?}
%  Yes, possibly, But NVCC discovers this I have been told.

\begin{small} 
\begin{verbatim} 
iters :: EWord32 
         -> EWord32 
         -> TProgram (SPush Thread EWord8)
iters bid tid =
  return $ fmap extract 
         $ seqUntil (f bid tid) cond  (0,0,1)
  where
    extract (_,_,c) = ((w32ToW8 c) `mod` 16) * 16
\end{verbatim}
\end{small}

%% \begin{small} 
%% \begin{verbatim} 
%% iters bid tid = 
%%   liftM extract $ seqUntil (f bid tid) cond  (0,0,1)
%%   where
%%     extract (_,_,c) = ((word32ToWord8 c) `mod` 16) * 16
%% \end{verbatim}
%% \end{small}
The final step is to run the iterations for each pixel location. 
This is done by using the {\tt generate} function: 
\begin{small}
\begin{Verbatim}[samepage=true] 
generate :: ASize l
            => l
            -> (EWord32 -> Program t (SPush t b))
            -> Push (Step t)  l b
\end{Verbatim}
\end{small}

\noindent
{\tt generate} is a function that can be used at the thread or block level.
It takes a function from index to computation for that index and executes a 
number of such computations at the level above. Here, it is used to compute 
an element per thread, a block,  and then to execute a number of such blocks over a grid. 
%at both 
%the block and grid level.

% \emph{*** generate needs to be explained!}

\begin{small} 
\begin{verbatim} 
genRect :: MemoryOps b
           => EWord32
           -> Word32
           -> (EWord32 -> EWord32 
                 -> TProgram (SPush Thread b))
           -> DPush Grid b 
genRect bs ts p = 
   generate bs (\bid -> return $ generate ts (p bid))
\end{verbatim}
\end{small} %$

The largest possible image that can be obtained using this particular 
method is 1024$\times$1024. This is because of the maximum threads per block 
limitation of current GPUs. The code below generates a 512$\times$512 element 
image. 
\begin{small} 
\begin{verbatim}
mandel = genRect 512 512 iters 
\end{verbatim}
\end{small}

This case study illustrates the use of the {\tt generate} function at 
two different levels of the GPU hierarchy. We also make use of 
iterations within each thread.

%% \begin{figure} 
%% \includegraphics[width=.50\linewidth]{mandel}
%% \includegraphics[width=.50\linewidth]{plate1}
%% \caption{ \emph{To the left:} Image of the Mandelbrot set with the parameters 
%% from table \ref{fig:mandeltable}. row ``mandel''.\newline
%% \emph{To the right:} A zoom in on the Mandelbrot image using the parameters 
%% from row ``mandel1''. }
%% \label{fig:mandel}
%% \end{figure}

%% \begin{figure} 
%% \includegraphics[width=.50\linewidth]{plate2}
%% \includegraphics[width=.50\linewidth]{plate3}
%% \caption{\emph{To the left:} Mandelbrot image using parameters from row ``mandel2'' in fig. \ref{fig:mandeltable}. \newline
%% \emph{To the right:} Parameters from row ``mandel3''.}
%% \label{fig:mandel2}
%% \end{figure}

\begin{figure} 
\begin{small}
\begin{tabular}{| l | r | r | r | r | r | }
\hline 
  Image & xmax & xmin & ymax & ymin & ms \\ \hline 
  mandel & 1.2 & -2.0 & 1.2 & -1.2 & 2.65 \\ \hline
  mandel1 & -0.690906 & -0691060 & 0.387228 & 0.387103 & 1.78 \\ \hline
  mandel2 & -0.723005 & -0.793114 & 0.140974 & 0.037822 & 6.79 \\ \hline
  mandel3 & -0.745388 & -0.745464 & 0.113030 & 0.112967 & 5.7 \\ \hline
\end{tabular}
\caption{ Running times for the Mandelbrot program.
The parameters ({\tt xmax}, {\tt xmin}, {\tt ymax}, {\tt ymin}) correspond
to zooming in on different areas of the generated images.}
%This table contains the parameters used when generating the 
%images in figures \ref{fig:mandel} and \ref{fig:mandel2}. The last column 
%contains the time it took to generate the images in ms. The timing results 
%where obtained using an GT650m GPU. }
\label{fig:mandeltable} 
\end{small}

\end{figure}

%% ------------------------------------------------------------------------
%% Reduction
%% ------------------------------------------------------------------------
\subsection{Reduction}  
\label{sec:redOpt}

\input{reductionfigs.tex}

In this section, we implement a series of reduction kernels. The Obsidian 
reductions are implemented to take an associative operator as a parameter. 
In the benchmarking, the operation used will be addition and the elements 
will be $32$ bit integers.

%\emph{ *** Probably good to say what type the elements of the arrays
%in the benchmarked CUDA has}

Some of the reduction 
kernels will also require that the operation is commutative. 
%A typical reduction is to sum up the elements of an array, here this is used 
%as the running example. In this section we implement a series of reduction 
%kernels. 
To illustrate the kind of low level control that an 
Obsidian programmer has over expressing details of a kernel, we show a series 
of reduction kernels, each with different optimisations applied. Many of the 
optimisations that are applied to the kernels are influenced by a 
tutorial presentation from NVIDIA~\cite{RedOpt}.

%The figures \ref{fig:red1} and \ref{fig:redSeq} also illustrate various 
%tradeoffs that can be made while implemented reduction.

%\emph{*** refer to fig:red1 and fig:redSeq in the right place.}

The kernels implemented in the following subsections use a variant of {\tt force}, 
called {\tt unsafeForce} that attempts to be clever about the insertion 
of synchronisation into the generated code. If the array that is being forced 
is shorter than the warp width, a synchronisation is not needed. It is called 
unsafe because when using push arrays there is still no absolute guarantee 
that the threads do not later write to indices that are used by threads that belong to another 
warp. Programmers should use {\tt unsafeForce} with care; it 
is however safe together with pull arrays.

This section focuses on local reduction kernels. The construction of large 
reduction algorithms from these kernels will be illustrated in 
section~\ref{sec:Benchmarks}.

\subsubsection{Reduction 1}

Our first attempt at reduction combines adjacent elements repeatedly.
This approach is illustrated on the left of Figure \ref{fig:red1}. 
In Obsidian, this entails splitting the array into its even and its odd elements and using {\tt zipWith} to combine these. 
This procedure is then repeated until there is only one element left. This 
kernel will work for arrays whose length is a power of two.

\begin{small} 
\begin{Verbatim}[samepage = true] 
red1 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red1 f arr
  | len arr == 1 = return $ push arr
  | otherwise    = 
    do
      let (a1,a2) = evenOdds arr
      arr' <- unsafeForce $ zipWith f a1 a2
      red1 f arr'   
\end{Verbatim}
\end{small}

The above code describes what one block of threads does. To spread 
this computation out over many blocks and thus perform many simultaneous 
reductions, {\tt pConcatMap} is used (as before):

\begin{small} 
\begin{Verbatim}[samepage = true] 
mapRed1 :: MemoryOps a 
              => (a -> a -> a) 
              -> DPull (SPull a) 
              -> DPush Grid a
mapRed1 f = pConcatMap $ red1 f
\end{Verbatim}
\end{small} %$

This kernel does not perform well (as can be seen in Figure~\ref{fig:reducetable}). 
This may be attributed to the memory access pattern used. Remember that
one gets better performance on memory access when consecutive threads
access consecutive elements, which happens if
each thread accesses elements that are some stride apart.

\subsubsection{Reduction 2}

{\tt red2} lets each thread access elements that are further apart. It does 
this by halving the input array and then using {\tt zipWith} on the halves (see Figure~ \ref{fig:red1}). 
This choice can only be made if the operator is commutative.

\begin{small} 
\begin{Verbatim}[samepage = true] 
red2 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red2 f arr
  | len arr == 1 = return $ push arr
  | otherwise    = 
    do
      let (a1,a2) = halve arr
      arr' <- unsafeForce $ zipWith f a1 a2
      red2 f arr'   
\end{Verbatim}
\end{small}

%This leads to the kind of access pattern that can be seen in the right half 
%of figure \ref{fig:red1}.

\subsubsection{Reduction 3}

The two previous implementations of reduce write the final value into shared 
memory (as there is a {\tt force} 
in the very last stage). This means that the last element is stored into shared 
memory and then directly copied into global memory. This can be avoided by cutting 
the recursion off at length 2 instead of 1, and performing the last operation 
without issuing a {\tt force}.

\begin{small} 
\begin{Verbatim}[samepage = true] 
red3 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red3 f arr
  | len arr == 2 = 
      return $ 
      push $ 
      singleton $ f (arr ! 0) (arr ! 1) 
  | otherwise    = 
    do
      let (a1,a2) = halve arr
      arr' <- unsafeForce $ zipWith f a1 a2
      red3 f arr'
\end{Verbatim}
\end{small}

\subsubsection{Reduction 4}

Now we have a set of three basic ways to implement reduction  and can start experimenting with adding 
sequential, per thread, computation. {\tt red4} uses
{\tt seqReduce}, which is provided by the Obsidian library and 
implements a sequential reduction that turns into a for loop in the generated 
CUDA code. The input array is split into 
chunks of 8 that are reduced sequentialy. The partial
results are reduced using the previously implemented ({\tt red3}).

\begin{small}
\begin{Verbatim}[samepage = true] 
red4 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red4 f arr =
  do
    arr' <- force $  
            pConcatMap (return . seqReduce f) 
                       (splitUp 8 arr)
    red3 f arr' 
\end{Verbatim}
\end{small}%$

As can be seen by the running times in Figure~\ref{fig:reducetable}, this optimisation 
did not come out well. The problem is that it reintroduces memory coalescing issues (see Figure \ref{fig:redSeq}).

\subsubsection{Reduction 5}

With {\tt red5}, the coalescing problem is dealt with by defining a new 
function to split up the array into sub arrays. The idea is that the 
elements in the inner arrays should be drawn from the original array in a strided fashion.

\begin{small} 
\begin{Verbatim}[samepage = true] 
coalesce :: Word32 -> SPull a -> SPull (SPull a)
coalesce n arr =
  mkPull s $ \i ->
   mkPull n $ \j -> arr ! (i + fromIntegral s * j)
  where
    s = (len arr) `div` n 
\end{Verbatim}
\end{small}

\noindent
With {\tt coalesce} in place of {\tt splitUp}, {\tt red5} can be defined as:

\begin{small} 
\begin{Verbatim}[samepage = true] 
red5 :: MemoryOps a
           => (a -> a -> a)
           -> SPull a
           -> BProgram (SPush Block a)
red5 f arr =
  do
    arr' <- force $  
            pConcatMap (return . seqReduce f)
                       (coalesce 8 arr)
    red3 f arr'

\end{Verbatim}
\end{small}%$

\subsubsection{Reductions 6 and 7}

Lastly, we try to push the tradeoff between number of threads and sequential 
work per thread further. {\tt red6} and {\tt red7} represent  
changing {\tt red5} to reduce 16 and 32 elements in the sequential phase.
The performance of the fastest of these kernels is very satisfactory, at a level where the kernel is {\em memory bound}, that is constrained
by memory bandwidth.

\subsubsection{Larger local reductions} 
Figure \ref{fig:reducetablel} shows running time results for kernels that are 
very similar to the ones described above, but with a larger number 
of elements per local reduction (that is per block). This means that all the kernels need 
to introduce a small amount of sequentiality at the beginning. This is 
done by introducing a new force function:

\begin{small}
\begin{Verbatim}[samepage=true] 
unsafeForce' arr | len arr > 1024 = return arr
                 | otherwise      = force  arr 
\end{Verbatim} 
\end{small}

\noindent
{\tt unsafeForce'} exploits the fact that {\tt force} doesn't actually
change the elements of the array. 
Skipping the write into shared memory should still give the desired effect,
so replacing it with \verb!return! is safe. Two parallel phases are fused into one parallel phase that performs sequential work.

This is a new way of introducing sequential work into kernels; unlike {\tt seqReduce}, it does not result in a for loop, but rather in an unrolled loop in the kernel.

%% \begin{small} 
%% \begin{Verbatim}[samepage = true]

%% naiveReduceLocal :: MemoryOps a
%%                     => (a -> a -> a)
%%                     -> SPull a
%%                     -> BProgram (SPush Block a)
%% naiveReduceLocal f arr
%%   | len arr == 1 = return $ push Block arr
%%   | otherwise    = 
%%     do
%%       let (a1,a2) = halve arr
%%       arr' <- force $ zipWith f a1 a2
%%       naiveReduceLocal f arr'   
%% \end{Verbatim}
%% \end{small}

% --------------------------------------------------------------------------- 
% Measurements 
% ---------------------------------------------------------------------------
%% \begin{figure} 
%% \begin{small}
%% \begin{tabular}{| l | c | c | c | c | r | }
%% \hline 
%%   Kernel & Threads & Shared mem. & Sync opt. & Occupancy & ms \\ \hline 
%%   naive & 1024 & 16384 & N & 94.3  & 8.2 \\ \hline
%%   warp &  1024 & 16384 & Y & 94.4  & 8.2 \\ \hline
%%   seq1 & 512 & 4096 &  Y & 94.6   & 4.4 \\ \hline
%%   seq2 & 256 & 2048 & Y & 96.7 & 2.5 \\ \hline
%%   seq2b & 256 & 2048 & Y & 91.2  & 10.8 \\ \hline
%% \end{tabular}
%% \caption{ This table contains the execution time of a grid of 8192 reductions 
%% of each 2048 elements. The timing results where obtained using an GT650m GPU. }
%% \label{fig:reducetable} 
%% \end{small} 
%% \end{figure}

\begin{figure} 
\begin{small}
\centering
\begin{tabular}{| l | c | c | c | c | r | }
\hline 
  Kernel & Threads & Shared mem. & ms & ms*\\ \hline 
  red1 & 1024 & 16384 & 9.619 & 1.678 \\ \hline
  red2 & 1024 & 16384 & 9.174 & 1.838 \\ \hline
  red3 & 1024 & 16384 & 8.795 & 1.855 \\ \hline
  red4 & 256 & 2048  & 10.839 & 2.120  \\ \hline
  red5 & 256 & 2048  & 2.608 &  0.445 \\ \hline
  red6 & 128 & 1024  & 2.471 &  0.442 \\ \hline
  red7 & 64 & 512  & 2.524 &    0.439 \\ \hline
\end{tabular}
\caption{ Execution time of a grid of 8192 reductions , each of
2048 elements on 
and on a GTX680 (ms*). The timing figures were obtained using the NVIDIA Profiler.}
\label{fig:reducetable} 
\end{small} 
\end{figure}

\begin{figure} 
\begin{small}
\centering
\begin{tabular}{| l | c | c | c | c | r | }
\hline 
  Kernel & Threads & Shared mem. & ms & ms*\\ \hline 
  red1l & 1024 & 16384 & 8.883 & 1.762 \\ \hline
  red2l & 1024 & 16384 & 5.721 & 1.030  \\ \hline
  red3l & 1024 & 16384 & 5.377 & 0.980  \\ \hline
  red4l & 256 & 2048  & 11.202 & 2.201 \\ \hline
  red5l & 256 & 2048  & 2.476  & 0.440 \\ \hline
  red6l & 128 & 1024  & 2.471  & 0.442 \\ \hline
  red7l & 64 & 512  & 4.712    & 0.752\\ \hline
\end{tabular}
\caption{ Execution time of a grid of 4096 reductions, each of 4096 elements on a GT650m GPU (ms)
and on a GTX680 (ms*). The timing figures were obtained using the NVIDIA Profiler.}
\label{fig:reducetablel} 
\end{small} 
\end{figure}

%% \begin{small} 
%% \begin{Verbatim}[samepage = true] 
%% reduceLocal :: MemoryOps a 
%%                => (a -> a -> a) 
%%                -> SPull a 
%%                -> BProgram (SPush Block a)
%% reduceLocal f arr | len arr == 2 = return $ push Block arr'
%%   where arr' = mkPullArray 1 $ \_ -> f (arr ! 0) (arr ! 1)
%% reduceLocal f arr | len arr < 32 =
%%   do
%%     let (a1,a2) = halve arr
%%     arr' <- write $ zipWith f a1 a2
%%     reduceLocal f arr'
%% reduceLocal f arr  = 
%%   do
%%     let (a1,a2) = halve arr
%%     arr' <- force $ zipWith f a1 a2
%%     reduceLocal f arr'

%% \end{Verbatim}
%% \end{small}

%% \begin{small} 
%% \begin{Verbatim}[samepage = true] 
%% coalesce :: Word32 -> SPull a -> SPull (SPull a)
%% coalesce n (Pull m ixf) 
%%   = Pull (m `div` n) $ \i -> 
%%       Pull n $ \j -> ixf (i + (sizeConv (m `div` n)) * j)

%% reduce :: MemoryOps a 
%%           => (a -> a -> a) 
%%           -> SPull a 
%%           -> Program Block (SPush Block a)
%% reduce f arr =
%%   do
%%      arr' <- force $  mapT (return . seqReduce f) (coalesce 8 arr)
%%      reduceLocal f arr'

%% reductions2 :: (ASize l, MemoryOps b)
%%                => (b -> b -> b)
%%                -> Pull l (SPull b)
%%                -> Push Grid l b
%% reductions2 f = mapG $ reduce f
%% \end{Verbatim}
%% \end{small}

%% \emph{*** Coalesced memory accesses}

%% \emph{*** Sequential computations}

%% \emph{*** Do not sync if inside a Warp}

\subsection{Scan}

Scan computes all the prefix sums of a sequence of values using 
a binary associative operator (and is familiar to Haskell
programmers as the {\tt scanl1} function).

Given an array of values $a_0,a_1,\ldots,a_n$ and associative operator $\oplus$,
the scan operation computes a new array:

\begin{equation*} 
%\begin{array}{rcr} z & = & a \\ f(x,y,z) & = & x + y + z \end{array} 
  \begin{array} {rcl} 
    s_0 & = & a_0 \\
    s_1 & = & a_0 \oplus a_1 \\ 
    & \ldots & \\
    s_n & = & a_0 \oplus a_1 \oplus \ldots \oplus a_n
  \end{array}
\end{equation*}

\input{scanfigs.tex}

%Currently algorithms like scan pose a challenge to Obsidian.
Figure \ref{fig:Sklansky}
shows a standard divide and conquer decomposition of scan. Data
flows from top to bottom and boxes with two inputs are
operators. At each level, exactly half of the boxes are operators
and in an imperative language the algorithm would naturally be implemented
in-place.
%From this figure we can 
%read that half of the elements are altered in each parallel level of the algorithm 
%while the other half are untouched.
Since we cannot express in-place algorithms 
currently in Obsidian, this means that we need to copy unchanged values into a new array 
during each phase. The memory mapping gives us a ping-ponging implementation, but an in-place implementation would likely be faster.
%\emph{*** Too vague. Do we know??}

Also, the threads now do two different things 
(copy, or perform operation). 
One can have as many threads as elements, but then each must have
a conditional to decide whether to be a copy or operation thread.
%This can be implemented either by launching as many threads 
%as elements and performing a conditional in each thread to decide if this is a copy thread 
%or an operation thread. 
Or we can launch half as many threads and have each of them perform 
both a copy and an operation. 
We will show code for both of these options; the first is
easier to implement.
%We address these issues further in section \ref{sec:Future}, on future work.

The Obsidian code below implements the scan network from Figure \ref{fig:Sklansky}, using as many threads as there are elements.
% and uses a 
%conditional to decide whether to copy or perform operations.

\begin{small} 
\begin{Verbatim}[samepage=true]
sklansky :: (Choice a, MemoryOps a)
            => Int
            -> (a -> a -> a)
            -> SPull a
            -> BProgram (SPush Block a)
sklansky 0 op arr = return $ push arr
sklansky n op arr =
  do 
    let arr1 = binSplit (n-1) (fan op) arr
    arr2 <- force arr1
    sklansky (n-1) op arr2
\end{Verbatim} 
\end{small} % $

This is a kernel generator;
the (Haskell) {\tt Int} parameter can be used to generate kernels of various sizes by setting 
it to the log base two of the desired array size.

The {\tt binSplit} combinator used in {\tt sklansky} is part of the Obsidian library 
and used to implement divide and conquer algorithms. It divides an array recursively in half 
a number of times (first parameter) and applies a computation to each part (second parameter). 
The operation applied in this case is {\tt fan}:

\begin{small}
\begin{Verbatim}[samepage=true]
fan :: Choice a
       => (a -> a -> a)
       -> SPull a
       -> SPull a
fan op arr =  a1 `conc`  fmap (op c) a2 
    where 
      (a1,a2) = halve arr
      c = a1 ! fromIntegral (len a1 - 1)
\end{Verbatim}
\end{small}

It is the array concatenation ({\tt conc}) used in this function that introduces 
conditionals into the generated code.

\subsubsection{two elements per thread}

%Here, we try to optimise out implementation of Scan and also try to introduce 
%sequential work. By introducing sequential work per thread we can perform 
%larger instances of scan per GPU block. This is good since there is a maximum 
%number of threads in each block.

%Obsidian provides the means to drop down to very low-level programming when 
%the higher level primitives do not operate as well as desired. The code listing 
%below illustrates this.

Both to avoid conditionals and to allow
for larger scans per block, we move to two elements per thread.
Each phase of the algorithm is a parallel for loop that is executed 
by half as many threads as there are elements to scan. The body of the loop
performs one operation and one copy, using bit-twiddling to compute indices.
Note the use of two {\em write functions} in sequence.
Similar patterns were used in our implementations of sorting networks~\cite{Obsidian}, for similar reasons.

%This code is a generator that given an integer that 
%refers to which parallel level of the algorithm we are interested in. It then 
%creates two separate parallel for loops; each running half as many threads as there 
%are elements. The first loop performs the operations between element and the 
%second just copies data.

\begin{small} 
\begin{Verbatim}[samepage=true] 
phase :: Int 
         -> (a -> a -> a) 
         -> SPull a 
         -> SPush Block a
phase i f arr =
  Push l $ \wf -> ForAll sl2 $ \tid ->
  do
    let ix1 = insertZero i tid
        ix2 = flipBit i ix1
        ix3 = zeroBits i ix2 - 1
    wf (arr ! ix1) ix1
    wf (f (arr ! ix3) (arr ! ix2) ) ix2
  where
    l = len arr
    l2 = l `div` 2
    sl2 = sizeConv l2
\end{Verbatim} 
\end{small}

For an input of length $2^n$, $n$ phases
are composed as follows:

\begin{small} 
\begin{Verbatim}[samepage=true] 
sklansky2 :: MemoryOps a 
             => Int
             -> (a -> a -> a)
             -> SPull a
             -> BProgram (SPush Block a)
sklansky2 l f = compose [phase i f | i <- [0..(l-1)]]
\end{Verbatim}
\end{small}

{\tt compose} takes a Haskell list of programs and composes 
them in sequence, forcing intermediate arrays between each step.
%Using Haskell for this kind of meta-programming is something that
%we plan to investigate further.

\begin{small} 
\begin{Verbatim}[samepage=true] 
compose :: MemoryOps a
           => [SPull a -> SPush Block  a] 
           -> SPull  a
           -> BProgram (SPush Block  a)
compose [f] arr = return $ f arr
compose (f:fs) arr = 
  do
    let arr1 = f arr
    arr2 <- force arr1
    compose fs arr2
\end{Verbatim} 
\end{small} % $

Comparing the two kernels {\tt sklansky} and {\tt sklansky2} in the 
NVIDIA profiler indicates that {\tt sklansky2}, while being faster than 
{\tt sklansky} in many cases, has a worse memory loading behaviour. This 
indicates that tweaking the way data is loaded into shared memory may 
be beneficial in that kernel.

\begin{small} 
\begin{Verbatim}[samepage=true] 
sklansky3 :: MemoryOps a
             => Int
             -> (a -> a -> a)
             -> SPull a
             -> BProgram (SPush Block a)
sklansky3 l f arr =
  do
    im <- force$ load 2 arr 
    compose [phase i f | i <- [0..(l-1)]] im 
\end{Verbatim}
\end{small}  % $

Here we use {\tt load 2} to realise loading of 2 elements per thread 
but in a strided way that is more likely to lead to a good memory access 
pattern. This function is an example of one of the custom ways to create a 
push array from a pull array mentioned in section \ref{sec:interplay}.
The results of these optimisations are shown in Figure~\ref{fig:scangraphs}.
To make a seriously fast local and global scan, we would need to add
more sequential work, as we did in the reductions. We do not yet know
if our inability to do in-place updates prevents us from competing with
hand-tuned CUDA scan code. 
%We hope to report more experiments in the final
%version of the paper.

%% Not shown here are the implementations of {\tt rv}, {\tt cv} and {\tt o}. 
%% {\tt rv} computes from a thread id one of the indices from which to read 
%% an element and perform operation on. {\tt o} provides the second element 
%% that also corresponds to the location of where to write the target element. 
%% {\tt cv} computes from the thread id an index that is supposed to be copied 
%% straight from the input array to the output array. The implementation 
%% of these functions are left out since they are highly specific to the 
%% algorithm.

%% \begin{figure} 
%% \begin{small}
%% \centering
%% \begin{tabular}{| l | c | c | r | }
%% \hline 
%%   Kernel    & Threads & ms & ms*    \\ \hline 
%%   sklansky  & 1024    & x1 & x2     \\ \hline
%%   sklansky2 & 1024    & y1 & y2     \\ \hline

%% \end{tabular}
%% \caption{CAPTION CAPTION CAPTION}
%% \label{fig:scantable} 
%% \end{small} 
%% \end{figure} 
\begin{figure}
\begin{minipage}{.5\linewidth}
\begin{tikzpicture} [scale = 0.5]
\begin{axis} [xmax = 3, 
              xmin = 0, 
              ymax = 6,
              ymin = 0,
              % ytick = data,
              xtick = {0,1,2,3},
              xticklabels = {128,256,512,1024},
              legend entries = {Sklansky, Sklansky2, Sklansky3},
              legend pos = north west,
              xlabel = \#Elements, 
              ylabel = Time in ms,
              title = 4096 Blocks,
  ]         
  \addplot[color = black, 
           mark  = *,
           mark size = 2pt] coordinates {
    (0, 0.364 )
    (1, 0.847 )
    (2, 2.098 ) 
    (3, 5.762 )
  };
  \addplot[color = black, 
           mark  = square*,
           mark size = 2pt] coordinates {
    (0, 0.408 )
    (1, 0.564 )
    (2, 1.292 ) 
    (3, 3.336 )
  };
  \addplot[color = black, 
           mark  = diamond*,
           mark size = 2pt] coordinates {
    (0, 0.341 )
    (1, 0.514 )
    (2, 1.179 ) 
    (3, 2.957 )
  };

\end{axis}
\end{tikzpicture} 
\end{minipage}
\begin{minipage}{.5\linewidth}
\begin{tikzpicture} [scale = 0.5]
\begin{axis} [xmax = 3, 
              xmin = 0, 
              ymax = 6,
              ymin = 0,
              % ytick = data,
              xtick = {0,1,2,3},
              xticklabels = {128,256,512,1024},
              legend entries = {Sklansky, Sklansky2, Sklansky3},
              legend pos = north west,
              xlabel = \#Elements,
              title = N/BlockSize Blocks
  ]         
  \addplot[color = black, 
           mark  = *,
           mark size = 2pt] coordinates {
    (0, 2.882 )
    (1, 3.372 )
    (2, 4.194 ) 
    (3, 5.762 )
  };
  \addplot[color = black, 
           mark  = square*,
           mark size = 2pt] coordinates {
    (0, 3.226 )
    (1, 2.242 )
    (2, 2.576 ) 
    (3, 3.336 )
  };
  \addplot[color = black, 
           mark  = diamond*,
           mark size = 2pt] coordinates {
    (0, 2.723 )
    (1, 2.041 )
    (2, 2.352 ) 
    (3, 2.957 )
  };

\end{axis}
\end{tikzpicture}

\end{minipage}
\caption{\emph{left:}The time it takes to run 4096 blocks of varying sizes using our
different implementations of the Sklansky kernel. \newline
\emph{Right:} varies the number of blocks so that the number of elements is the 
same in each experiment. These results where obtained on an NVIDIA GT 650m using the 
NVIDIA Profiler.}
\label{fig:scangraphs}
\end{figure}

\section{Combining kernels to solve large problems} 
\label{sec:Benchmarks}

With Obsidian, we can experiment with details during the implementation 
of local computations. In section \ref{sec:CASESTUDIES}, we saw that the 
description of a local kernel involves its behavior when spread out 
over many blocks. However, solving large problems must sometimes make 
use of many different local kernels or the same local kernel used 
repeatedly. For now, we turn to CUDA for describing these algorithms. 
Here the procedure of making use of combinations of kernels is explained 
using large reduction as an example.

\subsection{Large reductions}

We implement reduction of large arrays by running local kernels on blocks 
of the input array. If the local kernel reduces $ELT\_PER\_BLOCK$ elements to $1$ then 
this first step reduces $BLOCKS * ELT\_PER\_BLOCK$ elements into $BLOCKS$ partial results. 
The procedure is then repeated on the $BLOCKS$ elements until there is one 
value.

Chosing {\tt red5l} from section \ref{sec:redOpt} and a block-size 
and local size of 4096 gives a 16 Million element reduction in only two stages.

The CUDA code we need to write by hand is very simple and only describes the 
actual order of the kernels we wish to launch. We make no changes to the 
kernels themselves. The listing below shows an outline
of the CUDA code for launching {\tt red5l} to give $2^{24}$ (= 16777216)
element reduction.

%% The listing in appendix \ref{app:A} shows an outline
%% of the CUDA code for launching {\tt red5l} to give $2^{24}$ (= 16777216)
%% element reduction.   
%% \pagebreak

\begin{small}
\begin{Verbatim}[samepage=true]

/* Define sizes to work on */ 
#define BLOCKS 4096
#define ELT_PER_BLOCK 4096
#define N (BLOCKS * ELT_PER_BLOCK)

int main(void) {

  /* Allocate host and device arrays */ 
  float *input, result; 
  float *dinput, *dresult;

  input  = (float*)malloc(N*sizeof(float));

  cudaMalloc((void**)&dinput,N*sizeof(float));
  cudaMalloc((void**)&dresult,BLOCKS*sizeof(float));

  /* Generate input data */
  ...

  /* Copy data to device */
  cudaMemcpy(dinput,
             input,
             N*sizeof(float),
             cudaMemcpyHostToDevice);

  /* Launch kernels */ 
  red5l<<<BLOCKS,256,512*sizeof(float)>>>(dinput,
                                             N,
                                             dresult);

  red5l<<<1,256,512*sizeof(float)>>>(dresult,
                                        4096,
                                        dresult);

  /* Copy result from device */ 
  cudaMemcpy(&result,
             dresult,
             1*sizeof(float),
             cudaMemcpyDeviceToHost);

  /* Process data on host side, if needed */
  ...

  return SUCCESS; 
\end{Verbatim} 
\end{small}

This is just a very simple example of what CUDA code can look like. Part of 
the motivation for showing it is to make the presentation complete. 
We currently write this kind of CUDA skeleton by hand; section \ref{sec:Future} 
describes future work to improve this situation.

\begin{figure}
\centering 
\begin{small}
\begin{tabular}{| l | r | r | r | }
\hline 
  System            & Elements & ms     & ms*   \\ \hline 
  Obsidian and CUDA & $2^{24}$  & 2.480  & 0.446 \\ \hline
  Accelerate        & $2^{24}$  & 2.716  & 0.478 \\ \hline
\end{tabular}
\caption{Running times of $2^{24}$ element reduction using Obsidian and CUDA 
or Accelerate. The results were obtained on a NVIDIA GT650m GPU (ms) 
and on a NVIDIA GTX680 GPU (ms*).}

\label{fig:CUDATABLE} 
\end{small}
\end{figure}

Table \ref{fig:CUDATABLE} shows the running time for the Obsidian together with 
CUDA $2^{24}$ element reduction compared to Data.Array.Accelerate. 
Again, the kernel generated from Obsidian is memory bound.
These values are 
interesting since they indicate that the Accelerate reduction skeleton is 
very well optimised, while maintaining generality. We expected this, since 
reduction corresponds directly to the Accelerate Fold skeleton, which
we believe to have been manually optimised as a one-time effort. It is also 
interesting that adding the second phase of reduction, in this case 
one instance of 4096 elements, only negligibly influences the running time.
%Remember that in section \ref{sec:redOpt}, {\tt red5l} runs in 2.5ms as well. 
%Choosing to round to one decimal place obscures any influence by that last phase.

%\subsection{Scans} 
%\cite{OLA} \cite{LargeScan}

% ---------------------------------------------------------------------------
% RELATED WORK 
% --------------------------------------------------------------------------- 
\subsection{Related work}

There are many languages and libraries for GPU programming. Starting at the 
low-level end of the spectrum we have CUDA ~\cite{CUDA}. CUDA is NVIDIA's 
name for the programming model and extended C language for their GPUs. 
It is 
the capabilities of CUDA that we seek to match with Obsidian,
while giving the programmer the benefits of having Haskell as a meta programming 
language. 
%% a high-level and 
%% functional programming language. 
%% \emph {*** Comment from Mary: I think this is not the right place to talk about
%% these limitations. Is it possible (nonetheless) to list what parts of
%% CUDA are (er) lifted to the Obsidian level? Just saying ``benefits of a high
%% level functional programming language seems too vague. We need to list them.]
%% }

%% There 
%% are low-level features of CUDA that we, as of yet, do not have control over 
%% from Obsidian. This is mainly related to in-\emph{warp} computations using 
%% registers and shuffle instructions to avoid using shared memory. A warp is a 
%% group of 32 threads (on current generation GPUs), that operate in a true SIMD 
%% fashion. In CUDA, being an imperative language with mutable arrays, in-place 
%% algorithms can also be easily expressed. Apart from these limitations it is 
%% the capabilities of CUDA that we seek to match with Obsidian,
%% while offering the programmer some of the benefits of a high-level and 
%% functional programming language.

While remaining in the imperative world, but going all the way to the 
other end of the high-level - low-level spectrum, we have the NVIDIA
Thrust Library~\cite{Thrust}.
Thrust offers a programming model where details of GPU architecture are 
completely abstracted away. Here, the programmer expresses algorithms 
using building blocks like: \emph{Sort}, \emph{Scan} and \emph{Reduce}.

Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
~\cite{accelerate}. The abstraction level is comparable to that of Thrust. In 
other words, Accelerate hides most GPU details from the programmer. Accelerate 
provides a set of operations (that are parallel and suitable for GPU execution, 
much like in Thrust) implemented as skeletons. Recent work has permitted
the optimisation of Accelerate programs using fusion techniques to decrease
the number of kernel invocations needed (see reference ~\cite{ACCOPT}).
%Accelerate programs are being optimised using using fusion 
%techniques to decrease the number of kernel invocations needed. 
It seems to us that when using Accelerate the programmer has no control over how to 
decompose his computation onto the GPU or how to make use of shared memory 
resources. For many users, remaining entirely within Haskell will be
a big attraction of Accelerate.

Nikola ~\cite{NIKOLA} is another language embedded in Haskell that occupies the 
same place as Accelerate and Thrust on the abstraction level spectrum.

The systems above are all for flat data-parallelism, Bergstrom 
and Reppy are attempting nested data-parallelism by implementing a 
compiler for the NESL language for GPUs ~\cite{NESL}.

%\emph{*** Related work non-Haskell so as not to be too insular. Copperhead}

The Copperhead \cite{Copperhead} system compiles a subset of Python to run 
on GPUs. Much like other languages mentioned here, Copperhead identifies 
usages of certain parallel primitives that can be executed in parallel 
on the GPU (such as reduce, scan and map). But Copperhead also 
allows the expression of nested data-parallelism and is in that 
way different from both Accelerate and Obsidian.

In reference ~\cite{Finance}, Oancea et al. use manual transformations
to study a set of 
compiler optimisations for generating efficient GPU code from high-level 
and functional programs based on {\tt map}, {\tt reduce} and {\tt scan}. 
They tackle performance problems related to GPU programming, such as bad 
memory access patterns and diverging branches.
Obsidian enables easy exploration of decisions related to these issues.
%These are concepts that we try 
%to give the Obsidian programmer language support to reason about.

In the implementation of Obsidian we use a very simple and direct approach 
to monad reification. Related to this work is the approach based on a 
continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{NEIL} is also related. 
Our approach to monad reification is simpler but less general.

% ---------------------------------------------------------------------------
% FUTURE WORK 
% --------------------------------------------------------------------------- 
\section{Future Work} 
\label{sec:Future}
%\emph{*** TEXT LEADING INTO THE THREE SUBSECTIONS}

\subsection{Increase low level control}

The capabilities of the GPU are changing and evolving. For example, it is now 
possible to do warp local computation that exchange values between threads 
using a set of {\tt shuffle} instructions. These kernels do not need to use 
shared memory to the same extent as the ones we generate. It would be interesting 
to try to incorporate these capabilities Obsidian. However, this 
would mean that plans to target both OpenCL and CUDA would have to be retired.

In reference ~\cite{CSORT}, we experiment with atomic operations 
and are forced to write our code in a very imperative style. The addition of mutable 
arrays and a set of operations on them could improve this situation 
while giving up purity. We will explore what the addition of mutable arrays to 
Obsidian would mean as future work.

Being able to do in-place operations on arrays is beneficial to some algorithms. 
There are cases where in-place operation both simplifies implementation and improves performance.
Feldspar, an EDSL for Digital Signal Processing being developed in our group at Chalmers, has mutable arrays. In that case, it was Fast Fourier Transform
that pushed us towards this choice. Scan, in which elements must be copied
unchanged between arrays, while others are operated upon, may also push us towards mutable arrays. We need
to do more experiments before deciding whether or not to take this step,
particularly as we value the simpilicity of the current version of Obsidian.
%In 
%the case of scan, we need to copy many elements unchanged from one array to 
%another while performing operations on the other elements. We need a way to 
%implement in-place operations in Obsidian, maybe based on the mutable array 
%proposition above.

\subsection{To automate or not to automate} 
Obsidian makes use of types that model the GPU hierarchy. This limits 
the kind of programs one can express but also ensures that what we can write 
is suitable for the GPU. An alternative is to remove these types and at the 
same time make the compiler more clever. This would move some of the 
required GPU detail knowledge from the programmer into compiler transformations. 
We have recently started a project to investigate this change of Obsidian
together with a Master's student.

\subsection{To optimise or not to optimise} 
The version of Obsidian described here does not try to use any compiler optimisations techniques. Instead, we are expecting that the 
CUDA compiler will apply a good set of techniques, from common subexpression 
elimination to more GPU specific transformations. As future work, we 
will investigate to what extent this reliance on the CUDA
compiler is justified. Also, we want to 
see if there are high-level optimisations that we can perform more easily 
on our intermediate representations of GPU kernels than what is possible 
on low-level CUDA code.

\subsection{Kernel Coordination}

In section \ref{sec:Benchmarks}, we implemented large reduction using 
CUDA and kernels generated using Obsidian. This is something we want to 
be able to avoid. We envision a kernel coordination language embedded 
in Haskell where the programmer can express the kind of sequential 
and parallel compositions of kernels; note that this is different from 
composing two local computations into one, which we already have the 
capability to do. Having an embedded language for expressing full 
algorithms that make use of many different kernels would further 
improve the usability of Obsidian. We have started a project with the 
goal of implementing this language together with a Master's student.

%\subsection{Exploiting Haskell, metaprogramming}
%We see opportunities for exploiting the fact that Obsidian is embedded
%in Haskell during kernel generation. Methods similar to the ``clever circuits'' techniques used in Lava may be interesting to explore. The use of search
%to find algorithmic decompositions that are well matched to the GPU is
%also an avenue that we would like to explore.

\subsection{What should the combinators be?}
In this latest version of Obsidian, we have at last got a system
that permits the construction of high performance kernels. But it is
clear that we need to think hard about the set of combinators that
we use to describe common patterns of parallel and sequential computation.
Our current choice is a bit ad hoc and needs to be generalised and made
more systematic. We will also explore the use of search to find algorithmic 
decompositions that are well matched to the GPU, exploiting the fact 
that Obsidian is embedded in a powerful host language.

% ---------------------------------------------------------------------------
% CONCLUSION
% --------------------------------------------------------------------------- 
\section{Conclusion}

%\emph{*** Come back to the contributions from the first page}

%\emph{*** Say something about how getting good performance
%out of machines with strange memory hierarchies is a
%general, important and pressing problem.}

Obsidian lends itself well to the kind of experimentation with low level 
GPU details that allow for the implementation of efficient kernels. This is 
illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
used in the NVIDIA tutorial ~\cite{RedOpt}. The case study also show hows we 
can compose kernels and thus reuse prior effort. 
%For example, {\tt red4} 
%to {\tt red7} all make use of {\tt red3} for their parallel phases.

The use of GPU hierarchy generic functions makes the kernel code concise. The 
{\tt pMap} and {\tt pZipWith} functions are applicable both at block and grid 
level and are compiled into suitable parallel loops. The types that are used 
to model the GPU hierarchy also rule out any programs that we cannot 
very easily compile to the GPU. An alternative to this was discussed in 
the Future work section \ref{sec:Future}.

Our monad reification method certainly simplified
the implementation of Obsidian, and showed that the method works
well in practice (see reference ~\cite{BB} for more details of the method).

While other approaches to GPU programming in higher level languages tend to deliberately abstract away from the details of the GPU, we persist in our aim of exposing architectural details of
the machine and giving the programmer fine control.
This is partly because trying to provide simple
but effective programming idioms is an interesting challenge.
More importantly, though, we are fascinated
by the problem of how to assist
programmers in making the subtle algorithmic decisions needed
to program
parallel machines with strange, programmer controlled
 memory hierarchies, and strange
constraints on memory access patterns. This problem is by no means
confined to GPUs, and it is both difficult and pressing.
We hope that Obsidian can form the basis for work in this area.

%We hinted at our simple and direct method of monad reification. This is described 
%in more detail in reference ~\cite{BB}. 
%In Obsidian, we do not currently make 
%explicit use of the compositional nature of this method.

% ---------------------------------------------------------------------------
% CONCLUSION 
% --------------------------------------------------------------------------- 
%\section{Conclusion}

% ---------------------------------------------------------------------------
% APPENDICES
% --------------------------------------------------------------------------- 
%\appendix
%\section{CUDA code}
%\label{app:A}

%This is the text of the appendix, if you need one.

% ---------------------------------------------------------------------------
% ACKNOWLEDGEMENTS
% --------------------------------------------------------------------------- 
\acks
The work on monad reification presented in this paper is joint 
work between Josef Svenningsson and Joel Svensson.

Push arrays were invented by Koen Claessen. The implementation 
of push arrays in Obsidian is targeted at 
GPUs and restricted compared to Koen's more general idea. Koen has 
also been a source of important insights and tips that have
improved this work greatly.

This research has been funded by the Swedish Foundation for
Strategic Research (which funds the Resource Aware Functional 
Programming (RAW FP) Project) and by the
Swedish Research Council.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}
\bibliography{paper}
% The bibliography should be embedded for final submission.

%\begin{thebibliography}{}
%\softraggedright

%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...

%\end{thebibliography}

\end{document}

### 2013-10-02 src2/hl/paperThesis.tex
apa

--- 2013-10-02 src2/hl/paper-restruct.tex
@@
 
-\documentclass[preprint]{sigplanconf}
+%\documentclass[preprint]{sigplanconf}
 
@@
 
-\usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{placeins}
-\usepackage{tikz}
-\usepackage{pgfplots}
-\usetikzlibrary{shapes,snakes,positioning}
-\usepackage{fancyvrb}
-\usepackage[T1]{fontenc}
+%\usepackage{amsmath}
+%\usepackage{graphicx}
+%\usepackage{placeins}
+%\usepackage{tikz}
+%\usepackage{pgfplots}
+%\usetikzlibrary{shapes,snakes,positioning}
+%\usepackage{fancyvrb}
+%\usepackage[T1]{fontenc}
 %\newcounter{arraycard}
 
-\begin{document}
-
-\conferenceinfo{WXYZ '05}{date, City.} 
-\copyrightyear{2005} 
-\copyrightdata{[to be supplied]}
-
-\titlebanner{DRAFT}        % These are ignored unless
-\preprintfooter{Obsidian: GPU kernel generation}   % 'preprint' option specified.
+%\begin{document}
+
+%\conferenceinfo{WXYZ '05}{date, City.} 
+%\copyrightyear{2005} 
+%\copyrightdata{[to be supplied]}
+
+%\titlebanner{DRAFT}        % These are ignored unless
+%\preprintfooter{Obsidian: GPU kernel generation}   % 'preprint' option specified.
 
@@
 %\title{An Embedded Language for Resource Aware \\ GPU Kernel Programming}  
-\title{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
+%\title{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
 
@@
 
-\authorinfo{Bo Joel Svensson \and Mary Sheeran} 
-           {Chalmers University of Technology}
-           {\{joels,ms\} at chalmers dot se}
-
-\maketitle
+%\authorinfo{Bo Joel Svensson \and Mary Sheeran} 
+%           {Chalmers University of Technology}
+%           {\{joels,ms\} at chalmers dot se}
+
+\newcommand{\subsubsubsection}[1]{\paragraph{#1:}}
+%% \vspace{1ex}
+
+%% \noindent\emph{\bf#1}\newline
+
+%% \vspace{1ex}
+%% \noindent
+%% }
+
+%\maketitle
 % ---------------------------------------------------------------------------
@@
 % --------------------------------------------------------------------------- 
-\begin{abstract}
+\subsection*{abstract}
+%\begin{abstract}
 
@@
 
-\end{abstract}
+%\end{abstract}
 
 %\category{CR-number}{subcategory}{third-level}
-\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
-\category{D.3.4}{Programming Languages}{Processors}[Code generation]
+%\category{D.3.2}{Programming Languages}{Language Classifications}[Applicative (functional) languages; Concurrent, distributed, and parallel languages]
+%\category{D.3.4}{Programming Languages}{Processors}[Code generation]
 
@@
 %term1, term2
-\terms
-Languages, Performance
+%\terms
+%Languages, Performance
 
@@
 %keyword1, keyword2
-\keywords
-Data parallelism, array programming, embedded language
+%\keywords
+%Data parallelism, array programming, embedded language
 
@@
 % --------------------------------------------------------------------------- 
-\section{Introduction}
+\subsection{Introduction}
 
@@
 % --------------------------------------------------------------------------- 
-\subsection{Motivation and contributions}
+\subsubsection{Motivation and contributions}
 
@@
 % --------------------------------------------------------------------------- 
-\subsection{The GPU and CUDA}
+\subsubsection{The GPU and CUDA}
 \label{sec:GPU}
@@
 % --------------------------------------------------------------------------- 
-\section{A taste of Obsidian programming}
+\subsection{A taste of Obsidian programming}
 
@@
 {\em Push} arrays. A pull array naturally represents gather operations,
-while a push array captures scattering operations. Reference \cite{Obsidian} 
+while a push array captures scattering operations. Reference \cite{Obsidian-Expressive} 
 describes the introduction of push arrays into Obsidian.
@@
 % --------------------------------------------------------------------------- 
-\section{Obsidian internals}
+\subsection{Obsidian internals}
 
@@
 % --------------------------------------------------------------------------- 
-\subsection{Scalars}
+\subsubsection{Scalars}
 
@@
 % --------------------------------------------------------------------------- 
-\subsection{Pull arrays}
+\subsubsection{Pull arrays}
 
@@
 in many other embedded languages. The 
-embedded language Pan ~\cite{ELLIJFP} used a similar representation for 
+embedded language Pan ~\cite{COMPILEEDSL} used a similar representation for 
 images and it has been used since in Obsidian, Feldspar ~\cite{FELDSPAR} 
@@
 % --------------------------------------------------------------------------- 
-\subsection{Thread, Block and Grid Computations} 
+\subsubsection{Thread, Block and Grid Computations} 
 \label{sec:Program}
@@
 This data type also represents the lowest level programming interface 
-exposed by Obsidian. In reference ~\cite{CSORT} we make use of these low-level 
+exposed by Obsidian. In reference ~\cite{JOSEFJOELCSORT} we make use of these low-level 
 capabilities to implement histograms using atomic operations (not shown here).
@@
 % --------------------------------------------------------------------------- 
-\subsection{Push arrays} 
+\subsubsection{Push arrays} 
 \label{sec:PushArrays}
@@
 % --------------------------------------------------------------------------- 
-\subsection{Push and pull array interplay} 
+\subsubsection{Push and pull array interplay} 
 \label{sec:interplay}
@@
 % --------------------------------------------------------------------------- 
-\subsection{Compilation to CUDA}
+\subsubsection{Compilation to CUDA}
 \label{sec:Compile}
@@
 % --------------------------------------------------------------------------- 
-\section{Case studies} 
+\subsection{Case studies} 
 \label{sec:CASESTUDIES}
@@
 %% ------------------------------------------------------------------------
-\subsection{Reduction}  
+\subsubsection{Reduction}  
 \label{sec:redOpt}
 
-\input{reductionfigs.tex}
+\input{./hl/reductionfigs.tex}
 
@@
 optimisations that are applied to the kernels are influenced by a 
-tutorial presentation from NVIDIA~\cite{RedOpt}.
+tutorial presentation from NVIDIA~\cite{reduction}.
 
@@
 
-\subsubsection{Reduction 1}
+\subsubsubsection{Reduction 1}
 
@@
 
-\subsubsection{Reduction 2}
+\subsubsubsection{Reduction 2}
 
@@
 
-\subsubsection{Reduction 3}
+\subsubsubsection{Reduction 3}
 
@@
 
-\subsubsection{Reduction 4}
+\subsubsubsection{Reduction 4}
 
@@
 
-\subsubsection{Reduction 5}
+\subsubsubsection{Reduction 5}
 
@@
 
-\subsubsection{Reductions 6 and 7}
+\subsubsubsection{Reductions 6 and 7}
 
@@
 
-\subsubsection{Larger local reductions} 
+\subsubsubsection{Larger local reductions} 
 Figure \ref{fig:reducetablel} shows running time results for kernels that are 
@@
 
-\subsection{Scan}
+\subsubsection{Scan}
 
@@
 
-\input{scanfigs.tex}
+\input{./hl/scanfigs.tex}
 
@@
 
-\subsubsection{two elements per thread}
+\subsubsubsection{Two elements per thread}
 
@@
 Note the use of two {\em write functions} in sequence.
-Similar patterns were used in our implementations of sorting networks~\cite{Obsidian}, for similar reasons.
+Similar patterns were used in our implementations of sorting networks~\cite{Obsidian-Expressive}, for similar reasons.
 
@@
 
-\section{Combining kernels to solve large problems} 
+\subsection{Combining kernels to solve large problems} 
 \label{sec:Benchmarks}
@@
 
-\subsection{Large reductions}
+\subsubsection{Large reductions}
 
@@
 % --------------------------------------------------------------------------- 
-\subsection{Related work}
+\subsubsection{Related work}
 
@@
 other end of the high-level - low-level spectrum, we have the NVIDIA
-Thrust Library~\cite{Thrust}.
+Thrust Library~\cite{THRUST}.
 Thrust offers a programming model where details of GPU architecture are 
@@
 Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
-~\cite{accelerate}. The abstraction level is comparable to that of Thrust. In 
+~\cite{ACCELERATEDAMP11}. The abstraction level is comparable to that of Thrust. In 
 other words, Accelerate hides most GPU details from the programmer. Accelerate 
@@
 
-The Copperhead \cite{Copperhead} system compiles a subset of Python to run 
+The Copperhead \cite{copperhead} system compiles a subset of Python to run 
 on GPUs. Much like other languages mentioned here, Copperhead identifies 
@@
 to monad reification. Related to this work is the approach based on a 
-continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{NEIL} is also related. 
+continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{sculthorpe2013constrained} is also related. 
 Our approach to monad reification is simpler but less general.
@@
 % --------------------------------------------------------------------------- 
-\section{Future Work} 
+\subsection{Future Work} 
 \label{sec:Future}
@@
 
-\subsection{Increase low level control}
+\subsubsection{Increase low level control}
 
@@
 
-\subsection{To automate or not to automate} 
+\subsubsection{To automate or not to automate} 
 Obsidian makes use of types that model the GPU hierarchy. This limits 
@@
 
-\subsection{To optimise or not to optimise} 
+\subsubsection{To optimise or not to optimise} 
 The version of Obsidian described here does not try to use any compiler optimisations techniques. Instead, we are expecting that the 
@@
 
-\subsection{Kernel Coordination}
+\subsubsection{Kernel Coordination}
 
@@
 % --------------------------------------------------------------------------- 
-\section{Conclusion}
+\subsection{Conclusion}
 
@@
 illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
-used in the NVIDIA tutorial ~\cite{RedOpt}. The case study also show hows we 
+used in the NVIDIA tutorial ~\cite{reduction}. The case study also show hows we 
 can compose kernels and thus reuse prior effort. 
@@
 % --------------------------------------------------------------------------- 
-\acks
+\subsection*{Acknowledgments}
 The work on monad reification presented in this paper is joint 
@@
 
-\bibliographystyle{abbrvnat}
-\bibliography{paper}
+%\bibliographystyle{abbrvnat}
+%\bibliography{paper}
 % The bibliography should be embedded for final submission.
@@
 
-\end{document}
+%\end{document}

### 2013-10-02 src2/ifl/GPUs.tex
apa

@@
-\section{Graphics Processing Units, accessible high performance parallel computing}
+\subsection{Graphics Processing Units, accessible high performance parallel computing}
 \label{sec:gpu}
@@
 future in which each individual core is considerably more powerful than in
-today's GPUs~\cite{IntelLarrabee}.
+today's GPUs~\cite{larrabee}.
 

### 2013-10-02 src2/ifl/combinators.tex
apa

@@
-\section{Connection patterns for hardware design and parallel programming}\label{sec:combinators}
+\subsection{Connection patterns for hardware design and parallel programming}\label{sec:combinators}
 
@@
 For example, an important pattern is parallel prefix or scan.
-Given inputs $[x_0, x_1 \ldots x_{n-1}]$, the prefix problem is to compute each $x_0 \circ x_1 \circ \ldots \circ x_j$ for $0 \leq j < n$, for $\circ$ an associative, but not necessarily commutative, operator.
+Given inputs\newline $[x_0, x_1 \ldots x_{n-1}]$, the prefix problem is to compute each $x_0 \circ x_1 \circ \ldots \circ x_j$ for $0 \leq j < n$, for $\circ$ an associative, but not necessarily commutative, operator.
 For example, the prefix sum of {\tt [1..10]} is {\tt [1,3,6,10,15,21,28,36,45,55]}.
@@
 
-To express the construction in Lava, we make use of two connection patterns.
+To express the construction in Lava, we make use of two connection patterns.\newline
 {\tt two :: ([a] -> [b]) -> [a] -> [b]} applies its component to
@@
 that result from these applications.
-Thus, {\tt two (sklansky plus)} applied to {\tt [1..10]} gives {\tt [1,3,6,10,15,6,13,21,30,40]}.
+Thus, {\tt two (sklansky plus)} applied to {\tt [1..10]} gives {\tt [1,3,6,} {\tt10,15,6,13,21,30,40]}.
 Left-to-right serial composition has type

### 2013-10-02 src2/ifl/obsidian.tex
apa

@@
 % ------------------------------------------------------------------------------
-\section{Obsidian: a domain specific embedded language for GPU programming}
+\subsection{Obsidian: a domain specific embedded language for GPU programming}
 \label{sec:obsidian}
@@
 scan_add_kernel2 :: GArr IntE -> W (GArr IntE) 
-scan_add_kernel2 = cache ->- sync ->- sklansky (+) ->- wb ->- 
-                     sync
+scan_add_kernel2 = cache ->- sync ->- sklansky (+) ->- 
+                   wb ->- sync
 \end{code}
@@
 \begin{code}
-execute :: ExecMode -> (GArr (Exp a) -> W (GArr (Exp b)) -> 
-              [Exp a] -> IO [Exp b]
+execute :: ExecMode 
+         -> (GArr (Exp a) -> W (GArr (Exp b)) 
+         -> [Exp a] -> IO [Exp b]
 \end{code}
@@
 % ------------------------------------------------------------------------------
-\subsection{Implementation}
+\subsubsection{Implementation}
 
@@
 This follows the approach from Compiling Embedded Languages 
-\cite{Elliott03:CompileDSEL-JFP}. However, 
+\cite{COMPILEEDSL}. However, 
 to obtain a typed environment in which to operate, phantom types are used.
@@
 \begin {code}
-(\ix -> (E (BinOp Add (Index (Variable "input") ix) (LitInt 1)))) 
+(\ix -> (E (BinOp Add (Index (Variable "input") ix) 
+                      (LitInt 1)))) 
 \end{code}
@@
 representing a thread Id. The result is an expression looking as follows:
-\begin {code}
-E (BinOp Add (Index (Variable "input") (Variable "tid")) (LitInt 1))
+\begin{code}
+E (BinOp Add (Index (Variable "input") (Variable "tid")) 
+             (LitInt 1))
 \end{code}

### 2013-10-02 src2/hl/paperThesis.tex
apa

@@
 actual order of the kernels we wish to launch. We make no changes to the 
-kernels themselves. The listing below shows an outline
+kernels themselves. The listing in figure~\ref{fig:cudareductioncode} shows an outline
 of the CUDA code for launching {\tt red5l} to give $2^{24}$ (= 16777216)
@@
 %% \pagebreak
-
+\begin{figure} 
 \begin{small}
@@
 \end{small}
+\caption{CUDA code that combines reduction kernels into reduction of a large array.}
+\label{fig:cudareductioncode}
+\end{figure}
 

### 2013-10-03 src2/thesis.tex
apa
update

@@
 \usepackage{tikz,pgflibraryshapes}
+\usetikzlibrary{shapes,snakes,positioning}
+\usepackage{pgfplots}
 \usepackage{booktabs}
@@
 
-\newcommand{\thesistitle}{Not decided upon}
+\newcommand{\thesistitle}{Embedded Languages for Data Parallel Computation}
 \newcommand{\dept}{Department of Computer Science and Engineering}
@@
 
+% ---------------------------------------------------------------------------
+% Papers 
+% ---------------------------------------------------------------------------
+\newcommand{\paperBB}{Simple and Compositional Reification of Monadic Embedded Languages}
+\newcommand{\paperIFL}{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
+\newcommand{\paperPAPP}{GPGPU Kernel Implementation and Refinement using Obsidian} 
+\newcommand{\paperEXPRESSIVE}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+\newcommand{\paperCSORT}{Counting and Occurrence Sort for GPUs using an Embedded Language}
+\newcommand{\paperHL}{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
+\newcommand{\paperARBB}{Programming Future Parallel Architectures with Haskell and Intel ArBB}
+\newcommand{\paperEMBARBB}{Parallel Programming in Haskell Almost for Free}
+
+% ---------------------------------------------------------------------------
+% BEGIN ! 
+% ---------------------------------------------------------------------------
 \begin{document}
@@
 
-\noindent\copyright {\sc {Bo Joel Svensson}}, 2013 \\*[5mm]
+\noindent\copyright {\sc {Bo Joel Svensson}}, 2013 \\ 
 \vspace{1cm}
 
-\noindent Technical Report ..D\\
+\noindent Technical Report 101D\\
 \noindent ISSN                 \\
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 
@@
 \cleardoublepage
+\clearpage
 
@@
 \fancyhead[LE,RO]{\thepage}
+
+\chapter{About this thesis}
+
+This thesis contains an introduction and a selection of papers. The introduction 
+gives background and context to the papers.
+
+\section{Papers}
+
+The papers contained in this thesis falls in three different categories. 
+\begin{itemize} 
+\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
+\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
+\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
+\end{itemize}
+
+\subsection{EDSL implementation papers}
+
+\subsubsection{\paperBB}
+
+apa apa apa bepa cepa apa
+
+\subsection{GPU Programming papers}
+
+\subsubsection{\paperIFL}
+
+\subsubsection{\paperPAPP}
+
+\subsubsection{\paperEXPRESSIVE}
+
+\subsubsection{\paperCSORT}
+
+\subsubsection{\paperHL}
+
+\subsubsection{Retargetable parallel programming}
+
+\subsubsection{\paperARBB}
+
+\subsubsection{\paperEMBARBB}
 
@@
 
+\chapter{EDSL implementation papers}
+\label{chap:EDSLImplementation}
+% ---------------------------------------------------------------------------
+% 
+% ---------------------------------------------------------------------------
+\cleardoublepage
+
+\section{\paperBB}
+\label{sec:paperBB}
+%\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
+
+\begin{center} 
+Josef Svenningsson, Bo Joel Svensson
+\end{center}
+
+\input{./bb/paperThesis}
+
 \chapter{GPU programming papers}
+\label{chap:GPUProgramming}
 
@@
 
-\section{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
+\section{\paperIFL}
+\label{sec:paperIFL}
 %\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
@@
 
-\section{Obsidian: GPU Computing Using Haskell}
+\section{\paperPAPP}
+\label{sec:paperPAPP}
 %\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}
@@
 
-\section{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+\section{\paperEXPRESSIVE}
+\label{sec:paperEXPRESSIVE}
 %\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
@@
 
-\section{Counting and Occurrence Sort for GPUs using an Embedded Language}
+\section{\paperCSORT}
+\label{sec:paperCSORT}
 %\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}
@@
 
-\section{Simple and Compositional Reification of Monadic Embedded Languages}
+\section{\paperHL}
+\label{sec:paperHL}
 %\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
@@
 \begin{center} 
-Josef Svenningsson, Bo Joel Svensson
+Bo Joel Svensson, Mary Sheeran
 \end{center}
 
-\input{./bb/paperThesis}
+\input{./hl/paperThesis}
 
@@
 \chapter{Retargetable parallel programming papers}
+\label{chap:ArBB}
 % ---------------------------------------------------------------------------
@@
 
-\section{Programming Future Parallel Architectures with Haskell and Intel ArBB}
+\section{\paperARBB}
+\label{sec:paperARBB}
 %\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
@@
 
-\section{Parallel Programming in Haskell Almost for Free}
+\section{\paperEMBARBB}
+\label{sec:paperEMBARBB}
 %\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}

### 2013-10-06 src2/arbb/paperThesis.tex
hacking

@@
 years we have seen difficult-to-program chips suffer 
-% \cite{cell} 
+% \citearbb{cell} 
 (e.g. Cell)
@@
 GPU vendors strive to enable more traditional programming features
-\cite{fermi} (e.g. C++).  There is an increasing
+\citearbb{fermi} (e.g. C++).  There is an increasing
 tension between ease of programming and efficiency.
@@
 small embedded devices are most power efficient if their processors and operating systems
-omit programming features such as virtual memory and threads \cite{tinyos}.  
+omit programming features such as virtual memory and threads \citearbb{tinyos}.  
 At the other end of the power spectrum, GPU's
@@
  new EDSL implementation, called \accarbb{}, that combines
-existing systems, Accelerate \cite{ACCELERATEDAMP11} and ArBB \cite{ArBB}, to produce a unified
+existing systems, Accelerate \citearbb{ACCELERATEDAMP11} and ArBB \citearbb{ArBB}, to produce a unified
 high-level programming environment equally suited to multicore,
-vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \cite{larrabee}).
+vectorized CPUs, as to GPUs and other accelerators (such as Intel MIC chips \citearbb{larrabee}).
 %
 \if{0}
-ArBB is based on the RapidMind \cite{RapidMind} system which was an embedded 
+ArBB is based on the RapidMind \citearbb{RapidMind} system which was an embedded 
 DSL targeting multicore CPUs as well as GPUs. ArBB does not \new{yet} include a GPU 
@@
 architectures.  
-StreamIt\cite{streamit} is a good example.
-% (for example, StreamIt\cite{streamit})
-% The high-water mark in this area is perhaps  StreamIt\cite{streamit}.
+StreamIt\citearbb{streamit} is a good example.
+% (for example, StreamIt\citearbb{streamit})
+% The high-water mark in this area is perhaps  StreamIt\citearbb{streamit}.
 %
 
-% It has been noted \cite{not-sure} that 
-% Yet the few DSLs that gain popularity lose simplicity\cite{not-sure}.  
-% Yet popular DSLs rarely remain simple \cite{not-sure}.  
+% It has been noted \citearbb{not-sure} that 
+% Yet the few DSLs that gain popularity lose simplicity\citearbb{not-sure}.  
+% Yet popular DSLs rarely remain simple \citearbb{not-sure}.  
 DSLs may start out simple and focused, but if they gain popularity
@@
 differs from the more common usage of preprocessors and 
-% Lisp/Scheme macros \cite{scheme-macros}
+% Lisp/Scheme macros \citearbb{scheme-macros}
 macros, which typically add extra
@@
 performance-oriented EDSLs are often simple, first-order
-languages without pointers \cite{wavescript, ACCELERATEDAMP11}.
+languages without pointers \citearbb{wavescript, ACCELERATEDAMP11}.
 
@@
 As a research area, EDSLs and two-stage DSLs have been actively pursued for at least a
-decade \cite{VERTIGO, wavescript} but are gaining steam
+decade \citearbb{VERTIGO, wavescript} but are gaining steam
 recently 
-%\cite{sejits, stanford-ppl} 
-\cite{copperhead, stanford-ppl} 
+%\citearbb{sejits, stanford-ppl} 
+\citearbb{copperhead, stanford-ppl} 
 and are beginning to appear in
-commercial products \cite{ArBB}.
+commercial products \citearbb{ArBB}.
 %
@@
 programming solutions for domains such as machine learning and
-rendering \cite{stanford-ppl}.
+rendering \citearbb{stanford-ppl}.
   Moreover, EDSLs need not be hosted by esoteric research languages---Intel's
  ArBB embeds an array language in C++ and Berkeley's
-Copperhead \cite{copperhead} generates CUDA from simple Python code.
+Copperhead \citearbb{copperhead} generates CUDA from simple Python code.
 
@@
 %% {\em Accelerate} (Data.Array.Accelerate) is an existing EDSL for 
-%% high-level data-parallel programming in Haskell~\cite{Accelerate}. 
+%% high-level data-parallel programming in Haskell~\citearbb{Accelerate}. 
 The Accelerate programming model consists of collective 
@@
 function. 
-To [partially] automate the creation of these bindings we used the C2HS system \cite{C2HS}. The 
+To [partially] automate the creation of these bindings we used the C2HS system \citearbb{C2HS}. The 
 \arbbvmH{} bindings are very low-level.
@@
   operations over successively halving array sizes.  Essentially, cut
-  the array in half, combine the halves, repeat. In ~\cite{reduction}
+  the array in half, combine the halves, repeat. In ~\citearbb{reduction}
   this approach is explained in the context of CUDA. 
@@
 Black-Scholes option pricing is a finance-related benchmark that has been used in similar
-DSLs targeting GPUs \cite{ACCELERATEDAMP11, NIKOLA}. Since we are re-using the 
+DSLs targeting GPUs \citearbb{ACCELERATEDAMP11, NIKOLA}. Since we are re-using the 
 Accelerate front-end, we can directly use the Black-Scholes benchmark that 
@@
 incurring extra data-movement and possibly extra JIT compilation
-\cite{wheres-the-data-paper}.  Specifically, we see that the
+\citearbb{wheres-the-data-paper}.  Specifically, we see that the
 Accelerate Black-Scholes program from Figure~\ref{fig:blackscholes}
@@
 
-OpenCL~\cite{opencl08} is a programming model very similar to CUDA but with the aspiration 
+OpenCL~\citearbb{opencl08} is a programming model very similar to CUDA but with the aspiration 
 to offer both acceleration of computations on GPUs or to multicore CPUs. 
@@
 
-Microsoft Accelerator~\cite{ACCELERATOR} is an embedded language with similar 
+Microsoft Accelerator~\citearbb{ACCELERATOR} is an embedded language with similar 
 aspirations as ArBB, that is, to target a diverse range of architectures using the 
@@
 %Most work on partitioning workloads between 
-Many CPU and GPU comparisons, and some CPU/GPU workload partitioners \cite{merge}, rely on
+Many CPU and GPU comparisons, and some CPU/GPU workload partitioners \citearbb{merge}, rely on
 redundant hand-written versions of all kernels
-% \cite{cpu-gpu-partitioning} 
-(though some systems like Qilin \cite{qilin} allow a single source code).
+% \citearbb{cpu-gpu-partitioning} 
+(though some systems like Qilin \citearbb{qilin} allow a single source code).
 It is difficult in this kind of scenario
@@
 For example, comparing unoptimized serial CPU implementations vs. GPU
-ones is not informative \cite{debunking-dubey}.
+ones is not informative \citearbb{debunking-dubey}.
 In \systemname{} controlling for effort
@@
 Section~\ref{sec:accelerate-arbb} is one approach to this problem, as
-are the optimizations studied in the Copperhead \cite{copperhead} and
-Obsidian \cite{JSLIC} projects.
+are the optimizations studied in the Copperhead \citearbb{copperhead} and
+Obsidian \citearbb{JSLIC} projects.
 %
@@
 %\small
-% \bibliographystyle{plainnat}
-%\bibliographystyle{IEEEtran}
-%\bibliography{IEEEabrv,haskell_arbb}
+\bibliographystylearbb{alpha}
+\bibliographyarbb{thesis}
 %}

### 2013-10-06 src2/bb/body.tex
hacking

@@
 interest in functional programming using Haskell and has read ``The Haskell 
-school of expression'' \cite{HUDAK}. He gets the idea to implement 
+school of expression'' \citebb{HUDAK}. He gets the idea to implement 
 a language for robot control embedded in Haskell. \studname{} realises that the 
@@
 %an embedded domain specific language (EDSL) for robot control. Inspired 
-%by the book ``the Haskell school of expression'' \cite{HUDAK} but instead of 
+%by the book ``the Haskell school of expression'' \citebb{HUDAK} but instead of 
 %evaluating his robot EDSL, \studname{} wants to compile the language to a form 
@@
 %% \noindent For hints towards a solution to this I suggest that you follow 
-%% this \underline{link} \cite{GillBlog}. Another idea is that you refer 
+%% this \underline{link} \citebb{GillBlog}. Another idea is that you refer 
 %% to the article ``Generic monadic constructs for embedded languages'' 
-%% \cite{Generic}.
+%% \citebb{Generic}.
 %% \newline \newline
@@
 it works. For example, in this paper \direct{\docname{} shows
-  \studname{} the paper \cite{Generic}}, the authors use a
+  \studname{} the paper \citebb{Generic}}, the authors use a
 continuation monad to be able to reify the monadic constructs of their
@@
 described in a separate paper \direct{\docname{} pulls out the paper
-  \cite{axelsson2012generic} from a pile of papers}. It's hard to do
+  \citebb{axelsson2012generic} from a pile of papers}. It's hard to do
 an apples to apples comparison between Syntactic and your
@@
 \speak{\docname{}} There are also the papers
-\cite{Farmer,sculthorpe2013constrained} which solves
+\citebb{Farmer,sculthorpe2013constrained} which solves
 the problem in a manner that is more similar to yours. They also have explicit
@@
 it could be combined with other data types using techniques like Data
-Types \'{a} la Carte \cite{Wouter} or CompData \cite{CompData}. This
+Types \'{a} la Carte \citebb{Wouter} or CompData \citebb{CompData}. This
 way, a language can be designed piece by piece.You can then select the
@@
 solution is very similar to how variable binding is handled in the
-Syntactic library \cite{syntactic}.
+Syntactic library \citebb{syntactic}.
 \newline \newline
@@
 %A consequence of the compositionality is that we can use techniques
-%such as CompData \cite{} to decompose the definition of DSLs into
+%such as CompData \citebb{} to decompose the definition of DSLs into
 %smaller reusable building blocks. Figure \ref{fig:compdata} shows a
@@
 currently used in the implementation of Obsidian, an EDSL for general
-purpose GPU programming \cite{Obsidian-Expressive}. Obsidian has language
+purpose GPU programming \citebb{Obsidian-Expressive}. Obsidian has language
 constructs which are considerably more advanced than the robot
@@
 %can turn, move forward and has a sensor to detect whether it is facing
-%an obstacle or not. Inspired by the robot language in \cite{} you come
+%an obstacle or not. Inspired by the robot language in \citebb{} you come
 %up with a set of primitives shown in figure

### 2013-10-06 src2/csort/Obsidian.tex
hacking

@@
 between low-level control and useful high-level abstractions. Over time, this 
-has led to a number of different versions of Obsidian. In reference \cite{JSLIC}, 
+has led to a number of different versions of Obsidian. In reference \citecsort{JSLIC}, 
 two versions of Obsidian differing by having a {\em monadic} or a limited {\em arrow} 
@@
 The Obsidian implementation of pull arrays was influenced by Pan's representation 
-of images as functions from coordinates to colour values \cite{PAN}. A pull array 
+of images as functions from coordinates to colour values \citecsort{PAN}. A pull array 
 is a function from index to value and an associated length.
 
-Push arrays were added to Obsidian first in reference \cite{Obsidian-Expressive}. A push array
+Push arrays were added to Obsidian first in reference \citecsort{Obsidian-Expressive}. A push array
 specifies where elements are to end up in a result array, rather than where 
@@
 Push arrays have also been adopted by others and are used in the implementation 
-of filters in reference~\cite{FPCDSL} and in the new Nikola\footnote{github.com/mainland/nikola/blob/master/src/Data/Array/Nikola/Repr/Push.hs}.
+of filters in reference~\citecsort{FPCDSL} and in the new Nikola\footnote{github.com/mainland/nikola/blob/master/src/Data/Array/Nikola/Repr/Push.hs}.
 
@@
 %% as the data scattering operations used later. For more information about 
-%% push arrays, see reference \cite{PUSH}.
+%% push arrays, see reference \citecsort{PUSH}.
 
@@
 %is the addition of a new kind of array representation that we call 
-%{\em push} arrays \cite{PUSH}. Previous versions of Obsidian 
-%(described in \cite{JSLIC}) represented arrays as functions 
+%{\em push} arrays \citecsort{PUSH}. Previous versions of Obsidian 
+%(described in \citecsort{JSLIC}) represented arrays as functions 
 %from indices to values,we call these {\em pull} arrays. Another addition 
@@
 The resulting prefix sum is implemented as shown in reference 
-\cite{ScanCUDA}, see figure \ref{fig:prefixsum}. There are three steps 
+\citecsort{ScanCUDA}, see figure \ref{fig:prefixsum}. There are three steps 
 involved. First local prefix sums are computed; the rightmost elements elements are also stored in a separate, auxiliary, array. Following this, 
@@
 of an array is needed. The prefix sum algorithm that we implement here is 
-based on divide and conquer and originates from Sklansky \cite{Sklansky}. 
+based on divide and conquer and originates from Sklansky \citecsort{Sklansky}. 
 % A kernel that performs the distribution (part C in the figure) is also needed.
@@
 %a separate array. This separate array can be used to compute values to adjust 
-%the local sums by. This follows the schema developed in \cite{LargeScan}.
+%the local sums by. This follows the schema developed in \citecsort{LargeScan}.
 

### 2013-10-06 src2/csort/discussion.tex
apa
hacking

@@
 If the desired effect is to sort and remove duplicates 
-(for an example of this see \cite{REMOVEDUPS}) then 
+(for an example of this see \citecsort{REMOVEDUPS}) then 
 there is a clear benefit to using occurrence sort
@@
 In this paper, we are entirely focused on the implementation and
-performance of kernels. Reference \cite{CSORT} goes into the details
+performance of kernels. Reference \citecsort{CSORT} goes into the details
 of performing data transfers in parallel with ongoing computations. We
@@
 also the reason why we haven't performed any benchmarks against the algorithm in
-\cite{CSORT}, because it would compare very different things. As a
+\citecsort{CSORT}, because it would compare very different things. As a
 piece of future work, it would be interesting to add the capability of
@@
 is suitable only up to some point. A related algorithm, known as radix 
-sort \cite{Knuth}, addresses this issue by using multiple counting sort 
+sort \citecsort{Knuth}, addresses this issue by using multiple counting sort 
 like passes, one for each digit position and thus limits this auxiliary 
@@
 %% generating back end of such an EDSL. This EDSL is called PEGGY and 
-%% is described in \cite{cole2012beauty}.
+%% is described in \citecsort{cole2012beauty}.
 

### 2013-10-06 src2/csort/intro.tex
hacking

--- 2013-10-01 src2/csort/intro.tex
@@
 
-These algorithms are explored in Obsidian \cite{JSLIC,PUSH}, a domain
+These algorithms are explored in Obsidian \citecsort{JSLIC,Obsidian-Expressive}, a domain
 specific language targeting GPU programming. The goal of Obsidian is
@@
   outperforming the sorting implementation in the library
-  Thrust\cite{THRUST} in many cases.
+  Thrust\citecsort{THRUST} in many cases.
 \item Occurrence sort is shown to be particularly
@@
 
-Sorting has applications in the computer graphics field \cite{sintorn}. 
+Sorting has applications in the computer graphics field \citecsort{SINTORNSORT}. 
 Example instances of sorting and duplicate element removal in computer graphics 
-can be identified in references \cite{Olsson} and \cite{Karras}. Another example 
+can be identified in references \citecsort{Olsson} and \citecsort{Karras}. Another example 
 of where sorting and the removal of duplicates have applications is 
-in databases \cite{REMOVEDUPS}.
+in databases \citecsort{REMOVEDUPS}.
 
 %% Sorting on the GPU is interesting both for computer graphics
-%% applications \cite{SINTORN} and for using the GPU to offload the
-%% processor, for example in database applications \cite{REMOVEDUPS}.
+%% applications \citecsort{SINTORN} and for using the GPU to offload the
+%% processor, for example in database applications \citecsort{REMOVEDUPS}.
 
-The paper \cite{CSORT} implements counting sort in CUDA, and also optimisations 
+The paper \citecsort{CSORT} implements counting sort in CUDA, and also optimisations 
 that overlap computations with transferring data to and from the GPU.
@@
 programming. Other such approaches include Accelerate
-\cite{Acc} and Nikola \cite{NIKOLA}. Accelerate and Nikola both take an even
+\citecsort{ACCELERATEDAMP11} and Nikola \citecsort{NIKOLA}. Accelerate and Nikola both take an even
 higher level approach compared to Obsidian and abstract more from GPU
 details. Another example of providing a high-level interface to
-programming the GPU is the C++ library Thrust \cite{THRUST}.
-%\cite{FELDSPAR}
+programming the GPU is the C++ library Thrust \citecsort{THRUST}.
+%\citecsort{FELDSPAR}
 
@@
 %% compares the performance of the counting sort algorithm we implement here to 
-%% the NVIDIA Thrust \cite{THRUST} library. The Obsidian generated kernels perform
+%% the NVIDIA Thrust \citecsort{THRUST} library. The Obsidian generated kernels perform
 %% very well at relatively light work effort.
@@
 
-%\cite{CSORT} 
-%\cite{FELDSPAR}
+%\citecsort{CSORT} 
+%\citecsort{FELDSPAR}

### 2013-10-06 src2/embarbb/implementation.tex
hacking

--- 2013-09-30 src2/embarbb/implementation.tex
@@
 The {\tt d} parameter is used for a type level representation of the 
-dimensionality of a vector, as in the Repa library~\cite{REPA}. The 
+dimensionality of a vector, as in the Repa library~\citeemb{REPA}. The 
 dimensionality is encoded using the following types together with 
@@
 {\tt map} operation specifies elementwise application of a function over 
-vectors, like NESL's {\em apply-to-each}~\cite{NESL}, so it corresponds 
+vectors, like NESL's {\em apply-to-each}~\citeemb{NESL}, so it corresponds 
 to Haskell's {\tt map} and {\tt zipWith}. Both {\tt Call} and {\tt Map} 
 take a {\tt(R GenRecord)}. This is inspired by the delayed expressions 
-that enable the implementation of {\tt vapply} in Nikola~\cite{NIKOLA}. 
+that enable the implementation of {\tt vapply} in Nikola~\citeemb{NIKOLA}. 
 The {\tt R} is the reification monad used to create DAGs (directed acyclic 
@@
 A typed interface to the language is supplied using the same phantom types 
-method as used in {\em pan}~\cite{ELLIJFP} and many other Haskell embedded 
+method as used in {\em pan}~\citeemb{COMPILEEDSL} and many other Haskell embedded 
 languages since then. The choice to use an untyped {\tt Expr} datatype was 
@@
 detection is performed, and a directed acyclic graph (DAG) is created. 
-The method of sharing detection used is based on the StableNames method~\cite{Gill}.
+The method of sharing detection used is based on the StableNames method~\citeemb{Gill}.
 

### 2013-10-06 src2/embarbb/introduction.tex
hacking

@@
 
-Intel's Array Building Blocks (ArBB) is one approach to the problem of how to increase productivity for programmers who need to exploit hardware parallelism, but for whom very low level parallel programming (as typically found in High Performance Computing) is too difficult or time consuming~\cite{ARBB2011}.
+Intel's Array Building Blocks (ArBB) is one approach to the problem of how to increase productivity for programmers who need to exploit hardware parallelism, but for whom very low level parallel programming (as typically found in High Performance Computing) is too difficult or time consuming~\citeemb{ARBB2011}.
 ArBB is a retargettable dynamic compilation framework that is provided as an embedded language in C++.
@@
 Many of the ideas in ArBB feel familiar to functional programmers.
-For example, when referring to collection data types, reference~\cite{ARBB2011} states that ``Expressions over collections always act as if a completely new collection were created and assignment between collections always behaves as if the entire input collection were copied.''
+For example, when referring to collection data types, reference~\citeemb{ARBB2011} states that ``Expressions over collections always act as if a completely new collection were created and assignment between collections always behaves as if the entire input collection were copied.''
 This rather functional view of data is accompanied by extensive optimisation that (as the paper states) removes almost all copying.
@@
 %The compilation 
-%process also tries to fuse away intermediate datastructures~\cite{ARBB2011}.
+%process also tries to fuse away intermediate datastructures~\citeemb{ARBB2011}.
 
@@
 heavy use of the C++ template machinery. There is also a lower level 
-interface to ArBB, the ArBB Virtual Machine C API~\cite{arbbvm}. The low level C API is not
+interface to ArBB, the ArBB Virtual Machine C API~\citeemb{arbbvm}. The low level C API is not
 intended for use directly by applications programmers, but rather as a base 
-for implementations of alternative ArBB frontends (embeddings)~\cite{ARBB2011}.
+for implementations of alternative ArBB frontends (embeddings)~\citeemb{ARBB2011}.
 The ArBB VM implements the compilation and parallelisation.
-Reference~\cite{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
-steps towards using ArBB as a backend to the embedded language Accelerate~\cite{ACCELERATEDAMP11}.
+Reference~\citeemb{joelryan} presents Haskell bindings to the ArBB C API as well as initial 
+steps towards using ArBB as a backend to the embedded language Accelerate~\citeemb{ACCELERATEDAMP11}.
 

### 2013-10-06 src2/embarbb/related.tex
apa
hacking

@@
 \subsubsection{Data.Array.Accelerate} 
-Accelerate~\cite{ACCELERATEDAMP11} is an embedded language for general purpose, 
+Accelerate~\citeemb{ACCELERATEDAMP11} is an embedded language for general purpose, 
 flat data parallel computations on GPUs.
@@
 into flat data parallelism. 
-A recent article about DPH is reference~\cite{DPH}.
+A recent article about DPH is reference~\citeemb{DPH}.
 
@@
 developed at Chalmers, ELTE University and Ericsson 
-(the telecom company)~\cite{FELDSPAR2010}. The functional {\tt while} loop 
+(the telecom company)~\citeemb{FELDSPAR2010}. The functional {\tt while} loop 
 of EmbArBB is inspired by the same concept in Feldspar.
@@
 %The most recent release of Feldspar uses the {\em Syntactic} library
-%for manipulating abstract syntax~\cite{SyntacticICFP12,FELDSYNTACTIC}. 
+%for manipulating abstract syntax~\citeemb{SyntacticICFP12,FELDSYNTACTIC}. 
 %We will consider switching to using the Syntactic library in the continued
@@
 \subsubsection{Nikola} 
-Nikola is an embedded language for GPU programming\cite{NIKOLA}, also
+Nikola is an embedded language for GPU programming\citeemb{NIKOLA}, also
 in Haskell.  
@@
 The embedding used in Nikola makes use of an untyped expression 
-data structure wrapped up in phantom types, in the style of Pan~\cite{ELLIJFP}.
+data structure wrapped up in phantom types, in the style of Pan~\citeemb{COMPILEEDSL}.
 The EmbArBB embedding works in the same way. 
@@
 \subsubsection{Repa} 
-Repa is a library for regular, shape polymorphic parallel arrays~\cite{REPA}. Repa 
+Repa is a library for regular, shape polymorphic parallel arrays~\citeemb{REPA}. Repa 
 uses a concept of {\tt delayed} arrays to obtain fusion of operations, such as the 
@@
 %\subsection{Obsidian}
-%Really ? \cite{Obsidian}
+%Really ? \citeemb{Obsidian}
 
@@
 We wish to explore ways to make use of the fact that we have an array programming library embedded in a sophisticated, strongly typed host language.
-In our earlier work on the embedded hardware description language Lava, we investigated various approaches to exploiting the host language during netlist synthesis~\cite{LavaMultipliers}, and we have also experimented with the use of search and dynamic programming in generating parallel prefix (or scan) networks~\cite{SheeranJFP}.
+In our earlier work on the embedded hardware description language Lava, we investigated various approaches to exploiting the host language during netlist synthesis~\citeemb{LavaMultipliers}, and we have also experimented with the use of search and dynamic programming in generating parallel prefix (or scan) networks~\citeemb{SheeranJFP}.
 We intend to apply similar methods to the development of data parallel programs once the EmbArBB implementation is more complete and stable.
 
-%It would be interesting to explore ways of using the host language in more sophisticated ways, rather as we have done for hardware generation in Lava~\cite{LavaMultipliers}.
+%It would be interesting to explore ways of using the host language in more sophisticated ways, rather as we have done for hardware generation in Lava~\citeemb{LavaMultipliers}.
 
@@
 in a Masters course on parallel functional programming that has recently been
-introduced at Chalmers~\cite{TFPIEPFP}. The first instance of the course (in Spring 2012)
+introduced at Chalmers~\citeemb{TFPIEPFP}. The first instance of the course (in Spring 2012)
 covered Blelloch's NESL language, with its
-associated cost model~\cite{NESL}, but did not provide any satisfactory way for the students
+associated cost model~\citeemb{NESL}, but did not provide any satisfactory way for the students
 to experience real, nested data parallel programming. This was due not only to a lack of suitable hardware, but also to deficiencies in the available tools.

### 2013-10-06 src2/embarbb/results.tex
apa
hacking

@@
 \subsection{Benchmarks}
-\label{sec:Benchmarks}
+\label{sec:EmbArBBBenchmarks}
 %\FloatBarrier
@@
 parallel execution of the resulting code, making use of multiple
-cores, but not of SIMD parallelism~\cite{REPA}.
+cores, but not of SIMD parallelism~\citeemb{REPA}.
 The Repa versions of the benchmarks used in the comparison 

### 2013-10-06 src2/expressive/applications.tex
apa
hacking

@@
 Our small kernels are constructed in the form of sorting and merging {\em networks},
-building on Batcher's bitonic merger~\cite{Batcher} and
-on the periodic balanced merger~\cite{PeriodicBalanced}.
+building on Batcher's bitonic merger~\citeexp{Batcher} and
+on the periodic balanced merger~\citeexp{PeriodicBalanced}.
 We chose also to implement the large sorter used  in benchmarking the
@@
 construction and we have earlier explored ways to describe
-and analyse it in both (our) Ruby and in Lava~\cite{sortsRuby,LAVASORTER}.
+and analyse it in both (our) Ruby and in Lava~\citeexp{sortsRuby,LAVASORTER}.
 Here, we consider iterative descriptions using similar combinators.
@@
 order.
-\pagebreak
+%\pagebreak
 %% ** TODO More explanation needed.
@@
 In order to go faster still, we resort to building a different sorting network, of
-exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\cite{PeriodicBalanced}.
+exactly the same size as the bitonic sorter, but based instead on the balanced period merger of Dowd et al~\citeexp{PeriodicBalanced}.
 This involves the introduction of one new combinator that can be seen as a mixture
@@
 That is, we replaced the loop containing {\tt iSwap} above by
-\pagebreak
+% \pagebreak
 \begin{codesize}

### 2013-10-06 src2/expressive/discussion.tex
hacking

@@
 The series of kernels also illustrates how the use of combinators brings a form of reuse, and makes design exploration easier. Our experience of using similar combinators in the Lava hardware
-description language~\cite{LAVASORTER} was that a relatively small set of combinators went a long way. So, although we introduced three
+description language~\citeexp{LAVASORTER} was that a relatively small set of combinators went a long way. So, although we introduced three
 combinators here, {\tt ilv}, {\tt vee} and {\tt ilvVee}, which

### 2013-10-06 src2/expressive/intro.tex
apa
hacking

--- 2013-09-30 src2/expressive/intro.tex
@@
 When an Obsidian program is run, a representation of the program is created 
-as a syntax tree. For more information on EDSL implementation see~\cite{COMPILEEDSL}.
+as a syntax tree. For more information on EDSL implementation see~\citeexp{COMPILEEDSL}.
 The program representation generated when running an Obsidian program 
@@
 Our approach is different from that of other Haskell DSELs targetting
-GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. 
+GPUs~\citeexp{ACCELERATEDAMP11,NIKOLA,BARRACUDA}. 
 We do not try to abstract away from 
@@
 {\tt map}, {\tt reduce} and {\tt zipWith} as built in skeletons, implemented
-with the help of small, predefined, hand-tuned CUDA kernels~\cite{ACCELERATE}. 
+with the help of small, predefined, hand-tuned CUDA kernels~\citeexp{ACCELERATEDAMP11}. 
 Obsidian, on the other hand, allows the user to experiment with the
@@
 allows the programmer to guide code generation and control parallelism and thread
-use~\cite{JSLIC}.
+use~\citeexp{JSLIC}.
 
 %Today there are number of EDSLs written in Haskell targeting 
-%GPUs~\cite{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
+%GPUs~\citeexp{ACCELERATE,NIKOLA,BARRACUDA}. We justify our work on yet 
 %another by having a slightly different approach. We do not try to abstract away from 
@@
 %cursory, for a more complete view of the various versions and approaches 
-%we have tried, see~\cite{JSLIC}.
+%we have tried, see~\citeexp{JSLIC}.
 %More details on this later.
@@
 Using this array representation in a DSEL is not new; the first occurence that we know of is in
-Pan~\cite{PAN}. Similar array representations have also later been 
-used in Feldspar~\cite{FELDSPAR}, and more recently also in the Repa library~\cite{REPA}.
+Pan~\citeexp{PAN}. Similar array representations have also later been 
+used in Feldspar~\citeexp{FELDSPAR}, and more recently also in the Repa library~\citeexp{REPA}.
 Functions for indexing and getting the length of arrays are as follows:
@@
 The generated code for arrays of length eight is 
-\pagebreak
+%\pagebreak
 %The code is generated with the 
@@
 %For a more in depth view of the parts of Obsidian that has been showed 
-%in this section see~\cite{JSLIC}.
+%in this section see~\citeexp{JSLIC}.

### 2013-10-06 src2/expressive/push.tex
apa
hacking

@@
-\section{Push Arrays}
+\subsection{Push Arrays}
 
@@
 
-\subsection {What are Push Arrays?}
+\subsubsection {What are Push Arrays?}
 
@@
 For another example of using continuations and a more complete
-description of their meaning and application, see~\cite{POORKOEN}.
+description of their meaning and application, see~\citeexp{POORKOEN}.
 
@@
 in each thread of the GPU; using sequential computations per thread has 
-been shown to be beneficial~\cite{OLAMARCUS}. But for the general case 
+been shown to be beneficial~\citeexp{OLAMARCUS}. But for the general case 
 of arbitrary nesting, some method of flattening is needed. 

### 2013-10-06 src2/hl/paperThesis.tex
hacking

@@
 For key building blocks like scan and reduce, we see a trend towards
-the use of fewer and much more complicated kernels~\cite{merrill}, so
+the use of fewer and much more complicated kernels~\citehl{merrill}, so
 that the host code is often relatively simple. We wish to support the construction of these sophisticated kernels.
@@
 a domain specific language that could 
-benefit from GPU acceleration. Reference ~\cite{BEAUTY} is an example of this.   
+benefit from GPU acceleration. Reference ~\citehl{BEAUTY} is an example of this.   
 %One such experiment can be found in 
@@
 
-The GPUs we target with Obsidian are NVIDIA GPUs that support CUDA~\cite{CUDA}.
+The GPUs we target with Obsidian are NVIDIA GPUs that support CUDA~\citehl{CUDA}.
 CUDA is the NVIDIA C-dialect for data-parallel programming on their GPUs. 
 We try to stay within the subset of functionality that is common 
-between CUDA and OpenCL~\cite{OpenCL}, as we may in future want to generate 
+between CUDA and OpenCL~\citehl{OpenCL}, as we may in future want to generate 
 OpenCL programs.
@@
 {\em Push} arrays. A pull array naturally represents gather operations,
-while a push array captures scattering operations. Reference \cite{Obsidian-Expressive} 
+while a push array captures scattering operations. Reference \citehl{Obsidian-Expressive} 
 describes the introduction of push arrays into Obsidian.
@@
 in many other embedded languages. The 
-embedded language Pan ~\cite{COMPILEEDSL} used a similar representation for 
-images and it has been used since in Obsidian, Feldspar ~\cite{FELDSPAR} 
-and Repa ~\cite{REPA}. In Obsidian, there is a {\tt force} function that,
+embedded language Pan ~\citehl{COMPILEEDSL} used a similar representation for 
+images and it has been used since in Obsidian, Feldspar ~\citehl{FELDSPAR} 
+and Repa ~\citehl{REPA}. In Obsidian, there is a {\tt force} function that,
 when applied to a pull array, writes the elements of that array 
@@
 This data type also represents the lowest level programming interface 
-exposed by Obsidian. In reference ~\cite{JOSEFJOELCSORT} we make use of these low-level 
+exposed by Obsidian. In reference ~\citehl{JOSEFJOELCSORT} we make use of these low-level 
 capabilities to implement histograms using atomic operations (not shown here).
@@
 for the {\tt Program} data type is not entirely straightforward. Our approach 
-is described in further detail in reference ~\cite{BB}; the presentation 
+is described in further detail in reference ~\citehl{BB}; the presentation 
 here focuses instead on the type directed compilation of the {\tt ForAll} 
@@
 the {\tt Program}. This follows the same monad reification technique as described 
-in reference~\cite{BB}.
+in reference~\citehl{BB}.
 
@@
 presented here is based on a sequential C program from reference 
-~\cite{Fractals}. However, the problem is embarrassingly parallel.
+~\citehl{Fractals}. However, the problem is embarrassingly parallel.
 
@@
 optimisations that are applied to the kernels are influenced by a 
-tutorial presentation from NVIDIA~\cite{reduction}.
+tutorial presentation from NVIDIA~\citehl{reduction}.
 
@@
 Note the use of two {\em write functions} in sequence.
-Similar patterns were used in our implementations of sorting networks~\cite{Obsidian-Expressive}, for similar reasons.
+Similar patterns were used in our implementations of sorting networks~\citehl{Obsidian-Expressive}, for similar reasons.
 
@@
 %\subsection{Scans} 
-%\cite{OLA} \cite{LargeScan}
+%\citehl{OLA} \citehl{LargeScan}
 
@@
 There are many languages and libraries for GPU programming. Starting at the 
-low-level end of the spectrum we have CUDA ~\cite{CUDA}. CUDA is NVIDIA's 
+low-level end of the spectrum we have CUDA ~\citehl{CUDA}. CUDA is NVIDIA's 
 name for the programming model and extended C language for their GPUs. 
@@
 other end of the high-level - low-level spectrum, we have the NVIDIA
-Thrust Library~\cite{THRUST}.
+Thrust Library~\citehl{THRUST}.
 Thrust offers a programming model where details of GPU architecture are 
@@
 Data.Array.Accelerate is a language embedded in Haskell for GPU programming 
-~\cite{ACCELERATEDAMP11}. The abstraction level is comparable to that of Thrust. In 
+~\citehl{ACCELERATEDAMP11}. The abstraction level is comparable to that of Thrust. In 
 other words, Accelerate hides most GPU details from the programmer. Accelerate 
@@
 the optimisation of Accelerate programs using fusion techniques to decrease
-the number of kernel invocations needed (see reference ~\cite{ACCOPT}).
+the number of kernel invocations needed (see reference ~\citehl{ACCOPT}).
 %Accelerate programs are being optimised using using fusion 
@@
 
-Nikola ~\cite{NIKOLA} is another language embedded in Haskell that occupies the 
+Nikola ~\citehl{NIKOLA} is another language embedded in Haskell that occupies the 
 same place as Accelerate and Thrust on the abstraction level spectrum.
@@
 and Reppy are attempting nested data-parallelism by implementing a 
-compiler for the NESL language for GPUs ~\cite{NESL}.
+compiler for the NESL language for GPUs ~\citehl{NESL}.
 
@@
 
-The Copperhead \cite{copperhead} system compiles a subset of Python to run 
+The Copperhead \citehl{copperhead} system compiles a subset of Python to run 
 on GPUs. Much like other languages mentioned here, Copperhead identifies 
@@
 
-In reference ~\cite{Finance}, Oancea et al. use manual transformations
+In reference ~\citehl{Finance}, Oancea et al. use manual transformations
 to study a set of 
@@
 to monad reification. Related to this work is the approach based on a 
-continuation monad by Persson et al~\cite{Generic}. Recent work by Sculthorpe et al. in reference ~\cite{sculthorpe2013constrained} is also related. 
+continuation monad by Persson et al~\citehl{Generic}. Recent work by Sculthorpe et al. in reference ~\citehl{sculthorpe2013constrained} is also related. 
 Our approach to monad reification is simpler but less general.
@@
 
-In reference ~\cite{CSORT}, we experiment with atomic operations 
+In reference ~\citehl{CSORT}, we experiment with atomic operations 
 and are forced to write our code in a very imperative style. The addition of mutable 
@@
 illustrated in section \ref{sec:redOpt}, where we manage to apply techniques
-used in the NVIDIA tutorial ~\cite{reduction}. The case study also show hows we 
+used in the NVIDIA tutorial ~\citehl{reduction}. The case study also show hows we 
 can compose kernels and thus reuse prior effort. 
@@
 the implementation of Obsidian, and showed that the method works
-well in practice (see reference ~\cite{BB} for more details of the method).
+well in practice (see reference ~\citehl{BB} for more details of the method).
 
@@
 %We hinted at our simple and direct method of monad reification. This is described 
-%in more detail in reference ~\cite{BB}. 
+%in more detail in reference ~\citehl{BB}. 
 %In Obsidian, we do not currently make 
@@
 
-%\bibliographystyle{abbrvnat}
-%\bibliography{paper}
+\bibliographystylehl{alpha}
+\bibliographyhl{thesis}
+
 % The bibliography should be embedded for final submission.

### 2013-10-06 src2/ifl/GPUs.tex
hacking

@@
 demands. For a recent survey of the hardware, programming methods and tools,
-and successful applications, the reader is referred to~\cite{GPUComputing}.
+and successful applications, the reader is referred to~\citeifl{GPUComputing}.
 Figure~\ref{fig:image}, taken from that paper, and due to NVIDIA, shows the
@@
 
-See reference~\cite{GPUComputing} for information about
+See reference~\citeifl{GPUComputing} for information about
 the very similar AMD GPU architecture. We have used the NVIDIA architecture,
@@
 future in which each individual core is considerably more powerful than in
-today's GPUs~\cite{larrabee}.
+today's GPUs~\citeifl{larrabee}.
 
@@
 CUDA -- an extension of C designed to allow developers to exploit the power
-of GPUs. Reference~\cite{GPUCudaLuebke} gives a very brief but illuminating
+of GPUs. Reference~\citeifl{GPUCudaLuebke} gives a very brief but illuminating
 introduction to CUDA for potential new users. The idea is that the user

### 2013-10-06 src2/ifl/combinators.tex
hacking

@@
 Inspired by Backus' FP language, Sheeran's early work on $\mu$FP made use of
-{\em combining forms} with geometric interpretations~\cite{SheeranLFP84}. This approach
+{\em combining forms} with geometric interpretations~\citeifl{SheeranLFP84}. This approach
 to capturing circuit regularity was also influenced by contact
-with designers of regular array circuits in industry -- see reference~\cite{SheeranJUCS} for an overview of this and much other work on functional programming and hardware design.
+with designers of regular array circuits in industry -- see reference~\citeifl{SheeranJUCS} for an overview of this and much other work on functional programming and hardware design.
 
-Later work on (our) Ruby considered the use of binary relations, rather than functions in specifying hardware~\cite{LyngbyRuby}.
-Lava builds upon these ideas, but also gains much in expressiveness and flexibility by being embedded in Haskell~\cite{lavaICFP,ClaessenThesis}.
+Later work on (our) Ruby considered the use of binary relations, rather than functions in specifying hardware~\citeifl{LyngbyRuby}.
+Lava builds upon these ideas, but also gains much in expressiveness and flexibility by being embedded in Haskell~\citeifl{lavaICFP,ClaessenThesis}.
 The user writes what look like circuit descriptions, but are in fact circuit {\em generators}. Commonly used {\em connection patterns} are captured by higher order functions.
@@
 Circuit descriptions are {\em run} (in fact symbolically evaluated) in order to produce an intermediate representation, which is in turn written out in various formats (for fixed size instances).
-So this is an example of {\em staged programming}~\cite{Taha03}.
+So this is an example of {\em staged programming}~\citeifl{Taha03}.
 
-The Sklansky construction is one way to implement parallel prefix, and there are many others, see for instance Hinze's excellent survey~\cite{Hinze}.
+The Sklansky construction is one way to implement parallel prefix, and there are many others, see for instance Hinze's excellent survey~\citeifl{Hinze}.
 Those who develop prefix algorithms suitable for hardware implementation use a standard notation to represent
@@
 about implementing algorithms on GPUs. This literature does not seem to have yet been scrutinised by the Data Parallel Programming or GPGPU communities.
-This is possibly because GPUs are moving closer to simply being data parallel machines, and so work on library functions has taken inspiration from earlier work on Data Parallel Programming, such as Blelloch's NESL~\cite{NESL}.
+This is possibly because GPUs are moving closer to simply being data parallel machines, and so work on library functions has taken inspiration from earlier work on Data Parallel Programming, such as Blelloch's NESL~\citeifl{NESL}.
 But some of the restrictions from the early data parallel machines no longer hold today; for instance broadcasting a value to many processors was expensive in the past, but is much easier to do on modern GPUs.

### 2013-10-06 src2/ifl/conclusion.tex
apa
hacking

@@
-\section{Conclusion} \label{sec:conc}
+\subsection{Conclusion} \label{sec:conc}
 
 Obsidian provides a good interface for experimenting with algorithms on GPUs. 
-The earlier version described in~\cite{JMT} showed that it is possible to 
+The earlier version described in~\citeifl{JMT} showed that it is possible to 
 generate efficient CUDA code from the kind of high level descriptions we are 

### 2013-10-06 src2/ifl/related.tex
hacking

@@
 {Pan}, an embedded language for image synthesis developed by 
-Conal Elliot~\cite{PAN}. Because of the computational complexity of image generation, C 
+Conal Elliot~\citeifl{PAN}. Because of the computational complexity of image generation, C 
 code is generated. This C code can then be compiled by an optimising compiler. 
@@
 ``Compiling Embedded Languages'', describing the implementation of Pan have been 
-used in the implementation of Obsidian~\cite{COMPILEEDSL}.
+used in the implementation of Obsidian~\citeifl{COMPILEEDSL}.
 
 \subsubsection{Related work on GPU and GPGPU programming languages}
-We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\cite{Lucas08}.
+We cannot attempt an exhaustive description of GPU programming languages here, but refer the reader to a recent PhD thesis by Philipp Lucas, which contains an enlightening survey~\citeifl{Lucas08}.
 Lucas distinguishes carefully between languages (such as CG and HLSL) that aim to raise the level of abstraction at which graphics-oriented GPU programs are written, and those that attempt to abstract the entire GPU, and so must also provide a means to express the placing of programs on the GPU, feeding such programs with data, reading the results back to the CPU, and so on, as well as deciding to what extent the programmer should be involved
@@
 In the first group of graphics-oriented languages, we include PyGPU and Vertigo.
-PyGPU is a language for image processing embedded in Python~\cite{PyGPU}. 
+PyGPU is a language for image processing embedded in Python~\citeifl{PyGPU}. 
 PyGPU uses the introspective abilities of Python and thus bypasses
@@
 Programs written in PyGPU can be compiled and run on a GPU. 
-Vertigo is another embedded language by Conal Elliot~\cite{VERTIGO}. It
+Vertigo is another embedded language by Conal Elliot~\citeifl{VERTIGO}. It
 is a language for 3D graphics that targets the DirectX 8.1 shader model, and
 can be used to describe geometry, shaders and to generate textures. 
-%Each sublanguage is given formal semantics \cite{VERTIGO}. From programs 
+%Each sublanguage is given formal semantics \citeifl{VERTIGO}. From programs 
 %written in Vertigo, assembly language programs are generated for execution on a 
@@
 language that targets both GPUs and SIMD capable CPUs -- with the aim being
-a combination of a high degree of abstraction and a close resemblance to traditional programming languages~\cite{Lucas08}.
-BrookGPU (which is usually just called Brook) is a classic example of a language~\cite{Brook} designed to raise the level of abstraction at which GPGPU programming is done.
+a combination of a high degree of abstraction and a close resemblance to traditional programming languages~\citeifl{Lucas08}.
+BrookGPU (which is usually just called Brook) is a classic example of a language~\citeifl{Brook} designed to raise the level of abstraction at which GPGPU programming is done.
 It is an extension of C with embedded kernels, aimed at arithmetic-intense data parallel computations.
@@
 Brook had significant impact in that it raised the level of abstraction at which GPGPU programming can be done.
-The language Sh also aimed to raise the level of abstraction at which GPUs were programmed~\cite{Sh}.
+The language Sh also aimed to raise the level of abstraction at which GPUs were programmed~\citeifl{Sh}.
 Sh was an embedded language in C++, so our work is close in spirit to it.
-Sh has since evolved into the RapidMind development platform~\cite{RapidMind}, which now supports multicores and Cell processors as well as GPUs.
+Sh has since evolved into the RapidMind development platform~\citeifl{RapidMind}, which now supports multicores and Cell processors as well as GPUs.
 The RapidMind programming model has arrays as first class types. It
@@
 A recent PhD thesis by Jansen asserts that there are some problems with RapidMind's use of macros to embed the GPU programming language in C++, including the inability to pass
-kernels (or shader programs) as classes~\cite{JansenThesis}; the thesis proposes GPU++ and claims improvement over previous approaches, particularly through the exploitation of automatic partitioning of the programs onto the available GPU hardware, and through compiler optimisations that improve runtime performance.
+kernels (or shader programs) as classes~\citeifl{JansenThesis}; the thesis proposes GPU++ and claims improvement over previous approaches, particularly through the exploitation of automatic partitioning of the programs onto the available GPU hardware, and through compiler optimisations that improve runtime performance.
 
-Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\cite{ACCELERATOR}.
-Data Parallel Haskell~\cite{DPH} incorporates Nested Data Parallelism in the style of NESL~\cite{NESL} into Haskell. GPUGen, like Obsidian, aims
-to support GPGPU programming from Haskell~\cite{GPUGenTalk}.
+Microsoft's Accelerator project moves even closer to general purpose programming by doing away with the kernel notion and simply expressing programs in a data parallel style, using functions on arrays~\citeifl{ACCELERATOR}.
+Data Parallel Haskell~\citeifl{DPH} incorporates Nested Data Parallelism in the style of NESL~\citeifl{NESL} into Haskell. GPUGen, like Obsidian, aims
+to support GPGPU programming from Haskell~\citeifl{GPUGenTalk}.
 It works by translating Haskell's intermediate language, Core, into CUDA, for collective data operations such as scan, fold and map.
@@
 Our impression is that we wish to expose considerably more detail about the GPU to the programmer, but we do not yet have sufficient information about GPUGen to be able to do a more complete comparison.
-Finally, we mention the Spiral project, which develops methods and tools for automatically generating high performance libraries for a variety of platforms, in domains such a signal processing, multiplication and sorting~\cite{Spiral}.
+Finally, we mention the Spiral project, which develops methods and tools for automatically generating high performance libraries for a variety of platforms, in domains such a signal processing, multiplication and sorting~\citeifl{Spiral}.
 The tuning of an algorithm for a given platform is expressed as an optimisation problem, and the domain specific mathematical structure of the algorithm is used
@@
 to express how the memory hierarchy is to be used, where a relatively abstract description of
-the platform, viewed as a tree of processing nodes and memories, is a parameter~\cite{Sequoia}.
+the platform, viewed as a tree of processing nodes and memories, is a parameter~\citeifl{Sequoia}.
 Thus, programmers should write very generic code, which can be compiled for many different platforms.

### 2013-10-06 src2/ifl/results.tex
hacking

@@
 Obsidian to implement sorters. For the sorters, the generated C code 
-performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\cite{LAVASORTER},
+performs quite well. One periodic sorting network, called {\em Vsort}, implemented in Obsidian in the style of the sorters presented in reference~\citeifl{LAVASORTER},
 has a running time of 95$\mu$secs for 256 elements (the largest size that we can cope with in a single kernel). This running time can be compared to the 

### 2013-10-06 src2/papp/paperThesis.tex
apa
hacking

@@
 NVIDIA CUDA code is generated from these high level descriptions, and 
-passed to the {\tt nvcc} compiler~\cite{CUDAGuide2.0}. Currently, we 
+passed to the {\tt nvcc} compiler~\citepapp{CUDAGuide2.0}. Currently, we 
 consider only the generation of single kernels, and not their coordination.
@@
 %Obsidian generates NVIDIA CUDA code that is passed to the {\tt nvcc} 
-%compiler \cite{CUDAGuide2.0}.
+%compiler \citepapp{CUDAGuide2.0}.
 
@@
 %of Obsidian itself and a discussion of how the CUDA code is generated can 
-%be found in \cite{JSTECH}.
+%be found in \citepapp{JSTECH}.
 
@@
 GPUs come with hundreds of processing elements capable of managing thousands 
-of threads \cite{Brief8800}. The Obsidian project is about exploring ways 
+of threads \citepapp{Brief8800}. The Obsidian project is about exploring ways 
 to program these new machines.
@@
 GPUs (GPGPU), capable of generating code for modern NVIDIA GPUs. 
-In~\cite{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
+In~\citepapp{JSMSKC_IFL08} we describe a version of Obsidian that uses a monadic 
 interface. In the current version of Obsidian the monad has been replaced 
-by another data structure, closely related to Arrows~\cite{Arrows}, representing GPU programs. Although the details are 
+by another data structure, closely related to Arrows~\citepapp{Arrows}, representing GPU programs. Although the details are 
 left out because of space limitations, the use of this GPU program representation 
 can be seen in subsection~\ref{sec:GPUPrograms}. For a description of the implementation 
-see~\cite{JSTECH} and for more general information on embedded language 
-implementation see~\cite{COMPILEEDSL}.
+see~\citepapp{JSTECH} and for more general information on embedded language 
+implementation see~\citepapp{COMPILEEDSL}.
 
@@
 The kind of processing that is common in graphics falls in the data-parallel 
-category \cite{GEMS2}.
+category \citepapp{GEMS2}.
 % One example of a computation from graphics is the 
@@
 name CUDA (Compute Unified Device Architecture) 
-\cite{CUDAGuide2.0}. CUDA offers the GPGPU programmer a C %,wwwcuda,Brief8800
+\citepapp{CUDAGuide2.0}. CUDA offers the GPGPU programmer a C %,wwwcuda,Brief8800
 compiler and libraries for CUDA enabled GPUs (NVIDIA 8000 series and above).
@@
 on performance; the two diverging paths will in fact be executed sequentially 
-\cite{BestPrac}.
+\citepapp{BestPrac}.
 %There are however, more threads in a
@@
 %take different paths within a warp have a negative effect on performance;
-%the two diverging paths will in fact be executed sequentially \cite{BestPrac}.
+%the two diverging paths will in fact be executed sequentially \citepapp{BestPrac}.
 
@@
 the language features resemble those of Lava, a hardware description and 
-verification language \cite{lavaICFP}. The justification to use language 
+verification language \citepapp{lavaICFP}. The justification to use language 
 constructs similar to that of a hardware description language came from 
 the observation that GPGPU algorithms often were explained using circuit-like 
-pictures, see for example \cite{GEMS3}.
+pictures, see for example \citepapp{GEMS3}.
 
@@
 %literature where GPGPU programmers implement their kernels to operate on very 
-%specific array sizes \cite{GEMS3}. In a later stage, when the kernels are 
+%specific array sizes \citepapp{GEMS3}. In a later stage, when the kernels are 
 %composed into algorithms on large arrays, support for variable length 
@@
 program.% such as those in subsection~\ref{sec:ArrayLanguage}.
-The type constructor {\tt :->} is a variant of an arrow, which in turn is a generalisation of a monad~\cite{JSTECH}.
+The type constructor {\tt :->} is a variant of an arrow, which in turn is a generalisation of a monad~\citepapp{JSTECH}.
 %
@@
 \begin{verbatim}
-execute :: (Flatten a, Flatten b) => (Arr a :-> Arr b) -> [a] -> IO [b]
+execute :: (Flatten a, Flatten b) 
+         => (Arr a :-> Arr b) -> [a] -> IO [b]
 \end{verbatim}
@@
 \begin{verbatim} 
-Sync (fmap (toFData . f)) (Sync ((fmap toFData) . rev . (fmap fromFData)) (Pure (fmap g)))   
+Sync (fmap (toFData . f)) 
+     (Sync ((fmap toFData) . rev . (fmap fromFData)) 
+           (Pure (fmap g)))   
 \end{verbatim}
@@
 
-Sorting is a popular function to place on the GPU, see~\cite{SINTORNSORT,SORT3}.
+Sorting is a popular function to place on the GPU, see~\citepapp{SINTORNSORT,SORT3}.
 {\em Odd-Even merge sort}, due to Batcher, uses a merger component called the Odd-Even merger. This 
@@
 and {\tt odds}. For a detailed description of these combinators see
-~\cite{LAVASORTER}.
-
-\begin{small}
-\begin{verbatim}
-mergeOE :: (Choice a, Flatten a) => Int -> ((a,a) -> (a,a)) -> (Arr a :-> Arr a)
+~\citepapp{LAVASORTER}.
+
+\begin{small}
+\begin{verbatim}
+mergeOE :: (Choice a, Flatten a) 
+         => Int -> ((a,a) -> (a,a)) -> (Arr a :-> Arr a)
 mergeOE 1 f = pure (evens f)
@@
 
-*Obsidian> execute GPU (mergeOE 3 cmp) ([1,3,5,7,2,4,6,8] :: [IntE]) 
+*Obsidian> let input = ([1,3,5,7,2,4,6,8] :: [IntE]) 
+*Obsidian> execute GPU (mergeOE 3 cmp) input
 [1,2,3,4,5,6,7,8]
@@
 This subsection shows the implementation of a parallel prefix (also called scan) 
-kernel, known as {\tt sklansky} after J. Sklansky \cite{Sklansky}.
+kernel, known as {\tt sklansky} after J. Sklansky \citepapp{Sklansky}.
 This kernel will then be optimised step-by-step using Obsidian. This 
@@
 This is different from the normal {\tt sync} in the way it assigns work to threads. 
-For a more thorough explanation of {\tt syncHow} see ~\cite{JSTECH}.
-For an explanation of the parallel-prefix operation see~\cite{BlellochTR90}.
+For a more thorough explanation of {\tt syncHow} see ~\citepapp{JSTECH}.
+For an explanation of the parallel-prefix operation see~\citepapp{BlellochTR90}.
 
@@
 \begin{verbatim}
-sklansky :: (Flatten a, Choice a) => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
+sklansky :: (Flatten a, Choice a) 
+          => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
 sklansky 0 op = pure id
-sklansky n op = two (sklansky (n-1) op) ->- pure (fan op) ->- sync
+sklansky n op = two (sklansky (n-1) op) ->- 
+                pure (fan op) ->- sync
 \end{verbatim}
@@
 \begin{verbatim}
-sklansky1 :: (Flatten a, Choice a) => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
+sklansky1 :: (Flatten a, Choice a) 
+           => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
 sklansky1 0 op = pure id
-sklansky1 n op = two (sklansky1 (n-1) op) ->- pure (fan op) ->- syncHow (pairNth 256) 
+sklansky1 n op = two (sklansky1 (n-1) op) ->- 
+                 pure (fan op) ->- syncHow (pairNth 256) 
 \end{verbatim}
@@
 \begin{verbatim} 
-sklansky2 :: (Flatten a, Choice a) => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
+sklansky2 :: (Flatten a, Choice a) 
+           => Int -> (a -> a -> a) -> (Arr a :-> Arr a) 
 sklansky2 0 op = pure id
@@
 the parallel prefix of a large array. This is done in the same way as in
-~\cite{ScanCUDA}.
+~\citepapp{ScanCUDA}.
 The table below shows the results of using the different parallel prefix 
@@
 with the CUDA SDK. This kernel is based on the Brent-Kung network and its 
-implementation is shown in~\cite{ScanCUDA}. The {\tt NVIDIA SDK} 
+implementation is shown in~\citepapp{ScanCUDA}. The {\tt NVIDIA SDK} 
 kernel is highly optimized with regards to its memory access pattern and 
@@
 are based where obtained using the CUDA profiler supplied as part of the 
-CUDA toolkit~\cite{wwwcuda}.
+CUDA toolkit~\citepapp{wwwcuda}.
 
@@
 The question of how to program these machines arises. NVIDIA's answer is CUDA
-\cite{CUDAGuide2.0}. CUDA supplies a slightly extended version of C in which 
+\citepapp{CUDAGuide2.0}. CUDA supplies a slightly extended version of C in which 
 the programmer can specify GPU kernels and the controlling CPU program in the 
@@
 There are a number of other C/C++ based languages that target 
-GPGPU programmers, for example Brook\cite{Brook} and 
-RapidMind\cite{RapidMind}. Brook, CUDA and RapidMind are major improvements from 
+GPGPU programmers, for example Brook\citepapp{Brook} and 
+RapidMind\citepapp{RapidMind}. Brook, CUDA and RapidMind are major improvements from 
 what was previously available for the programmer interested in general purpose 
@@
 Higher level approaches are also being investigated. PyGPU embedds a GPGPU 
-programming language in Python\cite{PyGPU}. PyGPU makes use of Python's 
+programming language in Python\citepapp{PyGPU}. PyGPU makes use of Python's 
 introspective abilities to generate efficient code.
 
-Like Obsidian, Accelerate is embedded in Haskell \cite{GPUGEN}. Accelerate 
+Like Obsidian, Accelerate is embedded in Haskell \citepapp{GPUGEN}. Accelerate 
 however, is higher level language than Obsidian. Where the purpose of 
@@
 
-Vertigo is another GPU programming language embedded in Haskell\cite{VERTIGO}.
+Vertigo is another GPU programming language embedded in Haskell\citepapp{VERTIGO}.
 Vertigo can be used to describe procedureal surfaces and textures and generates
@@
 that target not only GPUs but also other platforms, such as multicore machines.
-Of particular interest to us is Microsoft's Accelerator project~\cite{ACCELERATOR}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
+Of particular interest to us is Microsoft's Accelerator project~\citepapp{ACCELERATOR}. Interestingly, Accelerator includes the same restriction to ``hardware-like'' algorithms as Obsidian does.
 %However, Vertigo is targeting graphical applications.
@@
 %Data-Parallel Haskell extends Haskell with parallel arrays and operations 
-%on parallel arrays\cite{DPH}. Data-Parallel Haskell implements the nested 
+%on parallel arrays\citepapp{DPH}. Data-Parallel Haskell implements the nested 
 %data-parallel paradigm and is in that sense following in the path of 
-%NESL\cite{NESL}.
+%NESL\citepapp{NESL}.
 
 %\begin{itemize}
-%\item Chapel \cite{Chapel}
-%\item DPH \cite{DPH}
-%\item Fortress \cite{fortress}
-%\item GPUGen \cite{GPUGEN}
-%\item Lava \cite{lavaTutorial}
-%\item RapidMind \cite{RapidMind}
-%\item Sequoia \cite{SEQUOIA}
-%\item X10 \cite{X10}
+%\item Chapel \citepapp{Chapel}
+%\item DPH \citepapp{DPH}
+%\item Fortress \citepapp{fortress}
+%\item GPUGen \citepapp{GPUGEN}
+%\item Lava \citepapp{lavaTutorial}
+%\item RapidMind \citepapp{RapidMind}
+%\item Sequoia \citepapp{SEQUOIA}
+%\item X10 \citepapp{X10}
 %\end{itemize}
@@
 
-%\bibliographystyle{elsarticle-num}
-%\bibliography{obs_okt}
+\bibliographystylepapp{alpha}
+\bibliographypapp{thesis}
 

### 2013-10-06 src2/embarbb/conclusion.tex
apa
update

@@
-\section{Conclusion} 
+\subsection{Conclusion} 
 We have shown that Intel's Array Building Blocks (ArBB) provides
 an interface that is well suited to functional programming. 
-The programs are quite close to mathematical specifications, in the style of NESL~\cite{NESL}.
+The programs are quite close to mathematical specifications, in the style of NESL~\citeemb{NESL}.
 We have only just completed the embedding of the part of ArBB that deals with nested vectors, and we need to tackle many more case studies.

### 2013-10-07 src2/thesis.tex
hacking
apa
update

@@
 \usepackage{booktabs}
+%\usepackage{etoolbox}
+
+\usepackage{multibib} 
+\newcites{bb}{Bibliography} 
+\newcites{ifl}{Bibliography}
+\newcites{papp}{Bibliography}
+\newcites{exp}{Bibliography}
+\newcites{hl}{Bibliography}
+\newcites{csort}{Bibliography}
+\newcites{emb}{Bibliography}
+\newcites{arbb}{Bibliography}
+% \newcites{t}{Bibliography}
+%\newcites[ifl}{Bibliography}
+
+%\patchcmd{\thebibliography}{\section*}{\section}{}{}
+
+\makeatletter
+\renewenvironment{thebibliography}[1]
+     {\section*{\bibname}%
+      %\@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
+      \list{\@biblabel{\@arabic\c@enumiv}}%
+           {\settowidth\labelwidth{\@biblabel{#1}}%
+            \leftmargin\labelwidth
+            \advance\leftmargin\labelsep
+            \@openbib@code
+            \usecounter{enumiv}%
+            \let\p@enumiv\@empty
+            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
+      \sloppy
+      \clubpenalty4000
+      \@clubpenalty \clubpenalty
+      \widowpenalty4000%
+      \sfcode`\.\@m}
+     {\def\@noitemerr
+       {\@latex@warning{Empty `thebibliography' environment}}%
+      \endlist}
+\makeatother
+
 %%
@@
 
+% ---------------------------------------------------------------------------
+%  ACKS !!! 
+% ---------------------------------------------------------------------------
 \section*{Acknowledgments}
@@
 Programming (RAW FP) Project) and by the Swedish Research Council.
+
+\clearpage
+
+% ---------------------------------------------------------------------------
+%  Guide for Oponent, grading committee
+% ---------------------------------------------------------------------------
+\section*{About this thesis}
+
+This thesis contains an introduction and a selection of papers. The introduction 
+gives background and context to the papers.
+
+The papers contained in this thesis falls in three categories. 
+\begin{itemize} 
+\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
+\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
+\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
+\end{itemize} 
+%In the following sections, each paper is given a short description. Hopefully 
+%there is enough information to wet the apetite.
+
+\subsection*{EDSL implementation papers}
+
+\subsubsection*{\paperBB}
+
+Authors: Josef Svenningsson and Bo Joel Svensson
+
+\vspace{5mm}
+
+\noindent We present a simple and compositional method for refication 
+of monads (or compilation of a monadic embedded language). I implemented 
+Obsidian using the method described here. The novelty and 
+usefulness of this simple way to reify monads was recognised by Josef
+and lead to the writing of this paper.
+
+\subsection*{GPU Programming papers}
+
+\subsubsection{\paperIFL}
+
+Authors: Bo Joel Svensson, Mary Sheeran and Koen Claessen \newline
+
+\vspace{5mm}
+
+\subsubsection{\paperPAPP}
+
+Authors: Bo Joel Svensson, Koen Claessen and Mary Sheeran \newline
+
+\vspace{5mm}
+
+\subsubsection{\paperEXPRESSIVE}
+
+Authors: Koen Claessen, Mary Sheeran and  Bo Joel Svensson \newline
+
+\vspace{5mm}
+
+\subsubsection{\paperCSORT}
+
+Authors: Josef Svenningsson, Bo Joel Svensson and Mary Sheeran \newline
+
+\vspace{5mm}
+
+\subsubsection{\paperHL}
+
+Authors: Bo Joel Svensson and Mary Sheeran \newline
+\noindent \emph{This work has not been published.}
+\vspace{5mm}
+
+\noindent
+
+In this paper the most recent additions and improvements of Obsidian are 
+described.
+
+\subsubsection{Retargetable parallel programming}
+
+\subsubsection{\paperARBB}
+
+Authors: Bo Joel Svensson and Ryan R. Newton
+
+\vspace{5mm}
+
+\noindent This position paper describes the work I did while on a 3 month internship at 
+Intel. The main part of this work was implementation of Haskell bindings 
+for the now retired Intel ArBB system.
+
+The paper also describes work we did towards implementing a backend for the 
+Accelerate system using ArBB. We reached a proof-of-concept implementation 
+capable of running some simple Accelerate programs using the ArBB backend.
+However, much work would remain to be able to run the full scope of Accelerate 
+programs on ArBB.
+
+\subsubsection{\paperEMBARBB}
+
+Authors: Bo Joel Svensson and Mary Sheeran
+
+\vspace{5mm}
+
+\noindent Here we explore another way to provide the capabilities of ArBB to the 
+Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
+we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
+Simply put, this means providing a more Haskell-programmer-friendly interface compared
+to the raw ArBB bindings (which are a collection of functions in the IO monad).
+
+\subsection*{Guide}
 
@@
 
-\chapter{About this thesis}
-
-This thesis contains an introduction and a selection of papers. The introduction 
-gives background and context to the papers.
-
-\section{Papers}
-
-The papers contained in this thesis falls in three different categories. 
-\begin{itemize} 
-\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
-\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
-\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
-\end{itemize}
-
-\subsection{EDSL implementation papers}
-
-\subsubsection{\paperBB}
-
-apa apa apa bepa cepa apa
-
-\subsection{GPU Programming papers}
-
-\subsubsection{\paperIFL}
-
-\subsubsection{\paperPAPP}
-
-\subsubsection{\paperEXPRESSIVE}
-
-\subsubsection{\paperCSORT}
-
-\subsubsection{\paperHL}
-
-\subsubsection{Retargetable parallel programming}
-
-\subsubsection{\paperARBB}
-
-\subsubsection{\paperEMBARBB}
-
+% --------------------------------------------------------------------------- %
+% --------------------------------------------------------------------------- %
+%
+%  INTRODUCTION 
+%
+% --------------------------------------------------------------------------- %
+% --------------------------------------------------------------------------- %
 \chapter{Introduction}
 
-\section{Plan} 
-\begin{itemize} 
-  \item The Scope
-  \item The Problem 
-  \item The Methods
-  \item The Results 
-\end{itemize}
+\section{The scope}
 
-\begin{itemize} 
-  \item The Context
-  \item Lift head from details 
-  \item Why are there interesting questions around parallelism and GPUs
-  \item The Story
-  \item Reflect on leassons learnt along the way 
-  \item Tell a story around the papers 
-  \item Future Work
-\end{itemize}
+\noindent\emph{Data-Parallelism, Structured-Compositional programming}
+
+\noindent\emph{GPUs}
+
+\noindent\emph{The context, Lift head from details}
+
+\subsection{Related work}
+
+\noindent\emph{GPU Programmming languages}
+
+\noindent\emph{data-parallel programming languages}
+
+\noindent\emph{Embedded languages}
+
+\noindent\emph{Monad reification methods}
+
+\section{The problem}
+
+\emph{what is so interesting about Parallelism and GPUs}
+
+\section{The method}
+
+\noindent\emph{Obsidian, Embedded languages}
+
+\noindent\emph{Array representations}
+
+\noindent\emph{Compositionality}
+
+\noindent\emph{Hierarchies}
+
+\section{The results}
+
+\emph{ How well did we do?}
+
+\section{Future work}
+
+\noindent\emph{Find a niche for Obsidian}
+
+\noindent\emph{Improve on Warp level capabilities}
+
+\noindent\emph{Improve expressivity, code performance}
+
+\section{Discussion}
 
 \clearpage{}
-%\section*{Additional Publications}
 
-%Papers not included in thesis
-
-%\bibliographystyle{plain}
-%\bibliography{intro}
+% --------------------------------------------------------------------------- %
+% --------------------------------------------------------------------------- %
+%
+%  PAPERS 
+%
+% --------------------------------------------------------------------------- %
+% --------------------------------------------------------------------------- %
 
@@
 
-\bibliographystyle{plain}
+% ---------------------------------------------------------------------------
+% nocites 
+% ---------------------------------------------------------------------------
+\nocite{*}
+
+\makeatletter
+\renewenvironment{thebibliography}[1]
+     {\chapter*{\bibname}%
+      \@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
+      \list{\@biblabel{\@arabic\c@enumiv}}%
+           {\settowidth\labelwidth{\@biblabel{#1}}%
+            \leftmargin\labelwidth
+            \advance\leftmargin\labelsep
+            \@openbib@code
+            \usecounter{enumiv}%
+            \let\p@enumiv\@empty
+            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
+      \sloppy
+      \clubpenalty4000
+      \@clubpenalty \clubpenalty
+      \widowpenalty4000%
+      \sfcode`\.\@m}
+     {\def\@noitemerr
+       {\@latex@warning{Empty `thebibliography' environment}}%
+       \endlist}
+\makeatother
+
+\bibliographystyle{alpha}
 \bibliography{thesis}

### 2013-10-08 src2/thesis.tex
update

@@
 
+The first real paper about Obsidian, our embedded language for GPU 
+programming. In this paper we identify similarities between 
+conenction patterns in hardware and parallel computations on GPUs. 
+We introduce a structured and compositional way to write data-parallel 
+programs on a GPU.
+
+Basic parallel building blocks are implemented in a monadic style and 
+composed using a sequential composition operator {\tt ->-}.
+
 \subsubsection{\paperPAPP}
@@
 
+\emph{a new program representation, based vaguely on Arrows}
+
 \subsubsection{\paperEXPRESSIVE}
@@
 \vspace{5mm}
+
+This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
+array representation target specific performance problems that we want 
+to solve with Obsidian.
 
@@
 % --------------------------------------------------------------------------- %
-\chapter{Introduction}
+\chapter{\thesistitle}
 
-\section{The scope}
+\section{Introduction}
+
+Computers today are increasingly parallel. With each new generation of 
+general purpose processor or GPU we get more parallel resources. To 
+fully make use of modern CPUs and GPUs we need to fully exploit their 
+parallelism.
+
+Modern CPUs are parallel in a number of different ways. They discover instruction
+level parallelism(ILP), they have multiple processor cores and vector units for 
+Single Instruction Multiple Data parallelism. The exact configurations, number
+of cores and number of vector units, vary over available processors.
+
+\noindent\emph{More text, lead into something about ArBB and forward ref to ArBB papers}
 
@@
 
-\noindent\emph{GPUs}
+Graphics Processors (GPUs), trade programmer convenience for increased parallelism. 
+For example, where CPUs dedicate much chip real estate for caches, branch prediction and 
+ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
+GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
+perspective and memory access patterns can make or break the performance of an 
+application.
 
-\noindent\emph{The context, Lift head from details}
+\noindent\emph{More text, lead into something about compositional, high-level, but 
+expressive enough programming models for GPU programming and ref to Obsidian Chapter}
 
-\subsection{Related work}
+%\noindent\emph{GPUs}
 
-\noindent\emph{GPU Programmming languages}
+%\subsection{Data-parallel programming on GPUs}
 
-\noindent\emph{data-parallel programming languages}
-
-\noindent\emph{Embedded languages}
-
-\noindent\emph{Monad reification methods}
+%\subsection{Data-parallel programming on CPUs}
 
@@
 
-\section{The method}
+\section{An embedded languages approach to data-parallelism}
 
@@
 \emph{ How well did we do?}
+
+\subsection{Related work}
+
+\noindent\emph{GPU Programmming languages}
+
+\noindent\emph{data-parallel programming languages}
+
+\noindent\emph{Embedded languages}
+
+\noindent\emph{Monad reification methods}
 

### 2013-10-09 src2/thesis.tex
apa

@@
 
-\subsection*{GPU Programming papers}
+\subsection*{GPU programming papers}
 
@@
 
-Computers today are increasingly parallel. With each new generation of 
-general purpose processor or GPU we get more parallel resources. To 
-fully make use of modern CPUs and GPUs we need to fully exploit their 
-parallelism.
+This thesis is about applying embedded language techniques to programming 
+multi- and many-core computers. Computers today are becoming more and more  
+parallel. With each new generation of 
+general purpose processor (CPU) or graphics processor (GPU) we get more 
+parallel resources. To fully make use of modern CPUs and GPUs we need to fully 
+exploit their parallelism.
 
 Modern CPUs are parallel in a number of different ways. They discover instruction
-level parallelism(ILP), they have multiple processor cores and vector units for 
-Single Instruction Multiple Data parallelism. The exact configurations, number
-of cores and number of vector units, vary over available processors.
+level parallelism (ILP), they have multiple processor cores and vector units for 
+Single Instruction Multiple Data parallelism. The exact configuration, number
+of cores and number of vector units, vary over available processor models. 
+To obtain maximum performance, with a specific target processor in mind, the programmer
+would use low-level intrinsics in order to get SIMD parallelism and some threading library
+for parallelism over the available cores. One problem with this is that the resulting program 
+would be specialised to a specific processor model.
+
+In chapter~\ref{chap:ArBB},
 
@@
 
-\section{An embedded languages approach to data-parallelism}
+\section{Embedded languages}
+
+\section{Obsidian: An embedded language for GPU kernel implementation}
 
@@
 \noindent\emph{Hierarchies}
+
+\section{Embedding Intel Array Building Blocks}
+
+\noindent\emph{Short section about the approach taken with EmbArBB}
 
@@
 
-\chapter{EDSL implementation papers}
+\chapter{EDSL Implementation Papers}
 \label{chap:EDSLImplementation}
@@
 
-\chapter{GPU programming papers}
+\chapter{GPU Programming Papers}
 \label{chap:GPUProgramming}
@@
 
-\chapter{Retargetable parallel programming papers}
+\chapter{Retargetable Parallel Programming Papers}
 \label{chap:ArBB}

### 2013-10-10 src2/thesis.tex
apa

@@
 \newcites{arbb}{Bibliography}
-% \newcites{t}{Bibliography}
+\newcites{t}{Bibliography}
 %\newcites[ifl}{Bibliography}
@@
 
- \textbf{Keywords:} Keyword1, Keyword2
+ \textbf{Keywords:} Data-parallelism, GPU, Embedded languages
 
@@
 \vspace{5mm}
-This research has been funded by the Swedish Foundation for
+\noindent This research has been funded by the Swedish Foundation for
 Strategic Research (which funds the Resource Aware Functional 
@@
 To obtain maximum performance, with a specific target processor in mind, the programmer
-would use low-level intrinsics in order to get SIMD parallelism and some threading library
+would use low-level intrinsics in order to get SIMD parallelism and a threading library
 for parallelism over the available cores. One problem with this is that the resulting program 
-would be specialised to a specific processor model.
-
-In chapter~\ref{chap:ArBB},
-
-\noindent\emph{More text, lead into something about ArBB and forward ref to ArBB papers}
-
-\noindent\emph{Data-Parallelism, Structured-Compositional programming}
+would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
+system approached this problem by providing a high level programming interface~\citet{ARBB2011}
+. This interface is based on a set of parallel operations on arrays, such as map, folds 
+and scans. ArBB was implemented as an embedded language in C++ but a low-level C interface to 
+its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
+target the available CPU or Accelerator (Such as the Intel Larrabee). In 
+chapter~\ref{chap:ArBB}, there are papers describing work towards embedding Intel 
+ArBB in Haskell.
 
 Graphics Processors (GPUs), trade programmer convenience for increased parallelism. 
-For example, where CPUs dedicate much chip real estate for caches, branch prediction and 
+For example, where CPUs dedicate large portions of chip area for caches, branch prediction and 
 ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
@@
 
-\noindent\emph{More text, lead into something about compositional, high-level, but 
-expressive enough programming models for GPU programming and ref to Obsidian Chapter}
+Intel ArBB is part of a movement towards thinking of structure and patterns in 
+parallel programming. Other languages and libraries with similar aspirations 
+are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
+In the book Structured Parallel Programming by Michael McCool et al, a case is 
+made for thinking in terms of buidling blocks for parallel programming~\citet{STRUCTURED}. 
+They emphasize the importance of collective array operations such as map, fold, scans, 
+permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
+Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
+the patterns as abstractions, but most are possible to express in each of the languages. 
+The patterns, while very powerful as abstractions, also provide tools for thinking about 
+parallel algorithms.
 
-%\noindent\emph{GPUs}
+The patterns are powerful abstractions, but it is also important not to be locked into 
+a single implementation of each pattern. This could have negative performance impacts. 
+In ArBB for example the patterns used are specialised to the current hardware during JIT. 
+Languages like OpenCL and CUDA do not provide abstractions of these patterns. The programmer 
+needs to implement them using lower level features making CUDA and OpenCL code less composable 
+and harder to reuse. As an example, look at the slides from NVIDIA in reference~\citet{reduction} 
+where a reduction primitive is implemented in CUDA. First a naive reduction kernel is implement 
+and it is refined over many steps, taking many characteristics of the GPU into mind. What is 
+is that says that we do not need to redo this exploration with each new GPU generation 
+and discover a new optimal sollution?
 
-%\subsection{Data-parallel programming on GPUs}
+Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
+compared to CUDA and provide a more composable interface. But we also want to keep 
+the possible for the programmer to implement patterns in different ways. A balance 
+must be struck between providing enough low-level control to get at the performance 
+and still make building blocks implemented by the programmer reusable and composable. 
+We want to give more compositional tools for the programmer to experiment with while 
+exploring the space of various ways to implement a given pattern or primitive. 
+In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.
 
-%\subsection{Data-parallel programming on CPUs}
+% \section{The problem}
 
-\section{The problem}
-
-\emph{what is so interesting about Parallelism and GPUs}
+% \emph{what is so interesting about Parallelism and GPUs}
 
 \section{Embedded languages}
+
+This thesis applies embedded language methodology to data-parallel programming. An 
+embedded language is implemented as a library in some host language, Haskell in this case.
 
@@
 \section{Discussion}
+
+\bibliographystylet{alpha}
+\bibliographyt{thesis}
+\addcontentsline{toc}{section}{Bibliography}
 

### 2013-10-16 src2/bb/body.tex
apa
updated writing

@@
   Cond      :: Program BoolE 
+               -> Program () 
                -> Program ()
@@
 described in a separate paper \direct{\docname{} pulls out the paper
-  \citebb{axelsson2012generic} from a pile of papers}. It's hard to do
+  \citebb{SyntacticICFP12} from a pile of papers}. It's hard to do
 an apples to apples comparison between Syntactic and your

### 2013-10-16 src2/thesis.tex
updated writing

@@
 \usepackage{tikz,pgflibraryshapes}
-\usetikzlibrary{shapes,snakes,positioning}
+\usetikzlibrary{shapes,snakes,positioning,matrix}
 \usepackage{pgfplots}
 \usepackage{booktabs}
+\usepackage{enumerate}
 %\usepackage{etoolbox}
@@
 
-\newcommand{\thesistitle}{Embedded Languages for Data Parallel Computation}
+\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Computation}
 \newcommand{\dept}{Department of Computer Science and Engineering}
@@
 % ---------------------------------------------------------------------------
-\newcommand{\paperBB}{Simple and Compositional Reification of Monadic Embedded Languages}
-\newcommand{\paperIFL}{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
-\newcommand{\paperPAPP}{GPGPU Kernel Implementation and Refinement using Obsidian} 
-\newcommand{\paperEXPRESSIVE}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
-\newcommand{\paperCSORT}{Counting and Occurrence Sort for GPUs using an Embedded Language}
-\newcommand{\paperHL}{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
-\newcommand{\paperARBB}{Programming Future Parallel Architectures with Haskell and Intel ArBB}
-\newcommand{\paperEMBARBB}{Parallel Programming in Haskell Almost for Free}
+\newcommand{\paperA}{Paper A}
+\newcommand{\paperATitle}{Simple and Compositional Reification of Monadic Embedded Languages}
+\newcommand{\paperB}{Paper B}
+\newcommand{\paperBTitle}{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
+\newcommand{\paperC}{Paper C} 
+\newcommand{\paperCTitle}{GPGPU Kernel Implementation and Refinement using Obsidian} 
+\newcommand{\paperD}{Paper D}
+\newcommand{\paperDTitle}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+\newcommand{\paperE}{Paper E}
+\newcommand{\paperETitle}{Counting and Occurrence Sort for GPUs using an Embedded Language}
+\newcommand{\paperF}{Paper F}
+\newcommand{\paperFTitle}{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
+\newcommand{\paperG}{Paper G}
+\newcommand{\paperGTitle}{Programming Future Parallel Architectures with Haskell and Intel ArBB}
+\newcommand{\paperH}{Paper H}
+\newcommand{\paperHTitle}{Parallel Programming in Haskell Almost for Free}
 
@@
 
- \textbf{Keywords:} Data-parallelism, GPU, Embedded languages
+\textbf{Keywords:} Data-parallelism, Functional Programming, Embedded languages, GPU
 
@@
 % ---------------------------------------------------------------------------
-%  Guide for Oponent, grading committee
+%  Guide for Opponent, grading committee
 % ---------------------------------------------------------------------------
-\section*{About this thesis}
+\section*{Publications}
 
-This thesis contains an introduction and a selection of papers. The introduction 
-gives background and context to the papers.
+This thesis includes the following publications:
 
-The papers contained in this thesis falls in three categories. 
-\begin{itemize} 
-\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
-\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
-\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
-\end{itemize} 
-%In the following sections, each paper is given a short description. Hopefully 
-%there is enough information to wet the apetite.
+\begin{enumerate}[A.] 
+\item Josef Svenningsson and Bo Joel Svensson.
+Simple and Compositional Reification of Monadic Embedded Languages, 2013.
+\emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
+\item Bo Joel Svensson, Mary Sheeran and Koen Claessen.
+Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors.
+In \emph{Proc.
+of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
+\item Bo Joel Svensson, Koen Claessen and Mary Sheeran.
+GPGPU kernel implementation and refinement using Obisidan.
+\emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
+\item Koen Claessen, Mary Sheeran and Bo Joel Svensson.
+Expressive Array Constructs in an Embedded GPU Kernel Programming Language.
+In \emph{Proceedings of the 7th workshop of Declarative aspects and applications of multicore programming}, DAMP '12.
+ACM.
+\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran.
+Efficient Counting Sort Implementations using an Embedded GPU Programming Language.
+\emph{2nd Workshop of Functional High-Performance Computing, FHPC '13.}
+\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
+\item Bo Joel Svensson and Ryan Newton.
+Programming Future Parallel Architectures with Haskell and ArBB.
+\emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
+\item Bo Joel Svensson and Mary Sheeran.
+Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks.
+\emph{Proceedings of the 1st ACM SIGPLAN Workshop of Functional High-Performance computing, FHPC '12.}
+\end{enumerate}
 
-\subsection*{EDSL implementation papers}
+\clearpage
+\subsection*{Personal contributions}
 
-\subsubsection*{\paperBB}
+The following describes my personal contributions to the papers included 
+in this thesis.
 
-Authors: Josef Svenningsson and Bo Joel Svensson
+\subsubsection{Paper A}
+I intuitively applied the monad reification method as part of the implementation 
+of Obsidian. Josef Svenningsson identified its importance and suggested we write 
+about it. The paper has two distinct parts. Sections one and two describe the 
+method as I used it. And section four makes the compositional aspect of the method 
+explicit.
 
-\vspace{5mm}
+The code and examples in section one and two are my work. Section four we developed together 
+but is mostly Josef's expertise. For section four we also got very much and valuable help 
+from Emil Axelsson.
 
-\noindent We present a simple and compositional method for refication 
-of monads (or compilation of a monadic embedded language). I implemented 
-Obsidian using the method described here. The novelty and 
-usefulness of this simple way to reify monads was recognised by Josef
-and lead to the writing of this paper.
+For the rest of the paper, just substituting my name for Benny and Josef's for Bj\"orn
+will give a good indication of our contributions.
 
-\subsection*{GPU programming papers}
+\subsubsection{Papers B, C, D and E} 
+The Obsidian language has grown out of discussions between Mary Sheeran, Koen Claessen and me. 
+When it comes to implementation of the ideas we had in these discussions, I have been responsible. 
+Push arrays, first appearing in paper D, were invented by Koen Claessen. However, the 
+implementation of them in the setting of Obsidian was done by me. Paper D also have large 
+contributions by Mary Sheeran when it comes to examples and implementing them.
 
-\subsubsection{\paperIFL}
+Paper E, was the first I wrote with Josef Svenningsson who had been thinking about an interesting
+variation of counting sort and wanted to try implementing it. Together we made the required changes
+to Obsidian. Me and Josef are equal partners in this paper and got much valuable advice from 
+Mary Sheeran.
 
-Authors: Bo Joel Svensson, Mary Sheeran and Koen Claessen \newline
+\subsubsection{Papers G and H} 
+Here, again, most implementation effort was done by me. For paper G this work was performed 
+under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
+paper. Programming and running time experimentation was performed by me while getting much 
+valuable advice and guidance from Ryan.
 
-\vspace{5mm}
+The work behind paper H was done by me as a hobby project and my aspirations were not 
+high. I just wanted to experiment with some embedded languages techniques that we had 
+been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
+the right amount of pressure leading to us encompassing much more of ArBB functionality than 
+I had in mind. The paper's text is joint work; most examples as well as benchmarking is my work
+but the sparse matrix vector multiplication example is entirely Mary's.
 
-The first real paper about Obsidian, our embedded language for GPU 
-programming. In this paper we identify similarities between 
-conenction patterns in hardware and parallel computations on GPUs. 
-We introduce a structured and compositional way to write data-parallel 
-programs on a GPU.
-
-Basic parallel building blocks are implemented in a monadic style and 
-composed using a sequential composition operator {\tt ->-}.
-
-\subsubsection{\paperPAPP}
-
-Authors: Bo Joel Svensson, Koen Claessen and Mary Sheeran \newline
-
-\vspace{5mm}
-
-\emph{a new program representation, based vaguely on Arrows}
-
-\subsubsection{\paperEXPRESSIVE}
-
-Authors: Koen Claessen, Mary Sheeran and  Bo Joel Svensson \newline
-
-\vspace{5mm}
-
-This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
-array representation target specific performance problems that we want 
-to solve with Obsidian.
-
-\subsubsection{\paperCSORT}
-
-Authors: Josef Svenningsson, Bo Joel Svensson and Mary Sheeran \newline
-
-\vspace{5mm}
-
-\subsubsection{\paperHL}
-
-Authors: Bo Joel Svensson and Mary Sheeran \newline
-\noindent \emph{This work has not been published.}
-\vspace{5mm}
-
-\noindent
-
-In this paper the most recent additions and improvements of Obsidian are 
-described.
-
-\subsubsection{Retargetable parallel programming}
-
-\subsubsection{\paperARBB}
-
-Authors: Bo Joel Svensson and Ryan R. Newton
-
-\vspace{5mm}
-
-\noindent This position paper describes the work I did while on a 3 month internship at 
-Intel. The main part of this work was implementation of Haskell bindings 
-for the now retired Intel ArBB system.
-
-The paper also describes work we did towards implementing a backend for the 
-Accelerate system using ArBB. We reached a proof-of-concept implementation 
-capable of running some simple Accelerate programs using the ArBB backend.
-However, much work would remain to be able to run the full scope of Accelerate 
-programs on ArBB.
-
-\subsubsection{\paperEMBARBB}
-
-Authors: Bo Joel Svensson and Mary Sheeran
-
-\vspace{5mm}
-
-\noindent Here we explore another way to provide the capabilities of ArBB to the 
-Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
-we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
-Simply put, this means providing a more Haskell-programmer-friendly interface compared
-to the raw ArBB bindings (which are a collection of functions in the IO monad).
-
-\subsection*{Guide}
+% \subsection*{Guide}
 
@@
 general purpose processor (CPU) or graphics processor (GPU) we get more 
-parallel resources. To fully make use of modern CPUs and GPUs we need to fully 
+parallel resources. To fully make use of modern CPUs and GPUs we need to
 exploit their parallelism.
@@
 level parallelism (ILP), they have multiple processor cores and vector units for 
-Single Instruction Multiple Data parallelism. The exact configuration, number
-of cores and number of vector units, vary over available processor models. 
+Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
+of cores and number of vector units, varies over available processor models. 
 To obtain maximum performance, with a specific target processor in mind, the programmer
-would use low-level intrinsics in order to get SIMD parallelism and a threading library
+can use low-level intrinsics in order to get SIMD parallelism and a threading library
 for parallelism over the available cores. One problem with this is that the resulting program 
 would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
-system approached this problem by providing a high level programming interface~\citet{ARBB2011}
-. This interface is based on a set of parallel operations on arrays, such as map, folds 
+system approached this problem by providing a high level programming interface~\citet{ARBB2011}.
+This interface is based on a set of parallel operations on arrays, such as map, folds 
 and scans. ArBB was implemented as an embedded language in C++ but a low-level C interface to 
 its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
-target the available CPU or Accelerator (Such as the Intel Larrabee). In 
+target the available CPU or accelerator (Such as the Intel Larrabee). In 
 chapter~\ref{chap:ArBB}, there are papers describing work towards embedding Intel 
@@
 
-Graphics Processors (GPUs), trade programmer convenience for increased parallelism. 
+Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
 For example, where CPUs dedicate large portions of chip area for caches, branch prediction and 
@@
 are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
-In the book Structured Parallel Programming by Michael McCool et al, a case is 
-made for thinking in terms of buidling blocks for parallel programming~\citet{STRUCTURED}. 
-They emphasize the importance of collective array operations such as map, fold, scans, 
+In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
+made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
+They emphasise the importance of collective array operations such as map, fold, scans, 
 permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
@@
 a single implementation of each pattern. This could have negative performance impacts. 
-In ArBB for example the patterns used are specialised to the current hardware during JIT. 
+In ArBB, for example, the patterns used are specialised to the current hardware during JIT. 
 Languages like OpenCL and CUDA do not provide abstractions of these patterns. The programmer 
@@
 is that says that we do not need to redo this exploration with each new GPU generation 
-and discover a new optimal sollution?
+and discover a new optimal solution?
 
@@
 compared to CUDA and provide a more composable interface. But we also want to keep 
-the possible for the programmer to implement patterns in different ways. A balance 
+the possibility for the programmer to implement patterns in different ways. A balance 
 must be struck between providing enough low-level control to get at the performance 
@@
 This thesis applies embedded language methodology to data-parallel programming. An 
-embedded language is implemented as a library in some host language, Haskell in this case.
+embedded language is implemented as a library in a host language, Haskell in this case. 
+The embedded language approach is beneficial since we are not entirely sure how 
+to program current and future highly parallel computers. Being embedded allows 
+us more rapid prototyping of ideas. Other benefits of being embedded often listed 
+are: making use of the host language's type system, parser and libraries (to some extent) 
+and to have the host language as a power macro language for your embedded language.
 
+A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
+by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embedd a compiler 
+backend and generate efficient code in some target language. One key aspect of this 
+approach is that the embedded language is a library of functions that creates and 
+composes Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
+well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
+embedding that builds ASTs is called a deep embedding. Shallow embeddings on the 
+other hand do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
+Josef Svenningsson and Emil Axelsson combines the two methods, shallow and deep, 
+in order to simplify the AST data type (less constructors) and make more extensible 
+embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
+and deep embeddings while EmbArBB is a more traditional deep embedding.
+
+The embedded language approach has become popular to use as a way of raising the 
+level of abstraction in fields where the default is to use low-level languages. 
+For example, there is Lava and Wired in the fields of hardware design and 
+verification and Feldspar for digital signal processing\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
+programming there are numerous embedded approaches using various host languages. 
+But to mention two that use Haskell we have Accelerate and Nikola 
+\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
+goal of being parallel across platforms \citet{ARBB2011}.
+
+\section{Wider context}
+\FloatBarrier
+
+In figure~\ref{fig:researchmatrix} I place our work in relation to other languages,
+embedded languages and libraries for parallel programming. The systems are roughly divided 
+into three groups based on their level of abstraction. There are low-level languages, here 
+I place languages that are imperative and C-like. In the middle layer I place languages 
+that go a bit further, they have higher level abstractions and are more easily composable. 
+In the highest level I place languages that completely abstract away from details of the 
+hardware they run on. The division is not free from personal bias and the placement of languages
+in boxes was not always easy.
+
+The figure also divides the languages according to what hardware they support. There 
+are CPU specific languages, GPU specific languages and languages that support or aspire 
+to support CPU,GPU and accelerator (Larrabee) execution.
+
+With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
+CUDA the programmer has full controll of how to divide the computation amongst threads and 
+blocks. While with Accelerate, Nikola and Thrust the programmer would use high-level patterns
+without direct insight into how the application would be mapped onto threads and blocks of the 
+GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
+but give the programmer more control of what the computation will look like on the GPU. This 
+is done by allowing the programmer to specify and compose thread-level and block-level code. 
+Recently the CUB library arrived that has a similar idea, that GPU programmers need to 
+have block/thread level control to get maximum performance, while maintaining a higher level 
+and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
+we were alone at targetting this level.
+
+The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
+compose to build his application. I place the ArBB/EmbArBB box slightly lower than 
+Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
+they are not general primitives. Where Accelerate have a single higher order reduction 
+operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. I will go into 
+what benefits there are to having ArBB embedded in Haskell compared to C++ in 
+section~\ref{sec:EmbArBB}.
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[align=center, scale=2.5]
+% \draw[help lines] (0,0) grid (3,3);
+
+\draw[->] (-0.5,0) -- (-0.5,3) node[anchor=east] {Abstraction level};
+
+\draw (0,3) node[anchor=south west] {CPUs};
+\draw (1,3) node[anchor=south west] {GPUs};
+%\draw (2,3) node[anchor=south west] {Accelerator};
+\draw (2,3) node[anchor=south west] {CPUs, GPUs\\ Accelerators};
+
+\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
+\node[anchor=west] (acc) at (1,1.7) {CUB};
+\node[anchor=west] (repa) at (0,2.6) {DPH\\Nesl\\Repa\\Meta-repa};
+\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
+\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
+
+\node[anchor=west] (cuda) at (1,0.3) {CUDA};
+\node[anchor=west] (opencl) at (2,0.3) {OpenCL};
+
+\node[anchor=west] (OpenMP) at (0,0.3) {OpenMP\\PThreads};
+
+\foreach \x in {0,...,2} { 
+  \foreach \y in {0,...,2} { 
+    \draw (\x,\y) rectangle (\x+1,\y+1);
+  }
+}
+
+\draw[red!75,rounded corners,ultra thick] (1,0.4) rectangle (2,2.4);
+\node[red!75,rotate=45,thick] (obs) at (1.5,1.5) {Obsidian};
+
+\draw[orange!90,rounded corners,ultra thick] (2,1.4) rectangle (3,2.6);
+\node[orange!90,thick] (arbb) at (2.5,2) {EmbArBB\\ArBB};
+
+\end{tikzpicture}
+\end{center}
+\caption{Placing our work in the landscape of languages and libraries for parallel 
+general purpose programming of CPUs, GPUs or both. }
+\label{fig:researchmatrix} 
+\end{figure}
+
+\FloatBarrier
 \section{Obsidian: An embedded language for GPU kernel implementation}
 
-\noindent\emph{Obsidian, Embedded languages}
+Obsidian is aspiring to raise the level of abstraction of GPU kernel implementation 
+while maintaining enough control over details that influence performance on a GPU. 
+We try to reach these goals by offering abstractions to the programmer that are 
+compositional and parallel. We also ensure that these can easily be compiled 
+to NVIDIA CUDA code. We have two different array representations, each with 
+its strengths and weaknesses. We also feel that it is important that the programmer 
+is able to use local shared memory in computations and has the ability to influence 
+memory access patterns. What these array representations and abstractions are 
+will be explained in the following sections with references to the papers 
+that introduce them or explains them the best.
 
-\noindent\emph{Array representations}
+\subsection{Deep embedding: The {\tt Program} data type}
 
-\noindent\emph{Compositionality}
+Obsidian uses a deeply embedded {\tt Program} data type that represents GPU Kernels. 
+A CUDA GPU has a hierarchy of parallel resources. At the bottom there are threads, each 
+executing sequential programs. There are groups of threads (of 32) called {\em Warps} that 
+execute in lock-step. Warps are the scheduled unit of work on a GPU. There are {\em Blocks} 
+of threads, a set of threads that run as a group and share local (shared) memory. And 
+lastly there is a {\em Grid} of blocks that specifies the total number of threads involved 
+in a computation (Number\_of\_Blocks * Number\_of\_Threads). The deeply embedded program 
+data type of Obsidian models this hierarchy by parameterising programs on a hierarchy 
+level type parameter (the  {\tt t} parameter below).
 
-\noindent\emph{Hierarchies}
+\begin{verbatim} 
+data Program t a where
+\end{verbatim}
+
+The {\tt t} parameter can be either {\tt Thread}, {\tt Warp}, {\tt Block} or {\tt Grid}.
+These types are related to each other via a {\tt Step} type constructor that represents 
+going upward one step in the hierarchy.
+
+\begin{verbatim} 
+data Step a -- A step in the hierarchy
+data Zero
+\end{verbatim}
+
+The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
+{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
+in the hierarchy leading to warp programming being a special case.
+
+As an example of how the hierarchy level parameter is used I show the {\tt ForAll} constructor 
+from the {\tt Program} data type.  This constructor represents parallelism either over threads 
+or blocks.
+
+\begin{verbatim}
+ForAll :: EWord32 
+            -> (EWord32 -> Program t ())
+            -> Program (Step t) ()
+\end{verbatim}
+
+{\tt ForAll} takes a number of parallel iterations, a body representation by higher order 
+syntax. The body is a {\tt Program} at a some level {\tt t} while the resulting program 
+is at a level step above. This means a thread program can be turned into a block program 
+by using{\tt ForAll} or that a block program can be turned into a grid program.
+
+Information about the {\tt Program} data type can be found in paper F, \paperFTitle, in 
+section \ref{sec:paperF}.
+
+%\noindent\emph{Obsidian, Embedded languages}
+
+\subsection{Scalar language}
+
+\subsection{Array representation}
+
+\subsubsection{Pull arrays}
+
+\subsubsection{Push arrays}
+
+%\subsubsection{Fusion for free}
+
+\subsection{Compositionality}
+
+\subsection{Results}
+
+\subsection{Examples} 
+\noindent\emph{Example program, use something from latest paper}
+
+\subsection{Compilation to CUDA}
 
@@
 
+\subsection{Background} 
+\noindent\emph{What does ArBB do, what are the operations and datastructures it support}
+
+\noindent\emph{What does EmbArBB do}
+
+\subsection{EmbArBB: Embedding ArBB in Haskell}
+\label{sec:EmbArBB}
+
 \noindent\emph{Short section about the approach taken with EmbArBB}
 
-\section{The results}
+\subsection{The results}
 
@@
 
-\subsection{Related work}
+\section{Related work}
 
-\noindent\emph{GPU Programmming languages}
+\subsection{Parallelism}
 
-\noindent\emph{data-parallel programming languages}
+\noindent\emph{DPH}\citet{DPH} \newline
+\noindent\emph{D.A.Accelerate}\citet{ACCELERATEDAMP11} \newline
+\noindent\emph{Nikola}\citet{NIKOLA}\newline
+\noindent\emph{Repa}\citet{REPA}\newline
+\noindent\emph{Meta-Repa}\citet{METAREPA}\newline
+\noindent\emph{ArBB}\citet{ARBB2011}\newline
+\noindent\emph{OpenCL}\citet{OpenCL}\newline
+\noindent\emph{CUDA}\citet{CUDA}\newline
+\noindent\emph{Microsoft Accelerator}\citet{ACCELERATOR}\newline
+\noindent\emph{Copperhead}\citet{copperhead}\newline
+\noindent\emph{Delite}\citet{DELITE}\newline
+\noindent\emph{A monad for deterministic parallelism (R. Newton)} \citet{MonadPar} \newline
+\noindent\emph{Nesl}\citet{NESL} \newline
 
-\noindent\emph{Embedded languages}
+\subsection{Embedded domain specific languages}
 
-\noindent\emph{Monad reification methods}
+\noindent\emph{Feldspar}\citet{FELDSPAR2010} \newline
+\noindent\emph{Lava}\citet{lavaICFP} \newline
+\noindent\emph{Vertigo}\citet{VERTIGO}  \newline
+
+\subsection{Compiling Embedded Languages}
+
+\noindent\emph{Generic monad constructs (josef + emil + anders}\citet{Generic} \newline
+\noindent\emph{The Constrained Monad Problem (Gill)}\citet{sculthorpe2013constrained} \newline
+\noindent\emph{Compiling Embedded Languages}\citet{COMPILEEDSL} \newline
 
@@
 
+\noindent\emph{What contributions are made by Obsidian}
+
+\noindent\emph{What works well with Obsidian}
+
+\noindent\emph{EmbArBB contributions}
+
+% ---------------------------------------------------------------------------
+% Guide 
+\section{Thesis overview}
+
+%This thesis contains an introduction and a selection of papers. The introduction 
+%gives background and context to the papers.
+
+The papers contained in this thesis fall in three categories. 
+\begin{itemize} 
+\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
+\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
+\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
+\end{itemize} 
+%In the following sections, each paper is given a short description. Hopefully 
+%there is enough information to wet the appetite.
+
+\subsection{EDSL implementation papers}
+
+\subsubsection{\paperA: \\ \paperATitle}
+
+Authors: Josef Svenningsson and Bo Joel Svensson
+
+\vspace{5mm}
+
+\noindent This paper describes a simple and compositional method for 
+reification of monads. This is useful for compilation of 
+monadic embedded languages. We use a simple robot control language 
+as the example. Robot programs can be expressed using Haskell do notation 
+and is compiled down to a simple first order program representation. 
+There is also a graphical simulator available that runs the compiled 
+code. Robot language and simulator is available at github: \url{github.com/svenssonjoel/Robot}.
+
+\subsection{GPU programming papers}
+
+\subsubsection{\paperB: \\ \paperBTitle}
+
+Authors: Bo Joel Svensson, Mary Sheeran and Koen Claessen \newline
+
+\vspace{5mm}
+
+The first real paper about Obsidian, our embedded language for GPU 
+programming. In this paper we identify similarities between 
+connection patterns in hardware and parallel computations on GPUs. 
+We introduce a structured and compositional way to write data-parallel 
+programs on a GPU.
+
+Basic parallel building blocks are implemented in a monadic style and 
+composed using a sequential composition operator {\tt ->-}.
+
+\subsubsection{\paperC: \\ \paperCTitle}
+
+Authors: Bo Joel Svensson, Koen Claessen and Mary Sheeran \newline
+
+\vspace{5mm}
+
+In this paper we describe a version of Obsidian implemented with a 
+a different internal representation of GPU programs. We realised that 
+most of our kernels have similar shape, they are parallel operations 
+composed in sequence interspersed with barrier synchronisations. The 
+internal representation used here captures this case exactly. Programming 
+is now done is a style resembling arrows.
+
+\subsubsection{\paperD: \\ \paperDTitle}
+
+Authors: Koen Claessen, Mary Sheeran and  Bo Joel Svensson \newline
+
+\vspace{5mm}
+
+This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
+array representation target specific performance problems that we want 
+to solve with Obsidian. Here we have also gone back to using monadic 
+style for programming in Obsidian.
+
+\subsubsection{\paperE: \\ \paperETitle}
+
+Authors: Josef Svenningsson, Bo Joel Svensson and Mary Sheeran \newline
+
+\vspace{5mm}
+
+This paper contains a sorting case study performed in Obsidian. It 
+is however the algorithms that are in focus here. We describe 
+an interesting variation on Counting sort that removes duplicate 
+elements and is a nice fit for GPUs because of its little need for 
+synchronisation.
+
+In order to implement these sorting algorithms we added atomic operations 
+to Obsidian.
+
+\subsubsection{\paperF: \\ \paperFTitle}
+
+Authors: Bo Joel Svensson and Mary Sheeran \newline
+\noindent \emph{This work has not yet been published.}
+\vspace{5mm}
+
+\noindent
+
+In this paper the most recent additions and improvements of Obsidian are 
+described. Obsidian Programs are now parameterised on Hierarchy level. This forces 
+the programmer into writing programs that we know how to compile to a GPU easily 
+(by limiting how parallel operations can be nested). The paper also gives a detailed  
+optimisation story using Obsidian for implementing reduction kernels.
+
+\subsubsection{Retargetable parallel programming}
+
+\subsubsection{\paperG: \\ \paperGTitle}
+
+Authors: Bo Joel Svensson and Ryan R. Newton
+
+\vspace{5mm}
+
+\noindent This position paper describes the work I did while on a 3 month internship at 
+Intel. The main part of this work was implementation of Haskell bindings 
+for the now retired Intel ArBB system.
+
+The paper also describes work we did towards implementing a backend for the 
+Accelerate system using ArBB. We reached a proof-of-concept implementation 
+capable of running some simple Accelerate programs using the ArBB backend.
+However, much work would remain to be able to run the full scope of Accelerate 
+programs on ArBB.
+
+\subsubsection{\paperH: \\ \paperHTitle}
+
+Authors: Bo Joel Svensson and Mary Sheeran
+
+\vspace{5mm}
+
+\noindent Here we explore another way to provide the capabilities of ArBB to the 
+Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
+we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
+Simply put, this means providing a more Haskell-programmer-friendly interface compared
+to the raw ArBB bindings (which are a collection of functions in the IO monad).
+
 \bibliographystylet{alpha}
 \bibliographyt{thesis}
-\addcontentsline{toc}{section}{Bibliography}
+%\addcontentsline{toc}{section}{Bibliography}
 
@@
 
-\section{\paperBB}
-\label{sec:paperBB}
+\section[\paperATitle]{\paperA: \\ \paperATitle}
+\label{sec:paperA}
 %\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
+
+% \paperATitle
 
@@
 
-\section{\paperIFL}
-\label{sec:paperIFL}
+\section[\paperBTitle]{\paperB: \\ \paperBTitle}
+\label{sec:paperB}
 %\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
 %Parallel Programming of Graphics Processors}
+
+%\paperBTitle
 
@@
 
-\section{\paperPAPP}
-\label{sec:paperPAPP}
+\section[\paperCTitle]{\paperC: \\ \paperCTitle}
+\label{sec:paperC}
 %\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}
+
+%\paperCTitle
 
@@
 
-\section{\paperEXPRESSIVE}
-\label{sec:paperEXPRESSIVE}
+\section[\paperDTitle]{\paperD: \\ \paperDTitle}
+\label{sec:paperD}
 %\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
+
+%\paperDTitle
 
@@
 
-\section{\paperCSORT}
-\label{sec:paperCSORT}
+\section[\paperETitle]{\paperE: \\ \paperETitle}
+\label{sec:paperE}
 %\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}
+
+% \paperETitle
 
@@
 
-\section{\paperHL}
-\label{sec:paperHL}
+\section[\paperFTitle]{\paperF: \\ \paperFTitle}
+\label{sec:paperF}
 %\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
+
+% \paperFTitle
 
@@
 
-\section{\paperARBB}
-\label{sec:paperARBB}
+\section[\paperGTitle]{\paperG: \\ \paperGTitle}
+\label{sec:paperG}
 %\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
+
+% \paperGTitle
 
@@
 
-\section{\paperEMBARBB}
-\label{sec:paperEMBARBB}
+\section[\paperHTitle]{\paperH: \\ \paperHTitle}
+\label{sec:paperH}
 %\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
+
+% \paperHTitle
 

### 2013-10-17 src2/thesis.tex
apa

@@
 Paper E, was the first I wrote with Josef Svenningsson who had been thinking about an interesting
-variation of counting sort and wanted to try implementing it. Together we made the required changes
-to Obsidian. Me and Josef are equal partners in this paper and got much valuable advice from 
-Mary Sheeran.
+variation of counting sort and wanted to implement it. Together we added atomic operations to 
+the deep embedding and I modified the Obsidian code generator to accomodate them. Sections 
+1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
+the benchmarking. Section 8 is joint work.
+%Me and Josef are equal partners in this paper and got much valuable advice from 
+%Mary Sheeran.
 
@@
 but the sparse matrix vector multiplication example is entirely Mary's.
+
+%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a fuctioning 
+%code generating backend u
 
@@
 a single implementation of each pattern. This could have negative performance impacts. 
-In ArBB, for example, the patterns used are specialised to the current hardware during JIT. 
-Languages like OpenCL and CUDA do not provide abstractions of these patterns. The programmer 
-needs to implement them using lower level features making CUDA and OpenCL code less composable 
-and harder to reuse. As an example, look at the slides from NVIDIA in reference~\citet{reduction} 
-where a reduction primitive is implemented in CUDA. First a naive reduction kernel is implement 
-and it is refined over many steps, taking many characteristics of the GPU into mind. What is 
-is that says that we do not need to redo this exploration with each new GPU generation 
-and discover a new optimal solution?
+In ArBB, for example, the patterns used are specialised to the current hardware during JIT 
+compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
+The programmer needs to implement them using lower level features making CUDA and OpenCL 
+code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
+reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
+naive reduction kernel is implement and it is refined over many steps, taking many 
+characteristics of the GPU into mind. The end result is a reduction kernel that performs 
+very well on the chosen GPU. But as GPUs evolve the design space may need to be explored again. 
+%What is 
+%is that says that we do not need to redo this exploration with each new GPU generation 
+%and discover a new optimal solution?
 
@@
 
-%\noindent\emph{Obsidian, Embedded languages}
-
 \subsection{Scalar language}
@@
 \subsubsection{Pull arrays}
+\emph{fusion for free}
 
 \subsubsection{Push arrays}
-
-%\subsubsection{Fusion for free}
+\emph{fusion for free}
 
@@
 
-\subsection{Background} 
-\noindent\emph{What does ArBB do, what are the operations and datastructures it support}
+Intel ArBB~\citet{ARBB2011} is a system for high-level data-parallel programming with the 
+ability to generate code for a variety of different hardware confgurations. It is implemented 
+as an embedded language in C++. One motivation for ArBB that is mentioned in~\citet{ARBB2011} is 
+that while high-performance computing specialist are highly competent at implementing 
+kernels using low-level programming techniques, mainstream developers are not. ArBB offers 
+a more composable way to write programs that make use of core and vector (SIMD) parallelism and 
+doing this while using a familiar language, C++.
 
-\noindent\emph{What does EmbArBB do}
+One of the greatest strengths of ArBB, as I see it, is that it also comes with a low-level C 
+interface. The main purpose of this C interface is to make it easier to use the ArBB system 
+from other languages. Many languages have foreign function interfaces that are geared towards C.
+This gives ArBB a level of language independence on top of its cross platform abilities.
+
+ArBB is based on a set of parallel primitives (or structures) on dense 1,2 and 3 dimensional 
+arrays and nested arrays: 
+\begin{itemize} 
+\item {\bf Reductions}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
+\item {\bf Scans}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
+\item {\bf Sorting}      {\tt sort} 
+\item {\bf Permutations} {\tt gather}, {\tt scatter}, {\tt reverse} ...
+\item {\bf Sequential controlflow} {\tt \_for}, {\tt \_while} 
+\item {\bf Map, zipWith, stencil}  {\tt map} 
+\end{itemize}
+
+These operations (and many more, there are about 120 operations in total) can be used 
+by the programmer when writing programs using ArBB. Note that there are sequential control 
+operations with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
+language and calling the ArBB functions does not actually execute them, it just ads a node to 
+some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
+ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
+This is exactly the way these things work in Haskell but it may be even more surprising to a
+C++ programmer. Before an ArBB function can be computed it needs to be captured. For this 
+ArBB has a {\tt call} function that is used to run functions using ArBB functionality. The 
+first time {\tt call} is used on a function pointer that function is executed and the ArBB 
+functions used in it builds an AST (we say the function has been captured). The AST is then 
+compiled and specialised for the hardware available. Any subsequent calls of a function using 
+ArBB functionality does not lead to a recompilation, the first call caches this code.
+
+The first step of our work with embedding Intel ArBB functionality was to implement very direct 
+and low-level bindings to the C interface. This provides a Haskell function (in the IO Monad) 
+for each of the functions in the ArBB C interface. More details and results of this work 
+can be found in paper G in section~\ref{sec:paperG}.
+
+%\noindent\emph{What does ArBB do, what are the operations and datastructures it support}
+
+%\noindent\emph{What does EmbArBB do}
 
@@
 
+The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
+is not useful for application implementation. A higher-level interface to ArBB to offer 
+the Haskell programmer is needed. Our first attempt was to implement an Accelerate backend 
+using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
+more general language and there was an API missmatch between them. These problems 
+are mentioned in in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
+to the Haskell in a way that is useful and simpler to implement, we start working on EmbArBB.
+
+With EmbArBB we improve the interfacing with Haskell by implementing a deeply embedded language. 
+Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings creating 
+function objects that be run.
+
 \noindent\emph{Short section about the approach taken with EmbArBB}
 
-\subsection{The results}
+\emph{ How well did we do?}
 
-\emph{ How well did we do?}
+%\subsection{The results}
+
+\subsection{Current status of ArBB and EmbArBB}
+
+Since our work with ArBB, Intel has unfortunately retired the ArBB system. This 
+leaves EmbArBB without a functioning code generating backend.
 
@@
 
-\subsubsection{Retargetable parallel programming}
+\subsection{Retargetable parallel programming}
 
@@
 to the raw ArBB bindings (which are a collection of functions in the IO monad).
+
+\subsection{Contributions}
+
+This thesis contributes to the follwoing areas:
+
+\begin{itemize} 
+\item Embedded language design and implementation. 
+\item General Purpose programming using a GPU (GPGPU).
+\item High level data-parallel programming across devices.
+\end{itemize}
+
+\subsubsection{Embedded language design and implementation}
+
+\subsubsection{General Purpose programming using a GPU}
+
+\subsubsection{High level data-parallel programming across devices}
 

### 2013-10-20 src2/thesis.tex
small update

@@
 variation of counting sort and wanted to implement it. Together we added atomic operations to 
-the deep embedding and I modified the Obsidian code generator to accomodate them. Sections 
+Obsidian and I modified the code generator to accomodate them. Sections 
 1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
@@
 exploit their parallelism.
+
+While it is sometimes possible to automatically parallelise programs originally 
+written for serial execution, it is becoming clear that explicitly parallel 
+programming models are needed for making use of all parallel resources~\cite{STRUCTURED}. 
+This also means that it is becoming important to think in parallel from the beginning.
 
@@
 would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
-system approached this problem by providing a high level programming interface~\citet{ARBB2011}.
+system approach this problem by providing a high level programming interface~\citet{ARBB2011}.
 This interface is based on a set of parallel operations on arrays, such as map, folds 
@@
 Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings creating 
-function objects that be run.
-
-\noindent\emph{Short section about the approach taken with EmbArBB}
+function objects that can then be run.
+
+\subsubsection{Dense and nested arrays}
+
+ArBB supports one, two and three dimensional arrays, called dense vectors. In EmbArBB we 
+represent these with a data type, {\tt DVector}, parameterised by dimensionality and 
+element type. The dimension parameter can be either {\tt Dim1}, {\tt Dim2} or {\tt Dim3}. 
+There are also nested vectors, in EmbArBB called {\tt NVector}, parameterised on element type.
+
+\subsubsection{Deep embedding}
+
+\begin{verbatim}
+data Expr = Lit Literal
+          | Var Variable
+
+          | Index0 Expr 
+          | ResIndex Expr Int
+
+          | Map (R GenRecord) [Expr]  
+          | Call (R GenRecord) [Expr]
+
+          | While ([Expr] -> Expr)  ([Expr] -> [Expr])  [Expr]
+
+          | While2 ([Expr] -> Expr)  
+                   ([Expr] -> [Expr])  
+                   [Expr}
+
+          | If Expr Expr Expr
+
+          | Op Op [Expr]   
+\end{verbatim}
+
+\subsubsection{compilation}
+
+\emph{Sharing detection}
+
+\emph{let ArBB do the optimisations}
+
+\emph{having deep embedding means we could compile to something else}
+
+\subsubsection{Evaluation}
 
 \emph{ How well did we do?}
-
-%\subsection{The results}
 

### 2013-10-23 src2/thesis.tex
update

@@
 
-\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Computation}
+\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Programming}
 \newcommand{\dept}{Department of Computer Science and Engineering}
@@
 \item Bo Joel Svensson, Koen Claessen and Mary Sheeran.
-GPGPU kernel implementation and refinement using Obisidan.
+GPGPU kernel implementation and refinement using Obsidian.
 \emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
@@
 Push arrays, first appearing in paper D, were invented by Koen Claessen. However, the 
-implementation of them in the setting of Obsidian was done by me. Paper D also have large 
+implementation of them in the setting of Obsidian was done by me. Paper D also has large 
 contributions by Mary Sheeran when it comes to examples and implementing them.
 
-Paper E, was the first I wrote with Josef Svenningsson who had been thinking about an interesting
+Paper E was the first I wrote with Josef Svenningsson, who had been thinking about an interesting
 variation of counting sort and wanted to implement it. Together we added atomic operations to 
-Obsidian and I modified the code generator to accomodate them. Sections 
+Obsidian and I modified the code generator to accommodate them. Sections 
 1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
@@
 \subsubsection{Papers G and H} 
-Here, again, most implementation effort was done by me. For paper G this work was performed 
+Here, again, most implementation effort was done by me. For paper G, this work was performed 
 under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
-paper. Programming and running time experimentation was performed by me while getting much 
+paper. Programming and running time experimentation was performed by me, while getting much 
 valuable advice and guidance from Ryan.
@@
 been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
-the right amount of pressure leading to us encompassing much more of ArBB functionality than 
-I had in mind. The paper's text is joint work; most examples as well as benchmarking is my work
+the right amount of pressure, leading to us encompassing much more of ArBB functionality than 
+I had in mind. The paper's text is joint work; most examples as well as benchmarking are my work,
 but the sparse matrix vector multiplication example is entirely Mary's.
 
-%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a fuctioning 
+%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a functioning 
 %code generating backend u
@@
 parallel. With each new generation of 
-general purpose processor (CPU) or graphics processor (GPU) we get more 
-parallel resources. To fully make use of modern CPUs and GPUs we need to
+general purpose processor (CPU) or graphics processor (GPU), we get more 
+parallel resources. To fully make use of modern CPUs and GPUs, we need to
 exploit their parallelism.
@@
 programming models are needed for making use of all parallel resources~\cite{STRUCTURED}. 
-This also means that it is becoming important to think in parallel from the beginning.
+This also means that it is important to think in parallel from the beginning.
+
+In this thesis the focus is on data-parallelism, as opposed to task- or 
+control-parallelism. Data parallelism is applicable where the input data can 
+be divided into chunks and operated upon independently. \emph{more, and maybe not here}
 
@@
 would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
-system approach this problem by providing a high level programming interface~\citet{ARBB2011}.
+system approaches this problem by providing a high level programming interface~\citet{ARBB2011}.
 This interface is based on a set of parallel operations on arrays, such as map, folds 
-and scans. ArBB was implemented as an embedded language in C++ but a low-level C interface to 
+and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
 its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
-target the available CPU or accelerator (Such as the Intel Larrabee). In 
-chapter~\ref{chap:ArBB}, there are papers describing work towards embedding Intel 
-ArBB in Haskell.
+target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
+knights Ferry, Xeon Phi). In chapter~\ref{chap:ArBB}, there are papers describing work 
+towards embedding Intel ArBB in Haskell.
 
 Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
-For example, where CPUs dedicate large portions of chip area for caches, branch prediction and 
+For example, where CPUs dedicate large portions of chip area to caches, branch prediction and 
 ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
@@
 compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
-The programmer needs to implement them using lower level features making CUDA and OpenCL 
+The programmer needs to implement them using lower level features, making CUDA and OpenCL 
 code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
 reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
-naive reduction kernel is implement and it is refined over many steps, taking many 
-characteristics of the GPU into mind. The end result is a reduction kernel that performs 
-very well on the chosen GPU. But as GPUs evolve the design space may need to be explored again. 
+naive reduction pattern is implemented and then refined over many steps, taking many 
+characteristics of the GPU into mind. The end result is an implementation of the 
+reduction pattern that performs very well on the chosen GPU. But as GPUs evolve, the 
+design space may need to be explored again. 
 %What is 
@@
 must be struck between providing enough low-level control to get at the performance 
-and still make building blocks implemented by the programmer reusable and composable. 
+and still making building blocks implemented by the programmer reusable and composable. 
 We want to give more compositional tools for the programmer to experiment with while 
@@
 
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% Embedded languages
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 \section{Embedded languages}
@@
 are: making use of the host language's type system, parser and libraries (to some extent) 
-and to have the host language as a power macro language for your embedded language.
+and to have the host language as a powerful macro language for your embedded language.
 
@@
 backend and generate efficient code in some target language. One key aspect of this 
-approach is that the embedded language is a library of functions that creates and 
-composes Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
+approach is that the embedded language is a library of functions that create and 
+compose Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
 well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
@@
 other hand do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
-Josef Svenningsson and Emil Axelsson combines the two methods, shallow and deep, 
-in order to simplify the AST data type (less constructors) and make more extensible 
+Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
+in order to simplify the AST data type (fewer constructors) and make more extensible 
 embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
-and deep embeddings while EmbArBB is a more traditional deep embedding.
+and deep embeddings, while EmbArBB is a more traditional deep embedding.
 
@@
 For example, there is Lava and Wired in the fields of hardware design and 
-verification and Feldspar for digital signal processing\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
+verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
 programming there are numerous embedded approaches using various host languages. 
@@
 goal of being parallel across platforms \citet{ARBB2011}.
-
-\section{Wider context}
-\FloatBarrier
-
-In figure~\ref{fig:researchmatrix} I place our work in relation to other languages,
-embedded languages and libraries for parallel programming. The systems are roughly divided 
-into three groups based on their level of abstraction. There are low-level languages, here 
-I place languages that are imperative and C-like. In the middle layer I place languages 
-that go a bit further, they have higher level abstractions and are more easily composable. 
-In the highest level I place languages that completely abstract away from details of the 
-hardware they run on. The division is not free from personal bias and the placement of languages
-in boxes was not always easy.
-
-The figure also divides the languages according to what hardware they support. There 
-are CPU specific languages, GPU specific languages and languages that support or aspire 
-to support CPU,GPU and accelerator (Larrabee) execution.
-
-With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
-CUDA the programmer has full controll of how to divide the computation amongst threads and 
-blocks. While with Accelerate, Nikola and Thrust the programmer would use high-level patterns
-without direct insight into how the application would be mapped onto threads and blocks of the 
-GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
-but give the programmer more control of what the computation will look like on the GPU. This 
-is done by allowing the programmer to specify and compose thread-level and block-level code. 
-Recently the CUB library arrived that has a similar idea, that GPU programmers need to 
-have block/thread level control to get maximum performance, while maintaining a higher level 
-and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
-we were alone at targetting this level.
-
-The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
-compose to build his application. I place the ArBB/EmbArBB box slightly lower than 
-Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
-they are not general primitives. Where Accelerate have a single higher order reduction 
-operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. I will go into 
-what benefits there are to having ArBB embedded in Haskell compared to C++ in 
-section~\ref{sec:EmbArBB}.
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[align=center, scale=2.5]
-% \draw[help lines] (0,0) grid (3,3);
-
-\draw[->] (-0.5,0) -- (-0.5,3) node[anchor=east] {Abstraction level};
-
-\draw (0,3) node[anchor=south west] {CPUs};
-\draw (1,3) node[anchor=south west] {GPUs};
-%\draw (2,3) node[anchor=south west] {Accelerator};
-\draw (2,3) node[anchor=south west] {CPUs, GPUs\\ Accelerators};
-
-\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
-\node[anchor=west] (acc) at (1,1.7) {CUB};
-\node[anchor=west] (repa) at (0,2.6) {DPH\\Nesl\\Repa\\Meta-repa};
-\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
-\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
-
-\node[anchor=west] (cuda) at (1,0.3) {CUDA};
-\node[anchor=west] (opencl) at (2,0.3) {OpenCL};
-
-\node[anchor=west] (OpenMP) at (0,0.3) {OpenMP\\PThreads};
-
-\foreach \x in {0,...,2} { 
-  \foreach \y in {0,...,2} { 
-    \draw (\x,\y) rectangle (\x+1,\y+1);
-  }
-}
-
-\draw[red!75,rounded corners,ultra thick] (1,0.4) rectangle (2,2.4);
-\node[red!75,rotate=45,thick] (obs) at (1.5,1.5) {Obsidian};
-
-\draw[orange!90,rounded corners,ultra thick] (2,1.4) rectangle (3,2.6);
-\node[orange!90,thick] (arbb) at (2.5,2) {EmbArBB\\ArBB};
-
-\end{tikzpicture}
-\end{center}
-\caption{Placing our work in the landscape of languages and libraries for parallel 
-general purpose programming of CPUs, GPUs or both. }
-\label{fig:researchmatrix} 
-\end{figure}
-
-\FloatBarrier
+Related work and information about the tools, libraries and language that make the 
+wider context surrounding Obsidian is described in section~\ref{sec:relatedwork}.
+
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% Obsidian
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 \section{Obsidian: An embedded language for GPU kernel implementation}
 
-Obsidian is aspiring to raise the level of abstraction of GPU kernel implementation 
+Obsidian aspires to raise the level of abstraction of GPU kernel implementation 
 while maintaining enough control over details that influence performance on a GPU. 
@@
 its strengths and weaknesses. We also feel that it is important that the programmer 
-is able to use local shared memory in computations and has the ability to influence 
+be able to use local shared memory in computations and has the ability to influence 
 memory access patterns. What these array representations and abstractions are 
-will be explained in the following sections with references to the papers 
-that introduce them or explains them the best.
+will be explained in the following sections, with references to the papers 
+that introduce them or explain them the best.
 
@@
 {\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
-in the hierarchy leading to warp programming being a special case.
+in the hierarchy leading to warp programming being a special case. One benefit a warp has is 
+that threads can communicate via shared memory without using synchronisation primitives. There 
+is no programmer support for warp level programming in CUDA (as there is for block level 
+programs). The programmer needs to take care to ensure that the communication patterns are 
+within a warp and is then free to remove synchronisations. If warps where to be put in the 
+program level hierarchy of Obsidian, its place would be between the thread and block level. 
+This forces the programmer to go via warps when putting together a block computation even 
+if the communication pattern is not suited for that, leading to inconvenience with no gain.
 
@@
 
-{\tt ForAll} takes a number of parallel iterations, a body representation by higher order 
-syntax. The body is a {\tt Program} at a some level {\tt t} while the resulting program 
-is at a level step above. This means a thread program can be turned into a block program 
-by using{\tt ForAll} or that a block program can be turned into a grid program.
+{\tt ForAll} takes a number of parallel iterations, a body represented by a function 
+from an index to a program. The body is a {\tt Program} at a some level {\tt t} while the 
+resulting program is at a level step above. This means a thread program can be turned into 
+a block program by using {\tt ForAll} or that a block program can be turned into a grid program.
 
@@
 
-\subsection{Scalar language}
-
-\subsection{Array representation}
+\subsection{Scalars}
+
+Scalars and operations on scalars are represented by an expression data type ({\tt Exp a}) in 
+Obsidian. This is another example of a deep embedding; the expression data type contains 
+constructors for literal values, arithmetic operations, and conditionals.
+
+Haskell's type class system allows us to overload arithmetic operations such as {\tt (+)}, 
+{\tt (-)} and {\tt (*)} by making expressions an instance of {\tt Num}. This allows the 
+Obsidian arithmetic expressions to look exactly like corresponding native Haskell arithmetic.
+
+The approach taken to embedd the scalar language is very similar to what is described 
+in~\citet{COMPILEEDSL}; but we use a GADT to obtained typed expressions rather than phantom types.
+
+\subsection{Arrays}
+
+Obsidian has two array representations. These array representations are implemented as 
+shallow embeddings on top of the expression and program data types. This means that these 
+arrays will disappear during Haskell evaluation of the Obsidian program.
 
 \subsubsection{Pull arrays}
-\emph{fusion for free}
+
+Pull arrays have been part of Obsidian from the very beginning~\citet{JMT} but called either 
+{\tt Array} or just {\tt Arr}. A pull array represents an array as an indexing function; given 
+an index it provides an element:
+
+\begin{verbatim} 
+data Pull s a = Pull {pullLen :: s, 
+                      pullFun :: EWord32 -> a}
+\end{verbatim}
+
+Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
+parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
+or dynamic (length represented by an expression, an unknown length). When creating block and 
+warp level computations we want to know the exact lengths in order to avoid generating code 
+riddled with conditionals concerning array lengths. Conditionals are problematic on GPUs and 
+while general programs that run for any size is possible when performance is crucial a fixed 
+size is recommended.
+
+One benefit of pull arrays are that they give fusion of all operations on them for free. for example {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:
+
+\begin{verbatim} 
+map   f (Pull n ixf) = Pull n (f . ixf)
+\end{verbatim}
+
+Now, we see that {\tt map f . map g} is the same as {\tt map f . g} since both result in {\tt f}
+and {\tt g} becoming composed onto the indexing function. No intermediate array is built in 
+memory.
+
+Another benefit of pull arrays are that they are parallel. Simply put, early Obsidian was pull 
+arrays in addition to a way to compute the array and store its elements to memory 
+(called {\tt sync} in early versions of Obsidian and later {\tt force}). A very 
+direct way to compile a pull array to a GPU is to launch as many threads as there are 
+elements in the array and apply the indexing function to the thread id. Now each thread 
+will have computed an element of the array and can store that to memory.
 
 \subsubsection{Push arrays}
-\emph{fusion for free}
-
-\subsection{Compositionality}
+
+Push arrays where added to Obsidian as a complement to pull arrays. Pull arrays are 
+good because they are so simple to understand and easy to parallelise. However certain
+operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
+or interleaving of pull arrays result in conditionals in the indexing function. The 
+interleaving case, which is worst, leads to diverging branching in every warp. If branches 
+diverge the GPU will execute turn off the processing elements taking one path and 
+allow the others to progress; then turning to the other branch. The computation of the 
+branches has become serialised and compute resources are wasted.
+
+\begin{verbatim}
+data Push p s a =
+  Push s ((a -> EWord32 -> TProgram ()) -> Program p ())
+\end{verbatim}
 
@@
 
-\subsection{Examples} 
-\noindent\emph{Example program, use something from latest paper}
+\emph{point to results in each paper}
+
+\emph{provide information around the results in papers}
 
@@
 
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% ArBB work
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 \section{Embedding Intel Array Building Blocks}
@@
 Intel ArBB~\citet{ARBB2011} is a system for high-level data-parallel programming with the 
-ability to generate code for a variety of different hardware confgurations. It is implemented 
+ability to generate code for a variety of different hardware configurations. It is implemented 
 as an embedded language in C++. One motivation for ArBB that is mentioned in~\citet{ARBB2011} is 
-that while high-performance computing specialist are highly competent at implementing 
+that while high-performance computing specialists are highly competent at implementing 
 kernels using low-level programming techniques, mainstream developers are not. ArBB offers 
@@
 
-ArBB is based on a set of parallel primitives (or structures) on dense 1,2 and 3 dimensional 
-arrays and nested arrays: 
+ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
 \begin{itemize} 
-\item {\bf Reductions}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
-\item {\bf Scans}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
-\item {\bf Sorting}      {\tt sort} 
-\item {\bf Permutations} {\tt gather}, {\tt scatter}, {\tt reverse} ...
-\item {\bf Sequential controlflow} {\tt \_for}, {\tt \_while} 
-\item {\bf Map, zipWith, stencil}  {\tt map} 
+\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
+\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
+\item {\bf Sorting:}      {\tt sort} 
+\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
+\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
+\item {\bf Map, zipWith, stencil:}  {\tt map} 
 \end{itemize}
@@
 These operations (and many more, there are about 120 operations in total) can be used 
-by the programmer when writing programs using ArBB. Note that there are sequential control 
-operations with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
+by the programmer when writing programs using ArBB. Note that there are sequential looping 
+constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
 language and calling the ArBB functions does not actually execute them, it just ads a node to 
@@
 ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
-This is exactly the way these things work in Haskell but it may be even more surprising to a
-C++ programmer. Before an ArBB function can be computed it needs to be captured. For this 
-ArBB has a {\tt call} function that is used to run functions using ArBB functionality. The 
-first time {\tt call} is used on a function pointer that function is executed and the ArBB 
-functions used in it builds an AST (we say the function has been captured). The AST is then 
-compiled and specialised for the hardware available. Any subsequent calls of a function using 
-ArBB functionality does not lead to a recompilation, the first call caches this code.
+This is exactly the way these things work in Haskell embedded languages, but it may be even 
+more surprising to a C++ programmer. Before an ArBB function can be computed it needs to be 
+captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
+functionality. The first time {\tt call} is used on a function pointer that function is 
+executed and the ArBB functions used in it build an AST (we say the function has been 
+captured). The AST is then compiled and specialised for the hardware available. Any 
+subsequent calls of a function using ArBB functionality do not lead to a recompilation, 
+the first call caches this compiled code.
 
@@
 using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
-more general language and there was an API missmatch between them. These problems 
+more richer language and there was an API mismatch between Accelerate and ArBB. These problems 
 are mentioned in in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
@@
 
+\subsubsection{Operations on arrays}
+
 \subsubsection{Deep embedding}
 
-\begin{verbatim}
-data Expr = Lit Literal
-          | Var Variable
-
-          | Index0 Expr 
-          | ResIndex Expr Int
-
-          | Map (R GenRecord) [Expr]  
-          | Call (R GenRecord) [Expr]
-
-          | While ([Expr] -> Expr)  ([Expr] -> [Expr])  [Expr]
-
-          | While2 ([Expr] -> Expr)  
-                   ([Expr] -> [Expr])  
-                   [Expr}
-
-          | If Expr Expr Expr
-
-          | Op Op [Expr]   
-\end{verbatim}
-
-\subsubsection{compilation}
-
-\emph{Sharing detection}
-
-\emph{let ArBB do the optimisations}
-
-\emph{having deep embedding means we could compile to something else}
+EmbArBB is a traditional deep embedding, there is an AST data type with constructors 
+representing each ArBB language feature. It is true that a deep embedding of ArBB 
+leads to duplication of effort. The low-level operations exported by the 
+ArBB C interface already construct an AST within the ArBB system. One possible benefit 
+of the deep embedding is that it buys independence from the ArBB backend. It is possible 
+to compile the AST generated on the Haskell side to some other target language.
+
+When compiling EmbArBB to ArBB a sharing detection pass is performed. This pass detects 
+sharing in the AST using the method of A. Gill~\citet{Gill}. One benefit of this pass 
+is that it reduces the numbers of calls into the ArBB C library. The efficiency of code 
+generated is probably not affected since ArBB would discover that sharing in a Common 
+Subexpression Elimination (CSE) pass. Another potential benefit is that sharing detection 
+on the Haskell side results in a simpler AST being built on the ArBB side; potentially making 
+the job for the ArBB optimisation passes simpler. What effects these 
+operations truly had on performance has not been investigated. Actually, applying a sharing 
+detection technique was one of my personal motivations for working on EmbArBB. In Obsidian 
+we have not seen the need for this technique.
+
+One benefit of a deep embedding is that transformations can be performed on the AST. Applying 
+any optimisations in EmbArBB would very likely duplicate effort already put into ArBB. 
+Therefore we apply no further transformations after sharing detection. However, if EmbArBB 
+would be used to generate code in some other target language a set of optimisations 
+would be needed.
 
@@
 
-\emph{ How well did we do?}
+In paper H ,section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
+Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
+C++ performance is not distinguishable from EmbArBB's indicating that our Haskell embedding 
+adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
+the code, that code is identical no matter if the AST was build via Haskell or C++ calls. 
+Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
+takes to go through the process of making those ArBB C API calls. Although if the generated 
+code is used many times the caching ArBB does of generated code means that the initial code 
+generation cost can be amortised.
 
@@
 Since our work with ArBB, Intel has unfortunately retired the ArBB system. This 
-leaves EmbArBB without a functioning code generating backend.
-
+leaves EmbArBB without a functioning code generating backend. This makes the choice of a
+deep embedding (in hindsight) the correct one. EmbArBB could be resurrected by implementing 
+a new code generating backend.
+
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% Related work
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 \section{Related work} 
+\label{sec:relatedwork}
+\FloatBarrier
+
+In figure~\ref{fig:researchmatrix} I our work is placed in relation to other languages,
+embedded languages and libraries for parallel programming. The systems are roughly divided 
+into three groups, based on their level of abstraction. There are low-level languages, here 
+I place languages that are imperative and C-like. In the middle layer I place languages 
+that go a bit further, they have higher level abstractions and are more easily composable. 
+In the highest level I place languages that completely abstract away from details of the 
+hardware they run on. The division is not free from personal bias and the placement of languages
+in boxes was not always easy.
+
+The figure also divides the languages according to what hardware they support. There 
+are CPU specific languages, GPU specific languages and languages that support or aspire 
+to support CPU,GPU and accelerator (Larrabee) execution.
+
+With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
+CUDA, the programmer has full control of how to divide the computation amongst threads and 
+blocks. Using Accelerate, Nikola and Thrust the programmer would use high-level patterns
+without direct insight into how the application would be mapped onto threads and blocks of the 
+GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
+but give the programmer more control of what the computation will look like on the GPU. This 
+is done by allowing the programmer to specify and compose thread-level and block-level code. 
+Recently, the CUB library arrived that is built on a similar idea, that GPU programmers need to 
+have block/thread level control to get maximum performance, while maintaining a higher level 
+and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
+we were alone at targeting this level.
+
+The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
+composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
+Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
+they are not general primitives. Where Accelerate have a single higher order reduction 
+operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
+having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
+%I will go into 
+%what benefits there are to having ArBB embedded in Haskell compared to C++ in
+
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[align=center, scale=2.5]
+% \draw[help lines] (0,0) grid (3,3);
+
+\draw[->] (-0.5,0) -- (-0.5,3) node[anchor=east] {Abstraction level};
+
+\draw (0,3) node[anchor=south west] {CPUs};
+\draw (1,3) node[anchor=south west] {GPUs};
+%\draw (2,3) node[anchor=south west] {Accelerator};
+\draw (2,3) node[anchor=south west] {CPUs, GPUs\\ Accelerators};
+
+\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
+\node[anchor=west] (acc) at (1,1.7) {CUB};
+\node[anchor=west] (repa) at (0,2.6) {DPH\\Nesl\\Repa\\Meta-repa};
+\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
+\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
+
+\node[anchor=west] (cuda) at (1,0.3) {CUDA};
+\node[anchor=west] (opencl) at (2,0.3) {OpenCL};
+
+\node[anchor=west] (OpenMP) at (0,0.3) {OpenMP\\PThreads};
+
+\foreach \x in {0,...,2} { 
+  \foreach \y in {0,...,2} { 
+    \draw (\x,\y) rectangle (\x+1,\y+1);
+  }
+}
+
+\draw[red!75,rounded corners,ultra thick] (1,0.4) rectangle (2,2.4);
+\node[red!75,rotate=45,thick] (obs) at (1.5,1.5) {Obsidian};
+
+\draw[orange!90,rounded corners,ultra thick] (2,1.4) rectangle (3,2.6);
+\node[orange!90,thick] (arbb) at (2.5,2) {EmbArBB\\ArBB};
+
+\end{tikzpicture}
+\end{center}
+\caption{Placing our work in the landscape of languages and libraries for parallel 
+general purpose programming of CPUs, GPUs or both. }
+\label{fig:researchmatrix} 
+\end{figure}
 
@@
 
+\FloatBarrier
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% Future work
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 \section{Future work}
@@
 
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% Discussion
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+
 \section{Discussion}
@@
 % ---------------------------------------------------------------------------
-% Guide 
+% ---------------------------------------------------------------------------
+%
+% Thesis overview
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 \section{Thesis overview}
@@
 
-\subsection{Contributions}
-
-This thesis contributes to the follwoing areas:
-
-\begin{itemize} 
-\item Embedded language design and implementation. 
-\item General Purpose programming using a GPU (GPGPU).
-\item High level data-parallel programming across devices.
-\end{itemize}
-
-\subsubsection{Embedded language design and implementation}
-
-\subsubsection{General Purpose programming using a GPU}
-
-\subsubsection{High level data-parallel programming across devices}
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% BIBLIOGRAPHY
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
 

### 2013-10-24 src2/thesis.tex
update

@@
 %
+% GPU Programming 
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+\section{GPUs and CUDA}
+
+In 2006 NVIDIA released their first Compute Unified Device Architechture (CUDA) 
+GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
+for general purpose workloads. A dialect of C, called CUDA C, was released specifically 
+to simplify implementation of general purpose algorithms on GPUs.
+
+A CUDA GPU is built around a single kind of processor (opposed to the different 
+kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
+MultiProcessors), all contain a number of function units called SPs (Streaming Processors). 
+Each MP also contains local memory, called shared memory since it can be accessed 
+by all of the SPs in that MP. The number of MPs vary over the available GPUs; cheaper 
+GPUs have as few as one MP, and as you go up in price the number of MPs increase.
+
+Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
+2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
+a {\em Warp}, that are executed in lock-step (SIMD style execution). Threads are 
+also divided into {\em Blocks}; the threads within a block can communicate using the 
+local shared memory. The maximum block size is currently 1024 threads. Since 
+the number of threads in a block can be higher than the warp size, a barrier 
+synchronisation mechanism exists to ensure that threads within a block 
+can communicate safely via the shared memory.
+
+Papers B, C, E and F (sections~\ref{sec:paperB},~\ref{sec:paperC},~\ref{sec:paperE},
+~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
+about them ,in relation to Obsdian.
+
+\subsection{CUDA programming}
+
+The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
+writes what is called {\em kernels}, in CUDA jargon a kernel refers to a sequential C program 
+that is parameterised on a threads identity. The kernel program is executed in parallel across 
+all the threads of a block and many such blocks of threads can execute in parallel on the 
+available MPs. The programmer needs to decompose the algorithm and data in such a way what 
+communication will be local to the blocks.
+
+Below is a very simple CUDA example to give some indication of what a kernel and 
+the code to launch that kernel looks like.
+
+\begin{small}
+\begin{verbatim} 
+__global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {
+
+  r[threadIdx.x] = i1[threadIdx.x] + i2[threadIdx.x];
+
+} 
+\end{verbatim}
+\end{small}
+
+This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
+elementwise sums into the result array, {\tt r}. {\tt threadIdx} identifies 
+a thread within a block. There is also a {\tt blockIdx} variable that identifies 
+the block a thread belongs to but that is not used in this case.
+
+In CUDA there is a notion of {\em host} and {\em device}, the device is the GPU and 
+its memory (Global memory, usually a few Gigabytes). The host refers to the system housing 
+the GPU. The host runs a control program that launches kernels on the GPU. First memory 
+needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
+the host systems memory into device memory.
+
+\begin{small}
+\begin{verbatim} 
+int main(void) {
+
+  int32_t h_i1[32], h_i2[32], h_r[32];
+
+  int32_t *d_i1, *d_i2, *d_r;
+
+  for (int i = 0; i < 32; ++i) { 
+    h_i1[i] = i;
+    h_i2[i] = 32 - i;
+  }
+
+  // Allocate device memory
+  cudaMalloc((void**)&d_i1,32*sizeof(int32_t));
+  cudaMalloc((void**)&d_i2,32*sizeof(int32_t));
+  cudaMalloc((void**)&d_r,32*sizeof(int32_t));
+
+  // Copy inputs to device
+  cudaMemcpy(d_i1,
+             h_i1,
+             32*sizeof(int32_t),
+             cudaMemcpyHostToDevice);
+  cudaMemcpy(d_i2,
+             h_i2,
+             32*sizeof(int32_t),
+             cudaMemcpyHostToDevice);
+
+  // Launch the kernel
+  kernel<<<1,32,0>>>(d_i1,d_i2,d_r);
+
+  // Copy result to host
+  cudaMemcpy(h_r,
+             d_r,
+             32*sizeof(int32_t),
+             cudaMemcpyDeviceToHost);
+
+  for (int i = 0; i < 32; ++i) { 
+    printf("%d ",h_r[i]);
+  }
+
+  return 0;
+}
+\end{verbatim} 
+\end{small}
+
+The syntax for launching a kernel is:
+
+\begin{small}
+\begin{verbatim}
+name<<<#Blocks, #Threads/Block, #BytesShared>>>(arg1,..,argN);
+\end{verbatim}
+\end{small}
+
+In this case one block of 32 threads using no shared memory, is launched.
+
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
 % Embedded languages
@@
 The approach taken to embedd the scalar language is very similar to what is described 
-in~\citet{COMPILEEDSL}; but we use a GADT to obtained typed expressions rather than phantom types.
+in~\citet{COMPILEEDSL}; but we use a Generalised Algebraic Datatype (GADT) to obtained 
+typed expressions rather than phantom types.
 
@@
 we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
-Simply put, this means providing a more Haskell-programmer-friendly interface compared
-to the raw ArBB bindings (which are a collection of functions in the IO monad).
+We call this embedded language EmbArBB and it provides a more Haskell-programmer-friendly 
+interface compared to the raw ArBB bindings (which are a collection of functions 
+in the IO monad).
 

### 2013-10-24 src2/thesis.tex
update

@@
 \begin{small}
-\begin{verbatim} 
+\begin{Verbatim}[samepage=true]
 __global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {
 
-  r[threadIdx.x] = i1[threadIdx.x] + i2[threadIdx.x];
+  unsigned int gid = blockIdx.x * 
+                     blockDim.x + 
+                     threadIdx.x;
+
+  r[gid] = i1[gid] + i2[gid];
 
 }
-\end{verbatim}
+
+\end{Verbatim}
 \end{small}
@@
 This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
-elementwise sums into the result array, {\tt r}. {\tt threadIdx} identifies 
+elementwise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
 a thread within a block. There is also a {\tt blockIdx} variable that identifies 
-the block a thread belongs to but that is not used in this case.
+the block a thread and {\tt blockDim.x} specifies the number of threads per block.
 
@@
 \begin{small}
-\begin{verbatim} 
+\begin{Verbatim}[samepage=true]
 int main(void) {
@@
 }
-\end{verbatim} 
+\end{Verbatim} 
 \end{small}
@@
 \begin{small}
-\begin{verbatim}
-name<<<#Blocks, #Threads/Block, #BytesShared>>>(arg1,..,argN);
-\end{verbatim}
+\begin{Verbatim}[samepage=true]
+name<<<Blocks, Threads/Block, BytesShared>>>(arg1,..,argN);
+\end{Verbatim}
 \end{small}
 
-In this case one block of 32 threads using no shared memory, is launched.
+\noindent In this example above one block of 32 threads using no shared memory, is launched.
+
+A kernel that uses shared memory takes a form similar to whats shown below.
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+__global__ void kernel2(int32_t *i, int32_t *r) {
+
+  extern __shared__ int32_t sm[];
+
+  unsigned int tid = threadIdx.x; 
+  unsigned int gid = blockIdx.x * 
+                     blockDim.x + 
+                     threadIdx.x;
+
+  sm[tid] = i[gid]; 
+  __syncthreads();
+
+  /* Compute on sm */
+
+  __syncthreads();
+  r[gid] = sm[tid];
+
+}
+\end{Verbatim} 
+\end{small}
+
+\noindent A portion of the input array is stored into shared memory. A 
+barrier synchronisation is used to ensure that all threads within the block 
+has stored its value into shared memory (only necessary if threads will 
+communicate across warp boundaries). After computing on the data locally 
+it is written back into global memory.
 

### 2013-10-25 src2/thesis.tex
update

@@
 Expressive Array Constructs in an Embedded GPU Kernel Programming Language.
-In \emph{Proceedings of the 7th workshop of Declarative aspects and applications of multicore programming}, DAMP '12.
+In \emph{Proceedings of the 7th workshop on Declarative aspects and applications of multicore programming}, DAMP '12.
 ACM.
@@
 Efficient Counting Sort Implementations using an Embedded GPU Programming Language.
-\emph{2nd Workshop of Functional High-Performance Computing, FHPC '13.}
+\emph{2nd Workshop on Functional High-Performance Computing, FHPC '13.}
 \item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
@@
 Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks.
-\emph{Proceedings of the 1st ACM SIGPLAN Workshop of Functional High-Performance computing, FHPC '12.}
+\emph{Proceedings of the 1st ACM SIGPLAN Workshop on Functional High-Performance computing, FHPC '12.}
 \end{enumerate}
@@
 
-The code and examples in section one and two are my work. Section four we developed together 
-but is mostly Josef's expertise. For section four we also got very much and valuable help 
+The code and examples in section one and two are my work. The contents of section four was 
+developed by Josef Svenningsson and me. For section four we also got very much and valuable help 
 from Emil Axelsson.
@@
 \subsubsection{Papers G and H} 
-Here, again, most implementation effort was done by me. For paper G, this work was performed 
+The work behind paper G, work was performed by me 
 under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
@@
 and still making building blocks implemented by the programmer reusable and composable. 
-We want to give more compositional tools for the programmer to experiment with while 
+We want to provide more compositional tools for the programmer to experiment with while 
 exploring the space of various ways to implement a given pattern or primitive. 
@@
 
-A CUDA GPU is built around a single kind of processor (opposed to the different 
+A CUDA GPU is built around a single kind of processor (as opposed to the different 
 kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
-MultiProcessors), all contain a number of function units called SPs (Streaming Processors). 
+MultiProcessors) all contain a number of function units called SPs (Streaming Processors). 
 Each MP also contains local memory, called shared memory since it can be accessed 
-by all of the SPs in that MP. The number of MPs vary over the available GPUs; cheaper 
-GPUs have as few as one MP, and as you go up in price the number of MPs increase.
+by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
+GPUs have as few as one MP, and as you go up in price the number of MPs increases.
 
@@
 2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
-a {\em Warp}, that are executed in lock-step (SIMD style execution). Threads are 
+{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
 also divided into {\em Blocks}; the threads within a block can communicate using the 
@@
 ~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
-about them ,in relation to Obsdian.
+about them in relation to Obsdian.
 
@@
 The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
-writes what is called {\em kernels}, in CUDA jargon a kernel refers to a sequential C program 
-that is parameterised on a threads identity. The kernel program is executed in parallel across 
+writes what are called {\em kernels}; in CUDA jargon a kernel refers to a sequential C program 
+that is parameterised on a thread's identity. The kernel program is executed in parallel across 
 all the threads of a block and many such blocks of threads can execute in parallel on the 
-available MPs. The programmer needs to decompose the algorithm and data in such a way what 
+available MPs. The programmer needs to decompose the algorithm and data in such a way that 
 communication will be local to the blocks.
@@
 The embedded language approach is beneficial since we are not entirely sure how 
-to program current and future highly parallel computers. Being embedded allows 
-us more rapid prototyping of ideas. Other benefits of being embedded often listed 
-are: making use of the host language's type system, parser and libraries (to some extent) 
+to program current and future highly parallel computers. Implementing embedded languages allows 
+us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
+are: making use of the host language's type system for guaranteeing generating correctly typed 
+target code, parser and libraries (to some extent) 
 and to have the host language as a powerful macro language for your embedded language.
@@
 A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
-by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embedd a compiler 
+by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embed a compiler 
 backend and generate efficient code in some target language. One key aspect of this 
@@
 well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
-embedding that builds ASTs is called a deep embedding. Shallow embeddings on the 
-other hand do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
+embedding that builds ASTs is called a deep embedding. Shallow embeddings, on the 
+other hand, do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
 Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
@@
 level of abstraction in fields where the default is to use low-level languages. 
-For example, there is Lava and Wired in the fields of hardware design and 
+For example, Lava and Wired in the fields of hardware design and 
 verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
-programming there are numerous embedded approaches using various host languages. 
-But to mention two that use Haskell we have Accelerate and Nikola 
+programming, there are numerous embedded approaches using various host languages. 
+Two languages for GPU programming embedded in Haskell are  Accelerate and Nikola 
 \citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
@@
 Related work and information about the tools, libraries and language that make the 
-wider context surrounding Obsidian is described in section~\ref{sec:relatedwork}.
+wider context surrounding Obsidian can be found in section~\ref{sec:relatedwork}.
 
@@
 
-Obsidian aspires to raise the level of abstraction of GPU kernel implementation 
+Obsidian aspires to raise the level of abstraction of GPU kernel implementation, 
 while maintaining enough control over details that influence performance on a GPU. 
@@
 its strengths and weaknesses. We also feel that it is important that the programmer 
-be able to use local shared memory in computations and has the ability to influence 
+be able to use local shared memory in computations and have the ability to influence 
 memory access patterns. What these array representations and abstractions are 
@@
 programs). The programmer needs to take care to ensure that the communication patterns are 
-within a warp and is then free to remove synchronisations. If warps where to be put in the 
+within a warp and is then free to remove synchronisations. If warps were to be put in the 
 program level hierarchy of Obsidian, its place would be between the thread and block level. 
-This forces the programmer to go via warps when putting together a block computation even 
+This would force the programmer to go via warps when putting together a block computation even 
 if the communication pattern is not suited for that, leading to inconvenience with no gain.
 
-As an example of how the hierarchy level parameter is used I show the {\tt ForAll} constructor 
+As an example of how the hierarchy level parameter is used, I show the {\tt ForAll} constructor 
 from the {\tt Program} data type.  This constructor represents parallelism either over threads 
@@
 
-The approach taken to embedd the scalar language is very similar to what is described 
-in~\citet{COMPILEEDSL}; but we use a Generalised Algebraic Datatype (GADT) to obtained 
-typed expressions rather than phantom types.
+The approach taken to embed the scalar language is very similar to what is described 
+in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
+Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
+types. Earlier versions of Obsidian (section~\ref{sec:paperB} used phantom types.
 
@@
 
-Push arrays where added to Obsidian as a complement to pull arrays. Pull arrays are 
+Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
 good because they are so simple to understand and easy to parallelise. However certain

### 2013-10-25 src2/thesis.tex
update

@@
 
-In 2006 NVIDIA released their first Compute Unified Device Architechture (CUDA) 
+In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
 GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
@@
 ~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
-about them in relation to Obsdian.
+about them in relation to Obsidian.
 
@@
 This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
-elementwise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
+element wise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
 a thread within a block. There is also a {\tt blockIdx} variable that identifies 
@@
 it is written back into global memory.
+
+\subsection{Writing efficient GPU kernels}
+
+Writing efficient GPU kernels is hard. There are many details that the CUDA 
+programmer must be aware of in order to make optimal use of the GPU. 
+the NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
+to learn more about these details. Here is a list with explanations of some 
+of the practices that the guide deem most important.
+
+\begin{itemize}
+
+\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
+and GPU is over 20 times of the bandwidth between host memory and the GPU device via the PCIe bus.
+This means that, first of all one must evaluate if the computation to be perform is significant 
+enough to warrant a transfer to the GPU. If it computation is not arithmetic intensive it may be 
+better to do the work on the CPU. However, if the data is already in device memory and 
+the operation you are about to perform on it is more efficiently handled by the CPU, it may still 
+be better to let the GPU do the work. In other words, compute near the data unless transferring it 
+pays off.
+
+\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
+GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
+what these access patterns are very some depending on the GPU. For a representative model of 
+GPU the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
+to as many transactions as cache lines touched (128b l1 cache lines). This means that strided 
+accesses from within a warp are not favoured.
+
+\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
+many 400 to 600 clock cycles if the data is not cached. The programmer should 
+use shared memory as much as possible to avoid redundant loading from global memory.
+Shared memory could also be used to coalesce reads from global memory. Shared memory is 
+a limited resource in the MP, so if a kernel uses very much shared memory it means 
+fewer such blocks can be active on the MP.
+
+\item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
+scheduled unit on a GPU. If a warp accesses memory (and will be waiting for it arrive during 
+600 cycles) it will be swapped out and another warp will take its place. This means that 
+the number of available warps decides how well latency can be hidden. A block that is not a 
+multiple of 32 leads to wasting GPU resources, some processing elements will stand idle during 
+the execution of the not full warps.
+
+\item {\bf Avoid different execution paths within a block:} Any branching instructions that 
+causes the execution path diverge within a warp will affect performance. When the execution 
+paths within a warp diverges the computation is serialised.
+
+\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
+when synchronising within any conditionally executed code; every thread must reach this barrier.
+Failing to ensure this will likely lead to the kernel producing incorrect answers. Note that 
+this is a major limitation to the composability of CUDA functions. If it is safe or 
+not to call some function at a certain place in your code, can only be decided by inspecting 
+that functions implementation. 
+\end{itemize}
+
+The list above provides some very general guide lines of what to think about when 
+writing GPU kernels. The lesson is that fast memory near the processing element 
+should be preferred over slow memory far away. However, this comes at the effort 
+of decomposing computation and data in a way that makes this possible.
+
+%\citet{FERMIOPT}
+%\citet{BestPrac} 
+%\citet{merrill}
 

### 2013-10-26 src2/thesis.tex
update

@@
 and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
-its capabilities was also offered. ArBB programs are JIT-ed (Just-In-Time compiled) to 
+its capabilities was also provided. ArBB programs are JIT-ed (Just-In-Time compiled) to 
 target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
@@
 GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
-for general purpose workloads. A dialect of C, called CUDA C, was released specifically 
+for general purpose programming. A dialect of C, called CUDA C, was released specifically 
 to simplify implementation of general purpose algorithms on GPUs.
@@
 also divided into {\em Blocks}; the threads within a block can communicate using the 
-local shared memory. The maximum block size is currently 1024 threads. Since 
+shared memory. The maximum block size is currently 1024 threads. Since 
 the number of threads in a block can be higher than the warp size, a barrier 
@@
 a thread within a block. There is also a {\tt blockIdx} variable that identifies 
-the block a thread and {\tt blockDim.x} specifies the number of threads per block.
+the block a thread belongs to and {\tt blockDim.x} specifies the number of threads per block.
 
@@
 needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
-the host systems memory into device memory.
+the host's system memory into device memory.
 
@@
 use shared memory as much as possible to avoid redundant loading from global memory.
-Shared memory could also be used to coalesce reads from global memory. Shared memory is 
+Shared memory can also be used to coalesce reads from global memory. Shared memory is 
 a limited resource in the MP, so if a kernel uses very much shared memory it means 
@@
 \item {\bf Avoid different execution paths within a block:} Any branching instructions that 
-causes the execution path diverge within a warp will affect performance. When the execution 
+causes the execution paths to diverge within a warp will affect performance. When the execution 
 paths within a warp diverges the computation is serialised.
@@
 when synchronising within any conditionally executed code; every thread must reach this barrier.
-Failing to ensure this will likely lead to the kernel producing incorrect answers. Note that 
+Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
 this is a major limitation to the composability of CUDA functions. If it is safe or 

### 2013-10-26 src2/thesis.tex
update

@@
 
+\subsection{Compilation to CUDA}
+
 \subsection{Results}
@@
 \emph{provide information around the results in papers}
-
-\subsection{Compilation to CUDA}
 
@@
 
-In figure~\ref{fig:researchmatrix} I our work is placed in relation to other languages,
-embedded languages and libraries for parallel programming. The systems are roughly divided 
+\subsection{Languages for data-parallel programming}
+
+In figure~\ref{fig:researchmatrix} our work is placed in relation to other languages,
+embedded languages and libraries for parallel programming. The languages and libraries consider 
+are data-parallel; languages for control parallelism or distributed message passing system are 
+not considered. The systems are roughly divided 
 into three groups, based on their level of abstraction. There are low-level languages, here 
@@
 are CPU specific languages, GPU specific languages and languages that support or aspire 
-to support CPU,GPU and accelerator (Larrabee) execution.
-
+to support CPU,GPU and accelerator (Larrabee, Xeon Phi) execution.
+
+% ---------------------------------------------------------------------------
+% Obsidian, Accelerate, Nikola, Thrust, CUB 
+% ---------------------------------------------------------------------------
 With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
-CUDA, the programmer has full control of how to divide the computation amongst threads and 
-blocks. Using Accelerate, Nikola and Thrust the programmer would use high-level patterns
-without direct insight into how the application would be mapped onto threads and blocks of the 
-GPU. Using Obsidian we want to bridge that gap by both being more composable than CUDA 
-but give the programmer more control of what the computation will look like on the GPU. This 
-is done by allowing the programmer to specify and compose thread-level and block-level code. 
-Recently, the CUB library arrived that is built on a similar idea, that GPU programmers need to 
-have block/thread level control to get maximum performance, while maintaining a higher level 
-and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as far as I am aware, 
-we were alone at targeting this level.
-
+CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
+the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
+Nikola~\citet{NIKOLA} 
+and Thrust~\citet{THRUST} the programmer would use 
+high-level patterns without direct insight into how the application would be mapped onto 
+threads and blocks of the GPU. Using Obsidian we want to bridge that gap by both being 
+more composable than CUDA but give the programmer more control of what the computation will 
+look like on the GPU. This is done by allowing the programmer to specify and compose 
+thread-level and block-level code. Recently, the CUB library arrived that is built on a 
+similar idea, that GPU programmers need to have block/thread level control to get maximum 
+performance, while maintaining a higher level and more Thrust-like programmer 
+interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
+this level.
+
+% ---------------------------------------------------------------------------
+% ArBB EmbArBB
+% ---------------------------------------------------------------------------
 The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
@@
 
+% ---------------------------------------------------------------------------
+% Microsoft Accelerator
+% ---------------------------------------------------------------------------
+Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
+a data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
+operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
+compile to multithreaded CPU code and make use of SIMD units of modern 
+CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.
+
+% ---------------------------------------------------------------------------
+% Python systems
+% ---------------------------------------------------------------------------
+There are many embedded languages implemented in Python that makes use of GPUs 
+to improve performance. One example is Copperhead~\citet{copperhead}. Copperhead 
+supports flat and nested data-parallelism via primitives such as {\tt map} and 
+{\tt reduce}.
+
+% ---------------------------------------------------------------------------
+% Repa, Meta-Repa 
+% ---------------------------------------------------------------------------
+
+Repa (Regular Shape-polymorphic Arrays)~\citet{REPA}, is a Haskell library for 
+data-parallel programming. 
+Repa uses the same array representation as Obsidian, in Repa called a delayed
+array, which is a function from index to element. However, in Repa arrays have shape, 
+they can be of any dimensionality and there shape-polymorphic functions on these. 
+For example {\tt map} can be applied to an array of any shape.
+
+Meta-Repa is a reimplementation of the Repa library using deep embedded language 
+techniques. Meta-Repa programs build ASTs that are then compiled (using template 
+Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
+The Meta-Repa approach to implementing the Repa library 
+gives inlining of operations for free, where Repa programs needs to be annotated 
+with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
+needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
+This is related to our work on EmbArBB that also uses an existing library (ArBB) 
+for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
+of embedding prexisting DSLs.
+
+% ---------------------------------------------------------------------------
+% DPH, NESL 
+% ---------------------------------------------------------------------------
+Data-Parallel Haskell~\citet{DPH} and Nesl~\citet{NESL} are languages for 
+nested data-parallelism. Both run on CPUs, but there is also work towards 
+a GPU version of Nesl~\citet{NestedGPU}.
+
+% ---------------------------------------------------------------------------
+% Delite, Vertigo 
+% ---------------------------------------------------------------------------
+
+Delite~\citet{DELITE} is an infrastructure for implementation of domain 
+specific embedded languages. There is an extensible internal representation (IR,AST); 
+the programmer extends some base IR with domain specific constructs. The Delite system 
+has been used to implement domain specific embedded languages for machine learning, rendering, 
+physics and other domains. Delite, being a framework and the languages implemented using it 
+very domains specific, was impossible to place in the matrix (figure~\ref{fig:researchmatrix}).
+
+\subsection{Embedded domain specific languages}
+
+\noindent\emph{Feldspar}\citet{FELDSPAR2010} \newline
+\noindent\emph{Lava}\citet{lavaICFP} \newline
+\noindent\emph{Vertigo}\citet{VERTIGO}  \newline
+
+\subsection{Compiling Embedded Languages}
+
+\noindent\emph{Generic monad constructs (josef + emil + anders}\citet{Generic} \newline
+\noindent\emph{The Constrained Monad Problem (Gill)}\citet{sculthorpe2013constrained} \newline
+\noindent\emph{Compiling Embedded Languages}\citet{COMPILEEDSL} \newline
+
 \begin{figure}
@@
 
-\subsection{Parallelism}
-
-\noindent\emph{DPH}\citet{DPH} \newline
-\noindent\emph{D.A.Accelerate}\citet{ACCELERATEDAMP11} \newline
-\noindent\emph{Nikola}\citet{NIKOLA}\newline
-\noindent\emph{Repa}\citet{REPA}\newline
-\noindent\emph{Meta-Repa}\citet{METAREPA}\newline
-\noindent\emph{ArBB}\citet{ARBB2011}\newline
-\noindent\emph{OpenCL}\citet{OpenCL}\newline
-\noindent\emph{CUDA}\citet{CUDA}\newline
-\noindent\emph{Microsoft Accelerator}\citet{ACCELERATOR}\newline
-\noindent\emph{Copperhead}\citet{copperhead}\newline
-\noindent\emph{Delite}\citet{DELITE}\newline
-\noindent\emph{A monad for deterministic parallelism (R. Newton)} \citet{MonadPar} \newline
-\noindent\emph{Nesl}\citet{NESL} \newline
-
-\subsection{Embedded domain specific languages}
-
-\noindent\emph{Feldspar}\citet{FELDSPAR2010} \newline
-\noindent\emph{Lava}\citet{lavaICFP} \newline
-\noindent\emph{Vertigo}\citet{VERTIGO}  \newline
-
-\subsection{Compiling Embedded Languages}
-
-\noindent\emph{Generic monad constructs (josef + emil + anders}\citet{Generic} \newline
-\noindent\emph{The Constrained Monad Problem (Gill)}\citet{sculthorpe2013constrained} \newline
-\noindent\emph{Compiling Embedded Languages}\citet{COMPILEEDSL} \newline
+%\subsection{Parallelism}
+
+%\noindent\emph{DPH}\citet{DPH} \newline
+%\noindent\emph{D.A.Accelerate}\citet{ACCELERATEDAMP11} \newline
+%\noindent\emph{Nikola}\citet{NIKOLA}\newline
+%\noindent\emph{Repa}\citet{REPA}\newline
+%\noindent\emph{Meta-Repa}\citet{METAREPA}\newline
+%\noindent\emph{ArBB}\citet{ARBB2011}\newline
+%\noindent\emph{OpenCL}\citet{OpenCL}\newline
+%\noindent\emph{CUDA}\citet{CUDA}\newline
+%\noindent\emph{Microsoft Accelerator}\citet{ACCELERATOR}\newline
+%\noindent\emph{Copperhead}\citet{copperhead}\newline
+%\noindent\emph{Delite}\citet{DELITE}\newline
+%\noindent\emph{A monad for deterministic parallelism (R. Newton)} \citet{MonadPar} \newline
+%\noindent\emph{Nesl}\citet{NESL} \newline
+
+%\subsection{Embedded domain specific languages}
 
@@
 \noindent\emph{What contributions are made by Obsidian} 
+Looking at the related work, we se that almost all high level GPU programming 
+languages provide abstractions of parallel primitives, aggregate operations, 
+map-folds-zipwiths, giving justification to  our investigation of GPU programming without 
+these.
 

### 2013-10-27 src2/thesis.tex
update

@@
 
-In this thesis the focus is on data-parallelism, as opposed to task- or 
-control-parallelism. Data parallelism is applicable where the input data can 
-be divided into chunks and operated upon independently. \emph{more, and maybe not here}
+This thesis focuses on data-parallelism, as opposed to control- or 
+task-parallelism. Data parallelism is applicable where the input data can 
+be divided into chunks and operated upon independently. Data-parallelism is scalable
+meaning that usually as problem size increase, the potential for parallelism increase.
 
@@
 The patterns, while very powerful as abstractions, also provide tools for thinking about 
-parallel algorithms.
+parallel algorithms. The ArBB library support flat regular data-parallelism, but also 
+irregular and to some degree nested data-parallelism.
 
@@
 In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian. 
+Obsidian is focused on flat and regular data-parallelism.
 
@@
 
+Push arrays, like pull arrays, are functions. A Push array is a higher order 
+function, taking a function from value and index to a thread program and giving 
+as result a program at some level in the hierarchy. The function passed to the 
+push array function we call a write-function. Below is the definition of 
+push arrays that we use in Obsidian. Push arrays are parameterised on element 
+type ({\tt a}), size type ({\tt s}) and program hierarchy level ({\tt p}).
+
 \begin{verbatim}
@@
 
+When working with pull arrays, the consumer of such an array decides how to 
+iterate over the array and what elements to compute. This is done applying the 
+push array function to those indices that are interesting to the consumer. Push 
+arrays work in the opposite way. The Push array function (the result program) contain 
+the iteration schema; the consumer of the push only decides what write function 
+to supply. The iteration schema used is decided upon when creating a push arrays. 
+The code below is an example of how to create a push array from a pull array.
+
+\begin{verbatim} 
+convertToPush :: Pull Word32 e 
+               -> Push Block Word32 e 
+convertToPush (Pull n ixf) =
+    Push n $ \wf -> 
+      forAll (fromIntegral n) $ \i -> wf (ixf i) i
+\end{verbatim}
+
+The function above takes a pull array and creates a push array with a parallel 
+iteration schema over threads within a block. There are other iteration schemas to 
+chose from such as sequential, combinbations of sequential and paralell and parallel 
+across both threads and blocks (two level parallelism).
+
+Push arrays give many of the same benefits that pull arrays have. Using pull 
+arrays we also get fusion of operations as default. The implementation of map 
+(shown below) illustrates how maping a function {\tt f} over a push array results 
+in an application of {\tt f} in the write-function.
+
+\begin{verbatim} 
+map f (Push s p) = Push s $ \wf -> p (\e ix -> wf (f e) ix)
+\end{verbatim}
+
 \subsection{Compilation to CUDA}
 
+When compiling Obsidian programs to CUDA, we only need to be concerned with the 
+{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype 
+we know that the AST can be quite directly translated to CUDA. We know that there will 
+be at most one level of nestedness in the parallelism (several blocks, each of several threads). 
+This means that we do not need to do any transformations on the AST before generating CUDA code.
+
 \subsection{Results}
 
-\emph{point to results in each paper}
-
-\emph{provide information around the results in papers}
+We have implemented basic algorithms such as sorting, merging, fractal generation, reductions and parallel prefix (scans). Often we have managed to get very good performance, close to that 
+of hand optimised CUDA code. However, there are also cases where we just cant seem to reach 
+the level of performance that we would want. One such example is parallel prefix, an 
+application that has been haunting us from the very beginning. Scan is problematic to us because 
+it is best implemented as an in-place algorithm while our array representations do 
+not allow for in-place updated (there is no memory associated with them to update). 
+We are trying to improve this situation by adding a third array representation to Obsidian,
+mutable arrays.
+
+Paper F (section~\ref{sec:paperF}), tells a story about how to implement and tune 
+a reduction kernel in Obsidian. At the end the implemented reduction kernel is 
+faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
+reduction in accelerate corresponds to just one single hand optimised skeleton. We would have 
+been happy just getting close.
+
+In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrance sort 
+using Obsidian. We use very low-level features of Obsidian and end up producing 
+code that performs very well.
+
+Paper D (section~\ref{sec:paperD}), introduces push arrays and uses them in the 
+implementation of sorting networks to quite good results. With push arrays we 
+where able to experiment with more detailed decompositions (what thread computes 
+what value) of the algorithms.
 
@@
 
+EmbArBB provides many of the ArBB operations on Arrays butprovides a Repa like 
+shape-polymorphic interface to them. For example a reduction operation can take a two  
+dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
+it takes a one dimensional vector and returns a single value (zero dimensional vector).
+
+\begin{verbatim} 
+addReduce :: Num a 
+           => Exp USize 
+           -> Exp (DVector (t:.Int) a) 
+           -> Exp (DVector t a) 
+\end{verbatim}
+
+The {\tt addReduce} function takes an input vector that is atleast one dimensional 
+({\tt t:.Int}) and returns a vector that has one less dimension ({\tt t}).
+
 \subsubsection{Deep embedding}
@@
 
-\noindent\emph{Feldspar}\citet{FELDSPAR2010} \newline
-\noindent\emph{Lava}\citet{lavaICFP} \newline
-\noindent\emph{Vertigo}\citet{VERTIGO}  \newline
+Feldspar is an embedded language for digital signal processors~\citet{FELDSPAR}. Apart 
+from using the syntactic library~\citet{SyntacticICFP12}, it is quite similar to Obsidian. 
+There is a deeply embedded core language (called Core) and vectors as a shallow embedding 
+on top of that, just like in Obsidian.
+
+When we first started working on Obsidian we used Lava~\citet{LavaICFP} as a inspiration. 
+We wanted to mimic the Lava programming style but generate GPU code. At first this worked quite 
+well and most Obsidian programs superficially looked quite like Lava programs. We have diverged 
+from this aspiration since realising that GPUs are quite a different story and performance 
+requires more control of GPU architecture specifics.
 
@@
 
-\noindent\emph{Generic monad constructs (josef + emil + anders}\citet{Generic} \newline
-\noindent\emph{The Constrained Monad Problem (Gill)}\citet{sculthorpe2013constrained} \newline
-\noindent\emph{Compiling Embedded Languages}\citet{COMPILEEDSL} \newline
+In paper A~(\ref{sec:paperA}) we describe our method for compiling monadic embedded languages. 
+This means we need to take a monadic computation and observe its structure in order to generate
+code. This work is closely related to a method described in ``The Constrained Monad Problem''
+by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In this paper they solve a more 
+general problem than just compilation but our method is nice because it is easy to understand 
+and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
+also solves the problem of compiling monads using a continuation monad.
+
+``Compiling Embedded Languages'' by Elliott and Finne has been a constant source of information 
+and something we have refered to in almost all our work. This is, in my opinion, the paper to turn 
+to for anyone starting out with compiled embedded languages.
 
@@
 
-\noindent\emph{Find a niche for Obsidian}
-
-\noindent\emph{Improve on Warp level capabilities}
-
-\noindent\emph{Improve expressivity, code performance}
+Obsidian is a language for GPU kernel implementation. It is not as well integrated 
+into Haskell as for example Accelerate. There are ways to execute Obsidian kernels 
+from within Haskell; the recommended workflow, however, is to generate the CUDA  
+as text and write the host code by hand. This can be seen both a shortcoming 
+and a feature. It is hard to integrate Accelerate programs in larger applications 
+written in other languages. While when using Obsidian a kernel is generated and 
+can be used as part of for example a library.
+
+Obsidian is also a quite low-level language. The programmer needs to know GPU details 
+and is allowed to write programs at a level almost as low as CUDA, if it is desired. 
+Paper E (section~\ref{sec:paperE}) contains examples of programming at this low level. 
+One potential use of Obsidian could be as the code generating backend of high-level 
+and very domain specific embedded languages. This could be a niche for Obsidian.
+
+Here and there in the thesis it has been hinted that Obsidian has warp level 
+capabilites and that these are not as integrated into the hierarchy as block, 
+thread and grid are. The warp abstractions offered also requires more work in 
+the code generating backend than the others, that correspond directly to CUDA concepts. 
+The problem is that CUDA does not have a warp abstraction. Warps needs to be 
+better integrated with the rest of Obsidian and the code generation required 
+for warps needs more testing before being trustworthy and useful.
+
+%\noindent\emph{Improve on Warp level capabilities}
+
+%\noindent\emph{Improve expressivity, code performance}
 
@@
 
-\noindent\emph{What contributions are made by Obsidian} 
-Looking at the related work, we se that almost all high level GPU programming 
+This thesis contains two approaches to embedding a language for data-parallel programming 
+in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
+and at little effort from us, very good performance is gained. The performance of 
+EmbArBB is entirely due to all the optimisation expertice and effort that invested in 
+the ArBB system. We get all that expertice for free. What EmbArBB provide is a 
+way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB is based on a 
+set of parallel primitives for the programmer to use and compose. Obsidian takes 
+a different approach. Built in primitives are very few and very low level, such as 
+sequential and parallel for loops. On top of the low-level capabilities a library 
+of high-level primitives can be built. The key is that we want the programmer to 
+be able to specify how to implement these higher level abstractions.
+
+Looking at the related work, we see that almost all high level GPU programming 
 languages provide abstractions of parallel primitives, aggregate operations, 
-map-folds-zipwiths, giving justification to  our investigation of GPU programming without 
-these.
-
-\noindent\emph{What works well with Obsidian}
-
-\noindent\emph{EmbArBB contributions}
+maps, folds and reductions, giving justification to our investigation of GPU 
+programming without these. We believe that the GPU architecture is still volatile 
+enough that there be no ``the right'' way of implementing these primitives. As of 
+today, the programmer still needs to explore the design space and find a sollution 
+that performs well on a particular GPU.
+
+Performance of Obsidian kernels is often good and the Obsidian code that they 
+are generated from is often shorted than corresponding CUDA programs. Obsidian 
+programs differ from CUDA programs by describing whole programs. With this I mean 
+that a CUDA programmer needs to take a leap from the single threaded program 
+parameterised on thread Id, to what actually will happen when this program is 
+run by hundreds or thousands threads at once. The extensive indexing arithmetic 
+often present in CUDA programs can make this leap a hard one. Similar programs 
+in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
+bounded parallel loops.
+
+Push arrays where invented by Koen Claessen and first implemented in Obsidian 
+to provide more control over where computation takes place (which thread computes 
+which element). Push arrays has since been experimented on by others and show up 
+in Nikola~\citet{NIKOLAPUSH}, been experimented upon by the people behind Feldspar 
+and appear in StreamHS~\citet{FPCDSL}.
 

### 2013-10-27 src2/thesis.tex
update

@@
 us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
-are: making use of the host language's type system for guaranteeing generating correctly typed 
-target code, parser and libraries (to some extent) 
-and to have the host language as a powerful macro language for your embedded language.
+are include making use of the host language's type system for guaranteeing generating correctly typed 
+target code. Already existing libraries in the host language (if they are general enough) 
+can be used in the embedded language. The host language can also be used as a powerful macro 
+language for your embedded language.
 

### 2013-10-27 src2/thesis.tex
update

@@
 This means that we do not need to do any transformations on the AST before generating CUDA code. 
+Paper F (section~\ref{sec:paperF}) contains technical details of the compilation process.
 
@@
 Here and there in the thesis it has been hinted that Obsidian has warp level 
-capabilites and that these are not as integrated into the hierarchy as block, 
-thread and grid are. The warp abstractions offered also requires more work in 
+capabilites and that these are not as integrated into the hierarchy as the block, 
+thread and grid levels are. The warp abstractions offered also requires more work in 
 the code generating backend than the others, that correspond directly to CUDA concepts. 

### 2013-10-28 src2/thesis.tex
update

@@
 task-parallelism. Data parallelism is applicable where the input data can 
-be divided into chunks and operated upon independently. Data-parallelism is scalable
-meaning that usually as problem size increase, the potential for parallelism increase.
+be divided into chunks and operated upon independently. Data-parallelism is scalable,
+meaning that the potential for parallelism increases with problem size.
 
@@
 The patterns, while very powerful as abstractions, also provide tools for thinking about 
-parallel algorithms. The ArBB library support flat regular data-parallelism, but also 
+parallel algorithms. The ArBB library supports flat regular data-parallelism, but also 
 irregular and to some degree nested data-parallelism.
@@
 
-\noindent In this example above one block of 32 threads using no shared memory, is launched.
-
-A kernel that uses shared memory takes a form similar to whats shown below.
+\noindent In the example above one block of 32 threads, using no shared memory, is launched.
+
+A kernel that uses shared memory takes the form:
 
@@
 \noindent A portion of the input array is stored into shared memory. A 
-barrier synchronisation is used to ensure that all threads within the block 
+barrier synchronisation is used to ensure that each threads within the block 
 has stored its value into shared memory (only necessary if threads will 
-communicate across warp boundaries). After computing on the data locally 
+communicate across warp boundaries). After computing on the data locally, 
 it is written back into global memory.
@@
 programmer must be aware of in order to make optimal use of the GPU. 
-the NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
-to learn more about these details. Here is a list with explanations of some 
-of the practices that the guide deem most important.
+The NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
+to learn more about these details. Here is a list, with explanations of some 
+of the practices that the guide deems most important.
 
@@
 \item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
-and GPU is over 20 times of the bandwidth between host memory and the GPU device via the PCIe bus.
-This means that, first of all one must evaluate if the computation to be perform is significant 
-enough to warrant a transfer to the GPU. If it computation is not arithmetic intensive it may be 
+and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
+This means that, first of all one must evaluate if the computation to be performed is significant 
+enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
 better to do the work on the CPU. However, if the data is already in device memory and 
-the operation you are about to perform on it is more efficiently handled by the CPU, it may still 
+the operation that is about to perform on it is more efficiently handled by the CPU, it may still 
 be better to let the GPU do the work. In other words, compute near the data unless transferring it 
@@
 GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
-what these access patterns are very some depending on the GPU. For a representative model of 
-GPU the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
+what these access patterns are varies somewhat depending on the GPU. For a typical model of 
+GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
 to as many transactions as cache lines touched (128b l1 cache lines). This means that strided 
@@
 \item {\bf Minimise use of global memory:} Accessing global device memory can take as 
-many 400 to 600 clock cycles if the data is not cached. The programmer should 
+many as 400 to 600 clock cycles if the data is not cached. The programmer should 
 use shared memory as much as possible to avoid redundant loading from global memory.
@@
 600 cycles) it will be swapped out and another warp will take its place. This means that 
-the number of available warps decides how well latency can be hidden. A block that is not a 
-multiple of 32 leads to wasting GPU resources, some processing elements will stand idle during 
-the execution of the not full warps.
-
-\item {\bf Avoid different execution paths within a block:} Any branching instructions that 
+the number of available warps decides how well memory latency can be hidden. A block that is 
+not a multiple of 32 leads to wasting GPU resources; some processing elements will stand 
+idle during the execution of the not full warps.
+
+\item {\bf Avoid different execution paths within a block:} Any branching instruction that 
 causes the execution paths to diverge within a warp will affect performance. When the execution 
-paths within a warp diverges the computation is serialised.
+paths within a warp diverge, the computation is serialised.
 
@@
 Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
-this is a major limitation to the composability of CUDA functions. If it is safe or 
-not to call some function at a certain place in your code, can only be decided by inspecting 
-that functions implementation. 
+this is a major limitation to the composability of CUDA functions. Wheather or not it is safe to 
+call some function in a certain place in your code can be decided only by inspecting that function's 
+implementation. 
+%f it is safe or 
+%not to call some function at a certain place in your code, can only be decided by inspecting 
+%that functions implementation. 
 \end{itemize}
 
-The list above provides some very general guide lines of what to think about when 
+The list above provides some very general guidelines on what to think about when 
 writing GPU kernels. The lesson is that fast memory near the processing element 
@@
 us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
-are include making use of the host language's type system for guaranteeing generating correctly typed 
-target code. Already existing libraries in the host language (if they are general enough) 
+are include making use of the host language's type system to guarantee that correctly typed target code is generated. 
+Already existing libraries in the host language (if they are general enough) 
 can be used in the embedded language. The host language can also be used as a powerful macro 
@@
 level of abstraction in fields where the default is to use low-level languages. 
-For example, Lava and Wired in the fields of hardware design and 
-verification and Feldspar for digital signal processing~\citet{LAVA,Wired,FELDSPAR2010}. For GPU 
-programming, there are numerous embedded approaches using various host languages. 
-Two languages for GPU programming embedded in Haskell are  Accelerate and Nikola 
-\citet{ACCELERATEDAMP11, NIKOLA}. And there is Intel ArBB, embedded in C++, with the 
-goal of being parallel across platforms \citet{ARBB2011}.
-Related work and information about the tools, libraries and language that make the 
-wider context surrounding Obsidian can be found in section~\ref{sec:relatedwork}.
+For example, Lava and Wired~\citet{LAVA,Wired} raises the level of abstraction in the 
+fields of hardware design and verification and Feldspar does the same in the area of  digital 
+signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
+embedded approaches using various host languages. Two languages for GPU programming 
+embedded in Haskell are  Accelerate and Nikola \citet{ACCELERATEDAMP11, NIKOLA}. 
+And there is Intel ArBB, embedded in C++, with the goal of being parallel across 
+platforms \citet{ARBB2011}. Related work and information about the tools, libraries 
+and language that make the wider context surrounding Obsidian can be found in 
+section~\ref{sec:relatedwork}.
 
@@
 
-{\tt ForAll} takes a number of parallel iterations, a body represented by a function 
-from an index to a program. The body is a {\tt Program} at a some level {\tt t} while the 
-resulting program is at a level step above. This means a thread program can be turned into 
-a block program by using {\tt ForAll} or that a block program can be turned into a grid program.
+{\tt ForAll} takes a number of parallel iterations, and a body represented by a function 
+from an index to a program. The body is a {\tt Program} at some level {\tt t} while the 
+resulting program is a step above. This means that a thread program can be turned into 
+a block program by using {\tt ForAll}, or that a block program can be turned into a grid program.
 
@@
 Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
-types. Earlier versions of Obsidian (section~\ref{sec:paperB} used phantom types.
+types. Earlier versions of Obsidian (section~\ref{sec:paperB}) used phantom types.
 
@@
 or dynamic (length represented by an expression, an unknown length). When creating block and 
-warp level computations we want to know the exact lengths in order to avoid generating code 
-riddled with conditionals concerning array lengths. Conditionals are problematic on GPUs and 
-while general programs that run for any size is possible when performance is crucial a fixed 
-size is recommended.
-
-One benefit of pull arrays are that they give fusion of all operations on them for free. for example {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:
+warp level computations, we want to know the exact lengths in order to avoid generating code 
+riddled with conditionals concerning array lengths. It is possible to write kernels that 
+are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
+conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
+when performance is crucial. 
+%Conditionals are problematic on GPUs and 
+%while general programs that run for any size is possible when performance is crucial a fixed 
+%size is recommended.
+
+One benefit of pull arrays is that they give fusion of all operations on them for free. 
+For example, {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:
 
@@
 
-Now, we see that {\tt map f . map g} is the same as {\tt map f . g} since both result in {\tt f}
+Now, we see that \verb!map f . map g! is the same as \verb!map (f . g)! since both result in {\tt f}
 and {\tt g} becoming composed onto the indexing function. No intermediate array is built in 
@@
 Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
-good because they are so simple to understand and easy to parallelise. However certain
+good because they are so simple to understand and easy to parallelise. However, certain
 operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
@@
 interleaving case, which is worst, leads to diverging branching in every warp. If branches 
-diverge the GPU will execute turn off the processing elements taking one path and 
+diverge, the GPU will turn off the processing elements taking one path and 
 allow the others to progress; then turning to the other branch. The computation of the 
-branches has become serialised and compute resources are wasted.
+branches is serialised and compute resources are wasted.
 
@@
     Push n $ \wf -> 
-      forAll (fromIntegral n) $ \i -> wf (ixf i) i
+      ForAll (fromIntegral n) $ \i -> wf (ixf i) i
 \end{verbatim}
@@
 iteration schema over threads within a block. There are other iteration schemas to 
-chose from such as sequential, combinbations of sequential and paralell and parallel 
+chose from such as sequential, combinbations of sequential and parallel
 across both threads and blocks (two level parallelism).
 
-Push arrays give many of the same benefits that pull arrays have. Using pull 
+Push arrays give many of the same benefits that pull arrays have. Using push 
 arrays we also get fusion of operations as default. The implementation of map 
@@
 When compiling Obsidian programs to CUDA, we only need to be concerned with the 
-{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype 
+{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype, 
 we know that the AST can be quite directly translated to CUDA. We know that there will 
@@
 
-We have implemented basic algorithms such as sorting, merging, fractal generation, reductions and parallel prefix (scans). Often we have managed to get very good performance, close to that 
-of hand optimised CUDA code. However, there are also cases where we just cant seem to reach 
+We have implemented basic algorithms such as sorting, merging, fractal generation, reductions 
+and parallel prefix (scans). Often we have managed to get very good performance, close to that 
+of hand optimised CUDA code. However, there are also cases where we just cannot seem to reach 
 the level of performance that we would want. One such example is parallel prefix, an 
 application that has been haunting us from the very beginning. Scan is problematic to us because 
-it is best implemented as an in-place algorithm while our array representations do 
-not allow for in-place updated (there is no memory associated with them to update). 
+it is best implemented as an in-place algorithm, while our array representations do 
+not allow for in-place update; there is no memory associated with them to update. 
 We are trying to improve this situation by adding a third array representation to Obsidian,
@@
 
-Paper F (section~\ref{sec:paperF}), tells a story about how to implement and tune 
-a reduction kernel in Obsidian. At the end the implemented reduction kernel is 
+Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
+a reduction kernel in Obsidian. At the end, the implemented reduction kernel is 
 faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
-reduction in accelerate corresponds to just one single hand optimised skeleton. We would have 
+reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
 been happy just getting close.
@@
 
-Paper D (section~\ref{sec:paperD}), introduces push arrays and uses them in the 
+Paper D (section~\ref{sec:paperD}) introduces push arrays and uses them in the 
 implementation of sorting networks to quite good results. With push arrays we 
-where able to experiment with more detailed decompositions (what thread computes 
+were able to experiment with more detailed decompositions (what thread computes 
 what value) of the algorithms.
@@
 
-EmbArBB provides many of the ArBB operations on Arrays butprovides a Repa like 
-shape-polymorphic interface to them. For example a reduction operation can take a two  
+EmbArBB provides many of the ArBB operations on vectors but provides a 
+shape-polymorphic interface to them. For example, a reduction operation can take a two  
 dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
@@
 
-The {\tt addReduce} function takes an input vector that is atleast one dimensional 
+The {\tt addReduce} function takes an input vector that is at least one dimensional 
 ({\tt t:.Int}) and returns a vector that has one less dimension ({\tt t}).
@@
 
-EmbArBB is a traditional deep embedding, there is an AST data type with constructors 
+EmbArBB is a traditional deep embedding; there is an AST data type with constructors 
 representing each ArBB language feature. It is true that a deep embedding of ArBB 
@@
 
-In paper H ,section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
+In paper H, section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
 Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
@@
 adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
-the code, that code is identical no matter if the AST was build via Haskell or C++ calls. 
+the code, that code is identical no matter if the AST was built via Haskell or C++ calls. 
 Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
-takes to go through the process of making those ArBB C API calls. Although if the generated 
+takes to go through the process of making those ArBB C API calls. However, if the generated 
 code is used many times the caching ArBB does of generated code means that the initial code 
@@
 
-In figure~\ref{fig:researchmatrix} our work is placed in relation to other languages,
-embedded languages and libraries for parallel programming. The languages and libraries consider 
-are data-parallel; languages for control parallelism or distributed message passing system are 
+In figure~\ref{fig:researchmatrix}, our work is placed in relation to other languages,
+embedded languages and libraries for parallel programming. The languages and libraries considered 
+are data-parallel; languages for control parallelism or distributed message passing systems are 
 not considered. The systems are roughly divided 
-into three groups, based on their level of abstraction. There are low-level languages, here 
+into three groups, based on their level of abstraction. There are low-level languages; here 
 I place languages that are imperative and C-like. In the middle layer I place languages 
-that go a bit further, they have higher level abstractions and are more easily composable. 
-In the highest level I place languages that completely abstract away from details of the 
+that have higher level abstractions and are more easily composable. 
+at the highest level, I place languages that completely abstract away from details of the 
 hardware they run on. The division is not free from personal bias and the placement of languages
@@
 % ---------------------------------------------------------------------------
-With Obsidian we try to target the sparsely occupied area of mid-level GPU programming. With 
+With Obsidian, we try to target the sparsely occupied area of mid-level GPU programming. With 
 CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
@@
 high-level patterns without direct insight into how the application would be mapped onto 
-threads and blocks of the GPU. Using Obsidian we want to bridge that gap by both being 
-more composable than CUDA but give the programmer more control of what the computation will 
+threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
+more composable than CUDA and giving the programmer control of what the computation will 
 look like on the GPU. This is done by allowing the programmer to specify and compose 
@@
 Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
-they are not general primitives. Where Accelerate have a single higher order reduction 
-operation, ArBB/EmbArBB have {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
+they are not general primitives. Where Accelerate has a single higher order reduction 
+operation, ArBB/EmbArBB has {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
 having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
@@
 Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
-a data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
+data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
 operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
@@
 Repa uses the same array representation as Obsidian, in Repa called a delayed
-array, which is a function from index to element. However, in Repa arrays have shape, 
-they can be of any dimensionality and there shape-polymorphic functions on these. 
-For example {\tt map} can be applied to an array of any shape.
-
-Meta-Repa is a reimplementation of the Repa library using deep embedded language 
+array, which is a function from index to element. However, in Repa arrays have shape; 
+they can be of any dimensionality and there are shape-polymorphic functions on these. 
+For example, {\tt map} can be applied to an array of any shape.
+
+Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
 techniques. Meta-Repa programs build ASTs that are then compiled (using template 
@@
 The Meta-Repa approach to implementing the Repa library 
-gives inlining of operations for free, where Repa programs needs to be annotated 
+gives inlining of operations for free, where Repa programs need to be annotated 
 with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
@@
 
-When we first started working on Obsidian we used Lava~\citet{LavaICFP} as a inspiration. 
+When we first started working on Obsidian, we used Lava~\citet{LavaICFP} as inspiration. 
 We wanted to mimic the Lava programming style but generate GPU code. At first this worked quite 
@@
 
-In paper A~(\ref{sec:paperA}) we describe our method for compiling monadic embedded languages. 
-This means we need to take a monadic computation and observe its structure in order to generate
+In paper A~(section \ref{sec:paperA}) we describe our method for compiling monadic embedded 
+languages. 
+This means to take a monadic computation and observe its structure in order to generate
 code. This work is closely related to a method described in ``The Constrained Monad Problem''
-by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In this paper they solve a more 
-general problem than just compilation but our method is nice because it is easy to understand 
+by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper they solve a more 
+general problem than just compilation, but our method is nice because it is easy to understand 
 and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
@@
 from within Haskell; the recommended workflow, however, is to generate the CUDA  
-as text and write the host code by hand. This can be seen both a shortcoming 
+as text and write the host code by hand. This can be seen as both a shortcoming 
 and a feature. It is hard to integrate Accelerate programs in larger applications 
-written in other languages. While when using Obsidian a kernel is generated and 
+written in other languages; when using Obsidian a kernel is generated and 
 can be used as part of for example a library.
@@
 and at little effort from us, very good performance is gained. The performance of 
-EmbArBB is entirely due to all the optimisation expertice and effort that invested in 
-the ArBB system. We get all that expertice for free. What EmbArBB provide is a 
-way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB is based on a 
+EmbArBB is entirely due to all the optimisation expertice and effort that was invested in 
+the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
+way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
 set of parallel primitives for the programmer to use and compose. Obsidian takes 
 a different approach. Built in primitives are very few and very low level, such as 
-sequential and parallel for loops. On top of the low-level capabilities a library 
+sequential and parallel for loops. On top of the low-level capabilities, a library 
 of high-level primitives can be built. The key is that we want the programmer to 
@@
 enough that there be no ``the right'' way of implementing these primitives. As of 
-today, the programmer still needs to explore the design space and find a sollution 
+today, the programmer still needs to explore the design space and find a solution 
 that performs well on a particular GPU.
@@
 Performance of Obsidian kernels is often good and the Obsidian code that they 
-are generated from is often shorted than corresponding CUDA programs. Obsidian 
+are generated from is often shorter than corresponding CUDA programs. Obsidian 
 programs differ from CUDA programs by describing whole programs. With this I mean 
@@
 
-Push arrays where invented by Koen Claessen and first implemented in Obsidian 
+Push arrays were invented by Koen Claessen and first implemented in Obsidian 
 to provide more control over where computation takes place (which thread computes 
 which element). Push arrays has since been experimented on by others and show up 
-in Nikola~\citet{NIKOLAPUSH}, been experimented upon by the people behind Feldspar 
-and appear in StreamHS~\citet{FPCDSL}.
+in Nikola~\citet{NIKOLAPUSH}, Feldspar and appear in StreamHS~\citet{FPCDSL}.
 

### 2013-10-29 src2/thesis.tex
update

@@
 \subsection{Writing efficient GPU kernels} 
+\label{sec:efficient}
 
@@
 \subsection{Results}
+\FloatBarrier
 
@@
 
+In section \ref{sec:efficient}, guidelines for writing efficient CUDA code is listed. 
+Here we revisit these guidelines and explain how they are approached by Obsidian and 
+the Obsidian programmer.
+
+\begin{itemize}
+
+\item {\bf Minimise data transfers between host and device:} Data transfers
+between host and device are out of Obsidian's scope. If Obsidian is used to create 
+a kernel for a computation, the programmer has already decided that a GPU should be used.
+
+\item {\bf Ensure that global memory accesses are coalesced:} Inputs to an Obsidian 
+program are pull arrays. Pull arrays can be arbitrarily permuted; the permutation 
+used will decide the access pattern. If this method can be applied or not partly also 
+depends on the algorithm being implemented.
+
+\item {\bf Minimise use of global memory:} Intermediate results can be computed into 
+shared memory. Arrays are stored into shared memory when the programmer uses a 
+{\tt force} primitive.
+
+\item {\bf Use a multiple of 32 threads per block:} Obsidian requires regularity even more 
+than CUDA does. Since we focus on generating high performance kernels, we rule out generation 
+of kernels that work on different input sizes. The choise of how many threads to use per block 
+is in the hands of the Obsidian program. The method for making that choise has changed over time. 
+Earlier, the number of threads needed by kernel was decided by the size of the arrays it 
+computed. This has recently changed is now set by parameters to the code generation process.
+
+\item {\bf Avoid different execution paths within a block:} Sometimes divergent code cannot 
+be avoided. However, It is important not to have threads diverge unnecessarily. When using pull 
+arrays alone this was often a problem; with the addition of push arrays the situation improved. 
+Using push arrays a divergent computation can be explicitly split up into phases, each with all 
+threads following the same execution path. This means that fewer threads are required to do the 
+same work.
+
+\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} In Obsidian it is impossible 
+for a barrier synchronisation to appear within conditional code. Conditionals can be introduced 
+either by a conditional expression or a {\tt Cond} operation in the program datatype. The Cond 
+operation chooses between two thread programs and the synchronisation operation is at the block 
+level in the hierarchy. So a synchronisation cannot occur in conditional code. 
+%f it is safe or 
+%not to call some function at a certain place in your code, can only be decided by inspecting 
+%that functions implementation. 
+\end{itemize}
+
+%% In the book ``Structured Parallel Programming''~\citet{STRUCTURED}, there is a list showing language 
+%% support for various parallel patterns. In figure~\ref{fig:caps} this list augmented with a column for Obsidian. 
+%% This figure is added to indicate relative expressive power of the different languages. The most 
+%% interesting comparison is between Obsidian and OpenCL that both have a programming model geared towards
+%% GPUs; even though OpenCL can compile for both CPU and GPU execution. The paragraphs below 
+%% take a closer look at well Obsidian performes in the patterns.
+
+%% \subsubsection{Map}
+
+%% Map, to a  functional programmer is a higher order function that takes a pure 
+%% function and applies it to all elements of a collection (for example a list). 
+%% That meaning of map is easily implementable in Obsidian on both pull and push arrays. 
+%% Obsidian is also capable, via its parallel for loops, to map sequential computations 
+%% over threads in a block and block computation over many blocks. This is probably 
+%% more in the spirit of what the authors refer to in~\citet{STRUCTURED}.
+
+%% \subsubsection{Stencil}
+
+%% Stencil computations can be expressed in Obsidian, with some manual labour. 
+%% A stencil computation would have the form of a pull array to pull array function. 
+%% Pull arrays allow reading at position. 
+%% Stencil operations where more than one output position is touched could be 
+%% implemented as a pull array to push array function.
+%% One problem with stencil computations is the borders. If every thread needs to 
+%% execute a conditional, checking if it is at the border, performance would suffer. 
+%% Using Obsidian the programmer would likely generate more than one kernel; a fast 
+%% kernel for the internal points and one or more border kernel.
+
+%% \subsubsection{Reduction and scan} 
+%% Reductions can easily be implemented in Obsidian and with very good performance. 
+%% Obsidian also seems particularly well suited for experimenting with the kind 
+%% of trade offs that are interesting when implementing reduction kernels.
+
+%% Scans can also be implemented and the code is elegant.
+
+%% \subsubsection{Pack and expand}
+
+%% \begin{figure}[!ht]
+%% \begin{small} 
+%% \begin{tabular}{|l|l|l|l|l|l|l|}
+%% Parallel Pattern & TBB & Cilk Plus & OpenMP & ArBB & OpenCL & Obsidian \\
+%% \hline 
+%% Parallel Nesting & F   & F         &        &      &        &          \\
+%% Map              & F   & F         & F      & F    & F      & I        \\
+%% Stencil          & I   & I         & I      & F    & I      & I        \\ 
+%% Workpile         & F   &           &        &      & I      &          \\
+%% Reduction        & F   & F         & F      & F    & I      & I        \\ 
+%% Scan             & F   & I         & I      & F    & I      & I        \\ 
+%% Fork-Join        & F   & F         & I      &      &        &          \\ 
+%% Recurrance       &     & P         &        &      &        &          \\ 
+%% Superscalar seq. &     &           &        &      & F      &          \\
+
+%% Pack             & I   &  I        &  I     & F    & I      & I        \\ 
+%% Expand           & I   &  I        &  I     & I    & I      & *        \\
+%% Pipeline         & F   &  I        &  I     &      &        &          \\
+%% Geometric decomp. & I  &  I        &  I     & I    & I      & I        \\
+%% Search           &  I  &  I        &  I     & I    & I      & I        \\
+%% Category reduction & I &  I        &  I     & I    & I      & I        \\
+%% Gather           & I   &  F        &  I     & F    & I      & I        \\
+%% Atomic scatter   & F   &  I        &  I     &      & I      & I        \\
+%% Permutation scatter & F & F        &  F     & F    & F      & I        \\ 
+%% Merge scatter    &  I  &  I        &  I     & F    & I      & *        \\
+%% Priority scatter &     &           &        &      &        & *        \\
+
+%% \end{tabular}
+%% \end{small}
+%% \label{fig:caps}
+%% \caption{This figure lists what parallel patterns a number of languages support. The list 
+%% is copied from ``Structured Parallel Programming''~{\protect\citet{STRUCTURED}} and augmented with a 
+%% column for Obsidian. \\
+%% {\bf Key:} F - means the pattern is supported directly with an abstraction. P - indicates 
+%% that the pattern can be implemented in terms of other patterns appearing below. I - means 
+%% that the pattern can be easily implemented. Lastly, (*) indicates some patterns 
+%% that can be implemented in Obsidian but currently not efficiently so.} 
+%% \end{figure}
+
+%% In the technical report ``The Landscape of Parallel Computing Research: A View 
+%% from Berkeley''~\citet{dwarfs}, 13 computational kernels (called dwarfs) of importance is identified. 
+%% The paper also proposes that we use these 13 dwarfs to evaluate parallel programming 
+%% models and languages. The dwarfs are listed below, together with thoughts on how 
+%% well they can be tackled by Obsdian.
+
+%% \begin{itemize} 
+%% \item {\bf Dense Linear Algebra:} Obsidian works well for expressing vector-vector 
+%% operations. It is also possible to express operations such as matrix multiplication, 
+%% even though Obsidians support for multidimensional arrays is limited. It is possible 
+%% to use pull arrays of pull arrays, but not push arrays of push arrays.
+
+%% \item {\bf Sparse Linear Algebra:}
+
+%% \item {\bf Spectral Methods (FFT):} FFTs are implemented in a way that is 
+%% similar to sorting networks. A so-called butterfly network is used. We know 
+%% that we can express these network structures in Obsidian.
+
+%% \item {\bf N-Body Methods:}
+
+%% \item {\bf Structured Grids:} Regular grids should not be a problem for Obsidian. 
+%% Obsidian has no built in support for stencils, the combination of pull and push array 
+%% still provides the capability to express such computations.
+
+%% \item {\bf Unstructured Grids:} May be hard problem for GPUs in general. The paper 
+%% states that these problems have the property that ``Updates typically involve multiple 
+%% levels of memory reference indirection''. We know that the GPU thrives when memory access 
+%% patters are predictable and regular, which is the opposite scenario.
+
+%% \end{itemize}
+
+\FloatBarrier 
 % ---------------------------------------------------------------------------
@@
 \label{sec:paperF}
+
 %\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
@@
 Bo Joel Svensson, Mary Sheeran
+\end{center}
+
+\begin{center}
+{\large \bf This paper will be revised before printing the Thesis} 
 \end{center}

### 2013-10-30 src2/thesis.tex
update
p

@@
 which element). Push arrays has since been experimented on by others and show up 
-in Nikola~\citet{NIKOLAPUSH}, Feldspar and appear in StreamHS~\citet{FPCDSL}.
+in Nikola~\citet{NIKOLAPUSH} and Feldspar. Push arrays appear in StreamHS~\citet{FPCDSL} 
+and ~\citet{MOA} as part of a multi-dimensional array calculus in Standard ML.
 

### 2013-10-30 src2/thesis.tex
update

--- 2013-10-29 src2/thesis.tex
@@
 
+\vspace{5mm} 
+\emph{This section will be revised. I want to add more code examples.} 
+\vspace{5mm}
+
 Obsidian aspires to raise the level of abstraction of GPU kernel implementation, 
@@
 the Obsidian programmer.
+
+\vspace{5mm}
+
+\emph{This list below will be revised. More details or code will be added.}
+
+\vspace{5mm}
 
@@
 
+\vspace{5mm}
+
+\emph{This section will revised.}
+
+\vspace{5mm}
+
 This thesis contains two approaches to embedding a language for data-parallel programming 
@@
 which element). Push arrays has since been experimented on by others and show up 
-in Nikola~\citet{NIKOLAPUSH}, Feldspar and appear in StreamHS~\citet{FPCDSL}.
+in Nikola~\citet{NIKOLAPUSH} and Feldspar~\citet{FELDSPARPUSH}. Push arrays appear 
+in StreamHS~\citet{FPCDSL} and in ~\citet{MOA} as part of a multi-dimensional array 
+calculus in Standard ML.
 

### 2013-11-03 src2/hl/paperThesis.tex
update

@@
     implement some GPU hierarchy generic functions (section \ref{sec:Compile}).
-  \item A simple and compositional method for reification of monads in the 
-    setting of a compiled embedded language (section \ref{sec:Compile}).
-
+ % \item A simple and compositional method for reification of monads in the 
+ %   setting of a compiled embedded language (section \ref{sec:Compile}).
 \end{itemize}
@@
 concepts. A GPU can come with as few as one of these multiprocessors.
-The two GPUs used in our measurements are the GTX 680, which has eight multiprocessors, with a total of 1536 processing cores, and the GT 650M, a laptop GPU, which has 384 such cores.
-On these cores, groups of 32 threads called {\em warps} are scheduled. There are a number 
-of warp scheduling units per multiprocessor. Within a warp, threads execute in 
-lockstep (SIMD); diverging branches, that is those that take different 
-paths on different threads within a warp, are serialised, 
-leading to performance penalties.
+The two GPUs used in our measurements are the GTX 680, which has eight 
+multiprocessors, with a total of 1536 processing cores, and the GT 650M, a 
+laptop GPU, which has 384 such cores. On these cores, groups of 32 threads 
+called {\em warps} are scheduled. There are a number of warp scheduling units 
+per multiprocessor. Within a warp, threads execute in lockstep (SIMD); 
+diverging branches, that is those that take different paths on different 
+threads within a warp, are serialised, leading to performance penalties.
 
@@
 The threads within a block can use the shared memory of the multiprocessor 
-to communicate with each other. A synchronisation primitive, {\tt \_\_syncthreads()}, gives all the threads within a block a coherent view
-of the shared memory. There is no similar 
-synchronisation primitive between threads of different blocks.
+to communicate with each other. A synchronisation primitive, {\tt \_\_syncthreads()}, 
+gives all the threads within a block a coherent view of the shared memory. 
+There is no similar synchronisation primitive between threads of different blocks.
 
@@
 in terms of the unique identifier for that block and thread.
-Some access patterns allow memory reads to be \emph{coalesced}, while others do not, giving very poor performance. The patterns that lead to good performance vary somewhat
-between different GPU generations, but regular, consecutive accesses by consecutive threads within a warp are best.
-
-A CUDA program is expressed at two levels. Kernels are data-parallel programs that run on the GPU. They are launched by the controlling 
-program, which runs on the CPU of the host machine.
+Some access patterns allow memory reads to be \emph{coalesced}, while others 
+do not, giving very poor performance. The patterns that lead to good performance 
+vary somewhat between different GPU generations, but regular, consecutive accesses 
+by consecutive threads within a warp are best.
+
+A CUDA program is expressed at two levels. Kernels are data-parallel programs 
+that run on the GPU. They are launched by the controlling program, which runs 
+on the CPU of the host machine.
 
@@
 \end{small}
-\caption{Code generated from the {\tt reverses} program with the local reverse instantiated with 
-{\tt reverseL} and {\tt reverseL'}. 
- {\tt reverse'} stores the array in shared memory before copying it to the global result array.} 
+\caption{Code generated from the {\tt reverses} and {\tt reverses'} programs.
+ {\tt reverses'} stores the array in shared memory before copying it to the global result array.} 
 \label{fig:rev1} 
@@
 The following Obsidian code shows how to split a large array up into 
-chunks of 256 elements ({\tt splitUp 256 arr}) and how the local reverse 
-kernel is mapped, in parallel, over blocks using {\tt pConcatMap}.
+chunks of 256 elements ({\tt splitUp 256 arr}) and how the local {\tt reverseL} 
+and {\tt reverseL'} kernels are  mapped, in parallel, over blocks using {\tt pConcatMap}.
 
@@
 reverses :: DPull a -> DPush Grid a
-reverses arr = pConcatMap reverseL $ splitUp 256 arr 
+reverses arr = pConcatMap (pJoin . reverseLocal)
+                          (splitUp 256 arr)
+
+reverses' :: MemoryOps a => DPull a -> DPush Grid a
+reverses' arr = pConcatMap (pJoin . reverseLocal')
+                           (splitUp 256 arr) 
 \end{verbatim} 
@@
 pConcatMap :: ASize l 
-              => (a1 -> Program t (SPush t a)) 
+            => (a1 -> SPush t a) 
             -> Pull l a1 
@@
 into a push array at the level above {\tt t}. Here, this is used to apply a block 
-level computation, giving a grid computation.
+level computation, giving a grid computation. The {\tt pJoin} function 
+has type \verb!Program t (Push t s a) -> Push t s a!; it creates a new push array 
+that computes the same results as the input computation.
 
@@
 \begin{verbatim} 
-largeReverse :: DPull a
-                -> Push Grid EWord32 a
+largeReverse :: DPull b
+              -> Push Grid EWord32 b
 largeReverse arr =
-  pConcat $ reverse $ pMap reverseL $ splitUp 256 arr 
+  pConcat $ reverse $ fmap (pJoin . reverseLocal)
+                      (splitUp 256 arr) 
 \end{verbatim} 
@@
 when applied to a pull array, writes the elements of that array 
-into shared memory using as many threads as the pull array is long. Hence, 
-the {\tt force} function can be used to express parallelism in Obsidian.
+into shared memory using as many threads as needed from those available. 
+If the array is larger than the number of threads made available, the 
+force operation is split into two or more passes. This is a change compared
+to earlier versions of Obsidian where the length of arrays we could force 
+where limited to the maximum number of threads. In this version, the 
+threads are, in a sense, virtual. The mapping from virtual threads to 
+real threads is done as a step in the compilation.
 
@@
 (a runtime value). Static lengths are used for local (or block) 
-computations, with those lengths determining the number of threads needed and the shared memory consumption. Dynamic 
+computations, with those lengths determining shared memory consumption. Dynamic 
 lengths are used for global computations (across many blocks). After specifying 
@@
 The {\tt Program} data type has a parameter, {\tt t}, that can be 
-either {\tt Thread}, {\tt Block} or {\tt Grid}, corresponding to the GPU hierarchy, as shown in Figure~\ref{fig:Hierarchy}.
+either {\tt Thread}, {\tt Block} or {\tt Grid}, corresponding to 
+the GPU hierarchy, as shown in Figure~\ref{fig:Hierarchy}.
 
@@
 to convert a push array into a pull array. 
-The type of {\tt force} is: 
+With some class constraints removed for simplicity, the type of {\tt force} is: 
 \begin{small} 
 \begin{Verbatim}[samepage=true] 
-force :: (Write p, MemoryOps a) 
-         => p Word32 a 
+force :: ( ... ) 
+       => arr Word32 a 
        -> Program Block (Pull Word32 a)
@@
 \noindent
-where {\tt p} is an array (push or pull) with elements that can be stored into memory. 
+where {\tt arr} is an array (push or pull) with elements that can be stored into memory. 
 Note that the return type of {\tt force} is a program yielding a pull array. This pull 
@@
 
-Converting a pull array to a push array is cheap and is done using a function called {\tt push} that behaves differently (sequentially or in parallel) at different levels
+Converting a pull array to a push array is cheap and is done using a function called 
+{\tt push} that behaves differently (sequentially or in parallel) at different levels
 of the GPU hierarchy. This is implemented as a type class:
@@
 
-After compiling the monadic {\tt Program} into a list of statements, 
-some characteristics are collected.
-
-The first analysis done on the {\tt IM} 
-discovers the number of threads needed per block to run the program. This 
-is done by a pass over the {\tt IM} and looking at the size  
-of each parallel {\tt SForAll} loop. The largest {\tt SForAll} loop present 
-decides how many threads are needed. The implementation of any other 
-{\tt SForAll} loop will be wrapped in a conditional that turns 
-off any unused threads.
-
-Note that there is no similar analysis to discover the maximum extents 
-of {\tt SForAllBlocks} loops. The reason is that the restriction
-to kernels of the form:
-
-\begin{small} 
-\begin{Verbatim}[samepage=true]
-myKernel :: a0 -> ... -> aN -> DPush Grid b 
-\end{Verbatim} 
-\end{small}
-
-\noindent means that there can be no sequence of {\tt SForAllBlocks} loops in the program. In other 
-words, an Obsidian Program for which we can generate code will contain at most 
-one {\tt SForAllBlocks} loop. Some care has also to be taken with
-the language constructs that result in grid push arrays, so that none of these 
-creates illegal programs.
+%% After compiling the monadic {\tt Program} into a list of statements, 
+%% some characteristics are collected.
+
+%% The first analysis done on the {\tt IM} 
+%% discovers the number of threads needed per block to run the program. This 
+%% is done by a pass over the {\tt IM} and looking at the size  
+%% of each parallel {\tt SForAll} loop. The largest {\tt SForAll} loop present 
+%% decides how many threads are needed. The implementation of any other 
+%% {\tt SForAll} loop will be wrapped in a conditional that turns 
+%% off any unused threads.
+
+%% Note that there is no similar analysis to discover the maximum extents 
+%% of {\tt SForAllBlocks} loops. The reason is that the restriction
+%% to kernels of the form:
+
+%% \begin{small} 
+%% \begin{Verbatim}[samepage=true]
+%% myKernel :: a0 -> ... -> aN -> DPush Grid b 
+%% \end{Verbatim} 
+%% \end{small}
+
+%% \noindent means that there can be no sequence of {\tt SForAllBlocks} loops in the program. In other 
+%% words, an Obsidian Program for which we can generate code will contain at most 
+%% one {\tt SForAllBlocks} loop. Some care has also to be taken with
+%% the language constructs that result in grid push arrays, so that none of these 
+%% creates illegal programs.
 
@@
 
-In the current {\tt IM}, all arrays in shared memory will have unique 
+In the {\tt IM}, all arrays in shared memory will have unique 
 names. These names are obtained via the value supply during the reification 
@@
 restrictions. After the arrays are assigned addresses in the shared memory,
-a renaming phase replaces array names with these locations. This is done 
-while also going from the {\tt IM} data type to a third internal representation 
-of the program. This third representation is a C-like AST extended with some 
-GPU specific concepts (just as CUDA is an extended C). We strive to make this representation sufficiently general to be turned into 
-either CUDA code or OpenCL code, but currently only output CUDA.
+a renaming phase replaces array names with these locations.
+
+The last step in compilation process uses the Language.C.Quote library to 
+turn the {\tt IM} into CUDA code. This compilation stage takes a {\em configuration} 
+as parameter. The configuration tells how many actual threads per block the 
+CUDA code should be generated for. At this stage parallel for loops (\tt SForAll) 
+are turned into more than one parallel for loop in sequence, if needed.
+
+%This is done 
+%while also going from the {\tt IM} data type to a third internal representation 
+%of the program. This third representation is a C-like AST extended with some 
+%GPU specific concepts (just as CUDA is an extended C). We strive to make this representation sufficiently general to be turned into 
+%either CUDA code or OpenCL code, but currently only output CUDA.
 
@@
 
-The following case studies show first a typical application, the Mandelbrot fractal, in which the required work is naturally highly parallel.
-The remaining case studies consider key building blocks, reduce and scan, that have data-flow graphs involving much more communication.
-Then, the programmer must find ways to decompose the algorithm in order to fit within
-the constraints imposed by the GPU architecture. These two case studies illustrate the use of Obsidian for exploring the design space when implementing a particular algorithm.
+The following case studies show first a typical application, the Mandelbrot 
+fractal, in which the required work is naturally highly parallel.
+The remaining case studies consider key building blocks, reduce and scan, 
+that have data-flow graphs involving much more communication. Then, the 
+programmer must find ways to decompose the algorithm in order to fit within
+the constraints imposed by the GPU architecture. These two case studies illustrate 
+the use of Obsidian for exploring the design space when implementing a particular algorithm.
 
@@
           => l 
-            -> (EWord32 -> Program t (SPush t b))
+          -> (EWord32 -> SPush t b) 
           -> Push (Step t) l b
@@
            -> Word32
-           -> (EWord32 -> EWord32 
-                 -> TProgram (SPush Thread b))
+           -> (EWord32 -> EWord32 -> SPush Thread b)
            -> DPush Grid b 
-genRect bs ts p = 
-   generate bs (\bid -> return $ generate ts (p bid))
+genRect bs ts p = generate bs $
+                  \bid -> generate ts $ p bid 
 \end{verbatim}
@@
 \begin{verbatim}
-mandel = genRect 512 512 iters 
+mandel = genRect 512 512 (\b -> pJoin . iters b)
 \end{verbatim}
@@
               -> DPush Grid a
-mapRed1 f = pConcatMap $ red1 f
+mapRed1 f = pConcatMap $ pJoin . red1 f
 \end{Verbatim}

### 2013-11-04 src2/hl/paperThesis.tex
update

@@
 the low-level details that are needed for performance.
-So far,
-we have chosen not to try to apply automatic transformations or compiler 
+
+So far, we have chosen not to try to apply automatic transformations or compiler 
 optimisation techniques to a GPU architecture agnostic program and hope 
@@
 correct decisions into the hands of the programmer. From Obsidian programs, 
-CUDA kernels are generated, while host code must still be written in CUDA.
+CUDA kernels are generated; host code can be written in Obsidian, but is still 
+preferably written in CUDA.
+
 For key building blocks like scan and reduce, we see a trend towards
@@
 as parameter. The configuration tells how many actual threads per block the 
-CUDA code should be generated for. At this stage parallel for loops (\tt SForAll) 
+CUDA code should be generated for. At this stage parallel for loops ({\tt SForAll})
 are turned into more than one parallel for loop in sequence, if needed.
@@
 
-In reference ~\citehl{CSORT}, we experiment with atomic operations 
+In reference ~\citehl{JOSEFJOELCSORT}, we experiment with atomic operations 
 and are forced to write our code in a very imperative style. The addition of mutable 

### 2013-11-04 src2/bb/body.tex
update

@@
 
-\pagebreak
+% \pagebreak
 \begin{small} 
@@
 \end{small}
+
+\pagebreak
 
@@
     s = newNameSupply
-
+\end{verbatim}
+\end{small}
+
+\begin{small}
+\begin{verbatim}
 compile :: NameSupply -> Program a -> (a, Prg) 
@@
 compile s TurnLeft  = ((),PTurnLeft)
+\end{verbatim}
+\end{small}
+
+\pagebreak
+
+\begin{small}
+\begin{verbatim}
 compile s Sensor    = (Var nom,PSensor nom)
@@
 \end{verbatim}
-\pagebreak
+% \pagebreak
 \begin{verbatim}

### 2013-11-06 src2/hl/paperThesis.tex
update

@@
 \begin{figure} 
-\def\s{.50}
+\def\s{.75}
 \begin{tikzpicture}[scale=(\s)]
@@
 
-\node [wbox,inner ysep=15pt,right=0cm of gd.north east, anchor=north west] (obs1) {% 
+\node [wbox,inner ysep=19.5pt,right=0cm of gd.north east, anchor=north west] (obs1) {% 
   \begin{minipage}{0.35\linewidth}
@@
 
-\node [wbox,inner ysep=15pt,below=0cm of obs1.south] (obs2) {% 
+\node [wbox,inner ysep=14pt,below=0pt of obs1.south] (obs2) {% 
   \begin{minipage}{0.35\linewidth}
@@
 };
-
-\node [wbox,inner ysep=9.5pt, below=0cm of obs2.south] (obs3) {% 
+%below=0cm of obs2.south
+\node [wbox,inner ysep=13.5pt, below=0pt of obs2.south] (obs3) {% 
   \begin{minipage}{0.35\linewidth}
-        {\small seqFor, seqUntil \\ seqReduce\\ seqScan, SPush, SPull}
+        {\small seqFor, seqUntil, seqReduce\\ seqScan, SPush, SPull}
     \end{minipage}
@@
              -> Exp a 
-
+\end{Verbatim}
+\end{small}
+
+\pagebreak
+
+\begin{small}
+\begin{Verbatim}             
   If      :: Scalar a 
@@
 \bibliographyhl{thesis}
+\addcontentsline{toc}{subsection}{Bibliography}
 

### 2013-11-06 src2/thesis-backup.tex
update

--- 2013-10-30 src2/thesis.tex
@@
 \documentclass[a4paper]{book}
+
+\usepackage[paperwidth=169mm, paperheight=239mm]{geometry}
 
@@
 
+\pagebreak
+
 The syntax for launching a kernel is:
@@
 us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
-are include making use of the host language's type system to guarantee that correctly typed target code is generated. 
+include making use of the host language's type system to guarantee that correctly typed target code is generated. 
 Already existing libraries in the host language (if they are general enough) 
@@
 \node[anchor=west] (repa) at (0,2.6) {DPH\\Nesl\\Repa\\Meta-repa};
-\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
+\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB\\ Feldspar};
 \node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
@@
 
-\begin{center}
-{\large \bf This paper will be revised before printing the Thesis} 
-\end{center}
+%\begin{center}
+%{\large \bf This paper will be revised before printing the Thesis} 
+%\end{center}
 
@@
 % ---------------------------------------------------------------------------
-\nocite{*}
-
-\makeatletter
-\renewenvironment{thebibliography}[1]
-     {\chapter*{\bibname}%
-      \@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
-      \list{\@biblabel{\@arabic\c@enumiv}}%
-           {\settowidth\labelwidth{\@biblabel{#1}}%
-            \leftmargin\labelwidth
-            \advance\leftmargin\labelsep
-            \@openbib@code
-            \usecounter{enumiv}%
-            \let\p@enumiv\@empty
-            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
-      \sloppy
-      \clubpenalty4000
-      \@clubpenalty \clubpenalty
-      \widowpenalty4000%
-      \sfcode`\.\@m}
-     {\def\@noitemerr
-       {\@latex@warning{Empty `thebibliography' environment}}%
-       \endlist}
-\makeatother
-
-\bibliographystyle{alpha}
-\bibliography{thesis}
-\addcontentsline{toc}{chapter}{Bibliography}
+%% \nocite{*}
+
+%% \makeatletter
+%% \renewenvironment{thebibliography}[1]
+%%      {\chapter*{\bibname}%
+%%       \@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
+%%       \list{\@biblabel{\@arabic\c@enumiv}}%
+%%            {\settowidth\labelwidth{\@biblabel{#1}}%
+%%             \leftmargin\labelwidth
+%%             \advance\leftmargin\labelsep
+%%             \@openbib@code
+%%             \usecounter{enumiv}%
+%%             \let\p@enumiv\@empty
+%%             \renewcommand\theenumiv{\@arabic\c@enumiv}}%
+%%       \sloppy
+%%       \clubpenalty4000
+%%       \@clubpenalty \clubpenalty
+%%       \widowpenalty4000%
+%%       \sfcode`\.\@m}
+%%      {\def\@noitemerr
+%%        {\@latex@warning{Empty `thebibliography' environment}}%
+%%        \endlist}
+%% \makeatother
+
+%% \bibliographystyle{alpha}
+%% \bibliography{thesis}
+%% \addcontentsline{toc}{chapter}{Bibliography}
 

### 2013-11-06 src2/thesis.tex
update

@@
 
-This thesis is about applying embedded language techniques to programming 
-multi- and many-core computers. Computers today are becoming more and more  
-parallel. With each new generation of 
-general purpose processor (CPU) or graphics processor (GPU), we get more 
-parallel resources. To fully make use of modern CPUs and GPUs, we need to
-exploit their parallelism.
-
-While it is sometimes possible to automatically parallelise programs originally 
-written for serial execution, it is becoming clear that explicitly parallel 
-programming models are needed for making use of all parallel resources~\cite{STRUCTURED}. 
-This also means that it is important to think in parallel from the beginning.
-
-This thesis focuses on data-parallelism, as opposed to control- or 
-task-parallelism. Data parallelism is applicable where the input data can 
-be divided into chunks and operated upon independently. Data-parallelism is scalable,
-meaning that the potential for parallelism increases with problem size.
-
-Modern CPUs are parallel in a number of different ways. They discover instruction
-level parallelism (ILP), they have multiple processor cores and vector units for 
-Single Instruction Multiple Data (SIMD)  parallelism. The exact configuration, number
-of cores and number of vector units, varies over available processor models. 
-To obtain maximum performance, with a specific target processor in mind, the programmer
-can use low-level intrinsics in order to get SIMD parallelism and a threading library
-for parallelism over the available cores. One problem with this is that the resulting program 
-would be specialised to a specific processor model. The Intel Array Building Blocks (ArBB) 
-system approaches this problem by providing a high level programming interface~\citet{ARBB2011}.
-This interface is based on a set of parallel operations on arrays, such as map, folds 
-and scans. ArBB was implemented as an embedded language in C++, but a low-level C interface to 
-its capabilities was also provided. ArBB programs are JIT-ed (Just-In-Time compiled) to 
-target the available CPU or accelerator (such as the Intel Larrabee~\citet{Larrabee}, 
-knights Ferry, Xeon Phi). In chapter~\ref{chap:ArBB}, there are papers describing work 
-towards embedding Intel ArBB in Haskell.
+This thesis applies embedded language methodology to data-parallel programming of 
+massively parallel hardware. Embedded languages are languages that are implemented 
+as a library for some existing host language. In this case, the functional programming 
+language Haskell is used as host language.
+
+Computers today are increasingly parallel. Each new generation of general purpose processors
+(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
+of how to program these computers most efficiently, both from a productivity and 
+a performance point of view, arise. Explicit parallel programming is hard and automatic 
+means of parallelisation, while giving a sequential programming model to the programmer, 
+often misses the target when it comes to performance. {\bf \emph{source??}}
+
+The embedded language approach to programming language implementation is beneficial 
+since we are not entirely sure of how to program current and future highly parallel 
+computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
+We can experiment with what abstractions to expose to the programmer, without much 
+effort. Other benefits of embedded languages often mentioned are: making use 
+of the host language's library functions, if they are general enough. The host language 
+can be used as a powerful macro language for the embedded language. An embedded 
+language also provide a more leveled learning curve to programmers already familiar 
+with the host language. For these programmers using an embedded language is more like  
+learning about a new library.
+
+This thesis focuses on data-parallelism, as opposed to task or control parallelism.
+Data-parallelism is applicable where the input data can be divided into chunks
+and operated upon independently. Data-parallelism is scalable; the potential 
+for parallelism increases with the problem size. Two different embedded languages 
+for data-parallel programming are presented in this thesis: Obsidian, for 
+general purpose computations on GPUs and EmbArBB for parallel programming across 
+platforms, CPU, GPU and other accelerators.
 
 Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
-For example, where CPUs dedicate large portions of chip area to caches, branch prediction and 
-ILP, GPUs put additional cores. This comes at a price of increased programmer effort.
-GPUs have programmer managed shared memories, conditionals are troublesome from a performance 
+For example, where CPUs dedicate large portions of chip area to caches, branch 
+prediction and Instruction level parallelism (ILP), GPUs put additional cores. 
+This comes at a price of increased programmer effort. GPUs have programmer managed 
+shared memories, conditionals are troublesome from a performance 
 perspective and memory access patterns can make or break the performance of an 
-application.
-
-Intel ArBB is part of a movement towards thinking of structure and patterns in 
-parallel programming. Other languages and libraries with similar aspirations 
-are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
-In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
-made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
-They emphasise the importance of collective array operations such as map, fold, scans, 
-permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
-Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
-the patterns as abstractions, but most are possible to express in each of the languages. 
-The patterns, while very powerful as abstractions, also provide tools for thinking about 
-parallel algorithms. The ArBB library supports flat regular data-parallelism, but also 
-irregular and to some degree nested data-parallelism.
-
-The patterns are powerful abstractions, but it is also important not to be locked into 
-a single implementation of each pattern. This could have negative performance impacts. 
-In ArBB, for example, the patterns used are specialised to the current hardware during JIT 
-compilation. Languages like OpenCL and CUDA do not provide abstractions of these patterns. 
-The programmer needs to implement them using lower level features, making CUDA and OpenCL 
-code less composable and harder to reuse. As an example, look at the slides from NVIDIA in 
-reference~\citet{reduction} where a reduction primitive is implemented in CUDA. First a 
-naive reduction pattern is implemented and then refined over many steps, taking many 
-characteristics of the GPU into mind. The end result is an implementation of the 
-reduction pattern that performs very well on the chosen GPU. But as GPUs evolve, the 
-design space may need to be explored again. 
-%What is 
-%is that says that we do not need to redo this exploration with each new GPU generation 
-%and discover a new optimal solution?
-
-Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
-compared to CUDA and provide a more composable interface. But we also want to keep 
-the possibility for the programmer to implement patterns in different ways. A balance 
-must be struck between providing enough low-level control to get at the performance 
-and still making building blocks implemented by the programmer reusable and composable. 
-We want to provide more compositional tools for the programmer to experiment with while 
-exploring the space of various ways to implement a given pattern or primitive. 
-In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian. 
-Obsidian is focused on flat and regular data-parallelism.
-
-% \section{The problem}
-
-% \emph{what is so interesting about Parallelism and GPUs}
+application. With Obsidian we try to find the right level of abstraction for 
+programming GPUs. We try to strike a balance between low-level control of details 
+that influence performance and high-level abstractions to increase productivity and 
+make programs more terse. The end goal with Obsidian is a language for writing 
+high performance GPU programs that compare well to code written by CUDA 
+(a C-dialect for GPU programming) experts, provided that both programmers have 
+similar GPU expertise. The Obsidian programmer would however do so in a more 
+composable manner. Obsidian programs are compiled into CUDA.
+
+Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
+good use of Obsidian requires that the programmer is willing to make an effort 
+of learning about GPUs and how to program them efficiently. However, another case 
+is a programmer who does not want to invest that effort but wants to write a high-level 
+parallel program and have it run reasonably well on different parallel architectures. 
+In this case a language with even higher level of abstraction is preferable. It then 
+becomes the job of a more advanced compiler to turn that into code for the targetted 
+platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
+operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
+has no power over the parallel strategy that should be used to implement these 
+operations but instead rely on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
+Haskell embedding of Intel ArBB that was released as an embedded language in C++.
+
+{\bf {\emph old text from here to end}}
 

### 2013-11-06 src2/thesis.tex
update

@@
 % ---------------------------------------------------------------------------
+%ditt ISBN är:  978-91-7385-939-4
+%ditt Ny serie nr:  3620
+%ISSN för publ.serien Doktorsavhandlingar vid Chalmers tekniska högskola : 0346-718X
 
@@
 \noindent Bo Joel Svensson \\
-\noindent ISBN \\
+\noindent ISBN 978-91-7385-939-4 \\
 
@@
 
+\noindent Doktorsavhandlingar vid Chalmers tekniska högskola \\
+\noindent Ny serie nr 3620 \\
+\noindent ISSN 0346-718X \\
+
+\vspace{1cm}
+
 \noindent Technical Report 101D\\
-\noindent ISSN                 \\
 \noindent \dept \\
@@
 make programs more terse. The end goal with Obsidian is a language for writing 
-high performance GPU programs that compare well to code written by CUDA 
+high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
 (a C-dialect for GPU programming) experts, provided that both programmers have 
@@
 Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
+EmbArBB relies entirely on the ArBB system for code generation and optimisation.
+
+\section {Obsidian: An embedded language for GPU kernel implementation}
+
+In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
+GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
+for general purpose programming. A dialect of C, called CUDA C, was released 
+specifically to simplify the implementation of general purpose computations on GPUs.
+
+Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
+non-graphics applications via the graphics API. The computation needed to be expressed 
+in graphics terms and the result computed through an act of rendering. Of course, since the 
+available abstractions targetted at the graphics domain, this added extra
+burden  on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
+there is a dialect of C, extended with GPU specific functionality.
+
+Having a language specifically made for general purpose programming on the GPU, such as 
+CUDA, is a big step up from earlier. But CUDA still offers a very low-level interface to 
+programming the GPU, which is preferred in a situation where the programmer needs to 
+think about hardware specific details to get the best performance. This CUDA provides, 
+but at the price that it is hard to build upon, abstract and reuse prior programming 
+effort. This is where we are trying to go one step further with Obsidian.
+
+\subsection{GPU hardware}
+
+Before going into programming of GPUs, it is important to have some background. The 
+GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
+of the details that make GPU programming hard is more apparent when looking at the 
+underlying hardware.
+
+A CUDA GPU is built around a single kind of processor (as opposed to the different 
+kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
+MultiProcessors) all contain a number of cores called SPs (Streaming Processors). 
+Each MP also contains local memory, called shared memory since it can be accessed 
+by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
+GPUs have as few as one MP, and as you go up in price the number of MPs increases.
+
+Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
+2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
+{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
+also divided into {\em Blocks}; the threads within a block can communicate using the 
+shared memory. The maximum block size is currently 1024 threads. Since 
+the number of threads in a block can be higher than the warp size, a barrier 
+synchronisation mechanism exists to ensure that threads within a block 
+can communicate safely via the shared memory.
+
+\subsection{Writing efficient GPU Kernels} 
+ \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}
 
@@
 \node[anchor=west] (acc) at (1,1.7) {CUB};
-\node[anchor=west] (repa) at (0,2.6) {DPH\\Nesl\\Repa\\Meta-repa};
-\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB\\ Feldspar};
+\node[anchor=west] (repa) at (0,2.5) {DPH\\Nesl\\Repa\\Meta-repa\\Feldspar};
+\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
 \node[anchor=west] (accelerator) at (2,1.7) {Accelerator};

### 2013-11-06 src2/thesis.tex
update

@@
 synchronisation mechanism exists to ensure that threads within a block 
-can communicate safely via the shared memory.
+can communicate safely via the shared memory. Blocks are also grouped into a grid that 
+is the collection of blocks executing the same program.
+
+\subsection{GPU programming}
+
+The CUDA programming model reflects the hierarchical architecture of the GPU. The 
+programmer writes what are called \emph{kernels}; in CUDA jargon a kernel refers to a sequential 
+C program that is parameterised on a thread's identity. The kernel program is executed 
+in parallel across all the threads of a block and many such blocks of threads can execute 
+in parallel on the available MPs. Kernels may use shared memory that is local to a block; 
+the programmer needs to decompose the algorithm and data in such a way that communication 
+will be local to the blocks.
+
+The program below is a very simple example of a CUDA kernel that computes elementwise sums.
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+__global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {
+
+  unsigned int gid = blockIdx.x * 
+                     blockDim.x + 
+                     threadIdx.x;
+
+  r[gid] = i1[gid] + i2[gid];
+
+}
+
+\end{Verbatim}
+\end{small}
+
+\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
+in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
+{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
+a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
+the blocks and of the grid is set by a controling program upon launch of the kernel. This 
+controling program is run by the CPU of the computer housing the GPU. the kernel is launched 
+using the following syntax:
+
+\begin{small} 
+\begin{Verbatim}[samepage=true]
+kernel<<<1,32,0>>>(d_i1,d_i2,d_r);
+\end{Verbatim}
+\end{small}
+
+\noindent The line of code above launches the kernel on one block of 32 threads. The zero 
+indicates that the kernel uses no shared memory. If the kernel did use shared memory, the 
+amount needed should have been specified there. Figure~\ref{fig:CUDACODE1} shows the complete 
+sequence of commands for a minimal example using the kernel specified above.
+
+In Obsidian the same kernel (the local computation) is implemented as follows:
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+
+\end{Verbatim]
+\end{small}
 
@@
 
+\begin{figure} 
 \begin{small}
@@
 \end{small}
+
+\caption{Host code for launching a simple CUDA kernel on the GPU.}
+\label{fig:CUDACODE1}
+
+\end{figure}
 

### 2013-11-07 src2/thesis.tex
update

@@
 This thesis applies embedded language methodology to data-parallel programming of 
-massively parallel hardware. Embedded languages are languages that are implemented 
+massively parallel hardware. An Embedded language is a language that is implemented 
 as a library for some existing host language. In this case, the functional programming 
-language Haskell is used as host language.
+language, Haskell is used as host language.
 
@@
 Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
-non-graphics applications via the graphics API. The computation needed to be expressed 
+non-graphics applications via the graphics API. The desired computation was expressed 
 in graphics terms and the result computed through an act of rendering. Of course, since the 
-available abstractions targetted at the graphics domain, this added extra
+available abstractions targeted the graphics domain, this added extra
 burden on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
-there is a dialect of C, extended with GPU specific functionality.
+there is a dialect of C, extended with a parallel programming model suitable for GPUs.
 
@@
 programming the GPU, which is preferred in a situation where the programmer needs to 
-think about hardware specific details to get the best performance. This CUDA provides, 
+think about hardware specific details to get the best performance. CUDA provides this, 
 but at the price that it is hard to build upon, abstract and reuse prior programming 
@@
 \begin{Verbatim}[samepage=true]
-__global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {
+__global__ void vecAdd(int32_t *i1, int32_t *i2, int32_t *r) {
 
@@
 \begin{Verbatim}[samepage=true]
-kernel<<<1,32,0>>>(d_i1,d_i2,d_r);
+vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
 \end{Verbatim}
@@
 \begin{Verbatim}[samepage=true]
-
-\end{Verbatim]
+vecAdd :: Num a => SPull a -> SPull a -> SPull a
+vecAdd = zipWith (+)
+\end{Verbatim}
+\end{small}
+
+\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
+as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
+we can generate CUDA kernels for any type that supports {\tt (+)}, is in the {\tt Num} class.
+
+There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
+we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
+which would mean conditionals in the code and conditionals are problematic from a performance 
+point of view. Another difference is that in Obsidian the programmer specifies the entire 
+computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
+in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
+threads provided at kernel launch. This means that the Obsidian programmer needs to 
+write a program that performs this mapping:
+
+\begin{small} 
+\begin{Verbatim}[samepage=true]
+vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
+vecAddG i1 i2  =
+  pConcat $
+  fmap push $
+  zipWith vecAdd
+          (splitUp 32 i1)
+          (splitUp 32 i2)  
+\end{Verbatim}
 \end{small}
@@
 % ---------------------------------------------------------------------------
-\section{GPUs and CUDA}
-
-In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
-GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs 
-for general purpose programming. A dialect of C, called CUDA C, was released specifically 
-to simplify implementation of general purpose algorithms on GPUs.
-
-A CUDA GPU is built around a single kind of processor (as opposed to the different 
-kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
-MultiProcessors) all contain a number of function units called SPs (Streaming Processors). 
-Each MP also contains local memory, called shared memory since it can be accessed 
-by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
-GPUs have as few as one MP, and as you go up in price the number of MPs increases.
-
-Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
-2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
-{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
-also divided into {\em Blocks}; the threads within a block can communicate using the 
-shared memory. The maximum block size is currently 1024 threads. Since 
-the number of threads in a block can be higher than the warp size, a barrier 
-synchronisation mechanism exists to ensure that threads within a block 
-can communicate safely via the shared memory.
-
-Papers B, C, E and F (sections~\ref{sec:paperB},~\ref{sec:paperC},~\ref{sec:paperE},
-~\ref{sec:paperF}) contain more information about GPUs, and the assumptions we make 
-about them in relation to Obsidian.
 
@@
   // Launch the kernel
-  kernel<<<1,32,0>>>(d_i1,d_i2,d_r);
+  vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
 

### 2013-11-07 src2/thesis.tex
update

@@
 
-\textbf{Keywords:} Data-parallelism, Functional Programming, Embedded languages, GPU
+\textbf{Keywords:} Data-parallelism, Embedded languages, Functional Programming, Graphics Processing Units
 
@@
 
+% -------------------------------------------------------------------------
+%  INTRO SECTION OF INTRO CHAPTER
+% -------------------------------------------------------------------------
+
 \section{Introduction}
@@
 This thesis applies embedded language methodology to data-parallel programming of 
-massively parallel hardware. An Embedded language is a language that is implemented 
+massively parallel hardware. An embedded language is a language that is implemented 
 as a library for some existing host language. In this case, the functional programming 
-language, Haskell is used as host language.
+language Haskell is used as host language.
 
@@
 of how to program these computers most efficiently, both from a productivity and 
-a performance point of view, arise. Explicit parallel programming is hard and automatic 
+a performance point of view, arises. Explicit parallel programming is hard and automatic 
 means of parallelisation, while giving a sequential programming model to the programmer, 
@@
 
-The embedded language approach to programming language implementation is beneficial 
+The embedded language approach to programming language implementation is beneficial, 
 since we are not entirely sure of how to program current and future highly parallel 
@@
 can be used as a powerful macro language for the embedded language. An embedded 
-language also provide a more leveled learning curve to programmers already familiar 
-with the host language. For these programmers using an embedded language is more like  
+language also provides a more leveled learning curve to programmers already familiar 
+with the host language. For these programmers, using an embedded language is more like  
 learning about a new library.
@@
 For example, where CPUs dedicate large portions of chip area to caches, branch 
-prediction and Instruction level parallelism (ILP), GPUs put additional cores. 
+prediction and instruction level parallelism (ILP), GPUs put additional cores. 
 This comes at a price of increased programmer effort. GPUs have programmer managed 
@@
 perspective and memory access patterns can make or break the performance of an 
-application. With Obsidian we try to find the right level of abstraction for 
+application. With Obsidian, we try to find the right level of abstraction for 
 programming GPUs. We try to strike a balance between low-level control of details 
@@
 good use of Obsidian requires that the programmer is willing to make an effort 
-of learning about GPUs and how to program them efficiently. However, another case 
+to learn about GPUs and how to program them efficiently. However, another case 
 is a programmer who does not want to invest that effort but wants to write a high-level 
 parallel program and have it run reasonably well on different parallel architectures. 
-In this case a language with even higher level of abstraction is preferable. It then 
+In this case, a language with even higher level of abstraction is preferable. It then 
 becomes the job of a more advanced compiler to turn that into code for the targetted 
@@
 has no power over the parallel strategy that should be used to implement these 
-operations but instead rely on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
+operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
 Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
 EmbArBB relies entirely on the ArBB system for code generation and optimisation.
+
+% -------------------------------------------------------------------------
+% Obsidian !
+% -------------------------------------------------------------------------
 
@@
 available abstractions targeted the graphics domain, this added extra
-burden on the programmer. CUDA was NVIDIA's first response to this problem. With CUDA 
-there is a dialect of C, extended with a parallel programming model suitable for GPUs.
+burden on the programmer. CUDA was NVIDIA's first response to this problem. CUDA 
+is a dialect of C, extended with a parallel programming model suitable for GPUs.
 
@@
 effort. This is where we are trying to go one step further with Obsidian.
+
+% -------------------------------------------------------------------------
+% GPU Hardware 
+% -------------------------------------------------------------------------
 
@@
 
+% -------------------------------------------------------------------------
+%  GPU Programming
+% -------------------------------------------------------------------------
+
 \subsection{GPU programming}
@@
 a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
-the blocks and of the grid is set by a controling program upon launch of the kernel. This 
-controling program is run by the CPU of the computer housing the GPU. the kernel is launched 
+the blocks and of the grid are set by a controling program upon launch of the kernel. This 
+controling program is run by the CPU of the computer housing the GPU. The kernel is launched 
 using the following syntax:
@@
 
-\noindent The line of code above launches the kernel on one block of 32 threads. The zero 
+\noindent This line of code launches the kernel on one block of 32 threads. The zero 
 indicates that the kernel uses no shared memory. If the kernel did use shared memory, the 
@@
 
-In Obsidian the same kernel (the local computation) is implemented as follows:
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-vecAdd :: Num a => SPull a -> SPull a -> SPull a
-vecAdd = zipWith (+)
-\end{Verbatim}
-\end{small}
-
-\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
-as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
-we can generate CUDA kernels for any type that supports {\tt (+)}, is in the {\tt Num} class.
-
-There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
-we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
-which would mean conditionals in the code and conditionals are problematic from a performance 
-point of view. Another difference is that in Obsidian the programmer specifies the entire 
-computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
-in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
-threads provided at kernel launch. This means that the Obsidian programmer needs to 
-write a program that performs this mapping:
-
-\begin{small} 
-\begin{Verbatim}[samepage=true]
-vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
-vecAddG i1 i2  =
-  pConcat $
-  fmap push $
-  zipWith vecAdd
-          (splitUp 32 i1)
-          (splitUp 32 i2)  
-\end{Verbatim}
-\end{small}
-
-\subsection{Writing efficient GPU Kernels} 
- \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}
-
-{\bf {\emph old text from here to end}}
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% GPU Programming 
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-
-\subsection{CUDA programming}
-
-The CUDA programming model reflects the hierarchical architecture of GPUs. The programmer 
-writes what are called {\em kernels}; in CUDA jargon a kernel refers to a sequential C program 
-that is parameterised on a thread's identity. The kernel program is executed in parallel across 
-all the threads of a block and many such blocks of threads can execute in parallel on the 
-available MPs. The programmer needs to decompose the algorithm and data in such a way that 
-communication will be local to the blocks.
-
-Below is a very simple CUDA example to give some indication of what a kernel and 
-the code to launch that kernel looks like.
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-__global__ void kernel(int32_t *i1, int32_t *i2, int32_t *r) {
-
-  unsigned int gid = blockIdx.x * 
-                     blockDim.x + 
-                     threadIdx.x;
-
-  r[gid] = i1[gid] + i2[gid];
-
-}
-
-\end{Verbatim}
-\end{small}
-
-This kernel takes two input arrays, {\tt i1} and {\tt i2}, and computes 
-element wise sums into the result array, {\tt r}. The {\tt threadIdx} variable identifies 
-a thread within a block. There is also a {\tt blockIdx} variable that identifies 
-the block a thread belongs to and {\tt blockDim.x} specifies the number of threads per block.
-
-In CUDA there is a notion of {\em host} and {\em device}, the device is the GPU and 
-its memory (Global memory, usually a few Gigabytes). The host refers to the system housing 
-the GPU. The host runs a control program that launches kernels on the GPU. First memory 
-needs to be allocated in the device to hold the inputs and outputs. Then data is copied from 
-the host's system memory into device memory.
+% -------------------------------------------------------------------------
+% CUDACODE1 FIGURE
+% -------------------------------------------------------------------------
 
@@
 
-\pagebreak
-
-The syntax for launching a kernel is:
+% -------------------------------------------------------------------------
+% 
+% -------------------------------------------------------------------------
+
+In Obsidian, the same kernel (the local computation) is implemented as follows:
 
@@
 \begin{Verbatim}[samepage=true]
-name<<<Blocks, Threads/Block, BytesShared>>>(arg1,..,argN);
+vecAdd :: Num a => SPull a -> SPull a -> SPull a
+vecAdd = zipWith (+)
 \end{Verbatim}
@@
 
-\noindent In the example above one block of 32 threads, using no shared memory, is launched.
-
-A kernel that uses shared memory takes the form:
+\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
+as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
+we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
+In Haskell we know this is true for types that are member of the {\tt Num} typeclass.
+
+There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
+we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
+which would mean conditionals in the code. Conditionals are problematic from a performance 
+point of view. Another difference is that in Obsidian the programmer specifies the entire 
+computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
+in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
+threads provided at kernel launch. This means that the Obsidian programmer needs to 
+write a program that performs this mapping:
 
@@
 \begin{Verbatim}[samepage=true]
-__global__ void kernel2(int32_t *i, int32_t *r) {
+vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
+vecAddG i1 i2  =
+  pConcat $
+  zipWith vecAdd
+          (splitUp 32 i1)
+          (splitUp 32 i2)  
+\end{Verbatim}
+\end{small} % $
+
+The {\tt vecAddG} program specifies a computation on large arrays; the large input arrays 
+are split up into pieces that are operated upon by the local computation, {\tt vecAdd}, 
+specified above. Each input array passed to this function must have length a multiple of 32.
+Since the {\tt splitUp} operation splits the arrays into chunks of 32 elements, the body 
+of the generated kernel will be constructed specifically for this size. {\tt zipWith} is 
+used to map the local computation onto each chunk of the input arrays. The result is an 
+array of arrays that is flattened with the {\tt pConcat} function. The CUDA code below 
+is generated from the {\tt vecAddG} program:
+
+\begin{small}
+\begin{Verbatim}[samepage=true] 
+extern "C" __global__ void vecAdd(int32_t* input0, uint32_t n0,
+                                  int32_t* input1, uint32_t n1,
+                                  int32_t* output2)
+{
+    uint32_t tid = threadIdx.x;
+
+    if (blockIdx.x < min(n0 / 32U, n1 / 32U)) {
+        output2[blockIdx.x * 32U + tid] = 
+            input0[blockIdx.x * 32U + tid] +
+            input1[blockIdx.x * 32U + tid];
+    }
+}
+\end{Verbatim}
+\end{small}
+
+There are some superficial differences between the code generated using Obsidian and the 
+handwritten example above. One such difference is that the index computation is not shared. 
+Obsidian relies on the CUDA compiler to discover such sharing via a common subexpression 
+elimination optimisation pass. Another difference is that Obsidian has inserted a conditional 
+that only executes the body for certain blocks. Discussion of the causes for that conditional 
+is postponed until a later section {\bf \emph{POINT TO SECTION}}. A real difference is that 
+the kernel is generated for a fixed size, here there is a constant (32) where in the handwritten
+code we used {\tt blockDim.x}. If those differences are set aside, the two kernels are identical. 
+Figure~\ref{fig:OBSIDIANHOST1} shows how the {\tt vecAddG} kernel can be launched from 
+within Haskell, using Obsidian's rudimentary support for launching GPU computations.
+
+An alternative way to implement {\tt vecAddG} is to inline the local computation 
+directly. This makes the entire program look like this.
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
+vecAddG i1 i2  =
+  pConcat $
+  zipWith (zipWith (+)) 
+          (splitUp 32 i1)
+          (splitUp 32 i2)
+\end{Verbatim} 
+\end{small} % $
+
+\noindent The generated code looks exactly as before.
+
+% -------------------------------------------------------------------------
+% OBSIDIANHOST1 FIGURE
+% -------------------------------------------------------------------------
+
+\begin{figure} 
+\begin{small}
+\begin{Verbatim}[samepage=true]
+performVecAdd =
+  withCUDA $
+  do
+    kern <- capture 32 (vecAddG :: DPull EInt32
+                                 -> DPull EInt32
+                                 -> DPush Grid EInt32)
+
+    useVector (V.fromList [0..32::Int32]) $ \i1 ->
+      useVector (V.fromList [0..32::Int32]) $ \i2 -> 
+        allocaVector 32 $ \(o :: CUDAVector Int32) ->
+        do
+          fill o 0
+          o <== (1,kern) <> i1 <> i2
+
+          r <- peekCUDAVector o
+          lift $ putStrLn $ show r
+\end{Verbatim} 
+\end{small}
+
+\caption{Host code for launching a simple Obsidian kernel on the GPU. \\ 
+ {\tt capture} compiles an Obsidian program into CUDA. At this point the 
+  types of the Obsidian program need to be made concrete. The number, 32, used as 
+ an argument to {\tt capture} specifies that generated code should be specialised 
+ for 32 threads per block. }
+\label{fig:OBSIDIANHOST1}
+\end{figure}
+
+A CUDA kernel that uses shared memory takes the form:
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+__global__ void kernel(int32_t *i, int32_t *r) {
 
@@
 \noindent A portion of the input array is stored into shared memory. A 
-barrier synchronisation is used to ensure that each threads within the block 
+barrier synchronisation is used to ensure that each thread within the block 
 has stored its value into shared memory (only necessary if threads will 
 communicate across warp boundaries). After computing on the data locally, 
-it is written back into global memory.
+it is written back into global memory. This kernel sets up two aliases for 
+thread identities. A thread's global identity is called {\tt gid} and its local 
+identity is called {\tt tid}. The global ID is used to index into the large global 
+array; indexing into local shared memory is done using the local ID.
+
+In Obsidian, local computations can also use shared memory. The Obsidian arrays 
+we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
+correspond to actual data in memory. Pull and push arrays, rather represent ways 
+to compute arrays. More information about these array representations is available 
+in section {\bf \emph{REF SECTION}}. The Obsidian programmer can store intermediate 
+arrays in shared memory by using a {\tt force} operation. The 
+{\tt force} operation computes an array and stores it into shared memory.
+
+The following kernel multiplies each element in an array by two and then adds one:
+
+\begin{small}
+\begin{Verbatim}[samepage=true] 
+kernel :: Num a => SPull a -> SPull a
+kernel = fmap (+1) . fmap (*2)
+\end{Verbatim}
+\end{small}
+
+\noindent This kernel computes its result without any storage of intermediate results. 
+In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
+are applied without any intermediate storage.
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
+                                  int32_t* output1)
+{
+    uint32_t tid = threadIdx.x;
+
+    if (blockIdx.x < n0 / 32U) {
+        output1[blockIdx.x * 32U + tid] = 
+          input0[blockIdx.x * 32U + tid] * 2 + 1;
+    }
+}
+\end{Verbatim}
+\end{small}
+
+Storing an intermediate result between the two operations can be accomplished by 
+changing the kernel as follows.
+
+\begin{small}
+\begin{Verbatim}[samepage=true]
+kernelF :: (MemoryOps a, Num a)
+         => SPull a -> BProgram (SPull a)
+kernelF = liftM (fmap (+1)) . force . fmap (*2)
+\end{Verbatim} 
+\end{small}
+
+\FloatBarrier 
+\subsection{Writing efficient GPU Kernels} 
+ \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}
+
+{\bf {\emph old text from here to end}}
+
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% GPU Programming 
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+
+\subsection{CUDA programming}
+
+A kernel that uses shared memory takes the form:
 

### 2013-11-08 src2/thesis.tex
update

@@
 
-A CUDA kernel that uses shared memory takes the form:
+\pagebreak
+
+\noindent A CUDA kernel that uses shared memory takes the form:
 
@@
 
+This change introduced some noise. The reason for this is that Obsidian programs that 
+use memory are monadic. The {\tt BProgram} in the result type is a monad. Another difference 
+is that we need to be able to write the array elements to memory; in Obsidian this means 
+that the element type must be in the {\tt MemoryOps} class.
+
 \FloatBarrier 
 \subsection{Writing efficient GPU Kernels} 
- \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}
-
-{\bf {\emph old text from here to end}}
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% GPU Programming 
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-
-\subsection{CUDA programming}
-
-A kernel that uses shared memory takes the form:
-
-\subsection{Writing efficient GPU kernels} 
 \label{sec:efficient}
@@
 to learn more about these details. Here is a list, with explanations of some 
-of the practices that the guide deems most important.
+of the practices that the guide deems most important. The list also contains 
+information about how these best practices influence our work on Obsidian.
 
@@
 
+Data transfers between the host memory and device (GPU) memory is out of Obsidian's scope. 
+If Obsidian is used to generate kernels for a computation, the programmer has already decided that 
+a GPU should be used. However, even after making the decision to use the GPU the programmer 
+may need to consider how to transfer the data. In some cases it is possible to overlap 
+computation on the GPU and data transfer. These kinds of optimisations are considered 
+when writing the C program that launches the CUDA kernels. Obsidian does have rudimentary 
+support for launching kernels from within Haskell but currently we consider this mostly a
+tool for quickly being able to run a kernel and see what it does, not as a means to implement 
+applications. This part of Obsidian is future work.
+
 \item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
@@
 
+Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
+the permutation used will decide the access pattern. If this method can be applied or 
+not partly also depends on the algorithm being implemented.
+
 \item {\bf Minimise use of global memory:} Accessing global device memory can take as 
@@
 
+Intermediate arrays are stored into shared memory when the Obsidian programmer uses 
+the {\tt force} function.
+
 \item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
@@
 
+Obsidian requires regularity even more than CUDA does. Since we focus on generating high 
+performance kernels, we rule out generation of kernels that work on different sizes (per block). 
+The choice of how many threads to use per block is in the hands of the Obsidian programmer.
+The method for making that choice has changed over time. Earlier, the number of threads 
+needed by a kernel was decided by the size of the arrays it computed. This has recently changed
+and is now set by a parameter passed to the code generation process.
+
 \item {\bf Avoid different execution paths within a block:} Any branching instruction that 
-causes the execution paths to diverge within a warp will affect performance. When the execution 
-paths within a warp diverge, the computation is serialised.
+causes the execution to branch into different code (in CUDA jargon called a diverging branch) 
+within a warp will affect performance. When the execution paths within a warp diverge, the 
+computation is serialised.
+
+Sometimes divergent code cannot be avoided. However, it is important not to have threads 
+diverge unnecessarily. When using pull arrays alone this was often a problem; with the 
+addition of push arrays the situation improved. Using push arrays some divergent computations 
+can be explicitly split up into phases, each phase with all threads following the 
+same execution path. This means that fewer threads are required to do the same work.
 
@@
 this is a major limitation to the composability of CUDA functions. Wheather or not it is safe to 
-call some function in a certain place in your code can be decided only by inspecting that function's 
-implementation. 
+call some function in a certain place in your code can be decided only by inspecting that 
+function's implementation.
+
+In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
+conditional code. Conditionals can be introduced either by an {\tt ifThenElse} function 
+at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
+monad. The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
+thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs ({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a synchronisation can 
+only appear in an Obsidian block program. More information about programs and the GPU hierarchy 
+is available in section~\ref{sec:OBSIMP}.
+
 %f it is safe or 
@@
 
-%\citet{FERMIOPT}
-%\citet{BestPrac} 
-%\citet{merrill}
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+%
+% Obsidian Implementation 
+%
+% ---------------------------------------------------------------------------
+% ---------------------------------------------------------------------------
+\subsection{Obsidian Implementation}
+\label{sec:OBSIMP}
+
+{\bf {\emph old text from here to end}}
 

### 2013-11-09 src2/thesis.tex
update

@@
 
+% ---------------------------------------------------------------------------
+% ABSTRACT !!  
+% ---------------------------------------------------------------------------
 \section*{Abstract}
-Abstract abstract
+
+Computers today are becoming more and more parallel. General purpose processors (CPUs) 
+have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
+data-parallelism. Graphics processors (GPUs) brings massive parallelism at the cost of 
+being harder to program than CPUs. This thesis applies embedded language methodology 
+to data-parallel programming. Two embedded languages are presented, Obsidian for general 
+purpose GPU programming and EmbArBB for data-parallel programming across platforms.
+
+Each generation of CPU and GPU bringing more parallel resources. The question of how to write 
+programs for these processors efficiently arise. We are after efficiency both in programmer 
+productivity and in application performance. Using embedded languages allows us to experiment
+with what abstractions to present to the programmer at relatively little effort.
+
+Obsidian is an embedded language for general purpose programming of GPUs. We try to strike 
+a balance between high level, productivity increasing abstractions and low-level 
+control needed for performance. Obsidian programs are compiled into CUDA C code.
+
+In cases where the programmer is not interested in low-level details 
+or wants the program to run over a range of hardware a higher level language can be used.
+EmbArBB is a Haskell embedding or the Intel ArBB system. EmbArBB relies on the ArBB system 
+to generate code (via a Just-In-Time compiler) to a range of hardware.
+
+EmbArBB embeds a preexisting library for data-parallelism into Haskell and we obtain 
+very good performance at little implementation effort. This performance comes from the 
+expertise and effort put into the ArBB system and that we get for free. Embedding ArBB is 
+a way to provide these benefits to the Haskell programmer and a way to increase usefulness
+of an existing system by opening it up to a wider audience. Obsidian is very different, it 
+is not based on a set of high-level parallel primitives. The Obsidian programmer 
+can implement these primitives in different ways, select the best one and then  abstract 
+upon that. We have obtained very good performance in case studies involving reductions.
 
@@
 \section*{Acknowledgments}
+
+I would like to thank my supervisor Mary Sheeran. She is a very suportive, motivating and 
+understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
+and tips he has provided over he years. I cannot imagine that this would have worked with 
+any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
+supervisors. Working with Josef has been a very fun and rewarding experience.
+
+I thank Manuel Chakravarty for letting me visit his group at the University of New South Wales 
+and Trevor McDonell and Sean Lee for taking care of me during my visit. I look back 
+very fondly to my stay in Australia.
+
+For being a great supervisor during my three month internship at Intel, I thank Ryan Newton. 
+These where three very intense months away from family but at the same time a very rewarding 
+experience.
+
+I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions.
+
+Thanks to my thesis opponent Stephen Edwards for valuable comments on a draft of this thesis.
+
+Finally I want to thank my family and friends. Thanks to my sisters and parents for their support.
+I thank my friends for being understanding during times when I have been very self-absorbed. 
+I am very grateful to my wife Andrea for supporting and believing in me and our daughter Noemi 
+for bringing much joy to my breaks from work while writing this thesis; soon she is old enough 
+for real LEGO.
 
@@
 means of parallelisation, while giving a sequential programming model to the programmer, 
-often misses the target when it comes to performance. {\bf \emph{source??}}
+often misses the target when it comes to performance. The book ``Structured Parallel 
+Programming'' by Michael McCool et al argues for the need of explicit parallel programming 
+models~\citet{STRUCTURED}.
 
@@
 \subsection{GPU programming} 
+\label{GPUPROGRAMMING}
 
@@
 handwritten example above. One such difference is that the index computation is not shared. 
-Obsidian relies on the CUDA compiler to discover such sharing via a common subexpression 
-elimination optimisation pass. Another difference is that Obsidian has inserted a conditional 
-that only executes the body for certain blocks. Discussion of the causes for that conditional 
-is postponed until a later section {\bf \emph{POINT TO SECTION}}. A real difference is that 
-the kernel is generated for a fixed size, here there is a constant (32) where in the handwritten
-code we used {\tt blockDim.x}. If those differences are set aside, the two kernels are identical. 
-Figure~\ref{fig:OBSIDIANHOST1} shows how the {\tt vecAddG} kernel can be launched from 
-within Haskell, using Obsidian's rudimentary support for launching GPU computations.
+Obsidian relies on the CUDA compiler to discover such sharing via a Common Subexpression 
+Elimination (CSE) optimisation pass. Another difference is that Obsidian has inserted a 
+conditional that only executes the body for certain blocks. Discussion of the causes for 
+that conditional is postponed until a later section~\ref{sec:OBSARRAYS}. A real 
+difference is that the kernel is generated for a fixed size, here there is a constant (32) 
+where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
+aside, the two kernels are identical. Figure~\ref{fig:OBSIDIANHOST1} shows how the 
+{\tt vecAddG} kernel can be launched from within Haskell, using Obsidian's rudimentary 
+support for launching GPU computations.
 
@@
 to compute arrays. More information about these array representations is available 
-in section {\bf \emph{REF SECTION}}. The Obsidian programmer can store intermediate 
-arrays in shared memory by using a {\tt force} operation. The 
-{\tt force} operation computes an array and stores it into shared memory.
+in section~\ref{sec:OBSARRAYS}. The Obsidian programmer can store intermediate 
+arrays in shared memory by using a {\tt force} operation. 
+%The 
+%{\tt force} operation computes an array and stores it into shared memory.
 
@@
 
-\noindent This kernel computes its result without any storage of intermediate results. 
+\noindent This kernel computes its result without any storage of intermediate results;
+the code is equivalent to \verb!fmap ((+1).(*2))!. 
 In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
@@
 that the element type must be in the {\tt MemoryOps} class.
+
+The code below is generated using the {\tt kernelF} program:
+
+\begin{small} 
+\begin{Verbatim}[samepage=true]
+extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
+                                  int32_t* output1)
+{
+    extern __shared__ uint8_t sbase[];
+    uint32_t tid = threadIdx.x;
+
+    if (blockIdx.x < n0 / 32U) {
+        ((int32_t*) sbase)[tid] = input0[blockIdx.x * 32U + tid] * 2;
+        __syncthreads();
+        output1[blockIdx.x * 32U + tid] = ((int32_t*) sbase)[tid] + 1;
+    }
+}
+\end{Verbatim}
+\end{small}
+
+\noindent This code uses a shared memory array called {\tt sbase}. There 
+is also a call to {\tt \_\_syncthreads} to ensure that all the writes have 
+completed before proceeding. In this case the synchronisation is not really
+necessary since all threads belong to the same warp. There is a {\tt unsafeforce} 
+operation that naively removes synchronisations if the array being forced is short 
+enough. More advanced analysis could also be performed to remove synchronisations 
+whenever communication is entirely warp local. We do not apply that optimisation, 
+but work on warp level computations is in progress and would put this ability 
+in the hands of the programmer.
 
@@
 In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
-conditional code. Conditionals can be introduced either by an {\tt ifThenElse} function 
+conditional code. Conditionals can be introduced either by an \\{\tt ifThenElse} function 
 at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
-monad. The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
-thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs ({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a synchronisation can 
-only appear in an Obsidian block program. More information about programs and the GPU hierarchy 
-is available in section~\ref{sec:OBSIMP}.
+monad; we have seen an example of the program monad in section~\ref{sec:GPUPROGRAMMING}. 
+The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
+thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs 
+({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a 
+synchronisation can only appear in an Obsidian block program. More information about programs 
+and the GPU hierarchy is available in section~\ref{sec:OBSIMP}.
 
@@
 
-{\bf {\emph old text from here to end}}
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Embedded languages
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\section{Embedded languages}
-
-This thesis applies embedded language methodology to data-parallel programming. An 
-embedded language is implemented as a library in a host language,  Haskell in this case. 
-The embedded language approach is beneficial since we are not entirely sure how 
-to program current and future highly parallel computers. Implementing embedded languages allows 
-us more rapid prototyping of ideas. Other benefits of embedded languages often listed 
-include making use of the host language's type system to guarantee that correctly typed target code is generated. 
-Already existing libraries in the host language (if they are general enough) 
-can be used in the embedded language. The host language can also be used as a powerful macro 
-language for your embedded language.
-
-A very good paper, and highly relevant to our approach, is ``Compiling Embedded Languages''
-by Conal Elliott et al~\citet{COMPILEEDSL}. This paper describes how to embed a compiler 
-backend and generate efficient code in some target language. One key aspect of this 
-approach is that the embedded language is a library of functions that create and 
-compose Abstract Syntax Trees (ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as 
-well as EmbArBB (chapter \ref{chap:ArBB}) are implemented in a similar way. An 
-embedding that builds ASTs is called a deep embedding. Shallow embeddings, on the 
-other hand, do not build ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' 
-Josef Svenningsson and Emil Axelsson combine the two methods, shallow and deep, 
-in order to simplify the AST data type (fewer constructors) and make more extensible 
-embedded languages \citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
-and deep embeddings, while EmbArBB is a more traditional deep embedding.
+Obsidian is implemented as an embedded language using Haskell as host. Obsidian 
+is a compiled embedded language; a very good paper that is highly relevant to our 
+approach, is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
+This paper describes how to embed a compiler backend and generate efficient code 
+in some target language. One key aspect of this approach is that the embedded 
+language is a library of functions that create and compose Abstract Syntax Trees 
+(ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as well as EmbArBB 
+(chapter \ref{chap:ArBB}) are implemented in a similar way. An embedding that builds 
+ASTs is called a deep embedding. Shallow embeddings, on the other hand, do not build 
+ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' Josef Svenningsson and 
+Emil Axelsson combine the two methods, shallow and deep, in order to simplify the 
+AST data type (fewer constructors) and make more extensible embedded languages 
+\citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
+and deep embeddings.
 
@@
 For example, Lava and Wired~\citet{LAVA,Wired} raises the level of abstraction in the 
-fields of hardware design and verification and Feldspar does the same in the area of  digital 
-signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
+fields of hardware design and verification and Feldspar does the same in the area of  
+digital signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
 embedded approaches using various host languages. Two languages for GPU programming 
@@
 
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Obsidian
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\section{Obsidian: An embedded language for GPU kernel implementation}
-
-\vspace{5mm} 
-\emph{This section will be revised. I want to add more code examples.} 
-\vspace{5mm}
-
-Obsidian aspires to raise the level of abstraction of GPU kernel implementation, 
-while maintaining enough control over details that influence performance on a GPU. 
-We try to reach these goals by offering abstractions to the programmer that are 
-compositional and parallel. We also ensure that these can easily be compiled 
-to NVIDIA CUDA code. We have two different array representations, each with 
-its strengths and weaknesses. We also feel that it is important that the programmer 
-be able to use local shared memory in computations and have the ability to influence 
-memory access patterns. What these array representations and abstractions are 
-will be explained in the following sections, with references to the papers 
-that introduce them or explain them the best.
-
-\subsection{Deep embedding: The {\tt Program} data type}
+\subsubsection{Deep embedding: The {\tt Program} data type}
 
@@
 
-\subsection{Scalars}
+\subsubsection{Scalars}
 
@@
 
-\subsection{Arrays}
+\subsubsection{Arrays} 
+\label{sec:OBSARRAYS}
 
@@
 
-\subsubsection{Pull arrays}
+\paragraph{Pull arrays:}
 
@@
 parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
-or dynamic (length represented by an expression, an unknown length). When creating block and 
-warp level computations, we want to know the exact lengths in order to avoid generating code 
+or dynamic (length represented by an expression, an unknown length). When creating block 
+level computations, we want to know the exact lengths in order to avoid generating code 
 riddled with conditionals concerning array lengths. It is possible to write kernels that 
@@
 conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
-when performance is crucial. 
+when performance is crucial. Dynamic lengths are used for grid level computations, this allows 
+a local computation of a fixed size to be run in parallel over arrays a multiple of the size that 
+the local computation handles. In generated code this is currently visible as a conditional 
+on {\tt blockIdx.x}. However, this is still considered work in progress; the conditional is 
+only really necessary for Obsidian kernels that have more than one output array and the outputs
+are of different length. 
 %Conditionals are problematic on GPUs and 
@@
 
-\subsubsection{Push arrays}
+\paragraph{Push arrays:}
 
@@
 
-\subsection{Compilation to CUDA}
+\subsubsection{Compilation to CUDA}
 
@@
 the Obsidian programmer.
-
-\vspace{5mm}
-
-\emph{This list below will be revised. More details or code will be added.}
-
-\vspace{5mm}
-
-\begin{itemize}
-
-\item {\bf Minimise data transfers between host and device:} Data transfers
-between host and device are out of Obsidian's scope. If Obsidian is used to create 
-a kernel for a computation, the programmer has already decided that a GPU should be used.
-
-\item {\bf Ensure that global memory accesses are coalesced:} Inputs to an Obsidian 
-program are pull arrays. Pull arrays can be arbitrarily permuted; the permutation 
-used will decide the access pattern. If this method can be applied or not partly also 
-depends on the algorithm being implemented.
-
-\item {\bf Minimise use of global memory:} Intermediate results can be computed into 
-shared memory. Arrays are stored into shared memory when the programmer uses a 
-{\tt force} primitive.
-
-\item {\bf Use a multiple of 32 threads per block:} Obsidian requires regularity even more 
-than CUDA does. Since we focus on generating high performance kernels, we rule out generation 
-of kernels that work on different input sizes. The choise of how many threads to use per block 
-is in the hands of the Obsidian program. The method for making that choise has changed over time. 
-Earlier, the number of threads needed by kernel was decided by the size of the arrays it 
-computed. This has recently changed is now set by parameters to the code generation process.
-
-\item {\bf Avoid different execution paths within a block:} Sometimes divergent code cannot 
-be avoided. However, It is important not to have threads diverge unnecessarily. When using pull 
-arrays alone this was often a problem; with the addition of push arrays the situation improved. 
-Using push arrays a divergent computation can be explicitly split up into phases, each with all 
-threads following the same execution path. This means that fewer threads are required to do the 
-same work.
-
-\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} In Obsidian it is impossible 
-for a barrier synchronisation to appear within conditional code. Conditionals can be introduced 
-either by a conditional expression or a {\tt Cond} operation in the program datatype. The Cond 
-operation chooses between two thread programs and the synchronisation operation is at the block 
-level in the hierarchy. So a synchronisation cannot occur in conditional code. 
-%f it is safe or 
-%not to call some function at a certain place in your code, can only be decided by inspecting 
-%that functions implementation. 
-\end{itemize}
-
-%% In the book ``Structured Parallel Programming''~\citet{STRUCTURED}, there is a list showing language 
-%% support for various parallel patterns. In figure~\ref{fig:caps} this list augmented with a column for Obsidian. 
-%% This figure is added to indicate relative expressive power of the different languages. The most 
-%% interesting comparison is between Obsidian and OpenCL that both have a programming model geared towards
-%% GPUs; even though OpenCL can compile for both CPU and GPU execution. The paragraphs below 
-%% take a closer look at well Obsidian performes in the patterns.
-
-%% \subsubsection{Map}
-
-%% Map, to a  functional programmer is a higher order function that takes a pure 
-%% function and applies it to all elements of a collection (for example a list). 
-%% That meaning of map is easily implementable in Obsidian on both pull and push arrays. 
-%% Obsidian is also capable, via its parallel for loops, to map sequential computations 
-%% over threads in a block and block computation over many blocks. This is probably 
-%% more in the spirit of what the authors refer to in~\citet{STRUCTURED}.
-
-%% \subsubsection{Stencil}
-
-%% Stencil computations can be expressed in Obsidian, with some manual labour. 
-%% A stencil computation would have the form of a pull array to pull array function. 
-%% Pull arrays allow reading at position. 
-%% Stencil operations where more than one output position is touched could be 
-%% implemented as a pull array to push array function.
-%% One problem with stencil computations is the borders. If every thread needs to 
-%% execute a conditional, checking if it is at the border, performance would suffer. 
-%% Using Obsidian the programmer would likely generate more than one kernel; a fast 
-%% kernel for the internal points and one or more border kernel.
-
-%% \subsubsection{Reduction and scan} 
-%% Reductions can easily be implemented in Obsidian and with very good performance. 
-%% Obsidian also seems particularly well suited for experimenting with the kind 
-%% of trade offs that are interesting when implementing reduction kernels.
-
-%% Scans can also be implemented and the code is elegant.
-
-%% \subsubsection{Pack and expand}
-
-%% \begin{figure}[!ht]
-%% \begin{small} 
-%% \begin{tabular}{|l|l|l|l|l|l|l|}
-%% Parallel Pattern & TBB & Cilk Plus & OpenMP & ArBB & OpenCL & Obsidian \\
-%% \hline 
-%% Parallel Nesting & F   & F         &        &      &        &          \\
-%% Map              & F   & F         & F      & F    & F      & I        \\
-%% Stencil          & I   & I         & I      & F    & I      & I        \\ 
-%% Workpile         & F   &           &        &      & I      &          \\
-%% Reduction        & F   & F         & F      & F    & I      & I        \\ 
-%% Scan             & F   & I         & I      & F    & I      & I        \\ 
-%% Fork-Join        & F   & F         & I      &      &        &          \\ 
-%% Recurrance       &     & P         &        &      &        &          \\ 
-%% Superscalar seq. &     &           &        &      & F      &          \\
-
-%% Pack             & I   &  I        &  I     & F    & I      & I        \\ 
-%% Expand           & I   &  I        &  I     & I    & I      & *        \\
-%% Pipeline         & F   &  I        &  I     &      &        &          \\
-%% Geometric decomp. & I  &  I        &  I     & I    & I      & I        \\
-%% Search           &  I  &  I        &  I     & I    & I      & I        \\
-%% Category reduction & I &  I        &  I     & I    & I      & I        \\
-%% Gather           & I   &  F        &  I     & F    & I      & I        \\
-%% Atomic scatter   & F   &  I        &  I     &      & I      & I        \\
-%% Permutation scatter & F & F        &  F     & F    & F      & I        \\ 
-%% Merge scatter    &  I  &  I        &  I     & F    & I      & *        \\
-%% Priority scatter &     &           &        &      &        & *        \\
-
-%% \end{tabular}
-%% \end{small}
-%% \label{fig:caps}
-%% \caption{This figure lists what parallel patterns a number of languages support. The list 
-%% is copied from ``Structured Parallel Programming''~{\protect\citet{STRUCTURED}} and augmented with a 
-%% column for Obsidian. \\
-%% {\bf Key:} F - means the pattern is supported directly with an abstraction. P - indicates 
-%% that the pattern can be implemented in terms of other patterns appearing below. I - means 
-%% that the pattern can be easily implemented. Lastly, (*) indicates some patterns 
-%% that can be implemented in Obsidian but currently not efficiently so.} 
-%% \end{figure}
-
-%% In the technical report ``The Landscape of Parallel Computing Research: A View 
-%% from Berkeley''~\citet{dwarfs}, 13 computational kernels (called dwarfs) of importance is identified. 
-%% The paper also proposes that we use these 13 dwarfs to evaluate parallel programming 
-%% models and languages. The dwarfs are listed below, together with thoughts on how 
-%% well they can be tackled by Obsdian.
-
-%% \begin{itemize} 
-%% \item {\bf Dense Linear Algebra:} Obsidian works well for expressing vector-vector 
-%% operations. It is also possible to express operations such as matrix multiplication, 
-%% even though Obsidians support for multidimensional arrays is limited. It is possible 
-%% to use pull arrays of pull arrays, but not push arrays of push arrays.
-
-%% \item {\bf Sparse Linear Algebra:}
-
-%% \item {\bf Spectral Methods (FFT):} FFTs are implemented in a way that is 
-%% similar to sorting networks. A so-called butterfly network is used. We know 
-%% that we can express these network structures in Obsidian.
-
-%% \item {\bf N-Body Methods:}
-
-%% \item {\bf Structured Grids:} Regular grids should not be a problem for Obsidian. 
-%% Obsidian has no built in support for stencils, the combination of pull and push array 
-%% still provides the capability to express such computations.
-
-%% \item {\bf Unstructured Grids:} May be hard problem for GPUs in general. The paper 
-%% states that these problems have the property that ``Updates typically involve multiple 
-%% levels of memory reference indirection''. We know that the GPU thrives when memory access 
-%% patters are predictable and regular, which is the opposite scenario.
-
-%% \end{itemize}
 
@@
 
+\begin{figure}
+\begin{center}
+\begin{tikzpicture}[align=center, scale=2.5]
+% \draw[help lines] (0,0) grid (3,3);
+
+\draw[->] (-0.5,0) -- (-0.5,3) node[anchor=east] {Abstraction level};
+
+\draw (0,3) node[anchor=south west] {CPUs};
+\draw (1,3) node[anchor=south west] {GPUs};
+%\draw (2,3) node[anchor=south west] {Accelerator};
+\draw (2,3) node[anchor=south west] {CPUs, GPUs\\ Accelerators};
+
+\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
+\node[anchor=west] (acc) at (1,1.7) {CUB};
+\node[anchor=west] (repa) at (0,2.5) {DPH\\Nesl\\Repa\\Meta-repa\\Feldspar};
+\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
+\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
+
+\node[anchor=west] (cuda) at (1,0.3) {CUDA};
+\node[anchor=west] (opencl) at (2,0.3) {OpenCL};
+
+\node[anchor=west] (OpenMP) at (0,0.3) {OpenMP\\PThreads};
+
+\foreach \x in {0,...,2} { 
+  \foreach \y in {0,...,2} { 
+    \draw (\x,\y) rectangle (\x+1,\y+1);
+  }
+}
+
+\draw[red!75,rounded corners,ultra thick] (1,0.4) rectangle (2,2.4);
+\node[red!75,rotate=45,thick] (obs) at (1.5,1.5) {Obsidian};
+
+\draw[orange!90,rounded corners,ultra thick] (2,1.4) rectangle (3,2.6);
+\node[orange!90,thick] (arbb) at (2.5,2) {EmbArBB\\ArBB};
+
+\end{tikzpicture}
+\end{center}
+\caption{Placing our work in the landscape of languages and libraries for parallel 
+general purpose programming of CPUs, GPUs or both. }
+\label{fig:researchmatrix} 
+\end{figure}
+
 % ---------------------------------------------------------------------------
@@
 they are not general primitives. Where Accelerate has a single higher order reduction 
-operation, ArBB/EmbArBB has {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
+operation, ArBB/EmbArBB has \\ {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
 having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
@@
 to for anyone starting out with compiled embedded languages.
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[align=center, scale=2.5]
-% \draw[help lines] (0,0) grid (3,3);
-
-\draw[->] (-0.5,0) -- (-0.5,3) node[anchor=east] {Abstraction level};
-
-\draw (0,3) node[anchor=south west] {CPUs};
-\draw (1,3) node[anchor=south west] {GPUs};
-%\draw (2,3) node[anchor=south west] {Accelerator};
-\draw (2,3) node[anchor=south west] {CPUs, GPUs\\ Accelerators};
-
-\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
-\node[anchor=west] (acc) at (1,1.7) {CUB};
-\node[anchor=west] (repa) at (0,2.5) {DPH\\Nesl\\Repa\\Meta-repa\\Feldspar};
-\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
-\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
-
-\node[anchor=west] (cuda) at (1,0.3) {CUDA};
-\node[anchor=west] (opencl) at (2,0.3) {OpenCL};
-
-\node[anchor=west] (OpenMP) at (0,0.3) {OpenMP\\PThreads};
-
-\foreach \x in {0,...,2} { 
-  \foreach \y in {0,...,2} { 
-    \draw (\x,\y) rectangle (\x+1,\y+1);
-  }
-}
-
-\draw[red!75,rounded corners,ultra thick] (1,0.4) rectangle (2,2.4);
-\node[red!75,rotate=45,thick] (obs) at (1.5,1.5) {Obsidian};
-
-\draw[orange!90,rounded corners,ultra thick] (2,1.4) rectangle (3,2.6);
-\node[orange!90,thick] (arbb) at (2.5,2) {EmbArBB\\ArBB};
-
-\end{tikzpicture}
-\end{center}
-\caption{Placing our work in the landscape of languages and libraries for parallel 
-general purpose programming of CPUs, GPUs or both. }
-\label{fig:researchmatrix} 
-\end{figure}
 
@@
 
-Obsidian is a language for GPU kernel implementation. It is not as well integrated 
-into Haskell as for example Accelerate. There are ways to execute Obsidian kernels 
-from within Haskell; the recommended workflow, however, is to generate the CUDA  
-as text and write the host code by hand. This can be seen as both a shortcoming 
-and a feature. It is hard to integrate Accelerate programs in larger applications 
-written in other languages; when using Obsidian a kernel is generated and 
-can be used as part of for example a library.
+Obsidian is a language for GPU kernel implementation, not complete applications. 
+It is not as well integrated into Haskell as for example Accelerate. There are 
+ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
+section~\ref{sec:GPUPROGRAMMING}); the recommended workflow, 
+however, is to generate the CUDA as text and write the host code by hand. This 
+can be seen as both a shortcoming and a feature. It is hard to integrate 
+Accelerate programs in larger applications written in other languages; when 
+using Obsidian a kernel is generated and can be used as part of for example a library.
 
@@
 
-Here and there in the thesis it has been hinted that Obsidian has warp level 
-capabilites and that these are not as integrated into the hierarchy as the block, 
-thread and grid levels are. The warp abstractions offered also requires more work in 
-the code generating backend than the others, that correspond directly to CUDA concepts. 
-The problem is that CUDA does not have a warp abstraction. Warps needs to be 
-better integrated with the rest of Obsidian and the code generation required 
-for warps needs more testing before being trustworthy and useful.
+Warp level computations is work in progress and that these are not as well integrated 
+into the hierarchy as the block, thread and grid levels are. The warp abstractions 
+offered also requires more work in the code generating backend than the others, 
+that correspond directly to CUDA concepts. The problem is that CUDA does not have 
+a warp abstraction. Warps needs to be better integrated with the rest of Obsidian 
+and the code generation required for warps needs more testing before being 
+trustworthy and useful. When operational, warp level abstractions will help 
+the programmer make better use of the GPU.
 
@@
 
-\vspace{5mm}
-
-\emph{This section will revised.}
-
-\vspace{5mm}
+%\vspace{5mm}
+
+% \emph{This section will revised.}
+
+%\vspace{5mm}
 
@@
 and at little effort from us, very good performance is gained. The performance of 
-EmbArBB is entirely due to all the optimisation expertice and effort that was invested in 
+EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
 the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
@@
 that a CUDA programmer needs to take a leap from the single threaded program 
-parameterised on thread Id, to what actually will happen when this program is 
+parameterised on thread ID, to what actually will happen when this program is 
 run by hundreds or thousands threads at once. The extensive indexing arithmetic 

### 2013-11-10 src2/hl/paperThesis.tex
update

@@
 reverses :: DPull a -> DPush Grid a
-reverses arr = pConcatMap (pJoin . reverseLocal)
+reverses arr = pConcatMap reverseLocal
                           (splitUp 256 arr)
@@
 reverses' :: MemoryOps a => DPull a -> DPush Grid a
-reverses' arr = pConcatMap (pJoin . reverseLocal')
+reverses' arr = pConcatMap reverseLocal'
                            (splitUp 256 arr) 
@@
 \noindent
-The {\tt pConcatMap} combinator has the following type: 
+The {\tt pConcatMap} combinator is overloaded; here it is applied 
+at the following type: 
 %takes a local computation that results in a push array.
@@
 pConcatMap :: ASize l 
-            => (a1 -> SPush t a) 
+            => (a1 -> Program t (SPush t a))
             -> Pull l a1 
@@
 into a push array at the level above {\tt t}. Here, this is used to apply a block 
-level computation, giving a grid computation. The {\tt pJoin} function 
-has type \verb!Program t (Push t s a) -> Push t s a!; it creates a new push array 
-that computes the same results as the input computation.
+level computation, giving a grid computation.
 
@@
 largeReverse arr =
-  pConcat $ reverse $ fmap (pJoin . reverseLocal)
+  pConcat $ reverse $ fmap reverseLocal
                       (splitUp 256 arr) 
@@
 \begin{verbatim} 
-iters :: EWord32 
-         -> EWord32 
-         -> TProgram (SPush Thread EWord8)
-iters bid tid =
-  return $ fmap extract 
-         $ seqUntil (f bid tid) cond  (0,0,1)
+iters :: EWord32 -> EWord32 -> SPush Thread EWord8
+iters bid tid = fmap extract $
+                seqUntil (f bid tid) cond (0,0,1)
   where
@@
 \begin{verbatim} 
-genRect :: MemoryOps b
-           => EWord32
+genRect :: EWord32
           -> Word32
@@
 \begin{verbatim}
-mandel = genRect 512 512 (\b -> pJoin . iters b)
+mandel = genRect 512 512 iters
 \end{verbatim}
@@
               -> DPush Grid a
-mapRed1 f = pConcatMap $ pJoin . red1 f
+mapRed1 f = pConcatMap $ red1 f
 \end{Verbatim}
@@
     arr' <- force $  
-            pConcatMap (return . seqReduce f) 
+            pConcatMap (seqReduce f) 
                        (splitUp 8 arr)
@@
     arr' <- force $  
-            pConcatMap (return . seqReduce f)
+            pConcatMap (seqReduce f)
                        (coalesce 8 arr)
@@
 changing {\tt red5} to reduce 16 and 32 elements in the sequential phase.
-The performance of the fastest of these kernels is very satisfactory, at a level where the kernel is {\em memory bound}, that is constrained
-by memory bandwidth.
+The performance of the fastest of these kernels is very satisfactory, at 
+a level where the kernel is {\em memory bound}, that is constrained by 
+memory bandwidth.
 

### 2013-11-10 src2/thesis.tex
update

@@
 have multiple processing cores and Single Instruction Multiple Data (SIMD) units for 
-data-parallelism. Graphics processors (GPUs) brings massive parallelism at the cost of 
+data-parallelism. Graphics processors (GPUs) bring massive parallelism at the cost of 
 being harder to program than CPUs. This thesis applies embedded language methodology 
@@
 
-Each generation of CPU and GPU bringing more parallel resources. The question of how to write 
-programs for these processors efficiently arise. We are after efficiency both in programmer 
+CPUs and GPUs get more parallel resources with each new generation. The question of how to 
+program these processors efficiently arises. We are after efficiency both in programmer 
 productivity and in application performance. Using embedded languages allows us to experiment
@@
 In cases where the programmer is not interested in low-level details 
-or wants the program to run over a range of hardware a higher level language can be used.
+or wants the program to run over a range of hardware, a higher level language can be used.
 EmbArBB is a Haskell embedding or the Intel ArBB system. EmbArBB relies on the ArBB system 
@@
 a way to provide these benefits to the Haskell programmer and a way to increase usefulness
-of an existing system by opening it up to a wider audience. Obsidian is very different, it 
+of an existing system by opening it up to a wider audience. Obsidian is very different; it 
 is not based on a set of high-level parallel primitives. The Obsidian programmer 
-can implement these primitives in different ways, select the best one and then  abstract 
-upon that. We have obtained very good performance in case studies involving reductions.
+can implement these primitives in different ways and then select the best one. 
+We have obtained very good performance in case studies involving reductions.
 
@@
 understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
-and tips he has provided over he years. I cannot imagine that this would have worked with 
+and tips he has provided over the years. I cannot imagine that this would have worked with 
 any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
-supervisors. Working with Josef has been a very fun and rewarding experience.
+supervisors. Working with Josef has been fun and rewarding.
 
@@
   pConcat $
+  fmap push $ 
   zipWith vecAdd
@@
 \end{Verbatim}
-\end{small} % $
+\end{small}
 
@@
   pConcat $
+  fmap push $
   zipWith (zipWith (+)) 
@@
 \end{Verbatim} 
-\end{small} % $
+\end{small}
 

### 2013-11-10 src2/thesis.tex
spell pass

@@
 
-I would like to thank my supervisor Mary Sheeran. She is a very suportive, motivating and 
+I would like to thank my supervisor Mary Sheeran. She is a very supportive, motivating and 
 understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
@@
 
+I have often turned to Emil Axelsson with technical questions and he has always been
+happy to help. For this I am grateful.
+
 I thank Manuel Chakravarty for letting me visit his group at the University of New South Wales 
@@
 
+Thanks to my thesis opponent Stephen Edwards for valuable comments on a draft of this thesis.
+
 I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions. 
-
-Thanks to my thesis opponent Stephen Edwards for valuable comments on a draft of this thesis.
+I also thank my office mates, Michal and Nicholas, for their company. I thank Eva Axelsson, 
+who I have felt most comfortable to turn to for help and information, even when she was not 
+the appropriate channel.
 
@@
 In this case, a language with even higher level of abstraction is preferable. It then 
-becomes the job of a more advanced compiler to turn that into code for the targetted 
+becomes the job of a more advanced compiler to turn that into code for the targeted 
 platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
@@
 a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
-the blocks and of the grid are set by a controling program upon launch of the kernel. This 
-controling program is run by the CPU of the computer housing the GPU. The kernel is launched 
+the blocks and of the grid are set by a controlling program upon launch of the kernel. This 
+controlling program is run by the CPU of the computer housing the GPU. The kernel is launched 
 using the following syntax:
@@
 iteration schema over threads within a block. There are other iteration schemas to 
-chose from such as sequential, combinbations of sequential and parallel
+chose from such as sequential, combinations of sequential and parallel
 across both threads and blocks (two level parallelism).
@@
 arrays we also get fusion of operations as default. The implementation of map 
-(shown below) illustrates how maping a function {\tt f} over a push array results 
+(shown below) illustrates how mapping a function {\tt f} over a push array results 
 in an application of {\tt f} in the write-function.
@@
 
-In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrance sort 
+In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrence sort 
 using Obsidian. We use very low-level features of Obsidian and end up producing 
@@
 for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
-of embedding prexisting DSLs.
+of embedding preexisting DSLs.
 
@@
 ``Compiling Embedded Languages'' by Elliott and Finne has been a constant source of information 
-and something we have refered to in almost all our work. This is, in my opinion, the paper to turn 
+and something we have referred to in almost all our work. This is, in my opinion, the paper to turn 
 to for anyone starting out with compiled embedded languages.
@@
 ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
-section~\ref{sec:GPUPROGRAMMING}); the recommended workflow, 
+section~\ref{sec:GPUPROGRAMMING}); the recommended work flow, 
 however, is to generate the CUDA as text and write the host code by hand. This 
@@
 EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
-the ArBB system. We get all that expertice for free. What EmbArBB provides is a 
+the ArBB system. We get all that expertise for free. What EmbArBB provides is a 
 way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 

### 2013-11-11 src2/thesis.tex
update

@@
 a balance between high level, productivity increasing abstractions and low-level 
-control needed for performance. Obsidian programs are compiled into CUDA C code.
+control needed for performance. The Obsidian programming model mirrors the GPU arhcitecture 
+and the programmer is constrained into writing GPU-friendly code. By supplying hierarchy level 
+polymorphic library functions, these constraints does not feel obtrusive. Obsidian programs are 
+compiled into CUDA C code. This compilation is based on a simple and elegant monad reification 
+technique.
 
@@
 can implement these primitives in different ways and then select the best one. 
-We have obtained very good performance in case studies involving reductions.
+We have obtained very good performance in case studies involving reductions. Obsidian 
+programs are also more terse and composable, compared to CUDA.
 

### 2013-11-11 src2/thesis.tex
spell pass

@@
 a balance between high level, productivity increasing abstractions and low-level 
-control needed for performance. The Obsidian programming model mirrors the GPU arhcitecture 
+control needed for performance. The Obsidian programming model mirrors the GPU architecture 
 and the programmer is constrained into writing GPU-friendly code. By supplying hierarchy level 
@@
 
-I have often turned to Emil Axelsson with technical questions and he has always been
+I have often turned to Emil Axelsson with technical questions, and he has always been
 happy to help. For this I am grateful.
@@
 I thank my friends for being understanding during times when I have been very self-absorbed. 
-I am very grateful to my wife Andrea for supporting and believing in me and our daughter Noemi 
+I am very grateful to my wife Andrea for supporting and believing in me and to our daughter Noemi 
 for bringing much joy to my breaks from work while writing this thesis; soon she is old enough 

### 2013-11-11 src2/hl/paperThesis.tex
update

@@
 \hline 
-  Kernel & Threads & Shared mem. & ms & ms*\\ \hline 
-  red1 & 1024 & 16384 & 9.619 & 1.678 \\ \hline
-  red2 & 1024 & 16384 & 9.174 & 1.838 \\ \hline
-  red3 & 1024 & 16384 & 8.795 & 1.855 \\ \hline
-  red4 & 256 & 2048  & 10.839 & 2.120  \\ \hline
-  red5 & 256 & 2048  & 2.608 &  0.445 \\ \hline
-  red6 & 128 & 1024  & 2.471 &  0.442 \\ \hline
-  red7 & 64 & 512  & 2.524 &    0.439 \\ \hline
+  Kernel & Threads & Shared mem. & ms \\ \hline 
+  red1 & 1024 & 16384 & 2.085 \\ \hline
+  red2 & 1024 & 16384 & 1.748 \\ \hline
+  red3 & 1024 & 16384 & 1.741 \\ \hline
+  red4 & 256 & 2048  & 2.113  \\ \hline
+  red5 & 256 & 2048  & 0.455  \\ \hline
+  red6 & 128 & 1024  & 0.441  \\ \hline
+  red7 & 64 & 512    & 0.441  \\ \hline
 \end{tabular}
@@
 2048 elements on 
-and on a GTX680 (ms*). The timing figures were obtained using the NVIDIA Profiler.}
+and on a GTX680. The timing figures were obtained using the NVIDIA Profiler.}
 \label{fig:reducetable} 
@@
 \hline 
-  Kernel & Threads & Shared mem. & ms & ms*\\ \hline 
-  red1l & 1024 & 16384 & 8.883 & 1.762 \\ \hline
-  red2l & 1024 & 16384 & 5.721 & 1.030  \\ \hline
-  red3l & 1024 & 16384 & 5.377 & 0.980  \\ \hline
-  red4l & 256 & 2048  & 11.202 & 2.201 \\ \hline
-  red5l & 256 & 2048  & 2.476  & 0.440 \\ \hline
-  red6l & 128 & 1024  & 2.471  & 0.442 \\ \hline
-  red7l & 64 & 512  & 4.712    & 0.752\\ \hline
+  Kernel & Threads & Shared mem. & ms \\ \hline 
+  red1l & 1024 & 16384 & 1.567   \\ \hline
+  red2l & 1024 & 16384 & 1.073   \\ \hline
+  red3l & 1024 & 16384 & 0.976   \\ \hline
+  red4l & 256 & 2048  &  2.240   \\ \hline
+  red5l & 256 & 2048  &  0.443   \\ \hline
+  red6l & 128 & 1024  &  0.440   \\ \hline
+  red7l & 64 & 512  &    0.440   \\ \hline
 \end{tabular}
-\caption{ Execution time of a grid of 4096 reductions, each of 4096 elements on a GT650m GPU (ms)
-and on a GTX680 (ms*). The timing figures were obtained using the NVIDIA Profiler.}
+\caption{ Execution time of a grid of 4096 reductions, each of 4096 elements on 
+a GTX680. The timing figures were obtained using the NVIDIA Profiler.}
 \label{fig:reducetablel} 
@@
 \hline 
-  System            & Elements & ms     & ms*   \\ \hline 
-  Obsidian and CUDA & $2^{24}$  & 2.480  & 0.446 \\ \hline
-  Accelerate        & $2^{24}$  & 2.716  & 0.478 \\ \hline
+  System            & Elements & ms    \\ \hline 
+  Obsidian and CUDA & $2^{24}$  & 0.449 \\ \hline
+  Accelerate        & $2^{24}$  & 0.478 \\ \hline
 \end{tabular}
 \caption{Running times of $2^{24}$ element reduction using Obsidian and CUDA 
-or Accelerate. The results were obtained on a NVIDIA GT650m GPU (ms) 
-and on a NVIDIA GTX680 GPU (ms*).}
+or Accelerate. The results were obtained on a NVIDIA GTX680 GPU.}
 

### 2013-11-11 src2/thesis.tex
update

@@
 CPUs and GPUs get more parallel resources with each new generation. The question of how to 
-program these processors efficiently arises. We are after efficiency both in programmer 
+efficiently program these processors arises. We are after efficiency both in programmer 
 productivity and in application performance. Using embedded languages allows us to experiment
@@
 control needed for performance. The Obsidian programming model mirrors the GPU architecture 
-and the programmer is constrained into writing GPU-friendly code. By supplying hierarchy level 
-polymorphic library functions, these constraints does not feel obtrusive. Obsidian programs are 
-compiled into CUDA C code. This compilation is based on a simple and elegant monad reification 
-technique.
+and the programmer is constrained into writing GPU-friendly code. Hierarchy level 
+polymorphic library functions are supplied to make these constraints feel less obtrusive. 
+Obsidian programs are compiled into CUDA C code. This compilation is based on a simple 
+and elegant monad reification technique.
 
@@
 The work behind paper H was done by me as a hobby project and my aspirations were not 
-high. I just wanted to experiment with some embedded languages techniques that we had 
+high. I just wanted to experiment with some embedded language techniques that we had 
 been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
@@
 Computers today are increasingly parallel. Each new generation of general purpose processors
-(CPUs) and of graphics processors (GPUs), brings more parallelism. The question 
+(CPUs) and of graphics processors (GPUs) brings more parallelism. The question 
 of how to program these computers most efficiently, both from a productivity and 
@@
 often misses the target when it comes to performance. The book ``Structured Parallel 
-Programming'' by Michael McCool et al argues for the need of explicit parallel programming 
+Programming'' by Michael McCool et al argues the need for explicit parallel programming 
 models~\citet{STRUCTURED}.
@@
 We can experiment with what abstractions to expose to the programmer, without much 
-effort. Other benefits of embedded languages often mentioned are: making use 
-of the host language's library functions, if they are general enough. The host language 
-can be used as a powerful macro language for the embedded language. An embedded 
-language also provides a more leveled learning curve to programmers already familiar 
-with the host language. For these programmers, using an embedded language is more like  
-learning about a new library.
+effort. There are many other benefits of embedded languages: making use 
+of the host language's library functions (if they are general enough), the host language 
+can be used as a powerful macro language for the embedded language, providing a more 
+leveled learning curve to programmers already familiar with the host language. 
+For programmers who are already proficient in the host language, learning an embedded 
+language is more like picking up a new library.
 
@@
 general purpose computations on GPUs and EmbArBB for parallel programming across 
-platforms, CPU, GPU and other accelerators.
-
-Graphics Processors (GPUs) trade programmer convenience for increased parallelism. 
+platforms, such as CPUs, GPUs and other accelerators.
+
+Graphics Processors trade programmer convenience for increased parallelism. 
 For example, where CPUs dedicate large portions of chip area to caches, branch 
@@
 (a C-dialect for GPU programming) experts, provided that both programmers have 
-similar GPU expertise. The Obsidian programmer would however do so in a more 
+similar GPU expertise. The Obsidian programmer would however write programs in a more 
 composable manner. Obsidian programs are compiled into CUDA.
@@
 operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
-Haskell embedding of Intel ArBB that was released as an embedded language in C++. 
-EmbArBB relies entirely on the ArBB system for code generation and optimisation.
+Haskell embedding of Intel ArBB~\citet{ARBB2011} that was released as an embedded language 
+in C++. EmbArBB relies entirely on the ArBB system for code generation and optimisation.
 
@@
 
-In 2006 NVIDIA released their first Compute Unified Device Architecture (CUDA) 
+In 2006, NVIDIA released their first Compute Unified Device Architecture (CUDA) 
 GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
-for general purpose programming. A dialect of C, called CUDA C, was released 
+for general purpose programming~\citet{wwwcuda}. A dialect of C, called CUDA C, was released 
 specifically to simplify the implementation of general purpose computations on GPUs.
@@
 GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
-of the details that make GPU programming hard is more apparent when looking at the 
+of the details that make GPU programming hard are more apparent when looking at the 
 underlying hardware.
@@
 also divided into {\em Blocks}; the threads within a block can communicate using the 
-shared memory. The maximum block size is currently 1024 threads. Since 
-the number of threads in a block can be higher than the warp size, a barrier 
-synchronisation mechanism exists to ensure that threads within a block 
-can communicate safely via the shared memory. Blocks are also grouped into a grid that 
+shared memory. The maximum block size is currently 1024 threads. Threads within 
+a warp can communicate via the shared memory without using any synchronisation 
+primitive. However, if communication takes place across warps synchronisation is 
+necessary. A barrier synchronisation mechanism exists to ensure that all threads within a block 
+have reached the same position in the code. Blocks are also grouped into a grid that 
 is the collection of blocks executing the same program.
@@
 \subsection{GPU programming} 
-\label{GPUPROGRAMMING}
+\label{sec:GPUPROGRAMMING}
 
 The CUDA programming model reflects the hierarchical architecture of the GPU. The 
-programmer writes what are called \emph{kernels}; in CUDA jargon a kernel refers to a sequential 
+programmer writes what are called \emph{kernels}; in CUDA jargon, a kernel refers to a sequential 
 C program that is parameterised on a thread's identity. The kernel program is executed 
@@
 
-In Obsidian, the same kernel (the local computation) is implemented as follows:
+\noindent In Obsidian, the same kernel (the local computation) is implemented as follows:
 
@@
 we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
-In Haskell we know this is true for types that are member of the {\tt Num} typeclass.
+In Haskell we know that types that belong to the {\tt Num} typeclass has a {\tt (+)} operation.
 
 There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
-we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds check 
+we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds checks 
 which would mean conditionals in the code. Conditionals are problematic from a performance 
@@
 conditional that only executes the body for certain blocks. Discussion of the causes for 
-that conditional is postponed until a later section~\ref{sec:OBSARRAYS}. A real 
-difference is that the kernel is generated for a fixed size, here there is a constant (32) 
+that conditional is postponed until section~\ref{sec:OBSARRAYS}. A real 
+difference is that the kernel is generated for a fixed size; here there is a constant (32) 
 where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
@@
 we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
-correspond to actual data in memory. Pull and push arrays, rather represent ways 
+correspond to actual data in memory. Pull and push arrays rather represent ways 
 to compute arrays. More information about these array representations is available 
@@
     if (blockIdx.x < n0 / 32U) {
-        ((int32_t*) sbase)[tid] = input0[blockIdx.x * 32U + tid] * 2;
+        ((int32_t*) sbase)[tid] = 
+           input0[blockIdx.x * 32U + tid] * 2;
         __syncthreads();
-        output1[blockIdx.x * 32U + tid] = ((int32_t*) sbase)[tid] + 1;
+        output1[blockIdx.x * 32U + tid] = 
+           ((int32_t*) sbase)[tid] + 1;
     }
@@
 \noindent This code uses a shared memory array called {\tt sbase}. There 
-is also a call to {\tt \_\_syncthreads} to ensure that all the writes have 
+is also a call to \\ {\tt \_\_syncthreads} to ensure that all the writes have 
 completed before proceeding. In this case the synchronisation is not really
-necessary since all threads belong to the same warp. There is a {\tt unsafeforce} 
+necessary since all threads belong to the same warp. There is an {\tt unsafeforce} 
 operation that naively removes synchronisations if the array being forced is short 
@@
 and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
-This means that, first of all one must evaluate if the computation to be performed is significant 
+This means that, first of all, one must evaluate if the computation to be performed is significant 
 enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
 better to do the work on the CPU. However, if the data is already in device memory and 
-the operation that is about to perform on it is more efficiently handled by the CPU, it may still 
-be better to let the GPU do the work. In other words, compute near the data unless transferring it 
-pays off.
+the operation that is about to be performed on it is more efficiently handled by the CPU, it 
+may still be better to let the GPU do the work. In other words, compute near the data unless 
+transferring it pays off.
 
@@
 GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
-to as many transactions as cache lines touched (128b l1 cache lines). This means that strided 
+to as many transactions as cache lines touched (128 Byte l1 cache lines). This means that strided 
 accesses from within a warp are not favoured.
@@
 Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
-the permutation used will decide the access pattern. If this method can be applied or 
-not partly also depends on the algorithm being implemented.
+the permutation used will decide the access pattern. If global memory accesses can 
+be coalesced or not partly also depends on the algorithm being implemented.
 
@@
 Shared memory can also be used to coalesce reads from global memory. Shared memory is 
-a limited resource in the MP, so if a kernel uses very much shared memory it means 
+a limited resource in the MP, so if a kernel uses very much shared memory, 
 fewer such blocks can be active on the MP.
@@
 not a multiple of 32 leads to wasting GPU resources; some processing elements will stand 
-idle during the execution of the not full warps.
+idle during the execution of the incomplete warps.
 
@@
 Sometimes divergent code cannot be avoided. However, it is important not to have threads 
-diverge unnecessarily. When using pull arrays alone this was often a problem; with the 
-addition of push arrays the situation improved. Using push arrays some divergent computations 
+diverge unnecessarily. When using pull arrays alone, this was often a problem; with the 
+addition of push arrays, the situation improved. Using push arrays, some divergent computations 
 can be explicitly split up into phases, each phase with all threads following the 
@@
 
-In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
+In Obsidian, it is impossible for a barrier synchronisation to appear within diverging 
 conditional code. Conditionals can be introduced either by an \\{\tt ifThenElse} function 
@@
 thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs 
-({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a 
+({\tt GProgram}). the {\tt Cond} conditional chooses between two thread programs while a 
 synchronisation can only appear in an Obsidian block program. More information about programs 
@@
 writing GPU kernels. The lesson is that fast memory near the processing element 
-should be preferred over slow memory far away. However, this comes at the effort 
+should be preferred over slow memory far away. Favouring fast memory demands the effort 
 of decomposing computation and data in a way that makes this possible.
@@
 is a compiled embedded language; a very good paper that is highly relevant to our 
-approach, is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
+approach is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
 This paper describes how to embed a compiler backend and generate efficient code 
@@
 level of abstraction in fields where the default is to use low-level languages. 
-For example, Lava and Wired~\citet{LAVA,Wired} raises the level of abstraction in the 
+For example, Lava and Wired~\citet{LAVA,Wired} raise the level of abstraction in the 
 fields of hardware design and verification and Feldspar does the same in the area of  
@@
 platforms \citet{ARBB2011}. Related work and information about the tools, libraries 
-and language that make the wider context surrounding Obsidian can be found in 
+and languages that make the wider context surrounding Obsidian can be found in 
 section~\ref{sec:relatedwork}.
@@
 Obsidian uses a deeply embedded {\tt Program} data type that represents GPU Kernels. 
-A CUDA GPU has a hierarchy of parallel resources. At the bottom there are threads, each 
+A CUDA GPU has a hierarchy of parallel resources. At the bottom there are threads,  
 executing sequential programs. There are groups of threads (of 32) called {\em Warps} that 
@@
 {\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
-in the hierarchy leading to warp programming being a special case. One benefit a warp has is 
+in the hierarchy, leading to warp programming being a special case. One benefit a warp has is 
 that threads can communicate via shared memory without using synchronisation primitives. There 
@@
 
-Scalars and operations on scalars are represented by an expression data type ({\tt Exp a}) in 
+Scalars and operations on scalars are represented by an expression data type \\ ({\tt Exp a}) in 
 Obsidian. This is another example of a deep embedding; the expression data type contains 
@@
 in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
-Datatype (GADT) to obtain typed expressions. An alternative to the GADT, is to use phantom 
-types. Earlier versions of Obsidian (section~\ref{sec:paperB}) used phantom types.
+Datatype (GADT) to obtain typed expressions. An alternative to using a GADT is to use phantom 
+types, which earlier versions of Obsidian did (section~\ref{sec:paperB}).
 
@@
 Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
-parameter is used to be able to represent arrays with both static (known at Haskell runtime) 
+parameter is used to be able to represent arrays with static (known at Haskell runtime) 
 or dynamic (length represented by an expression, an unknown length). When creating block 
 level computations, we want to know the exact lengths in order to avoid generating code 
-riddled with conditionals concerning array lengths. It is possible to write kernels that 
+riddled with conditionals concerning array lengths. In CUDA, it is possible to write kernels that 
 are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
 conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
-when performance is crucial. Dynamic lengths are used for grid level computations, this allows 
-a local computation of a fixed size to be run in parallel over arrays a multiple of the size that 
-the local computation handles. In generated code this is currently visible as a conditional 
-on {\tt blockIdx.x}. However, this is still considered work in progress; the conditional is 
-only really necessary for Obsidian kernels that have more than one output array and the outputs
-are of different length. 
+when performance is crucial. In Obsidian, we have chosen to disallow implementation of kernels 
+that operate on different sizes locally.  Dynamic lengths are used for grid level computations; 
+this allows a local computation of a fixed size to be run in parallel over arrays a multiple
+of the size that the local computation handles. In generated code, this is currently visible 
+as a conditional on {\tt blockIdx.x}. However, this is still considered work in progress; the 
+conditional is only really necessary for Obsidian kernels that have more than one output array 
+and the outputs are of different length. 
 %Conditionals are problematic on GPUs and 
@@
 
-Another benefit of pull arrays are that they are parallel. Simply put, early Obsidian was pull 
+Another benefit of pull arrays is that they are parallel. Simply put, early Obsidian was pull 
 arrays in addition to a way to compute the array and store its elements to memory 
@@
 operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
-or interleaving of pull arrays result in conditionals in the indexing function. The 
+and interleaving of pull arrays results in conditionals in the indexing function. The 
 interleaving case, which is worst, leads to diverging branching in every warp. If branches 
-diverge, the GPU will turn off the processing elements taking one path and 
-allow the others to progress; then turning to the other branch. The computation of the 
+diverge, the GPU turns the processing elements taking one path off and 
+allow the others to progress; then turn to the other branch. The computation of the 
 branches is serialised and compute resources are wasted.
@@
 When working with pull arrays, the consumer of such an array decides how to 
-iterate over the array and what elements to compute. This is done applying the 
+iterate over the array and what elements to compute. This is done by applying the 
 push array function to those indices that are interesting to the consumer. Push 
-arrays work in the opposite way. The Push array function (the result program) contain 
+arrays work in the opposite way. The Push array (the result program) contains 
 the iteration schema; the consumer of the push only decides what write function 
-to supply. The iteration schema used is decided upon when creating a push arrays. 
+to supply. The iteration schema used is decided upon when creating a push array. 
 The code below is an example of how to create a push array from a pull array.
@@
 
-Push arrays give many of the same benefits that pull arrays have. Using push 
-arrays we also get fusion of operations as default. The implementation of map 
+Push arrays give many of the same benefits as pull arrays. Using push 
+arrays, we also get fusion of operations as default. The implementation of map 
 (shown below) illustrates how mapping a function {\tt f} over a push array results 
@@
 Paper D (section~\ref{sec:paperD}) introduces push arrays and uses them in the 
-implementation of sorting networks to quite good results. With push arrays we 
+implementation of sorting networks, with quite good results. With push arrays we 
 were able to experiment with more detailed decompositions (what thread computes 
 what value) of the algorithms.
-
-In section \ref{sec:efficient}, guidelines for writing efficient CUDA code is listed. 
-Here we revisit these guidelines and explain how they are approached by Obsidian and 
-the Obsidian programmer.
 

### 2013-11-11 src2/thesis.tex
update

@@
 Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
-the permutation used will decide the access pattern. If global memory accesses can 
-be coalesced or not partly also depends on the algorithm being implemented.
+the permutation used will decide the access pattern. In section~\ref{sec:OBSARRAYS}, there 
+is more information about pull arrays. If global memory accesses can 
+be coalesced or not partly also depends on the algorithm being implemented. The next paragraph 
+suggests a trick for coalescing memory accesses, in some cases, even if the algorithm is 
+unsuitable. Then it becomes a trade between bad memory access pattern and using more 
+shared memory.
 
 \item {\bf Minimise use of global memory:} Accessing global device memory can take as 
-many as 400 to 600 clock cycles if the data is not cached. The programmer should 
-use shared memory as much as possible to avoid redundant loading from global memory.
-Shared memory can also be used to coalesce reads from global memory. Shared memory is 
-a limited resource in the MP, so if a kernel uses very much shared memory, 
-fewer such blocks can be active on the MP.
+many as 400 to 600 clock cycles if the data is not cached, this is 10 - 20 times as many cycles 
+as an access to shared memory. The programmer should use shared memory as much as possible 
+to avoid redundant loading from global memory. Shared memory can also be used to coalesce 
+reads from global memory. Shared memory is a limited resource in the MP, so if a kernel 
+uses very much shared memory, fewer such blocks can be active on the MP.
 
@@
 can be explicitly split up into phases, each phase with all threads following the 
-same execution path. This means that fewer threads are required to do the same work.
+same execution path. This means that fewer threads are required to do the same work. Push arrays 
+are explained in more detail in section~\ref{sec:OBSARRAYS}.
 

### 2013-11-11 src2/hl/paperThesis.tex
update

@@
 
-\subsection{Fractals}
+\subsubsection{Fractals}
 
@@
 
-\subsection{Combining kernels to solve large problems} 
+\subsubsection{Combining kernels to solve large problems} 
 \label{sec:Benchmarks}
@@
 
-\subsection{What should the combinators be?}
+\subsubsection{What should the combinators be?}
 In this latest version of Obsidian, we have at last got a system

### 2013-11-11 src2/thesis.tex
update

@@
 \item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
-\item {\bf Map, zipWith, stencil:}  {\tt map} 
+\item {\bf Map, zipWith, stencil:} are all supported via a single primitive called {\tt map}.
 \end{itemize}
@@
 constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
-language and calling the ArBB functions does not actually execute them, it just ads a node to 
+language and calling the ArBB functions does not actually execute them, it just adds a node to 
 some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
@@
 This is exactly the way these things work in Haskell embedded languages, but it may be even 
-more surprising to a C++ programmer. Before an ArBB function can be computed it needs to be 
+more surprising to a C++ programmer. Before an ArBB function can be computed, it needs to be 
 captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
-functionality. The first time {\tt call} is used on a function pointer that function is 
+functionality. The first time {\tt call} is used on a function pointer, that function is 
 executed and the ArBB functions used in it build an AST (we say the function has been 
 captured). The AST is then compiled and specialised for the hardware available. Any 
-subsequent calls of a function using ArBB functionality do not lead to a recompilation, 
-the first call caches this compiled code.
+subsequent calls of a function using ArBB functionality do not lead to a recompilation; 
+as the first call caches this compiled code.
 
@@
 The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
-is not useful for application implementation. A higher-level interface to ArBB to offer 
-the Haskell programmer is needed. Our first attempt was to implement an Accelerate backend 
+are not useful for application implementation. We need a higher-level interface to ArBB to offer 
+to the Haskell programmer. Our first attempt was to implement an Accelerate backend 
 using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
-more richer language and there was an API mismatch between Accelerate and ArBB. These problems 
-are mentioned in in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
-to the Haskell in a way that is useful and simpler to implement, we start working on EmbArBB.
-
-With EmbArBB we improve the interfacing with Haskell by implementing a deeply embedded language. 
-Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings creating 
+richer language and there was an API mismatch between Accelerate and ArBB. These problems 
+are mentioned in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
+to the Haskell programmer in a way that is useful and simpler to implement, we started 
+working on EmbArBB.
+
+EmbArBB improves the interfacing with Haskell by implementing a deeply embedded language. 
+Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings, creating 
 function objects that can then be run.
@@
 
-ArBB supports one, two and three dimensional arrays, called dense vectors. In EmbArBB we 
+ArBB supports one, two and three dimensional arrays, called dense vectors. In \\ EmbArBB, we 
 represent these with a data type, {\tt DVector}, parameterised by dimensionality and 
@@
 
-EmbArBB provides many of the ArBB operations on vectors but provides a 
-shape-polymorphic interface to them. For example, a reduction operation can take a two  
-dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
-it takes a one dimensional vector and returns a single value (zero dimensional vector).
+EmbArBB provides many of the ArBB operations on vectors via a 
+shape-polymorphic interface. For example, a reduction operation can take a two  
+dimensional vector and reduce all rows or columns and provide a one dimensional result. 
+It can also take a one dimensional vector and return a single value (zero dimensional vector).
 
@@
 
-When compiling EmbArBB to ArBB a sharing detection pass is performed. This pass detects 
+When compiling EmbArBB to ArBB, a sharing detection pass is performed. This pass detects 
 sharing in the AST using the method of A. Gill~\citet{Gill}. One benefit of this pass 
-is that it reduces the numbers of calls into the ArBB C library. The efficiency of code 
-generated is probably not affected since ArBB would discover that sharing in a Common 
+is that it reduces the number of calls into the ArBB C library. The efficiency of code 
+generated is probably not affected, since ArBB would discover that sharing in a Common 
 Subexpression Elimination (CSE) pass. Another potential benefit is that sharing detection 
-on the Haskell side results in a simpler AST being built on the ArBB side; potentially making 
+on the Haskell side results in a simpler AST being built on the ArBB side, potentially making 
 the job for the ArBB optimisation passes simpler. What effects these 
 operations truly had on performance has not been investigated. Actually, applying a sharing 
-detection technique was one of my personal motivations for working on EmbArBB. In Obsidian 
-we have not seen the need for this technique.
+detection technique was one of my personal motivations for working on EmbArBB. In Obsidian, 
+we have not seen the need for this technique. In Obsidian, sharing is in the hands of the 
+programmer, via the {\tt force} operation. On a GPU, communication is more expensive than 
+computation, so whether to share a computation or not becomes a tool for the programmer to use.
 
@@
 any optimisations in EmbArBB would very likely duplicate effort already put into ArBB. 
-Therefore we apply no further transformations after sharing detection. However, if EmbArBB 
+Therefore, we apply no further transformations after sharing detection. However, if EmbArBB 
 would be used to generate code in some other target language a set of optimisations 
@@
 In paper H, section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
-Repa library. EmbArBB compares favourably to Repa and is often a lot faster. EmbArBB and the ArBB 
-C++ performance is not distinguishable from EmbArBB's indicating that our Haskell embedding 
-adds no extra overhead. Of course this is expected, since once the ArBB system has generated 
+Repa library. EmbArBB compares favourably to Repa and is often a lot faster. The ArBB 
+C++ performance is not distinguishable from EmbArBB's, indicating that our Haskell embedding 
+adds no extra overhead. Of course, this is expected, since once the ArBB system has generated 
 the code, that code is identical no matter if the AST was built via Haskell or C++ calls. 
-Missing is a comparison of how well EmbArBB and C++ compares in regards to the time it actually 
-takes to go through the process of making those ArBB C API calls. However, if the generated 
-code is used many times the caching ArBB does of generated code means that the initial code 
-generation cost can be amortised.
+A comparison of how well EmbArBB and C++ compares in regards to the time it actually 
+takes to go through the process of making those ArBB C API calls has not been performed. 
+However, if the generated code is used many times, ArBB's caching of generated code means 
+that the initial code generation cost can be amortised.
 
@@
 into three groups, based on their level of abstraction. There are low-level languages; here 
-I place languages that are imperative and C-like. In the middle layer I place languages 
+I place languages that are imperative and C-like. In the middle, layer I place languages 
 that have higher level abstractions and are more easily composable. 
@@
 Nikola~\citet{NIKOLA} 
-and Thrust~\citet{THRUST} the programmer would use 
-high-level patterns without direct insight into how the application would be mapped onto 
-threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
+and Thrust~\citet{THRUST}, the programmer uses 
+high-level patterns, without direct insight into how the application is mapped onto 
+the threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
 more composable than CUDA and giving the programmer control of what the computation will 
@@
 The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
-composes to build his application. I place the ArBB and EmbArBB in a box slightly lower than 
-Accelerate since while ArBB/EmbArBB also have built in reduction primitives, just like Accelerate, 
+composes to build his application. I place ArBB and EmbArBB in a box slightly lower than 
+Accelerate,  since while ArBB/EmbArBB also have built in reduction primitives, just like 
+Accelerate, 
 they are not general primitives. Where Accelerate has a single higher order reduction 
@@
 % ---------------------------------------------------------------------------
-Microsoft Accelerator~\citet{ACCELERATOR}, is a language embedded in C\#. Accelerator has
+Microsoft Accelerator~\citet{ACCELERATOR} is a language embedded in C\#. Accelerator has
 data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
@@
 % ---------------------------------------------------------------------------
-There are many embedded languages implemented in Python that makes use of GPUs 
+There are many embedded languages implemented in Python that make use of GPUs 
 to improve performance. One example is Copperhead~\citet{copperhead}. Copperhead 
@@
 
-Repa (Regular Shape-polymorphic Arrays)~\citet{REPA}, is a Haskell library for 
+Repa (Regular Shape-polymorphic Arrays)~\citet{REPA} is a Haskell library for 
 data-parallel programming. 
@@
 techniques. Meta-Repa programs build ASTs that are then compiled (using template 
-Haskell) to Haskell code using low-level parts of the Repa library for its parallelism.
+Haskell) to Haskell code that uses low-level parts of the Repa library for its parallelism.
 The Meta-Repa approach to implementing the Repa library 
@@
 physics and other domains. Delite, being a framework and the languages implemented using it 
-very domains specific, was impossible to place in the matrix (figure~\ref{fig:researchmatrix}).
+very domains specific, was impossible to place in the matrix (Figure~\ref{fig:researchmatrix}).
 
@@
 
-Feldspar is an embedded language for digital signal processors~\citet{FELDSPAR}. Apart 
+Feldspar is an embedded language for digital signal processing~\citet{FELDSPAR}. Apart 
 from using the syntactic library~\citet{SyntacticICFP12}, it is quite similar to Obsidian. 
@@
 
-In paper A~(section \ref{sec:paperA}) we describe our method for compiling monadic embedded 
-languages. 
-This means to take a monadic computation and observe its structure in order to generate
-code. This work is closely related to a method described in ``The Constrained Monad Problem''
-by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper they solve a more 
+In paper A~(section \ref{sec:paperA}), we describe our method for compiling monadic embedded 
+languages. This work is closely related to a method described in ``The Constrained Monad Problem''
+by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper, they solve a more 
 general problem than just compilation, but our method is nice because it is easy to understand 
-and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
-also solves the problem of compiling monads using a continuation monad.
+and solves a single problem and does it well. In ``Generic monad constructs'', Persson et al 
+also solve the problem of compiling monads using a continuation monad.
 
@@
 It is not as well integrated into Haskell as for example Accelerate. There are 
-ways to execute Obsidian kernels from within Haskell (we have seen examples of this in 
+ways to execute Obsidian kernels from within Haskell (we have seen an example of this in 
 section~\ref{sec:GPUPROGRAMMING}); the recommended work flow, 
@@
 
-Obsidian is also a quite low-level language. The programmer needs to know GPU details 
+Obsidian is also quite a low-level language. The programmer needs to know GPU details 
 and is allowed to write programs at a level almost as low as CUDA, if it is desired. 
@@
 
-Warp level computations is work in progress and that these are not as well integrated 
-into the hierarchy as the block, thread and grid levels are. The warp abstractions 
-offered also requires more work in the code generating backend than the others, 
+Obsidian's support for warp level computations is work in progress. Currently, warp level 
+computations are not as well integrated into the hierarchy as the block, thread and grid 
+levels are. The warp abstractions 
+offered also require more work in the code generating backend than those 
 that correspond directly to CUDA concepts. The problem is that CUDA does not have 
@@
 in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
-and at little effort from us, very good performance is gained. The performance of 
+and, at little effort from us, very good performance is gained. The performance of 
 EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
@@
 programming without these. We believe that the GPU architecture is still volatile 
-enough that there be no ``the right'' way of implementing these primitives. As of 
+enough that there is no ``the right'' way of implementing these primitives. As of 
 today, the programmer still needs to explore the design space and find a solution 
@@
 are generated from is often shorter than corresponding CUDA programs. Obsidian 
-programs differ from CUDA programs by describing whole programs. With this I mean 
-that a CUDA programmer needs to take a leap from the single threaded program 
+programs differ from CUDA programs by describing whole programs. A CUDA programmer 
+needs to take a leap from the single threaded program 
 parameterised on thread ID, to what actually will happen when this program is 
-run by hundreds or thousands threads at once. The extensive indexing arithmetic 
+run by hundreds or thousands of threads at once. The extensive indexing arithmetic 
 often present in CUDA programs can make this leap a hard one. Similar programs 
@@
 as the example. Robot programs can be expressed using Haskell do notation 
-and is compiled down to a simple first order program representation. 
+and are compiled down to a simple first order program representation. 
 There is also a graphical simulator available that runs the compiled 
-code. Robot language and simulator is available at github: \url{github.com/svenssonjoel/Robot}.
+code. The robot language and simulator are available at github: \\ 
+\url{github.com/svenssonjoel/Robot}.
 
@@
 
-The first real paper about Obsidian, our embedded language for GPU 
-programming. In this paper we identify similarities between 
+This is the first real paper about Obsidian, our embedded language for GPU 
+programming. In this paper, we identify similarities between 
 connection patterns in hardware and parallel computations on GPUs. 
@@
 a different internal representation of GPU programs. We realised that 
-most of our kernels have similar shape, they are parallel operations 
+most of our kernels have similar shape; they are parallel operations 
 composed in sequence interspersed with barrier synchronisations. The 
@@
 This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
-array representation target specific performance problems that we want 
+array representation targets specific performance problems that we want 
 to solve with Obsidian. Here we have also gone back to using monadic 
@@
 an interesting variation on Counting sort that removes duplicate 
-elements and is a nice fit for GPUs because of its little need for 
+elements and is a nice fit for GPUs because of its reduced need for 
 synchronisation.
 
-In order to implement these sorting algorithms we added atomic operations 
+In order to implement these sorting algorithms, we added atomic operations 
 to Obsidian.
@@
 
-In this paper the most recent additions and improvements of Obsidian are 
+In this paper, the most recent additions and improvements to Obsidian are 
 described. Obsidian Programs are now parameterised on Hierarchy level. This forces
 the programmer into writing programs that we know how to compile to a GPU easily 
-(by limiting how parallel operations can be nested). The paper also gives a detailed  
-optimisation story using Obsidian for implementing reduction kernels.
+(by limiting how parallel operations can be nested). The central part of this paper 
+describes how to implement and optimise reduction kernels using Obsidian.
 

### 2013-11-11 src2/thesis.tex
update

@@
 deep embedding (in hindsight) the correct one. EmbArBB could be resurrected by implementing 
-a new code generating backend.
+a new code generating backend. However, a complete reimplementation of the ArBB API would be 
+a massive undertaking.
 
@@
 and very domain specific embedded languages. This could be a niche for Obsidian.
+
+Obsidian constrains the programmer into writing programs that we can easily compile to 
+CUDA. This is done by modelling the GPU programming in types attached to programs. A 
+currently ongoing master's project is investigating what effects removing these 
+types have. It is clear that if the programmer is not constrained to writing flat 
+(or very limited nested) programs, transformations are needed on the AST before 
+generating CUDA code.
 

### 2013-11-12 src2/hl/paperThesis.tex
update

@@
 concepts. A GPU can come with as few as one of these multiprocessors.
-The two GPUs used in our measurements are the GTX 680, which has eight 
-multiprocessors, with a total of 1536 processing cores, and the GT 650M, a 
-laptop GPU, which has 384 such cores. On these cores, groups of 32 threads 
-called {\em warps} are scheduled. There are a number of warp scheduling units 
-per multiprocessor. Within a warp, threads execute in lockstep (SIMD); 
-diverging branches, that is those that take different paths on different 
-threads within a warp, are serialised, leading to performance penalties.
+The GPUs used in our measurements are the GTX 680, which has eight 
+multiprocessors, with a total of 1536 processing cores, the GTX 670 with 1344 cores, 
+and the GT 650M, a laptop GPU, which has 384 such cores. On these cores, 
+groups of 32 threads called {\em warps} are scheduled. There are a number 
+of warp scheduling units per multiprocessor. Within a warp, threads execute 
+in lockstep (SIMD); diverging branches, that is those that take different 
+paths on different threads within a warp, are serialised, leading to 
+performance penalties.
 
@@
 {\em Grid} computations. The programmer specifies a sequential computation 
-to be carried out by each thread. Many instances of a sequential computation,  
+to be carried out by each thread. Many instances of a sequential computation  
 run in parallel over threads, to form a block. A number of such blocks run in 
@@
 function {\tt reverse}. The {\tt push} function converts a pull array to a push 
-array.
+array. {\tt SPull} and {\tt SPush} arrays have static lengths, this is explained in more 
+detail on page \pageref{pag:HLSTATIC}.
 
@@
 
+\pagebreak
 \noindent
@@
 For example, a kernel that outputs data to two different global
-arrays cannot be expressed.
+arrays cannot currently be expressed. Lifting this restriction 
+is work in progress.
 
@@
 force operation is split into two or more passes. This is a change compared
-to earlier versions of Obsidian where the length of arrays we could force 
-where limited to the maximum number of threads. In this version, the 
-threads are, in a sense, virtual. The mapping from virtual threads to 
-real threads is done as a step in the compilation.
+to earlier versions of Obsidian, where the length of arrays we could force 
+were limited to the maximum number of threads per block (1024 threads). In 
+this version, the threads are, in a sense, virtual. The mapping from virtual 
+threads to real threads is done as a step in the compilation.
 
@@
 
-There are type synonyms for the dynamic and static versions of  pull arrays:
+There are type synonyms for the dynamic and static versions of  pull arrays: \label{pag:HLSTATIC}
 
@@
 
-The programs that can be specified in Obsidian
-are represented in a {\tt Program} data type.
+The programs that can be specified in Obsidian are represented in a {\tt Program} data type.
 This data type also represents the lowest level programming interface 
-exposed by Obsidian. In reference ~\citehl{JOSEFJOELCSORT} we make use of these low-level 
-capabilities to implement histograms using atomic operations (not shown here).
+exposed by Obsidian. In reference ~\citehl{JOSEFJOELCSORT} we make use of the low-level 
+capabilities provided by that interface to implement histograms using atomic 
+operations (not shown here).
 
@@
 {\tt ForAll} executes a program body at the level below several times in parallel. 
-If the body of the {\tt ForAll} loop is a {\tt Thread} computation the {\tt ForAll} loop 
+If the body of the {\tt ForAll} loop is a {\tt Thread} computation, the {\tt ForAll} loop 
 itself is a {\tt Block} computations. Likewise, if the body is a {\tt Block} computation, 
@@
 that uses that write-function a number of times. Note that the 
-write-function produces a {\tt TProgram} (short for {\tt Program Thread}), This 
-restricts our push arrays from representing nested data parallelism. 
+write-function produces a {\tt TProgram} (short for {\tt Program Thread}). 
+As a result, push arrays cannot represent nested data parallelism. 
 This concept is best illustrated by showing how to turn a pull array into a push array.
@@
 The {\tt ASize} class has instances for both the static and dynamic lengths 
-and is used to obtain conversion to a representation used internally; it 
-provides a function called {\tt sizeConv} which can convert both static and 
-dynamic lengths to a {\tt Word32} expression.
+and is used to obtain conversion to a representation used internally; 
+the function {\tt sizeConv} converts both static and dynamic lengths to a {\tt Word32} expression.
 
@@
 
-The last step in compilation process uses the Language.C.Quote library to 
+The last step in the compilation process uses the Language.C.Quote library to 
 turn the {\tt IM} into CUDA code. This compilation stage takes a {\em configuration} 
@@
 \caption{ Execution time of a grid of 8192 reductions , each of
-2048 elements on 
-and on a GTX680. The timing figures were obtained using the NVIDIA Profiler.}
+2048 elements and on a GTX680. The timing figures were obtained using the NVIDIA Profiler.}
 \label{fig:reducetable} 
@@
 \begin{minipage}{.5\linewidth}
-\begin{tikzpicture} [scale = 0.5]
+\begin{tikzpicture} [scale = 0.6]
 \begin{axis} [xmax = 3, 
               xmin = 0, 
-              ymax = 6,
+              ymax = 1.5,
               ymin = 0,
@@
            mark size = 2pt] coordinates {
-    (0, 0.364 )
-    (1, 0.847 )
-    (2, 2.098 ) 
-    (3, 5.762 )
+    (0, 0.096 )
+    (1, 0.214 )
+    (2, 0.517 ) 
+    (3, 1.382 )
   };
@@
            mark size = 2pt] coordinates {
-    (0, 0.408 )
-    (1, 0.564 )
-    (2, 1.292 ) 
-    (3, 3.336 )
+    (0, 0.076 )
+    (1, 0.126 )
+    (2, 0.286 ) 
+    (3, 0.714 )
   };
@@
            mark size = 2pt] coordinates {
-    (0, 0.341 )
-    (1, 0.514 )
-    (2, 1.179 ) 
-    (3, 2.957 )
+    (0, 0.077 )
+    (1, 0.127 )
+    (2, 0.286 ) 
+    (3, 0.694 )
   };
@@
            mark size = 2pt] coordinates {
-    (0, 2.882 )
-    (1, 3.372 )
-    (2, 4.194 ) 
-    (3, 5.762 )
+    (0, 0.715 )
+    (1, 0.832 )
+    (2, 1.025 ) 
+    (3, 1.381 )
   };
@@
            mark size = 2pt] coordinates {
-    (0, 3.226 )
-    (1, 2.242 )
-    (2, 2.576 ) 
-    (3, 3.336 )
+    (0, 0.551 )
+    (1, 0.475 )
+    (2, 0.562 ) 
+    (3, 0.715 )
   };
@@
            mark size = 2pt] coordinates {
-    (0, 2.723 )
-    (1, 2.041 )
-    (2, 2.352 ) 
-    (3, 2.957 )
+    (0, 0.560 )
+    (1, 0.487 )
+    (2, 0.565 ) 
+    (3, 0.692 )
   };
@@
 \emph{Right:} varies the number of blocks so that the number of elements is the 
-same in each experiment. These results where obtained on an NVIDIA GT 650m using the 
-NVIDIA Profiler.}
+same in each experiment. These results where obtained on an NVIDIA GTX 670.}
 \label{fig:scangraphs}
@@
 Obsidian makes use of types that model the GPU hierarchy. This limits 
-the kind of programs one can express but also ensures that what we can write 
-is suitable for the GPU. An alternative is to remove these types and at the 
+the kind of programs that can be expressed, ensuring that the resulting programs are
+suitable for the GPU. An alternative is to remove these types and at the 
 same time make the compiler more clever. This would move some of the 
 required GPU detail knowledge from the programmer into compiler transformations. 
-We have recently started a project to investigate this change of Obsidian
-together with a Master's student.
+We have recently started a project to investigate this change to Obsidian,
+together with a Masters student.
 
 \subsubsection{To optimise or not to optimise} 
-The version of Obsidian described here does not try to use any compiler optimisations techniques. Instead, we are expecting that the 
+The version of Obsidian described here does not try to use any compiler 
+optimisation techniques. Instead, we are expecting that the 
 CUDA compiler will apply a good set of techniques, from common subexpression 
@@
 improve the usability of Obsidian. We have started a project with the 
-goal of implementing this language together with a Master's student.
+goal of implementing this language, together with a Masters student.
 

### 2013-11-12 src2/thesis.tex
update

@@
 and parallel prefix (scans). Often we have managed to get very good performance, close to that 
-of hand optimised CUDA code. However, there are also cases where we just cannot seem to reach 
-the level of performance that we would want. One such example is parallel prefix, an 
-application that has been haunting us from the very beginning. Scan is problematic to us because 
-it is best implemented as an in-place algorithm, while our array representations do 
-not allow for in-place update; there is no memory associated with them to update. 
-We are trying to improve this situation by adding a third array representation to Obsidian,
-mutable arrays.
+of hand optimised CUDA code. GPU execution of scan operations is something that has 
+been worked on extensively in the GPGPU community and very efficient implementations 
+exist~\citet{merrill,ScanCUDA,OLAMARCUS,EFFICIENTSCAN}. In Paper F (section~\ref{sec:paperF}), 
+we only study scan kernels (scan of small arrays) and measure their speed relative to eachother. 
+Compared to a scan kernel supplied with the CUDA API, our kernel fastest kernel is slighty more 
+than three times slower. Performance needs to be improved at the kernel level before we attempt 
+composing them into large scan operations. We hope to improve this situation with the 
+addition of the warp level abstractions in Obsidian (this is future work, 
+section~\ref{sec:thesisFUTURE}).
 
 Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
-a reduction kernel in Obsidian. At the end, the implemented reduction kernel is  
-faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
-reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
-been happy just getting close.
+a reduction kernel in Obsidian. At the end, the implemented reduction kernel 
+reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwith. 
+%faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
+%reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
+%been happy just getting close.
 
@@
 \section{Future work} 
+\label{sec:thesisFUTURE}
 

### 2013-11-12 blad/page1.tex
update

\thispagestyle{empty}

\begin{center}

\includegraphics{ChalmGUtextsvEng.eps}

\vfill

{\huge Embedded Languages for Data-Parallel Programming}

\vfill

{\LARGE \bf Bo Joel Svensson}

\vfill
The defence of this thesis for the degree of Doctor of Philosophy\linebreak
will be held in room EA, D&IT building,\linebreak
R\"annv\"agen 6B, Chalmers University of Technology,\linebreak
on {\bf December 16, 2013} {\bf at 10.00}.
\linebreak\linebreak
The defence will be held in English.

\vfill

Opponent: Professor Stephen A. Edwards,\linebreak
Department of Computer Science\linebreak
Columbia University\linebreak

\vfill

Department of Computer Science and Engineering\\
Chalmers University of Technology and G\"oteborg University\\
SE-412 96 G\"oteborg, Sweden\\
Telephone: +46 (0)31-772 1000

\vfill

\includegraphics{ChalmGUmarke.eps}

\end{center}

\newpage

### 2013-11-12 blad/page2.tex
update

--- 2013-11-11 src2/thesis.tex
@@
-\documentclass[a4paper]{book}
-
-\usepackage[paperwidth=169mm, paperheight=239mm]{geometry}
-
-%\usepackage[latin9]{inputenc}
-\usepackage[utf8x]{inputenc}
-\usepackage[pdftex]{graphicx}
-\usepackage{pdfpages}
-\usepackage[british]{babel}
-\usepackage{placeins}
-\usepackage{dialogue}
-\usepackage{fancyvrb}
-
-\usepackage{t1enc}         
-\usepackage{times}
-\usepackage{url}
-\usepackage{enumerate}
-
-%\usepackage{code}
-\usepackage{hyperref}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{mathtools}
-%\usepackage{lingmacros}
-\usepackage{subfig}
-\usepackage{fancyhdr}
-
-\pagestyle{plain}
-%\pagenumbering{arabic}
-
-%stuff
-\usepackage{latexsym}
-\usepackage{tikz,pgflibraryshapes}
-\usetikzlibrary{shapes,snakes,positioning,matrix}
-\usepackage{pgfplots}
-\usepackage{booktabs}
-\usepackage{enumerate}
-%\usepackage{etoolbox}
-
-\usepackage{multibib} 
-\newcites{bb}{Bibliography} 
-\newcites{ifl}{Bibliography}
-\newcites{papp}{Bibliography}
-\newcites{exp}{Bibliography}
-\newcites{hl}{Bibliography}
-\newcites{csort}{Bibliography}
-\newcites{emb}{Bibliography}
-\newcites{arbb}{Bibliography}
-\newcites{t}{Bibliography}
-%\newcites[ifl}{Bibliography}
-
-%\patchcmd{\thebibliography}{\section*}{\section}{}{}
-
-\makeatletter
-\renewenvironment{thebibliography}[1]
-     {\section*{\bibname}%
-      %\@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
-      \list{\@biblabel{\@arabic\c@enumiv}}%
-           {\settowidth\labelwidth{\@biblabel{#1}}%
-            \leftmargin\labelwidth
-            \advance\leftmargin\labelsep
-            \@openbib@code
-            \usecounter{enumiv}%
-            \let\p@enumiv\@empty
-            \renewcommand\theenumiv{\@arabic\c@enumiv}}%
-      \sloppy
-      \clubpenalty4000
-      \@clubpenalty \clubpenalty
-      \widowpenalty4000%
-      \sfcode`\.\@m}
-     {\def\@noitemerr
-       {\@latex@warning{Empty `thebibliography' environment}}%
-      \endlist}
-\makeatother
-
-%%
-
-%% \newcommand{\bequ}{\begin{quote}}
-%% \newcommand{\enqu}{\end{quote}}
-%% \newcommand{\eop}[1]{\mbox{\sl #1}}
-%% \newcommand{\sube}[2]{#1_{#2}}
-%% \newcommand{\id}[1]{{\it #1}}
-%% \newcommand{\kwd}[1]    {\mbox{\textbf{#1}}}
-%% \newcommand{\str}[1]{{\id "#1"}}
-%% \newcommand{\lemon}{\emph{lemon}}
-%% \newcommand{\textt}[1]{\texttt{#1}}
-%% \newcommand{\txttt}[1]{\texttt{#1}}
-
-%\renewcommand*{\thesection}{\arabic{section}}
-
-\newcommand{\thesistitle}{Embedded Languages for Data-Parallel Programming}
-\newcommand{\dept}{Department of Computer Science and Engineering}
-\newcommand{\uni}{Chalmers University of Technology and G\"oteborg University}
-\newcommand{\group}{Functional Programming Research Group}
-
-% ---------------------------------------------------------------------------
-% Papers 
-% ---------------------------------------------------------------------------
-\newcommand{\paperA}{Paper A}
-\newcommand{\paperATitle}{Simple and Compositional Reification of Monadic Embedded Languages}
-\newcommand{\paperB}{Paper B}
-\newcommand{\paperBTitle}{Obsidian: A Domain Specific Embedded Language for Parallel Programming of Graphics Processors}
-\newcommand{\paperC}{Paper C} 
-\newcommand{\paperCTitle}{GPGPU Kernel Implementation and Refinement using Obsidian} 
-\newcommand{\paperD}{Paper D}
-\newcommand{\paperDTitle}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
-\newcommand{\paperE}{Paper E}
-\newcommand{\paperETitle}{Counting and Occurrence Sort for GPUs using an Embedded Language}
-\newcommand{\paperF}{Paper F}
-\newcommand{\paperFTitle}{A High-Level Embedded Language for Low-Level GPU Kernel Programming}
-\newcommand{\paperG}{Paper G}
-\newcommand{\paperGTitle}{Programming Future Parallel Architectures with Haskell and Intel ArBB}
-\newcommand{\paperH}{Paper H}
-\newcommand{\paperHTitle}{Parallel Programming in Haskell Almost for Free}
-
-% ---------------------------------------------------------------------------
-% BEGIN ! 
-% ---------------------------------------------------------------------------
-\begin{document}
-
-\begin{titlepage}
-\begin{centering}
-{\sc Thesis for the Degree of Doctor of Philosophy}
-\vspace{30ex}
-
-{\LARGE\bf\thesistitle}
-
-\vspace{7ex}
-
-\large Bo Joel Svensson
-\vfill
-\includegraphics[width=120mm]{./img/ChalmGUtextsvEng}\\[5mm]
-\includegraphics[height=4cm]{./img/ChalmGUmarke}\\
-
-\vspace{1cm}
-\normalsize
-{\sc \dept}\\
-{\sc \uni}\\
-G\"oteborg, Sweden 2013
-
-\end{centering}
-\end{titlepage}
-
-% ---------------------------------------------------------------------------
-% Tryckortssida
-% ---------------------------------------------------------------------------
-%ditt ISBN är:  978-91-7385-939-4
-%ditt Ny serie nr:  3620
-%ISSN för publ.serien Doktorsavhandlingar vid Chalmers tekniska högskola : 0346-718X
-
-\quad \vfill
-
-{\noindent\large\bf\thesistitle} \\
-\noindent Bo Joel Svensson \\
-\noindent ISBN 978-91-7385-939-4 \\
-
-\vspace{1cm}
-
-\noindent\copyright {\sc {Bo Joel Svensson}}, 2013 \\ 
-\vspace{1cm}
-
-\noindent Doktorsavhandlingar vid Chalmers tekniska högskola \\
-\noindent Ny serie nr 3620 \\
-\noindent ISSN 0346-718X \\
-
-\vspace{1cm}
-
-\noindent Technical Report 101D\\
-\noindent \dept \\
-\noindent \group\\
-
-\vspace{1cm}
-
-\noindent \uni \\
-\noindent SE--412~96~~G\"oteborg, Sweden\\
-\noindent Phone: +46 (0)31--772~1000 \\
-
-\vspace{1cm}
-
-\noindent Printed in Sweden\\
-\noindent Chalmers Reproservice\\
-\noindent G\"oteborg, Sweden 2013
-
 \thispagestyle{empty}
 
-\clearpage
-\pagenumbering{roman}
-
-% ---------------------------------------------------------------------------
-% ABSTRACT !!  
-% ---------------------------------------------------------------------------
 \section*{Abstract}
 
+\vspace{5mm}
+
+\begin{small}
 Computers today are becoming more and more parallel. General purpose processors (CPUs) 
@@
 
-\clearpage
-
-% ---------------------------------------------------------------------------
-%  ACKS !!! 
-% ---------------------------------------------------------------------------
-\section*{Acknowledgments}
-
-I would like to thank my supervisor Mary Sheeran. She is a very supportive, motivating and 
-understanding supervisor. I also thank my co-supervisor Koen Claessen for all the great advice 
-and tips he has provided over the years. I cannot imagine that this would have worked with 
-any other supervisors. Thanks to Josef Svenningsson, a late addition to my troop of 
-supervisors. Working with Josef has been fun and rewarding.
-
-I have often turned to Emil Axelsson with technical questions, and he has always been
-happy to help. For this I am grateful.
-
-I thank Manuel Chakravarty for letting me visit his group at the University of New South Wales 
-and Trevor McDonell and Sean Lee for taking care of me during my visit. I look back 
-very fondly to my stay in Australia.
-
-For being a great supervisor during my three month internship at Intel, I thank Ryan Newton. 
-These where three very intense months away from family but at the same time a very rewarding 
-experience.
-
-Thanks to my thesis opponent Stephen Edwards for valuable comments on a draft of this thesis.
-
-I want to thank Erik, Ola, Markus and Viktor for our occasional lunches and GPU discussions. 
-I also thank my office mates, Michal and Nicholas, for their company. I thank Eva Axelsson, 
-who I have felt most comfortable to turn to for help and information, even when she was not 
-the appropriate channel.
-
-Finally I want to thank my family and friends. Thanks to my sisters and parents for their support.
-I thank my friends for being understanding during times when I have been very self-absorbed. 
-I am very grateful to my wife Andrea for supporting and believing in me and to our daughter Noemi 
-for bringing much joy to my breaks from work while writing this thesis; soon she is old enough 
-for real LEGO.
-
-\vspace{5mm}
-\noindent This research has been funded by the Swedish Foundation for
-Strategic Research (which funds the Resource Aware Functional 
-Programming (RAW FP) Project) and by the Swedish Research Council.
-
-\clearpage
-
-% ---------------------------------------------------------------------------
-%  Guide for Opponent, grading committee
-% ---------------------------------------------------------------------------
-\section*{Publications}
-
-This thesis includes the following publications:
-
-\begin{enumerate}[A.] 
-\item Josef Svenningsson and Bo Joel Svensson.
-Simple and Compositional Reification of Monadic Embedded Languages, 2013.
-\emph{18th ACM SIGPLAN International Conference of Functional Programming, ICFP 2013.}
-\item Bo Joel Svensson, Mary Sheeran and Koen Claessen.
-Obsidian: A Domain Specific Embedded Language for General Purpose Parallel Programming of Graphics Processors.
-In \emph{Proc.
-of Implementation and Applications of Functional Languages (IFL)}, Lecture Notes in Computer Science, Springer Verlag, March 2009.
-\item Bo Joel Svensson, Koen Claessen and Mary Sheeran.
-GPGPU kernel implementation and refinement using Obsidian.
-\emph{Practical Aspects of High-level Parallel Programming, PaPP 2010.} Procedia Computer Science.
-\item Koen Claessen, Mary Sheeran and Bo Joel Svensson.
-Expressive Array Constructs in an Embedded GPU Kernel Programming Language.
-In \emph{Proceedings of the 7th workshop on Declarative aspects and applications of multicore programming}, DAMP '12.
-ACM.
-\item Josef Svenningsson, Bo Joel Svensson and Mary Sheeran.
-Efficient Counting Sort Implementations using an Embedded GPU Programming Language.
-\emph{2nd Workshop on Functional High-Performance Computing, FHPC '13.}
-\item Bo Joel Svensson and Mary Sheeran. A High-Level Embedded Language for Low-Level GPU Kernel Programming. {\bf\emph{This work has not yet been published.}} 
-\item Bo Joel Svensson and Ryan Newton.
-Programming Future Parallel Architectures with Haskell and ArBB.
-\emph{Future Architectural Support for Parallel Programming (FASPP), in conjunction with ISCA '11.}
-\item Bo Joel Svensson and Mary Sheeran.
-Parallel Programming in Haskell Almost for Free: an embedding of Intel's Array Building Blocks.
-\emph{Proceedings of the 1st ACM SIGPLAN Workshop on Functional High-Performance computing, FHPC '12.}
-\end{enumerate}
-
-\clearpage
-\subsection*{Personal contributions}
-
-The following describes my personal contributions to the papers included 
-in this thesis.
-
-\subsubsection{Paper A}
-I intuitively applied the monad reification method as part of the implementation 
-of Obsidian. Josef Svenningsson identified its importance and suggested we write 
-about it. The paper has two distinct parts. Sections one and two describe the 
-method as I used it. And section four makes the compositional aspect of the method 
-explicit.
-
-The code and examples in section one and two are my work. The contents of section four was 
-developed by Josef Svenningsson and me. For section four we also got very much and valuable help 
-from Emil Axelsson.
-
-For the rest of the paper, just substituting my name for Benny and Josef's for Bj\"orn
-will give a good indication of our contributions.
-
-\subsubsection{Papers B, C, D and E} 
-The Obsidian language has grown out of discussions between Mary Sheeran, Koen Claessen and me. 
-When it comes to implementation of the ideas we had in these discussions, I have been responsible. 
-Push arrays, first appearing in paper D, were invented by Koen Claessen. However, the 
-implementation of them in the setting of Obsidian was done by me. Paper D also has large 
-contributions by Mary Sheeran when it comes to examples and implementing them.
-
-Paper E was the first I wrote with Josef Svenningsson, who had been thinking about an interesting
-variation of counting sort and wanted to implement it. Together we added atomic operations to 
-Obsidian and I modified the code generator to accommodate them. Sections 
-1 and 2 in the paper are written by Josef. Sections 3 through 7 are mostly my work, including 
-the benchmarking. Section 8 is joint work.
-%Me and Josef are equal partners in this paper and got much valuable advice from 
-%Mary Sheeran.
-
-\subsubsection{Papers G and H} 
-The work behind paper G, work was performed by me 
-under supervision of Ryan R. Newton who is also mostly responsible for the writing of the 
-paper. Programming and running time experimentation was performed by me, while getting much 
-valuable advice and guidance from Ryan.
-
-The work behind paper H was done by me as a hobby project and my aspirations were not 
-high. I just wanted to experiment with some embedded language techniques that we had 
-been avoiding (had no need for) in Obsidian. Mary discovered what I was doing and applied 
-the right amount of pressure, leading to us encompassing much more of ArBB functionality than 
-I had in mind. The paper's text is joint work; most examples as well as benchmarking are my work,
-but the sparse matrix vector multiplication example is entirely Mary's.
-
-%Intel has unfortunately retired the ArBB project. This leaves EmbArBB without a functioning 
-%code generating backend u
-
-% \subsection*{Guide}
-
-\tableofcontents
-
-\cleardoublepage
-\clearpage
-
-\pagenumbering{arabic}
-\pagestyle{fancy}
-\fancyfoot{}
-\fancyhead[LO]{}
-\fancyhead[RO]{\leftmark}
-\renewcommand{\headrulewidth}{0.0pt}
-\fancyhead[LE,RO]{\thepage}
-
-% --------------------------------------------------------------------------- %
-% --------------------------------------------------------------------------- %
-%
-%  INTRODUCTION 
-%
-% --------------------------------------------------------------------------- %
-% --------------------------------------------------------------------------- %
-\chapter{\thesistitle}
-
-% -------------------------------------------------------------------------
-%  INTRO SECTION OF INTRO CHAPTER
-% -------------------------------------------------------------------------
-
-\section{Introduction}
-
-This thesis applies embedded language methodology to data-parallel programming of 
-massively parallel hardware. An embedded language is a language that is implemented 
-as a library for some existing host language. In this case, the functional programming 
-language Haskell is used as host language.
-
-Computers today are increasingly parallel. Each new generation of general purpose processors
-(CPUs) and of graphics processors (GPUs) brings more parallelism. The question 
-of how to program these computers most efficiently, both from a productivity and 
-a performance point of view, arises. Explicit parallel programming is hard and automatic 
-means of parallelisation, while giving a sequential programming model to the programmer, 
-often misses the target when it comes to performance. The book ``Structured Parallel 
-Programming'' by Michael McCool et al argues the need for explicit parallel programming 
-models~\citet{STRUCTURED}.
-
-The embedded language approach to programming language implementation is beneficial, 
-since we are not entirely sure of how to program current and future highly parallel 
-computers. Implementing an embedded language allows us more rapid prototyping of ideas. 
-We can experiment with what abstractions to expose to the programmer, without much 
-effort. There are many other benefits of embedded languages: making use 
-of the host language's library functions (if they are general enough), the host language 
-can be used as a powerful macro language for the embedded language, providing a more 
-leveled learning curve to programmers already familiar with the host language. 
-For programmers who are already proficient in the host language, learning an embedded 
-language is more like picking up a new library.
-
-This thesis focuses on data-parallelism, as opposed to task or control parallelism.
-Data-parallelism is applicable where the input data can be divided into chunks
-and operated upon independently. Data-parallelism is scalable; the potential 
-for parallelism increases with the problem size. Two different embedded languages 
-for data-parallel programming are presented in this thesis: Obsidian, for 
-general purpose computations on GPUs and EmbArBB for parallel programming across 
-platforms, such as CPUs, GPUs and other accelerators.
-
-Graphics Processors trade programmer convenience for increased parallelism. 
-For example, where CPUs dedicate large portions of chip area to caches, branch 
-prediction and instruction level parallelism (ILP), GPUs put additional cores. 
-This comes at a price of increased programmer effort. GPUs have programmer managed 
-shared memories, conditionals are troublesome from a performance 
-perspective and memory access patterns can make or break the performance of an 
-application. With Obsidian, we try to find the right level of abstraction for 
-programming GPUs. We try to strike a balance between low-level control of details 
-that influence performance and high-level abstractions to increase productivity and 
-make programs more terse. The end goal with Obsidian is a language for writing 
-high performance GPU programs that compare well to code written by CUDA~\citet{CUDA}  
-(a C-dialect for GPU programming) experts, provided that both programmers have 
-similar GPU expertise. The Obsidian programmer would however write programs in a more 
-composable manner. Obsidian programs are compiled into CUDA.
-
-Obsidian is geared towards the programmer who is interested in GPU hardware. Making 
-good use of Obsidian requires that the programmer is willing to make an effort 
-to learn about GPUs and how to program them efficiently. However, another case 
-is a programmer who does not want to invest that effort but wants to write a high-level 
-parallel program and have it run reasonably well on different parallel architectures. 
-In this case, a language with even higher level of abstraction is preferable. It then 
-becomes the job of a more advanced compiler to turn that into code for the targeted 
-platforms. The EmbArBB language targets this programmer. EmbArBB exposes a set of high-level 
-operations on arrays, such as {\tt map}, {\tt reduce} and {\tt scan}. The programmer 
-has no power over the parallel strategy that should be used to implement these 
-operations but instead relies on a JIT-compiler (Just In Time-compiler). EmbArBB is a 
-Haskell embedding of Intel ArBB~\citet{ARBB2011} that was released as an embedded language 
-in C++. EmbArBB relies entirely on the ArBB system for code generation and optimisation.
-
-% -------------------------------------------------------------------------
-% Obsidian !
-% -------------------------------------------------------------------------
-
-\section {Obsidian: An embedded language for GPU kernel implementation}
-
-In 2006, NVIDIA released their first Compute Unified Device Architecture (CUDA) 
-GPU. With CUDA, NVIDIA also made it a lot easier for programmers to use GPUs
-for general purpose programming~\citet{wwwcuda}. A dialect of C, called CUDA C, was released 
-specifically to simplify the implementation of general purpose computations on GPUs.
-
-Before CUDA, programmers still tried (very ingeniously) to make use of GPUs for 
-non-graphics applications via the graphics API. The desired computation was expressed 
-in graphics terms and the result computed through an act of rendering. Of course, since the 
-available abstractions targeted the graphics domain, this added extra
-burden on the programmer. CUDA was NVIDIA's first response to this problem. CUDA 
-is a dialect of C, extended with a parallel programming model suitable for GPUs.
-
-Having a language specifically made for general purpose programming on the GPU, such as 
-CUDA, is a big step up from earlier. But CUDA still offers a very low-level interface to 
-programming the GPU, which is preferred in a situation where the programmer needs to 
-think about hardware specific details to get the best performance. CUDA provides this, 
-but at the price that it is hard to build upon, abstract and reuse prior programming 
-effort. This is where we are trying to go one step further with Obsidian.
-
-% -------------------------------------------------------------------------
-% GPU Hardware 
-% -------------------------------------------------------------------------
-
-\subsection{GPU hardware}
-
-Before going into programming of GPUs, it is important to have some background. The 
-GPU programming model exposed by CUDA very much mirrors the underlying hardware. Some 
-of the details that make GPU programming hard are more apparent when looking at the 
-underlying hardware.
-
-A CUDA GPU is built around a single kind of processor (as opposed to the different 
-kinds of processors found in earlier GPUs). The processors in the GPU (called MPs, 
-MultiProcessors) all contain a number of cores called SPs (Streaming Processors). 
-Each MP also contains local memory, called shared memory since it can be accessed 
-by all of the SPs in that MP. The number of MPs varies over the available GPUs; cheaper 
-GPUs have as few as one MP, and as you go up in price the number of MPs increases.
-
-Each MP of the GPU can manage a large number of threads; on today's GPUs up to 
-2048 threads can run on a single MP. The GPU schedules threads in groups of 32, called 
-{\em Warps}, that are executed in lock-step (SIMD style execution). Threads are 
-also divided into {\em Blocks}; the threads within a block can communicate using the 
-shared memory. The maximum block size is currently 1024 threads. Threads within 
-a warp can communicate via the shared memory without using any synchronisation 
-primitive. However, if communication takes place across warps synchronisation is 
-necessary. A barrier synchronisation mechanism exists to ensure that all threads within a block 
-have reached the same position in the code. Blocks are also grouped into a grid that 
-is the collection of blocks executing the same program.
-
-% -------------------------------------------------------------------------
-%  GPU Programming
-% -------------------------------------------------------------------------
-
-\subsection{GPU programming} 
-\label{sec:GPUPROGRAMMING}
-
-The CUDA programming model reflects the hierarchical architecture of the GPU. The 
-programmer writes what are called \emph{kernels}; in CUDA jargon, a kernel refers to a sequential 
-C program that is parameterised on a thread's identity. The kernel program is executed 
-in parallel across all the threads of a block and many such blocks of threads can execute 
-in parallel on the available MPs. Kernels may use shared memory that is local to a block; 
-the programmer needs to decompose the algorithm and data in such a way that communication 
-will be local to the blocks.
-
-The program below is a very simple example of a CUDA kernel that computes elementwise sums.
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-__global__ void vecAdd(int32_t *i1, int32_t *i2, int32_t *r) {
-
-  unsigned int gid = blockIdx.x * 
-                     blockDim.x + 
-                     threadIdx.x;
-
-  r[gid] = i1[gid] + i2[gid];
-
-}
-
-\end{Verbatim}
 \end{small}
-
-\noindent The kernel takes two input arrays, {\tt i1} and {\tt i2}; the result is stored 
-in {\tt r}. Note how the kernel makes use of three variables, {\tt threadIdx.x}, 
-{\tt blockIdx.x} and {\tt gridDim.x}. These variables identify a thread uniquely within 
-a grid. Here they are used to obtain an index into the provided arrays. The dimensions of 
-the blocks and of the grid are set by a controlling program upon launch of the kernel. This 
-controlling program is run by the CPU of the computer housing the GPU. The kernel is launched 
-using the following syntax:
-
-\begin{small} 
-\begin{Verbatim}[samepage=true]
-vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
-\end{Verbatim}
-\end{small}
-
-\noindent This line of code launches the kernel on one block of 32 threads. The zero 
-indicates that the kernel uses no shared memory. If the kernel did use shared memory, the 
-amount needed should have been specified there. Figure~\ref{fig:CUDACODE1} shows the complete 
-sequence of commands for a minimal example using the kernel specified above.
-
-% -------------------------------------------------------------------------
-% CUDACODE1 FIGURE
-% -------------------------------------------------------------------------
-
-\begin{figure} 
-\begin{small}
-\begin{Verbatim}[samepage=true]
-int main(void) {
-
-  int32_t h_i1[32], h_i2[32], h_r[32];
-
-  int32_t *d_i1, *d_i2, *d_r;
-
-  for (int i = 0; i < 32; ++i) { 
-    h_i1[i] = i;
-    h_i2[i] = 32 - i;
-  }
-
-  // Allocate device memory
-  cudaMalloc((void**)&d_i1,32*sizeof(int32_t));
-  cudaMalloc((void**)&d_i2,32*sizeof(int32_t));
-  cudaMalloc((void**)&d_r,32*sizeof(int32_t));
-
-  // Copy inputs to device
-  cudaMemcpy(d_i1,
-             h_i1,
-             32*sizeof(int32_t),
-             cudaMemcpyHostToDevice);
-  cudaMemcpy(d_i2,
-             h_i2,
-             32*sizeof(int32_t),
-             cudaMemcpyHostToDevice);
-
-  // Launch the kernel
-  vecAdd<<<1,32,0>>>(d_i1,d_i2,d_r);
-
-  // Copy result to host
-  cudaMemcpy(h_r,
-             d_r,
-             32*sizeof(int32_t),
-             cudaMemcpyDeviceToHost);
-
-  for (int i = 0; i < 32; ++i) { 
-    printf("%d ",h_r[i]);
-  }
-
-  return 0;
-}
-\end{Verbatim} 
-\end{small}
-
-\caption{Host code for launching a simple CUDA kernel on the GPU.}
-\label{fig:CUDACODE1}
-
-\end{figure}
-
-% -------------------------------------------------------------------------
-% 
-% -------------------------------------------------------------------------
-
-\noindent In Obsidian, the same kernel (the local computation) is implemented as follows:
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-vecAdd :: Num a => SPull a -> SPull a -> SPull a
-vecAdd = zipWith (+)
-\end{Verbatim}
-\end{small}
-
-\noindent The {\tt vecAdd} program takes two {\tt SPull} input arrays and gives one 
-as result. The Obsidian {\tt vecAdd} is more general than the CUDA version. From it 
-we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
-In Haskell we know that types that belong to the {\tt Num} typeclass has a {\tt (+)} operation.
-
-There are some important differences between CUDA and Obsidian programs. One is that with Obsidian 
-we generate kernels for fixed size local computations. This is to avoid doing out-of-bounds checks 
-which would mean conditionals in the code. Conditionals are problematic from a performance 
-point of view. Another difference is that in Obsidian the programmer specifies the entire 
-computation, including how it is mapped over the blocks of the GPU. This is done implicitly 
-in CUDA, based on the indexing arithmetic used in the kernel and the number of blocks and 
-threads provided at kernel launch. This means that the Obsidian programmer needs to 
-write a program that performs this mapping:
-
-\begin{small} 
-\begin{Verbatim}[samepage=true]
-vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
-vecAddG i1 i2  =
-  pConcat $
-  fmap push $ 
-  zipWith vecAdd
-          (splitUp 32 i1)
-          (splitUp 32 i2)  
-\end{Verbatim}
-\end{small}
-
-The {\tt vecAddG} program specifies a computation on large arrays; the large input arrays 
-are split up into pieces that are operated upon by the local computation, {\tt vecAdd}, 
-specified above. Each input array passed to this function must have length a multiple of 32.
-Since the {\tt splitUp} operation splits the arrays into chunks of 32 elements, the body 
-of the generated kernel will be constructed specifically for this size. {\tt zipWith} is 
-used to map the local computation onto each chunk of the input arrays. The result is an 
-array of arrays that is flattened with the {\tt pConcat} function. The CUDA code below 
-is generated from the {\tt vecAddG} program:
-
-\begin{small}
-\begin{Verbatim}[samepage=true] 
-extern "C" __global__ void vecAdd(int32_t* input0, uint32_t n0,
-                                  int32_t* input1, uint32_t n1,
-                                  int32_t* output2)
-{
-    uint32_t tid = threadIdx.x;
-
-    if (blockIdx.x < min(n0 / 32U, n1 / 32U)) {
-        output2[blockIdx.x * 32U + tid] = 
-            input0[blockIdx.x * 32U + tid] +
-            input1[blockIdx.x * 32U + tid];
-    }
-}
-\end{Verbatim}
-\end{small}
-
-There are some superficial differences between the code generated using Obsidian and the 
-handwritten example above. One such difference is that the index computation is not shared. 
-Obsidian relies on the CUDA compiler to discover such sharing via a Common Subexpression 
-Elimination (CSE) optimisation pass. Another difference is that Obsidian has inserted a 
-conditional that only executes the body for certain blocks. Discussion of the causes for 
-that conditional is postponed until section~\ref{sec:OBSARRAYS}. A real 
-difference is that the kernel is generated for a fixed size; here there is a constant (32) 
-where in the handwritten code we used {\tt blockDim.x}. If those differences are set 
-aside, the two kernels are identical. Figure~\ref{fig:OBSIDIANHOST1} shows how the 
-{\tt vecAddG} kernel can be launched from within Haskell, using Obsidian's rudimentary 
-support for launching GPU computations.
-
-An alternative way to implement {\tt vecAddG} is to inline the local computation 
-directly. This makes the entire program look like this.
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-vecAddG :: Num a => DPull a -> DPull a -> DPush Grid a
-vecAddG i1 i2  =
-  pConcat $
-  fmap push $
-  zipWith (zipWith (+)) 
-          (splitUp 32 i1)
-          (splitUp 32 i2)
-\end{Verbatim} 
-\end{small}
-
-\noindent The generated code looks exactly as before.
-
-% -------------------------------------------------------------------------
-% OBSIDIANHOST1 FIGURE
-% -------------------------------------------------------------------------
-
-\begin{figure} 
-\begin{small}
-\begin{Verbatim}[samepage=true]
-performVecAdd =
-  withCUDA $
-  do
-    kern <- capture 32 (vecAddG :: DPull EInt32
-                                 -> DPull EInt32
-                                 -> DPush Grid EInt32)
-
-    useVector (V.fromList [0..32::Int32]) $ \i1 ->
-      useVector (V.fromList [0..32::Int32]) $ \i2 -> 
-        allocaVector 32 $ \(o :: CUDAVector Int32) ->
-        do
-          fill o 0
-          o <== (1,kern) <> i1 <> i2
-
-          r <- peekCUDAVector o
-          lift $ putStrLn $ show r
-\end{Verbatim} 
-\end{small}
-
-\caption{Host code for launching a simple Obsidian kernel on the GPU. \\ 
- {\tt capture} compiles an Obsidian program into CUDA. At this point the 
-  types of the Obsidian program need to be made concrete. The number, 32, used as 
- an argument to {\tt capture} specifies that generated code should be specialised 
- for 32 threads per block. }
-\label{fig:OBSIDIANHOST1}
-\end{figure}
-
-\pagebreak
-
-\noindent A CUDA kernel that uses shared memory takes the form:
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-__global__ void kernel(int32_t *i, int32_t *r) {
-
-  extern __shared__ int32_t sm[];
-
-  unsigned int tid = threadIdx.x; 
-  unsigned int gid = blockIdx.x * 
-                     blockDim.x + 
-                     threadIdx.x;
-
-  sm[tid] = i[gid]; 
-  __syncthreads();
-
-  /* Compute on sm */
-
-  __syncthreads();
-  r[gid] = sm[tid];
-
-}
-\end{Verbatim} 
-\end{small}
-
-\noindent A portion of the input array is stored into shared memory. A 
-barrier synchronisation is used to ensure that each thread within the block 
-has stored its value into shared memory (only necessary if threads will 
-communicate across warp boundaries). After computing on the data locally, 
-it is written back into global memory. This kernel sets up two aliases for 
-thread identities. A thread's global identity is called {\tt gid} and its local 
-identity is called {\tt tid}. The global ID is used to index into the large global 
-array; indexing into local shared memory is done using the local ID.
-
-In Obsidian, local computations can also use shared memory. The Obsidian arrays 
-we have seen in the examples, {\tt Pull} and {\tt Push} arrays do not directly 
-correspond to actual data in memory. Pull and push arrays rather represent ways 
-to compute arrays. More information about these array representations is available 
-in section~\ref{sec:OBSARRAYS}. The Obsidian programmer can store intermediate 
-arrays in shared memory by using a {\tt force} operation. 
-%The 
-%{\tt force} operation computes an array and stores it into shared memory.
-
-The following kernel multiplies each element in an array by two and then adds one:
-
-\begin{small}
-\begin{Verbatim}[samepage=true] 
-kernel :: Num a => SPull a -> SPull a
-kernel = fmap (+1) . fmap (*2)
-\end{Verbatim}
-\end{small}
-
-\noindent This kernel computes its result without any storage of intermediate results;
-the code is equivalent to \verb!fmap ((+1).(*2))!. 
-In the generated CUDA code below, it can be seen that the (*2) and (+1) operations 
-are applied without any intermediate storage.
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
-                                  int32_t* output1)
-{
-    uint32_t tid = threadIdx.x;
-
-    if (blockIdx.x < n0 / 32U) {
-        output1[blockIdx.x * 32U + tid] = 
-          input0[blockIdx.x * 32U + tid] * 2 + 1;
-    }
-}
-\end{Verbatim}
-\end{small}
-
-Storing an intermediate result between the two operations can be accomplished by 
-changing the kernel as follows.
-
-\begin{small}
-\begin{Verbatim}[samepage=true]
-kernelF :: (MemoryOps a, Num a)
-         => SPull a -> BProgram (SPull a)
-kernelF = liftM (fmap (+1)) . force . fmap (*2)
-\end{Verbatim} 
-\end{small}
-
-This change introduced some noise. The reason for this is that Obsidian programs that 
-use memory are monadic. The {\tt BProgram} in the result type is a monad. Another difference 
-is that we need to be able to write the array elements to memory; in Obsidian this means 
-that the element type must be in the {\tt MemoryOps} class.
-
-The code below is generated using the {\tt kernelF} program:
-
-\begin{small} 
-\begin{Verbatim}[samepage=true]
-extern "C" __global__ void kernel(int32_t* input0, uint32_t n0,
-                                  int32_t* output1)
-{
-    extern __shared__ uint8_t sbase[];
-    uint32_t tid = threadIdx.x;
-
-    if (blockIdx.x < n0 / 32U) {
-        ((int32_t*) sbase)[tid] = 
-           input0[blockIdx.x * 32U + tid] * 2;
-        __syncthreads();
-        output1[blockIdx.x * 32U + tid] = 
-           ((int32_t*) sbase)[tid] + 1;
-    }
-}
-\end{Verbatim}
-\end{small}
-
-\noindent This code uses a shared memory array called {\tt sbase}. There 
-is also a call to \\ {\tt \_\_syncthreads} to ensure that all the writes have 
-completed before proceeding. In this case the synchronisation is not really
-necessary since all threads belong to the same warp. There is an {\tt unsafeforce} 
-operation that naively removes synchronisations if the array being forced is short 
-enough. More advanced analysis could also be performed to remove synchronisations 
-whenever communication is entirely warp local. We do not apply that optimisation, 
-but work on warp level computations is in progress and would put this ability 
-in the hands of the programmer.
-
-\FloatBarrier 
-\subsection{Writing efficient GPU Kernels} 
-\label{sec:efficient}
-
-Writing efficient GPU kernels is hard. There are many details that the CUDA 
-programmer must be aware of in order to make optimal use of the GPU. 
-The NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
-to learn more about these details. Here is a list, with explanations of some 
-of the practices that the guide deems most important. The list also contains 
-information about how these best practices influence our work on Obsidian.
-
-\begin{itemize}
-
-\item {\bf Minimise data transfers between host and device:} The bandwidth between device memory 
-and GPU is much higher than the bandwidth between host memory and the GPU device via the PCIe bus.
-This means that, first of all, one must evaluate if the computation to be performed is significant 
-enough to warrant a transfer to the GPU. If the computation is not arithmetic intensive, it may be 
-better to do the work on the CPU. However, if the data is already in device memory and 
-the operation that is about to be performed on it is more efficiently handled by the CPU, it 
-may still be better to let the GPU do the work. In other words, compute near the data unless 
-transferring it pays off.
-
-Data transfers between the host memory and device (GPU) memory is out of Obsidian's scope. 
-If Obsidian is used to generate kernels for a computation, the programmer has already decided that 
-a GPU should be used. However, even after making the decision to use the GPU the programmer 
-may need to consider how to transfer the data. In some cases it is possible to overlap 
-computation on the GPU and data transfer. These kinds of optimisations are considered 
-when writing the C program that launches the CUDA kernels. Obsidian does have rudimentary 
-support for launching kernels from within Haskell but currently we consider this mostly a
-tool for quickly being able to run a kernel and see what it does, not as a means to implement 
-applications. This part of Obsidian is future work.
-
-\item {\bf Ensure that global memory accesses are coalesced:} Certain access patterns into 
-GPU device memory can be {\em coalesced} (combined into few memory transactions). Unfortunately 
-what these access patterns are varies somewhat depending on the GPU. For a typical model of 
-GPU, the rule is that concurrent memory accesses (by threads within a warp) will be coalesced 
-to as many transactions as cache lines touched (128 Byte l1 cache lines). This means that strided 
-accesses from within a warp are not favoured.
-
-Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
-the permutation used will decide the access pattern. In section~\ref{sec:OBSARRAYS}, there 
-is more information about pull arrays. If global memory accesses can 
-be coalesced or not partly also depends on the algorithm being implemented. The next paragraph 
-suggests a trick for coalescing memory accesses, in some cases, even if the algorithm is 
-unsuitable. Then it becomes a trade between bad memory access pattern and using more 
-shared memory.
-
-\item {\bf Minimise use of global memory:} Accessing global device memory can take as 
-many as 400 to 600 clock cycles if the data is not cached, this is 10 - 20 times as many cycles 
-as an access to shared memory. The programmer should use shared memory as much as possible 
-to avoid redundant loading from global memory. Shared memory can also be used to coalesce 
-reads from global memory. Shared memory is a limited resource in the MP, so if a kernel 
-uses very much shared memory, fewer such blocks can be active on the MP.
-
-Intermediate arrays are stored into shared memory when the Obsidian programmer uses 
-the {\tt force} function.
-
-\item {\bf Use a multiple of 32 threads per block:} Groups of 32 threads, warps, are the 
-scheduled unit on a GPU. If a warp accesses memory (and will be waiting for it arrive during 
-600 cycles) it will be swapped out and another warp will take its place. This means that 
-the number of available warps decides how well memory latency can be hidden. A block that is 
-not a multiple of 32 leads to wasting GPU resources; some processing elements will stand 
-idle during the execution of the incomplete warps.
-
-Obsidian requires regularity even more than CUDA does. Since we focus on generating high 
-performance kernels, we rule out generation of kernels that work on different sizes (per block). 
-The choice of how many threads to use per block is in the hands of the Obsidian programmer.
-The method for making that choice has changed over time. Earlier, the number of threads 
-needed by a kernel was decided by the size of the arrays it computed. This has recently changed
-and is now set by a parameter passed to the code generation process.
-
-\item {\bf Avoid different execution paths within a block:} Any branching instruction that 
-causes the execution to branch into different code (in CUDA jargon called a diverging branch) 
-within a warp will affect performance. When the execution paths within a warp diverge, the 
-computation is serialised.
-
-Sometimes divergent code cannot be avoided. However, it is important not to have threads 
-diverge unnecessarily. When using pull arrays alone, this was often a problem; with the 
-addition of push arrays, the situation improved. Using push arrays, some divergent computations 
-can be explicitly split up into phases, each phase with all threads following the 
-same execution path. This means that fewer threads are required to do the same work. Push arrays 
-are explained in more detail in section~\ref{sec:OBSARRAYS}.
-
-\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
-when synchronising within any conditionally executed code; every thread must reach this barrier.
-Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
-this is a major limitation to the composability of CUDA functions. Wheather or not it is safe to 
-call some function in a certain place in your code can be decided only by inspecting that 
-function's implementation.
-
-In Obsidian, it is impossible for a barrier synchronisation to appear within diverging 
-conditional code. Conditionals can be introduced either by an \\{\tt ifThenElse} function 
-at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
-monad; we have seen an example of the program monad in section~\ref{sec:GPUPROGRAMMING}. 
-The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
-thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs 
-({\tt GProgram}). the {\tt Cond} conditional chooses between two thread programs while a 
-synchronisation can only appear in an Obsidian block program. More information about programs 
-and the GPU hierarchy is available in section~\ref{sec:OBSIMP}.
-
-%f it is safe or 
-%not to call some function at a certain place in your code, can only be decided by inspecting 
-%that functions implementation. 
-\end{itemize}
-
-The list above provides some very general guidelines on what to think about when 
-writing GPU kernels. The lesson is that fast memory near the processing element 
-should be preferred over slow memory far away. Favouring fast memory demands the effort 
-of decomposing computation and data in a way that makes this possible.
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Obsidian Implementation 
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\subsection{Obsidian Implementation}
-\label{sec:OBSIMP}
-
-Obsidian is implemented as an embedded language using Haskell as host. Obsidian 
-is a compiled embedded language; a very good paper that is highly relevant to our 
-approach is ``Compiling Embedded Languages'' by Conal Elliott et al~\citet{COMPILEEDSL}. 
-This paper describes how to embed a compiler backend and generate efficient code 
-in some target language. One key aspect of this approach is that the embedded 
-language is a library of functions that create and compose Abstract Syntax Trees 
-(ASTs). Obsidian (chapter \ref{chap:GPUProgramming}) as well as EmbArBB 
-(chapter \ref{chap:ArBB}) are implemented in a similar way. An embedding that builds 
-ASTs is called a deep embedding. Shallow embeddings, on the other hand, do not build 
-ASTs. In ``Combining Deep and Shallow Embedding for EDSL'' Josef Svenningsson and 
-Emil Axelsson combine the two methods, shallow and deep, in order to simplify the 
-AST data type (fewer constructors) and make more extensible embedded languages 
-\citet{DEEPSHALLOW}. Obsidian also uses a combination of shallow 
-and deep embeddings.
-
-The embedded language approach has become popular to use as a way of raising the 
-level of abstraction in fields where the default is to use low-level languages. 
-For example, Lava and Wired~\citet{LAVA,Wired} raise the level of abstraction in the 
-fields of hardware design and verification and Feldspar does the same in the area of  
-digital signal processing~\citet{FELDSPAR2010}. For GPU programming, there are numerous 
-embedded approaches using various host languages. Two languages for GPU programming 
-embedded in Haskell are  Accelerate and Nikola \citet{ACCELERATEDAMP11, NIKOLA}. 
-And there is Intel ArBB, embedded in C++, with the goal of being parallel across 
-platforms \citet{ARBB2011}. Related work and information about the tools, libraries 
-and languages that make the wider context surrounding Obsidian can be found in 
-section~\ref{sec:relatedwork}.
-
-\subsubsection{Deep embedding: The {\tt Program} data type}
-
-Obsidian uses a deeply embedded {\tt Program} data type that represents GPU Kernels. 
-A CUDA GPU has a hierarchy of parallel resources. At the bottom there are threads,  
-executing sequential programs. There are groups of threads (of 32) called {\em Warps} that 
-execute in lock-step. Warps are the scheduled unit of work on a GPU. There are {\em Blocks} 
-of threads, a set of threads that run as a group and share local (shared) memory. And 
-lastly there is a {\em Grid} of blocks that specifies the total number of threads involved 
-in a computation (Number\_of\_Blocks * Number\_of\_Threads). The deeply embedded program 
-data type of Obsidian models this hierarchy by parameterising programs on a hierarchy 
-level type parameter (the  {\tt t} parameter below).
-
-\begin{verbatim} 
-data Program t a where
-\end{verbatim}
-
-The {\tt t} parameter can be either {\tt Thread}, {\tt Warp}, {\tt Block} or {\tt Grid}.
-These types are related to each other via a {\tt Step} type constructor that represents 
-going upward one step in the hierarchy.
-
-\begin{verbatim} 
-data Step a -- A step in the hierarchy
-data Zero
-\end{verbatim}
-
-The {\tt Thread} type is level {\tt Zero} ({\tt type Thread = Zero}). Then {\tt Block} is 
-{\tt Step Thread} and {\tt Grid} is {\tt Step Block}. Currently {\tt Warp} is not included 
-in the hierarchy, leading to warp programming being a special case. One benefit a warp has is 
-that threads can communicate via shared memory without using synchronisation primitives. There 
-is no programmer support for warp level programming in CUDA (as there is for block level 
-programs). The programmer needs to take care to ensure that the communication patterns are 
-within a warp and is then free to remove synchronisations. If warps were to be put in the 
-program level hierarchy of Obsidian, its place would be between the thread and block level. 
-This would force the programmer to go via warps when putting together a block computation even 
-if the communication pattern is not suited for that, leading to inconvenience with no gain.
-
-As an example of how the hierarchy level parameter is used, I show the {\tt ForAll} constructor 
-from the {\tt Program} data type.  This constructor represents parallelism either over threads 
-or blocks.
-
-\begin{verbatim}
-ForAll :: EWord32 
-            -> (EWord32 -> Program t ())
-            -> Program (Step t) ()
-\end{verbatim}
-
-{\tt ForAll} takes a number of parallel iterations, and a body represented by a function 
-from an index to a program. The body is a {\tt Program} at some level {\tt t} while the 
-resulting program is a step above. This means that a thread program can be turned into 
-a block program by using {\tt ForAll}, or that a block program can be turned into a grid program.
-
-Information about the {\tt Program} data type can be found in paper F, \paperFTitle, in 
-section \ref{sec:paperF}.
-
-\subsubsection{Scalars}
-
-Scalars and operations on scalars are represented by an expression data type \\ ({\tt Exp a}) in 
-Obsidian. This is another example of a deep embedding; the expression data type contains 
-constructors for literal values, arithmetic operations, and conditionals.
-
-Haskell's type class system allows us to overload arithmetic operations such as {\tt (+)}, 
-{\tt (-)} and {\tt (*)} by making expressions an instance of {\tt Num}. This allows the 
-Obsidian arithmetic expressions to look exactly like corresponding native Haskell arithmetic.
-
-The approach taken to embed the scalar language is very similar to what is described 
-in ``Compiling Embedded Languages''~\citet{COMPILEEDSL}; but we use a Generalised Algebraic 
-Datatype (GADT) to obtain typed expressions. An alternative to using a GADT is to use phantom 
-types, which earlier versions of Obsidian did (section~\ref{sec:paperB}).
-
-\subsubsection{Arrays} 
-\label{sec:OBSARRAYS}
-
-Obsidian has two array representations. These array representations are implemented as 
-shallow embeddings on top of the expression and program data types. This means that these 
-arrays will disappear during Haskell evaluation of the Obsidian program.
-
-\paragraph{Pull arrays:}
-
-Pull arrays have been part of Obsidian from the very beginning~\citet{JMT} but called either 
-{\tt Array} or just {\tt Arr}. A pull array represents an array as an indexing function; given 
-an index it provides an element:
-
-\begin{verbatim} 
-data Pull s a = Pull {pullLen :: s, 
-                      pullFun :: EWord32 -> a}
-\end{verbatim}
-
-Pull arrays are parameterised on both element ({\tt a}) and length ({\tt s}) type. The length 
-parameter is used to be able to represent arrays with static (known at Haskell runtime) 
-or dynamic (length represented by an expression, an unknown length). When creating block 
-level computations, we want to know the exact lengths in order to avoid generating code 
-riddled with conditionals concerning array lengths. In CUDA, it is possible to write kernels that 
-are not specialised for a certain size. This is accomplished by adding bounds checks. Since 
-conditionals  are problematic from a performance point of view, fixed size kernels are recommended 
-when performance is crucial. In Obsidian, we have chosen to disallow implementation of kernels 
-that operate on different sizes locally.  Dynamic lengths are used for grid level computations; 
-this allows a local computation of a fixed size to be run in parallel over arrays a multiple
-of the size that the local computation handles. In generated code, this is currently visible 
-as a conditional on {\tt blockIdx.x}. However, this is still considered work in progress; the 
-conditional is only really necessary for Obsidian kernels that have more than one output array 
-and the outputs are of different length. 
-%Conditionals are problematic on GPUs and 
-%while general programs that run for any size is possible when performance is crucial a fixed 
-%size is recommended.
-
-One benefit of pull arrays is that they give fusion of all operations on them for free. 
-For example, {\tt map f} is implemented on pull arrays by composing {\tt f} with the indexing function:
-
-\begin{verbatim} 
-map f (Pull n ixf) = Pull n (f . ixf)
-\end{verbatim}
-
-Now, we see that \verb!map f . map g! is the same as \verb!map (f . g)! since both result in {\tt f}
-and {\tt g} becoming composed onto the indexing function. No intermediate array is built in 
-memory.
-
-Another benefit of pull arrays is that they are parallel. Simply put, early Obsidian was pull 
-arrays in addition to a way to compute the array and store its elements to memory 
-(called {\tt sync} in early versions of Obsidian and later {\tt force}). A very 
-direct way to compile a pull array to a GPU is to launch as many threads as there are 
-elements in the array and apply the indexing function to the thread id. Now each thread 
-will have computed an element of the array and can store that to memory.
-
-\paragraph{Push arrays:}
-
-Push arrays were added to Obsidian as a complement to pull arrays. Pull arrays are 
-good because they are so simple to understand and easy to parallelise. However, certain
-operations on pull arrays yield code that is not suitable for GPU execution. Concatenation 
-and interleaving of pull arrays results in conditionals in the indexing function. The 
-interleaving case, which is worst, leads to diverging branching in every warp. If branches 
-diverge, the GPU turns the processing elements taking one path off and 
-allow the others to progress; then turn to the other branch. The computation of the 
-branches is serialised and compute resources are wasted.
-
-Push arrays, like pull arrays, are functions. A Push array is a higher order 
-function, taking a function from value and index to a thread program and giving 
-as result a program at some level in the hierarchy. The function passed to the 
-push array function we call a write-function. Below is the definition of 
-push arrays that we use in Obsidian. Push arrays are parameterised on element 
-type ({\tt a}), size type ({\tt s}) and program hierarchy level ({\tt p}).
-
-\begin{verbatim}
-data Push p s a =
-  Push s ((a -> EWord32 -> TProgram ()) -> Program p ())
-\end{verbatim}
-
-When working with pull arrays, the consumer of such an array decides how to 
-iterate over the array and what elements to compute. This is done by applying the 
-push array function to those indices that are interesting to the consumer. Push 
-arrays work in the opposite way. The Push array (the result program) contains 
-the iteration schema; the consumer of the push only decides what write function 
-to supply. The iteration schema used is decided upon when creating a push array. 
-The code below is an example of how to create a push array from a pull array.
-
-\begin{verbatim} 
-convertToPush :: Pull Word32 e 
-               -> Push Block Word32 e 
-convertToPush (Pull n ixf) =
-    Push n $ \wf -> 
-      ForAll (fromIntegral n) $ \i -> wf (ixf i) i
-\end{verbatim}
-
-The function above takes a pull array and creates a push array with a parallel 
-iteration schema over threads within a block. There are other iteration schemas to 
-chose from such as sequential, combinations of sequential and parallel
-across both threads and blocks (two level parallelism).
-
-Push arrays give many of the same benefits as pull arrays. Using push 
-arrays, we also get fusion of operations as default. The implementation of map 
-(shown below) illustrates how mapping a function {\tt f} over a push array results 
-in an application of {\tt f} in the write-function.
-
-\begin{verbatim} 
-map f (Push s p) = Push s $ \wf -> p (\e ix -> wf (f e) ix)
-\end{verbatim}
-
-\subsubsection{Compilation to CUDA}
-
-When compiling Obsidian programs to CUDA, we only need to be concerned with the 
-{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype, 
-we know that the AST can be quite directly translated to CUDA. We know that there will 
-be at most one level of nestedness in the parallelism (several blocks, each of several threads). 
-This means that we do not need to do any transformations on the AST before generating CUDA code. 
-Paper F (section~\ref{sec:paperF}) contains technical details of the compilation process.
-
-\subsection{Results}
-\FloatBarrier
-
-We have implemented basic algorithms such as sorting, merging, fractal generation, reductions 
-and parallel prefix (scans). Often we have managed to get very good performance, close to that 
-of hand optimised CUDA code. However, there are also cases where we just cannot seem to reach 
-the level of performance that we would want. One such example is parallel prefix, an 
-application that has been haunting us from the very beginning. Scan is problematic to us because 
-it is best implemented as an in-place algorithm, while our array representations do 
-not allow for in-place update; there is no memory associated with them to update. 
-We are trying to improve this situation by adding a third array representation to Obsidian,
-mutable arrays.
-
-Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
-a reduction kernel in Obsidian. At the end, the implemented reduction kernel is  
-faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
-reduction in Accelerate corresponds to just one single hand optimised skeleton. We would have 
-been happy just getting close.
-
-In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrence sort 
-using Obsidian. We use very low-level features of Obsidian and end up producing 
-code that performs very well.
-
-Paper D (section~\ref{sec:paperD}) introduces push arrays and uses them in the 
-implementation of sorting networks, with quite good results. With push arrays we 
-were able to experiment with more detailed decompositions (what thread computes 
-what value) of the algorithms.
-
-\FloatBarrier 
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% ArBB work
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\section{Embedding Intel Array Building Blocks}
-
-Intel ArBB~\citet{ARBB2011} is a system for high-level data-parallel programming with the 
-ability to generate code for a variety of different hardware configurations. It is implemented 
-as an embedded language in C++. One motivation for ArBB that is mentioned in~\citet{ARBB2011} is 
-that while high-performance computing specialists are highly competent at implementing 
-kernels using low-level programming techniques, mainstream developers are not. ArBB offers 
-a more composable way to write programs that make use of core and vector (SIMD) parallelism and 
-doing this while using a familiar language, C++.
-
-One of the greatest strengths of ArBB, as I see it, is that it also comes with a low-level C 
-interface. The main purpose of this C interface is to make it easier to use the ArBB system 
-from other languages. Many languages have foreign function interfaces that are geared towards C.
-This gives ArBB a level of language independence on top of its cross platform abilities.
-
-ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
-\begin{itemize} 
-\item {\bf Reductions:}   {\tt add\_reduce}, {\tt mul\_reduce}, {\tt max\_reduce} ... 
-\item {\bf Scans:}        {\tt add\_scan}, {\tt mul\_scan}, {\tt max\_scan} ... 
-\item {\bf Sorting:}      {\tt sort} 
-\item {\bf Permutations:} {\tt gather}, {\tt scatter}, {\tt reverse} ...
-\item {\bf Sequential loops:} {\tt \_for}, {\tt \_while} 
-\item {\bf Map, zipWith, stencil:} are all supported via a single primitive called {\tt map}.
-\end{itemize}
-
-These operations (and many more, there are about 120 operations in total) can be used 
-by the programmer when writing programs using ArBB. Note that there are sequential looping 
-constructs with special names, {\tt \_for} and {\tt \_while}. ArBB is a deeply embedded 
-language and calling the ArBB functions does not actually execute them, it just adds a node to 
-some internal representation. Therefore, the use of normal C++ for loops leads to unrolled 
-ArBB programs and a special {\tt \_for} loop is needed to get a loop in the generated program. 
-This is exactly the way these things work in Haskell embedded languages, but it may be even 
-more surprising to a C++ programmer. Before an ArBB function can be computed, it needs to be 
-captured. For this, ArBB has a {\tt call} function that is used to run functions using ArBB 
-functionality. The first time {\tt call} is used on a function pointer, that function is 
-executed and the ArBB functions used in it build an AST (we say the function has been 
-captured). The AST is then compiled and specialised for the hardware available. Any 
-subsequent calls of a function using ArBB functionality do not lead to a recompilation; 
-as the first call caches this compiled code.
-
-The first step of our work with embedding Intel ArBB functionality was to implement very direct 
-and low-level bindings to the C interface. This provides a Haskell function (in the IO Monad) 
-for each of the functions in the ArBB C interface. More details and results of this work 
-can be found in paper G in section~\ref{sec:paperG}.
-
-%\noindent\emph{What does ArBB do, what are the operations and datastructures it support}
-
-%\noindent\emph{What does EmbArBB do}
-
-\subsection{EmbArBB: Embedding ArBB in Haskell}
-\label{sec:EmbArBB}
-
-The ArBB Haskell bindings provide a very rudimentary interface to ArBB functionality and 
-are not useful for application implementation. We need a higher-level interface to ArBB to offer 
-to the Haskell programmer. Our first attempt was to implement an Accelerate backend 
-using the ArBB bindings. However, this had some hard problems to solve. Accelerate is a 
-richer language and there was an API mismatch between Accelerate and ArBB. These problems 
-are mentioned in paper G, section~\ref{sec:paperG}. As a way to offer ArBB functionality 
-to the Haskell programmer in a way that is useful and simpler to implement, we started 
-working on EmbArBB.
-
-EmbArBB improves the interfacing with Haskell by implementing a deeply embedded language. 
-Now, programs in EmbArBB create ASTs. These ASTs are compiled via the ArBB bindings, creating 
-function objects that can then be run.
-
-\subsubsection{Dense and nested arrays}
-
-ArBB supports one, two and three dimensional arrays, called dense vectors. In \\ EmbArBB, we 
-represent these with a data type, {\tt DVector}, parameterised by dimensionality and 
-element type. The dimension parameter can be either {\tt Dim1}, {\tt Dim2} or {\tt Dim3}. 
-There are also nested vectors, in EmbArBB called {\tt NVector}, parameterised on element type.
-
-\subsubsection{Operations on arrays}
-
-EmbArBB provides many of the ArBB operations on vectors via a 
-shape-polymorphic interface. For example, a reduction operation can take a two  
-dimensional vector and reduce all rows or columns and provide a one dimensional result. 
-It can also take a one dimensional vector and return a single value (zero dimensional vector).
-
-\begin{verbatim} 
-addReduce :: Num a 
-           => Exp USize 
-           -> Exp (DVector (t:.Int) a) 
-           -> Exp (DVector t a) 
-\end{verbatim}
-
-The {\tt addReduce} function takes an input vector that is at least one dimensional 
-({\tt t:.Int}) and returns a vector that has one less dimension ({\tt t}).
-
-\subsubsection{Deep embedding}
-
-EmbArBB is a traditional deep embedding; there is an AST data type with constructors 
-representing each ArBB language feature. It is true that a deep embedding of ArBB 
-leads to duplication of effort. The low-level operations exported by the 
-ArBB C interface already construct an AST within the ArBB system. One possible benefit 
-of the deep embedding is that it buys independence from the ArBB backend. It is possible 
-to compile the AST generated on the Haskell side to some other target language.
-
-When compiling EmbArBB to ArBB, a sharing detection pass is performed. This pass detects 
-sharing in the AST using the method of A. Gill~\citet{Gill}. One benefit of this pass 
-is that it reduces the number of calls into the ArBB C library. The efficiency of code 
-generated is probably not affected, since ArBB would discover that sharing in a Common 
-Subexpression Elimination (CSE) pass. Another potential benefit is that sharing detection 
-on the Haskell side results in a simpler AST being built on the ArBB side, potentially making 
-the job for the ArBB optimisation passes simpler. What effects these 
-operations truly had on performance has not been investigated. Actually, applying a sharing 
-detection technique was one of my personal motivations for working on EmbArBB. In Obsidian, 
-we have not seen the need for this technique. In Obsidian, sharing is in the hands of the 
-programmer, via the {\tt force} operation. On a GPU, communication is more expensive than 
-computation, so whether to share a computation or not becomes a tool for the programmer to use.
-
-One benefit of a deep embedding is that transformations can be performed on the AST. Applying 
-any optimisations in EmbArBB would very likely duplicate effort already put into ArBB. 
-Therefore, we apply no further transformations after sharing detection. However, if EmbArBB 
-would be used to generate code in some other target language a set of optimisations 
-would be needed.
-
-\subsubsection{Evaluation}
-
-In paper H, section~\ref{sec:paperH}, we benchmark EmbArBB against ArBB in C++ and the Haskell 
-Repa library. EmbArBB compares favourably to Repa and is often a lot faster. The ArBB 
-C++ performance is not distinguishable from EmbArBB's, indicating that our Haskell embedding 
-adds no extra overhead. Of course, this is expected, since once the ArBB system has generated 
-the code, that code is identical no matter if the AST was built via Haskell or C++ calls. 
-A comparison of how well EmbArBB and C++ compares in regards to the time it actually 
-takes to go through the process of making those ArBB C API calls has not been performed. 
-However, if the generated code is used many times, ArBB's caching of generated code means 
-that the initial code generation cost can be amortised.
-
-\subsection{Current status of ArBB and EmbArBB}
-
-Since our work with ArBB, Intel has unfortunately retired the ArBB system. This 
-leaves EmbArBB without a functioning code generating backend. This makes the choice of a
-deep embedding (in hindsight) the correct one. EmbArBB could be resurrected by implementing 
-a new code generating backend.
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Related work
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\section{Related work} 
-\label{sec:relatedwork}
-\FloatBarrier
-
-\subsection{Languages for data-parallel programming}
-
-In figure~\ref{fig:researchmatrix}, our work is placed in relation to other languages,
-embedded languages and libraries for parallel programming. The languages and libraries considered 
-are data-parallel; languages for control parallelism or distributed message passing systems are 
-not considered. The systems are roughly divided 
-into three groups, based on their level of abstraction. There are low-level languages; here 
-I place languages that are imperative and C-like. In the middle, layer I place languages 
-that have higher level abstractions and are more easily composable. 
-at the highest level, I place languages that completely abstract away from details of the 
-hardware they run on. The division is not free from personal bias and the placement of languages
-in boxes was not always easy.
-
-The figure also divides the languages according to what hardware they support. There 
-are CPU specific languages, GPU specific languages and languages that support or aspire 
-to support CPU,GPU and accelerator (Larrabee, Xeon Phi) execution.
-
-% ---------------------------------------------------------------------------
-% Obsidian, Accelerate, Nikola, Thrust, CUB 
-% ---------------------------------------------------------------------------
-With Obsidian, we try to target the sparsely occupied area of mid-level GPU programming. With 
-CUDA~\citet{wwwcuda} and OpenCL~\citet{OpenCL}, the programmer has full control of how to divide 
-the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
-Nikola~\citet{NIKOLA} 
-and Thrust~\citet{THRUST}, the programmer uses 
-high-level patterns, without direct insight into how the application is mapped onto 
-the threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
-more composable than CUDA and giving the programmer control of what the computation will 
-look like on the GPU. This is done by allowing the programmer to specify and compose 
-thread-level and block-level code. Recently, the CUB library arrived that is built on a 
-similar idea, that GPU programmers need to have block/thread level control to get maximum 
-performance, while maintaining a higher level and more Thrust-like programmer 
-interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
-this level.
-
-\begin{figure}
-\begin{center}
-\begin{tikzpicture}[align=center, scale=2.5]
-% \draw[help lines] (0,0) grid (3,3);
-
-\draw[->] (-0.5,0) -- (-0.5,3) node[anchor=east] {Abstraction level};
-
-\draw (0,3) node[anchor=south west] {CPUs};
-\draw (1,3) node[anchor=south west] {GPUs};
-%\draw (2,3) node[anchor=south west] {Accelerator};
-\draw (2,3) node[anchor=south west] {CPUs, GPUs\\ Accelerators};
-
-\node[anchor=west] (acc) at (1,2.7) {Accelerate\\Nikola\\Thrust};
-\node[anchor=west] (acc) at (1,1.7) {CUB};
-\node[anchor=west] (repa) at (0,2.5) {DPH\\Nesl\\Repa\\Meta-repa\\Feldspar};
-\node[anchor=west] (mixed) at (0,1.7) {Cilk+\\ TBB};
-\node[anchor=west] (accelerator) at (2,1.7) {Accelerator};
-
-\node[anchor=west] (cuda) at (1,0.3) {CUDA};
-\node[anchor=west] (opencl) at (2,0.3) {OpenCL};
-
-\node[anchor=west] (OpenMP) at (0,0.3) {OpenMP\\PThreads};
-
-\foreach \x in {0,...,2} { 
-  \foreach \y in {0,...,2} { 
-    \draw (\x,\y) rectangle (\x+1,\y+1);
-  }
-}
-
-\draw[red!75,rounded corners,ultra thick] (1,0.4) rectangle (2,2.4);
-\node[red!75,rotate=45,thick] (obs) at (1.5,1.5) {Obsidian};
-
-\draw[orange!90,rounded corners,ultra thick] (2,1.4) rectangle (3,2.6);
-\node[orange!90,thick] (arbb) at (2.5,2) {EmbArBB\\ArBB};
-
-\end{tikzpicture}
-\end{center}
-\caption{Placing our work in the landscape of languages and libraries for parallel 
-general purpose programming of CPUs, GPUs or both. }
-\label{fig:researchmatrix} 
-\end{figure}
-
-% ---------------------------------------------------------------------------
-% ArBB EmbArBB
-% ---------------------------------------------------------------------------
-The ArBB and EmbArBB systems also provide a set of parallel primitives that the programmer
-composes to build his application. I place ArBB and EmbArBB in a box slightly lower than 
-Accelerate,  since while ArBB/EmbArBB also have built in reduction primitives, just like 
-Accelerate, 
-they are not general primitives. Where Accelerate has a single higher order reduction 
-operation, ArBB/EmbArBB has \\ {\tt reduce\_add, reduce\_mul} and so on. Benefits of 
-having ArBB embedded in Haskell compared to C++ are discussed in section~\ref{sec:EmbArBB}.
-%I will go into 
-%what benefits there are to having ArBB embedded in Haskell compared to C++ in
-
-% ---------------------------------------------------------------------------
-% Microsoft Accelerator
-% ---------------------------------------------------------------------------
-Microsoft Accelerator~\citet{ACCELERATOR} is a language embedded in C\#. Accelerator has
-data-parallel arrays and a set of aggregate operations (operations on entire arrays). The 
-operations on data-parallel arrays are JIT-compiled to GPU code. Accelerator can also 
-compile to multithreaded CPU code and make use of SIMD units of modern 
-CPUs~\citet{ACCELERATORCPU}; this makes the Accelerator system related to ArBB.
-
-% ---------------------------------------------------------------------------
-% Python systems
-% ---------------------------------------------------------------------------
-There are many embedded languages implemented in Python that make use of GPUs 
-to improve performance. One example is Copperhead~\citet{copperhead}. Copperhead 
-supports flat and nested data-parallelism via primitives such as {\tt map} and 
-{\tt reduce}.
-
-% ---------------------------------------------------------------------------
-% Repa, Meta-Repa 
-% ---------------------------------------------------------------------------
-
-Repa (Regular Shape-polymorphic Arrays)~\citet{REPA} is a Haskell library for 
-data-parallel programming. 
-Repa uses the same array representation as Obsidian, in Repa called a delayed
-array, which is a function from index to element. However, in Repa arrays have shape; 
-they can be of any dimensionality and there are shape-polymorphic functions on these. 
-For example, {\tt map} can be applied to an array of any shape.
-
-Meta-Repa~\citet{METAREPA} is a reimplementation of the Repa library using deep embedded language 
-techniques. Meta-Repa programs build ASTs that are then compiled (using template 
-Haskell) to Haskell code that uses low-level parts of the Repa library for its parallelism.
-The Meta-Repa approach to implementing the Repa library 
-gives inlining of operations for free, where Repa programs need to be annotated 
-with pragmas to ensure inlining. On the other hand, in Meta-Repa the programmer 
-needs to explicitly state when something should not be inlined (using a {\tt force} primitive).
-This is related to our work on EmbArBB that also uses an existing library (ArBB) 
-for its parallelism. In reference~\citet{FPCDSL}, a case is made for this style 
-of embedding preexisting DSLs.
-
-% ---------------------------------------------------------------------------
-% DPH, NESL 
-% ---------------------------------------------------------------------------
-Data-Parallel Haskell~\citet{DPH} and Nesl~\citet{NESL} are languages for 
-nested data-parallelism. Both run on CPUs, but there is also work towards 
-a GPU version of Nesl~\citet{NestedGPU}.
-
-% ---------------------------------------------------------------------------
-% Delite, Vertigo 
-% ---------------------------------------------------------------------------
-
-Delite~\citet{DELITE} is an infrastructure for implementation of domain 
-specific embedded languages. There is an extensible internal representation (IR,AST); 
-the programmer extends some base IR with domain specific constructs. The Delite system 
-has been used to implement domain specific embedded languages for machine learning, rendering, 
-physics and other domains. Delite, being a framework and the languages implemented using it 
-very domains specific, was impossible to place in the matrix (Figure~\ref{fig:researchmatrix}).
-
-\subsection{Embedded domain specific languages}
-
-Feldspar is an embedded language for digital signal processing~\citet{FELDSPAR}. Apart 
-from using the syntactic library~\citet{SyntacticICFP12}, it is quite similar to Obsidian. 
-There is a deeply embedded core language (called Core) and vectors as a shallow embedding 
-on top of that, just like in Obsidian.
-
-When we first started working on Obsidian, we used Lava~\citet{LavaICFP} as inspiration. 
-We wanted to mimic the Lava programming style but generate GPU code. At first this worked quite 
-well and most Obsidian programs superficially looked quite like Lava programs. We have diverged 
-from this aspiration since realising that GPUs are quite a different story and performance 
-requires more control of GPU architecture specifics.
-
-\subsection{Compiling Embedded Languages}
-
-In paper A~(section \ref{sec:paperA}), we describe our method for compiling monadic embedded 
-languages. This work is closely related to a method described in ``The Constrained Monad Problem''
-by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In that paper, they solve a more 
-general problem than just compilation, but our method is nice because it is easy to understand 
-and solves a single problem and does it well. In ``Generic monad constructs'', Persson et al 
-also solve the problem of compiling monads using a continuation monad.
-
-``Compiling Embedded Languages'' by Elliott and Finne has been a constant source of information 
-and something we have referred to in almost all our work. This is, in my opinion, the paper to turn 
-to for anyone starting out with compiled embedded languages.
-
-%\subsection{Parallelism}
-
-%\noindent\emph{DPH}\citet{DPH} \newline
-%\noindent\emph{D.A.Accelerate}\citet{ACCELERATEDAMP11} \newline
-%\noindent\emph{Nikola}\citet{NIKOLA}\newline
-%\noindent\emph{Repa}\citet{REPA}\newline
-%\noindent\emph{Meta-Repa}\citet{METAREPA}\newline
-%\noindent\emph{ArBB}\citet{ARBB2011}\newline
-%\noindent\emph{OpenCL}\citet{OpenCL}\newline
-%\noindent\emph{CUDA}\citet{CUDA}\newline
-%\noindent\emph{Microsoft Accelerator}\citet{ACCELERATOR}\newline
-%\noindent\emph{Copperhead}\citet{copperhead}\newline
-%\noindent\emph{Delite}\citet{DELITE}\newline
-%\noindent\emph{A monad for deterministic parallelism (R. Newton)} \citet{MonadPar} \newline
-%\noindent\emph{Nesl}\citet{NESL} \newline
-
-%\subsection{Embedded domain specific languages}
-
-\FloatBarrier
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Future work
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\section{Future work}
-
-Obsidian is a language for GPU kernel implementation, not complete applications. 
-It is not as well integrated into Haskell as for example Accelerate. There are 
-ways to execute Obsidian kernels from within Haskell (we have seen an example of this in 
-section~\ref{sec:GPUPROGRAMMING}); the recommended work flow, 
-however, is to generate the CUDA as text and write the host code by hand. This 
-can be seen as both a shortcoming and a feature. It is hard to integrate 
-Accelerate programs in larger applications written in other languages; when 
-using Obsidian a kernel is generated and can be used as part of for example a library.
-
-Obsidian is also quite a low-level language. The programmer needs to know GPU details 
-and is allowed to write programs at a level almost as low as CUDA, if it is desired. 
-Paper E (section~\ref{sec:paperE}) contains examples of programming at this low level. 
-One potential use of Obsidian could be as the code generating backend of high-level 
-and very domain specific embedded languages. This could be a niche for Obsidian.
-
-Obsidian's support for warp level computations is work in progress. Currently, warp level 
-computations are not as well integrated into the hierarchy as the block, thread and grid 
-levels are. The warp abstractions 
-offered also require more work in the code generating backend than those 
-that correspond directly to CUDA concepts. The problem is that CUDA does not have 
-a warp abstraction. Warps needs to be better integrated with the rest of Obsidian 
-and the code generation required for warps needs more testing before being 
-trustworthy and useful. When operational, warp level abstractions will help 
-the programmer make better use of the GPU.
-
-%\noindent\emph{Improve on Warp level capabilities}
-
-%\noindent\emph{Improve expressivity, code performance}
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Discussion
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-
-\section{Discussion}
-
-%\vspace{5mm}
-
-% \emph{This section will revised.}
-
-%\vspace{5mm}
-
-This thesis contains two approaches to embedding a language for data-parallel programming 
-in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
-and, at little effort from us, very good performance is gained. The performance of 
-EmbArBB is entirely due to all the optimisation expertise and effort that was invested in 
-the ArBB system. We get all that expertise for free. What EmbArBB provides is a 
-way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB are based on a 
-set of parallel primitives for the programmer to use and compose. Obsidian takes 
-a different approach. Built in primitives are very few and very low level, such as 
-sequential and parallel for loops. On top of the low-level capabilities, a library 
-of high-level primitives can be built. The key is that we want the programmer to 
-be able to specify how to implement these higher level abstractions.
-
-Looking at the related work, we see that almost all high level GPU programming 
-languages provide abstractions of parallel primitives, aggregate operations, 
-maps, folds and reductions, giving justification to our investigation of GPU 
-programming without these. We believe that the GPU architecture is still volatile 
-enough that there is no ``the right'' way of implementing these primitives. As of 
-today, the programmer still needs to explore the design space and find a solution 
-that performs well on a particular GPU.
-
-Performance of Obsidian kernels is often good and the Obsidian code that they 
-are generated from is often shorter than corresponding CUDA programs. Obsidian 
-programs differ from CUDA programs by describing whole programs. A CUDA programmer 
-needs to take a leap from the single threaded program 
-parameterised on thread ID, to what actually will happen when this program is 
-run by hundreds or thousands of threads at once. The extensive indexing arithmetic 
-often present in CUDA programs can make this leap a hard one. Similar programs 
-in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
-bounded parallel loops.
-
-Push arrays were invented by Koen Claessen and first implemented in Obsidian 
-to provide more control over where computation takes place (which thread computes 
-which element). Push arrays has since been experimented on by others and show up 
-in Nikola~\citet{NIKOLAPUSH} and Feldspar~\citet{FELDSPARPUSH}. Push arrays appear 
-in StreamHS~\citet{FPCDSL} and in ~\citet{MOA} as part of a multi-dimensional array 
-calculus in Standard ML.
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% Thesis overview
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-\section{Thesis overview}
-
-%This thesis contains an introduction and a selection of papers. The introduction 
-%gives background and context to the papers.
-
-The papers contained in this thesis fall in three categories. 
-\begin{itemize} 
-\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
-\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
-\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
-\end{itemize} 
-%In the following sections, each paper is given a short description. Hopefully 
-%there is enough information to wet the appetite.
-
-\subsection{EDSL implementation papers}
-
-\subsubsection{\paperA: \\ \paperATitle}
-
-Authors: Josef Svenningsson and Bo Joel Svensson
-
-\vspace{5mm}
-
-\noindent This paper describes a simple and compositional method for 
-reification of monads. This is useful for compilation of 
-monadic embedded languages. We use a simple robot control language 
-as the example. Robot programs can be expressed using Haskell do notation 
-and are compiled down to a simple first order program representation. 
-There is also a graphical simulator available that runs the compiled 
-code. The robot language and simulator are available at github: \\ 
-\url{github.com/svenssonjoel/Robot}.
-
-\subsection{GPU programming papers}
-
-\subsubsection{\paperB: \\ \paperBTitle}
-
-Authors: Bo Joel Svensson, Mary Sheeran and Koen Claessen \newline
-
-\vspace{5mm}
-
-This is the first real paper about Obsidian, our embedded language for GPU 
-programming. In this paper, we identify similarities between 
-connection patterns in hardware and parallel computations on GPUs. 
-We introduce a structured and compositional way to write data-parallel 
-programs on a GPU.
-
-Basic parallel building blocks are implemented in a monadic style and 
-composed using a sequential composition operator {\tt ->-}.
-
-\subsubsection{\paperC: \\ \paperCTitle}
-
-Authors: Bo Joel Svensson, Koen Claessen and Mary Sheeran \newline
-
-\vspace{5mm}
-
-In this paper we describe a version of Obsidian implemented with a 
-a different internal representation of GPU programs. We realised that 
-most of our kernels have similar shape; they are parallel operations 
-composed in sequence interspersed with barrier synchronisations. The 
-internal representation used here captures this case exactly. Programming 
-is now done is a style resembling arrows.
-
-\subsubsection{\paperD: \\ \paperDTitle}
-
-Authors: Koen Claessen, Mary Sheeran and  Bo Joel Svensson \newline
-
-\vspace{5mm}
-
-This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
-array representation targets specific performance problems that we want 
-to solve with Obsidian. Here we have also gone back to using monadic 
-style for programming in Obsidian.
-
-\subsubsection{\paperE: \\ \paperETitle}
-
-Authors: Josef Svenningsson, Bo Joel Svensson and Mary Sheeran \newline
-
-\vspace{5mm}
-
-This paper contains a sorting case study performed in Obsidian. It 
-is however the algorithms that are in focus here. We describe 
-an interesting variation on Counting sort that removes duplicate 
-elements and is a nice fit for GPUs because of its reduced need for 
-synchronisation.
-
-In order to implement these sorting algorithms, we added atomic operations 
-to Obsidian.
-
-\subsubsection{\paperF: \\ \paperFTitle}
-
-Authors: Bo Joel Svensson and Mary Sheeran \newline
-\noindent \emph{This work has not yet been published.}
-\vspace{5mm}
-
-\noindent
-
-In this paper, the most recent additions and improvements to Obsidian are 
-described. Obsidian Programs are now parameterised on Hierarchy level. This forces
-the programmer into writing programs that we know how to compile to a GPU easily 
-(by limiting how parallel operations can be nested). The central part of this paper 
-describes how to implement and optimise reduction kernels using Obsidian.
-
-\subsection{Retargetable parallel programming}
-
-\subsubsection{\paperG: \\ \paperGTitle}
-
-Authors: Bo Joel Svensson and Ryan R. Newton
-
-\vspace{5mm}
-
-\noindent This position paper describes the work I did while on a 3 month internship at 
-Intel. The main part of this work was implementation of Haskell bindings 
-for the now retired Intel ArBB system.
-
-The paper also describes work we did towards implementing a backend for the 
-Accelerate system using ArBB. We reached a proof-of-concept implementation 
-capable of running some simple Accelerate programs using the ArBB backend.
-However, much work would remain to be able to run the full scope of Accelerate 
-programs on ArBB.
-
-\subsubsection{\paperH: \\ \paperHTitle}
-
-Authors: Bo Joel Svensson and Mary Sheeran
-
-\vspace{5mm}
-
-\noindent Here we explore another way to provide the capabilities of ArBB to the 
-Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
-we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
-We call this embedded language EmbArBB and it provides a more Haskell-programmer-friendly 
-interface compared to the raw ArBB bindings (which are a collection of functions 
-in the IO monad).
-
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-%
-% BIBLIOGRAPHY
-%
-% ---------------------------------------------------------------------------
-% ---------------------------------------------------------------------------
-
-\bibliographystylet{alpha}
-\bibliographyt{thesis}
-%\addcontentsline{toc}{section}{Bibliography}
-
-\clearpage{}
-
-% --------------------------------------------------------------------------- %
-% --------------------------------------------------------------------------- %
-%
-%  PAPERS 
-%
-% --------------------------------------------------------------------------- %
-% --------------------------------------------------------------------------- %
-
-\chapter{EDSL Implementation Papers}
-\label{chap:EDSLImplementation}
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperATitle]{\paperA: \\ \paperATitle}
-\label{sec:paperA}
-%\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
-
-% \paperATitle
-
-\begin{center} 
-Josef Svenningsson, Bo Joel Svensson
-\end{center}
-
-\input{./bb/paperThesis}
-
-\chapter{GPU Programming Papers}
-\label{chap:GPUProgramming}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperBTitle]{\paperB: \\ \paperBTitle}
-\label{sec:paperB}
-%\addcontentsline{toc}{chapter}{Obsidian: A Domain Specific Embedded Language for
-%Parallel Programming of Graphics Processors}
-
-%\paperBTitle
-
-\begin{center} 
-Bo Joel Svensson, Mary Sheeran, Koen Claessen
-\end{center}
-
-\input{./ifl/paperThesis}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperCTitle]{\paperC: \\ \paperCTitle}
-\label{sec:paperC}
-%\addcontentsline{toc}{chapter}{Obsidian: GPU Computing Using Haskell}
-
-%\paperCTitle
-
-\begin{center} 
-Bo Joel Svensson, Koen Claessen, Mary Sheeran
-\end{center}
-
-\input{./papp/paperThesis}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperDTitle]{\paperD: \\ \paperDTitle}
-\label{sec:paperD}
-%\addcontentsline{toc}{chapter}{Expressive Array Constructs in an Embedded GPU Kernel Programming Language}
-
-%\paperDTitle
-
-\begin{center} 
-Koen Claessen, Mary Sheeran, Bo Joel Svensson
-\end{center}
-
-\input{./expressive/paperThesis}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperETitle]{\paperE: \\ \paperETitle}
-\label{sec:paperE}
-%\addcontentsline{toc}{chapter}{Counting and Occurrence Sort for GPUs using an Embedded Language}
-
-% \paperETitle
-
-\begin{center} 
-Josef Svenningsson, Bo Joel Svensson, Mary Sheeran 
-\end{center}
-
-\input{./csort/paperThesis}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperFTitle]{\paperF: \\ \paperFTitle}
-\label{sec:paperF}
-
-%\addcontentsline{toc}{chapter}{Simple and Compositional Reification of Monadic Embedded Languages}
-
-% \paperFTitle
-
-\begin{center} 
-Bo Joel Svensson, Mary Sheeran
-\end{center}
-
-%\begin{center}
-%{\large \bf This paper will be revised before printing the Thesis} 
-%\end{center}
-
-\input{./hl/paperThesis}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-
-\chapter{Retargetable Parallel Programming Papers}
-\label{chap:ArBB}
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperGTitle]{\paperG: \\ \paperGTitle}
-\label{sec:paperG}
-%\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
-
-% \paperGTitle
-
-\begin{center} 
-Bo Joel Svensson, Ryan R. Newton
-\end{center}
-
-\input{./arbb/paperThesis}
-
-% ---------------------------------------------------------------------------
-% 
-% ---------------------------------------------------------------------------
-\cleardoublepage
-
-\section[\paperHTitle]{\paperH: \\ \paperHTitle}
-\label{sec:paperH}
-%\addcontentsline{toc}{chapter}{Parallel Programming in Haskell Almost for Free}
-
-% \paperHTitle
-
-\begin{center} 
-Bo Joel Svensson, Mary Sheeran
-\end{center}
-
-\input{./embarbb/paperThesis}
-
-\cleardoublepage
-
-% ---------------------------------------------------------------------------
-% nocites 
-% ---------------------------------------------------------------------------
-%% \nocite{*}
-
-%% \makeatletter
-%% \renewenvironment{thebibliography}[1]
-%%      {\chapter*{\bibname}%
-%%       \@mkboth{\MakeUppercase\bibname}{\MakeUppercase\bibname}%
-%%       \list{\@biblabel{\@arabic\c@enumiv}}%
-%%            {\settowidth\labelwidth{\@biblabel{#1}}%
-%%             \leftmargin\labelwidth
-%%             \advance\leftmargin\labelsep
-%%             \@openbib@code
-%%             \usecounter{enumiv}%
-%%             \let\p@enumiv\@empty
-%%             \renewcommand\theenumiv{\@arabic\c@enumiv}}%
-%%       \sloppy
-%%       \clubpenalty4000
-%%       \@clubpenalty \clubpenalty
-%%       \widowpenalty4000%
-%%       \sfcode`\.\@m}
-%%      {\def\@noitemerr
-%%        {\@latex@warning{Empty `thebibliography' environment}}%
-%%        \endlist}
-%% \makeatother
-
-%% \bibliographystyle{alpha}
-%% \bibliography{thesis}
-%% \addcontentsline{toc}{chapter}{Bibliography}
-
-\end{document}
+\newpage

### 2013-11-12 src2/thesis.tex
update

@@
 exist~\citet{merrill,ScanCUDA,OLAMARCUS,EFFICIENTSCAN}. In Paper F (section~\ref{sec:paperF}), 
-we only study scan kernels (scan of small arrays) and measure their speed relative to eachother. 
-Compared to a scan kernel supplied with the CUDA API, our kernel fastest kernel is slighty more 
+we only study scan kernels (scan of small arrays) and measure their speed relative to each other. 
+Compared to a scan kernel supplied with the CUDA API, our fastest kernel is slighty more 
 than three times slower. Performance needs to be improved at the kernel level before we attempt 
@@
 Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
-a reduction kernel in Obsidian. At the end, the implemented reduction kernel 
-reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwith. 
+a reduction kernel in Obsidian. Seven reductions kernels are implemented, the first 
+very simple and naive. Each following kernel, tweaks some aspect of the previous one. 
+Applying these changes is pleasingly simple using Obsidian; some of these changes 
+would mean a major rewrite in CUDA and some are even impossible. 
+The final kernel reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwith. 
+At this point, we have pushed Obsidian as far as possible with the current abstractions. 
+However, it is possible that with more work on warp level capabilites, this could be 
+improved a bit further.   
 %faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
@@
 Paper D (section~\ref{sec:paperD}) introduces push arrays and uses them in the 
-implementation of sorting networks, with quite good results. With push arrays we 
+implementation of sorting networks, with quite good results. With push arrays, we 
 were able to experiment with more detailed decompositions (what thread computes 
@@
 % ---------------------------------------------------------------------------
-\section{Thesis overview}
-
-%This thesis contains an introduction and a selection of papers. The introduction 
-%gives background and context to the papers.
-
-The papers contained in this thesis fall in three categories. 
-\begin{itemize} 
-\item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
-\item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
-\item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
-\end{itemize} 
-%In the following sections, each paper is given a short description. Hopefully 
-%there is enough information to wet the appetite.
-
-\subsection{EDSL implementation papers}
-
-\subsubsection{\paperA: \\ \paperATitle}
-
-Authors: Josef Svenningsson and Bo Joel Svensson
-
-\vspace{5mm}
-
-\noindent This paper describes a simple and compositional method for 
-reification of monads. This is useful for compilation of 
-monadic embedded languages. We use a simple robot control language 
-as the example. Robot programs can be expressed using Haskell do notation 
-and are compiled down to a simple first order program representation. 
-There is also a graphical simulator available that runs the compiled 
-code. The robot language and simulator are available at github: \\ 
-\url{github.com/svenssonjoel/Robot}.
-
-\subsection{GPU programming papers}
-
-\subsubsection{\paperB: \\ \paperBTitle}
-
-Authors: Bo Joel Svensson, Mary Sheeran and Koen Claessen \newline
-
-\vspace{5mm}
-
-This is the first real paper about Obsidian, our embedded language for GPU 
-programming. In this paper, we identify similarities between 
-connection patterns in hardware and parallel computations on GPUs. 
-We introduce a structured and compositional way to write data-parallel 
-programs on a GPU.
-
-Basic parallel building blocks are implemented in a monadic style and 
-composed using a sequential composition operator {\tt ->-}.
-
-\subsubsection{\paperC: \\ \paperCTitle}
-
-Authors: Bo Joel Svensson, Koen Claessen and Mary Sheeran \newline
-
-\vspace{5mm}
-
-In this paper we describe a version of Obsidian implemented with a 
-a different internal representation of GPU programs. We realised that 
-most of our kernels have similar shape; they are parallel operations 
-composed in sequence interspersed with barrier synchronisations. The 
-internal representation used here captures this case exactly. Programming 
-is now done is a style resembling arrows.
-
-\subsubsection{\paperD: \\ \paperDTitle}
-
-Authors: Koen Claessen, Mary Sheeran and  Bo Joel Svensson \newline
-
-\vspace{5mm}
-
-This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
-array representation targets specific performance problems that we want 
-to solve with Obsidian. Here we have also gone back to using monadic 
-style for programming in Obsidian.
-
-\subsubsection{\paperE: \\ \paperETitle}
-
-Authors: Josef Svenningsson, Bo Joel Svensson and Mary Sheeran \newline
-
-\vspace{5mm}
-
-This paper contains a sorting case study performed in Obsidian. It 
-is however the algorithms that are in focus here. We describe 
-an interesting variation on Counting sort that removes duplicate 
-elements and is a nice fit for GPUs because of its reduced need for 
-synchronisation.
-
-In order to implement these sorting algorithms, we added atomic operations 
-to Obsidian.
-
-\subsubsection{\paperF: \\ \paperFTitle}
-
-Authors: Bo Joel Svensson and Mary Sheeran \newline
-\noindent \emph{This work has not yet been published.}
-\vspace{5mm}
-
-\noindent
-
-In this paper, the most recent additions and improvements to Obsidian are 
-described. Obsidian Programs are now parameterised on Hierarchy level. This forces
-the programmer into writing programs that we know how to compile to a GPU easily 
-(by limiting how parallel operations can be nested). The central part of this paper 
-describes how to implement and optimise reduction kernels using Obsidian.
-
-\subsection{Retargetable parallel programming}
-
-\subsubsection{\paperG: \\ \paperGTitle}
-
-Authors: Bo Joel Svensson and Ryan R. Newton
-
-\vspace{5mm}
-
-\noindent This position paper describes the work I did while on a 3 month internship at 
-Intel. The main part of this work was implementation of Haskell bindings 
-for the now retired Intel ArBB system.
-
-The paper also describes work we did towards implementing a backend for the 
-Accelerate system using ArBB. We reached a proof-of-concept implementation 
-capable of running some simple Accelerate programs using the ArBB backend.
-However, much work would remain to be able to run the full scope of Accelerate 
-programs on ArBB.
-
-\subsubsection{\paperH: \\ \paperHTitle}
-
-Authors: Bo Joel Svensson and Mary Sheeran
-
-\vspace{5mm}
-
-\noindent Here we explore another way to provide the capabilities of ArBB to the 
-Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
-we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
-We call this embedded language EmbArBB and it provides a more Haskell-programmer-friendly 
-interface compared to the raw ArBB bindings (which are a collection of functions 
-in the IO monad).
+%% \section{Thesis overview}
+
+%% %This thesis contains an introduction and a selection of papers. The introduction 
+%% %gives background and context to the papers.
+
+%% The papers contained in this thesis fall in three categories. 
+%% \begin{itemize} 
+%% \item General EDSL methodology in chapter~\ref{chap:EDSLImplementation}. 
+%% \item GPU Programming using an EDSL in chapter~\ref{chap:GPUProgramming}. 
+%% \item Retargetable parallel programming in chapter~\ref{chap:ArBB}. 
+%% \end{itemize} 
+%% %In the following sections, each paper is given a short description. Hopefully 
+%% %there is enough information to wet the appetite.
+
+%% \subsection{EDSL implementation papers}
+
+%% \subsubsection{\paperA: \\ \paperATitle}
+
+%% Authors: Josef Svenningsson and Bo Joel Svensson
+
+%% \vspace{5mm}
+
+%% \noindent This paper describes a simple and compositional method for 
+%% reification of monads. This is useful for compilation of 
+%% monadic embedded languages. We use a simple robot control language 
+%% as the example. Robot programs can be expressed using Haskell do notation 
+%% and are compiled down to a simple first order program representation. 
+%% There is also a graphical simulator available that runs the compiled 
+%% code. The robot language and simulator are available at github: \\ 
+%% \url{github.com/svenssonjoel/Robot}.
+
+%% \subsection{GPU programming papers}
+
+%% \subsubsection{\paperB: \\ \paperBTitle}
+
+%% Authors: Bo Joel Svensson, Mary Sheeran and Koen Claessen \newline
+
+%% \vspace{5mm}
+
+%% This is the first real paper about Obsidian, our embedded language for GPU 
+%% programming. In this paper, we identify similarities between 
+%% connection patterns in hardware and parallel computations on GPUs. 
+%% We introduce a structured and compositional way to write data-parallel 
+%% programs on a GPU.
+
+%% Basic parallel building blocks are implemented in a monadic style and 
+%% composed using a sequential composition operator {\tt ->-}.
+
+%% \subsubsection{\paperC: \\ \paperCTitle}
+
+%% Authors: Bo Joel Svensson, Koen Claessen and Mary Sheeran \newline
+
+%% \vspace{5mm}
+
+%% In this paper we describe a version of Obsidian implemented with a 
+%% a different internal representation of GPU programs. We realised that 
+%% most of our kernels have similar shape; they are parallel operations 
+%% composed in sequence interspersed with barrier synchronisations. The 
+%% internal representation used here captures this case exactly. Programming 
+%% is now done is a style resembling arrows.
+
+%% \subsubsection{\paperD: \\ \paperDTitle}
+
+%% Authors: Koen Claessen, Mary Sheeran and  Bo Joel Svensson \newline
+
+%% \vspace{5mm}
+
+%% This paper introduces {\em Push} arrays, invented by Koen Claessen. This 
+%% array representation targets specific performance problems that we want 
+%% to solve with Obsidian. Here we have also gone back to using monadic 
+%% style for programming in Obsidian.
+
+%% \subsubsection{\paperE: \\ \paperETitle}
+
+%% Authors: Josef Svenningsson, Bo Joel Svensson and Mary Sheeran \newline
+
+%% \vspace{5mm}
+
+%% This paper contains a sorting case study performed in Obsidian. It 
+%% is however the algorithms that are in focus here. We describe 
+%% an interesting variation on Counting sort that removes duplicate 
+%% elements and is a nice fit for GPUs because of its reduced need for 
+%% synchronisation.
+
+%% In order to implement these sorting algorithms, we added atomic operations 
+%% to Obsidian.
+
+%% \subsubsection{\paperF: \\ \paperFTitle}
+
+%% Authors: Bo Joel Svensson and Mary Sheeran \newline
+%% \noindent \emph{This work has not yet been published.}
+%% \vspace{5mm}
+
+%% \noindent
+
+%% In this paper, the most recent additions and improvements to Obsidian are 
+%% described. Obsidian Programs are now parameterised on Hierarchy level. This forces
+%% the programmer into writing programs that we know how to compile to a GPU easily 
+%% (by limiting how parallel operations can be nested). The central part of this paper 
+%% describes how to implement and optimise reduction kernels using Obsidian.
+
+%% \subsection{Retargetable parallel programming}
+
+%% \subsubsection{\paperG: \\ \paperGTitle}
+
+%% Authors: Bo Joel Svensson and Ryan R. Newton
+
+%% \vspace{5mm}
+
+%% \noindent This position paper describes the work I did while on a 3 month internship at 
+%% Intel. The main part of this work was implementation of Haskell bindings 
+%% for the now retired Intel ArBB system.
+
+%% The paper also describes work we did towards implementing a backend for the 
+%% Accelerate system using ArBB. We reached a proof-of-concept implementation 
+%% capable of running some simple Accelerate programs using the ArBB backend.
+%% However, much work would remain to be able to run the full scope of Accelerate 
+%% programs on ArBB.
+
+%% \subsubsection{\paperH: \\ \paperHTitle}
+
+%% Authors: Bo Joel Svensson and Mary Sheeran
+
+%% \vspace{5mm}
+
+%% \noindent Here we explore another way to provide the capabilities of ArBB to the 
+%% Haskell programmer. Rather than trying to use ArBB as a backend to Accelerate 
+%% we provide a more direct mapping of ArBB's functionality to Haskell idioms. 
+%% We call this embedded language EmbArBB and it provides a more Haskell-programmer-friendly 
+%% interface compared to the raw ArBB bindings (which are a collection of functions 
+%% in the IO monad).
 

### 2013-11-12 src2/thesis.tex
spell pass

@@
 we can generate CUDA kernels for any element type that supports the {\tt (+)} operation. 
-In Haskell we know that types that belong to the {\tt Num} typeclass has a {\tt (+)} operation.
+In Haskell we know that types that belong to the {\tt Num} type class has a {\tt (+)} operation.
 
@@
 we only study scan kernels (scan of small arrays) and measure their speed relative to each other. 
-Compared to a scan kernel supplied with the CUDA API, our fastest kernel is slighty more 
+Compared to a scan kernel supplied with the CUDA API, our fastest kernel is slightly more 
 than three times slower. Performance needs to be improved at the kernel level before we attempt 
@@
 Paper F (section~\ref{sec:paperF}) tells a story about how to implement and tune 
-a reduction kernel in Obsidian. Seven reductions kernels are implemented, the first 
-very simple and naive. Each following kernel, tweaks some aspect of the previous one. 
+a reduction kernel in Obsidian. Seven reduction kernels are implemented, the first 
+very simple and naive. Each following kernel tweaks some aspect of the previous one. 
 Applying these changes is pleasingly simple using Obsidian; some of these changes 
-would mean a major rewrite in CUDA and some are even impossible. 
-The final kernel reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwith. 
-At this point, we have pushed Obsidian as far as possible with the current abstractions. 
-However, it is possible that with more work on warp level capabilites, this could be 
-improved a bit further.   
+would mean a major rewrite in CUDA. The final kernel is almost 5 times as fast as the first, 
+and reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwidth. The speed of any 
+kernel will always be limited by how fast the data can be pushed through the GPU; this performance 
+is quite satisfactory. In Paper F, a large reduction algorithm based on this kernel is 
+implemented that runs slightly faster than Accelerates fold skeleton. With this reduction 
+case study, we have pushed Obsidian as far as possible with the current abstractions. 
+However, it is possible that with more work on the warp level abstractions, this could be 
+improved further.   
 %faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
@@
 the computation amongst threads and blocks. Using Accelerate~\citet{ACCELERATEDAMP11}, 
-Nikola~\citet{NIKOLA} 
-and Thrust~\citet{THRUST}, the programmer uses 
-high-level patterns, without direct insight into how the application is mapped onto 
-the threads and blocks of the GPU. Using Obsidian, we want to bridge that gap by both being 
-more composable than CUDA and giving the programmer control of what the computation will 
-look like on the GPU. This is done by allowing the programmer to specify and compose 
-thread-level and block-level code. Recently, the CUB library arrived that is built on a 
-similar idea, that GPU programmers need to have block/thread level control to get maximum 
-performance, while maintaining a higher level and more Thrust-like programmer 
-interface~\citet{CUB}. Before CUB, and as far as I am aware, we were alone at targeting 
-this level.
+Nikola~\citet{NIKOLA} and Thrust~\citet{THRUST}, the programmer uses high-level patterns, 
+without direct insight into how the application is mapped onto the threads and blocks of 
+the GPU. Using Obsidian, we want to bridge that gap by both being more composable than 
+CUDA and giving the programmer control of what the computation will look like on the GPU. 
+This is done by allowing the programmer to specify and compose thread-level and block-level 
+code. Recently, the CUB library arrived that is built on a similar idea, that GPU programmers 
+need to have block/thread level control to get maximum performance, while maintaining a 
+higher level and more Thrust-like programmer interface~\citet{CUB}. Before CUB, and as 
+far as I am aware, we were alone at targeting this level.
+
+An in depth comparison of Obsidian and Accelerate is hard to perform. The reason for 
+this is that Obsidian is a language for kernels, while Accelerate is a language for 
+applications. Any larger GPU application consists of a number of kernels that 
+are cooperating. Obsidian only has very basic support for launching and coordinating 
+kernels for such applications, from within Haskell and Accelerate is limited in specifying 
+new specialised kernels.
 

### 2013-11-12 src2/thesis.tex
update

@@
 would mean a major rewrite in CUDA. The final kernel is almost 5 times as fast as the first, 
-and reaches a throughput of 140GB/sec on a GPU with 192GB/sec memory bandwidth. The speed of any 
+and reaches a throughput of 140GB/s on a GPU with 192GB/s memory bandwidth. The speed of any 
 kernel will always be limited by how fast the data can be pushed through the GPU; this performance 
@@
 However, it is possible that with more work on the warp level abstractions, this could be 
-improved further.   
+improved further. Our reduction case study was based on lecture slides from 
+NVIDIA~\citet{reducion}. Reference~\citet{merrill}, has much information about 
+how to think about performance when programming GPUs. 
 %faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
@@
 are cooperating. Obsidian only has very basic support for launching and coordinating 
-kernels for such applications, from within Haskell and Accelerate is limited in specifying 
+kernels for such applications from within Haskell, and Accelerate is limited in specifying 
 new specialised kernels.

### 2013-11-12 src2/embarbb/programming.tex
hacking
update

@@
 %nesting descriptor. 
-% BJS: I have seen this elsewhere but cannot find it! (wanted to \cite something)
+% BJS: I have seen this elsewhere but cannot find it! (wanted to \citeemb something)
 %The ArBB system offers the programmer {\em dense} and {\em nested} vectors 
@@
 only occur the first time a function is called. The compilation procedure
-is outlined in~\cite{ARBB2011}.
+is outlined in~\citeemb{ARBB2011}.
 
@@
 A function that adds a constant to all elements of a vector is
+
+\pagebreak
 
@@
 for any captured functions or vectors created in the ArBB monad.
+
+\pagebreak
 
@@
 
-% \FloatBarrier 
+\FloatBarrier 
 \subsubsection{Saxpy} 
@@
 
-\begin{verbatim} 
+\begin{Verbatim}[samepage=true]
 saxpy :: Num a 
@@
       ss = constVector (length x) s
-\end{verbatim}
+\end{Verbatim}
 
@@
 
-\begin{verbatim}
-main = 
-  withArBB $ 
+\begin{Verbatim}[samepage=true]
+main = withArBB $ 
   do 
@@
      liftIO$ putStrLn$ show r
-\end{verbatim}
+\end{Verbatim}
+
+% $
 
@@
 
-\begin{verbatim}
-main =  
-  withArBB $ 
+\begin{Verbatim}[samepage=true]
+main = withArBB $ 
   do 
@@
      liftIO$ putStrLn$ show r
-\end{verbatim}
+\end{Verbatim}
 %$ -- fixes syntax highlighting
@@
 
-\begin{verbatim}
+\begin{Verbatim}[samepage=true]
 histImage :: Exp (Vector Word32) 
@@
       white = 255
-\end{verbatim}
+\end{Verbatim}
+
+% $
 
@@
 to get parallelism both in the individual summations and between summations.
-This is a well known algorithm that can be traced back to Blelloch~\cite{NESL}.
+This is a well known algorithm that can be traced back to Blelloch~\citeemb{NESL}.
 

### 2013-11-12 src2/embarbb/results.tex
update

@@
 part is defined as follows 
-\begin{verbatim}
+\begin{Verbatim}[samepage=true]
 gradientX :: Monad m => Image -> m Image
@@
                          -1  0  1 |]
-
+\end{Verbatim}
+\begin{Verbatim}[samepage=true]
 gradientY :: Monad m => Image -> m Image
@@
                         -1 -2 -1 |] 
-\end{verbatim}
+\end{Verbatim}
 

### 2013-11-12 src2/ifl/related.tex
update

@@
 a combination of a high degree of abstraction and a close resemblance to traditional programming languages~\citeifl{Lucas08}.
-BrookGPU (which is usually just called Brook) is a classic example of a language~\citeifl{Brook} designed to raise the level of abstraction at which GPGPU programming is done.
+BrookGPU (which is usually just called Brook) is an example of a language~\citeifl{Brook} designed to raise the level of abstraction at which GPGPU programming is done.
 It is an extension of C with embedded kernels, aimed at arithmetic-intense data parallel computations.

### 2013-11-12 src2/thesis.tex
update

@@
 can be used as a powerful macro language for the embedded language, providing a more 
-leveled learning curve to programmers already familiar with the host language. 
+levelled learning curve to programmers already familiar with the host language. 
 For programmers who are already proficient in the host language, learning an embedded 
@@
 used to map the local computation onto each chunk of the input arrays. The result is an 
-array of arrays that is flattened with the {\tt pConcat} function. The CUDA code below 
-is generated from the {\tt vecAddG} program:
+array of arrays that is flattened with the {\tt pConcat} function.
+
+\pagebreak
+
+\noindent The CUDA code below is generated from the {\tt vecAddG} program:
 
@@
 \begin{verbatim} 
-map f (Push s p) = Push s $ \wf -> p (\e ix -> wf (f e) ix)
+map f (Push s p) = 
+  Push s $ \wf -> p (\e ix -> wf (f e) ix)
 \end{verbatim}
@@
 and reaches a throughput of 140GB/s on a GPU with 192GB/s memory bandwidth. The speed of any 
-kernel will always be limited by how fast the data can be pushed through the GPU; this performance 
-is quite satisfactory. In Paper F, a large reduction algorithm based on this kernel is 
-implemented that runs slightly faster than Accelerates fold skeleton. With this reduction 
-case study, we have pushed Obsidian as far as possible with the current abstractions. 
-However, it is possible that with more work on the warp level abstractions, this could be 
-improved further. Our reduction case study was based on lecture slides from 
-NVIDIA~\citet{reducion}. Reference~\citet{merrill}, has much information about 
-how to think about performance when programming GPUs. 
+kernel will always be limited by how fast the data can be pushed through the GPU
+(see reference~\citet{merrill} for an in depth discussion of kernel performance); this performance 
+is very satisfactory. In Paper F, a large reduction algorithm based on this kernel is 
+implemented that runs slightly faster than Accelerate's fold skeleton. Even being just nearly 
+as fast as Accelerate would in this case have been fine, since the fold skeleton is hand-tuned 
+CUDA code. With this reduction case study, we have pushed Obsidian as far as possible 
+with the current abstractions. However, it is possible that with more work on the warp 
+level abstractions, this could be improved further. Our reduction case study was based 
+on lecture slides from NVIDIA~\citet{reduction}.
 %faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
@@
 ArBB is based on a set of parallel primitives (or structures) on dense one, two and three dimensional vectors and nested vectors: 
+\begin{small}
 \begin{itemize} 
@@
 \end{itemize}
-
-These operations (and many more, there are about 120 operations in total) can be used 
+\end{small}
+
+\noindent These operations (and many more, there are about 120 operations in total) can be used 
 by the programmer when writing programs using ArBB. Note that there are sequential looping 

### 2013-11-12 src2/embarbb/programming.tex
update

@@
 steps are necessary before it can be run on real data. The code 
-below shows how to set up a 4x4 scaling matrix and a vector to 
+in figure~\ref{fig:EMB_ARBBFIG} shows how to set up a 4x4 scaling matrix and a vector to 
 multiply by that matrix using ArBB. This entails allocating 
@@
 device that is common in general purpose GPU (GPGPU) programming . 
-
-%[samepage=true] 
-\begin{Verbatim} 
+\begin{figure}
+\begin{small}
+\begin{Verbatim} [samepage=true] 
 int main(void)
@@
 \end{Verbatim}
+\end{small}
+
+\caption{ArBB glue code.}
+\label{fig:EMB_ARBBFIG}
+
+\end{figure}
 

### 2013-11-13 src2/arbb/paperThesis.tex
updated writing
update

@@
 Our first \systemname{} prototype adapts an existing EDSL called {\em
-  Accelerate} (Data.Array.Accelerate).  Accelerate targets high-level
+  Accelerate}.  Accelerate targets high-level
 data-parallel programming in Haskell.
@@
 aspirations as ArBB, that is, to target a diverse range of architectures using the 
-same source code. Accelerate can be used from the C\# language or the 
+same source code. Accelerator can be used from the C\# language or the 
 functional F\# language and targets GPUs or of CPUs and their
@@
 \bibliographyarbb{thesis}
+\addcontentsline{toc}{subsection}{Bibliography}
 %}
